### 2013-12-27 paper/draft1.tex
Added files related to paper

\documentclass[twocolumn]{revtex4}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{float}
%\usepackage{forloop}
\usepackage{tikz}
\usepackage{mhchem}

\begin{document}

\title{Intrinsic Variables and State Reconstruction in Multiscale Simulations}

%\author{}
%\email{}
%\affiliation{}

\date{\today}

\begin{abstract}
Finding informative low-dimensional descriptions of high dimensional simulation data
(like the ones arising in Molecular Dynamics or kinetic Monte Carlo simulations of
physical/chemical processes) is crucial in understanding physical phenomena, and can
also dramatically assist in accelerating the simulations themselves.
%
In this paper we discuss and illustrate the use of nonlinear intrinsic variables (NIVs)
in the mining of high-dimensional multiscale simulation data.
%
In particular we focus on the way that NIVs allow us to functionally merge different
simulation ensembles, and {\em different partial observations of these ensembles}, as well
as to infer variables not explicitly observed/measured.
%
The approach relies on certain simple features of the underlying process variability to
filter out measurement noise and systematically recover a unique reference coordinate frame.
%
We illustrate the approach through two distinct sets of multiscale atomistic simulations: a fast/slow
time scale Stochastic Simulation Algorithm simulation of enzyme kinetics, and a molecular
dynamics simulation of alanine dipeptide in explicit water.

\end{abstract}

\keywords{Intrinsic variables, partial observations, complex dynamical systems}

\maketitle

\section{Introduction}
The last decade has witnessed extensive advances in dimensionality reduction techniques:
finding meaningful low-dimensional descriptions of high-dimensional data.
%
These developments have the potential to significantly enable the computational exploration
of phyiscochemical problems.
%
If the (high-dimensional) data $Y(t)$ arise from, say, a
molecular dynamics simulation of a macromolecule in solution, or from the stochastic
simulation of a complex chemical reaction scheme, the detection of a few good, coarse-grained
``reduction coordinates" $x(t)$ is invaluable in understanding/predicting system behavior.
%
enhancing, acceleration...

%
While the benefits from such reduced descriptions are manifest, a crucial shortcoming of
such data-driven reduction coordinates is their dependence on the specific data set processed,
and not only on the physical model in question.
%
It is well known that, even in the simple linear case of Principal Component Analysis,
different data sets on the same low-dimensional hyperplane in the ambient space $Y$
will lead to different basis vectors spanning it - in effect, to different $x$.
%
And while this can be easily fixed by a linear transformation, the problem becomes
exacerbated when the low-dimensional space is curved (a manifold, rather than a hyperplane)
and when different data ensembles are obtained using different instrumental modalities
(say, when one wants to merge molecular dynamics data with spectral information).
%
Clearly, the ability to systematically construct a {\em unique, consistent} reduction
coordinate set, shared by all different measurement ensembles and observation modalities -
what we call a (Nonlinear) Intrinsic Variable Set, is invaluable.
%
Embedding in such a coordinate set allows us to naturally merge different observations of the same system;
more importantly, it enables the construction of an empirical mapping between these different
observation ensembles, allowing us to complete partial measurements through the construction
of nonlinear observers of unmeasured variables.
%
This latter empirical mapping requires good interpolation tools in embedding space; to this
end we will demonstrate here the use of a multiscale Laplacian Pyramid approach \cite{rabin2012heterogeneous}.
%
We chose two distinct multiscale illustrative examples. The first is a Gillespie Stochastic Simulation
Algorithm (SSA)  of two Goldbeter-Koshland modules in an enzyme kinetics model \cite{zagaris2012stability}; separation of time scales is known,
in certain parameter regimes, to reduce this model to a two-dimensional effective description.
The second example is the molecular dynamics
simulation (in explicit water) of a simple peptide fragment (alanine dipeptide) whose folding
dynamics are known to be described through a small set of physical observables (two dihedral angles).
%
The paper is structured as follows: in Section \ref{sec:NIV} we present Nonlinear Intrinsic Variable formulation and
the associated inference method. Section \ref{sec:LapPyr} contains our discussion of Laplacian Pyramids that
will be used for the completion of partial observations. Section \ref{sec:examples} contain the results
of applying the approach to simulation data from our two illustrative examples. We conclude with
a summary and discussion of open issues in Section \ref{sec:conclusions}.

\section{Nonlinear Intrinsic Variables} \label{sec:NIV}

\begin{figure*}[ht]
\includegraphics[width=0.7\textwidth]{IntrinsicEmbeddingIllustration2.pdf}
\caption{Illustration of the nonlinear embedding that yields an intrinsic representation independent of measurement modality. (Bottom) The underlying variables in which the diffusion parameters are independent with unit variance. The circle illustrates the end points of the short bursts that create a sphere around the initial point on the manifold. (Top Lft) The first set of observed variables. The ellipse illustrates the end points of the short bursts as observed via the first observation modality. (Top right) The second set of observed variables. The ellipse illustrates the end points of the short bursts as observed via the second observation modality.}
\end{figure*}

{\bf To do:} \\
Block diagram/algorithm

\section{Laplacian Pyramids} \label{sec:LapPyr}

{\bf To do:} \\
A flow chart/block diagram of the method (possibly with the formulas).\\
Maybe an illustration of the multiscale reconstruction of functions.

\begin{figure*}[ht]
\includegraphics[width=0.7\textwidth]{LapPyr_illustration.pdf}
\caption{Illustration of overall methodology}
\end{figure*}

\section{Models and Results} \label{sec:examples}

\subsection{Chemical Reaction Network}

We first consider a chemical reaction network of multiple enzyme-substrate interactions \cite{zagaris2012stability}.
%
The equations that govern the network are\\
\begin{center}
\ce{$E$ + $S$ <=>[e_1][e_{-1}] $E:S$ ->[e_2][] $E$ + $S^{*}$}\\
\ce{$S$ + $E$ <=>[b_1][b_{-1}] $S:E$ ->[b_2][] $S$ + $E^{*}$}\\
\ce{$D$ + $S^{*}$ <=>[d_1][d_{-1}] $D:S^{*}$ ->[d_2][] $D$ + $S$}\\
\ce{$F$ + $E^{*}$ <=>[f_1][f_{-1}] $F:E^{*}$ ->[f_2][] $F$ + $E$}\\
\end{center}

The $^{*}$ denote activated forms of the various species, and the : denote complexes formed between two species; the complexes $E:S$ and $S:E$ are not equivalent.

There are 10 species in this reaction system.
%
However, one can write four conservation equations (since total $E$, $S$, $D$, and $F$ are all conserved) to reduce the system to 6 dimensions.
%
We assume each reaction follows an elementary rate law.
%
We use the parameters $b_1=5$, $d_1=0.0009$, $e_1=0.1$, $f_1=0.1$, $b_{-1} = 10.6$, $d_{-1}=0.05$, $e_{-1}=0.5$, $f_{-1} =0.01$, $b_2=0.4$, $d_2=0.85$, $e_2=0.05$, and $f_2=2$.
%
We take $S_T=E_T=D_T=1$, and $F_T=0.02$, where $S_T$, $E_T$, $D_T$, and $F_T$ are total concentrations of $S$, $E$, $D$, and $F$, respectively.
%
In this parameter regime, the system exhibits a separation of timescales, so that initial conditions quickly approach a two-dimensional manifold.
%
%The relevant timescales around the steady state ($-1/\lambda_i$, where $\lambda_i$ are the eigenvalues of the Hessian) are 1176, 9.731, 1.594, 1.111, 0.4975, 0.06498.

Although chemical reaction systems are typically thought of as governed by a system of ODEs, the ODEs are only an approximation that holds in the limit of a large number of molecules.
%
When the number of molecules is small, the system is inherently stochastic and its dynamics can be simulated using the Gillespie Stochastic Simulation Algorithm (SSA) \cite{gillespie1977exact}.
%
We can control the level of noise in the system by adjusting the volume $V$ (and therefore adjusting the number of molecules); in our simulations, we take $V=10^5$.
%
The volume is small enough so that we can still see appreciable stochasticity in small simulation bursts, but large enough so that the underlying two-dimensional manifold is (relatively) smooth.

We generate 3000 random initial conditions (enforcing that all concentrations must be non-negative),
and then integrate each point forward for 10 time units using the Gillespie algorithm.
%
10 time units is sufficiently long for the initial points to converge to the two-dimensional manifold,
but not long enough for the points to converge to the final steady state or to a one-dimensional curve.
%
We take these final points to be representative of points on the manifold; three-dimensional projections of the resulting manifold are shown in Figure \ref{fig:rxn_manifolds}.
%
From each manifold point, we run 20 short simulations, each for 0.2 time units.
%
We then use these short simulation ``bursts'' to estimate the local covariances for NIV.

\begin{figure*}[ht]
  \includegraphics[height=6cm]{rxn_figures/rxn_manifold1.jpg}\llap{\raisebox{3.5cm}{\includegraphics[height=2.5cm]{rxn_figures/rxn_manifold1_2.jpg}}}
  \includegraphics[height=6cm]{rxn_figures/rxn_manifold2.jpg}\llap{\raisebox{3.5cm}{\includegraphics[height=2.5cm]{rxn_figures/rxn_manifold2_2.jpg}}} \\
  \includegraphics[height=6cm]{rxn_figures/rxn_manifold3.jpg}\llap{\raisebox{3.5cm}{\includegraphics[height=2.5cm]{rxn_figures/rxn_manifold3_2.jpg}}}
  \includegraphics[height=6cm]{rxn_figures/rxn_manifold4.jpg}\llap{\raisebox{3.5cm}{\includegraphics[height=2.5cm]{rxn_figures/rxn_manifold4_2.jpg}}} \\
    \caption{Projections of the manifold from stochastic simulation of the chemical reaction network. The insets show rotations of the projections to illustrate the two-dimensionality of the manifold.}
    \label{fig:rxn_manifolds}
\end{figure*}

We consider two different data sets from our simulations.
%
In data set 1, we only observe components $E$, $S$, $S:E$, and $F:E^{*}$; in data set 2, we only observe components $E$, $S$, $E:S$, and $D:S^{*}$.
%
We wish to use NIV together with Laplacian Pyramids to predict the values of $S:E$ and $F:E^{*}$ in data set 2.

Figure \ref{fig:rxn_embedding} shows the two-dimensional NIV embeddings for the two different data sets; the embeddings appear very consistent.
%
In addition, the two data sets contain 500 common data points.
%
We use these data points to compute the correlation between the embeddings for data set 1 and data set 2.
%
We obtain a correlation of 0.97 and 0.95 between the two data sets for the first and second NIVs, respectively, indicating that the two embeddings are in excellent agreement with each other.

\begin{figure*}[ht]
    \includegraphics[width=0.4\textwidth]{rxn_figures/rxn_NLICA1.jpg}
    \includegraphics[width=0.4\textwidth]{rxn_figures/rxn_NLICA2.jpg}
    \includegraphics[width=0.4\textwidth]{rxn_figures/rxn_NLICA_corr1.jpg}
    \includegraphics[width=0.4\textwidth]{rxn_figures/rxn_NLICA_corr2.jpg}
    \caption{(Top, left) NIV embedding obtained from data set 1 (observations of components E, S, S:E, and F:E$^{*}$), colored by $S$. (Top, right) NIV embedding obtained from data set 2 (observations of components E, S, E:S, and D:S$^{*}$), colored by $S$. (Bottom, left) Correlation of first NIV between two different embeddings (correlation=0.97). (Bottom, right)  Correlation of second NIV between two different embeddings (correlation=0.95).}
    \label{fig:rxn_embedding}
\end{figure*}

We then use Laplacian Pyramids to predict $S:E$ and $F:E^{*}$ in data set 2.
%
Because data set 1 and data set 2 are measured for different components, there is no simple way to estimate $S:E$ and $F:E^{*}$ from the two data sets in the observation space.
%
Instead, we must first project the data into the NIV space so that we can compute neighbors {\em between} the two data sets.
%
We use data set 1 to train a Laplacian Pyramids function from the two-dimensional NIV embedding to $S:E$ and $F:E^{*}$.
%
We then use this function to predict the values  of $S:E$ and $F:E^{*}$ for data set 2, using the computed NIV embedding for data set 2.
%
In this way, we are exploiting the fact that the NIV embedding is intrinsic and consistent between the two data sets, even though the two data sets contain measurements of different components.
%
The results of the Laplacian Pyramids prediction are seen in Figure \ref{fig:rxn_recon}.
%
The MSE between the true and predicted values is .....

\begin{figure*}[ht]
    \includegraphics[width=0.4\textwidth]{rxn_figures/chm4_from_Em.png}
    \includegraphics[width=0.4\textwidth]{rxn_figures/chm6_from_Em.png}
    \caption{Laplacian Pyramids (LapPyr) and nearest neighbor (nn) reconstructions of $S:E$ (left) and $F:E^{*}$ (right).}
    \label{fig:rxn_recon}
\end{figure*}

\subsection{Alanine Dipeptide}

Our second example comes from a simulation of a small peptide fragment.
%
Alanine dipeptide (Ala2) is often used as a ``prototypical'' protein for simulation studies.
%
We simulate the motion of Ala2 in explicit solvent using the AMBER 10 molecular simulation package \cite{case2008Amber} with an
optimized version \cite{best2009optimized} of the AMBER ff03 force field \cite{duan2003point}.
%
The molecule was solvated with 638 TIP3P water molecules \cite{jorgensen1983comparison}, with the particle mesh Ewald method used for long-range electrostatic interactions \cite{essmann1995smooth} and periodic boundary conditions.
%
The simulation is performed at constant volume and temperature (the NVT ensemble), with the temperature being maintained at 300~K with a Langevin thermostat \cite{loncharich1992langevin}.
%
Hydrogen bonds were fixed using the SHAKE algorithm \cite{ryckaert1977numerical}.
%
The two dihedral angles $\phi$ and $\psi$ are known to parameterize the free energy surface, and the free energy surface contains three minima (labeled A, B, and C, see Figure \ref{fig:ala_fes}).
%
Our simulations were concentrated around minimum B in the free energy surface, located at $\phi=-65^{\circ}$, $\psi=150^{\circ}$.
%
We start many simulations at $10^{\circ}$ away from the minimum, and allow the simulations to each run for 0.1~ps; we record the configuration every 1~fs.
%
Configurations are stored with all atoms except the hydrogens.

\begin{figure*}[ht]
    \includegraphics[width=0.6\textwidth]{FES.pdf}
    \includegraphics[width=0.3\textwidth]{molecule_labeled2.jpg}
    \caption{(Left) Free energy surface for Alanine Dipeptide. The minima are labeled A, B, and C, and the corresponding molecular configurations are shown. (Right) The molecular structure of alanine dipeptide, excluding the hydrogens. The atoms are numbered and the two dihedral angles $\phi$ and $\psi$ are indicated.}
    \label{fig:ala_fes}
\end{figure*}

We first wish to demonstrate the benefit of NIV over other nonlinear dimensionality reduction techniques, such as diffusion maps.
%
We compute the NIV embedding for 10,000 data points using only atoms 2, 4, 6, 8, and 10.
%
We then compute the NIV embedding for the same 10,000 data points using only atoms 5, 7, and 9.
%
We also do the same using diffusion maps instead of NIV.
%
The correlation between the NIV coordinates for the two data sets and the diffusion maps coordinates for the two data sets are shown in Figure \ref{fig:ala_corr}.
%
The correlation between the two NIV embeddings is higher than the correlation between the two diffusion maps embeddings.

\begin{figure*}[ht]
    \includegraphics[width=0.4\textwidth]{ala_figures_evenodd/NLICA_corr.jpg}
    \includegraphics[width=0.4\textwidth]{ala_figures_evenodd/DM_corr.jpg}
    \caption{Correlation between the second NIV (left) and the second DM (right) computed using the atoms 5, 7, and 9 and using the atoms 2, 4, 6, 8, and 10.}
    \label{fig:ala_corr}
\end{figure*}

\begin{table}
\begin{tabular}{| r || c | c |}
  \hline
  & Coordinate 1 & Coordinate 2 \\
  \hline
  \hline
  NLICA & 0.62 & 0.84 \\
  \hline
  DM & 0.54 & 0.60 \\
  \hline
\end{tabular}
\caption{Correlation between the embedding coordinates from NLICA and DMAPS using the atoms 5, 7, and 9 and using the atoms 2, 4, 6, 8, and 10.}
\end{table}

We then use NIV together with Laplacian Pyramids to predict the conformation of Ala2 when we only observe some of the atoms.
%
Our training data consists of 10,000 data points, each containing the coordinates of all the atoms in Ala2.
%
We compute the NIV embedding for the training data, and then train Laplacian Pyramids to go from the NIV embedding to the moelcular configuration.
%
The NIV embedding for the training data is shown in Figure \ref{fig:ala_embed}.
%
The embedding is three-dimensional; the first coordinate describes the flipping of atoms 1 and 3, the second coordinate describes the dihedral angle $\phi$, and the third coordinate describes the dihedral angle $\psi$.

We then take 12,000 test data points.
%
Each test point contains the coordinates of {\em only} the odd atoms in Ala2.
%
2,000 of the test points are also contained in the training data set, so that we can evaluate the consistency of the two embeddings.
%
We compute the NIV embedding for the test data points, and then use the Laplacian Pyramids from the training data to predict the location of all the atoms for each point in the test data.

\begin{table}
\begin{tabular}{| r || c | c | c |}
  \hline
  & Coordinate 1 & Coordinate 2 & Coordinate 3 \\
  \hline
  \hline
  NLICA & 0.97 & 0.72 & 0.85 \\
  \hline
  DM & 0.95 & 0.76 & 0.74 \\
  \hline
\end{tabular}
\caption{Correlation between the embedding coordinates from NLICA and DMAPS using all and just odd coordinates.}
\end{table}

\begin{figure*}[ht]
    \includegraphics[width=0.3\textwidth]{ala_figures_allodd/ala2_embed1.jpg}
    \includegraphics[width=0.3\textwidth]{ala_figures_allodd/ala2_embed2.jpg}
    \includegraphics[width=0.3\textwidth]{ala_figures_allodd/ala2_embed3.jpg}
    \caption{The 3-dimensional NIV embedding for Alanine Dipeptide computed using all the atoms, colored by the y-coordinate of the first atom (left), the dihedral angle $\phi$ (center), and the dihedral angle $\psi$ (right). Each embedding is rotated so that the correlation between the colors and the relevant NIV can easily be seen.}
    \label{fig:ala_embed}
\end{figure*}

The embeddings for the two data sets are found to be fairly consistent, with correlations of 0.97, 0.72, 0.85 for the first, second, and third NIVs, respectively.
%
Figure \ref{fig:ala_recon} shows the reconstructed position versus true position for some selected atoms.
%
The atoms that have significant movement (such as atoms 1 and 3) exhibit a strong correlation between the true and reconstructed positions.
%
Figure \ref{fig:ala_molecules} shows molecular structures for the true and reconstructed configurations for some data points.
%
There is nice agreement between the true and reconstructed configurations.

\begin{figure*}[ht]
  \centering
    \foreach \atomnumber in {1,2,3,6,8} {
        \includegraphics[width=0.2\textwidth]{ala_figures_allodd/ala2_recon_corr_\atomnumber_1.jpg}
        \includegraphics[width=0.2\textwidth]{ala_figures_allodd/ala2_recon_corr_\atomnumber_2.jpg}
        \includegraphics[width=0.2\textwidth]{ala_figures_allodd/ala2_recon_corr_\atomnumber_3.jpg}\\
      }
  \caption{The correlation between the true position and the reconstructed position (using Laplacian Pyramids) for the test data. The columns correspond the x-, y-, and z-coordinates, and the rows correspond to atoms 1, 2, 3, 6, and 8.}
  \label{fig:ala_recon}
\end{figure*}

\begin{figure*}[ht]
    \centering
    \includegraphics[width=0.3\textwidth,trim=0.5in 3in 4in 0.5in, clip]{molec_figures/molecule300_balls.jpg}
    \includegraphics[width=0.3\textwidth,trim=0.5in 3in 4in 0.5in, clip]{molec_figures/molecule300_wire.jpg}\\
    \includegraphics[width=0.3\textwidth,trim=0.5in 3in 4in 0.5in, clip]{molec_figures/molecule600_balls.jpg}
    \includegraphics[width=0.3\textwidth,trim=0.5in 3in 4in 0.5in, clip]{molec_figures/molecule600_wire.jpg}\\
    \includegraphics[width=0.3\textwidth,trim=0.5in 3in 4in 0.5in, clip]{molec_figures/molecule900_balls.jpg}
    \includegraphics[width=0.3\textwidth,trim=0.5in 3in 4in 0.5in, clip]{molec_figures/molecule900_wire.jpg}
    \caption{True structure (black) and reconstructed structure (red) for three different data points. Each data point is shown in ``ball-and-stick'' representation (left) and ``wireframe'' representation (right) so that the discrepancies can easily be seen between the different configurations.}
    \label{fig:ala_molecules}
\end{figure*}

\section{Conclusions} \label{sec:conclusions}

\begin{acknowledgments}
The authors would like to thank...
\end{acknowledgments}

\bibliography{../../references/references}

\appendix

\section{Chemical Reaction Network Parameters}
The reaction network
\begin{center}
\ce{$E$ + $S$ <=>[e_1][e_{-1}] $E:S$ ->[e_2][] $E$ + $S^{*}$}\\
\ce{$S$ + $E$ <=>[b_1][b_{-1}] $S:E$ ->[b_2][] $S$ + $E^{*}$}\\
\ce{$D$ + $S^{*}$ <=>[d_1][d_{-1}] $D:S^{*}$ ->[d_2][] $D$ + $S$}\\
\ce{$F$ + $E^{*}$ <=>[f_1][f_{-1}] $F:E^{*}$ ->[f_2][] $F$ + $E$}\\
\end{center}

is governed by the following ODEs
\begin{equation}
    \begin{array}{rcl}
        \frac{d S}{dt} & = & -e_1 S E + e_{-1} E:S -b_1 S E \\
                        &   & + b_{-1} S:E + b_2 S:E + d_2 D:S^{*} \\
        \frac{d E}{dt} & = & -e_1 S E + e_{-1} E:S + e_2 E:S \\
                        &   & - b_1 S E +b_{-1} S:E + f_2 F:E^{*} \\
        \frac{d E:S}{dt}  & = & e_1 S E -e_{-1} E:S -e_2 E:S \\
        \frac{d S:E}{dt}& = & b_1 S E -b_{-1} S:E -b_2 S:E \\
        \frac{d S^{*}}{dt} & = & e_2 E:S -d_1 D S^{*} + d_{-1} D:S^{*} \\
        \frac{d E^{*}}{dt} & = & b_2 S:E -f_1 F E^{*} + d_{-1} F:E^{*} \\
        \frac{d D}{dt} & = & -d_1 D S^{*} + d_{-1} D:S^{*} + d_2 D:S^{*} \\
        \frac{dF}{dt} & = & -f_1 F E^{*} + f_{-1} F:E^{*} + f_2 F:E^{*} \\
        \frac{d D:S^{*}}{dt} & = & d_1 D S^{*} - d_{-1} D:S^{*} -d_2 D:S^{*} \\
        \frac{d F:E^{*}}{dt} & = & f_1 FE^{*} - f_{-1} F:E^{*} -f_2 F:E^{*}
    \end{array}
\end{equation}

We can write four balance equations 
\begin{equation}
    \begin{array}{rcl}
        S_T & = & S^{*} + S + E:S + S:E + D:S^{*} \\
        E_T & = & E^{*} + E + E:S + S:E + F:E^{*} \\
        D_T & = & D + D:S^{*} \\
        F_T & = & F + F:E^{*} 
    \end{array}
\end{equation}

We eliminate $S^{*}$, $E^{*}$, $D$, and $F$ from the system of ODEs.
%
We therefore obtain
\begin{equation}
    \begin{array}{rcl}
        \frac{d S}{dt} & = & -e_1 S E + e_{-1} E:S -b_1 S E \\
                        &   & + b_{-1} S:E + b_2 S:E + d_2 D:S^{*} \\
        \frac{d E}{dt} & = & -e_1 S E + e_{-1} E:S + e_2 E:S \\
                        &   & - b_1 S E +b_{-1} S:E + f_2 F:E^{*} \\
        \frac{d E:S}{dt}  & = & e_1 S E -e_{-1} E:S -e_2 E:S \\
        \frac{d S:E}{dt}& = & b_1 S E -b_{-1} S:E -b_2 S:E \\
        \frac{d D:S^{*}}{dt} & = & d_1 (D_T - D:S^{*}) (S_T - S - E:S - S:E - D:S^{*}) \\
                            &   & - d_{-1} D:S^{*} -d_2 D:S^{*} \\
        \frac{d F:E^{*}}{dt} & = & f_1 (F_T - F:E^{*}) (E_T - E - E:S - S:E - F:E^{*}) \\
                            &   & - f_{-1} F:E^{*} -f_2 F:E^{*}
    \end{array}
\end{equation}

Alternatively, we can write the rates for the 12 chemical reactions as
\begin{equation}
    \begin{array}{rcl}
        r_1 & = & e_1 S E \\
        r_2 & = & e_{-1} E:S \\
        r_3 & = & e_2 E:S \\
        r_4 & = & b_1 S E \\
        r_5 & = & b_{-1} S:E \\
        r_6 & = & b_2 S:E \\
        r_7 & = & d_1 (D_T - D:S^{*}) (S_T - S - E:S - S:E - D:S^{*}) \\
        r_8 & = & d_{-1} D:S^{*} \\
        r_9 & = & d_2 D:S^{*} \\
        r_{10} & = & f_1 (F_T - F:E^{*}) (E_T - E - E:S - S:E - F:E^{*}) \\
        r_{11} & = & f_{-1} F:E^{*} \\
        r_{12} & = & f_2 F:E^{*}
    \end{array}
\end{equation}

For the Gillespie SSA, we use these rates to adjust the {\em number} of each molecule, depending on which reaction occurs.
%
We use the parameters $b_1=5/V$, $d_1=0.0009/V$, $e_1=0.1/V$, $f_1=0.1/V$, $b_{-1} = 10.6$, $d_{-1}=0.05$, $e_{-1}=0.5$, $f_{-1} =0.01$, $b_2=0.4$, $d_2=0.85$, $e_2=0.05$, and $f_2=2$.
%
We take $S_T=E_T=D_T=1V$, and $F_T=0.02V$, where $S_T$, $E_T$, $D_T$, and $F_T$ are total number of $S$, $E$, $D$, and $F$, respectively.

\end{document}

### 2013-12-27 paper/draft2.tex
Added files related to paper

\documentclass[aip,jcp,reprint,twocolumn]{revtex4-1}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage{mhchem}

\usepackage[size=small]{caption}
%\usepackage{etoolbox}
%\AtBeginEnvironment{algorithm}{\noindent\hrulefill\par\nobreak\vskip-5pt}

\usepackage{newfloat}
\DeclareFloatingEnvironment[
    fileext=loa,
    listname=List of Algorithms,
    name=ALGORITHM,
    placement=tbhp,
]{algorithm}

\DeclareCaptionFormat{algorithms}{\vskip-15pt\hrulefill\par#1#2#3\vskip-6pt\hrulefill}
\captionsetup[algorithm]{singlelinecheck=off,format=algorithms}

%\usepackage{float}
%\floatstyle{ruled}
%\newfloat{algorithm}{t}{loa}
%\floatname{algorithm}{Algorithm}

\begin{document}

\title{Intrinsic Variables and State Reconstruction in Multiscale Simulations}

%\author{}
%\email{}
%\affiliation{}

\date{\today}

\begin{abstract}
Finding informative low-dimensional descriptions of high dimensional simulation data
(like the ones arising in molecular dynamics or kinetic Monte Carlo simulations of
physical/chemical processes) is crucial in understanding physical phenomena, and can
also dramatically assist in accelerating the simulations themselves.
%
In this paper, we discuss and illustrate the use of nonlinear intrinsic variables (NIVs)
in the mining of high-dimensional multiscale simulation data.
%
In particular, we focus on the way that NIVs allow us to functionally merge different
simulation ensembles, and {\em different partial observations of these ensembles}, as well
as to infer variables not explicitly measured.
%
The approach relies on certain simple features of the underlying process variability to
filter out measurement noise and systematically recover a unique reference coordinate frame.
%
We illustrate the approach through two distinct sets of multiscale atomistic simulations: a fast/slow
time scale simulation of an enzyme reaction network, and a molecular
dynamics simulation of alanine dipeptide in explicit water.

\end{abstract}

\keywords{Intrinsic variables, partial observations, complex dynamical systems}

\maketitle

\section{Introduction}
The last decade has witnessed extensive advances in dimensionality reduction techniques:
finding meaningful low-dimensional descriptions of high-dimensional data.
%
These developments have the potential to significantly enable the computational exploration
of phyiscochemical problems.
%
If the (high-dimensional) data $Y(t)$ arise from, say, a
molecular dynamics simulation of a macromolecule in solution, or from the stochastic
simulation of a complex chemical reaction scheme, the detection of a few good, coarse-grained
``reduction coordinates" $x(t)$ is invaluable in understanding and predicting system behavior.

%
While the benefits from such reduced descriptions are manifest, a crucial shortcoming of
such data-driven reduction coordinates is their dependence on the specific data set processed,
and not only on the physical model in question.
%
It is well known that, even in the simple linear case of Principal Component Analysis,
different data sets on the same low-dimensional hyperplane in the ambient space $Y$
will lead to different basis vectors - in effect, to different $x$.
%
While this can be easily fixed by a linear transformation, the problem becomes
exacerbated when the low-dimensional space is curved (a manifold, rather than a hyperplane)
and when different data ensembles are obtained using different instrumental modalities
(say, when one wants to merge molecular dynamics data with spectral information).
%
Clearly, the ability to systematically construct a {\em unique, consistent} reduction
coordinate set, shared by all measurement ensembles and observation modalities -
what we call a (Nonlinear) Intrinsic Variable Set - is invaluable.
%
Embedding data in such a coordinate system allows us to naturally merge different observations of the same system;
more importantly, it enables the construction of an empirical mapping between these different
observation ensembles, allowing us to complete partial measurements through the construction
of nonlinear observers of unmeasured variables.
%
This latter empirical mapping requires good interpolation tools in embedding space; to this
end we will demonstrate here the use of a multiscale Laplacian Pyramid approach \cite{rabin2012heterogeneous}.
%
We chose two distinct multiscale illustrative examples. The first is a simulation
of two Goldbeter-Koshland modules in an enzyme kinetics model \cite{zagaris2012stability} using Gillespie's Stochastic Simulation
Algorithm (SSA); in certain parameter regimes, separation of time scales is known
to reduce this model to an effective two-dimensional description.
%
Although this example is rather simple, it will serve as a nice introduction to our techniques and highlight the main features of the algorithms.
%
The second example is a molecular dynamics
simulation (in explicit water) of a simple peptide fragment (alanine dipeptide) whose folding
dynamics are known to be described through a small set of physical observables.
%
This example will demonstrate how our techniques outperform more common techniques, such as diffusion maps for dimensionality reduction
and nearest neighbor interpolation for reconstruction.
%
The paper is structured as follows: in Section \ref{sec:NIV} we present the Nonlinear Intrinsic Variable formulation and
the associated inference method. 
%
Section \ref{sec:LapPyr} contains our discussion of Laplacian Pyramids that
will be used for the completion of partial observations. 
%
Section \ref{sec:examples} contain the results
of applying the approach to simulation data from our two illustrative examples. 
%
We conclude with
a summary and discussion of open issues in Section \ref{sec:conclusions}.

\section{Nonlinear Intrinsic Variables} \label{sec:NIV}

\subsection{Overview}
Let $\mathbf{Y}(t)$ be a high-dimensional measured process in $\mathbb{R}^n$ consisting of $n$ observable variables.
The measured process is assumed to be a manifestation in an observable domain of a low-dimensional diffusion process. Thus, it can be expressed by
\begin{equation}
	\mathbf{Y}(t) = f(\mathbf{x}(t))
\end{equation}
where $f:\mathbb{R}^d \rightarrow \mathbb{R}^n$ is an unknown (possibly nonlinear) function and $\mathbf{x}(t)$ is a diffusion process that consists of $d$ underlying variables.
%
The dynamics of the diffusion process in each of its underlying variables are described by normalized stochastic differential equations as
\begin{equation}
	d x_i(t) = a_i (x_i(t)) dt + d w_i(t), \ i=1,\ldots,d
\end{equation}
where $a_i$ are unknown drift functions and $\dot{w}_i(t)$ are independent white noises TODO: should there be a dot here?.

Given a sequence of samples $\mathbf{Y}(t), \ t=1,\ldots,T$, we present an empirical method to construct a unique and consistent reduction coordinate set, represented here by $\mathbf{x}(t)$. 
%
By relying on the fact that the empirical method we will describe is independent of the observation function $f$, we present the notion of intrinsic modeling and refer to the coordinates of $\mathbf{x}(t)$ as a Nonlinear Intrinsic Variable (NIV) set.
%
The available samples $\mathbf{Y}(t)$ may be the product of different measurement functions $f$ in various observable domains, or they maybe partial measurements consisting of merely a subset of the coordinates of the observable domains.
%
The idea would be to empirically construct a NIV coordinate system driven entirely by measurements that is invariant to the observation modality (see Figure \ref{fig:IntrinsicIllustration} for illustration).

The presented method consists of the following main principles. 
%
(1) The underlying diffusion process implies that a short trajectory of successive samples mainly consists of diffusion noise, and hence, creates a ``sphere" of samples in the underlying domain. 
%
This sphere is mapped to an ellipse in the observable domain by the measurement function $f$. 
%
In this work, the identification of the associated ellipse of samples according to the time labels enables to explore and learn the tangent planes of the observable manifolds via the principal components of the covariance matrices of the samples in these ellipses. 
%
(2) The principal directions of the tangent planes are utilized to define a Riemannian metric that is shown to be locally invariant to the measurement function $f$. 
%
(3) The NIV are constructed through the eigenvalue decomposition of a Laplace operator that is built upon a pairwise affinity between the samples, defined using the Riemannian metric.

\begin{figure*}[ht]
\includegraphics[width=0.7\textwidth]{IntrinsicEmbeddingIllustration2}
\caption{Illustration of the nonlinear embedding that yields an intrinsic representation independent of the measurement modality. (Bottom) The underlying variables in which the diffusion parameters are independent with unit variance. The circle illustrates successive sample in a short burst/trajectory in time that create a sphere on the parametric manifold. (Top Left) The first set of observed variables. The ellipse illustrates the mapping of the sphere of the underlying samples into the observable domain via the first observation modality. In this domain, we illustrate that the observations might be partial, i.e., consist of merely a subset of the observed domain variables. (Top Right) The second set of observable variables. The ellipse illustrates the mapping of the sphere of the underlying samples into a different observable domain via a second observation modality.}
\label{fig:IntrinsicIllustration}
\end{figure*}

\subsection{Mahalanobis Distance}
\label{subsec:mahalanobis}

Let $\mathbf{C}(t)$ be the covariance matrix associated with the measured sample $\mathbf{Y}(t)$. In practice, the covariance matrix can be estimated from a short trajectory of samples in time around the sample $\mathbf{Y}(t)$ by
\begin{equation}
	\widehat{\mathbf{C}}(t) = \sum \limits _{\tau = t-L}^{t+L} (\mathbf{Y}(\tau) - \widehat{\boldsymbol{\mu}}(t))(\mathbf{Y}(\tau) - \widehat{\boldsymbol{\mu}}(t))^T
	\label{eq:cov}
\end{equation}
where $\widehat{\boldsymbol{\mu}}(t)$ is the empirical mean of the samples.

In particular, we define a Riemannian metric between a pair of samples using the associated covariance matrices as
\begin{align}
	d(\mathbf{Y}(t), \mathbf{Y}(\tau)) &= 2 (\mathbf{Y}(t) - \mathbf{Y}(\tau))^T(\mathbf{C}(t) + \mathbf{C}(\tau))^{-1} \nonumber \\
	&\times (\mathbf{Y}(t) - \mathbf{Y}(\tau)).
	\label{eq:mahalanobis}
\end{align}
This metric \eqref{eq:mahalanobis} is called the Mahalanobis distance.
%
As previously described, the covariance matrices convey the local variability of the measurements and are utilized to explore and learn the tangent planes of the observable manifold.
%
This information is then utilized in \eqref{eq:mahalanobis} to compare a pair of points according to the directions of their respected tangent planes.

The Mahalanobis distance is invariant under linear transformations.
%
Thus, by assuming that the observation function $f$ is bi-Lipschitz and smooth and by using local linearization of the function, i.e., $\mathbf{Y}(t) = \mathbf{J}(t) \mathbf{x}(t) + \boldsymbol{\epsilon}(t)$ where $\mathbf{J}(t)$ is the Jacobian of $f(\mathbf{x}(t))$ and $\boldsymbol{\epsilon}(t)$ is the residual consisting of higher-order terms, it was shown by Singer and Coifman \cite{singer2008non} that $\mathbf{C}(t) = \mathbf{J}(t)\mathbf{J}^T(t)$ and that the Mahalanobis distance approximates the Euclidean distance between the corresponding samples of the underlying process to a second order, i.e.,
\begin{equation}
	\| \mathbf{x}(t) - \mathbf{x}(\tau) \|^2 = d^2(\mathbf{Y}(t), \mathbf{Y}(\tau)) + \mathcal{O}(\| \mathbf{Y}(t) - \mathbf{Y}(\tau)\|^4).
\end{equation}
%
This result implies that the Mahalanobis distance is invariant to the measurement function $f$, and hence, 
it yields the same distances between samples obtained under different observation modalities or even partial observations.

\subsection{Laplace Operator}
The Mahalanobis distance described in Section \ref{subsec:mahalanobis} enables to compare the observations in terms of the intrinsic variables of the associated underlying diffusion process.
%
In this section, we recover the underlying process itself from the pairwise Euclidean distances through the eigenvectors of a Laplace operator.

Let $\mathbf{W}$ be a pairwise affinity matrix (kernel) based on a Gaussian, whose $(t,\tau)$-th element is given by
\begin{equation}
	W_{t,\tau} = \exp \left\{ - \frac{ d^2(\mathbf{Y}(t), \mathbf{Y}(\tau) )} {\varepsilon}\right\}
	\label{eq:kernel}
\end{equation}
where $\varepsilon$ is the kernel scale, which can be set according to \cite{coifman2008graph}.
%
Let $\mathbf{D}$ be a diagonal matrix whose diagonal elements are the sums of rows of $\mathbf{W}$, and let $\mathbf{W}^{\mathrm{norm}} = \mathbf{D}^{-1/2}\mathbf{W}\mathbf{D}^{-1/2}$ be a normalized kernel that shares its eigenvectors with the normalize graph-Laplacian $\mathbf{I}-\mathbf{W}^{\mathrm{norm}}$ \cite{chung1997spectral}. 
%
These eigenvectors $\psi_j$ reveal the underlying structure of the data TODO:cite. 
%
Specifically, the $i$-th coordinate of the $j$-th eigenvector can be associated with an intrinsic coordinate $j$ of the sample $\mathbf{x}(i)$ of the underlying process. 
%
The above steps to construct the nonlinear intrinsic variables are summarized in Algorithm \ref{algo}.

%TODO: do we need to describe the extension procedure?! - the recovery of the intrinsic variables through the eigenvalue decomposition (EVD) of the Laplace operator is computationally heavy...\\
TODO: we can discuss the probabilistic interpretation and explicitly state that we set the kernel scale to the median of the kernel.\\

Each eigenvector then describes an intrinsic variable for the data set of interest.
%
Because eigenvectors can be determined up to a scaling factor, we must scale the eigenvectors from different data sets in a way so that the resulting embeddings are consistent.
%
We first scale the eigenvectors so that the norm of the each eigenvector is equal to the cardinality of the data set.
%
Intuitively, a data set that has twice as many samples as another data set, but the same underlying density and shape, should have twice the norm.
%
Still, the embedding eigenevctors for two identical data sets could differ by a sign.
%
Reconciling the signs for the embeddings of different data sets can be done in several ways and is somewhat problem-specific.
%
For example, if the mean of the embedding is sufficiently far from 0, the signs can be chosen such that the mean of each eigenvector is greater than 0;
alternatively, we can construct the data sets such that a few points are common between both data sets, and the sign of each eigenvector can be chosen to optimize the consistency of the embeddings of these common points.

\begin{algorithm}[th!]
\caption{Nonlinear Intrinsic Variables Construction}
\begin{enumerate}

\item
Obtain a sequence of high-dimensional observation samples $\mathbf{Y}(t)$.

\item
Compute the empirical covariance matrix $\widehat{\mathbf{C}}(t)$ of each sample $\mathbf{Y}(t)$ in a short windows in time according to \eqref{eq:cov}.

\item
Using the samples and their associated covariance matrices, compute the Mahalanobis distance between the observations \eqref{eq:mahalanobis} .

\item
Build the pairwise affinity matrix $\mathbf{W}$ and the corresponding normalized kernel $\mathbf{W}^{(\mathrm{norm})}$ (\ref{eq:kernel}).

\item
Apply EVD on the normalized kernel and view the values of its principal eigenvector as the intrinsic variables of the given observations.

\end{enumerate}
\hrule
\label{algo}
\end{algorithm}

\section{Laplacian Pyramids} \label{sec:LapPyr}

Laplacian Pyramids (LP) is a multi-scale algorithm for approximating and extending an empirical function $f$ defined on a set of points.
%
The algorithm uses Laplacian kernels of decreasing widths to create multi-scale representations of $f$.
%
These representations can be easily extended to new data points.
%
This type of multi-scale representation was introduced by Burt and Adelson \cite{burt1983laplacian} for image coding,
and was later shown to be a tight frame by Do and Veterli \cite{do2003framing}.
%
Recently, LP was used to extend nonlinear embedding coordinates to new high-dimensional data points \cite{rabin2012heterogeneous}.
%
We will first review the LP algorithm for approximating and extending a one-dimensional function.
%
We will then describe the application of LP in the context of this work - extending a high-dimensional function that is known on a set of points which lie in a low-dimensional space.

Let $f: \Gamma \rightarrow \mathbb{R}^d$ be a function that is known on a subset of points $S \subset \Gamma$.
%
A coarse representation of $f$ is generated using a coarse smoothing operator $P_0$.
%
The smoothing operator $P_0$ is a normalized, coarse Laplacian kernel, defined by $p_0(t,\tau)= s_0^{-1}(t)w_0(t,\tau)$, where $w_0(t,\tau)=e^{-d^2(t,\tau) / \sigma_0}$. 
%
The pairwise distance $d(t,\tau)$ is typically the Euclidian distance. 
%
The parameter $\sigma_0$ is set to be large compared to the values of $d(t,\tau)$  and $s_0^{-1}(t)=\sum_{\tau \in S}w_0(t,\tau)$ is the normalizing term. 
%
The application of $P_0$ to $f$ yields a coarse representation of the function, which we denote by $f_0=P_0(f)$.

The difference $\delta_1 = f-P_0(f)$ is input for the next iteration of the algorithm, 
which uses the smoothing operator $P_1$, $P_1 \propto e^{-d^2(t,\tau)^2 / 2^{-1} \sigma_0}$, to construct a coarse representation of $\delta_1$.
%
The obtained  representation of $\delta_1$, $P_1(\delta_1)$ together with the result of the previous iteration $f_0 = P_0(f)$ 
yields a new, finer representation of $f$, $f_1 = P_0(f) + P_1(\delta_1)$.

In an iterative manner, multi-scales representations of the function $f$, denoted by $f_l$ are constructed (see Eq. \eqref{eq:LP_multi_scale}). 
%
As $l$ grows, the approximation is of a finer resolution because $P_l \propto e^{-d^2(t,\tau) / 2^{-l} \sigma_0}$ uses a Laplacian kernel of a finer width. 
%
The iterations stop when the difference between $f$ and $f_l$ is smaller than a pre-defined error threshold.

\begin{equation} \label{eq:LP_multi_scale}
 \begin{array}{cl}
\mbox{scale 0:} & f_0 = P_0(f) \\
\mbox{scale 1:} & f_1 = P_0(f) + P_1(\delta_1) \\
: & : \\
\mbox{scale l:} & f_l = P_0(f) + \sum_{k=1}^{l}P_k(\delta_k)\\
\end{array}
\end{equation}

The representations $f_0$, $f_1$,..$f_l$ can be extended to a new point $\tilde{t} \in \Gamma \backslash S $ by extending the operators $P_0, P_1,\ldots,P_l$. 
%
For example, $f_0(\tilde{t}) = \sum_{t \in S} p_0(\tilde{t}, t)f(t)$ and
$f_1(\tilde{t}) = f_0(\tilde{t}) + \sum_{(t \in S)}p_1(\tilde{t}, t)\delta_1(t)$.

\begin{widetext}
Figure \ref{fig:LP_ex} displays an illustrative example of the algorithm, when applied to the function 
 \begin{equation} \label{eq:LP_example}
f(x) = \left\{
\begin{array}{l l}
-0.02(x-4\pi)^2 + \sin(x) &  0 \le x \le 4\pi \\
-0.02(x-4\pi)^2 + \sin(x) + 0.5 \sin(3x) &  4\pi < x \le 7.5 \pi \\
-0.02(x-4\pi)^2 + \sin(x) + 0.5 \sin(3x) +0.25 \sin(9x) &  7.5 \pi < x \le 10 \pi
\end{array}
\right.
\end{equation}
that contains several scales. 
%
The coarse regions of the function ( $0 \le x \le 4\pi$ ) are well approximated by a small number of scales. 
%
As the function becomes more oscillatory ( $\pi \le x \le 7.5\pi$ and $7.5\pi \le x \le 10\pi$), 
a finer representation, (meaning a larger number of scales  $l$ )  is required to capture its behavior.
\end{widetext}

In this work, LP is applied to extend a high-dimensional function $f:\mathbb{R}^d \rightarrow \mathbb{R}^n$, which maps a set of points in the NIV space to their values $\mathbf{Y}(t)$ in the observable space.
%
Let $\Psi(t) = \left(\psi_1(t),\psi_2(t),\ldots,\psi_d(t)\right)$ be the set of NIV's that were constructed from the data samples $\mathbf{Y}(t)$, as described in Algorithm \ref{algo}. 
%
The values of function $f:\mathbb{R}^d \rightarrow \mathbb{R}^n$ are known on the subset $S = \{\Psi(t)\}$, with $f(\Psi(t)) = Y(t)$.

A na\"{\i}ve way to extend $f$ to a new data point $\Psi(\tilde{t})$ is find the point's nearest neighbors in NIV space and average their function values. 
%
A different, point-wise adaptive approach is described by Buchman {\em et al} \cite{buchman2011high}; 
high-dimensional hurricane tracks were estimated from low dimensional embedding coordinates using a weighted average of the points close to $\Psi(\tilde{t})$ in the embedded space. 
%
However, this point-wise adaptation requires setting the nearest neighborhood radius parameter for every point.

The LP algorithm automatically finds the nearest neighborhood radius for each new point $\Psi(\tilde{t})$. 
%
This radius will be large in smooth regions of the function, and small in regions in which $f$ contains higher frequency components. 
%
The LP approximation of a new, high dimensional value $Y(\tilde{t})$, is calculated by a weighted average of the function values that belong to the neighboring points. 
%
The weights are based on the pairwise distances in the intrinsic, low-dimensional space. 
%
In practice, a set of smoothing operators $P_0, P_1, \ldots, P_l$
\begin{equation} \label{eq:LP_multi_scale_app}
P_l \propto e^{-d(\Psi(t),\Psi(\tau))^2 / \sigma_0 \cdot 2^{-l}}
\end{equation}
are constructed and later extended to create the multiscale approximations as defined in Eq. \ref{eq:LP_multi_scale}.
%
The LP algorithm for inverse mapping is summarized in Algorithm \ref{algo_LP}.

\begin{algorithm}[th!]
\caption{Laplacian Pyramids for Inverse Mapping}
\begin{enumerate}

\item
Construct a set of smoothing operators $P_0, P_1, \ldots, P_l$  based on intrinsic pairwise distances $d(\Psi(t),\Psi(\tau))$.

\item
Use the smoothing operators to obtain a multiscale representation (see Eq. \ref{eq:LP_multi_scale_app}) of $f:\Psi(t) \rightarrow Y(t)$.

\item Given a new point $\Psi(\tilde{t})$ in NIV, extend the smoothing operators $P_0, P_1, \ldots, P_l$  by
$p_l(\Psi(\tilde{t}), \Psi(t)) = s_0^{-1}(\Psi(t))w_0(\Psi(\tilde{t}),\Psi(t))$.

\item
Use the extended smoothing operators to approximate the value of $f_l(\tilde{t}) = f_0(\tilde{t}) + \sum_{k}\sum_{(t \in S)}p_k(\tilde{t}, t)\delta_k(t)$

\item
The value $Y(\tilde{t})$ is approximated by $f_l(\Psi(\tilde{t}))$.

\end{enumerate}
\hrule
\label{algo_LP}
\end{algorithm}

\begin{figure*}[ht]
\includegraphics[width=0.2\textwidth]{LP_figures/Scale3_S2}
\includegraphics[width=0.2\textwidth]{LP_figures/Scale5_S2}
\includegraphics[width=0.2\textwidth]{LP_figures/Scale8_S2}
\includegraphics[width=0.2\textwidth]{LP_figures/Scale11_S2}
\caption{Approximation of function $f(x)$ as defined in \eqref{eq:LP_example} for scales 3, 5, 8, and 11 using Laplacian Pyramids (top, the true function is shown in blue and the Laplacian Pyramids approximation in shown in black), and the residual error in the approximation (bottom, shown in red).}
\label{fig:LP_ex}
\end{figure*}

\section{Models and Results} \label{sec:examples}

\subsection{Chemical Reaction Network}

We first consider a chemical reaction network of multiple enzyme-substrate interactions \cite{zagaris2012stability}.
%
The equations that govern the network are\\
\begin{center}
\ce{$E$ + $S$ <=>[e_1][e_{-1}] $E:S$ ->[e_2][] $E$ + $S^{*}$}\\
\ce{$S$ + $E$ <=>[b_1][b_{-1}] $S:E$ ->[b_2][] $S$ + $E^{*}$}\\
\ce{$D$ + $S^{*}$ <=>[d_1][d_{-1}] $D:S^{*}$ ->[d_2][] $D$ + $S$}\\
\ce{$F$ + $E^{*}$ <=>[f_1][f_{-1}] $F:E^{*}$ ->[f_2][] $F$ + $E$}\\
\end{center}

The $^{*}$ denote activated forms of the various species, and the : denote complexes formed between two species; the complexes $E:S$ and $S:E$ are not equivalent.

There are 10 species in this reaction system;
however, one can write four conservation equations (since total $E$, $S$, $D$, and $F$ are all conserved) to reduce the system to 6 dimensions.
%
%We assume each reaction follows an elementary rate law.
%
%We use the parameters $b_1=5$, $d_1=0.0009$, $e_1=0.1$, $f_1=0.1$, $b_{-1} = 10.6$, $d_{-1}=0.05$, $e_{-1}=0.5$, $f_{-1} =0.01$, $b_2=0.4$, $d_2=0.85$, $e_2=0.05$, and $f_2=2$.
%
%We take $S_T=E_T=D_T=1$, and $F_T=0.02$, where $S_T$, $E_T$, $D_T$, and $F_T$ are total concentrations of $S$, $E$, $D$, and $F$, respectively.
%
In the parameter regime we consider, the system exhibits a separation of timescales, so that initial conditions quickly approach a two-dimensional manifold;
details about the specific parameters can be found in Appendix \ref{app:rxn}.
%
%The relevant timescales around the steady state ($-1/\lambda_i$, where $\lambda_i$ are the eigenvalues of the Hessian) are 1176, 9.731, 1.594, 1.111, 0.4975, 0.06498.

Although the dynamics of chemical reaction networks are typically described by a system of ODEs, the ODEs are only an approximation that holds in the limit of a large number of molecules.
%
When the number of molecules is small, the system is inherently stochastic and its dynamics can be simulated using the Gillespie's Stochastic Simulation Algorithm (SSA) \cite{gillespie1977exact}.
%
We can control the level of noise by adjusting the volume $V$, and therefore adjusting the number of molecules, in the system (in our simulations, we take $V=10^5$).
%
We take the volume small enough so that we can still see appreciable stochasticity in small simulation bursts, but large enough so that the underlying two-dimensional manifold is (relatively) smooth.

We generate 3000 random initial conditions $\mathbf{Y}^{full}_0(1), \dots, \mathbf{Y}^{full}_0(3000) \in \mathbb{R}^6$, enforcing that all concentrations must be non-negative
(we take the order of the components in the vector $\mathbf{Y}^{full}_0(t)$ to be $S$, $E$, $E:S$, $S:E$, $D:S^{*}$, $F:E^{*}$).
%
We integrate each point $\mathbf{Y}^{full}_0(t)$ forward for 10 time units using the SSA to obtain point the $\mathbf{Y}^{full}(t) \in \mathbb{R}^6$;
10 time units is sufficiently long for the initial points to converge to the two-dimensional manifold,
but not long enough for the points to converge to a one-dimensional curve or to the final steady state.
%
We take $\mathbf{Y}^{full}(1), \dots, \mathbf{Y}^{full}(3000)$ to be representative points on the underlying two-dimensional manifold;
three-dimensional projections of the resulting manifold are shown in Figure \ref{fig:rxn_manifolds}.
%
From each manifold point $\mathbf{Y}^{full}(t)$, we run 20 short simulations, each for 0.2 time units.
%
We denote the endpoints from the short simulations as $\widetilde{\mathbf{Y}}^{full}(t, 1), \dots, \widetilde{\mathbf{Y}}^{full}(t, 20)$.

\begin{figure*}[ht]
  \includegraphics[height=6cm]{rxn_figures/rxn_manifold1}\llap{\raisebox{3.5cm}{\includegraphics[height=2.5cm]{rxn_figures/rxn_manifold1_2}}}
  \includegraphics[height=6cm]{rxn_figures/rxn_manifold2}\llap{\raisebox{3.5cm}{\includegraphics[height=2.5cm]{rxn_figures/rxn_manifold2_2}}} \\
  \includegraphics[height=6cm]{rxn_figures/rxn_manifold3}\llap{\raisebox{3.5cm}{\includegraphics[height=2.5cm]{rxn_figures/rxn_manifold3_2}}}
  \includegraphics[height=6cm]{rxn_figures/rxn_manifold4}\llap{\raisebox{3.5cm}{\includegraphics[height=2.5cm]{rxn_figures/rxn_manifold4_2}}} \\
    \caption{Projections of the manifold from stochastic simulation of the chemical reaction network. The insets show rotations of the projections to illustrate the two-dimensionality of the manifold.}
    \label{fig:rxn_manifolds}
\end{figure*}

We consider two different data sets from our simulations.
%
Data set 1, denoted $\mathbf{Y}_1(t) \in \mathbb{R}^4$, consists of $\mathbf{Y}^{full}(1), \dots, \mathbf{Y}^{full}(2000)$, restricted to components $S$, $E$, $S:E$, and $F:E^{*}$, i.e.
$$\mathbf{Y}_1(t) =
\left( \begin{array}{cccccc}
1 & 0 & 0 & 0 & 0 & 0 \\
0 & 1 & 0 & 0 & 0 & 0 \\
0 & 0 & 0 & 1 & 0 & 0 \\
0 & 0 & 0 & 0 & 0 & 1
\end{array} \right) \mathbf{Y}^{full}(t), \: t=1, \dots, 2000.$$
%
Data set 2, denoted $\mathbf{Y}_2(t) \in \mathbb{R}^4$, consists of $\mathbf{Y}^{full}(1500), \dots, \mathbf{Y}^{full}(3000)$, restricted to components $S$, $E$, $E:S$, and $D:S^{*}$, i.e.
$$\mathbf{Y}_2(t) =
\left( \begin{array}{cccccc}
1 & 0 & 0 & 0 & 0 & 0 \\
0 & 1 & 0 & 0 & 0 & 0 \\
0 & 0 & 1 & 0 & 0 & 0 \\
0 & 0 & 0 & 0 & 1 & 0
\end{array} \right)
\mathbf{Y}^{full}(t + 1499), \: t=1, \dots, 1501.$$
%
The endpoints of the simulation bursts for the two data sets, $\widetilde{\mathbf{Y}_1}(t, \tau)$ and $\widetilde{\mathbf{Y}_2}(t, \tau)$, are defined analogously.
%
We then estimate the covariances for each point in each data set as
\begin{equation}
\widehat{\mathbf{C}_j}(t) = \sum_{\tau=1}^{20} \left( \widetilde{\mathbf{Y}_j}(t, \tau) - \hat{\mathbf{\mu}}_j(t) \right)\left( \widetilde{\mathbf{Y}_j}(t, \tau) - \hat{\mathbf{\mu}}_j(t) \right)^T, i \in \{1, 2\}
\end{equation}
where $\hat{\mathbf{\mu}_j}$ is the empirical mean of $\widetilde{\mathbf{Y}_j}(t, 1), \dots, \widetilde{\mathbf{Y}_j}(t, 20)$.
%
We will use NIV together with Laplacian Pyramids to predict the values of $S:E$ and $F:E^{*}$ in data set 2.

Figure \ref{fig:rxn_embedding} shows the two-dimensional NIV embeddings for the two different data sets; the embeddings appear very consistent.
%
We also compute the correlation between the embedding coordinates for $\mathbf{Y}_1(1500), \dots, \mathbf{Y}_1(2000)$ and $\mathbf{Y}_2(1), \dots, \mathbf{Y}_2(501)$,
since both come from the same underlying points $\mathbf{Y}^{full}(1500), \dots, \mathbf{Y}^{full}(2000)$.
%
We obtain a correlation of 0.97 and 0.95 for the first and second NIVs, respectively, indicating that the two embeddings are in excellent agreement with each other.

\begin{figure*}[ht]
    \includegraphics[width=0.4\textwidth]{rxn_figures/rxn_NLICA1}
    \includegraphics[width=0.4\textwidth]{rxn_figures/rxn_NLICA2}
    \includegraphics[width=0.4\textwidth]{rxn_figures/rxn_NLICA_corr1}
    \includegraphics[width=0.4\textwidth]{rxn_figures/rxn_NLICA_corr2}
    \caption{(Top, left) NIV embedding obtained from data set 1 (observations of components E, S, S:E, and F:E$^{*}$), colored by $S$. (Top, right) NIV embedding obtained from data set 2 (observations of components E, S, E:S, and D:S$^{*}$), colored by $S$. (Bottom, left) Correlation of first NIV between two different embeddings (correlation=0.97). (Bottom, right)  Correlation of second NIV between two different embeddings (correlation=0.95).}
    \label{fig:rxn_embedding}
\end{figure*}

We then use Laplacian Pyramids to predict $S:E$ and $F:E^{*}$ for $\mathbf{Y}_2(t)$.
%
Because $\mathbf{Y}_1(t)$ and $\mathbf{Y}_2(t)$ are measured for different components, there is no simple way to estimate $S:E$ and $F:E^{*}$ from the data in the observation space.
%
Instead, we must first project the data into the NIV space so that we can compute neighbors {\em between} the two data sets.
%
We use $\mathbf{Y}_1(t)$ to train a Laplacian Pyramids function from the two-dimensional NIV embedding to $S:E$ and $F:E^{*}$.
%
We then use this function to predict the values  of $S:E$ and $F:E^{*}$ for $\mathbf{Y}_2(t)$, using the computed NIV embedding for $\mathbf{Y}_2(t)$.
%
In this way, we are exploiting the fact that the NIV embedding is intrinsic and consistent between the two data sets, even though the two data sets contain measurements of different chemical species.

The results of the Laplacian Pyramids prediction are shown in Figure \ref{fig:rxn_recon}.
%
The normalized mean-squared error between the true and predicted values, defined as $\frac{\langle (y_{true}-y_{pred})^2 \rangle}{\langle y_{true}^2 \rangle}$ is 0.0372 and 0.0287 for $S:E$ and $F:E^{*}$, respectively.
%
Therefore, we can effectively predict the unobserved components in the reaction network using NIV together with Laplacian Pyramids.

\begin{figure*}[ht]
    \includegraphics[width=0.4\textwidth]{rxn_figures/rxn_recon4}
    \includegraphics[width=0.4\textwidth]{rxn_figures/rxn_recon6}
    \caption{Laplacian Pyramids (LapPyr) reconstructions of $S:E$ (left) and $F:E^{*}$ (right).}
    \label{fig:rxn_recon}
\end{figure*}

\subsection{Alanine Dipeptide}

Our second example comes from a simulation of a small peptide fragment.
%
Alanine dipeptide (Ala2) is often used as a ``prototypical'' protein for simulation studies (TODO: cite Ala2 papers).
%
We simulate the motion of Ala2 in explicit solvent using the AMBER 10 molecular simulation package \cite{case2008Amber} with an
optimized version \cite{best2009optimized} of the AMBER ff03 force field \cite{duan2003point}.
%
The molecule is solvated with 638 TIP3P water molecules \cite{jorgensen1983comparison}, with the particle mesh Ewald method for long-range electrostatic interactions \cite{essmann1995smooth} and periodic boundary conditions.
%
The simulation is performed at constant volume and temperature (NVT ensemble), with the temperature being maintained at 300~K with a Langevin thermostat \cite{loncharich1992langevin}.
%
Hydrogen bonds are fixed using the SHAKE algorithm \cite{ryckaert1977numerical}.
%
The two dihedral angles $\phi$ and $\psi$ are known to parameterize the free energy surface, and the free energy surface contains three minima (labeled A, B, and C, see Figure \ref{fig:ala_fes}).
%
Our simulations are concentrated around minimum B in the free energy surface, located at $\phi=-65^{\circ}$, $\psi=150^{\circ}$.
%
We start many simulations at $10^{\circ}$ away from the minimum, and allow the simulations to each run for 0.1~ps, while recording the configuration of Ala2 every 1~fs (therefore, each trajectory is 100 points long).
%
Configurations are recorded with all atoms except the hydrogens.

\begin{figure*}[ht]
    \includegraphics[width=0.6\textwidth]{FES}
    \includegraphics[width=0.3\textwidth]{molecule_labeled2}
    \caption{(Left) Free energy surface for Alanine Dipeptide. The minima are labeled A, B, and C, and the corresponding molecular configurations are shown. (Right) The molecular structure of alanine dipeptide, excluding the hydrogens. The atoms are numbered and the two dihedral angles $\phi$ and $\psi$ are indicated.}
    \label{fig:ala_fes}
\end{figure*}

We first wish to demonstrate the benefit of NIV over other nonlinear dimensionality reduction techniques, such as diffusion maps.
%
We consider 10,000 data points from our simulation $\mathbf{Y}^{full}(1), \dots, \mathbf{Y}^{full}(10000) \in \mathbb{R}^{30}$; every 100 data points comes from a continuous simulation trajectory.
%
We construct two data sets:
%
$\mathbf{Y}_{even}(t), t=1, \dots, 10000$ is $\mathbf{Y}^{full}(t)$ restricted to atoms 2, 4, 6, 8, and 10,
and $\mathbf{Y}_{odd}(t), t=1, \dots, 10000$ is $\mathbf{Y}^{full}(t)$ restricted to atoms 5, 7, and 9 (see Figure \ref{fig:ala_fes} for the atom indexing).
%
We then compute the NIV and diffusion maps embeddings for $\mathbf{Y}_{even}(t)$ and $\mathbf{Y}_{odd}(t)$.
%
For NIV, we compute the covariances as in \eqref{eq:cov}, with $L=10$;
at the ends of each simulation trajectory, we take fewer points before or following the point of interest to estimate the covariance
(as neighboring points in our data set that do not come from the
same trajectories should not contribute to the estimate of the covariance).

The correlation between the NIV coordinates for the two data sets and the diffusion maps coordinates for the two data sets are shown in Figure \ref{fig:ala_corr}.
%
The correlation between the two NIV embeddings is higher than the correlation between the two diffusion maps embeddings,
demonstrating that it is often advantageous to use NIV over diffusion maps if one wishes to obtain a consistent embedding and ``merge'' data sets from different observation domains.

\begin{figure*}[ht]
    \includegraphics[width=0.4\textwidth]{ala_figures_evenodd/NLICA_corr}
    \includegraphics[width=0.4\textwidth]{ala_figures_evenodd/DM_corr}
    \caption{Correlation between the second NIV (left) and the second DM (right) computed using the atoms 2, 4, 6, 8, and 10 ($\psi_2^{even}, DM_2^{even}$) , and using atoms 5, 7, and 9 ($\psi_2^{odd}, DM_2^{odd}$).
    The correlations for the first (not shown) and second NIV coordinates are found to be 0.62 and 0.84, respectively.
    The correlations for the first (not shown) and second DM coordinates are found to be 0.54 and 0.60, respectively. }
    \label{fig:ala_corr}
\end{figure*}

%\begin{table}
%\begin{tabular}{| r || c | c |}
%  \hline
%  & Coordinate 1 & Coordinate 2 \\
%  \hline
%  \hline
%  NLICA & 0.62 & 0.84 \\
%  \hline
%  DM & 0.54 & 0.60 \\
%  \hline
%\end{tabular}
%\caption{Correlation between the embedding coordinates from NLICA and DMAPS using the atoms 5, 7, and 9 and using the atoms 2, 4, 6, 8, and 10.}
%\end{table}

We then use NIV together with Laplacian Pyramids to predict the conformation of Ala2 when we only observe some of the atoms.
%
We have 20000 data points $\mathbf{Y}^{full}(1), \dots, \mathbf{Y}^{full}(20000)$, where every 100 data points comes from one continuous simulation trajectory.
%
Our first data set (which will serve as our training data for Laplacian Pyramids), $\mathbf{Y}_{all}(t)$, consists of the first 10000 data points ($\mathbf{Y}_{all}(t) = \mathbf{Y}^{full}(t), t=1, \dots, 10000$).
%
Our second data set (which will serve as our test data), $\mathbf{Y}_{odd}(t)$, consists of the last 12000 data points restricted to {\em only} the odd atoms ($\mathbf{Y}_{odd}(t) = \mathbf{Y}^{full}(t+8000)$ restricted to the odd atoms, $t = 1, \dots, 12000$).
%
We compute the covariances as in \eqref{eq:cov} with $L=15$;
at the ends of each simulation trajectory, we take fewer points before or following the point of interest to estimate the covariance.
%
We compute the NIV embedding for the training data $\mathbf{Y}_{all}(t)$ and the test data $\mathbf{Y}_{odd}(t)$; we then use the Laplacian Pyramids from the training data to predict the location of all the atoms for each point in the test data.

The NIV embedding for the training data $Y_{all}(t)$ is shown in Figure \ref{fig:ala_embed}.
%
The embedding is three-dimensional; the first coordinate describes the flipping of atoms 1 and 3, the second coordinate describes the dihedral angle $\phi$, and the third coordinate describes the dihedral angle $\psi$.
%
We calculate the correlation between the embedding coordinates for $\mathbf{Y}_{all}(8001), \dots, \mathbf{Y}_{all}(10000)$, and $\mathbf{Y}_{odd}(1), \dots, \mathbf{Y}_{odd}(2000)$,
since they come from the same simulation data points $\mathbf{Y}^{full}(8001), \dots, \mathbf{Y}^{full}(10000)$.
%
The embeddings for the two data sets are found to be fairly consistent, with correlations of 0.97, 0.72, 0.85 for the first, second, and third NIVs, respectively.

%\begin{table}
%\begin{tabular}{| r || c | c | c |}
%  \hline
%  & Coordinate 1 & Coordinate 2 & Coordinate 3 \\
%  \hline
%  \hline
%  NLICA & 0.97 & 0.72 & 0.85 \\
%  \hline
%  DM & 0.95 & 0.76 & 0.74 \\
%  \hline
%\end{tabular}
%\caption{Correlation between the embedding coordinates from NLICA and DMAPS using all and just odd coordinates.}
%\end{table}

\begin{figure*}[ht]
    \includegraphics[width=0.3\textwidth]{ala_figures_allodd/ala2_embed1}
    \includegraphics[width=0.3\textwidth]{ala_figures_allodd/ala2_embed2}
    \includegraphics[width=0.3\textwidth]{ala_figures_allodd/ala2_embed3}
    \caption{The 3-dimensional NIV embedding for Alanine Dipeptide computed using $\mathbf{Y}_{all}(t)$, colored by the y-coordinate of the first atom (left), the dihedral angle $\phi$ (center), and the dihedral angle $\psi$ (right). Each embedding is rotated so that the correlation between the colors and the relevant NIV can easily be seen.}
    \label{fig:ala_embed}
\end{figure*}

Figure \ref{fig:ala_recon} shows the reconstructed position versus true position for some selected atoms.
%
The atoms that have significant movement (such as atoms 1 and 3) exhibit a strong correlation between the true and reconstructed positions.
%
Figure \ref{fig:ala_molecules} shows molecular structures for the true and reconstructed configurations for some data points.
%
There is nice agreement between the true and reconstructed configurations.

We would like to also demonstrate the effectiveness of Laplacian Pyramids over other reconstruction techniques.
%
We also predict the configurations from the NIV components using simple nearest-neighbor interpolation.
%
The average reconstruction error, scaled by the average bond length within the molecule, is shown in Figure \ref{fig:ala_mse}.
%
The Laplacian Pyramid approach outperforms simple nearest neighbor search for all of the atoms.

\begin{figure*}[ht]
  \centering
    \foreach \atomnumber in {1,2,3,6,8} {
        \includegraphics[width=0.2\textwidth]{ala_figures_allodd/ala2_recon_corr_\atomnumber_1}
        \includegraphics[width=0.2\textwidth]{ala_figures_allodd/ala2_recon_corr_\atomnumber_2}
        \includegraphics[width=0.2\textwidth]{ala_figures_allodd/ala2_recon_corr_\atomnumber_3}\\
      }
  \caption{The correlation between the true position and the reconstructed position (using Laplacian Pyramids) for the test data. The columns correspond the x-, y-, and z-coordinates, and the rows correspond to atoms 1, 2, 3, 6, and 8.}
  \label{fig:ala_recon}
\end{figure*}

\begin{figure*}[ht]
    \centering
    \includegraphics[width=0.3\textwidth,trim=0in 0.7in 0in 0in, clip]{molec_figures/molecule300_balls}
    \includegraphics[width=0.3\textwidth,trim=0in 0.7in 0in 0in, clip]{molec_figures/molecule300_wire}\\
    \includegraphics[width=0.3\textwidth,trim=0in 0.7in 0in 0in, clip]{molec_figures/molecule600_balls}
    \includegraphics[width=0.3\textwidth,trim=0in 0.7in 0in 0in, clip]{molec_figures/molecule600_wire}\\
    \includegraphics[width=0.3\textwidth,trim=0in 0.7in 0in 0in, clip]{molec_figures/molecule900_balls}
    \includegraphics[width=0.3\textwidth,trim=0in 0.7in 0in 0in, clip]{molec_figures/molecule900_wire}
    \caption{True structure (black) and reconstructed structure (red) for three different data points. Each data point is shown in ``ball-and-stick'' representation (left) and ``wireframe'' representation (right) so that the discrepancies can easily be seen between the different configurations.}
    \label{fig:ala_molecules}
\end{figure*}

\begin{figure*}
    \includegraphics[width=0.5\textwidth]{ala_figures_allodd/ala2_recon_mse}
    \caption{Mean squared error of reconstructed position, normalized by the average bond length, for each atom in Ala2.
    The positions were reconstructed using both Laplacian Pyramids (LapPyr) and nearest neighbor interpolation (nn).}
    \label{fig:ala_mse}
\end{figure*}

\section{Conclusions} \label{sec:conclusions}

\begin{acknowledgments}
The authors would like to thank...
\end{acknowledgments}

\bibliography{../../references/references}

\appendix

\section{Chemical Reaction Network Parameters} \label{app:rxn}

We consider the following network of chemical reactions:
\begin{center}
\ce{$E$ + $S$ <=>[e_1][e_{-1}] $E:S$ ->[e_2][] $E$ + $S^{*}$}\\
\ce{$S$ + $E$ <=>[b_1][b_{-1}] $S:E$ ->[b_2][] $S$ + $E^{*}$}\\
\ce{$D$ + $S^{*}$ <=>[d_1][d_{-1}] $D:S^{*}$ ->[d_2][] $D$ + $S$}\\
\ce{$F$ + $E^{*}$ <=>[f_1][f_{-1}] $F:E^{*}$ ->[f_2][] $F$ + $E$}\\
\end{center}

\begin{widetext}
In the limit of a large number of molecules, the dynamics of this network is governed by the following ODEs:
\begin{equation}
    \begin{array}{rcl}
        S^\prime & = & -e_1 S E + e_{-1} E:S -b_1 S E + b_{-1} S:E + b_2 S:E + d_2 D:S^{*} \\
        E^\prime & = & -e_1 S E + e_{-1} E:S + e_2 E:S - b_1 S E +b_{-1} S:E + f_2 F:E^{*} \\
        E:S^\prime & = & e_1 S E -e_{-1} E:S -e_2 E:S \\
        S:E^\prime & = & b_1 S E -b_{-1} S:E -b_2 S:E \\
        {S^{*}}^\prime & = & e_2 E:S -d_1 D S^{*} + d_{-1} D:S^{*} \\
        {E^{*}}^\prime & = & b_2 S:E -f_1 F E^{*} + d_{-1} F:E^{*} \\
        D^\prime & = & -d_1 D S^{*} + d_{-1} D:S^{*} + d_2 D:S^{*} \\
        F^\prime & = & -f_1 F E^{*} + f_{-1} F:E^{*} + f_2 F:E^{*} \\
        {D:S^{*}}^\prime & = & d_1 D S^{*} - d_{-1} D:S^{*} -d_2 D:S^{*} \\
        {F:E^{*}}^\prime & = & f_1 FE^{*} - f_{-1} F:E^{*} -f_2 F:E^{*}
    \end{array}
\end{equation}
\end{widetext}

We can write four balance equations for the conservation of total $S$, $E$, $D$, and $F$.
\begin{equation}
    \begin{array}{rcl}
        S_T & = & S^{*} + S + E:S + S:E + D:S^{*} \\
        E_T & = & E^{*} + E + E:S + S:E + F:E^{*} \\
        D_T & = & D + D:S^{*} \\
        F_T & = & F + F:E^{*}
    \end{array}
\end{equation}

\begin{widetext}
We choose to eliminate $S^{*}$, $E^{*}$, $D$, and $F$ from the system of ODEs.
%
We therefore obtain a system of 6 ODEs:
\begin{equation}
    \begin{array}{rcl}
        S^\prime & = & -e_1 S E + e_{-1} E:S -b_1 S E \\
                        &   & + b_{-1} S:E + b_2 S:E + d_2 D:S^{*} \\
        E^\prime & = & -e_1 S E + e_{-1} E:S + e_2 E:S \\
                        &   & - b_1 S E +b_{-1} S:E + f_2 F:E^{*} \\
        E:S^\prime  & = & e_1 S E -e_{-1} E:S -e_2 E:S \\
        S:E^\prime & = & b_1 S E -b_{-1} S:E -b_2 S:E \\
        {D:S^{*}}^\prime & = & d_1 (D_T - D:S^{*}) (S_T - S - E:S - S:E - D:S^{*}) - d_{-1} D:S^{*} -d_2 D:S^{*} \\
        {F:E^{*}}^\prime & = & f_1 (F_T - F:E^{*}) (E_T - E - E:S - S:E - F:E^{*}) - f_{-1} F:E^{*} -f_2 F:E^{*}
    \end{array}
\end{equation}
\end{widetext}

Alternatively, we can write the rates for the 12 chemical reactions as
\begin{equation}
    \begin{array}{rcl}
        r_1 & = & e_1 S E \\
        r_2 & = & e_{-1} E:S \\
        r_3 & = & e_2 E:S \\
        r_4 & = & b_1 S E \\
        r_5 & = & b_{-1} S:E \\
        r_6 & = & b_2 S:E \\
        r_7 & = & d_1 (D_T - D:S^{*}) (S_T - S - E:S - S:E - D:S^{*}) \\
        r_8 & = & d_{-1} D:S^{*} \\
        r_9 & = & d_2 D:S^{*} \\
        r_{10} & = & f_1 (F_T - F:E^{*}) (E_T - E - E:S - S:E - F:E^{*}) \\
        r_{11} & = & f_{-1} F:E^{*} \\
        r_{12} & = & f_2 F:E^{*}
    \end{array}
\end{equation}

For the Gillespie SSA, we use these rates to adjust the {\em number} of each molecule, depending on which reaction occurs.
%
We take the volume of the reactor $V=10^5$.
%
We use the parameters $b_1=5/V$, $d_1=0.0009/V$, $e_1=0.1/V$, $f_1=0.1/V$, $b_{-1} = 10.6$, $d_{-1}=0.05$, $e_{-1}=0.5$, $f_{-1} =0.01$, $b_2=0.4$, $d_2=0.85$, $e_2=0.05$, and $f_2=2$.
%
We take $S_T=E_T=D_T=1V$, and $F_T=0.02V$, where $S_T$, $E_T$, $D_T$, and $F_T$ are total number of $S$, $E$, $D$, and $F$, respectively.

\end{document}

### 2013-12-27 paper/draft3.tex
Added files related to paper

--- 2013-12-27 paper/draft2.tex
@@
-\documentclass[aip,jcp,reprint,twocolumn]{revtex4-1}
+\documentclass[aip,jcp,preprint]{revtex4-1}
 \usepackage{amssymb}
@@
 \usepackage{tikz}
-\usepackage{mhchem}
+\usepackage{natbib}
+%\usepackage{mhchem}
 
@@
 
-%\author{}
-%\email{}
-%\affiliation{}
+\author{Carmeline J. Dsilva}
+\email{cdsilva@princeton.edu}
+\affiliation{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, USA}
+
+\author{Ronen Talmon}
+\email{ronen.talmon@yale.edu}
+\affiliation{Department of Mathematics, Yale University, New Haven, Connecticut, USA}
+
+\author{Neta Rabin}
+\email{netar@afeka.ac.ilil}
+\affiliation{Department of Exact Sciences, Afeka Tel-Aviv Academic College of Engineering, Tel-Aviv, Israel}
+
+\author{Ronald R. Coifman}
+\email{coifman@math.yale.edu}
+\affiliation{Department of Mathematics, Yale University, New Haven, Connecticut, USA}
+
+\author{Ioannis G. Kevrekidis}
+\email{yannis@princeton.edu}
+\affiliation{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, USA}
+\affiliation{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, USA}
 
@@
 \begin{abstract}
-Finding informative low-dimensional descriptions of high dimensional simulation data
+Finding informative low-dimensional descriptions of high-dimensional simulation data
 (like the ones arising in molecular dynamics or kinetic Monte Carlo simulations of
-physical/chemical processes) is crucial in understanding physical phenomena, and can
+physical and chemical processes) is crucial to understanding physical phenomena, and can
 also dramatically assist in accelerating the simulations themselves.
 %
-In this paper, we discuss and illustrate the use of nonlinear intrinsic variables (NIVs)
+In this paper, we discuss and illustrate the use of nonlinear intrinsic variables (NIV)
 in the mining of high-dimensional multiscale simulation data.
 %
-In particular, we focus on the way that NIVs allow us to functionally merge different
+In particular, we focus on the way NIV allow us to functionally merge different
 simulation ensembles, and {\em different partial observations of these ensembles}, as well
@@
 %
-We illustrate the approach through two distinct sets of multiscale atomistic simulations: a fast/slow
-time scale simulation of an enzyme reaction network, and a molecular
-dynamics simulation of alanine dipeptide in explicit water.
+We illustrate the approach through two distinct sets of multiscale atomistic simulations:
+a simulation of an enzyme reaction network exhibiting both fats and slow time scales,
+ and a molecular dynamics simulation of alanine dipeptide in explicit water.
 
@@
 These developments have the potential to significantly enable the computational exploration
-of phyiscochemical problems.
+of physicochemical problems.
 %
-If the (high-dimensional) data $Y(t)$ arise from, say, a
-molecular dynamics simulation of a macromolecule in solution, or from the stochastic
+If the (high-dimensional) data $\mathbf{Y}(t)$ arise from, for example, a
+molecular dynamics simulation of a macromolecule in solution or the stochastic
 simulation of a complex chemical reaction scheme, the detection of a few good, coarse-grained
-``reduction coordinates" $x(t)$ is invaluable in understanding and predicting system behavior.
+``reduction coordinates" $\mathbf{x}(t)$ is invaluable in understanding and predicting system behavior.
 
@@
 It is well known that, even in the simple linear case of Principal Component Analysis,
-different data sets on the same low-dimensional hyperplane in the ambient space $Y$
-will lead to different basis vectors - in effect, to different $x$.
+different data sets on the same low-dimensional hyperplane in the ambient space
+will lead to different basis vectors - in effect, to different reduction coordinates $\mathbf{x}$.
 %
-While this can be easily fixed by a linear transformation, the problem becomes
+While this can be easily rectified by a linear transformation, the problem becomes
 exacerbated when the low-dimensional space is curved (a manifold, rather than a hyperplane)
-and when different data ensembles are obtained using different instrumental modalities
-(say, when one wants to merge molecular dynamics data with spectral information).
+and when different data sets are obtained using different instrumental modalities
+(such as when one wants to merge molecular dynamics data with spectral information).
 %
@@
 coordinate set, shared by all measurement ensembles and observation modalities -
-what we call a (Nonlinear) Intrinsic Variable Set - is invaluable.
+what we call Nonlinear Intrinsic Variables - is invaluable.
 %
@@
 %
-This latter empirical mapping requires good interpolation tools in embedding space; to this
-end we will demonstrate here the use of a multiscale Laplacian Pyramid approach \cite{rabin2012heterogeneous}.
+This empirical mapping requires good interpolation tools in embedding space; to this
+end, here we will demonstrate the use of a multiscale Laplacian Pyramid approach \cite{rabin2012heterogeneous}.
+
+We will demonstrate our methodologies with two distinct illustrative examples.
 %
-We chose two distinct multiscale illustrative examples. The first is a simulation
+The first is a simulation
 of two Goldbeter-Koshland modules in an enzyme kinetics model \cite{zagaris2012stability} using Gillespie's Stochastic Simulation
@@
 %
-This example will demonstrate how our techniques outperform more common techniques, such as diffusion maps for dimensionality reduction
-and nearest neighbor interpolation for reconstruction.
+This example will demonstrate how our techniques outperform more common techniques,
+such as diffusion maps for dimensionality reduction
+and nearest neighbor interpolation for interpolation.
 %
@@
 \end{equation}
-where $a_i$ are unknown drift functions and $\dot{w}_i(t)$ are independent white noises TODO: should there be a dot here?.
+where $a_i$ are unknown drift functions and $\dot{w}_i(t)$ are independent white noises.
 
@@
 %
-The idea would be to empirically construct a NIV coordinate system driven entirely by measurements that is invariant to the observation modality (see Figure \ref{fig:IntrinsicIllustration} for illustration).
+The idea is to empirically construct a NIV coordinate system driven entirely by measurements that is invariant to the observation function $f$ (see Figure \ref{fig:IntrinsicIllustration} for illustration).
 
@@
 %
-In this work, the identification of the associated ellipse of samples according to the time labels enables to explore and learn the tangent planes of the observable manifolds via the principal components of the covariance matrices of the samples in these ellipses.
+In this work, the identification of the associated ellipse of samples according to the time labels enables us
+to estimate the tangent planes of the observable manifolds via the principal components of the covariance matrices of the samples in these ellipses.
 %
@@
 \includegraphics[width=0.7\textwidth]{IntrinsicEmbeddingIllustration2}
-\caption{Illustration of the nonlinear embedding that yields an intrinsic representation independent of the measurement modality.
-(Bottom) The underlying variables in which the diffusion parameters are independent with unit variance.
-The circle illustrates successive sample in a short burst/trajectory in time that create a sphere on the parametric manifold.
+\caption{Illustration of the nonlinear embedding that yields an intrinsic representation independent of the measurement function $f$.
+(Bottom) The underlying variables in which the coordinates are independent with unit variance.
+The circle illustrates samples in a short trajectory in time that create a sphere on the manifold.
 (Top Left) The first set of observed variables.
-The ellipse illustrates the mapping of the sphere of the underlying samples into the observable domain via the first observation modality.
+The ellipse illustrates the mapping of the sphere of the underlying samples into the observable domain via the first observation function.
 In this domain, we illustrate that the observations might be partial, i.e., consist of merely a subset of the observed domain variables.
 (Top Right) The second set of observable variables.
-The ellipse illustrates the mapping of the sphere of the underlying samples into a different observable domain via a second observation modality.}
+The ellipse illustrates the mapping of the sphere of the underlying samples into a different observable domain via a second observation function.}
 \label{fig:IntrinsicIllustration}
@@
 In particular, we define a Riemannian metric between a pair of samples using the associated covariance matrices as
-\begin{align}
-	d(\mathbf{Y}(t), \mathbf{Y}(\tau)) &= 2 (\mathbf{Y}(t) - \mathbf{Y}(\tau))^T(\mathbf{C}(t) + \mathbf{C}(\tau))^{-1} \nonumber \\
-	&\times (\mathbf{Y}(t) - \mathbf{Y}(\tau)).
+\begin{equation}
+	d^2(\mathbf{Y}(t), \mathbf{Y}(\tau)) = 2 (\mathbf{Y}(t) - \mathbf{Y}(\tau))^T(\mathbf{C}(t) + \mathbf{C}(\tau))^{-1}(\mathbf{Y}(t) - \mathbf{Y}(\tau)).
 	\label{eq:mahalanobis}
-\end{align}
+\end{equation}
 This metric \eqref{eq:mahalanobis} is called the Mahalanobis distance.
@@
 This result implies that the Mahalanobis distance is invariant to the measurement function $f$, and hence,
-it yields the same distances between samples obtained under different observation modalities or even partial observations.
+it yields the same distances between samples obtained under different observation functions or even partial observations.
 
 \subsection{Laplace Operator}
-The Mahalanobis distance described in Section \ref{subsec:mahalanobis} enables to compare the observations in terms of the intrinsic variables of the associated underlying diffusion process.
+The Mahalanobis distance described in Section \ref{subsec:mahalanobis} enables us to compare observations in terms of the intrinsic variables of the associated underlying diffusion process.
 %
@@
 \end{equation}
-where $\varepsilon$ is the kernel scale, which can be set according to \cite{coifman2008graph}.
+where $\varepsilon$ is the kernel scale, which can be set according to \onlinecite{coifman2008graph, hein2005intrinsic}.
 %
-Let $\mathbf{D}$ be a diagonal matrix whose diagonal elements are the sums of rows of $\mathbf{W}$, and let $\mathbf{W}^{\mathrm{norm}} = \mathbf{D}^{-1/2}\mathbf{W}\mathbf{D}^{-1/2}$ be a normalized kernel that shares its eigenvectors with the normalize graph-Laplacian $\mathbf{I}-\mathbf{W}^{\mathrm{norm}}$ \cite{chung1997spectral}.
+Based on the kernel, we form a weighted graph, where the measurements $\mathbf{Y}(t)$ are the graph nodes and the weight of the edge connecting node $\mathbf{Y}(t)$ to node $\mathbf{Y}(\tau)$ is $W_{t,\tau}$.
 %
-These eigenvectors $\psi_j$ reveal the underlying structure of the data TODO:cite. 
+In particular, a Gaussian kernel exhibits a notion of locality by defining a neighborhood around each measurement $\mathbf{Y}(t)$ of radius $\varepsilon$, 
+i.e., measurements $\mathbf{Y}(\tau)$ s.t. $d^2(\mathbf{Y}(t), \mathbf{Y}(\tau) ) > \varepsilon$ are weakly connected to $\mathbf{Y}(t)$. 
+%
+In practice, we set $\varepsilon$ to be the median of the elements of the kernel matrix. 
+%
+According to the graph interpretation, it implies a well connected graph because each measurement is effectively connected to half of the other measurements.
+
+Let $\mathbf{D}$ be a diagonal matrix whose diagonal elements are the sums of rows of $\mathbf{W}$, and let $\mathbf{W}^{\mathrm{norm}} = \mathbf{D}^{-1/2}\mathbf{W}\mathbf{D}^{-1/2}$
+be a normalized kernel that shares its eigenvectors with the normalize graph-Laplacian $\mathbf{I}-\mathbf{W}^{\mathrm{norm}}$ \cite{chung1997spectral}.
+%
+The eigenvectors of $\mathbf{W}^{\mathrm{norm}}$, denoted $\psi_j$, reveal the underlying structure of the data \cite{coifman2005geometric}.
 %
@@
 %
+The eigenvectors are ordered such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_n|$, where $\lambda_j$ is the eigenvalue associated with eigenvector $\psi_j$. 
+%
+Because $\mathbf{W}^{\mathrm{norm}} \sim \mathbf{D}^{-1}\mathbf{W} $, and $\mathbf{D}^{-1}\mathbf{W}$ is row-stochastic, 
+$\lambda_1 = 1$ and $\psi_1$ is the diagonal of $\mathbf{D}^{1/2}$. 
+%
+The next few eigenvectors describe the geometry of the underlying manifold.
+%
 The above steps to construct the nonlinear intrinsic variables are summarized in Algorithm \ref{algo}.
-
-%TODO: do we need to describe the extension procedure?! - the recovery of the intrinsic variables through the eigenvalue decomposition (EVD) of the Laplace operator is computationally heavy...\\
-TODO: we can discuss the probabilistic interpretation and explicitly state that we set the kernel scale to the median of the kernel.\\
 
@@
 %
-Because eigenvectors can be determined up to a scaling factor, we must scale the eigenvectors from different data sets in a way so that the resulting embeddings are consistent.
+Because eigenvectors can be determined up to a scaling factor, we must scale the eigenvectors from different data sets such that the resulting embeddings are consistent.
 %
@@
 %
-Intuitively, a data set that has twice as many samples as another data set, but the same underlying density and shape, should have twice the norm.
+Intuitively, a data set that has twice as many samples as another data set, but the same underlying geometry, should have twice the norm.
 %
@@
 
-Laplacian Pyramids (LP) is a multi-scale algorithm for approximating and extending an empirical function $f$ defined on a set of points.
+Laplacian Pyramids (LP) is a multiscale algorithm for approximating and extending an empirical function $f$ defined on a set of points.
 %
-The algorithm uses Laplacian kernels of decreasing widths to create multi-scale representations of $f$.
-%
+The algorithm uses Laplacian kernels of decreasing widths to create multiscale representations of $f$;
 These representations can be easily extended to new data points.
 %
-This type of multi-scale representation was introduced by Burt and Adelson \cite{burt1983laplacian} for image coding,
+This type of multiscale representation was introduced by Burt and Adelson \cite{burt1983laplacian} for image coding,
 and was later shown to be a tight frame by Do and Veterli \cite{do2003framing}.
@@
 %
-We will first review the LP algorithm for approximating and extending a one-dimensional function.
-%
-We will then describe the application of LP in the context of this work - extending a high-dimensional function that is known on a set of points which lie in a low-dimensional space.
+We will first review the LP algorithm for approximating and extending a one-dimensional function,
+and then describe the application of LP in the context of this work - extending a high-dimensional function that is known on a set of points which lie in a low-dimensional space.
 
-Let $f: \Gamma \rightarrow \mathbb{R}^d$ be a function that is known on a subset of points $S \subset \Gamma$.
+Let $f: \Gamma \rightarrow \mathbb{R}^n$ be a function that is known on a subset of points $S \subset \Gamma$.
 %
@@
 %
-The smoothing operator $P_0$ is a normalized, coarse Laplacian kernel, defined by $p_0(t,\tau)= s_0^{-1}(t)w_0(t,\tau)$, where $w_0(t,\tau)=e^{-d^2(t,\tau) / \sigma_0}$. 
+The smoothing operator $P_0$ is a normalized, coarse Laplacian kernel, defined by $p_0(t,\tau)= s_0^{-1}(t)w_0(t,\tau)$,
+where $w_0(t,\tau)=e^{-d^2(t,\tau) / \sigma_0}$ and $s_0(t)=\sum_{\tau \in S}w_0(t,\tau)$ is the normalizing term.
 %
-The pairwise distance $d(t,\tau)$ is typically the Euclidian distance. 
-%
-The parameter $\sigma_0$ is set to be large compared to the values of $d(t,\tau)$  and $s_0^{-1}(t)=\sum_{\tau \in S}w_0(t,\tau)$ is the normalizing term. 
+The pairwise distance $d(t,\tau)$ is typically the Euclidian distance, and the parameter $\sigma_0$ is set to be large compared to the values of $d^2(t,\tau)$.
 %
@@
 
-The difference $\delta_1 = f-P_0(f)$ is input for the next iteration of the algorithm, 
+The difference $\delta_1 = f-P_0(f)$ is the input for the next iteration of the algorithm,
 which uses the smoothing operator $P_1$, $P_1 \propto e^{-d^2(t,\tau)^2 / 2^{-1} \sigma_0}$, to construct a coarse representation of $\delta_1$.
@@
 yields a new, finer representation of $f$, $f_1 = P_0(f) + P_1(\delta_1)$.
-
-In an iterative manner, multi-scales representations of the function $f$, denoted by $f_l$ are constructed (see Eq. \eqref{eq:LP_multi_scale}). 
 %
-As $l$ grows, the approximation is of a finer resolution because $P_l \propto e^{-d^2(t,\tau) / 2^{-l} \sigma_0}$ uses a Laplacian kernel of a finer width. 
-%
-The iterations stop when the difference between $f$ and $f_l$ is smaller than a pre-defined error threshold.
-
+In an iterative manner, multiscale representations of the function $f$, denoted $f_l$, are constructed.
 \begin{equation} \label{eq:LP_multi_scale}
@@
 \end{equation}
+As $l$ increases, the approximation becomes more refined because $P_l \propto e^{-d^2(t,\tau) / 2^{-l} \sigma_0}$ uses a Laplacian kernel of a finer width.
+%
+The iterations stop when the difference between $f$ and $f_l$ is smaller than a pre-defined error threshold.
 
-The representations $f_0$, $f_1$,..$f_l$ can be extended to a new point $\tilde{t} \in \Gamma \backslash S $ by extending the operators $P_0, P_1,\ldots,P_l$. 
+The representations $f_0, f_1, \dots, f_l$ can be extended to a new point $\tilde{t} \in \Gamma \backslash S $ by extending the operators $P_0, P_1,\ldots,P_l$.
 %
@@
 %
-The coarse regions of the function ( $0 \le x \le 4\pi$ ) are well approximated by a small number of scales. 
+The coarse regions of the function ($0 \le x \le 4\pi$) are well approximated by a small number of scales.
 %
-As the function becomes more oscillatory ( $\pi \le x \le 7.5\pi$ and $7.5\pi \le x \le 10\pi$), 
-a finer representation, (meaning a larger number of scales  $l$ )  is required to capture its behavior.
+As the function becomes more oscillatory ($\pi \le x \le 7.5\pi$ and $7.5\pi \le x \le 10\pi$),
+a finer representation (meaning a larger number of scales  $l$) is required to capture its behavior.
 \end{widetext}
@@
 %
-Let $\Psi(t) = \left(\psi_1(t),\psi_2(t),\ldots,\psi_d(t)\right)$ be the set of NIV's that were constructed from the data samples $\mathbf{Y}(t)$, as described in Algorithm \ref{algo}. 
+Let $\Psi(t) = \left(\psi_1(t),\psi_2(t),\ldots,\psi_d(t)\right)$ be the set of NIV that were constructed from the data samples $\mathbf{Y}(t)$, as described in Algorithm \ref{algo}.
 %
@@
 %
-The LP approximation of a new, high dimensional value $Y(\tilde{t})$, is calculated by a weighted average of the function values that belong to the neighboring points. 
+The LP approximation of a new, high dimensional value $Y(\tilde{t})$ is calculated by a weighted average of the function values that belong to the neighboring points.
 %
@@
 %
-In practice, a set of smoothing operators $P_0, P_1, \ldots, P_l$
+In practice, a set of smoothing operators $P_0, P_1, \ldots, P_l$, with
 \begin{equation} \label{eq:LP_multi_scale_app}
-P_l \propto e^{-d(\Psi(t),\Psi(\tau))^2 / \sigma_0 \cdot 2^{-l}}
+P_l \propto e^{-d^2(\Psi(t),\Psi(\tau)) / 2^{-l} \sigma_0},
 \end{equation}
@@
 \includegraphics[width=0.2\textwidth]{LP_figures/Scale11_S2}
-\caption{Approximation of function $f(x)$ as defined in \eqref{eq:LP_example} for scales 3, 5, 8, and 11 using Laplacian Pyramids (top, the true function is shown in blue and the Laplacian Pyramids approximation in shown in black), and the residual error in the approximation (bottom, shown in red).}
+\caption{(Top) Approximation of function $f(x)$ as defined in \eqref{eq:LP_example} for scales 3, 5, 8, and 11 using Laplacian Pyramids; the true function is shown in blue and the LP approximation is shown in black.
+(Bottom) The residual error in the LP approximation.}
 \label{fig:LP_ex}
@@
 The equations that govern the network are\\
-\begin{center}
-\ce{$E$ + $S$ <=>[e_1][e_{-1}] $E:S$ ->[e_2][] $E$ + $S^{*}$}\\
-\ce{$S$ + $E$ <=>[b_1][b_{-1}] $S:E$ ->[b_2][] $S$ + $E^{*}$}\\
-\ce{$D$ + $S^{*}$ <=>[d_1][d_{-1}] $D:S^{*}$ ->[d_2][] $D$ + $S$}\\
-\ce{$F$ + $E^{*}$ <=>[f_1][f_{-1}] $F:E^{*}$ ->[f_2][] $F$ + $E$}\\
-\end{center}
+\begin{equation}
+\begin{array}{rcl}
+E + S \overset{e_1}{\underset{e_{-1}}{\leftrightharpoons}} & E:S & \overset{e_2}{\rightarrow} E + S^{*} \\
+S + E \overset{b_1}{\underset{b_{-1}}{\leftrightharpoons}} & S:E & \overset{b_2}{\rightarrow} S + E^{*}\\
+D + S^{*} \overset{d_1}{\underset{d_{-1}}{\leftrightharpoons}} & D:S^{*} & \overset{d_2}{\rightarrow} D + S\\
+F + E^{*} \overset{f_1}{\underset{f_{-1}}{\leftrightharpoons}} & F:E^{*} & \overset{f_2}{\rightarrow} F + E
+\end{array}
+\end{equation}
 
-The $^{*}$ denote activated forms of the various species, and the : denote complexes formed between two species; the complexes $E:S$ and $S:E$ are not equivalent.
-
-There are 10 species in this reaction system;
-however, one can write four conservation equations (since total $E$, $S$, $D$, and $F$ are all conserved) to reduce the system to 6 dimensions.
+The ``$^{*}$'' denotes an activated form of a species, and the ``:'' denotes a complex formed between two species; the complexes $E:S$ and $S:E$ are not equivalent.
 %
-%We assume each reaction follows an elementary rate law.
+There are 10 species in this reaction system.
 %
-%We use the parameters $b_1=5$, $d_1=0.0009$, $e_1=0.1$, $f_1=0.1$, $b_{-1} = 10.6$, $d_{-1}=0.05$, $e_{-1}=0.5$, $f_{-1} =0.01$, $b_2=0.4$, $d_2=0.85$, $e_2=0.05$, and $f_2=2$.
+However, one can write four conservation equations (since total $E$, $S$, $D$, and $F$ are all conserved) to reduce the system to 6 dimensions.
 %
-%We take $S_T=E_T=D_T=1$, and $F_T=0.02$, where $S_T$, $E_T$, $D_T$, and $F_T$ are total concentrations of $S$, $E$, $D$, and $F$, respectively.
+In the parameter regime we consider, the system exhibits a separation of timescales, so that initial conditions quickly approach a two-dimensional manifold.
 %
-In the parameter regime we consider, the system exhibits a separation of timescales, so that initial conditions quickly approach a two-dimensional manifold;
-details about the specific parameters can be found in Appendix \ref{app:rxn}.
-%
-%The relevant timescales around the steady state ($-1/\lambda_i$, where $\lambda_i$ are the eigenvalues of the Hessian) are 1176, 9.731, 1.594, 1.111, 0.4975, 0.06498.
+Details about the specific parameters can be found in Appendix \ref{app:rxn}.
 
@@
 
-We generate 3000 random initial conditions $\mathbf{Y}^{full}_0(1), \dots, \mathbf{Y}^{full}_0(3000) \in \mathbb{R}^6$, enforcing that all concentrations must be non-negative
-(we take the order of the components in the vector $\mathbf{Y}^{full}_0(t)$ to be $S$, $E$, $E:S$, $S:E$, $D:S^{*}$, $F:E^{*}$).
+We generate 3000 random initial conditions $\mathbf{Y}_0(1), \dots, \mathbf{Y}_0(3000) \in \mathbb{R}^6$, enforcing that all concentrations must be non-negative
+(we take the order of the components in the vector $\mathbf{Y}_0(t)$ to be $S$, $E$, $E:S$, $S:E$, $D:S^{*}$, $F:E^{*}$).
 %
-We integrate each point $\mathbf{Y}^{full}_0(t)$ forward for 10 time units using the SSA to obtain point the $\mathbf{Y}^{full}(t) \in \mathbb{R}^6$;
+We integrate each point $\mathbf{Y}_0(t)$ forward for 10 time units using the SSA to obtain point the $\mathbf{Y}(t) \in \mathbb{R}^6$;
 10 time units is sufficiently long for the initial points to converge to the two-dimensional manifold,
@@
 %
-We take $\mathbf{Y}^{full}(1), \dots, \mathbf{Y}^{full}(3000)$ to be representative points on the underlying two-dimensional manifold;
+We consider $\mathcal{Y} = \{ \mathbf{Y}(t): t=1, \dots, 3000 \}$ to be representative points on the underlying two-dimensional manifold;
 three-dimensional projections of the resulting manifold are shown in Figure \ref{fig:rxn_manifolds}.
 %
-From each manifold point $\mathbf{Y}^{full}(t)$, we run 20 short simulations, each for 0.2 time units.
+From each manifold point $y \in \mathcal{Y}$, we run 20 short simulation ``bursts'', each for 0.2 time units.
 %
-We denote the endpoints from the short simulations as $\widetilde{\mathbf{Y}}^{full}(t, 1), \dots, \widetilde{\mathbf{Y}}^{full}(t, 20)$.
+We denote the endpoints from the short simulations as $\mathcal{Y}^{burst}(y)$.
 
@@
   \includegraphics[height=6cm]{rxn_figures/rxn_manifold4}\llap{\raisebox{3.5cm}{\includegraphics[height=2.5cm]{rxn_figures/rxn_manifold4_2}}} \\
-    \caption{Projections of the manifold from stochastic simulation of the chemical reaction network.
+    \caption{Projections of the data obtained from stochastic simulation of the chemical reaction network.
 The insets show rotations of the projections to illustrate the two-dimensionality of the manifold.}
@@
 %
-Data set 1, denoted $\mathbf{Y}_1(t) \in \mathbb{R}^4$, consists of $\mathbf{Y}^{full}(1), \dots, \mathbf{Y}^{full}(2000)$, restricted to components $S$, $E$, $S:E$, and $F:E^{*}$, i.e.
-$$\mathbf{Y}_1(t) =
+Data set 1, denoted $\mathcal{Y}_1$, consists of $\mathbf{Y}(1), \dots, \mathbf{Y}(2000)$, restricted to components $S$, $E$, $S:E$, and $F:E^{*}$, i.e.
+$$\mathcal{Y}_1 = \left\{
 \left( \begin{array}{cccccc}
@@
 0 & 0 & 0 & 0 & 0 & 1
-\end{array} \right) \mathbf{Y}^{full}(t), \: t=1, \dots, 2000.$$
+\end{array} \right) \mathbf{Y}(t) \in \mathbb{R}^4: t=1, \dots, 2000 \right\}.$$
 %
-Data set 2, denoted $\mathbf{Y}_2(t) \in \mathbb{R}^4$, consists of $\mathbf{Y}^{full}(1500), \dots, \mathbf{Y}^{full}(3000)$, restricted to components $S$, $E$, $E:S$, and $D:S^{*}$, i.e.
-$$\mathbf{Y}_2(t) =
+Data set 2, denoted $\mathcal{Y}_2$, consists of $\mathbf{Y}(1500), \dots, \mathbf{Y}(3000)$, restricted to components $S$, $E$, $E:S$, and $D:S^{*}$, i.e.
+$$\mathcal{Y}_2 = \left\{
 \left( \begin{array}{cccccc}
@@
 \end{array} \right)
-\mathbf{Y}^{full}(t + 1499), \: t=1, \dots, 1501.$$
+\mathbf{Y}(t) \in \mathbb{R}^4: t=1500, \dots, 3000 \right\}.$$
 %
-The endpoints of the simulation bursts for the two data sets, $\widetilde{\mathbf{Y}_1}(t, \tau)$ and $\widetilde{\mathbf{Y}_2}(t, \tau)$, are defined analogously.
+The endpoints of the simulation bursts for the two data sets, $\mathcal{Y}^{burst}_1(y)$ and $\mathcal{Y}^{burst}_2(y)$, are defined analogously.
 %
@@
 \begin{equation}
-\widehat{\mathbf{C}_j}(t) = \sum_{\tau=1}^{20} \left( \widetilde{\mathbf{Y}_j}(t, \tau) - \hat{\mathbf{\mu}}_j(t) \right)\left( \widetilde{\mathbf{Y}_j}(t, \tau) - \hat{\mathbf{\mu}}_j(t) \right)^T, i \in \{1, 2\}
+\widehat{\mathbf{C}_i}(y) = \sum_{z \in \mathcal{Y}^{burst}_i(y)} \left( z - \hat{\mathbf{\mu}}_i(y) \right)\left( z - \hat{\mathbf{\mu}}_i(y) \right)^T, y \in \mathcal{Y}_i, i \in \{1, 2\}
 \end{equation}
-where $\hat{\mathbf{\mu}_j}$ is the empirical mean of $\widetilde{\mathbf{Y}_j}(t, 1), \dots, \widetilde{\mathbf{Y}_j}(t, 20)$.
+where $\hat{\mathbf{\mu}}_i(y)$ is the empirical mean of $\mathcal{Y}^{burst}_i(y)$.
 %
@@
 %
-We also compute the correlation between the embedding coordinates for $\mathbf{Y}_1(1500), \dots, \mathbf{Y}_1(2000)$ and $\mathbf{Y}_2(1), \dots, \mathbf{Y}_2(501)$,
-since both come from the same underlying points $\mathbf{Y}^{full}(1500), \dots, \mathbf{Y}^{full}(2000)$.
+We also note that both $\mathcal{Y}_1$ and $\mathcal{Y}_2$ contain points that are projections of $\mathbf{Y}(1500), \dots, \mathbf{Y}(2000)$.
 %
-We obtain a correlation of 0.97 and 0.95 for the first and second NIVs, respectively, indicating that the two embeddings are in excellent agreement with each other.
+We therefore compute the correlation between the embedding coordinates for these points common to $\mathcal{Y}_1$ and $\mathcal{Y}_2$.
+%
+We obtain a correlation of 0.97 and 0.95 for the first and second NIV, respectively, indicating that the two embeddings are in excellent agreement with each other.
 
@@
 
-We then use Laplacian Pyramids to predict $S:E$ and $F:E^{*}$ for $\mathbf{Y}_2(t)$.
+We then use LP to predict $S:E$ and $F:E^{*}$ for $\mathcal{Y}_2$.
 %
-Because $\mathbf{Y}_1(t)$ and $\mathbf{Y}_2(t)$ are measured for different components, there is no simple way to estimate $S:E$ and $F:E^{*}$ from the data in the observation space.
+Because $\mathcal{Y}_1$ and $\mathcal{Y}_2$ are measured for different components, there is no simple way to estimate $S:E$ and $F:E^{*}$ directly in the observation space.
 %
@@
 %
-We use $\mathbf{Y}_1(t)$ to train a Laplacian Pyramids function from the two-dimensional NIV embedding to $S:E$ and $F:E^{*}$.
+We use $\mathcal{Y}_1$ to train an LP function from the two-dimensional NIV embedding to $S:E$ and $F:E^{*}$.
 %
-We then use this function to predict the values  of $S:E$ and $F:E^{*}$ for $\mathbf{Y}_2(t)$, using the computed NIV embedding for $\mathbf{Y}_2(t)$.
+We then use this function to predict the values  of $S:E$ and $F:E^{*}$ for $\mathcal{Y}_2$, using the computed NIV embedding for $\mathcal{Y}_2$.
 %
@@
 
-The results of the Laplacian Pyramids prediction are shown in Figure \ref{fig:rxn_recon}.
+The results of the LP prediction are shown in Figure \ref{fig:rxn_recon}.
 %
@@
 %
-Therefore, we can effectively predict the unobserved components in the reaction network using NIV together with Laplacian Pyramids.
+Therefore, we can effectively predict the unobserved components in the reaction network using NIV together with LP.
 
@@
     \includegraphics[width=0.4\textwidth]{rxn_figures/rxn_recon6}
-    \caption{Laplacian Pyramids (LapPyr) reconstructions of $S:E$ (left) and $F:E^{*}$ (right).}
+    \caption{LP reconstructions of $S:E$ (left) and $F:E^{*}$ (right).}
     \label{fig:rxn_recon}
@@
 %
-Alanine dipeptide (Ala2) is often used as a ``prototypical'' protein for simulation studies (TODO: cite Ala2 papers).
+Alanine dipeptide (Ala2) is often used as a ``prototypical'' protein for simulation studies
+\cite{apostolakis1999calculation, bolhuis2000reaction, chekmarev2004long, ma2005automatic, frewen2009exploration, ferguson2011integrating}.
 %
@@
 %
-The molecule is solvated with 638 TIP3P water molecules \cite{jorgensen1983comparison}, with the particle mesh Ewald method for long-range electrostatic interactions \cite{essmann1995smooth} and periodic boundary conditions.
+The molecule is solvated with 638 TIP3P water molecules \cite{jorgensen1983comparison}, with periodic boundary conditions and the particle mesh Ewald method used for long-range electrostatic interactions \cite{essmann1995smooth}.
 %
@@
     \includegraphics[width=0.3\textwidth]{molecule_labeled2}
-    \caption{(Left) Free energy surface for Alanine Dipeptide.
+    \caption{(Left) Free energy surface for Ala2.
 The minima are labeled A, B, and C, and the corresponding molecular configurations are shown.
-(Right) The molecular structure of alanine dipeptide, excluding the hydrogens.
+(Right) The molecular structure of Ala2, excluding the hydrogens.
 The atoms are numbered and the two dihedral angles $\phi$ and $\psi$ are indicated.}
@@
 
-We first wish to demonstrate the benefit of NIV over other nonlinear dimensionality reduction techniques, such as diffusion maps.
+We first wish to demonstrate the benefit of NIV over other nonlinear dimensionality reduction techniques, such as diffusion maps \cite{coifman2005geometric}.
 %
-We consider 10,000 data points from our simulation $\mathbf{Y}^{full}(1), \dots, \mathbf{Y}^{full}(10000) \in \mathbb{R}^{30}$; every 100 data points comes from a continuous simulation trajectory.
+We consider 10,000 data points from our simulation $\mathbf{Y}(1), \dots, \mathbf{Y}(10000) \in \mathbb{R}^{30}$; every 100 data points comes from a continuous simulation trajectory.
 %
@@
 %
-$\mathbf{Y}_{even}(t), t=1, \dots, 10000$ is $\mathbf{Y}^{full}(t)$ restricted to atoms 2, 4, 6, 8, and 10,
-and $\mathbf{Y}_{odd}(t), t=1, \dots, 10000$ is $\mathbf{Y}^{full}(t)$ restricted to atoms 5, 7, and 9 (see Figure \ref{fig:ala_fes} for the atom indexing).
+$\mathcal{Y}_{even} = \{\mathbf{Y}(t) \text{ restricted to atoms 2, 4, 6, 8, and 10}: t=1, \dots, 10000 \}$,
+and $\mathcal{Y}_{odd} = \{\mathbf{Y}(t) \text{ restricted to atoms 5, 7, and 9}: t=1, \dots, 10000 \}$ (see Figure \ref{fig:ala_fes} for the atom indexing).
 %
-We then compute the NIV and diffusion maps embeddings for $\mathbf{Y}_{even}(t)$ and $\mathbf{Y}_{odd}(t)$.
+We then compute the NIV and diffusion maps embeddings for $\mathcal{Y}_{even}$ and $\mathcal{Y}_{odd}$.
 %
@@
 (as neighboring points in our data set that do not come from the
-same trajectories should not contribute to the estimate of the covariance).
+same trajectories should not contribute to the estimate of the local covariance).
 
@@
 The correlation between the two NIV embeddings is higher than the correlation between the two diffusion maps embeddings,
-demonstrating that it is often advantageous to use NIV over diffusion maps if one wishes to obtain a consistent embedding and ``merge'' data sets from different observation domains.
+demonstrating that it is often advantageous to use NIV over diffusion maps if one wishes to obtain a consistent embedding and merge data sets from different observation domains.
 
@@
 
-We then use NIV together with Laplacian Pyramids to predict the conformation of Ala2 when we only observe some of the atoms.
+We then use NIV together with LP to predict the conformation of Ala2 when we only observe some of the atoms.
 %
-We have 20000 data points $\mathbf{Y}^{full}(1), \dots, \mathbf{Y}^{full}(20000)$, where every 100 data points comes from one continuous simulation trajectory.
+We have 20000 data points $\mathbf{Y}(1), \dots, \mathbf{Y}(20000)$, where every 100 data points comes from one continuous simulation trajectory.
 %
-Our first data set (which will serve as our training data for Laplacian Pyramids), $\mathbf{Y}_{all}(t)$, consists of the first 10000 data points ($\mathbf{Y}_{all}(t) = \mathbf{Y}^{full}(t), t=1, \dots, 10000$).
+Our first data set (which will serve as our training data for LP), $\mathcal{Y}_{all}$,
+consists of the first 10000 data points ($\mathcal{Y}_{all} = \{\mathbf{Y}(t): t=1, \dots, 10000\}$).
 %
-Our second data set (which will serve as our test data), $\mathbf{Y}_{odd}(t)$, consists of the last 12000 data points restricted to {\em only} the odd atoms ($\mathbf{Y}_{odd}(t) = \mathbf{Y}^{full}(t+8000)$ restricted to the odd atoms, $t = 1, \dots, 12000$).
+Our second data set (which will serve as our test data), $\mathcal{Y}_{odd}$, consists of the last 12000 data points restricted to {\em only} the odd atoms
+($\mathcal{Y}_{odd} = \{ \mathbf{Y}(t) \text{ restricted to the odd atoms}: t = 8001, \dots, 20000\}$).
 %
@@
 %
-We compute the NIV embedding for the training data $\mathbf{Y}_{all}(t)$ and the test data $\mathbf{Y}_{odd}(t)$; we then use the Laplacian Pyramids from the training data to predict the location of all the atoms for each point in the test data.
+We compute the NIV embedding for the training data $\mathcal{Y}_{all}$ and the test data $\mathcal{Y}_{odd}$; we then use LP interpolation from the training data to predict the location of all the atoms for each point in the test data.
 
-The NIV embedding for the training data $Y_{all}(t)$ is shown in Figure \ref{fig:ala_embed}.
+The NIV embedding for the training data $\mathcal{Y}_{all}$ is shown in Figure \ref{fig:ala_embed}.
 %
-The embedding is three-dimensional; the first coordinate describes the flipping of atoms 1 and 3, the second coordinate describes the dihedral angle $\phi$, and the third coordinate describes the dihedral angle $\psi$.
+The embedding is three-dimensional: the first coordinate describes the flipping of atoms 1 and 3, the second coordinate describes the dihedral angle $\phi$, and the third coordinate describes the dihedral angle $\psi$.
 %
-We calculate the correlation between the embedding coordinates for $\mathbf{Y}_{all}(8001), \dots, \mathbf{Y}_{all}(10000)$, and $\mathbf{Y}_{odd}(1), \dots, \mathbf{Y}_{odd}(2000)$,
-since they come from the same simulation data points $\mathbf{Y}^{full}(8001), \dots, \mathbf{Y}^{full}(10000)$.
+We calculate the correlation between the embedding coordinates for the points in $\mathcal{Y}_{all}$ and $\mathcal{Y}_{odd}$
+that come from the common simulation data points $\mathbf{Y}(8001), \dots, \mathbf{Y}(10000)$.
 %
-The embeddings for the two data sets are found to be fairly consistent, with correlations of 0.97, 0.72, 0.85 for the first, second, and third NIVs, respectively.
+The embeddings for the two data sets are found to be fairly consistent, with correlations of 0.97, 0.72, 0.85 for the first, second, and third NIV, respectively.
 
@@
     \includegraphics[width=0.3\textwidth]{ala_figures_allodd/ala2_embed3}
-    \caption{The 3-dimensional NIV embedding for Alanine Dipeptide computed using $\mathbf{Y}_{all}(t)$, colored by the y-coordinate of the first atom (left), the dihedral angle $\phi$ (center), and the dihedral angle $\psi$ (right).
+    \caption{The 3-dimensional NIV embedding for Ala2 computed using $\mathcal{Y}_{all}$, colored by the y-coordinate of the first atom (left), the dihedral angle $\phi$ (center), and the dihedral angle $\psi$ (right).
 Each embedding is rotated so that the correlation between the colors and the relevant NIV can easily be seen.}
@@
 %
-Figure \ref{fig:ala_molecules} shows molecular structures for the true and reconstructed configurations for some data points.
+Figure \ref{fig:ala_molecules} shows molecular structures for the true and reconstructed configurations for some data points;
+there is nice agreement between the true and reconstructed configurations.
+
+We would also like to demonstrate the effectiveness of LP over other reconstruction techniques.
 %
-There is nice agreement between the true and reconstructed configurations.
-
-We would like to also demonstrate the effectiveness of Laplacian Pyramids over other reconstruction techniques.
-%
-We also predict the configurations from the NIV components using simple nearest-neighbor interpolation.
+We predict the configurations from the NIV components using simple nearest-neighbor interpolation.
 %
@@
 %
-The Laplacian Pyramid approach outperforms simple nearest neighbor search for all of the atoms.
+The LP approach outperforms simple nearest neighbor search for all of the atoms.
 
@@
       }
-  \caption{The correlation between the true position and the reconstructed position (using Laplacian Pyramids) for the test data.
+  \caption{The correlation between the true position and the reconstructed position (using LP) for the test data.
 The columns correspond the x-, y-, and z-coordinates, and the rows correspond to atoms 1, 2, 3, 6, and 8.}
@@
     \caption{Mean squared error of reconstructed position, normalized by the average bond length, for each atom in Ala2.
-    The positions were reconstructed using both Laplacian Pyramids (LapPyr) and nearest neighbor interpolation (nn).}
+    The positions were reconstructed using both LP and nearest neighbor (nn) interpolation .}
     \label{fig:ala_mse}
@@
 We consider the following network of chemical reactions:
-\begin{center}
-\ce{$E$ + $S$ <=>[e_1][e_{-1}] $E:S$ ->[e_2][] $E$ + $S^{*}$}\\
-\ce{$S$ + $E$ <=>[b_1][b_{-1}] $S:E$ ->[b_2][] $S$ + $E^{*}$}\\
-\ce{$D$ + $S^{*}$ <=>[d_1][d_{-1}] $D:S^{*}$ ->[d_2][] $D$ + $S$}\\
-\ce{$F$ + $E^{*}$ <=>[f_1][f_{-1}] $F:E^{*}$ ->[f_2][] $F$ + $E$}\\
-\end{center}
+\begin{equation}
+\begin{array}{rcl}
+E + S \overset{e_1}{\underset{e_{-1}}{\leftrightharpoons}} & E:S & \overset{e_2}{\rightarrow} E + S^{*} \\
+S + E \overset{b_1}{\underset{b_{-1}}{\leftrightharpoons}} & S:E & \overset{b_2}{\rightarrow} S + E^{*}\\
+D + S^{*} \overset{d_1}{\underset{d_{-1}}{\leftrightharpoons}} & D:S^{*} & \overset{d_2}{\rightarrow} D + S\\
+F + E^{*} \overset{f_1}{\underset{f_{-1}}{\leftrightharpoons}} & F:E^{*} & \overset{f_2}{\rightarrow} F + E
+\end{array}
+\end{equation}
 
@@
 We take $S_T=E_T=D_T=1V$, and $F_T=0.02V$, where $S_T$, $E_T$, $D_T$, and $F_T$ are total number of $S$, $E$, $D$, and $F$, respectively.
+%
+In this parameter regime, the relevant timescales around the steady state ($-1/\lambda_i$, where $\lambda_i$ are the eigenvalues of the Hessian) are 1176, 9.731, 1.594, 1.111, 0.4975, 0.06498.
+%
+Therefore, we choose to integrate forward for 10 time units to find points on a 2-dimensional manifold.
 

### 2013-12-27 paper/draft4.tex
Added files related to paper

--- 2013-12-27 paper/draft3.tex
@@
 \usepackage{natbib}
-%\usepackage{mhchem}
 
@@
 We illustrate the approach through two distinct sets of multiscale atomistic simulations:
-a simulation of an enzyme reaction network exhibiting both fats and slow time scales,
+a stochastic simulation of an enzyme reaction network exhibiting both fast and slow time scales,
  and a molecular dynamics simulation of alanine dipeptide in explicit water.
@@
 The last decade has witnessed extensive advances in dimensionality reduction techniques:
-finding meaningful low-dimensional descriptions of high-dimensional data.
+finding meaningful low-dimensional descriptions of high-dimensional data (TODO: add references).
 %
@@
 simulation of a complex chemical reaction scheme, the detection of a few good, coarse-grained
-``reduction coordinates" $\mathbf{x}(t)$ is invaluable in understanding and predicting system behavior.
+``reduction coordinates" $\mathbf{x}(t)$ can be invaluable in understanding and predicting system behavior.
 
-%
 While the benefits from such reduced descriptions are manifest, a crucial shortcoming of
-such data-driven reduction coordinates is their dependence on the specific data set processed,
+ data-driven reduction coordinates is their dependence on the specific data set processed,
 and not only on the physical model in question.
 %
-It is well known that, even in the simple linear case of Principal Component Analysis,
+It is well known that, even in the simple linear case of Principal Component Analysis \cite{jolliffe2005principal},
 different data sets on the same low-dimensional hyperplane in the ambient space
@@
 and when different data sets are obtained using different instrumental modalities
-(such as when one wants to merge molecular dynamics data with spectral information).
+(such as when one wants to merge molecular dynamics data with, for example, spectral information TODO: add specific example of spectral information).
 %
@@
 %
-This empirical mapping requires good interpolation tools in embedding space; to this
+This empirical mapping also requires good interpolation tools in embedding space; to this
 end, here we will demonstrate the use of a multiscale Laplacian Pyramid approach \cite{rabin2012heterogeneous}.
@@
 The first is a simulation
-of two Goldbeter-Koshland modules in an enzyme kinetics model \cite{zagaris2012stability} using Gillespie's Stochastic Simulation
-Algorithm (SSA); in certain parameter regimes, separation of time scales is known
-to reduce this model to an effective two-dimensional description.
+of two Goldbeter-Koshland modules in an enzyme kinetics model using Gillespie's Stochastic Simulation
+Algorithm (SSA) \cite{gillespie1977exact}; 
+in certain parameter regimes, separation of time scales is known
+to reduce this model to an effective two-dimensional description \cite{zagaris2012stability}.
 %
-Although this example is rather simple, it will serve as a nice introduction to our techniques and highlight the main features of the algorithms.
+Although this example is rather simple, it will serve as an introduction to our techniques and highlight the main features of the algorithms.
 %
@@
 simulation (in explicit water) of a simple peptide fragment (alanine dipeptide) whose folding
-dynamics are known to be described through a small set of physical observables.
+dynamics are known to be described through a small set of physical observables \cite{bolhuis2000reaction}.
 %
-This example will demonstrate how our techniques outperform more common techniques,
+This example will demonstrate how our techniques can have advantages over more common techniques,
 such as diffusion maps for dimensionality reduction
-and nearest neighbor interpolation for interpolation.
+and nearest neighbor interpolation for reconstruction.
 %
@@
 Let $\mathbf{Y}(t)$ be a high-dimensional measured process in $\mathbb{R}^n$ consisting of $n$ observable variables.
-The measured process is assumed to be a manifestation in an observable domain of a low-dimensional diffusion process. Thus, it can be expressed by
+%
+The measured process is assumed (and this is a critical assumption) to be a manifestation in an observable domain of a low-dimensional diffusion process. Thus, it can be expressed by
 \begin{equation}
@@
 \end{equation}
-where $f:\mathbb{R}^d \rightarrow \mathbb{R}^n$ is an unknown (possibly nonlinear) function and $\mathbf{x}(t)$ is a diffusion process that consists of $d$ underlying variables.
+where $f:\mathbb{R}^d \rightarrow \mathbb{R}^n$ is an unknown (possibly nonlinear, $d \ll n$) function and $\mathbf{x}(t)$ is a diffusion process that consists of $d$ underlying variables.
 %
@@
 \end{equation}
-where $a_i$ are unknown drift functions and $\dot{w}_i(t)$ are independent white noises.
+where $a_i$ are unknown drift functions and $\dot{w}_i(t)$ are independent white noises (the independence is another critical assumption).
 
@@
 %
-By relying on the fact that the empirical method we will describe is independent of the observation function $f$, we present the notion of intrinsic modeling and refer to the coordinates of $\mathbf{x}(t)$ as a Nonlinear Intrinsic Variable (NIV) set.
+By relying on the fact that the empirical method we will describe is independent of the observation function $f$, 
+we refer to the coordinates of $\mathbf{x}(t)$ as Nonlinear Intrinsic Variables (NIV).
 %
-The available samples $\mathbf{Y}(t)$ may be the product of different measurement functions $f$ in various observable domains, or they maybe partial measurements consisting of merely a subset of the coordinates of the observable domains.
+The available samples $\mathbf{Y}(t)$ may be the result of different measurement functions $f$ in various observable domains, 
+or they may be partial measurements consisting of merely a subset of the coordinates of the observable domains.
 %
-The idea is to empirically construct a NIV coordinate system driven entirely by measurements that is invariant to the observation function $f$ (see Figure \ref{fig:IntrinsicIllustration} for illustration).
+The idea is to empirically construct a NIV coordinate system driven entirely by measurements that is invariant to the observation function $f$ (see Figure \ref{fig:IntrinsicIllustration} for a schematic illustration).
 
@@
 In this work, the identification of the associated ellipse of samples according to the time labels enables us
-to estimate the tangent planes of the observable manifolds via the principal components of the covariance matrices of the samples in these ellipses.
+to estimate the tangent planes of the observable manifolds {\em via the principal components of the covariance matrices of the samples in these ellipses}.
 %
-(2) The principal directions of the tangent planes are utilized to define a Riemannian metric that is shown to be locally invariant to the measurement function $f$.
+(2) The principal directions of the tangent planes are utilized to define a Riemannian metric that is shown to be {\em locally invariant to the measurement function $f$}.
 %
-(3) The NIV are constructed through the eigenvalue decomposition of a Laplace operator that is built upon a pairwise affinity between the samples, defined using the Riemannian metric.
+(3) The NIV are constructed through the eigenvalue decomposition of a Laplace operator that is built upon a pairwise affinity between the samples, defined using this Riemannian metric.
 
@@
 \caption{Illustration of the nonlinear embedding that yields an intrinsic representation independent of the measurement function $f$.
-(Bottom) The underlying variables in which the coordinates are independent with unit variance.
-The circle illustrates samples in a short trajectory in time that create a sphere on the manifold.
+(Bottom) The underlying variables in which the noises are independent with unit variance.
+The circle illustrates samples from, say, a short trajectory in time that sample a disc on the manifold.
 (Top Left) The first set of observed variables.
 The ellipse illustrates the mapping of the sphere of the underlying samples into the observable domain via the first observation function.
-In this domain, we illustrate that the observations might be partial, i.e., consist of merely a subset of the observed domain variables.
-(Top Right) The second set of observable variables.
-The ellipse illustrates the mapping of the sphere of the underlying samples into a different observable domain via a second observation function.}
+In this sketch, we illustrate that the observations might be {\em partial}, i.e., might consist of merely a subset of the observed domain variables.
+(Top Right) Second set of observable variables.
+The ellipse illustrates the mapping of the sphere of the underlying samples into this (different) observable domain via a second observation function.}
 \label{fig:IntrinsicIllustration}
@@
 where $\widehat{\boldsymbol{\mu}}(t)$ is the empirical mean of the samples.
-
-In particular, we define a Riemannian metric between a pair of samples using the associated covariance matrices as
+%
+We define a Riemannian metric between a pair of samples using the associated covariance matrices as
 \begin{equation}
-	d^2(\mathbf{Y}(t), \mathbf{Y}(\tau)) = 2 (\mathbf{Y}(t) - \mathbf{Y}(\tau))^T(\mathbf{C}(t) + \mathbf{C}(\tau))^{-1}(\mathbf{Y}(t) - \mathbf{Y}(\tau)).
+	d^2(\mathbf{Y}(t), \mathbf{Y}(\tau)) = 2 (\mathbf{Y}(t) - \mathbf{Y}(\tau))^T(\mathbf{C}(t) + \mathbf{C}(\tau))^{-1}(\mathbf{Y}(t) - \mathbf{Y}(\tau));
 	\label{eq:mahalanobis}
 \end{equation}
-This metric \eqref{eq:mahalanobis} is called the Mahalanobis distance.
+this is the Mahalanobis distance.
 %
@@
 %
-This information is then utilized in \eqref{eq:mahalanobis} to compare a pair of points according to the directions of their respected tangent planes.
-
+This information is then utilized in \eqref{eq:mahalanobis} to compare a pair of points according to the directions of their respective tangent planes.
+%
 The Mahalanobis distance is invariant under linear transformations.
 %
-Thus, by assuming that the observation function $f$ is bi-Lipschitz and smooth and by using local linearization of the function, i.e., $\mathbf{Y}(t) = \mathbf{J}(t) \mathbf{x}(t) + \boldsymbol{\epsilon}(t)$ where $\mathbf{J}(t)$ is the Jacobian of $f(\mathbf{x}(t))$ and $\boldsymbol{\epsilon}(t)$ is the residual consisting of higher-order terms, it was shown by Singer and Coifman \cite{singer2008non} that $\mathbf{C}(t) = \mathbf{J}(t)\mathbf{J}^T(t)$ and that the Mahalanobis distance approximates the Euclidean distance between the corresponding samples of the underlying process to a second order, i.e.,
+Thus, by assuming that the observation function $f$ is bi-Lipschitz and smooth and by using local linearization of the function, i.e., $\mathbf{Y}(t) = \mathbf{J}(t) \mathbf{x}(t) + \boldsymbol{\epsilon}(t)$ where $\mathbf{J}(t)$ is the Jacobian of $f(\mathbf{x}(t))$ and $\boldsymbol{\epsilon}(t)$ is the residual consisting of higher-order terms, it was shown by Singer and Coifman \cite{singer2008non} that $\mathbf{C}(t) = \mathbf{J}(t)\mathbf{J}^T(t)$ and that the Mahalanobis distance approximates the Euclidean distance between the corresponding samples of the underlying process to second order, i.e.,
 \begin{equation}
@@
 it yields the same distances between samples obtained under different observation functions or even partial observations. 
+%
+We would like to note that, in general, $f$ being bi-Lipschitz implies that $f$ is invertible.
+%
+However, in practice, determining whether $f$ contains sufficient information and is ``rich enough'' to completely determine the underlying process is a non-trivial task.
 
@@
 %
-In this section, we recover the underlying process itself from the pairwise Euclidean distances through the eigenvectors of a Laplace operator.
+In this section, we show how to recover the underlying process itself from the pairwise Euclidean distances through the eigenvectors of a Laplace operator.
 
@@
 \end{equation}
-where $\varepsilon$ is the kernel scale, which can be set according to \onlinecite{coifman2008graph, hein2005intrinsic}.
+where $\varepsilon$ is the kernel scale, which can be set according to References \onlinecite{coifman2008graph, hein2005intrinsic}.
 %
@@
 In particular, a Gaussian kernel exhibits a notion of locality by defining a neighborhood around each measurement $\mathbf{Y}(t)$ of radius $\varepsilon$,
-i.e., measurements $\mathbf{Y}(\tau)$ s.t. $d^2(\mathbf{Y}(t), \mathbf{Y}(\tau) ) > \varepsilon$ are weakly connected to $\mathbf{Y}(t)$. 
+i.e., measurements $\mathbf{Y}(\tau)$ such that $d^2(\mathbf{Y}(t), \mathbf{Y}(\tau) ) > \varepsilon$ are weakly connected to $\mathbf{Y}(t)$.
 %
@@
 %
-According to the graph interpretation, it implies a well connected graph because each measurement is effectively connected to half of the other measurements.
+According to the graph interpretation, this implies a well connected graph because each measurement is effectively connected to half of the other measurements \cite{rohrdanz2011determination}.
 
-Let $\mathbf{D}$ be a diagonal matrix whose diagonal elements are the sums of rows of $\mathbf{W}$, and let $\mathbf{W}^{\mathrm{norm}} = \mathbf{D}^{-1/2}\mathbf{W}\mathbf{D}^{-1/2}$
-be a normalized kernel that shares its eigenvectors with the normalize graph-Laplacian $\mathbf{I}-\mathbf{W}^{\mathrm{norm}}$ \cite{chung1997spectral}.
+Let $\mathbf{D}$ be a diagonal matrix whose elements are the sums of rows of $\mathbf{W}$, and let $\mathbf{W}^{\mathrm{norm}} = \mathbf{D}^{-1/2}\mathbf{W}\mathbf{D}^{-1/2}$
+be a normalized kernel that shares its eigenvectors with the normalized graph-Laplacian $\mathbf{I}-\mathbf{W}^{\mathrm{norm}} \:$ \cite{chung1997spectral}.
 %
@@
 %
+However, some eigenvectors can be higher harmonics of the same principal direction (similar to how the eigenfunctions $\cos x$ and $\cos 2x$ both characterize the $x$ direction), 
+and so one must check for correlations between the eigenvectors before selecting those that describe the underlying manifold geometry.
+%
 The above steps to construct the nonlinear intrinsic variables are summarized in Algorithm \ref{algo}.
@@
 %
-Still, the embedding eigenevctors for two identical data sets could differ by a sign.
+Still, the embedding eigenvectors for two identical data sets could differ by a sign.
 %
@@
 For example, if the mean of the embedding is sufficiently far from 0, the signs can be chosen such that the mean of each eigenvector is greater than 0;
-alternatively, we can construct the data sets such that a few points are common between both data sets, and the sign of each eigenvector can be chosen to optimize the consistency of the embeddings of these common points.
+alternatively, we can construct the data sets such that there is a common region between both data sets, and the sign of each eigenvector can be chosen to optimize the consistency of the embeddings of the common regions.
 
@@
 \item
-Compute the empirical covariance matrix $\widehat{\mathbf{C}}(t)$ of each sample $\mathbf{Y}(t)$ in a short windows in time according to \eqref{eq:cov}.
+Compute the empirical covariance matrix $\widehat{\mathbf{C}}(t)$ of each sample $\mathbf{Y}(t)$ {\em in a short window in time} according to \eqref{eq:cov}.
 
@@
 \item
-Build the pairwise affinity matrix $\mathbf{W}$ and the corresponding normalized kernel $\mathbf{W}^{(\mathrm{norm})}$ (\ref{eq:kernel}).
+Build the pairwise affinity matrix $\mathbf{W}$ and the corresponding normalized kernel $\mathbf{W}^{\mathrm{norm}}$ (\ref{eq:kernel}).
 
 \item
-Apply EVD on the normalized kernel and view the values of its principal eigenvector as the intrinsic variables of the given observations.
+Apply EVD on the normalized kernel and view the values of its principal eigenvectors as the intrinsic variables of the given observations.
 
@@
 
+In this work, we are not only interested in extracting the underlying variables $\mathbf{x}$ from some (partial) observations $\mathbf{Y}$,
+but also interested in extending high-dimensional functions on a set of points which lie in a low-dimensional space - predicting $\mathbf{Y}$ for new points $\mathbf{x}$.
+%
 Laplacian Pyramids (LP) is a multiscale algorithm for approximating and extending an empirical function $f$ defined on a set of points.
@@
 The algorithm uses Laplacian kernels of decreasing widths to create multiscale representations of $f$;
-These representations can be easily extended to new data points.
+these representations can be easily extended to new data points.
 %
@@
 We will first review the LP algorithm for approximating and extending a one-dimensional function,
-and then describe the application of LP in the context of this work - extending a high-dimensional function that is known on a set of points which lie in a low-dimensional space.
+and then describe the application of LP in extending high-dimensional functions.
 
@@
 %
-The smoothing operator $P_0$ is a normalized, coarse Laplacian kernel, defined by $p_0(t,\tau)= s_0^{-1}(t)w_0(t,\tau)$,
-where $w_0(t,\tau)=e^{-d^2(t,\tau) / \sigma_0}$ and $s_0(t)=\sum_{\tau \in S}w_0(t,\tau)$ is the normalizing term.
+The smoothing operator $P_0$ is a normalized, coarse Laplacian kernel, defined by $p_0(i, j)= s_0^{-1}(t)w_0(i, j)$,
+where $w_0(i, j)=e^{-d^2(i, j) / \sigma_0}$ and $s_0(t)=\sum_{j \in S}w_0(i, j)$ is the normalizing term.
 %
-The pairwise distance $d(t,\tau)$ is typically the Euclidian distance, and the parameter $\sigma_0$ is set to be large compared to the values of $d^2(t,\tau)$.
+The pairwise distance $d(i, j)$ is typically the Euclidean distance, and the parameter $\sigma_0$ is set to be large compared to the values of $d^2(i, j)$.
 %
@@
 The difference $\delta_1 = f-P_0(f)$ is the input for the next iteration of the algorithm,
-which uses the smoothing operator $P_1$, $P_1 \propto e^{-d^2(t,\tau)^2 / 2^{-1} \sigma_0}$, to construct a coarse representation of $\delta_1$.
+which uses the smoothing operator $P_1$, $P_1 \propto e^{-d^2(i, j)^2 / 2^{-1} \sigma_0}$, to construct a coarse representation of $\delta_1$.
 %
@@
 \end{equation}
-As $l$ increases, the approximation becomes more refined because $P_l \propto e^{-d^2(t,\tau) / 2^{-l} \sigma_0}$ uses a Laplacian kernel of a finer width.
+As $l$ increases, the approximation becomes more refined because $P_l \propto e^{-d^2(i, j) / 2^{-l} \sigma_0}$ uses a Laplacian kernel of a finer width.
 %
@@
 
-The representations $f_0, f_1, \dots, f_l$ can be extended to a new point $\tilde{t} \in \Gamma \backslash S $ by extending the operators $P_0, P_1,\ldots,P_l$.
+The representations $f_0, f_1, \dots, f_l$ can be extended to a new point $\tilde{i} \in \Gamma \backslash S $ by extending the operators $P_0, P_1,\ldots,P_l$.
 %
-For example, $f_0(\tilde{t}) = \sum_{t \in S} p_0(\tilde{t}, t)f(t)$ and
-$f_1(\tilde{t}) = f_0(\tilde{t}) + \sum_{(t \in S)}p_1(\tilde{t}, t)\delta_1(t)$.
+For example, $f_0(\tilde{i}) = \sum_{i \in S} p_0(\tilde{i}, i)f(i)$ and
+$f_1(\tilde{i}) = f_0(\tilde{i}) + \sum_{(i \in S)}p_1(\tilde{i}, i)\delta_1(i)$.
 
@@
 
-In this work, LP is applied to extend a high-dimensional function $f:\mathbb{R}^d \rightarrow \mathbb{R}^n$, which maps a set of points in the NIV space to their values $\mathbf{Y}(t)$ in the observable space.
+In this work, LP is applied to extend a high-dimensional function $f:\mathbb{R}^d \rightarrow \mathbb{R}^n$, 
+which maps a set of points in the NIV space to their values $\mathbf{Y}(i)$ in the observable space.
 %
-Let $\Psi(t) = \left(\psi_1(t),\psi_2(t),\ldots,\psi_d(t)\right)$ be the set of NIV that were constructed from the data samples $\mathbf{Y}(t)$, as described in Algorithm \ref{algo}.
+Let $\Psi(i) = \left(\psi_1(i),\psi_2(i),\ldots,\psi_d(i)\right)$ be the set of NIV that were constructed from the data samples $\mathbf{Y}(i)$, as described in Algorithm \ref{algo}.
 %
-The values of function $f:\mathbb{R}^d \rightarrow \mathbb{R}^n$ are known on the subset $S = \{\Psi(t)\}$, with $f(\Psi(t)) = Y(t)$.
+The values of function $f:\mathbb{R}^d \rightarrow \mathbb{R}^n$ are known on the subset $S = \{\Psi(i)\}$, with $f(\Psi(i)) = Y(i)$.
 
-A na\"{\i}ve way to extend $f$ to a new data point $\Psi(\tilde{t})$ is find the point's nearest neighbors in NIV space and average their function values.
+A na\"{\i}ve way to extend $f$ to a new data point $\Psi(\tilde{i})$ is find the point's nearest neighbors in NIV space and average their function values.
 %
-A different, point-wise adaptive approach is described by Buchman {\em et al} \cite{buchman2011high};
-high-dimensional hurricane tracks were estimated from low dimensional embedding coordinates using a weighted average of the points close to $\Psi(\tilde{t})$ in the embedded space.
+A different, point-wise adaptive approach is described by Buchman {\em et al} \cite{buchman2011high}:
+high-dimensional hurricane tracks were estimated from low dimensional embedding coordinates using a weighted average of the points close to $\Psi(\tilde{i})$ in the embedded space.
 %
 However, this point-wise adaptation requires setting the nearest neighborhood radius parameter for every point.
-
-The LP algorithm automatically finds the nearest neighborhood radius for each new point $\Psi(\tilde{t})$.
+%
+The LP algorithm automatically finds the nearest neighborhood radius for each new point $\Psi(\tilde{i})$.
 %
@@
 %
-The LP approximation of a new, high dimensional value $Y(\tilde{t})$ is calculated by a weighted average of the function values that belong to the neighboring points.
+The LP approximation of a new, high dimensional value $Y(\tilde{i})$ is calculated by a weighted average of the function values that belong to the neighboring points.
 %
@@
 \begin{equation} \label{eq:LP_multi_scale_app}
-P_l \propto e^{-d^2(\Psi(t),\Psi(\tau)) / 2^{-l} \sigma_0},
+P_l \propto e^{-d^2(\Psi(i),\Psi(j)) / 2^{-l} \sigma_0},
 \end{equation}
@@
 %
-The LP algorithm for inverse mapping is summarized in Algorithm \ref{algo_LP}.
+The LP algorithm for the inverse mapping is summarized in Algorithm \ref{algo_LP}.
 
@@
 \item
-Construct a set of smoothing operators $P_0, P_1, \ldots, P_l$  based on intrinsic pairwise distances $d(\Psi(t),\Psi(\tau))$.
+Construct a set of smoothing operators $P_0, P_1, \ldots, P_l$  based on intrinsic pairwise distances $d(\Psi(i),\Psi(j))$.
 
 \item
-Use the smoothing operators to obtain a multiscale representation (see Eq. \ref{eq:LP_multi_scale_app}) of $f:\Psi(t) \rightarrow Y(t)$.
+Use the smoothing operators to obtain a multiscale representation (see Eq. \ref{eq:LP_multi_scale_app}) of $f:\Psi(i) \rightarrow Y(i)$.
 
-\item Given a new point $\Psi(\tilde{t})$ in NIV, extend the smoothing operators $P_0, P_1, \ldots, P_l$  by
-$p_l(\Psi(\tilde{t}), \Psi(t)) = s_0^{-1}(\Psi(t))w_0(\Psi(\tilde{t}),\Psi(t))$.
+\item Given a new point $\Psi(\tilde{i})$ in NIV, extend the smoothing operators $P_0, P_1, \ldots, P_l$  by
+$p_l(\Psi(\tilde{i}), \Psi(i)) = s_0^{-1}(\Psi(i))w_0(\Psi(\tilde{i}),\Psi(i))$.
 
 \item
-Use the extended smoothing operators to approximate the value of $f_l(\tilde{t}) = f_0(\tilde{t}) + \sum_{k}\sum_{(t \in S)}p_k(\tilde{t}, t)\delta_k(t)$
+Use the extended smoothing operators to approximate the value of $f_l(\tilde{i}) = f_0(\tilde{i}) + \sum_{k}\sum_{(i \in S)}p_k(\tilde{i}, i)\delta_k(i)$
 
 \item
-The value $Y(\tilde{t})$ is approximated by $f_l(\Psi(\tilde{t}))$.
+The value $Y(\tilde{i})$ is approximated by $f_l(\Psi(\tilde{i}))$.
 
@@
 \includegraphics[width=0.2\textwidth]{LP_figures/Scale11_S2}
-\caption{(Top) Approximation of function $f(x)$ as defined in \eqref{eq:LP_example} for scales 3, 5, 8, and 11 using Laplacian Pyramids; the true function is shown in blue and the LP approximation is shown in black.
+\caption{(Top) Approximation of function $f(x)$ as defined in \eqref{eq:LP_example} for ``scales'' 3, 5, 8, and 11 using Laplacian Pyramids; the true function is shown in blue and the LP approximation is shown in black.
 (Bottom) The residual error in the LP approximation.}
@@
 
-\subsection{Chemical Reaction Network}
+\subsection{Chemical Reaction Network}\label{subsec:rxn_network}
 
@@
 %
-The equations that govern the network are\\
+The reaction steps that comprise the network are\\
 \begin{equation}
@@
 %
-However, one can write four conservation equations (since total $E$, $S$, $D$, and $F$ are all conserved) to reduce the system to 6 dimensions.
+However, one can write four conservation equations (since total $E$, $S$, $D$, and $F$ are all conserved) to reduce the system to 6 dimensions 
+(we take the order of dimensions to be $S$, $E$, $E:S$, $S:E$, $D:S^{*}$, $F:E^{*}$).
 %
-In the parameter regime we consider, the system exhibits a separation of timescales, so that initial conditions quickly approach a two-dimensional manifold.
+In the parameter regime we consider, the system of ODEs exhibits a separation of time scales, so that initial conditions quickly approach a two-dimensional manifold.
 %
-Details about the specific parameters can be found in Appendix \ref{app:rxn}.
+Details about the specific parameter values can be found in Appendix \ref{app:rxn}.
 
@@
 %
-When the number of molecules is small, the system is inherently stochastic and its dynamics can be simulated using the Gillespie's Stochastic Simulation Algorithm (SSA) \cite{gillespie1977exact}.
+When the number of molecules is small, the system is inherently stochastic and its dynamics can be simulated using the Gillespie Stochastic Simulation Algorithm (SSA) \cite{gillespie1977exact}.
 %
@@
 %
-We take the volume small enough so that we can still see appreciable stochasticity in small simulation bursts, but large enough so that the underlying two-dimensional manifold is (relatively) smooth.
+We take the volume small enough so that we can still observe appreciable stochasticity in small simulation bursts, but large enough so that the underlying two-dimensional manifold is (relatively) smooth.
 
-We generate 3000 random initial conditions $\mathbf{Y}_0(1), \dots, \mathbf{Y}_0(3000) \in \mathbb{R}^6$, enforcing that all concentrations must be non-negative
-(we take the order of the components in the vector $\mathbf{Y}_0(t)$ to be $S$, $E$, $E:S$, $S:E$, $D:S^{*}$, $F:E^{*}$).
+We generate 3000 random initial conditions $\mathbf{Y}_0(1), \dots, \mathbf{Y}_0(3000) \in \mathbb{R}^6$, enforcing that all concentrations must be non-negative.
 %
 We integrate each point $\mathbf{Y}_0(t)$ forward for 10 time units using the SSA to obtain point the $\mathbf{Y}(t) \in \mathbb{R}^6$;
-10 time units is sufficiently long for the initial points to converge to the two-dimensional manifold,
-but not long enough for the points to converge to a one-dimensional curve or to the final steady state.
+according to the timescales around the steady state calculated from the linearized ODEs, 10 time units is sufficiently long for the initial points in the ODE system to converge to the two-dimensional manifold,
+but not long enough for the points to converge to a one-dimensional curve or to the final steady state (see Appendix \ref{app:rxn} for more details). 
+%
+In our stochastic simulations, the initial points converge to an approximate two-dimensional mainfold in expected value (see Figure \ref{fig:rxn_manifolds}).
 %
@@
   \includegraphics[height=6cm]{rxn_figures/rxn_manifold4}\llap{\raisebox{3.5cm}{\includegraphics[height=2.5cm]{rxn_figures/rxn_manifold4_2}}} \\
-    \caption{Projections of the data obtained from stochastic simulation of the chemical reaction network.
-The insets show rotations of the projections to illustrate the two-dimensionality of the manifold.}
+    \caption{Projections of the data obtained from stochastic simulation of the chemical reaction network described in Section \ref{subsec:rxn_network}.
+The insets show rotations of the projections to illustrate the approximate two-dimensionality of the ``slow manifold''.}
     \label{fig:rxn_manifolds}
@@
 where $\hat{\mathbf{\mu}}_i(y)$ is the empirical mean of $\mathcal{Y}^{burst}_i(y)$.
+
+We first demonstrate that NIV produces the same embeddings for $\mathcal{Y}_1$ and $\mathcal{Y}_2$, even though the two data sets contain information of different chemical species/
 %
-We will use NIV together with Laplacian Pyramids to predict the values of $S:E$ and $F:E^{*}$ in data set 2.
-
-Figure \ref{fig:rxn_embedding} shows the two-dimensional NIV embeddings for the two different data sets; the embeddings appear very consistent.
+Figure \ref{fig:rxn_embedding} shows the two-dimensional NIV embeddings for the two different data sets; the embeddings appear consistent.
 %
@@
 We obtain a correlation of 0.97 and 0.95 for the first and second NIV, respectively, indicating that the two embeddings are in excellent agreement with each other.
+%
+We would like to note that both $\mathcal{Y}_1$ and $\mathcal{Y}_2$ are ``rich enough'' to recover the two-dimensional manifolds.
+%
+Clearly, other data sets consisting of different partial observations could be constructed such that that underlying manifold would not be recovered.
 
@@
     \includegraphics[width=0.4\textwidth]{rxn_figures/rxn_NLICA_corr2}
-    \caption{(Top, left) NIV embedding obtained from data set 1 (observations of components E, S, S:E, and F:E$^{*}$), colored by $S$.
-(Top, right) NIV embedding obtained from data set 2 (observations of components E, S, E:S, and D:S$^{*}$), colored by $S$.
+    \caption{(Top, left) NIV embedding obtained from $\mathcal{Y}_1$ (observations of components E, S, S:E, and F:E$^{*}$), colored by $S$.
+(Top, right) NIV embedding obtained from $\mathcal{Y}_2$ (observations of components E, S, E:S, and D:S$^{*}$), colored by $S$.
+Visually, we can see that the embeddings obtained from $\mathcal{Y}_1$ and $\mathcal{Y}_2$ are consistent, even though the two data sets consist of observations of different chemical species.
 (Bottom, left) Correlation of first NIV between two different embeddings (correlation=0.97).
-(Bottom, right)  Correlation of second NIV between two different embeddings (correlation=0.95).}
+(Bottom, right)  Correlation of second NIV between two different embeddings (correlation=0.95).
+We obtain a good quantitative agreement between the embedding coordinates for the two data sets.}
     \label{fig:rxn_embedding}
@@
 
-We then use LP to predict $S:E$ and $F:E^{*}$ for $\mathcal{Y}_2$.
+We then use NIV together with Laplacian Pyramids to predict the values of $S:E$ and $F:E^{*}$ for $\mathcal{Y}_2$.
 %
@@
 %
-Instead, we must first project the data into the NIV space so that we can compute neighbors {\em between} the two data sets.
+Instead, we must first project the data into the NIV space so that we can compute neighbors {\em between the two data sets}.
 %
@@
 Therefore, we can effectively predict the unobserved components in the reaction network using NIV together with LP. 
+%
 
@@
     \includegraphics[width=0.4\textwidth]{rxn_figures/rxn_recon6}
-    \caption{LP reconstructions of $S:E$ (left) and $F:E^{*}$ (right).}
+    \caption{LP reconstructions of $S:E$ (left) and $F:E^{*}$ (right) for $\mathcal{Y}_2$, using $\mathcal{Y}_1$ as training data.}
     \label{fig:rxn_recon}
@@
 
-Our second example comes from a simulation of a small peptide fragment.
+Our second example comes from the molecular dynamics simulation of a small peptide fragment.
 %
-Alanine dipeptide (Ala2) is often used as a ``prototypical'' protein for simulation studies
+Alanine dipeptide (Ala2) is often used as a ``prototypical'' protein caricature for simulation studies
 \cite{apostolakis1999calculation, bolhuis2000reaction, chekmarev2004long, ma2005automatic, frewen2009exploration, ferguson2011integrating}.
@@
 %
-The molecule is solvated with 638 TIP3P water molecules \cite{jorgensen1983comparison}, with periodic boundary conditions and the particle mesh Ewald method used for long-range electrostatic interactions \cite{essmann1995smooth}.
+The molecule is solvated with 638 TIP3P water molecules \cite{jorgensen1983comparison} 
+with periodic boundary conditions, and the particle mesh Ewald method is used for long-range electrostatic interactions \cite{essmann1995smooth}.
 %
@@
 %
-Hydrogen bonds are fixed using the SHAKE algorithm \cite{ryckaert1977numerical}.
+Hydrogen bond lengths are fixed using the SHAKE algorithm \cite{ryckaert1977numerical}.
 %
-The two dihedral angles $\phi$ and $\psi$ are known to parameterize the free energy surface, and the free energy surface contains three minima (labeled A, B, and C, see Figure \ref{fig:ala_fes}).
+The two dihedral angles $\phi$ and $\psi$ are known to parameterize the free energy surface, which contains three important minima (labeled A, B, and C, see Figure \ref{fig:ala_fes}).
 %
@@
     \includegraphics[width=0.3\textwidth]{molecule_labeled2}
-    \caption{(Left) Free energy surface for Ala2.
-The minima are labeled A, B, and C, and the corresponding molecular configurations are shown.
-(Right) The molecular structure of Ala2, excluding the hydrogens.
-The atoms are numbered and the two dihedral angles $\phi$ and $\psi$ are indicated.}
+    \caption{(Left) Free energy surface for Ala2. The relevant minima are labeled A, B, and C, and the corresponding molecular configurations are shown. 
+    (Right) Sample representative molecular structure of Ala2, excluding the hydrogens. The atoms are numbered and the two dihedral angles $\phi$ and $\psi$ are indicated.}
     \label{fig:ala_fes}
@@
 
-We first wish to demonstrate the benefit of NIV over other nonlinear dimensionality reduction techniques, such as diffusion maps \cite{coifman2005geometric}.
+We first compare the benefits of NIV over other nonlinear dimensionality reduction techniques, such as diffusion maps \cite{coifman2005geometric}.
 %
@@
 %
-For NIV, we compute the covariances as in \eqref{eq:cov}, with $L=10$;
-at the ends of each simulation trajectory, we take fewer points before or following the point of interest to estimate the covariance
-(as neighboring points in our data set that do not come from the
-same trajectories should not contribute to the estimate of the local covariance).
+For NIV, we compute the covariances as in \eqref{eq:cov}, with $L=10$.
+%at the ends of each simulation trajectory, we take fewer points before or following the point of interest to estimate the covariance
+%(as neighboring points in our data set that do not come from the
+%same trajectories should not contribute to the estimate of the local covariance).
 
@@
 %
-The correlation between the two NIV embeddings is higher than the correlation between the two diffusion maps embeddings,
-demonstrating that it is often advantageous to use NIV over diffusion maps if one wishes to obtain a consistent embedding and merge data sets from different observation domains.
+The correlation between the two NIV embeddings is higher than the correlation between the two diffusion maps embeddings.
+Therefore, it is often advantageous to use NIV over diffusion maps if one wishes to obtain a consistent embedding and merge data sets from different observation domains.
 
@@
 %
-We compute the covariances as in \eqref{eq:cov} with $L=15$;
-at the ends of each simulation trajectory, we take fewer points before or following the point of interest to estimate the covariance.
+We compute the covariances as in \eqref{eq:cov} with $L=15$.
+%at the ends of each simulation trajectory, we take fewer points before or following the point of interest to estimate the covariance.
 %
@@
 %
-The embedding is three-dimensional: the first coordinate describes the flipping of atoms 1 and 3, the second coordinate describes the dihedral angle $\phi$, and the third coordinate describes the dihedral angle $\psi$.
+The embedding is three-dimensional, and each coordinate can be (visually) associated with a physical variable: 
+the first coordinate describes the flipping of atoms 1 and 3, the second coordinate describes the dihedral angle $\phi$, and the third coordinate describes the dihedral angle $\psi$.
 %
@@
 
-Figure \ref{fig:ala_recon} shows the reconstructed position versus true position for some selected atoms.
+Figure \ref{fig:ala_recon} shows the reconstructed position from partial observation versus true position for some selected atoms.
 %
@@
 Figure \ref{fig:ala_molecules} shows molecular structures for the true and reconstructed configurations for some data points;
-there is nice agreement between the true and reconstructed configurations.
+there is qualitative agreement between the true and reconstructed configurations.
 
@@
 \section{Conclusions} \label{sec:conclusions}
+Nonlinear intrinsic variables have great potential applications in characterizing complex stochastic dynamical systems.
+%
+The intrinsicness of the embedding coordinates allows us to compare and merge data sets that come from different measurement functions of the same underlying system.
+%
+We can therefore merge data from coarse- and fine-scale simulations, as well as merge data from simulations with data from experiments.
+
+There are many open research questions and interesting research directions that may be explored.
+%
+In this work, we discuss data sets that consist of different partial observations of the same underlying process, but where the data is well-sampled and each data set covers the entire domain.
+%
+However, NIV can also be used to merge data sets when each data set samples only a portion of the domain, provided there is enough overlap to ``register'' the embeddings.
+%
+Merging multiple data sets that come from different portions of the domain would not only require scaling the embedding coordinates, 
+but also require shifting and permuting the embedding coordinates.
+
+In addition, we have stated that the measurement function $f$ has to be ``rich enough'' in order to recover the variables of the underlying stochastic process.
+%
+However, determining whether $f$ meets such criteria is a non-trivial task.
+%
+For many problems, one would like to say what types of measurement functions $f$ could be considered in order to build accurate reduced models.
+%
+Other issues, such as accurately estimating the covariances matrices required to compute the Mahalanobis distance, are also of current research interest.
+%
+We are confident that the exploration of these open questions will lead to many interesting applications for our methodologies.
 
 \begin{acknowledgments}
-The authors would like to thank...
+C. J. D. would like to acknowledge support from the US Department of Energy Computation Science Graduate Fellowship, grant number DE-FG02-97ER25308.
+%
+I. G. K. would like to acknowledge support from the US Department of Energy, grant numbers DE-FG02-10ER26024 and DE-FG02-09ER25877.
 \end{acknowledgments}
@@
 %
-Therefore, we choose to integrate forward for 10 time units to find points on a 2-dimensional manifold.
+Therefore, we choose to integrate forward for 10 time units to find points on a perceived two-dimensional manifold.
 

### 2013-12-27 paper/draft5.tex
Added files related to paper

--- 2013-12-27 paper/draft4.tex
@@
 %
-In particular, we focus on the way NIV allow us to functionally merge different
+In particular, we focus on the way NIV allows us to functionally merge different
 simulation ensembles, and {\em different partial observations of these ensembles}, as well
@@
 %
-We illustrate the approach through two distinct sets of multiscale atomistic simulations:
+We illustrate the approach through two distinct sets of multiscale {\color{red} atomistic} simulations:
 a stochastic simulation of an enzyme reaction network exhibiting both fast and slow time scales,
@@
 The last decade has witnessed extensive advances in dimensionality reduction techniques:
-finding meaningful low-dimensional descriptions of high-dimensional data (TODO: add references).
+finding meaningful low-dimensional descriptions of high-dimensional data {\color{red} TODO: add references}.
 %
@@
 and when different data sets are obtained using different instrumental modalities
-(such as when one wants to merge molecular dynamics data with, for example, spectral information TODO: add specific example of spectral information).
+(such as when one wants to merge molecular dynamics data with, for example, spectral information {\color{red} TODO: add specific example of spectral information}).
 %
@@
 %
-Because eigenvectors can be determined up to a scaling factor, we must scale the eigenvectors from different data sets such that the resulting embeddings are consistent.
+Because eigenvectors can be determined up to a scaling factor, we must scale the eigenvectors from different data sets so that the resulting embeddings are consistent.
 %
@@
 We integrate each point $\mathbf{Y}_0(t)$ forward for 10 time units using the SSA to obtain point the $\mathbf{Y}(t) \in \mathbb{R}^6$;
-according to the timescales around the steady state calculated from the linearized ODEs, 10 time units is sufficiently long for the initial points in the ODE system to converge to the two-dimensional manifold,
+from the time scales calculated from the linearized ODEs, 10 time units is sufficiently long for the initial points in the ODE system to converge to the two-dimensional manifold,
 but not long enough for the points to converge to a one-dimensional curve or to the final steady state (see Appendix \ref{app:rxn} for more details).
@@
 
-We first demonstrate that NIV produces the same embeddings for $\mathcal{Y}_1$ and $\mathcal{Y}_2$, even though the two data sets contain information of different chemical species/
+We first demonstrate that NIV produces the same embeddings for $\mathcal{Y}_1$ and $\mathcal{Y}_2$, even though the two data sets contain information of different chemical species.
 %
@@
 
-We first compare the benefits of NIV over other nonlinear dimensionality reduction techniques, such as diffusion maps \cite{coifman2005geometric}.
+We first examine the benefits of NIV over diffusion maps \cite{coifman2005geometric}, another nonlinear dimensionality reduction techniques.
 %
@@
 \section{Conclusions} \label{sec:conclusions}
-Nonlinear intrinsic variables have great potential applications in characterizing complex stochastic dynamical systems.
+We have used nonlinear intrinsic variables to analyze two complex stochastic molecular systems: a simulation of a chemical reaction network and a molecular dynamics simulation of alanine dipeptide.
 %
-The intrinsicness of the embedding coordinates allows us to compare and merge data sets that come from different measurement functions of the same underlying system.
+In both examples, we were able to uncover the intrinsic variables governing the underlying stochastic process, independent of the measurements or observations of the system.
 %
-We can therefore merge data from coarse- and fine-scale simulations, as well as merge data from simulations with data from experiments.
+The uniqueness of the embedding coordinates allowed us to compare and merge data sets from different measurement functions, and therefore allowed us to use Laplacian Pyramids to complete partial observations.
 
@@
 %
-In this work, we discuss data sets that consist of different partial observations of the same underlying process, but where the data is well-sampled and each data set covers the entire domain.
+In this work, we considered data sets that consist of different partial observations, but where each data set covers the entire domain and is well-sampled.
 %
-However, NIV can also be used to merge data sets when each data set samples only a portion of the domain, provided there is enough overlap to ``register'' the embeddings.
+However, NIV could also be used to merge data sets when each data set samples only a portion of the domain, provided there is enough overlap to ``register'' the embeddings.
 %
-Merging multiple data sets that come from different portions of the domain would not only require scaling the embedding coordinates, 
-but also require shifting and permuting the embedding coordinates.
+Merging data sets that come from different portions of the domain would not only require scaling the embedding coordinates,
+but also shifting and permuting the embedding coordinates.
+%
+The ability to merge data from different regions of the domain would then allow us to analyze systems where complete sampling is computationally intractable, such as molecular systems with high energy barriers separating regions of state space.
 
@@
 %
-For many problems, one would like to say what types of measurement functions $f$ could be considered in order to build accurate reduced models.
+For many problems, one would like to say what types of measurement functions $f$ are rich enough, as knowing which functions retain sufficient detail of the underlying stochastic process can aid us in building accurate reduced models.
 %
@@
 %
-We are confident that the exploration of these open questions will lead to many interesting applications for our methodologies.
+We are confident that the exploration of these open questions will lead to many interesting applications for our methodologies, such as merging data from different molecular simulations, or merging simulation data with experimental data.
 

### 2013-12-27 paper/draft6.tex
Added files related to paper

--- 2013-12-27 paper/draft5.tex
@@
 
+\makeatletter
+\renewcommand\@make@capt@title[2]{%
+ \@ifx@empty\float@link{\@firstofone}{\expandafter\href\expandafter{\float@link}}%
+  {\textbf{#1}}\@caption@fignum@sep#2\quad
+}%
+\makeatother
+
 %\usepackage{float}
@@
 
-\title{Intrinsic Variables and State Reconstruction in Multiscale Simulations}
+\title{Nonlinear Intrinsic Variables and State Reconstruction in Multiscale Simulations}
 
@@
 \author{Neta Rabin}
-\email{netar@afeka.ac.ilil}
+\email{netar@afeka.ac.il}
 \affiliation{Department of Exact Sciences, Afeka Tel-Aviv Academic College of Engineering, Tel-Aviv, Israel}
@@
 %
-We illustrate the approach through two distinct sets of multiscale {\color{red} atomistic} simulations:
+We illustrate the approach through two distinct sets of atomistic simulations:
 a stochastic simulation of an enzyme reaction network exhibiting both fast and slow time scales,
@@
 The last decade has witnessed extensive advances in dimensionality reduction techniques:
-finding meaningful low-dimensional descriptions of high-dimensional data {\color{red} TODO: add references}.
+finding meaningful low-dimensional descriptions of high-dimensional data \cite{Tenenbaum2000,Roweis2000,Donoho2003,Belkin2003,Coifman2006}.
 %
@@
 If the (high-dimensional) data $\mathbf{Y}(t)$ arise from, for example, a
-molecular dynamics simulation of a macromolecule in solution or the stochastic
+molecular dynamics simulation of a macromolecule in solution, or from the stochastic
 simulation of a complex chemical reaction scheme, the detection of a few good, coarse-grained
@@
 %
-While this can be easily rectified by a linear transformation, the problem becomes
+While this can be easily rectified by an affine transformation
+(see, by analogy, the discussion in \cite{lafon2006data}),
+the problem becomes
 exacerbated when the low-dimensional space is curved (a manifold, rather than a hyperplane)
@@
 coordinate set, shared by all measurement ensembles and observation modalities -
-what we call Nonlinear Intrinsic Variables - is invaluable.
+what we will call Nonlinear Intrinsic Variables - is invaluable.
 %
@@
 %
-This empirical mapping also requires good interpolation tools in embedding space; to this
-end, here we will demonstrate the use of a multiscale Laplacian Pyramid approach \cite{rabin2012heterogeneous}.
+%%%RONEN
+% we need in the sentence above to mention that we have the analog of a ``training set"
+%
+To construct this empirical mapping and the associated observers,
+accurate interpolation tools must be available in the embedding space; to this
+end, here we chose to demonstrate the use of a multiscale Laplacian Pyramid approach \cite{rabin2012heterogeneous}.
+%%
+%%RONEN
+%% remember to add all the other options, like kriging..... in the discussion
+%%
 
@@
 in certain parameter regimes, separation of time scales is known
-to reduce this model to an effective two-dimensional description \cite{zagaris2012stability}.
+to reduce the ODE model of this kinetic scheme to an effective two-dimensional description \cite{zagaris2012stability}.
 %
@@
 %
-This example will demonstrate how our techniques can have advantages over more common techniques,
+This example will allow us to compate our approach to more common techniques,
 such as diffusion maps for dimensionality reduction
-and nearest neighbor interpolation for reconstruction.
+and nearest neighbor interpolation for observation reconstrcution.
 %
-The paper is structured as follows: in Section \ref{sec:NIV} we present the Nonlinear Intrinsic Variable formulation and
+The remainder of the paper is structured as follows: in Section \ref{sec:NIV} we present the Nonlinear Intrinsic Variable formulation and
 the associated inference method.
@@
 %
-Section \ref{sec:examples} contain the results
-of applying the approach to simulation data from our two illustrative examples.
+In section \ref{sec:examples} the results
+of the application of the approach to simulation data from our two illustrative examples is presented and discussed.
 %
-We conclude with
-a summary and discussion of open issues in Section \ref{sec:conclusions}.
+We conclude with a summary and our perspective on open issues in Section \ref{sec:conclusions}.
 
@@
 %
-The measured process is assumed (and this is a critical assumption) to be a manifestation in an observable domain of a low-dimensional diffusion process. Thus, it can be expressed by
+The measured process is assumed (and this is a critical assumption) to be a manifestation, in an observable domain, of a low-dimensional diffusion process. Thus, it can be expressed as
 \begin{equation}
-	\mathbf{Y}(t) = f(\mathbf{x}(t))
+	\mathbf{Y}(t) = \mathbf{f}(\mathbf{x}(t))
 \end{equation}
-where $f:\mathbb{R}^d \rightarrow \mathbb{R}^n$ is an unknown (possibly nonlinear, $d \ll n$) function and $\mathbf{x}(t)$ is a diffusion process that consists of $d$ underlying variables.
+where $\mathbf{f}:\mathbb{R}^d \rightarrow \mathbb{R}^n$ is an unknown (possibly nonlinear, $d \ll n$) function and $\mathbf{x}(t)$ is a diffusion process that consists of $d$ underlying variables.
 %
-The dynamics of the diffusion process in each of its underlying variables are described by normalized stochastic differential equations as
+The dynamics of the diffusion process in terms of its underlying variables is described by normalized stochastic differential equations as
 \begin{equation}
@@
 \end{equation}
-where $a_i$ are unknown drift functions and $\dot{w}_i(t)$ are independent white noises (the independence is another critical assumption).
+where $a_i$ are unknown drift functions and $\dot{w}_i(t)$ are independent white noises (this independence is our second critical assumption).
 
@@
 %
-By relying on the fact that the empirical method we will describe is independent of the observation function $f$,
+By relying on the fact that the empirical method we will describe is independent of the observation function $\mathbf{f}$,
 we refer to the coordinates of $\mathbf{x}(t)$ as Nonlinear Intrinsic Variables (NIV).
 %
-The available samples $\mathbf{Y}(t)$ may be the result of different measurement functions $f$ in various observable domains,
+The available samples $\mathbf{Y}(t)$ may be the result of different measurement functions $\mathbf{f}$ in various observable domains,
 or they may be partial measurements consisting of merely a subset of the coordinates of the observable domains.
 %
-The idea is to empirically construct a NIV coordinate system driven entirely by measurements that is invariant to the observation function $f$ (see Figure \ref{fig:IntrinsicIllustration} for a schematic illustration).
+The idea is to empirically construct a NIV coordinate system driven entirely by measurements that is invariant to the observation function $\mathbf{f}$ 
+(see Figure \ref{fig:IntrinsicIllustration} for a schematic illustration).
+%
+We remark that the available data should be rich enough, i.e., consists of a sufficient amount of ``historical" data with sufficient variability, in order to get a full empirical model.
 
-The presented method consists of the following main principles.
+The method consists of the following main principles.
 %
-(1) The underlying diffusion process implies that a short trajectory of successive samples mainly consists of diffusion noise, and hence, creates a ``sphere" of samples in the underlying domain.
+(1) The underlying diffusion process implies that a short trajectory of successive samples mainly consists of diffusion noise,
+and hence, creates a ``sphere" of samples in the underlying domain.
+%%
+%%%RONEN
+% whether here or in the discussion there should be something about testing this
+% practically
 %
-This sphere is mapped to an ellipse in the observable domain by the measurement function $f$.
+This sphere is mapped to an ellipse in the observable domain by the measurement function $\mathbf{f}$.
 %
@@
 %
-(2) The principal directions of the tangent planes are utilized to define a Riemannian metric that is shown to be {\em locally invariant to the measurement function $f$}.
+(2) The principal directions of the tangent planes are utilized to define a Riemannian metric that is shown to be {\em locally invariant to the measurement function $\mathbf{f}$}.
 %
@@
 \includegraphics[width=0.7\textwidth]{IntrinsicEmbeddingIllustration2}
-\caption{Illustration of the nonlinear embedding that yields an intrinsic representation independent of the measurement function $f$.
+\caption{Illustration of the nonlinear embedding that yields an intrinsic representation independent of the measurement function $\mathbf{f}$.
 (Bottom) The underlying variables in which the noises are independent with unit variance.
@@
 %
-The Mahalanobis distance is invariant under linear transformations.
+The Mahalanobis distance is invariant under affine transformations.
 %
-Thus, by assuming that the observation function $f$ is bi-Lipschitz and smooth and by using local linearization of the function, i.e., $\mathbf{Y}(t) = \mathbf{J}(t) \mathbf{x}(t) + \boldsymbol{\epsilon}(t)$ where $\mathbf{J}(t)$ is the Jacobian of $f(\mathbf{x}(t))$ and $\boldsymbol{\epsilon}(t)$ is the residual consisting of higher-order terms, it was shown by Singer and Coifman \cite{singer2008non} that $\mathbf{C}(t) = \mathbf{J}(t)\mathbf{J}^T(t)$ and that the Mahalanobis distance approximates the Euclidean distance between the corresponding samples of the underlying process to second order, i.e.,
+Thus, by assuming that the observation function $\mathbf{f}$ is bi-Lipschitz and smooth and by using local linearization of the function, i.e., $\mathbf{Y}(t) = \mathbf{J}(t) \mathbf{x}(t) + \boldsymbol{\epsilon}(t)$ where $\mathbf{J}(t)$ is the Jacobian of $\mathbf{f}(\mathbf{x}(t))$ and $\boldsymbol{\epsilon}(t)$ is the residual consisting of higher-order terms, it was shown by Singer and Coifman \cite{singer2008non} that $\mathbf{C}(t) = \mathbf{J}(t)\mathbf{J}^T(t)$ and that the Mahalanobis distance approximates the Euclidean distance between the corresponding samples of the underlying process to second order, i.e.,
 \begin{equation}
@@
 %
-This result implies that the Mahalanobis distance is invariant to the measurement function $f$, and hence,
+This result implies that the Mahalanobis distance is invariant to the measurement function $\mathbf{f}$, and hence,
 it yields the same distances between samples obtained under different observation functions or even partial observations.
 %
-We would like to note that, in general, $f$ being bi-Lipschitz implies that $f$ is invertible.
+We would like to note that, in general, $\mathbf{f}$ being bi-Lipschitz implies that $\mathbf{f}$ is invertible.
+%%%
+%%%RONEN
+% this is nontrivial and it requires discussion
+%  about what manifold thing need to live on for this to be doable
 %
-However, in practice, determining whether $f$ contains sufficient information and is ``rich enough'' to completely determine the underlying process is a non-trivial task.
+However, in practice, determining whether $\mathbf{f}$ contains sufficient information and is ``rich enough'' to completely determine the underlying process is a non-trivial task.
+%
+In this work, we exploit the fact that $\mathbf{C}(t) = \mathbf{J}(t)\mathbf{J}^T(t)$, which implies that $\mathbf{C}(t)$ is an $n \times n$ positive semidefinite matrix of rank $d$, to empirically infer the dimension $d$.
+%
+According to the spectrum of the local covariance matrices and their corresponding spectral gaps, we approximate the rank of the matrices. 
+%
+Consistent rank estimates among the covariance matrices may imply that the measurements are ``rich enough", and hence, may be good indicators for the dimension $d$. 
+%
+In addition, since usually the dimension $d$ of the underlying process is smaller than the dimension of the measured process $n$, the covariance matrix is singular and non-invertible. 
+%
+Thus, in practice we use the pseudo-inverse to compute the inverse matrices in \eqref{eq:mahalanobis}.
 
@@
 In practice, we set $\varepsilon$ to be the median of the elements of the kernel matrix.
+%%%RONEN
+%   this is also not really clear, and "effectively connected to half" is
+%   also not true -- just need to be said carefully
 %
@@
 %
-The next few eigenvectors describe the geometry of the underlying manifold.
+The next few eigenvectors can be argued to describe the geometry of the underlying manifold \cite{coifman2005geometric}
 %
-However, some eigenvectors can be higher harmonics of the same principal direction (similar to how the eigenfunctions $\cos x$ and $\cos 2x$ both characterize the $x$ direction),
-and so one must check for correlations between the eigenvectors before selecting those that describe the underlying manifold geometry.
+However, some eigenvectors can be higher harmonics of the same principal direction along the data manifold.
+This is analogous to how the eigenfunctions $\cos x$ and $\cos 2x$ of the usual Laplacian in one spatial dimension
+and with no flux boundary conditions are one-to-one with the values of $x$ for $0 \le x \le 1$);
+so one must check for correlations between the eigenvectors before selecting those that describe the underlying manifold geometry.
 %
@@
 
-Each eigenvector then describes an intrinsic variable for the data set of interest.
+Ignoring these ``higher harmonics", each retained eigenvector then describes an intrinsic variable for the data set of interest.
 %
-Because eigenvectors can be determined up to a scaling factor, we must scale the eigenvectors from different data sets so that the resulting embeddings are consistent.
+Because eigenvectors can be determined up to a scaling factor,
+we must scale the eigenvectors from different data sets so that the resulting embeddings are consistent.
 %
-We first scale the eigenvectors so that the norm of the each eigenvector is equal to the cardinality of the data set.
+We first scale the eigenvectors so that $\|\psi_i\| = T$, where $T$ is the number of data points;
+intuitively, a data set that has twice as many samples as another data set, but the same underlying geometry, should have twice the norm.
 %
-Intuitively, a data set that has twice as many samples as another data set, but the same underlying geometry, should have twice the norm.
+Still, the computed embedding eigenvectors, even for two identical data sets, may differ by a sign.
 %
-Still, the embedding eigenvectors for two identical data sets could differ by a sign.
+Reconciling the signs for the embeddings of different data sets can be rationally done in several ways and is somewhat problem-specific.
 %
-Reconciling the signs for the embeddings of different data sets can be done in several ways and is somewhat problem-specific.
+For example, if the mean of the embedding is sufficiently far from 0, we can require $\langle \psi_i \rangle > 0$;
+%the signs can be chosen such that the mean of each eigenvector has a prescribed sign;
+alternatively, if there is a common region sampled by both data sets, the sign of each eigenvector can be chosen to optimize the consistency of the embeddings of the common region data.
 %
-For example, if the mean of the embedding is sufficiently far from 0, the signs can be chosen such that the mean of each eigenvector is greater than 0;
-alternatively, we can construct the data sets such that there is a common region between both data sets, and the sign of each eigenvector can be chosen to optimize the consistency of the embeddings of the common regions.
+We will return to the issue of embedding consistency for different data sets in our concluding discussion; for the moment,
+we will assume that our different sets sample the same region of data space in a representative enough way for the
+correspondence between the sequences of retained eigenvectors for different embeddings to be obvious.
 
@@
 \item
-Apply EVD on the normalized kernel and view the values of its principal eigenvectors as the intrinsic variables of the given observations.
+Apply eigenvector decomposition (EVD) on the normalized kernel and view the values of its principal eigenvectors (modulo the possibility of
+ ``higher harmonics", see text) as the intrinsic variables of the given observations.
 
@@
 
-\section{Laplacian Pyramids} \label{sec:LapPyr}
+\section{Laplacian Pyramids for Data Extension} \label{sec:LapPyr}
 
 In this work, we are not only interested in extracting the underlying variables $\mathbf{x}$ from some (partial) observations $\mathbf{Y}$,
-but also interested in extending high-dimensional functions on a set of points which lie in a low-dimensional space - predicting $\mathbf{Y}$ for new points $\mathbf{x}$.
+but also interested in extending high-dimensional functions on a set of points which lie in a low-dimensional space.
 %
-Laplacian Pyramids (LP) is a multiscale algorithm for approximating and extending an empirical function $f$ defined on a set of points.
+More specifically, thinking of the ambient space coordinates, $\mathbf{Y}$, as functions on the data,
+we want to estimate $\mathbf{Y}$ for new points $\mathbf{x}$.
 %
-The algorithm uses Laplacian kernels of decreasing widths to create multiscale representations of $f$;
+Laplacian Pyramids (LP) is a multiscale algorithm for approximating and extending an empirical function $\mathbf{f}$ defined on a set of points.
+%
+The algorithm uses Laplacian kernels of decreasing widths to create multiscale representations of $\mathbf{f}$;
 these representations can be easily extended to new data points.
@@
 %
+%%%RONEN
+% this is colloquially correct but inconsistent
+%
 We will first review the LP algorithm for approximating and extending a one-dimensional function,
@@
 
-Let $f: \Gamma \rightarrow \mathbb{R}^n$ be a function that is known on a subset of points $S \subset \Gamma$.
+Let $\mathbf{f}: \Gamma \rightarrow \mathbb{R}^n$ be a function that is known on a subset of points $S \subset \Gamma$.
 %
-A coarse representation of $f$ is generated using a coarse smoothing operator $P_0$.
+A coarse representation of $\mathbf{f}$ is generated using a coarse smoothing operator $P_0$.
 %
-The smoothing operator $P_0$ is a normalized, coarse Laplacian kernel, defined by $p_0(i, j)= s_0^{-1}(t)w_0(i, j)$,
+The smoothing operator $P_0$ is a normalized, coarse Laplacian kernel, defined by 
+\begin{equation}
+p_0(i, j)= s_0^{-1}(t)w_0(i, j),
+\end{equation}
 where $w_0(i, j)=e^{-d^2(i, j) / \sigma_0}$ and $s_0(t)=\sum_{j \in S}w_0(i, j)$ is the normalizing term.
@@
 %
-The application of $P_0$ to $f$ yields a coarse representation of the function, which we denote by $f_0=P_0(f)$.
+The application of $P_0$ to $\mathbf{f}$ yields a coarse representation of the function, which we denote by $\mathbf{f}_0=P_0(\mathbf{f})$.
 
-The difference $\delta_1 = f-P_0(f)$ is the input for the next iteration of the algorithm,
+The difference $\delta_1 = \mathbf{f}-P_0(\mathbf{f})$ is the input for the next iteration of the algorithm,
 which uses the smoothing operator $P_1$, $P_1 \propto e^{-d^2(i, j)^2 / 2^{-1} \sigma_0}$, to construct a coarse representation of $\delta_1$.
 %
-The obtained representation of $\delta_1$, $P_1(\delta_1)$ together with the result of the previous iteration $f_0 = P_0(f)$
-yields a new, finer representation of $f$, $f_1 = P_0(f) + P_1(\delta_1)$.
+The obtained representation of $\delta_1$, $P_1(\delta_1)$ together with the result of the previous iteration $\mathbf{f}_0 = P_0(\mathbf{f})$
+yields a new, finer representation of $\mathbf{f}$, $\mathbf{f}_1 = P_0(\mathbf{f}) + P_1(\delta_1)$.
 %
-In an iterative manner, multiscale representations of the function $f$, denoted $f_l$, are constructed.
+In an iterative manner, multiscale representations of the function $\mathbf{f}$, denoted $\mathbf{f}_l$, are constructed.
 \begin{equation} \label{eq:LP_multi_scale}
  \begin{array}{cl}
-\mbox{scale 0:} & f_0 = P_0(f) \\
-\mbox{scale 1:} & f_1 = P_0(f) + P_1(\delta_1) \\
+\mbox{scale 0:} & \mathbf{f}_0 = P_0(\mathbf{f}) \\
+\mbox{scale 1:} & \mathbf{f}_1 = P_0(\mathbf{f}) + P_1(\delta_1) \\
 : & : \\
-\mbox{scale l:} & f_l = P_0(f) + \sum_{k=1}^{l}P_k(\delta_k)\\
+\mbox{scale l:} & \mathbf{f}_l = P_0(\mathbf{f}) + \sum_{k=1}^{l}P_k(\delta_k)\\
 \end{array}
@@
 %
-The iterations stop when the difference between $f$ and $f_l$ is smaller than a pre-defined error threshold.
+The iterations stop when the difference between $\mathbf{f}$ and $\mathbf{f}_l$ is smaller than a pre-defined error threshold.
 
-The representations $f_0, f_1, \dots, f_l$ can be extended to a new point $\tilde{i} \in \Gamma \backslash S $ by extending the operators $P_0, P_1,\ldots,P_l$.
+The representations $\mathbf{f}_0, \mathbf{f}_1, \dots, \mathbf{f}_l$ can be extended to a new point $\tilde{i} \in \Gamma \backslash S $ by extending the operators $P_0, P_1,\ldots,P_l$.
 %
-For example, $f_0(\tilde{i}) = \sum_{i \in S} p_0(\tilde{i}, i)f(i)$ and
-$f_1(\tilde{i}) = f_0(\tilde{i}) + \sum_{(i \in S)}p_1(\tilde{i}, i)\delta_1(i)$.
+For example, $\mathbf{f}_0(\tilde{i}) = \sum_{i \in S} p_0(\tilde{i}, i)\mathbf{f}(i)$ and
+$\mathbf{f}_1(\tilde{i}) = \mathbf{f}_0(\tilde{i}) + \sum_{(i \in S)}p_1(\tilde{i}, i)\delta_1(i)$.
 
@@
  \begin{equation} \label{eq:LP_example}
-f(x) = \left\{
+\mathbf{f}(x) = \left\{
 \begin{array}{l l}
@@
 
-In this work, LP is applied to extend a high-dimensional function $f:\mathbb{R}^d \rightarrow \mathbb{R}^n$,
+In this work, LP is applied to extend a high-dimensional function $\mathbf{f}:\mathbb{R}^d \rightarrow \mathbb{R}^n$,
 which maps a set of points in the NIV space to their values $\mathbf{Y}(i)$ in the observable space.
@@
 %
-The values of function $f:\mathbb{R}^d \rightarrow \mathbb{R}^n$ are known on the subset $S = \{\Psi(i)\}$, with $f(\Psi(i)) = Y(i)$.
+The values of function $\mathbf{f}:\mathbb{R}^d \rightarrow \mathbb{R}^n$ are known on the subset $S = \{\Psi(i)\}$, with $\mathbf{f}(\Psi(i)) = \mathbf{Y}(i)$.
 
-A na\"{\i}ve way to extend $f$ to a new data point $\Psi(\tilde{i})$ is find the point's nearest neighbors in NIV space and average their function values.
+A na\"{\i}ve way to extend $\mathbf{f}$ to a new data point $\Psi(\tilde{i})$ is find the point's nearest neighbors in NIV space and average their function values.
 %
@@
 %
-The LP algorithm automatically finds the nearest neighborhood radius for each new point $\Psi(\tilde{i})$.
+The LP algorithm sets the appropriate nearest neighborhood radius for each new point $\Psi(\tilde{i})$.
 %
-This radius will be large in smooth regions of the function, and small in regions in which $f$ contains higher frequency components.
+This radius will be large in smooth regions of the function, and small in regions in which $\mathbf{f}$ contains higher frequency components.
 %
@@
 \item
-Use the smoothing operators to obtain a multiscale representation (see Eq. \ref{eq:LP_multi_scale_app}) of $f:\Psi(i) \rightarrow Y(i)$.
+Use the smoothing operators to obtain a multiscale representation (see Eq. \ref{eq:LP_multi_scale_app}) of $\mathbf{f}:\Psi(i) \rightarrow Y(i)$.
 
@@
 \item
-Use the extended smoothing operators to approximate the value of $f_l(\tilde{i}) = f_0(\tilde{i}) + \sum_{k}\sum_{(i \in S)}p_k(\tilde{i}, i)\delta_k(i)$
+Use the extended smoothing operators to approximate the value of $\mathbf{f}_l(\tilde{i}) = \mathbf{f}_0(\tilde{i}) + \sum_{k}\sum_{(i \in S)}p_k(\tilde{i}, i)\delta_k(i)$
 
 \item
-The value $Y(\tilde{i})$ is approximated by $f_l(\Psi(\tilde{i}))$.
+The value $Y(\tilde{i})$ is approximated by $\mathbf{f}_l(\Psi(\tilde{i}))$.
 
@@
 \includegraphics[width=0.2\textwidth]{LP_figures/Scale11_S2}
-\caption{(Top) Approximation of function $f(x)$ as defined in \eqref{eq:LP_example} for ``scales'' 3, 5, 8, and 11 using Laplacian Pyramids; the true function is shown in blue and the LP approximation is shown in black.
+\caption{(Top) Approximation of the function $\mathbf{f}(x)$ defined in \eqref{eq:LP_example} for ``scales'' 3, 5, 8, and 11 using Laplacian Pyramids; the true function is shown in blue and the LP approximation is shown in black.
 (Bottom) The residual error in the LP approximation.}
@@
 
-\subsection{Chemical Reaction Network}\label{subsec:rxn_network}
+\subsection{A Chemical Reaction Network}\label{subsec:rxn_network}
 
-We first consider a chemical reaction network of multiple enzyme-substrate interactions \cite{zagaris2012stability}.
+We first consider a chemical reaction network involving  multiple enzyme-substrate interactions \cite{zagaris2012stability}.
 %
@@
 However, one can write four conservation equations (since total $E$, $S$, $D$, and $F$ are all conserved) to reduce the system to 6 dimensions
-(we take the order of dimensions to be $S$, $E$, $E:S$, $S:E$, $D:S^{*}$, $F:E^{*}$).
+(which we order as $S$, $E$, $E:S$, $S:E$, $D:S^{*}$, $F:E^{*}$).
 %
-In the parameter regime we consider, the system of ODEs exhibits a separation of time scales, so that initial conditions quickly approach a two-dimensional manifold.
+In the parameter regime we consider, the ODE approximation of this scheme exhibits a separation of time scales, so that initial conditions quickly approach a two-dimensional manifold.
 %
@@
 
-Although the dynamics of chemical reaction networks are typically described by a system of ODEs, the ODEs are only an approximation that holds in the limit of a large number of molecules.
+Although the dynamics of chemical reaction networks are typically described by a system of ODEs, the ODEs are only an approximation that holds
+in the limit of a large number of molecules.
 %
-When the number of molecules is small, the system is inherently stochastic and its dynamics can be simulated using the Gillespie Stochastic Simulation Algorithm (SSA) \cite{gillespie1977exact}.
+When the number of molecules is small, the system is inherently stochastic and its dynamics can be simulated using the
+Gillespie Stochastic Simulation Algorithm (SSA) \cite{gillespie1977exact}; at intermediate molecule counts, the Chemical Langevin approximation
+becomes useful.
 %
-We can control the level of noise by adjusting the volume $V$, and therefore adjusting the number of molecules, in the system (in our simulations, we take $V=10^5$).
+We can control the level of noise in our SSA by adjusting the volume $V$, and therefore adjusting the number of molecules, in the system (in our simulations, we take $V=10^5$).
 %
@@
 
-We generate 3000 random initial conditions $\mathbf{Y}_0(1), \dots, \mathbf{Y}_0(3000) \in \mathbb{R}^6$, enforcing that all concentrations must be non-negative.
+We generate 3000 random initial conditions $\mathbf{Y}_0(1), \dots, \mathbf{Y}_0(3000) \in \mathbb{R}^6$, enforcing that all concentrations
+must be non-negative.
 %
-We integrate each point $\mathbf{Y}_0(t)$ forward for 10 time units using the SSA to obtain point the $\mathbf{Y}(t) \in \mathbb{R}^6$;
+We evolve each point $\mathbf{Y}_0(t)$ forward for 10 time units using the SSA to obtain a point $\mathbf{Y}(t) \in \mathbb{R}^6$;
 from the time scales calculated from the linearized ODEs, 10 time units is sufficiently long for the initial points in the ODE system to converge to the two-dimensional manifold,
@@
 %
-In our stochastic simulations, the initial points converge to an approximate two-dimensional mainfold in expected value (see Figure \ref{fig:rxn_manifolds}).
+In our stochastic simulations, the initial points appear to converge to an approximate two-dimensional manifold
+(in expected value, see Figure \ref{fig:rxn_manifolds}).
 %
-We consider $\mathcal{Y} = \{ \mathbf{Y}(t): t=1, \dots, 3000 \}$ to be representative points on the underlying two-dimensional manifold;
+We consider $\mathcal{Y} = \{ \mathbf{Y}(t): t=1, \dots, 3000 \}$ to be representative points ``on" this apparent two-dimensional manifold;
 three-dimensional projections of the resulting manifold are shown in Figure \ref{fig:rxn_manifolds}.
@@
 %
-Figure \ref{fig:rxn_embedding} shows the two-dimensional NIV embeddings for the two different data sets; the embeddings appear consistent.
+Figure \ref{fig:rxn_embedding} shows the two-dimensional NIV embeddings for the two different data sets; the embeddings appear visually consistent.
 %
@@
 %
-We obtain a correlation of 0.97 and 0.95 for the first and second NIV, respectively, indicating that the two embeddings are in excellent agreement with each other.
+We obtain a correlation of 0.97 and 0.95 for the first and second NIV, respectively, indicating that the two embeddings are in quantitative
+agreement with each other.
 %
-We would like to note that both $\mathcal{Y}_1$ and $\mathcal{Y}_2$ are ``rich enough'' to recover the two-dimensional manifolds.
+Note that both $\mathcal{Y}_1$ and $\mathcal{Y}_2$ are high-dimensional enough (``rich enough'') to allows us to recover
+the common underlying two-dimensional manifold.
 %
 Clearly, other data sets consisting of different partial observations could be constructed such that that underlying manifold would not be recovered.
-
+%
+%%%RONEN
+%  this sentiment should go somewhere else -- else if we keep here, we can say for example a 2d vector of observations
+%  that by bad luck inlcludes a conserved quantity
+%
 \begin{figure*}[ht]
@@
 
-We then use NIV together with Laplacian Pyramids to predict the values of $S:E$ and $F:E^{*}$ for $\mathcal{Y}_2$.
+We then use NIV together with Laplacian Pyramids to estimate the values of $S:E$ and $F:E^{*}$ for $\mathcal{Y}_2$.
 %
@@
 %
-The normalized mean-squared error between the true and predicted values, defined as $\frac{\langle (y_{true}-y_{pred})^2 \rangle}{\langle y_{true}^2 \rangle}$ is 0.0372 and 0.0287 for $S:E$ and $F:E^{*}$, respectively.
+The normalized mean-squared error between the true and estimated values, defined as $\frac{\langle (y_{true}-y_{pred})^2 \rangle}{\langle y_{true}^2 \rangle}$ is 0.0372 and 0.0287 for $S:E$ and $F:E^{*}$, respectively.
 %
-Therefore, we can effectively predict the unobserved components in the reaction network using NIV together with LP.
+Therefore, we can effectively estimate the unobserved components in the reaction network using NIV together with LP.
 %
@@
 
-We first examine the benefits of NIV over diffusion maps \cite{coifman2005geometric}, another nonlinear dimensionality reduction techniques.
+We first compare NIV with direct diffusion maps \cite{coifman2005geometric}, an established nonlinear dimensionality reduction technique.
 %
@@
 
-The correlation between the NIV coordinates for the two data sets and the diffusion maps coordinates for the two data sets are shown in Figure \ref{fig:ala_corr}.
+The correlation between the NIV coordinates for the two data sets and the diffusion map coordinates for the two data sets are shown in Figure \ref{fig:ala_corr}.
 %
-The correlation between the two NIV embeddings is higher than the correlation between the two diffusion maps embeddings.
-Therefore, it is often advantageous to use NIV over diffusion maps if one wishes to obtain a consistent embedding and merge data sets from different observation domains.
+The correlation between the two NIV embeddings is higher than the correlation between the two diffusion map embeddings.
+Therefore, it appears advantageous to use NIV over diffusion maps if one wishes to obtain a consistent embedding, and merge data sets from different observation domains (as long as the two main assumptions underpinning the NIV process hold).
 
@@
 %
-The embedding is three-dimensional, and each coordinate can be (visually) associated with a physical variable:
+The embedding is three-dimensional, and visual inspection reveals that each coordinate can be directly linked with one physical variable:
 the first coordinate describes the flipping of atoms 1 and 3, the second coordinate describes the dihedral angle $\phi$, and the third coordinate describes the dihedral angle $\psi$.
@@
 
-Figure \ref{fig:ala_recon} shows the reconstructed position from partial observation versus true position for some selected atoms.
+Figure \ref{fig:ala_recon} shows the reconstructed position from partial observation versus true position for certain selected atoms.
 %
-The atoms that have significant movement (such as atoms 1 and 3) exhibit a strong correlation between the true and reconstructed positions.
+Atoms that move appreciably within the data set (such as atoms 1 and 3) exhibit a strong correlation between the true and reconstructed positions.
 %
@@
 %
-In both examples, we were able to uncover the intrinsic variables governing the underlying stochastic process, independent of the measurements or observations of the system.
+In both examples, we were able to uncover the intrinsic variables governing the underling stochastic process, independent of the measurements or observations of the system.
 %
@@
 %
-For many problems, one would like to say what types of measurement functions $f$ are rich enough, as knowing which functions retain sufficient detail of the underlying stochastic process can aid us in building accurate reduced models.
+For many problems, one would like to say what types of measurement functions $f$ are rich enough, 
+as knowing which functions retain sufficient detail of the underlying stochastic process can aid us in building accurate reduced models.
 %

### 2013-12-27 paper/draft7.tex
Added files related to paper

--- 2013-12-27 paper/draft6.tex
@@
 While this can be easily rectified by an affine transformation
-(see, by analogy, the discussion in \cite{lafon2006data}),
+(see, by analogy, the discussion in Lafon {\em et al}\cite{lafon2006data}),
 the problem becomes
@@
 more importantly, it enables the construction of an empirical mapping between these different
-observation ensembles, allowing us to complete partial measurements through the construction
-of nonlinear observers of unmeasured variables.
-%
-%%%RONEN
-% we need in the sentence above to mention that we have the analog of a ``training set"
+observation ensembles, allowing us to complete partial measurements for a test data set from a training data set 
+that consists of {\em different} observations.
 %
@@
 %
-This example will allow us to compate our approach to more common techniques,
+This example will allow us to compare our approach to more common techniques,
 such as diffusion maps for dimensionality reduction
-and nearest neighbor interpolation for observation reconstrcution.
+and nearest neighbor interpolation for observation reconstruction.
 %
@@
 (Top Right) Second set of observable variables.
-The ellipse illustrates the mapping of the sphere of the underlying samples into this (different) observable domain via a second observation function.}
-\label{fig:IntrinsicIllustration}
+The ellipse illustrates the mapping of the sphere of the underlying samples into this (different) observable domain via a second observation function.
+\label{fig:IntrinsicIllustration}}
 \end{figure*}
@@
 %
-We first scale the eigenvectors so that $\|\psi_i\| = T$, where $T$ is the number of data points;
+We first scale the eigenvectors so that $\|\psi_i\|_2 = T$, where $T$ is the number of data points;
 intuitively, a data set that has twice as many samples as another data set, but the same underlying geometry, should have twice the norm.
@@
 \item
-Apply eigenvector decomposition (EVD) on the normalized kernel and view the values of its principal eigenvectors (modulo the possibility of
+Apply eigenvector decomposition on the normalized kernel and view the values of its principal eigenvectors (modulo the possibility of
  ``higher harmonics", see text) as the intrinsic variables of the given observations.
@@
 %
-Clearly, other data sets consisting of different partial observations could be constructed such that that underlying manifold would not be recovered.
-%
-%%%RONEN
-%  this sentiment should go somewhere else -- else if we keep here, we can say for example a 2d vector of observations
-%  that by bad luck inlcludes a conserved quantity
-%
 \begin{figure*}[ht]
@@
 %
-In this work, we considered data sets that consist of different partial observations, but where each data set covers the entire domain and is well-sampled.
+In this work, we considered data sets that consist of different partial observations, but where each data set samples the entire underlying manifold.
 %
-However, NIV could also be used to merge data sets when each data set samples only a portion of the domain, provided there is enough overlap to ``register'' the embeddings.
+However, NIV could also be used to merge data sets when each data set samples only a portion of the manifold, provided there is enough overlap to ``register'' the embeddings.
 %
-Merging data sets that come from different portions of the domain would not only require scaling the embedding coordinates,
+Merging data sets that come from different portions of the manifold would not only require scaling the embedding coordinates,
 but also shifting and permuting the embedding coordinates.
 %
-The ability to merge data from different regions of the domain would then allow us to analyze systems where complete sampling is computationally intractable, such as molecular systems with high energy barriers separating regions of state space.
+The ability to merge data from different regions would then allow us to analyze systems where complete sampling is computationally intractable, such as molecular systems with high energy barriers separating regions of state space.
 
-In addition, we have stated that the measurement function $f$ has to be ``rich enough'' in order to recover the variables of the underlying stochastic process.
+In addition, we have stated that the measurement function $\mathbf{f}$ has to be ``rich enough'' in order to recover the variables of the underlying stochastic process.
 %
-However, determining whether $f$ meets such criteria is a non-trivial task.
+However, determining whether $\mathbf{f}$ meets such criteria is a non-trivial task.
 %
-For many problems, one would like to say what types of measurement functions $f$ are rich enough, 
+For many problems, one would like to say what types of measurement functions $\mathbf{f}$ are rich enough, 
 as knowing which functions retain sufficient detail of the underlying stochastic process can aid us in building accurate reduced models.

### 2013-12-27 paper/draft8.tex
Added files related to paper

--- 2013-12-27 paper/draft7.tex
@@
 and when different data sets are obtained using different instrumental modalities
-(such as when one wants to merge molecular dynamics data with, for example, spectral information {\color{red} TODO: add specific example of spectral information}).
+(such as when one wants to merge molecular dynamics data with, for example, spectral information).
 %
@@
 more importantly, it enables the construction of an empirical mapping between these different
-observation ensembles, allowing us to complete partial measurements for a test data set from a training data set 
+observation ensembles, allowing us to complete partial measurements in a test data set from a training data set
 that consists of {\em different} observations.
@@
 accurate interpolation tools must be available in the embedding space; to this
-end, here we chose to demonstrate the use of a multiscale Laplacian Pyramid approach \cite{rabin2012heterogeneous}.
-%%
-%%RONEN
-%% remember to add all the other options, like kriging..... in the discussion
-%%
+end, we will demonstrate the use of a multiscale Laplacian Pyramid approach \cite{rabin2012heterogeneous}.
 
-We will demonstrate our methodologies with two distinct illustrative examples.
+We will illustrate our methodologies with two distinct examples.
 %
 The first is a simulation
-of two Goldbeter-Koshland modules in an enzyme kinetics model using Gillespie's Stochastic Simulation
+of two Goldbeter-Koshland modules in an enzyme kinetics model using the Gillespie Stochastic Simulation
 Algorithm (SSA) \cite{gillespie1977exact};
@@
 This example will allow us to compare our approach to more common techniques,
-such as diffusion maps for dimensionality reduction
+such as diffusion maps \cite{coifman2005geometric} for dimensionality reduction
 and nearest neighbor interpolation for observation reconstruction.
@@
 Section \ref{sec:LapPyr} contains our discussion of Laplacian Pyramids that
-will be used for the completion of partial observations.
+is used for the completion of partial observations.
 %
-In section \ref{sec:examples} the results
-of the application of the approach to simulation data from our two illustrative examples is presented and discussed.
+In Section \ref{sec:examples}, the results
+of the application of the approach to simulation data from our two illustrative examples are presented and discussed.
 %
@@
 \end{equation}
-where $\mathbf{f}:\mathbb{R}^d \rightarrow \mathbb{R}^n$ is an unknown (possibly nonlinear, $d \ll n$) function and $\mathbf{x}(t)$ is a diffusion process that consists of $d$ underlying variables.
+where $\mathbf{f}:\mathbb{R}^d \rightarrow \mathcal{M}$ is an unknown (possibly nonlinear, $d \ll n$) function, $\mathcal{M} \subset \mathbb{R}^n$ is a $d$-dimensional manifold, and $\mathbf{x}(t)$ is a diffusion process that consists of $d$ underlying variables.
 %
@@
 (see Figure \ref{fig:IntrinsicIllustration} for a schematic illustration).
-%
-We remark that the available data should be rich enough, i.e., consists of a sufficient amount of ``historical" data with sufficient variability, in order to get a full empirical model.
+
+We remark that the available data should be ``rich enough'', i.e., consist of a sufficient amount of historical data with adequate variability, in order to obtain the full empirical model.
 
@@
 (1) The underlying diffusion process implies that a short trajectory of successive samples mainly consists of diffusion noise,
-and hence, creates a ``sphere" of samples in the underlying domain.
+and hence, creates a ``sphere" of samples in the underlying domain $\mathbb{R}^d$.
 %%
@@
 %
-In this work, the identification of the associated ellipse of samples according to the time labels enables us
+In this work, the identification of the associated ellipse of samples according to the time trajectory of our data enables us
 to estimate the tangent planes of the observable manifolds {\em via the principal components of the covariance matrices of the samples in these ellipses}.
@@
 \end{equation}
-where $\widehat{\boldsymbol{\mu}}(t)$ is the empirical mean of the samples.
+where $\widehat{\boldsymbol{\mu}}(t)$ is the empirical mean of the short trajectory of samples.
 %
@@
 \begin{equation}
-	d^2(\mathbf{Y}(t), \mathbf{Y}(\tau)) = 2 (\mathbf{Y}(t) - \mathbf{Y}(\tau))^T(\mathbf{C}(t) + \mathbf{C}(\tau))^{-1}(\mathbf{Y}(t) - \mathbf{Y}(\tau));
+	d^2(\mathbf{Y}(t), \mathbf{Y}(\tau)) = 2 (\mathbf{Y}(t) - \mathbf{Y}(\tau))^T(\mathbf{C}(t) + \mathbf{C}(\tau))^{\dagger}(\mathbf{Y}(t) - \mathbf{Y}(\tau));
 	\label{eq:mahalanobis}
 \end{equation}
-this is the Mahalanobis distance.
+this is the Mahalanobis distance (and $^{\dagger}$ denotes a pseudoinverse, as discussed below).
 %
@@
 %
-We would like to note that, in general, $\mathbf{f}$ being bi-Lipschitz implies that $\mathbf{f}$ is invertible.
-%%%
-%%%RONEN
-% this is nontrivial and it requires discussion
-%  about what manifold thing need to live on for this to be doable
+We would like to note that, in general, $\mathbf{f}$ being bi-Lipschitz implies that $\mathbf{f}$ is invertible (on the $d$-dimensional
+manifold $\mathcal{M}$).
 %
@@
 %
-Consistent rank estimates among the covariance matrices may imply that the measurements are ``rich enough", and hence, may be good indicators for the dimension $d$. 
+Consistent rank estimates among these local covariance matrices are taken to imply that the measurements are ``rich enough", and hence, may be good indicators for the dimension $d$.
 %
-In addition, since usually the dimension $d$ of the underlying process is smaller than the dimension of the measured process $n$, the covariance matrix is singular and non-invertible. 
+Since the dimension $d$ of the underlying process is typically considerably smaller than the dimension of the measured process $n$,
+the covariance matrix is singular and non-invertible;
 %
-Thus, in practice we use the pseudo-inverse to compute the inverse matrices in \eqref{eq:mahalanobis}.
+thus, we use the pseudo-inverse in \eqref{eq:mahalanobis}.
 
@@
 In practice, we set $\varepsilon$ to be the median of the elements of the kernel matrix.
-%%%RONEN
-%   this is also not really clear, and "effectively connected to half" is
-%   also not true -- just need to be said carefully
 %
@@
 
-Let $\mathbf{D}$ be a diagonal matrix whose elements are the sums of rows of $\mathbf{W}$, and let $\mathbf{W}^{\mathrm{norm}} = \mathbf{D}^{-1/2}\mathbf{W}\mathbf{D}^{-1/2}$
+Let $\mathbf{D}$ be a diagonal matrix whose elements are the row sums of $\mathbf{W}$, and let $\mathbf{W}^{\mathrm{norm}} = \mathbf{D}^{-1/2}\mathbf{W}\mathbf{D}^{-1/2}$
 be a normalized kernel that shares its eigenvectors with the normalized graph-Laplacian $\mathbf{I}-\mathbf{W}^{\mathrm{norm}} \:$ \cite{chung1997spectral}.
@@
 %
-The next few eigenvectors can be argued to describe the geometry of the underlying manifold \cite{coifman2005geometric}
+The next few eigenvectors can be argued to describe the geometry of the underlying manifold \cite{coifman2005geometric}.
 %
 However, some eigenvectors can be higher harmonics of the same principal direction along the data manifold.
-This is analogous to how the eigenfunctions $\cos x$ and $\cos 2x$ of the usual Laplacian in one spatial dimension
+(this is analogous to how the eigenfunctions $\cos x$ and $\cos 2x$ of the usual Laplacian in one spatial dimension
 and with no flux boundary conditions are one-to-one with the values of $x$ for $0 \le x \le 1$);
-so one must check for correlations between the eigenvectors before selecting those that describe the underlying manifold geometry.
+one must check for correlations between the eigenvectors before selecting those that describe the underlying manifold geometry.
 %
@@
 
-Ignoring these ``higher harmonics", each retained eigenvector then describes an intrinsic variable for the data set of interest.
+Ignoring the higher harmonics, each retained eigenvector then describes an intrinsic variable for the data set of interest.
 %
-Because eigenvectors can be determined up to a scaling factor,
-we must scale the eigenvectors from different data sets so that the resulting embeddings are consistent.
+%Because eigenvectors can be determined up to a scaling factor,
+We must normalize the eigenvectors from different data sets so that the resulting embeddings are consistent.
 %
-We first scale the eigenvectors so that $\|\psi_i\|_2 = T$, where $T$ is the number of data points;
+We first scale the eigenvectors so that $\|\psi_i\| = T$, where $T$ is the number of data points;
 intuitively, a data set that has twice as many samples as another data set, but the same underlying geometry, should have twice the norm.
@@
 %
-More specifically, thinking of the ambient space coordinates, $\mathbf{Y}$, as functions on the data,
+More specifically, viewing the ambient space coordinates $\mathbf{Y}$ as functions on the data,
 we want to estimate $\mathbf{Y}$ for new points $\mathbf{x}$.
 %
-Laplacian Pyramids (LP) is a multiscale algorithm for approximating and extending an empirical function $\mathbf{f}$ defined on a set of points.
+Laplacian Pyramids (LP) is a multiscale algorithm for extending an empirical function $\mathbf{f}$ defined on a set of points
+to new points not in the dataset.
 %
@@
 Recently, LP was used to extend nonlinear embedding coordinates to new high-dimensional data points \cite{rabin2012heterogeneous}.
-%
-%%%RONEN
-% this is colloquially correct but inconsistent
 %
@@
 As the function becomes more oscillatory ($\pi \le x \le 7.5\pi$ and $7.5\pi \le x \le 10\pi$),
-a finer representation (meaning a larger number of scales  $l$) is required to capture its behavior.
+a finer representation, and a larger number of scales  $l$, is required to capture its behavior.
 \end{widetext}
 
-In this work, LP is applied to extend a high-dimensional function $\mathbf{f}:\mathbb{R}^d \rightarrow \mathbb{R}^n$,
+In this work, LP is applied to extend a high-dimensional function $\mathbf{f}:\mathbb{R}^d \rightarrow \mathcal{M}$,
 which maps a set of points in the NIV space to their values $\mathbf{Y}(i)$ in the observable space.
@@
 %
-The values of function $\mathbf{f}:\mathbb{R}^d \rightarrow \mathbb{R}^n$ are known on the subset $S = \{\Psi(i)\}$, with $\mathbf{f}(\Psi(i)) = \mathbf{Y}(i)$.
+The values of function $\mathbf{f}:\mathbb{R}^d \rightarrow \mathcal{M}$ are known on the subset $S = \{\Psi(i)\}$, with $\mathbf{f}(\Psi(i)) = \mathbf{Y}(i)$.
 
@@
 \item Given a new point $\Psi(\tilde{i})$ in NIV, extend the smoothing operators $P_0, P_1, \ldots, P_l$  by
-$p_l(\Psi(\tilde{i}), \Psi(i)) = s_0^{-1}(\Psi(i))w_0(\Psi(\tilde{i}),\Psi(i))$.
+$p_l(\Psi(\tilde{i}), \Psi(i)) = s_l^{-1}(\Psi(i)) w_l(\Psi(\tilde{i}),\Psi(i))$.
 
@@
 When the number of molecules is small, the system is inherently stochastic and its dynamics can be simulated using the
-Gillespie Stochastic Simulation Algorithm (SSA) \cite{gillespie1977exact}; at intermediate molecule counts, the Chemical Langevin approximation
+Gillespie Stochastic Simulation Algorithm (SSA) \cite{gillespie1977exact}; at intermediate molecule counts, the chemical Langevin approximation \cite{gillespie2000chemical}
 becomes useful.
@@
 We evolve each point $\mathbf{Y}_0(t)$ forward for 10 time units using the SSA to obtain a point $\mathbf{Y}(t) \in \mathbb{R}^6$;
-from the time scales calculated from the linearized ODEs, 10 time units is sufficiently long for the initial points in the ODE system to converge to the two-dimensional manifold,
+according to the time scales calculated from the linearized ODEs, 10 time units is sufficiently long for the initial points in the ODE system to converge to the two-dimensional manifold,
 but not long enough for the points to converge to a one-dimensional curve or to the final steady state (see Appendix \ref{app:rxn} for more details).
@@
 %
-Data set 1, denoted $\mathcal{Y}_1$, consists of $\mathbf{Y}(1), \dots, \mathbf{Y}(2000)$, restricted to components $S$, $E$, $S:E$, and $F:E^{*}$, i.e.
+Data set 1, denoted $\mathcal{Y}_1$, consists of $\mathbf{Y}(1), \dots, \mathbf{Y}(2000)$, restricted to components $S$, $E$, $S:E$, and $F:E^{*}$, i.e.,
 $$\mathcal{Y}_1 = \left\{
@@
 %
-Data set 2, denoted $\mathcal{Y}_2$, consists of $\mathbf{Y}(1500), \dots, \mathbf{Y}(3000)$, restricted to components $S$, $E$, $E:S$, and $D:S^{*}$, i.e.
+Data set 2, denoted $\mathcal{Y}_2$, consists of $\mathbf{Y}(1500), \dots, \mathbf{Y}(3000)$, restricted to components $S$, $E$, $E:S$, and $D:S^{*}$, i.e.,
 $$\mathcal{Y}_2 = \left\{
@@
 %
-Note that both $\mathcal{Y}_1$ and $\mathcal{Y}_2$ are high-dimensional enough (``rich enough'') to allows us to recover
+We would like to note that both $\mathcal{Y}_1$ and $\mathcal{Y}_2$ are sufficiently high-dimensional (``rich enough'') to allows us to recover
 the common underlying two-dimensional manifold.
@@
 %
-The normalized mean-squared error between the true and estimated values, defined as $\frac{\langle (y_{true}-y_{pred})^2 \rangle}{\langle y_{true}^2 \rangle}$ is 0.0372 and 0.0287 for $S:E$ and $F:E^{*}$, respectively.
+The normalized mean-squared errors between the true and estimated values for $S:E$ and $F:E^{*}$, defined as $\frac{\langle (y_{true}-y_{pred})^2 \rangle}{\langle y_{true}^2 \rangle}$, are 0.0372 and 0.0287 , respectively.
 %
@@
 %
-Our simulations are concentrated around minimum B in the free energy surface, located at $\phi=-65^{\circ}$, $\psi=150^{\circ}$.
+Our simulations are concentrated around minimum B in the free energy surface, located at $\phi \approx -65^{\circ}$, $\psi \approx 150^{\circ}$.
 %
@@
 The correlation between the two NIV embeddings is higher than the correlation between the two diffusion map embeddings.
-Therefore, it appears advantageous to use NIV over diffusion maps if one wishes to obtain a consistent embedding, and merge data sets from different observation domains (as long as the two main assumptions underpinning the NIV process hold).
+Therefore, it appears advantageous to use NIV over diffusion maps if one wishes to obtain a consistent embedding, and merge data sets from different observation domains (as long as the two main assumptions underpinning the NIV algorithm hold).
 
@@
 %
-We have 20000 data points $\mathbf{Y}(1), \dots, \mathbf{Y}(20000)$, where every 100 data points comes from one continuous simulation trajectory.
+We have 20000 data points $\mathbf{Y}(1), \dots, \mathbf{Y}(20000)$, where every 100 data points come from one continuous simulation trajectory.
 %
@@
 %
-Atoms that move appreciably within the data set (such as atoms 1 and 3) exhibit a strong correlation between the true and reconstructed positions.
+The strong correlation between the true and reconstructed positions is easier to appreciate for
+atoms that move substantially within the data set (such as atoms 1 and 3).
 %
-Figure \ref{fig:ala_molecules} shows molecular structures for the true and reconstructed configurations for some data points;
+Figure \ref{fig:ala_molecules} shows molecular structures for the true and reconstructed configurations for selected data points;
 there is qualitative agreement between the true and reconstructed configurations.
 
-We would also like to demonstrate the effectiveness of LP over other reconstruction techniques.
+For a brief comparison of LP over other reconstruction techniques, we also reconstruct configurations from the
+NIV components using simple nearest-neighbor interpolation.
 %
-We predict the configurations from the NIV components using simple nearest-neighbor interpolation.
-%
-The average reconstruction error, scaled by the average bond length within the molecule, is shown in Figure \ref{fig:ala_mse}.
-%
-The LP approach outperforms simple nearest neighbor search for all of the atoms.
+The average reconstruction error, scaled by the average bond length within the molecule, is shown in Figure \ref{fig:ala_mse};
+LP arguably outperforms simple nearest neighbor search for all of the atoms.
 
@@
 \section{Conclusions} \label{sec:conclusions}
-We have used nonlinear intrinsic variables to analyze two complex stochastic molecular systems: a simulation of a chemical reaction network and a molecular dynamics simulation of alanine dipeptide.
 %
-In both examples, we were able to uncover the intrinsic variables governing the underling stochastic process, independent of the measurements or observations of the system.
+We have used Nonlinear Intrinsic Variables to analyze two complex atomistic simulations: a stochastic simulation of a chemical reaction network and a molecular dynamics simulation of alanine dipeptide.
 %
-The uniqueness of the embedding coordinates allowed us to compare and merge data sets from different measurement functions, and therefore allowed us to use Laplacian Pyramids to complete partial observations.
+In both examples, we were able to uncover the intrinsic variables governing the underlying stochastic process, independent
+of the particular measurement or observation  of the system (under the conditions mentioned).
+%
+The uniqueness of the embedding coordinates allowed us to compare and merge data sets from different measurement functions,
+and therefore allowed us to use an interpolation/extension scheme (here Laplacian Pyramids) to complete partial observations.
+%
+Different interpolation techniques (e.g.
+kriging \cite{matheron1963principles, matheron1973intrinsic}, geometric harmonics \cite{coifman2006geometric},  versions of the Nystr\"{o}m extension) can and should
+be explored, since the performance of such techniques may well be problem dependent, especially for multiscale, complex simulation data.
 
-There are many open research questions and interesting research directions that may be explored.
+There are many open questions leading to interesting research directions to be explored.
 %
-In this work, we considered data sets that consist of different partial observations, but where each data set samples the entire underlying manifold.
+In this work, we considered data sets that consist of different partial observations, but in which each data set samples the entire underlying manifold in
+what we loosely referred to as a ``representative enough" way.
 %
@@
 Merging data sets that come from different portions of the manifold would not only require scaling the embedding coordinates,
-but also shifting and permuting the embedding coordinates.
+but also shifting and possibly permuting the embedding coordinates (in this spirit, see the discussion in Reference~\onlinecite{lafon2006data}).
 %
-The ability to merge data from different regions would then allow us to analyze systems where complete sampling is computationally intractable, such as molecular systems with high energy barriers separating regions of state space.
+The ability to merge data from different regions would then allow us to analyze systems where complete sampling is computationally intractable,
+such as molecular systems with several high energy barriers separating regions of state space.
 
-In addition, we have stated that the measurement function $\mathbf{f}$ has to be ``rich enough'' in order to recover the variables of the underlying stochastic process.
+%In addition, we have stated that the measurement function $\mathbf{f}$ has to be ``rich enough'' in order to recover the variables of the underlying stochastic process.
+%However, determining whether $\mathbf{f}$ meets such criteria is a non-trivial task.
+%For many problems, one would like to say what types of measurement functions $\mathbf{f}$ are rich enough,
+%as knowing which functions retain sufficient detail of the underlying stochastic process can aid us in building accurate reduced models.
 %
-However, determining whether $\mathbf{f}$ meets such criteria is a non-trivial task.
 %
-For many problems, one would like to say what types of measurement functions $\mathbf{f}$ are rich enough, 
-as knowing which functions retain sufficient detail of the underlying stochastic process can aid us in building accurate reduced models.
+Other issues, such as accurately estimating the covariance matrices required to compute the Mahalanobis distance, are also of current research interest.
 %
-Other issues, such as accurately estimating the covariances matrices required to compute the Mahalanobis distance, are also of current research interest.
+It is clearly necessary to link this type of calculation with modern estimation techniques for (multiscale) diffusions \cite{ait2002maximum, ait2003effects, ait2008closed}
+to test the appropriateness of the window sampling lengths selected; this will determine the accuracy of the noise covariance estimation by eliminating
+the bias due to drift variations.
 %
-We are confident that the exploration of these open questions will lead to many interesting applications for our methodologies, such as merging data from different molecular simulations, or merging simulation data with experimental data.
+We are confident that the exploration of these open questions will enable the use of our methodology in many  interesting applications,
+such as merging data from molecular simulations at different levels of granularity, or merging simulation data with experimental observations.
 
@@
 %
-Therefore, we choose to integrate forward for 10 time units to find points on a perceived two-dimensional manifold.
+Therefore, we choose to evolve forward for 10 time units to find points on a perceived two-dimensional manifold.
 

### 2013-12-27 paper/draft9.tex
Added files related to paper

--- 2013-12-27 paper/draft8.tex
@@
 \usepackage[size=small]{caption}
-%\usepackage{etoolbox}
-%\AtBeginEnvironment{algorithm}{\noindent\hrulefill\par\nobreak\vskip-5pt}
-
 \usepackage{newfloat}
@@
 \makeatother
-
-%\usepackage{float}
-%\floatstyle{ruled}
-%\newfloat{algorithm}{t}{loa}
-%\floatname{algorithm}{Algorithm}
 
@@
 While this can be easily rectified by an affine transformation
-(see, by analogy, the discussion in Lafon {\em et al}\cite{lafon2006data}),
+(see, by analogy, the discussion in Lafon {\em et al.}\cite{lafon2006data}),
 the problem becomes
@@
 %
-Clearly, the ability to systematically construct a {\em unique, consistent} reduction
-coordinate set, shared by all measurement ensembles and observation modalities -
-what we will call Nonlinear Intrinsic Variables - is invaluable.
+Clearly, the ability to systematically construct a {\em unique} and {\em consistent} reduction
+coordinate set, shared by all measurement ensembles and observation modalities, is invaluable.
+%
+We will call these coordinates Nonlinear Intrinsic Variables.
 %
@@
 %
-The measured process is assumed (and this is a critical assumption) to be a manifestation, in an observable domain, of a low-dimensional diffusion process. Thus, it can be expressed as
+We impose two critical assumptions. First, the measured process is assumed to be a manifestation in an observable domain of a low-dimensional diffusion process. Thus, it can be expressed by
 \begin{equation}
-	\mathbf{Y}(t) = \mathbf{f}(\mathbf{x}(t))
+	\mathbf{Y}(t) = f(\mathbf{x}(t)),
 \end{equation}
-where $\mathbf{f}:\mathbb{R}^d \rightarrow \mathcal{M}$ is an unknown (possibly nonlinear, $d \ll n$) function, $\mathcal{M} \subset \mathbb{R}^n$ is a $d$-dimensional manifold, and $\mathbf{x}(t)$ is a diffusion process that consists of $d$ underlying variables.
+where $f:\mathbb{R}^d \rightarrow \mathcal{M}$ is an unknown (possibly nonlinear) function, $\mathcal{M} \subset \mathbb{R}^n$ is a $d$-dimensional manifold,
+and $\mathbf{x}(t)$ is a diffusion process that consists of $d$ underlying variables (with $d \ll n$).
 %
-The dynamics of the diffusion process in terms of its underlying variables is described by normalized stochastic differential equations as
+Second, the dynamics of the diffusion process in each of its underlying variables are described by normalized stochastic differential equations as
 \begin{equation}
-	d x_i(t) = a_i (x_i(t)) dt + d w_i(t), \ i=1,\ldots,d
+	d x_i(t) = a_i (x_i(t)) dt + d w_i(t), \ i=1,\ldots,d,
 \end{equation}
-where $a_i$ are unknown drift functions and $\dot{w}_i(t)$ are independent white noises (this independence is our second critical assumption).
+where $a_i$ are unknown drift functions and $\dot{w}_i(t)$ are independent white noises.
+%
+The independence is our second critical assumption.
 
-Given a sequence of samples $\mathbf{Y}(t), \ t=1,\ldots,T$, we present an empirical method to construct a unique and consistent reduction coordinate set, represented here by $\mathbf{x}(t)$.
+Given a sequence of samples $\mathbf{Y}(t), \ t=1,\ldots,T$, we present an empirical method to construct a unique and consistent reduction coordinate set, represented here by $\mathbf{x}(t)$ \cite{singer2008non}.
 %
-By relying on the fact that the empirical method we will describe is independent of the observation function $\mathbf{f}$,
+Because the empirical method we will describe is independent of the observation function $\mathbf{f}$,
 we refer to the coordinates of $\mathbf{x}(t)$ as Nonlinear Intrinsic Variables (NIV).
@@
 (see Figure \ref{fig:IntrinsicIllustration} for a schematic illustration).
-
+%
 We remark that the available data should be ``rich enough'', i.e., consist of a sufficient amount of historical data with adequate variability, in order to obtain the full empirical model.
@@
 \begin{figure*}[ht]
-\includegraphics[width=0.7\textwidth]{IntrinsicEmbeddingIllustration2}
+\includegraphics[width=5in]{fig1}
 \caption{Illustration of the nonlinear embedding that yields an intrinsic representation independent of the measurement function $\mathbf{f}$.
@@
 \begin{equation}
-	\widehat{\mathbf{C}}(t) = \sum \limits _{\tau = t-L}^{t+L} (\mathbf{Y}(\tau) - \widehat{\boldsymbol{\mu}}(t))(\mathbf{Y}(\tau) - \widehat{\boldsymbol{\mu}}(t))^T
+	\widehat{\mathbf{C}}(t) = \sum \limits _{\tau = t-L}^{t+L} (\mathbf{Y}(\tau) - \widehat{\boldsymbol{\mu}}(t))(\mathbf{Y}(\tau) - \widehat{\boldsymbol{\mu}}(t))^T,
 	\label{eq:cov}
@@
 \begin{equation}
-	d^2(\mathbf{Y}(t), \mathbf{Y}(\tau)) = 2 (\mathbf{Y}(t) - \mathbf{Y}(\tau))^T(\mathbf{C}(t) + \mathbf{C}(\tau))^{\dagger}(\mathbf{Y}(t) - \mathbf{Y}(\tau));
+	d^2(\mathbf{Y}(t), \mathbf{Y}(\tau)) = 2 (\mathbf{Y}(t) - \mathbf{Y}(\tau))^T(\widehat{\mathbf{C}}(t) + \widehat{\mathbf{C}}(\tau))^{\dagger}(\mathbf{Y}(t) - \mathbf{Y}(\tau));
 	\label{eq:mahalanobis}
@@
 %
-Thus, by assuming that the observation function $\mathbf{f}$ is bi-Lipschitz and smooth and by using local linearization of the function, i.e., $\mathbf{Y}(t) = \mathbf{J}(t) \mathbf{x}(t) + \boldsymbol{\epsilon}(t)$ where $\mathbf{J}(t)$ is the Jacobian of $\mathbf{f}(\mathbf{x}(t))$ and $\boldsymbol{\epsilon}(t)$ is the residual consisting of higher-order terms, it was shown by Singer and Coifman \cite{singer2008non} that $\mathbf{C}(t) = \mathbf{J}(t)\mathbf{J}^T(t)$ and that the Mahalanobis distance approximates the Euclidean distance between the corresponding samples of the underlying process to second order, i.e.,
+Thus, by assuming that the observation function $\mathbf{f}$ is bi-Lipschitz and smooth, and by using local linearization of the function, i.e., $\mathbf{Y}(t) = \mathbf{J}(t) \mathbf{x}(t) + \boldsymbol{\epsilon}(t)$ where $\mathbf{J}(t)$ is the Jacobian of $\mathbf{f}(\mathbf{x}(t))$ and $\boldsymbol{\epsilon}(t)$ is the residual consisting of higher-order terms, it was shown by Singer and Coifman \cite{singer2008non} that $\mathbf{C}(t) = \mathbf{J}(t)\mathbf{J}^T(t)$ and that the Mahalanobis distance approximates the Euclidean distance between the corresponding samples of the underlying process to second order, i.e.,
 \begin{equation}
@@
 \begin{equation}
-	W_{t,\tau} = \exp \left\{ - \frac{ d^2(\mathbf{Y}(t), \mathbf{Y}(\tau) )} {\varepsilon}\right\}
+	W_{t,\tau} = \exp \left\{ - \frac{ d^2(\mathbf{Y}(t), \mathbf{Y}(\tau) )} {\varepsilon}\right\},
 	\label{eq:kernel}
 \end{equation}
-where $\varepsilon$ is the kernel scale, which can be set according to References \onlinecite{coifman2008graph, hein2005intrinsic}.
+where $\varepsilon$ is the kernel scale, which can be set according to Hein and Audibert \cite{hein2005intrinsic} and Coifman {\em et al.} \cite{coifman2008graph}.
 %
@@
 %
-In practice, we set $\varepsilon$ to be the median of the elements of the kernel matrix.
+In practice, we set $\varepsilon$ to be the median of the pairwise distances.
 %
-According to the graph interpretation, this implies a well connected graph because each measurement is effectively connected to half of the other measurements \cite{rohrdanz2011determination}.
+According to the graph interpretation, this implies a well-connected graph because each measurement is effectively connected to half of the other measurements \cite{rohrdanz2011determination}.
 
@@
 However, some eigenvectors can be higher harmonics of the same principal direction along the data manifold.
-(this is analogous to how the eigenfunctions $\cos x$ and $\cos 2x$ of the usual Laplacian in one spatial dimension
-and with no flux boundary conditions are one-to-one with the values of $x$ for $0 \le x \le 1$);
+This is analogous to how the eigenfunctions $\cos x$ and $\cos 2x$ of the usual Laplacian in one spatial dimension
+and with no flux boundary conditions are one-to-one with the values of $x$ for $0 \le x \le 1$;
 one must check for correlations between the eigenvectors before selecting those that describe the underlying manifold geometry.
@@
 We first scale the eigenvectors so that $\|\psi_i\| = T$, where $T$ is the number of data points;
-intuitively, a data set that has twice as many samples as another data set, but the same underlying geometry, should have twice the norm.
+intuitively, the norm of the embedding eigenvectors should scale with the size of the data set to make the embedding coordinates of each data point independent of the size of the data set.
+%twice as many samples as another data set, but with the same underlying geometry, should have twice the norm.
 %
@@
 We will return to the issue of embedding consistency for different data sets in our concluding discussion; for the moment,
-we will assume that our different sets sample the same region of data space in a representative enough way for the
-correspondence between the sequences of retained eigenvectors for different embeddings to be obvious.
+we will assume that our different sets sample the same region of data space in a representative enough way such that the
+correspondence between the sequences of retained eigenvectors for different embeddings is obvious.
 
@@
 \item
-Apply eigenvector decomposition on the normalized kernel and view the values of its principal eigenvectors (modulo the possibility of
- ``higher harmonics", see text) as the intrinsic variables of the given observations.
+Apply eigenvalue decomposition to the normalized kernel and view the values of its principal eigenvectors (modulo the possibility of
+ ``higher harmonics", see text) as the Nonlinear Intrinsic Variables (NIV) of the given observations.
 
@@
 %
-More specifically, viewing the ambient space coordinates $\mathbf{Y}$ as functions on the data,
+More specifically, viewing the ambient space coordinates $\mathbf{Y}$ as functions on the low-dimensional data $\mathbf{x} \in \mathbb{R}^d$,
 we want to estimate $\mathbf{Y}$ for new points $\mathbf{x}$.
@@
 \begin{equation}
-p_0(i, j)= s_0^{-1}(t)w_0(i, j),
+p_0(i, j)= s_0^{-1}(i)w_0(i, j),\: i, j \in \Gamma,
 \end{equation}
-where $w_0(i, j)=e^{-d^2(i, j) / \sigma_0}$ and $s_0(t)=\sum_{j \in S}w_0(i, j)$ is the normalizing term.
+where $w_0(i, j)=e^{-d^2(i, j) / \sigma_0}$ and $s_0(i)=\sum_{j \in S}w_0(i, j)$ is the normalizing term.
 %
@@
 %
-A different, point-wise adaptive approach is described by Buchman {\em et al} \cite{buchman2011high}:
+A different, point-wise adaptive approach is described by Buchman {\em et al.} \cite{buchman2011high}:
 high-dimensional hurricane tracks were estimated from low dimensional embedding coordinates using a weighted average of the points close to $\Psi(\tilde{i})$ in the embedded space.
@@
 %
-The LP algorithm sets the appropriate nearest neighborhood radius for each new point $\Psi(\tilde{i})$.
+The LP algorithm finds the appropriate nearest neighborhood radius for each new point $\Psi(\tilde{i})$.
 %
@@
 %
-The LP approximation of a new, high dimensional value $Y(\tilde{i})$ is calculated by a weighted average of the function values that belong to the neighboring points.
+The LP approximation of a new, high dimensional point $\mathbf{Y}(\tilde{i})$ is calculated by a weighted average of the function values that belong to the neighboring points.
 %
@@
 \item
-Construct a set of smoothing operators $P_0, P_1, \ldots, P_l$  based on intrinsic pairwise distances $d(\Psi(i),\Psi(j))$.
+Construct a set of smoothing operators $P_0, P_1, \ldots, P_l$  based on intrinsic pairwise distances $d(\Psi(i),\Psi(j))$ (where $d$ is typically the Euclidean distance).
 
@@
 \item
-Use the extended smoothing operators to approximate the value of $\mathbf{f}_l(\tilde{i}) = \mathbf{f}_0(\tilde{i}) + \sum_{k}\sum_{(i \in S)}p_k(\tilde{i}, i)\delta_k(i)$
+Use the extended smoothing operators to approximate the value of $Y(\tilde{i})$ as $Y(\tilde{i}) \approx \mathbf{f}_l(\Psi(\tilde{i})) = \mathbf{f}_0(\Psi(\tilde{i})) + \sum_{k}\sum_{(i \in S)}p_k(\Psi(\tilde{i}), \Psi(i))\delta_k(\Psi(i))$
 
-\item
-The value $Y(\tilde{i})$ is approximated by $\mathbf{f}_l(\Psi(\tilde{i}))$.
+%\item
+%The value $Y(\tilde{i})$ is approximated by $\mathbf{f}_l(\Psi(\tilde{i}))$.
 
@@
 \begin{figure*}[ht]
-\includegraphics[width=0.2\textwidth]{LP_figures/Scale3_S2}
-\includegraphics[width=0.2\textwidth]{LP_figures/Scale5_S2}
-\includegraphics[width=0.2\textwidth]{LP_figures/Scale8_S2}
-\includegraphics[width=0.2\textwidth]{LP_figures/Scale11_S2}
-\caption{(Top) Approximation of the function $\mathbf{f}(x)$ defined in \eqref{eq:LP_example} for ``scales'' 3, 5, 8, and 11 using Laplacian Pyramids; the true function is shown in blue and the LP approximation is shown in black.
-(Bottom) The residual error in the LP approximation.}
+\includegraphics[width=6in]{fig2}
+\caption{Approximation of the function $\mathbf{f}(x)$ defined in \eqref{eq:LP_example} using Laplacian Pyramids for ``scales'' (a) 3, (b) 5, (c) 8, and (d) 11.
+The top shows the true function in blue and the LP approximation is shown in black, and the bottom shows the residual error in the LP approximation.}
 \label{fig:LP_ex}
@@
 %
-In the parameter regime we consider, the ODE approximation of this scheme exhibits a separation of time scales, so that initial conditions quickly approach a two-dimensional manifold.
+We consider a parameter regime in which the ODE approximation of this scheme exhibits a separation of time scales, so that initial conditions quickly approach a two-dimensional manifold.
 %
@@
 %
-We can control the level of noise in our SSA by adjusting the volume $V$, and therefore adjusting the number of molecules, in the system (in our simulations, we take $V=10^5$).
+We can control the level of noise in our simulation by adjusting the volume $V$, and therefore, adjusting the number of molecules, in the system.
 %
-We take the volume small enough so that we can still observe appreciable stochasticity in small simulation bursts, but large enough so that the underlying two-dimensional manifold is (relatively) smooth.
+We take the volume small enough so that we can still observe appreciable stochasticity in small simulation bursts, but large enough (in our simulations, we take $V=10^5$) so that the underlying two-dimensional manifold is (relatively) smooth.
 
@@
 %
-We consider $\mathcal{Y} = \{ \mathbf{Y}(t): t=1, \dots, 3000 \}$ to be representative points ``on" this apparent two-dimensional manifold;
-three-dimensional projections of the resulting manifold are shown in Figure \ref{fig:rxn_manifolds}.
+We consider $\mathcal{Y} = \{ \mathbf{Y}(t): t=1, \dots, 3000 \}$ to be representative points ``on" this apparent two-dimensional manifold.
 %
-From each manifold point $y \in \mathcal{Y}$, we run 20 short simulation ``bursts'', each for 0.2 time units.
+From each manifold point $\mathbf{y} \in \mathcal{Y}$, we run 20 short simulation ``bursts'', each for 0.2 time units.
 %
-We denote the endpoints from the short simulations as $\mathcal{Y}^{burst}(y)$.
+We denote the endpoints from the short simulations as $\mathcal{Y}^{burst}(\mathbf{y})$.
 
 \begin{figure*}[ht]
-  \includegraphics[height=6cm]{rxn_figures/rxn_manifold1}\llap{\raisebox{3.5cm}{\includegraphics[height=2.5cm]{rxn_figures/rxn_manifold1_2}}}
-  \includegraphics[height=6cm]{rxn_figures/rxn_manifold2}\llap{\raisebox{3.5cm}{\includegraphics[height=2.5cm]{rxn_figures/rxn_manifold2_2}}} \\
-  \includegraphics[height=6cm]{rxn_figures/rxn_manifold3}\llap{\raisebox{3.5cm}{\includegraphics[height=2.5cm]{rxn_figures/rxn_manifold3_2}}}
-  \includegraphics[height=6cm]{rxn_figures/rxn_manifold4}\llap{\raisebox{3.5cm}{\includegraphics[height=2.5cm]{rxn_figures/rxn_manifold4_2}}} \\
+  \includegraphics[width=6in]{fig3}
     \caption{Projections of the data obtained from stochastic simulation of the chemical reaction network described in Section \ref{subsec:rxn_network}.
@@
 %
-The endpoints of the simulation bursts for the two data sets, $\mathcal{Y}^{burst}_1(y)$ and $\mathcal{Y}^{burst}_2(y)$, are defined analogously.
+The endpoints of the simulation bursts for the two data sets, $\mathcal{Y}^{burst}_1(\mathbf{y})$ and $\mathcal{Y}^{burst}_2(\mathbf{y})$, are defined analogously.
 %
@@
 \begin{equation}
-\widehat{\mathbf{C}_i}(y) = \sum_{z \in \mathcal{Y}^{burst}_i(y)} \left( z - \hat{\mathbf{\mu}}_i(y) \right)\left( z - \hat{\mathbf{\mu}}_i(y) \right)^T, y \in \mathcal{Y}_i, i \in \{1, 2\}
+\widehat{\mathbf{C}}_i(\mathbf{y}) = \sum_{\mathbf{z} \in \mathcal{Y}^{burst}_i(\mathbf{y})} \left( \mathbf{z} - \hat{\mathbf{\mu}}_i(\mathbf{y}) \right)\left( \mathbf{z} - \hat{\mathbf{\mu}}_i(\mathbf{y}) \right)^T, \mathbf{y} \in \mathcal{Y}_i, i \in \{1, 2\}
 \end{equation}
-where $\hat{\mathbf{\mu}}_i(y)$ is the empirical mean of $\mathcal{Y}^{burst}_i(y)$.
+where $\hat{\mathbf{\mu}}_i(\mathbf{y})$ is the empirical mean of $\mathcal{Y}^{burst}_i(\mathbf{y})$.
 
@@
 \begin{figure*}[ht]
-    \includegraphics[width=0.4\textwidth]{rxn_figures/rxn_NLICA1}
-    \includegraphics[width=0.4\textwidth]{rxn_figures/rxn_NLICA2}
-    \includegraphics[width=0.4\textwidth]{rxn_figures/rxn_NLICA_corr1}
-    \includegraphics[width=0.4\textwidth]{rxn_figures/rxn_NLICA_corr2}
-    \caption{(Top, left) NIV embedding obtained from $\mathcal{Y}_1$ (observations of components E, S, S:E, and F:E$^{*}$), colored by $S$.
-(Top, right) NIV embedding obtained from $\mathcal{Y}_2$ (observations of components E, S, E:S, and D:S$^{*}$), colored by $S$.
+    \includegraphics[width=6in]{fig4}
+    \caption{(a) NIV embedding obtained from $\mathcal{Y}_1$ (observations of components E, S, S:E, and F:E$^{*}$), colored by $S$.
+(b) NIV embedding obtained from $\mathcal{Y}_2$ (observations of components E, S, E:S, and D:S$^{*}$), colored by $S$.
 Visually, we can see that the embeddings obtained from $\mathcal{Y}_1$ and $\mathcal{Y}_2$ are consistent, even though the two data sets consist of observations of different chemical species.
-(Bottom, left) Correlation of first NIV between two different embeddings (correlation=0.97).
-(Bottom, right)  Correlation of second NIV between two different embeddings (correlation=0.95).
+(c) Correlation of first NIV between two different embeddings (correlation=0.97).
+(d)  Correlation of second NIV between two different embeddings (correlation=0.95).
 We obtain a good quantitative agreement between the embedding coordinates for the two data sets.}
@@
 %
-Instead, we must first project the data into the NIV space so that we can compute neighbors {\em between the two data sets}.
+Instead, we must first embed the data into the NIV space so that we can compute neighbors {\em between the two data sets}.
 %
@@
 \begin{figure*}[ht]
-    \includegraphics[width=0.4\textwidth]{rxn_figures/rxn_recon4}
-    \includegraphics[width=0.4\textwidth]{rxn_figures/rxn_recon6}
-    \caption{LP reconstructions of $S:E$ (left) and $F:E^{*}$ (right) for $\mathcal{Y}_2$, using $\mathcal{Y}_1$ as training data.}
+    \includegraphics[width=6in]{fig5}
+    \caption{LP reconstructions of (a) $S:E$ and (b) $F:E^{*}$ for $\mathcal{Y}_2$, using $\mathcal{Y}_1$ as training data.}
     \label{fig:rxn_recon}
@@
 \begin{figure*}[ht]
-    \includegraphics[width=0.6\textwidth]{FES}
-    \includegraphics[width=0.3\textwidth]{molecule_labeled2}
-    \caption{(Left) Free energy surface for Ala2. The relevant minima are labeled A, B, and C, and the corresponding molecular configurations are shown.
-    (Right) Sample representative molecular structure of Ala2, excluding the hydrogens. The atoms are numbered and the two dihedral angles $\phi$ and $\psi$ are indicated.}
+    \includegraphics[width=6in]{fig6}
+    \caption{(a) Free energy surface for Ala2. The relevant minima are labeled A, B, and C, and the corresponding molecular configurations are shown.
+    (b) Sample representative molecular structure of Ala2, excluding the hydrogens. The atoms are numbered and the two dihedral angles $\phi$ and $\psi$ are indicated.}
     \label{fig:ala_fes}
@@
 %
-We then compute the NIV and diffusion maps embeddings for $\mathcal{Y}_{even}$ and $\mathcal{Y}_{odd}$.
-%
-For NIV, we compute the covariances as in \eqref{eq:cov}, with $L=10$.
+We then compute the NIV and diffusion maps embeddings for $\mathcal{Y}_{even}$ and $\mathcal{Y}_{odd}$;
+for NIV, we compute the covariances as in \eqref{eq:cov}, with $L=10$.
 %at the ends of each simulation trajectory, we take fewer points before or following the point of interest to estimate the covariance
@@
 
-The correlation between the NIV coordinates for the two data sets and the diffusion map coordinates for the two data sets are shown in Figure \ref{fig:ala_corr}.
+The correlation between the NIV coordinates for the two data sets and the diffusion map (DM) coordinates for the two data sets are shown in Figure \ref{fig:ala_corr}.
 %
 The correlation between the two NIV embeddings is higher than the correlation between the two diffusion map embeddings.
-Therefore, it appears advantageous to use NIV over diffusion maps if one wishes to obtain a consistent embedding, and merge data sets from different observation domains (as long as the two main assumptions underpinning the NIV algorithm hold).
+Therefore, it appears advantageous to use NIV over diffusion maps if one wishes to obtain a consistent embedding and merge data sets from different observation domains (as long as the two main assumptions underpinning the NIV algorithm hold).
 
 \begin{figure*}[ht]
-    \includegraphics[width=0.4\textwidth]{ala_figures_evenodd/NLICA_corr}
-    \includegraphics[width=0.4\textwidth]{ala_figures_evenodd/DM_corr}
-    \caption{Correlation between the second NIV (left) and the second DM (right) computed using the atoms 2, 4, 6, 8, and 10 ($\psi_2^{even}, DM_2^{even}$) , and using atoms 5, 7, and 9 ($\psi_2^{odd}, DM_2^{odd}$).
+    \includegraphics[width=6in]{fig7}
+    \caption{(a) Correlation between the second NIV computed using the atoms 2, 4, 6, 8, and 10 ($\psi_2^{even}$) and the second NIV computed using atoms 5, 7, and 9 ($\psi_2^{odd}$). 
+    (b) Correlation between the second DM computed using the atoms 2, 4, 6, 8, and 10 ($DM_2^{even}$) and the second DM computed using atoms 5, 7, and 9 ($DM_2^{odd}$).
     The correlations for the first (not shown) and second NIV coordinates are found to be 0.62 and 0.84, respectively.
@@
 \begin{figure*}[ht]
-    \includegraphics[width=0.3\textwidth]{ala_figures_allodd/ala2_embed1}
-    \includegraphics[width=0.3\textwidth]{ala_figures_allodd/ala2_embed2}
-    \includegraphics[width=0.3\textwidth]{ala_figures_allodd/ala2_embed3}
-    \caption{The 3-dimensional NIV embedding for Ala2 computed using $\mathcal{Y}_{all}$, colored by the y-coordinate of the first atom (left), the dihedral angle $\phi$ (center), and the dihedral angle $\psi$ (right).
+    \includegraphics[width=6in]{fig8}
+    \caption{The 3-dimensional NIV embedding for Ala2 computed using $\mathcal{Y}_{all}$, colored by (a) the y-coordinate of the first atom, (b) the dihedral angle $\phi$, and (c) the dihedral angle $\psi$.
 Each embedding is rotated so that the correlation between the colors and the relevant NIV can easily be seen.}
@@
   \centering
-    \foreach \atomnumber in {1,2,3,6,8} {
-        \includegraphics[width=0.2\textwidth]{ala_figures_allodd/ala2_recon_corr_\atomnumber_1}
-        \includegraphics[width=0.2\textwidth]{ala_figures_allodd/ala2_recon_corr_\atomnumber_2}
-        \includegraphics[width=0.2\textwidth]{ala_figures_allodd/ala2_recon_corr_\atomnumber_3}\\
-      }
+        \includegraphics[width=4.5in]{fig9}
   \caption{The correlation between the true position and the reconstructed position (using LP) for the test data.
@@
     \centering
-    \includegraphics[width=0.3\textwidth,trim=0in 0.7in 0in 0in, clip]{molec_figures/molecule300_balls}
-    \includegraphics[width=0.3\textwidth,trim=0in 0.7in 0in 0in, clip]{molec_figures/molecule300_wire}\\
-    \includegraphics[width=0.3\textwidth,trim=0in 0.7in 0in 0in, clip]{molec_figures/molecule600_balls}
-    \includegraphics[width=0.3\textwidth,trim=0in 0.7in 0in 0in, clip]{molec_figures/molecule600_wire}\\
-    \includegraphics[width=0.3\textwidth,trim=0in 0.7in 0in 0in, clip]{molec_figures/molecule900_balls}
-    \includegraphics[width=0.3\textwidth,trim=0in 0.7in 0in 0in, clip]{molec_figures/molecule900_wire}
+    \includegraphics[width=4in]{fig10}
     \caption{True structure (black) and reconstructed structure (red) for three different data points.
@@
 \begin{figure*}
-    \includegraphics[width=0.5\textwidth]{ala_figures_allodd/ala2_recon_mse}
+    \includegraphics[width=3in]{fig11}
     \caption{Mean squared error of reconstructed position, normalized by the average bond length, for each atom in Ala2.
@@
 %
-In both examples, we were able to uncover the intrinsic variables governing the underlying stochastic process, independent
+In both examples, we were able to uncover the intrinsic variables governing the underlying stochastic process, which are independent
 of the particular measurement or observation  of the system (under the conditions mentioned).
@@
 Merging data sets that come from different portions of the manifold would not only require scaling the embedding coordinates,
-but also shifting and possibly permuting the embedding coordinates (in this spirit, see the discussion in Reference~\onlinecite{lafon2006data}).
+but also shifting and possibly permuting the embedding coordinates (in this spirit, see the discussion in Lafon {\em et al.} \cite{lafon2006data}).
 %
@@
 %
-Other issues, such as accurately estimating the covariance matrices required to compute the Mahalanobis distance, are also of current research interest.
+Other issues, such as accurately estimating the covariance matrices required for the computation of the Mahalanobis distance, are also of current research interest.
 %
@@
 
-We consider the following network of chemical reactions:
+We consider the following network of chemical reactions.
 \begin{equation}
@@
 \begin{widetext}
-In the limit of a large number of molecules, the dynamics of this network is governed by the following ODEs:
+In the limit of a large number of molecules, the dynamics of this network is governed by the following ODEs.
 \begin{equation}
@@
 %
-We therefore obtain a system of 6 ODEs:
+We therefore obtain a system of 6 ODEs.
 \begin{equation}

### 2013-12-27 paper/figures1.tex
Added files related to paper

\documentclass[12pt]{article}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{float}
\usepackage{forloop}
\usepackage{tikz}
\usepackage[bottom=1in,top=1in,left=1in,right=1in]{geometry}

\begin{document}

{\bf Keywords:} Intrinsic variables, partial observations, complex dynamical systems

\section{Introduction}

\begin{itemize}
    \item We know that manifold learning can lead to reduction in reducible complex dynamical systems

    \item We know from recent Ronen that if we do IV we can ``merge'' data sets

    \item We know from recent Neta that LapPyr (especially with new adaptive twists) can be used to interpolate function on sampled data and, in, particular, or reduced representations of sample data

    \item Here, we synthesize these three ideas to show that if we have partial measurements for ``complex but effectively low-dimensional'' systems, we can reconstruct/estimate the missing measurements based on a training set.

    \item This is important in both open loop sensing and in closed control loops based on the measurements of unavailable variables

    \item Two examples that arise in typical physicochemical computations, one from chemical reactions and one from MD

    \item Both examples are NOT SDEs but it is established/accepted that they can be effectively modeled as SDEs.

    \item Sometimes we can predict variables and sometimes we can predict statistics of variables --- maybe best to bring it up WITHIN the examples
\end{itemize}

\section{Methods}

\subsection{Intrinsic Modeling}
\begin{figure}[H]
\includegraphics[width=\textwidth]{IntrinsicEmbeddingIllustration2.pdf}
\caption{Illustration of the nonlinear embedding that yields an intrinsic representation independent of measurement modality. (Left) The underlying variables in which the diffusion parameters are independent with unit variance. The depicted points illustrate the end points of the short bursts that create a sphere around the initial point on the manifold. (Right Top) The first set of observed variables. The depicted points illustrate the end points of the short bursts as observed via the first observation modality. (Right Bottom) The second set of observed variables. The depicted points illustrate the end points of the short bursts as observed via the second observation modality.}
\end{figure}

{\bf To do:} \\
Block diagram/algorithm

\subsection{Laplacian Pyramids}

{\bf To do:} \\
A flow chart/block diagram of the method (possibly with the formulas).\\
Maybe an illustration of the multiscale reconstruction of functions.

\begin{figure}[H]
\includegraphics[width=\textwidth]{LapPyr_illustration.pdf}
\caption{Illustration of overall methodology}
\end{figure}

\section{Models and Results}

\subsection{Chemical Reaction Network}

\begin{figure}[H]
  \includegraphics[height=6cm]{rxn_figures/rxn_manifold1.jpg}\llap{\raisebox{3.5cm}{\includegraphics[height=2.5cm]{rxn_figures/rxn_manifold1_2.jpg}}}
  \includegraphics[height=6cm]{rxn_figures/rxn_manifold2.jpg}\llap{\raisebox{3.5cm}{\includegraphics[height=2.5cm]{rxn_figures/rxn_manifold2_2.jpg}}} \\
  \includegraphics[height=6cm]{rxn_figures/rxn_manifold3.jpg}\llap{\raisebox{3.5cm}{\includegraphics[height=2.5cm]{rxn_figures/rxn_manifold3_2.jpg}}}
  \includegraphics[height=6cm]{rxn_figures/rxn_manifold4.jpg}\llap{\raisebox{3.5cm}{\includegraphics[height=2.5cm]{rxn_figures/rxn_manifold4_2.jpg}}} \\
%    \includegraphics[width=0.5\textwidth]{rxn_figures/rxn_manifold1.jpg}
%    \includegraphics[width=0.5\textwidth]{rxn_figures/rxn_manifold2.jpg}\\
%    \includegraphics[width=0.5\textwidth]{rxn_figures/rxn_manifold3.jpg}
%    \includegraphics[width=0.5\textwidth]{rxn_figures/rxn_manifold4.jpg}
    \caption{Projections of manifold onto different chemical components. The insets show a different projection to illustrate the two-dimensionality of the manifold.}
\end{figure}

\begin{figure}[H]
    \includegraphics[width=0.5\textwidth]{rxn_figures/rxn_NLICA1.jpg}
    \includegraphics[width=0.5\textwidth]{rxn_figures/rxn_NLICA2.jpg}
    \includegraphics[width=0.5\textwidth]{rxn_figures/rxn_NLICA_corr1.jpg}
    \includegraphics[width=0.5\textwidth]{rxn_figures/rxn_NLICA_corr2.jpg}
    \caption{(Top) IV embedding obtained from observations of components 1, 2, 4, and 6 (left) embedding obtained from observations of components 1, 2, 3, and 5 (right), colored by component 1 (left) (Bottom) Correlation of first IV (left, correlation=0.97) and second IV (right, correlation=0.95) between two different embeddings.}
\end{figure}

\begin{figure}[H]
    \includegraphics[width=0.5\textwidth]{rxn_figures/chm4_from_Em.png}
    \includegraphics[width=0.5\textwidth]{rxn_figures/chm6_from_Em.png}
    \caption{Laplacian Pyramids (LapPyr) and nearest neighbor (nn) reconstructions of species 4 (left) and 6 (right). The nearest neighbors are computed both in the embedded space and in the ambient space, and in the embedded space.}
\end{figure}

The nearest neighbor reconstruction using neighbors in the ambient space is more accurate than the other two methods, implying that some of the error arises from inaccuracies in the embeddings.
%
However, we would like to note that computing nearest neighbors in the ambient space requires knowing how to map full observations to partial observations.
%
Therefore, in cases where we have partial {\em unknown} observations (i.e. we observe some chemicals, but we do not know which ones), we can still use the embedded space (intrinsic variables) to predict the remaining species.

\subsection{Alanine Dipeptide}

\begin{itemize}
  \item 10000 data points
  \item The NLICA embeddings were computed using odd and even atoms (atoms 1 and 3 were removed)
  \item Laplacian Pyramids were then used to predict the atom positions for the test data using the IV representation
\end{itemize}

\begin{figure}[H]
    \includegraphics[width=0.6\textwidth]{FES.pdf}
    \includegraphics[width=0.4\textwidth]{molecule_labeled2.jpg}
    \caption{(Left) Free energy surface for Alanine Dipeptide. The minima are labeled A, B, and C, and the corresponding molecular configurations are shown. (Right) The molecular structure of Alanine Dipeptide. The atoms are numbered and the two dihedral angles $\phi$ and $\psi$ are indicated.}
\end{figure}

The motion of Ala2 can be well described by three IV.
%
The first IV describes the flipping of the first and third atoms, the second IV describes the dihedral angle $\phi$, and the third IV describes the dihedral angle $\psi$.

\begin{figure}[H]
    \includegraphics[width=0.5\textwidth]{ala_figures_evenodd/NLICA_corr.jpg}
    \includegraphics[width=0.5\textwidth]{ala_figures_evenodd/DM_corr.jpg}
    \caption{(Left) Correlation between the second IV computed using the odd atoms and the even atoms. (Right) Correlation between the second DM computed using the odd atoms and the even atoms.}
\end{figure}

\begin{table}
\begin{tabular}{| r || c | c |}
  \hline
  & Coordinate 1 & Coordinate 2 \\
  \hline
  \hline
  NLICA & 0.62 & 0.84 \\
  \hline
  DM & 0.54 & 0.60 \\
  \hline
\end{tabular}
\caption{Correlation between the embedding coordates from NLICA and DMAPS using the odd and even atoms.}
\end{table}

\begin{table}
\begin{tabular}{| r || c | c | c |}
  \hline
  & Coordinate 1 & Coordinate 2 & Coordinate 3 \\
  \hline
  \hline
  NLICA & 0.97 & 0.72 & 0.85 \\
  \hline
  DM & 0.95 & 0.76 & 0.74 \\
  \hline
\end{tabular}
\caption{Correlation between the embedding coordinates from NLICA and DMAPS using all and just odd coordinates.}
\end{table}

\begin{figure}[H]
    \includegraphics[width=0.3\textwidth]{ala_figures_allodd/ala2_embed1.jpg}
    \includegraphics[width=0.3\textwidth]{ala_figures_allodd/ala2_embed2.jpg}
    \includegraphics[width=0.3\textwidth]{ala_figures_allodd/ala2_embed3.jpg}
    \caption{The 3-dimensional embedding for Alanine Dipeptide using all the atoms, colored by the y-coordinate of the first atom (left), the dihedral angle $\phi$ (center), and the dihedral angle $\psi$ (right).}
\end{figure}

\begin{figure}[H]
  \centering
%  \newcounter{atomnumber}
%  \forloop{atomnumber}{1}{\value{atomnumber} < 11}{
%    \includegraphics[width=0.15\textwidth]{ala_figures_allodd/ala2_recon_corr_\arabic{atomnumber}_1.jpg}
%    \includegraphics[width=0.15\textwidth]{ala_figures_allodd/ala2_recon_corr_\arabic{atomnumber}_2.jpg}
%    \includegraphics[width=0.15\textwidth]{ala_figures_allodd/ala2_recon_corr_\arabic{atomnumber}_3.jpg}\\
%  }
    \foreach \atomnumber in {1,2,3,6,8} {
        \includegraphics[width=0.2\textwidth]{ala_figures_allodd/ala2_recon_corr_\atomnumber_1.jpg}
        \includegraphics[width=0.2\textwidth]{ala_figures_allodd/ala2_recon_corr_\atomnumber_2.jpg}
        \includegraphics[width=0.2\textwidth]{ala_figures_allodd/ala2_recon_corr_\atomnumber_3.jpg}\\
      }
  \caption{The correlation between the true position and the reconstructed position for atoms 1, 2, 3, 6, and 8 (using Laplacian Pyramids) for each coordinate for the NLICA embedding of the odd data, using the NLICA embedding from all coordinates as training data.}
\end{figure}

\begin{figure}[H]
    \includegraphics[width=0.3\textwidth,trim=0.5in 3in 4in 0.5in, clip]{molec_figures/molecule300.jpg}
    \includegraphics[width=0.3\textwidth,trim=0.5in 3in 4in 0.5in, clip]{molec_figures/molecule600.jpg}
    \includegraphics[width=0.3\textwidth,trim=0.5in 3in 4in 0.5in, clip]{molec_figures/molecule900.jpg}    
    \caption{True structure (black) and reconstructed structure (red) for three different data points}
\end{figure}

\section{Conclusions}

\end{document}

### 2013-12-27 paper/jcp_paper.tex
Added files related to paper

--- 2013-12-27 paper/draft9.tex
@@
 %
-We first scale the eigenvectors so that $\|\psi_i\| = T$, where $T$ is the number of data points;
-intuitively, the norm of the embedding eigenvectors should scale with the size of the data set to make the embedding coordinates of each data point independent of the size of the data set.
+We first scale the eigenvectors so that $\|\psi_i\| = T$, where $T$ is the number of data points,
+to make the embedding coordinates invariant to the size of the data set.
 %twice as many samples as another data set, but with the same underlying geometry, should have twice the norm.
@@
 For example, if the mean of the embedding is sufficiently far from 0, we can require $\langle \psi_i \rangle > 0$;
-%the signs can be chosen such that the mean of each eigenvector has a prescribed sign;
 alternatively, if there is a common region sampled by both data sets, the sign of each eigenvector can be chosen to optimize the consistency of the embeddings of the common region data.
@@
 
-\bibliography{../../references/references}
+%\bibliography{../../references/references}
+%merlin.mbs aipnum4-1.bst 2010-07-25 4.21a (PWD, AO, DPC) hacked
+%Control: key (0)
+%Control: author (8) initials jnrlst
+%Control: editor formatted (1) identically to author
+%Control: production of article title (-1) disabled
+%Control: page (0) single
+%Control: year (1) truncated
+%Control: production of eprint (0) enabled
+\begin{thebibliography}{39}%
+\makeatletter
+\providecommand \@ifxundefined [1]{%
+ \@ifx{#1\undefined}
+}%
+\providecommand \@ifnum [1]{%
+ \ifnum #1\expandafter \@firstoftwo
+ \else \expandafter \@secondoftwo
+ \fi
+}%
+\providecommand \@ifx [1]{%
+ \ifx #1\expandafter \@firstoftwo
+ \else \expandafter \@secondoftwo
+ \fi
+}%
+\providecommand \natexlab [1]{#1}%
+\providecommand \enquote  [1]{``#1''}%
+\providecommand \bibnamefont  [1]{#1}%
+\providecommand \bibfnamefont [1]{#1}%
+\providecommand \citenamefont [1]{#1}%
+\providecommand \href@noop [0]{\@secondoftwo}%
+\providecommand \href [0]{\begingroup \@sanitize@url \@href}%
+\providecommand \@href[1]{\@@startlink{#1}\@@href}%
+\providecommand \@@href[1]{\endgroup#1\@@endlink}%
+\providecommand \@sanitize@url [0]{\catcode `\\12\catcode `\$12\catcode
+  `\&12\catcode `\#12\catcode `\^12\catcode `\_12\catcode `\%12\relax}%
+\providecommand \@@startlink[1]{}%
+\providecommand \@@endlink[0]{}%
+\providecommand \url  [0]{\begingroup\@sanitize@url \@url }%
+\providecommand \@url [1]{\endgroup\@href {#1}{\urlprefix }}%
+\providecommand \urlprefix  [0]{URL }%
+\providecommand \Eprint [0]{\href }%
+\providecommand \doibase [0]{http://dx.doi.org/}%
+\providecommand \selectlanguage [0]{\@gobble}%
+\providecommand \bibinfo  [0]{\@secondoftwo}%
+\providecommand \bibfield  [0]{\@secondoftwo}%
+\providecommand \translation [1]{[#1]}%
+\providecommand \BibitemOpen [0]{}%
+\providecommand \bibitemStop [0]{}%
+\providecommand \bibitemNoStop [0]{.\EOS\space}%
+\providecommand \EOS [0]{\spacefactor3000\relax}%
+\providecommand \BibitemShut  [1]{\csname bibitem#1\endcsname}%
+\let\auto@bib@innerbib\@empty
+%</preamble>
+\bibitem [{\citenamefont {Tenenbaum}, \citenamefont {de~Silva},\ and\
+  \citenamefont {Langford}(2000)}]{Tenenbaum2000}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {J.~B.}\ \bibnamefont
+  {Tenenbaum}}, \bibinfo {author} {\bibfnamefont {V.}~\bibnamefont {de~Silva}},
+  \ and\ \bibinfo {author} {\bibfnamefont {J.~C.}\ \bibnamefont {Langford}},\
+  }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {Science}\ }\textbf
+  {\bibinfo {volume} {260}},\ \bibinfo {pages} {2319} (\bibinfo {year}
+  {2000})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Roweis}\ and\ \citenamefont
+  {Saul}(2000)}]{Roweis2000}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {S.~T.}\ \bibnamefont
+  {Roweis}}\ and\ \bibinfo {author} {\bibfnamefont {L.~K.}\ \bibnamefont
+  {Saul}},\ }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal}
+  {Science}\ }\textbf {\bibinfo {volume} {260}},\ \bibinfo {pages} {2323}
+  (\bibinfo {year} {2000})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Donoho}\ and\ \citenamefont
+  {Grimes}(2003)}]{Donoho2003}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {D.~L.}\ \bibnamefont
+  {Donoho}}\ and\ \bibinfo {author} {\bibfnamefont {C.}~\bibnamefont
+  {Grimes}},\ }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {PNAS}\
+  }\textbf {\bibinfo {volume} {100}},\ \bibinfo {pages} {5591} (\bibinfo {year}
+  {2003})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Belkin}\ and\ \citenamefont
+  {Niyogi}(2003)}]{Belkin2003}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {M.}~\bibnamefont
+  {Belkin}}\ and\ \bibinfo {author} {\bibfnamefont {P.}~\bibnamefont
+  {Niyogi}},\ }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {Neural
+  Comput.}\ }\textbf {\bibinfo {volume} {15}},\ \bibinfo {pages} {1373}
+  (\bibinfo {year} {2003})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Coifman}\ and\ \citenamefont
+  {Lafon}(2006{\natexlab{a}})}]{Coifman2006}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {R.}~\bibnamefont
+  {Coifman}}\ and\ \bibinfo {author} {\bibfnamefont {S.}~\bibnamefont
+  {Lafon}},\ }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {Appl.
+  Comput. Harmon. Anal.}\ }\textbf {\bibinfo {volume} {21}},\ \bibinfo {pages}
+  {5} (\bibinfo {year} {2006}{\natexlab{a}})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Jolliffe}(2005)}]{jolliffe2005principal}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {I.}~\bibnamefont
+  {Jolliffe}},\ }\href@noop {} {\emph {\bibinfo {title} {Principal Component
+  Analysis}}}\ (\bibinfo  {publisher} {Wiley Online Library},\ \bibinfo {year}
+  {2005})\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Lafon}, \citenamefont {Keller},\ and\ \citenamefont
+  {Coifman}(2006)}]{lafon2006data}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {S.}~\bibnamefont
+  {Lafon}}, \bibinfo {author} {\bibfnamefont {Y.}~\bibnamefont {Keller}}, \
+  and\ \bibinfo {author} {\bibfnamefont {R.~R.}\ \bibnamefont {Coifman}},\
+  }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {IEEE Transactions
+  on Pattern Analysis and Machine Intelligence}\ }\textbf {\bibinfo {volume}
+  {28}},\ \bibinfo {pages} {1784} (\bibinfo {year} {2006})}\BibitemShut
+  {NoStop}%
+\bibitem [{\citenamefont {Rabin}\ and\ \citenamefont
+  {Coifman}(2012)}]{rabin2012heterogeneous}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {N.}~\bibnamefont
+  {Rabin}}\ and\ \bibinfo {author} {\bibfnamefont {R.}~\bibnamefont
+  {Coifman}},\ }in\ \href@noop {} {\emph {\bibinfo {booktitle} {Proceedings of
+  the 12th SIAM International Conference on Data Mining (SDM 2012), Anaheim,
+  California, USA}}}\ (\bibinfo {year} {2012})\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Gillespie}(1977)}]{gillespie1977exact}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {D.~T.}\ \bibnamefont
+  {Gillespie}},\ }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {The
+  Journal of Physical Chemistry}\ }\textbf {\bibinfo {volume} {81}},\ \bibinfo
+  {pages} {2340} (\bibinfo {year} {1977})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Zagaris}\ \emph {et~al.}(2012)\citenamefont
+  {Zagaris}, \citenamefont {Vandekerckhove}, \citenamefont {Gear},
+  \citenamefont {Kaper},\ and\ \citenamefont
+  {Kevrekidis}}]{zagaris2012stability}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {A.}~\bibnamefont
+  {Zagaris}}, \bibinfo {author} {\bibfnamefont {C.}~\bibnamefont
+  {Vandekerckhove}}, \bibinfo {author} {\bibfnamefont {C.~W.}\ \bibnamefont
+  {Gear}}, \bibinfo {author} {\bibfnamefont {T.~J.}\ \bibnamefont {Kaper}}, \
+  and\ \bibinfo {author} {\bibfnamefont {I.~G.}\ \bibnamefont {Kevrekidis}},\
+  }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {Discrete and
+  Continuous Dynamical Systems-Series A}\ }\textbf {\bibinfo {volume} {32}},\
+  \bibinfo {pages} {2759} (\bibinfo {year} {2012})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Bolhuis}, \citenamefont {Dellago},\ and\
+  \citenamefont {Chandler}(2000)}]{bolhuis2000reaction}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {P.~G.}\ \bibnamefont
+  {Bolhuis}}, \bibinfo {author} {\bibfnamefont {C.}~\bibnamefont {Dellago}}, \
+  and\ \bibinfo {author} {\bibfnamefont {D.}~\bibnamefont {Chandler}},\
+  }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {Proceedings of the
+  National Academy of Sciences}\ }\textbf {\bibinfo {volume} {97}},\ \bibinfo
+  {pages} {5877} (\bibinfo {year} {2000})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Coifman}\ \emph {et~al.}(2005)\citenamefont
+  {Coifman}, \citenamefont {Lafon}, \citenamefont {Lee}, \citenamefont
+  {Maggioni}, \citenamefont {Nadler}, \citenamefont {Warner},\ and\
+  \citenamefont {Zucker}}]{coifman2005geometric}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {R.~R.}\ \bibnamefont
+  {Coifman}}, \bibinfo {author} {\bibfnamefont {S.}~\bibnamefont {Lafon}},
+  \bibinfo {author} {\bibfnamefont {A.~B.}\ \bibnamefont {Lee}}, \bibinfo
+  {author} {\bibfnamefont {M.}~\bibnamefont {Maggioni}}, \bibinfo {author}
+  {\bibfnamefont {B.}~\bibnamefont {Nadler}}, \bibinfo {author} {\bibfnamefont
+  {F.}~\bibnamefont {Warner}}, \ and\ \bibinfo {author} {\bibfnamefont {S.~W.}\
+  \bibnamefont {Zucker}},\ }\href@noop {} {\bibfield  {journal} {\bibinfo
+  {journal} {Proceedings of the National Academy of Sciences}\ }\textbf
+  {\bibinfo {volume} {102}},\ \bibinfo {pages} {7426} (\bibinfo {year}
+  {2005})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Singer}\ and\ \citenamefont
+  {Coifman}(2008)}]{singer2008non}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {A.}~\bibnamefont
+  {Singer}}\ and\ \bibinfo {author} {\bibfnamefont {R.~R.}\ \bibnamefont
+  {Coifman}},\ }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal}
+  {Applied and Computational Harmonic Analysis}\ }\textbf {\bibinfo {volume}
+  {25}},\ \bibinfo {pages} {226} (\bibinfo {year} {2008})}\BibitemShut
+  {NoStop}%
+\bibitem [{\citenamefont {Hein}\ and\ \citenamefont
+  {Audibert}(2005)}]{hein2005intrinsic}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {M.}~\bibnamefont
+  {Hein}}\ and\ \bibinfo {author} {\bibfnamefont {J.~Y.}\ \bibnamefont
+  {Audibert}},\ }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {L.
+  De Raedt, S. Wrobel (Eds.), Proc. 22nd Int. Conf. Mach. Learn., ACM}\ ,\
+  \bibinfo {pages} {289}} (\bibinfo {year} {2005})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Coifman}\ \emph {et~al.}(2008)\citenamefont
+  {Coifman}, \citenamefont {Shkolnisky}, \citenamefont {Sigworth},\ and\
+  \citenamefont {Singer}}]{coifman2008graph}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {R.~R.}\ \bibnamefont
+  {Coifman}}, \bibinfo {author} {\bibfnamefont {Y.}~\bibnamefont {Shkolnisky}},
+  \bibinfo {author} {\bibfnamefont {F.~J.}\ \bibnamefont {Sigworth}}, \ and\
+  \bibinfo {author} {\bibfnamefont {A.}~\bibnamefont {Singer}},\ }\href@noop {}
+  {\bibfield  {journal} {\bibinfo  {journal} {IEEE Transactions on Image
+  Processing}\ }\textbf {\bibinfo {volume} {17}},\ \bibinfo {pages} {1891}
+  (\bibinfo {year} {2008})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Rohrdanz}\ \emph {et~al.}(2011)\citenamefont
+  {Rohrdanz}, \citenamefont {Zheng}, \citenamefont {Maggioni},\ and\
+  \citenamefont {Clementi}}]{rohrdanz2011determination}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {M.~A.}\ \bibnamefont
+  {Rohrdanz}}, \bibinfo {author} {\bibfnamefont {W.}~\bibnamefont {Zheng}},
+  \bibinfo {author} {\bibfnamefont {M.}~\bibnamefont {Maggioni}}, \ and\
+  \bibinfo {author} {\bibfnamefont {C.}~\bibnamefont {Clementi}},\ }\href@noop
+  {} {\bibfield  {journal} {\bibinfo  {journal} {The Journal of Chemical
+  Physics}\ }\textbf {\bibinfo {volume} {134}},\ \bibinfo {pages} {124116}
+  (\bibinfo {year} {2011})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Chung}(1997)}]{chung1997spectral}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {F.~R.}\ \bibnamefont
+  {Chung}},\ }\href@noop {} {\emph {\bibinfo {title} {Spectral Graph
+  Theory}}},\ Vol.~\bibinfo {volume} {92}\ (\bibinfo  {publisher} {AMS
+  Bookstore},\ \bibinfo {year} {1997})\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Burt}\ and\ \citenamefont
+  {Adelson}(1983)}]{burt1983laplacian}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {P.}~\bibnamefont
+  {Burt}}\ and\ \bibinfo {author} {\bibfnamefont {E.}~\bibnamefont {Adelson}},\
+  }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {Communications,
+  IEEE Transactions on}\ }\textbf {\bibinfo {volume} {31}},\ \bibinfo {pages}
+  {532} (\bibinfo {year} {1983})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Do}\ and\ \citenamefont
+  {Vetterli}(2003)}]{do2003framing}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {M.~N.}\ \bibnamefont
+  {Do}}\ and\ \bibinfo {author} {\bibfnamefont {M.}~\bibnamefont {Vetterli}},\
+  }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {Signal Processing,
+  IEEE Transactions on}\ }\textbf {\bibinfo {volume} {51}},\ \bibinfo {pages}
+  {2329} (\bibinfo {year} {2003})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Buchman}, \citenamefont {Lee},\ and\ \citenamefont
+  {Schafer}(2011)}]{buchman2011high}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {S.~M.}\ \bibnamefont
+  {Buchman}}, \bibinfo {author} {\bibfnamefont {A.~B.}\ \bibnamefont {Lee}}, \
+  and\ \bibinfo {author} {\bibfnamefont {C.~M.}\ \bibnamefont {Schafer}},\
+  }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {Statistical
+  Methodology}\ }\textbf {\bibinfo {volume} {8}},\ \bibinfo {pages} {18}
+  (\bibinfo {year} {2011})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Gillespie}(2000)}]{gillespie2000chemical}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {D.~T.}\ \bibnamefont
+  {Gillespie}},\ }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {The
+  Journal of Chemical Physics}\ }\textbf {\bibinfo {volume} {113}},\ \bibinfo
+  {pages} {297} (\bibinfo {year} {2000})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Apostolakis}, \citenamefont {Ferrara},\ and\
+  \citenamefont {Caflisch}(1999)}]{apostolakis1999calculation}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {J.}~\bibnamefont
+  {Apostolakis}}, \bibinfo {author} {\bibfnamefont {P.}~\bibnamefont
+  {Ferrara}}, \ and\ \bibinfo {author} {\bibfnamefont {A.}~\bibnamefont
+  {Caflisch}},\ }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {The
+  Journal of Chemical Physics}\ }\textbf {\bibinfo {volume} {110}},\ \bibinfo
+  {pages} {2099} (\bibinfo {year} {1999})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Chekmarev}, \citenamefont {Ishida},\ and\
+  \citenamefont {Levy}(2004)}]{chekmarev2004long}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {D.~S.}\ \bibnamefont
+  {Chekmarev}}, \bibinfo {author} {\bibfnamefont {T.}~\bibnamefont {Ishida}}, \
+  and\ \bibinfo {author} {\bibfnamefont {R.~M.}\ \bibnamefont {Levy}},\
+  }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {The Journal of
+  Physical Chemistry B}\ }\textbf {\bibinfo {volume} {108}},\ \bibinfo {pages}
+  {19487} (\bibinfo {year} {2004})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Ma}\ and\ \citenamefont
+  {Dinner}(2005)}]{ma2005automatic}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {A.}~\bibnamefont
+  {Ma}}\ and\ \bibinfo {author} {\bibfnamefont {A.~R.}\ \bibnamefont
+  {Dinner}},\ }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {The
+  Journal of Physical Chemistry B}\ }\textbf {\bibinfo {volume} {109}},\
+  \bibinfo {pages} {6769} (\bibinfo {year} {2005})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Frewen}, \citenamefont {Hummer},\ and\ \citenamefont
+  {Kevrekidis}(2009)}]{frewen2009exploration}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {T.~A.}\ \bibnamefont
+  {Frewen}}, \bibinfo {author} {\bibfnamefont {G.}~\bibnamefont {Hummer}}, \
+  and\ \bibinfo {author} {\bibfnamefont {I.~G.}\ \bibnamefont {Kevrekidis}},\
+  }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {The Journal of
+  Chemical Physics}\ }\textbf {\bibinfo {volume} {131}},\ \bibinfo {pages}
+  {134104} (\bibinfo {year} {2009})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Ferguson}\ \emph {et~al.}(2011)\citenamefont
+  {Ferguson}, \citenamefont {Panagiotopoulos}, \citenamefont {Debenedetti},\
+  and\ \citenamefont {Kevrekidis}}]{ferguson2011integrating}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {A.~L.}\ \bibnamefont
+  {Ferguson}}, \bibinfo {author} {\bibfnamefont {A.~Z.}\ \bibnamefont
+  {Panagiotopoulos}}, \bibinfo {author} {\bibfnamefont {P.~G.}\ \bibnamefont
+  {Debenedetti}}, \ and\ \bibinfo {author} {\bibfnamefont {I.~G.}\ \bibnamefont
+  {Kevrekidis}},\ }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal}
+  {The Journal of Chemical Physics}\ }\textbf {\bibinfo {volume} {134}},\
+  \bibinfo {pages} {135103} (\bibinfo {year} {2011})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Case}\ \emph {et~al.}(2008)\citenamefont {Case},
+  \citenamefont {Darden}, \citenamefont {T.~E.~Cheatham}, \citenamefont
+  {Simmerling}, \citenamefont {Wang}, \citenamefont {Duke}, \citenamefont
+  {Luo}, \citenamefont {Crowley}, \citenamefont {Walker}, \citenamefont
+  {Zhang}, \citenamefont {Merz}, \citenamefont {Wang}, \citenamefont {Hayik},
+  \citenamefont {Roitberg}, \citenamefont {Seabra}, \citenamefont {Kolossváry},
+  \citenamefont {Wong}, \citenamefont {Paesani}, \citenamefont {Vanicek},
+  \citenamefont {Wu}, \citenamefont {Brozell}, \citenamefont {Steinbrecher},
+  \citenamefont {Gohlke}, \citenamefont {Yang}, \citenamefont {Tan},
+  \citenamefont {Mongan}, \citenamefont {Hornak}, \citenamefont {Cui},
+  \citenamefont {Mathews}, \citenamefont {Seetin}, \citenamefont {Sagui},
+  \citenamefont {Babin},\ and\ \citenamefont {Kollman}}]{case2008Amber}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {D.~A.}\ \bibnamefont
+  {Case}}, \bibinfo {author} {\bibfnamefont {T.~A.}\ \bibnamefont {Darden}},
+  \bibinfo {author} {\bibfnamefont {I.}~\bibnamefont {T.~E.~Cheatham}},
+  \bibinfo {author} {\bibfnamefont {C.~L.}\ \bibnamefont {Simmerling}},
+  \bibinfo {author} {\bibfnamefont {J.}~\bibnamefont {Wang}}, \bibinfo {author}
+  {\bibfnamefont {R.~E.}\ \bibnamefont {Duke}}, \bibinfo {author}
+  {\bibfnamefont {R.}~\bibnamefont {Luo}}, \bibinfo {author} {\bibfnamefont
+  {M.}~\bibnamefont {Crowley}}, \bibinfo {author} {\bibfnamefont {R.~C.}\
+  \bibnamefont {Walker}}, \bibinfo {author} {\bibfnamefont {W.}~\bibnamefont
+  {Zhang}}, \bibinfo {author} {\bibfnamefont {K.~M.}\ \bibnamefont {Merz}},
+  \bibinfo {author} {\bibfnamefont {B.}~\bibnamefont {Wang}}, \bibinfo {author}
+  {\bibfnamefont {S.}~\bibnamefont {Hayik}}, \bibinfo {author} {\bibfnamefont
+  {A.}~\bibnamefont {Roitberg}}, \bibinfo {author} {\bibfnamefont
+  {G.}~\bibnamefont {Seabra}}, \bibinfo {author} {\bibfnamefont
+  {I.}~\bibnamefont {Kolossváry}}, \bibinfo {author} {\bibfnamefont {K.~F.}\
+  \bibnamefont {Wong}}, \bibinfo {author} {\bibfnamefont {F.}~\bibnamefont
+  {Paesani}}, \bibinfo {author} {\bibfnamefont {J.}~\bibnamefont {Vanicek}},
+  \bibinfo {author} {\bibfnamefont {X.}~\bibnamefont {Wu}}, \bibinfo {author}
+  {\bibfnamefont {S.~R.}\ \bibnamefont {Brozell}}, \bibinfo {author}
+  {\bibfnamefont {T.}~\bibnamefont {Steinbrecher}}, \bibinfo {author}
+  {\bibfnamefont {H.}~\bibnamefont {Gohlke}}, \bibinfo {author} {\bibfnamefont
+  {L.}~\bibnamefont {Yang}}, \bibinfo {author} {\bibfnamefont {C.}~\bibnamefont
+  {Tan}}, \bibinfo {author} {\bibfnamefont {J.}~\bibnamefont {Mongan}},
+  \bibinfo {author} {\bibfnamefont {V.}~\bibnamefont {Hornak}}, \bibinfo
+  {author} {\bibfnamefont {G.}~\bibnamefont {Cui}}, \bibinfo {author}
+  {\bibfnamefont {D.~H.}\ \bibnamefont {Mathews}}, \bibinfo {author}
+  {\bibfnamefont {M.~G.}\ \bibnamefont {Seetin}}, \bibinfo {author}
+  {\bibfnamefont {C.}~\bibnamefont {Sagui}}, \bibinfo {author} {\bibfnamefont
+  {V.}~\bibnamefont {Babin}}, \ and\ \bibinfo {author} {\bibfnamefont
+  {P.}~\bibnamefont {Kollman}},\ }\href@noop {} {\enquote {\bibinfo {title}
+  {{AMBER} 10},}\ }\bibinfo {howpublished} {University of California, San
+  Francisco} (\bibinfo {year} {2008})\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Best}\ and\ \citenamefont
+  {Hummer}(2009)}]{best2009optimized}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {R.~B.}\ \bibnamefont
+  {Best}}\ and\ \bibinfo {author} {\bibfnamefont {G.}~\bibnamefont {Hummer}},\
+  }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {Journal of
+  Physical Chemistry B}\ }\textbf {\bibinfo {volume} {113}},\ \bibinfo {pages}
+  {9004} (\bibinfo {year} {2009})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Duan}\ \emph {et~al.}(2003)\citenamefont {Duan},
+  \citenamefont {Wu}, \citenamefont {Chowdhury}, \citenamefont {Lee},
+  \citenamefont {Xiong}, \citenamefont {Zhang}, \citenamefont {Yang},
+  \citenamefont {Cieplak}, \citenamefont {Luo}, \citenamefont {Lee},
+  \citenamefont {Caldwell}, \citenamefont {Wang},\ and\ \citenamefont
+  {Kollman}}]{duan2003point}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {Y.}~\bibnamefont
+  {Duan}}, \bibinfo {author} {\bibfnamefont {C.}~\bibnamefont {Wu}}, \bibinfo
+  {author} {\bibfnamefont {S.}~\bibnamefont {Chowdhury}}, \bibinfo {author}
+  {\bibfnamefont {M.~C.}\ \bibnamefont {Lee}}, \bibinfo {author} {\bibfnamefont
+  {G.~M.}\ \bibnamefont {Xiong}}, \bibinfo {author} {\bibfnamefont
+  {W.}~\bibnamefont {Zhang}}, \bibinfo {author} {\bibfnamefont
+  {R.}~\bibnamefont {Yang}}, \bibinfo {author} {\bibfnamefont {P.}~\bibnamefont
+  {Cieplak}}, \bibinfo {author} {\bibfnamefont {R.}~\bibnamefont {Luo}},
+  \bibinfo {author} {\bibfnamefont {T.}~\bibnamefont {Lee}}, \bibinfo {author}
+  {\bibfnamefont {J.}~\bibnamefont {Caldwell}}, \bibinfo {author}
+  {\bibfnamefont {J.~M.}\ \bibnamefont {Wang}}, \ and\ \bibinfo {author}
+  {\bibfnamefont {P.}~\bibnamefont {Kollman}},\ }\href@noop {} {\bibfield
+  {journal} {\bibinfo  {journal} {Journal of Computational Chemistry}\ }\textbf
+  {\bibinfo {volume} {24}},\ \bibinfo {pages} {1999} (\bibinfo {year}
+  {2003})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Jorgensen}\ \emph {et~al.}(1983)\citenamefont
+  {Jorgensen}, \citenamefont {Chandrasekhar}, \citenamefont {Madura},
+  \citenamefont {Impey},\ and\ \citenamefont
+  {Klein}}]{jorgensen1983comparison}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {W.~L.}\ \bibnamefont
+  {Jorgensen}}, \bibinfo {author} {\bibfnamefont {J.}~\bibnamefont
+  {Chandrasekhar}}, \bibinfo {author} {\bibfnamefont {J.~D.}\ \bibnamefont
+  {Madura}}, \bibinfo {author} {\bibfnamefont {R.~W.}\ \bibnamefont {Impey}}, \
+  and\ \bibinfo {author} {\bibfnamefont {M.~L.}\ \bibnamefont {Klein}},\
+  }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {The Journal of
+  Chemical Physics}\ }\textbf {\bibinfo {volume} {79}},\ \bibinfo {pages} {926}
+  (\bibinfo {year} {1983})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Essmann}\ \emph {et~al.}(1995)\citenamefont
+  {Essmann}, \citenamefont {Perera}, \citenamefont {Berkowitz}, \citenamefont
+  {Darden}, \citenamefont {Lee},\ and\ \citenamefont
+  {Pedersen}}]{essmann1995smooth}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {U.}~\bibnamefont
+  {Essmann}}, \bibinfo {author} {\bibfnamefont {L.}~\bibnamefont {Perera}},
+  \bibinfo {author} {\bibfnamefont {M.~L.}\ \bibnamefont {Berkowitz}}, \bibinfo
+  {author} {\bibfnamefont {T.}~\bibnamefont {Darden}}, \bibinfo {author}
+  {\bibfnamefont {H.}~\bibnamefont {Lee}}, \ and\ \bibinfo {author}
+  {\bibfnamefont {L.~G.}\ \bibnamefont {Pedersen}},\ }\href@noop {} {\bibfield
+  {journal} {\bibinfo  {journal} {The Journal of Chemical Physics}\ }\textbf
+  {\bibinfo {volume} {103}},\ \bibinfo {pages} {8577} (\bibinfo {year}
+  {1995})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Loncharich}, \citenamefont {Brooks},\ and\
+  \citenamefont {Pastor}(1992)}]{loncharich1992langevin}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {R.~J.}\ \bibnamefont
+  {Loncharich}}, \bibinfo {author} {\bibfnamefont {B.~R.}\ \bibnamefont
+  {Brooks}}, \ and\ \bibinfo {author} {\bibfnamefont {R.~W.}\ \bibnamefont
+  {Pastor}},\ }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal}
+  {Biopolymers}\ }\textbf {\bibinfo {volume} {32}},\ \bibinfo {pages} {523}
+  (\bibinfo {year} {1992})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Ryckaert}, \citenamefont {Ciccotti},\ and\
+  \citenamefont {Berendsen}(1977)}]{ryckaert1977numerical}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {J.~P.}\ \bibnamefont
+  {Ryckaert}}, \bibinfo {author} {\bibfnamefont {G.}~\bibnamefont {Ciccotti}},
+  \ and\ \bibinfo {author} {\bibfnamefont {H.~J.~C.}\ \bibnamefont
+  {Berendsen}},\ }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal}
+  {Journal of Computational Physics}\ }\textbf {\bibinfo {volume} {23}},\
+  \bibinfo {pages} {327} (\bibinfo {year} {1977})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Matheron}(1963)}]{matheron1963principles}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {G.}~\bibnamefont
+  {Matheron}},\ }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal}
+  {Economic Geology}\ }\textbf {\bibinfo {volume} {58}},\ \bibinfo {pages}
+  {1246} (\bibinfo {year} {1963})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Matheron}(1973)}]{matheron1973intrinsic}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {G.}~\bibnamefont
+  {Matheron}},\ }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal}
+  {Advances in Applied Probability}\ ,\ \bibinfo {pages} {439}} (\bibinfo
+  {year} {1973})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Coifman}\ and\ \citenamefont
+  {Lafon}(2006{\natexlab{b}})}]{coifman2006geometric}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {R.~R.}\ \bibnamefont
+  {Coifman}}\ and\ \bibinfo {author} {\bibfnamefont {S.}~\bibnamefont
+  {Lafon}},\ }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {Applied
+  and Computational Harmonic Analysis}\ }\textbf {\bibinfo {volume} {21}},\
+  \bibinfo {pages} {31} (\bibinfo {year} {2006}{\natexlab{b}})}\BibitemShut
+  {NoStop}%
+\bibitem [{\citenamefont {A{\"\i}t-Sahalia}(2002)}]{ait2002maximum}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {Y.}~\bibnamefont
+  {A{\"\i}t-Sahalia}},\ }\href@noop {} {\bibfield  {journal} {\bibinfo
+  {journal} {Econometrica}\ }\textbf {\bibinfo {volume} {70}},\ \bibinfo
+  {pages} {223} (\bibinfo {year} {2002})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {A{\"\i}t-Sahalia}\ and\ \citenamefont
+  {Mykland}(2003)}]{ait2003effects}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {Y.}~\bibnamefont
+  {A{\"\i}t-Sahalia}}\ and\ \bibinfo {author} {\bibfnamefont {P.~A.}\
+  \bibnamefont {Mykland}},\ }\href@noop {} {\bibfield  {journal} {\bibinfo
+  {journal} {Econometrica}\ }\textbf {\bibinfo {volume} {71}},\ \bibinfo
+  {pages} {483} (\bibinfo {year} {2003})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {A{\"\i}t-Sahalia}(2008)}]{ait2008closed}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {Y.}~\bibnamefont
+  {A{\"\i}t-Sahalia}},\ }\href@noop {} {\bibfield  {journal} {\bibinfo
+  {journal} {The Annals of Statistics}\ ,\ \bibinfo {pages} {906}} (\bibinfo
+  {year} {2008})}\BibitemShut {NoStop}%
+\end{thebibliography}%
 

### 2013-12-27 paper/jcp_paper_revision.tex
Added files related to paper

--- 2013-12-27 paper/jcp_paper.tex
@@
 %
-These developments have the potential to significantly enable the computational exploration
+Often, high-dimensional data do not uniformly fill the entire ambient space, but rather lie on a lower-dimensional manifold.
+%
+Discovering a parameterization of the manifold therefore yields a compact representation of the data, and may lead to efficient analysis and processing schemes.
+%
+In particular, these developments have the potential to significantly enable the computational exploration
 of physicochemical problems.
+%
+Traditionally, state reduction in chemical reaction networks has been done using some {\em a priori} knowledge about separation of time scales,
+for example, using the quasi-steady state approximation to reduce a network of chemical reactions\cite{bowen1963singular}.
+%
+Alternatively, empirical knowledge about the system of interest has been used to develop coarse grained descriptions,
+such as grouping atoms of a macromolecule into larger ``beads'' to accelerate simulations\cite{monticelli2008martini, spiga2013electrostatic,izvekov2005systematic, saunders2013coarse}.
+%
+The use of manifold learning allows for the systematic discovery of data-driven, low-dimensional descriptions that circumvent the need for such {\em a priori} knowledge of the system.
 %
@@
 ``reduction coordinates" $\mathbf{x}(t)$ can be invaluable in understanding and predicting system behavior.
+%
+For example, such reduction coordinates may be the dihedral angles inferred from trajectories of the atoms of a macromolecule.
 
@@
 different data sets on the same low-dimensional hyperplane in the ambient space
-will lead to different basis vectors - in effect, to different reduction coordinates $\mathbf{x}$.
+will lead to different bases spanning the hyperplane - in effect, to different reduction coordinates $\mathbf{x}$.
 %
@@
 %
-Clearly, the ability to systematically construct a {\em unique} and {\em consistent} reduction
-coordinate set, shared by all measurement ensembles and observation modalities, is invaluable.
+
+In this paper, we propose a new method to embed data in a low-dimensional space.
+%
+This method is based on a kernel constructed from a Riemannian metric that is specially designed to exploit the noise/diffusion induced variability of the data in local neighborhoods.
+%
+By solving the appropriate eigenvector problem, local affinities from the kernel are integrated into a global coordinate system for the data.
+
+We will demonstrate this can produce a {\em unique} and {\em consistent} reduction
+coordinate set, shared by all measurement ensembles and observation modalities.
 %
@@
 
-\subsection{Overview}
+\subsection{Problem Formulation}
 Let $\mathbf{Y}(t)$ be a high-dimensional measured process in $\mathbb{R}^n$ consisting of $n$ observable variables.
@@
 \begin{equation}
-	d x_i(t) = a_i (x_i(t)) dt + d w_i(t), \ i=1,\ldots,d,
+	d x_i(t) = a_i (\mathbf{x}(t)) dt + d w_i(t), \ i=1,\ldots,d,
+	\label{eq:dynamics}
 \end{equation}
@@
 %
-We remark that the available data should be ``rich enough'', i.e., consist of a sufficient amount of historical data with adequate variability, in order to obtain the full empirical model.
+We remark that the available data should be ``rich enough'', i.e., consist of a sufficient amount of historical data with adequate variability, in order to obtain the full empirical model;
+this is further discussed and demonstrated in the experimental study.
 
@@
 and hence, creates a ``sphere" of samples in the underlying domain $\mathbb{R}^d$.
-%%
-%%%RONEN
-% whether here or in the discussion there should be something about testing this
-% practically
 %
@@
 
-Let $\mathbf{D}$ be a diagonal matrix whose elements are the row sums of $\mathbf{W}$, and let $\mathbf{W}^{\mathrm{norm}} = \mathbf{D}^{-1/2}\mathbf{W}\mathbf{D}^{-1/2}$
+Let $\mathbf{D}$ be a diagonal matrix whose elements are the row sums of $\mathbf{W}$, and let
+\begin{equation}\label{eq:kernel}
+\mathbf{W}^{\mathrm{norm}} = \mathbf{D}^{-1/2}\mathbf{W}\mathbf{D}^{-1/2}
+\end{equation}
 be a normalized kernel that shares its eigenvectors with the normalized graph-Laplacian $\mathbf{I}-\mathbf{W}^{\mathrm{norm}} \:$ \cite{chung1997spectral}.
+%
+The diagonal entries of $D$ serve as estimates for the density of data around each point.
 %
@@
 \end{algorithm}
+
+\subsection{Limiting Diffusion Operator} \label{sec:diff_limit}
+Assume that the dynamics of the underlying variables in \eqref{eq:dynamics} can be rewritten as
+\begin{equation} \label{eq:Langevin}
+	d\mathbf{x}(t) = - \nabla U(\mathbf{x}(t)) dt + d \mathbf{w}(t),
+\end{equation}
+where the drift term is expressed as a gradient of a potential $U$ and $\dot{\mathbf{w}}(t) \in \mathbb{R}^d$ is a vector consisting of independent white noises.
+%
+Nadler {\em et al.} \cite{nadler2006diffusion} showed that in the limit $\varepsilon \rightarrow 0$ and $T \rightarrow \infty$ the kernel $\mathbf{W}^{\mathrm{norm}}$ converges to the continuous backward Fokker-Planck operator
+\begin{equation}
+	\mathcal{L} = \Delta - \nabla U \cdot \nabla.
+	\label{eq:limiting_operator}
+\end{equation}
+%
+We remark that, while in diffusion maps \cite{coifman2005geometric}, the limiting operator is the Laplacian on the manifold on which the observations lie,
+here the Laplacian operator on the data that is based on the Mahalanobis distance converges to a limiting Laplacian operator on the
+manifold of the intrinsic variables $\mathbf{x}$.
+
+The convergence to the limiting operator implies that the eigenvectors of $\mathbf{W}^{\mathrm{norm}}$, which are used to represent the NIV, approximate the eigenfunctions of the limiting Fokker-Planck operator on the NIV manifold.
+%
+As such, according to \eqref{eq:limiting_operator}, they depend on the potential.
+%
+Thus, the constructed embedding is consistent as long as the subsets of observations $\mathbf{Y}(t)$ are
+adequate samples of the same (equilibrium) density.
+%
+For example, when looking at multiple data sets from a molecular simulation, the embeddings will be consistent provided each simulation sufficiently samples the same underling free energy surface.
+
+In order to merge observations from different regions,
+the fundamental question is, to what extent the respective sampling densities affect the coordinates we obtain.
+%
+In principle, we can decouple diffusion dynamics (and thus, equilibrium densities)
+from geometry by normalizing the diffusion kernel with the sample density.
+%
+This has been explored for diffusion maps independently of NIV \cite{coifman2005geometric}.
+%
+When the normalized kernel is defined as
+\begin{equation} \label{eq:norm_kernel}
+\widetilde{\mathbf{W}}^{\mathrm{norm}} = \mathbf{D}^{-1}\mathbf{W}\mathbf{D}^{-1},
+\end{equation}
+the limiting diffusion operator is the Laplace-Beltrami operator \cite{nadler2006diffusion}, given by $\widetilde{\mathcal{L}} = \Delta$.
+%
+In other words, this diffusion captures only the geometry of the observations and is invariant to the potential (or, alternatively,
+to the details of the sampling density).
+This kernel is similar to the kernel in \eqref{eq:kernel}, but $D$ appears with coefficient -1 rather than $-1/2$.
+
+Since local coordinates are based on the noise term in \eqref{eq:Langevin}, sets of measurements from different regions
+will essentially detect the same system of coordinates (the same variable ``axes");
+yet different sampling densities (arising, for example,
+from the drift term in \eqref{eq:Langevin}) will result in different {\em scalings} of these axes.
+%
+Applying the Laplace-Beltrami normalization in the NIV context will thus eliminate sampling density effects.
+%
+An alternative approach would find the scaling transformations that optimally ``register"
+data sampled from common regions (in the spirit of histogram reweighting \cite{ferrenberg1988new}).
 
@@
 \end{equation}
-are constructed and later extended to create the multiscale approximations as defined in Eq. \ref{eq:LP_multi_scale}.
+are constructed and later extended to create the multiscale approximations as defined in \eqref{eq:LP_multi_scale}.
 %
@@
 \item
-Use the smoothing operators to obtain a multiscale representation (see Eq. \ref{eq:LP_multi_scale_app}) of $\mathbf{f}:\Psi(i) \rightarrow Y(i)$.
+Use the smoothing operators to obtain a multiscale representation (see \eqref{eq:LP_multi_scale_app}) of $\mathbf{f}:\Psi(i) \rightarrow Y(i)$.
 
@@
 Use the extended smoothing operators to approximate the value of $Y(\tilde{i})$ as $Y(\tilde{i}) \approx \mathbf{f}_l(\Psi(\tilde{i})) = \mathbf{f}_0(\Psi(\tilde{i})) + \sum_{k}\sum_{(i \in S)}p_k(\Psi(\tilde{i}), \Psi(i))\delta_k(\Psi(i))$
-
-%\item
-%The value $Y(\tilde{i})$ is approximated by $\mathbf{f}_l(\Psi(\tilde{i}))$.
 
@@
 \end{equation}
-
 The ``$^{*}$'' denotes an activated form of a species, and the ``:'' denotes a complex formed between two species; the complexes $E:S$ and $S:E$ are not equivalent.
@@
 
-We generate 3000 random initial conditions $\mathbf{Y}_0(1), \dots, \mathbf{Y}_0(3000) \in \mathbb{R}^6$, enforcing that all concentrations
-must be non-negative.
+We generate 3000 initial conditions $\mathbf{Y}_0(1), \dots, \mathbf{Y}_0(3000) \in \mathbb{R}^6$ uniformly at random from the
+region of state space where all concentrations (respecting conservation laws) are non-negative.
 %
@@
 
+To determine whether a sufficient amount of data is available to obtain an accurate embedding, we compute the NIV embeddings for different numbers of data points
+(arising from different numbers of trajectories initialized randomly in the full concentration space).
+%
+Figure \ref{fig:rxn_convergence} shows the two-dimensional NIV embeddings and corresponding eigenvalue spectra of the kernel for the Gillespie SSA example computed from 10, 100, 500, and 1500 data points.
+%
+We observe that the eigenvalue spectra appear converged for 100 data points and above.
+%
+This implies that a sufficient amount of data is available merely to discover the intrinsic variables.
+%
+However, we observe that 100 points are insufficient to construct a self-consistent embedding {\em in the two variables discovered}.
+%
+In order to account for the different scaling of the NIV coordinate axes, we require at least 500 data points.
+%
+This last convergence is exemplified in Figure \ref{fig:rxn_convergence} (c) and (d).
+%
+The convergence of the spectrum of the kernel and the ultimate self-consistency of the NIV embedding is our empirical indicator that a sufficient amount of data is available.
+\begin{figure*}[ht]
+    \includegraphics[width=6in]{fig4b}
+    \caption{Two-dimensional NIV embeddings (top) and corresponding eigenvalue spectra (bottom) calculated from subsets of $\mathcal{Y}_1$ containing (a) 10 data points, (b) 100 data points, (c) 500 data points, and (d) 1500 data points.
+One can see that the embedding and spectrum in (a) is inconsistent with the other embeddings, and that, although the spectrum in (b) appears converged, the embedding is not converged until (c).
+}
+    \label{fig:rxn_convergence}
+\end{figure*}
+
 We then use NIV together with Laplacian Pyramids to estimate the values of $S:E$ and $F:E^{*}$ for $\mathcal{Y}_2$.
@@
 Our simulations are concentrated around minimum B in the free energy surface, located at $\phi \approx -65^{\circ}$, $\psi \approx 150^{\circ}$.
+%
+We choose to sample only around minimum B, as each data set must sample the same region of the potential energy surface to obtain consistent NIV, as described in Section \ref{sec:diff_limit}.
 %
@@
 for NIV, we compute the covariances as in \eqref{eq:cov}, with $L=10$.
-%at the ends of each simulation trajectory, we take fewer points before or following the point of interest to estimate the covariance
-%(as neighboring points in our data set that do not come from the
-%same trajectories should not contribute to the estimate of the local covariance).
 
@@
 
-%\begin{table}
-%\begin{tabular}{| r || c | c |}
-%  \hline
-%  & Coordinate 1 & Coordinate 2 \\
-%  \hline
-%  \hline
-%  NLICA & 0.62 & 0.84 \\
-%  \hline
-%  DM & 0.54 & 0.60 \\
-%  \hline
-%\end{tabular}
-%\caption{Correlation between the embedding coordinates from NLICA and DMAPS using the atoms 5, 7, and 9 and using the atoms 2, 4, 6, 8, and 10.}
-%\end{table}
-
 We then use NIV together with LP to predict the conformation of Ala2 when we only observe some of the atoms.
@@
 We compute the covariances as in \eqref{eq:cov} with $L=15$.
-%at the ends of each simulation trajectory, we take fewer points before or following the point of interest to estimate the covariance.
 %
@@
 The embeddings for the two data sets are found to be fairly consistent, with correlations of 0.97, 0.72, 0.85 for the first, second, and third NIV, respectively.
-
-%\begin{table}
-%\begin{tabular}{| r || c | c | c |}
-%  \hline
-%  & Coordinate 1 & Coordinate 2 & Coordinate 3 \\
-%  \hline
-%  \hline
-%  NLICA & 0.97 & 0.72 & 0.85 \\
-%  \hline
-%  DM & 0.95 & 0.76 & 0.74 \\
-%  \hline
-%\end{tabular}
-%\caption{Correlation between the embedding coordinates from NLICA and DMAPS using all and just odd coordinates.}
-%\end{table}
 
@@
 there is qualitative agreement between the true and reconstructed configurations.
+%
+The discrepancies between the true and reconstructed positions of atoms 9 and 10 (see Figure \ref{fig:ala_fes} for atom indexing) for some of the molecular structures are not surprising.
+%
+From Figure \ref{fig:ala_embed}, one can see that the first three NIV describe the flipping of atoms 1 and 3 and the dihedral angles $\phi$ and $\psi$.
+%
+However, atoms 9 and 10 do not participate in any of these physical quantities, and so there is little information about their positions contained in the first three NIV.
+%
+We suspect that using more NIV would result in a better description of atoms 9 and 10 and lead to more accurate structure reconstructions.
 
@@
 LP arguably outperforms simple nearest neighbor search for all of the atoms.
+%
+To determine why LP outperforms nearest neighbor search, we examine the errors at different scales in the LP algorithm.
+%
+The example described in Figure \ref{fig:LP_ex} demonstrates the appropriateness of the LP algorithm to signals with an intrinsic multiscale structure.
+%
+We perform a similar analysis using the data from the alanine dipeptide example, and show that the observed signal does exhibit a similar error structure,
+with large errors in some regions of NIV space and small errors in other regions;
+these results are shown in Figure \ref{fig:LapPyr_ala2_errors}.
+%
+We therefore conclude that the alanine dipeptide data does contain a measure of multiscale behavior, and, therefore, that LP is an appropriate algorithm for reconstruction.
 
@@
 
+\begin{figure*}[ht]
+    \centering
+    \includegraphics[width=3in]{fig10b}
+    \caption{The error in the reconstructed position of the z-coordinate of atom 10 as a function of $\psi_1$.
+One can see that the error is large when $\psi_1$ is large (similar to the error structure in Figure \ref{fig:LP_ex}), demonstrating that our data is indeed multiscale and that Laplacian Pyramids is an appropriate interpolation algorithm to use.}
+    \label{fig:LapPyr_ala2_errors}
+\end{figure*}
+
 \section{Conclusions} \label{sec:conclusions}
@@
 
-%In addition, we have stated that the measurement function $\mathbf{f}$ has to be ``rich enough'' in order to recover the variables of the underlying stochastic process.
-%However, determining whether $\mathbf{f}$ meets such criteria is a non-trivial task.
-%For many problems, one would like to say what types of measurement functions $\mathbf{f}$ are rich enough,
-%as knowing which functions retain sufficient detail of the underlying stochastic process can aid us in building accurate reduced models.
-%
-%
 Other issues, such as accurately estimating the covariance matrices required for the computation of the Mahalanobis distance, are also of current research interest.
@@
 %\bibliography{../../references/references}
+%\bibliography{references}
 %merlin.mbs aipnum4-1.bst 2010-07-25 4.21a (PWD, AO, DPC) hacked
@@
 %Control: production of eprint (0) enabled
-\begin{thebibliography}{39}%
+\begin{thebibliography}{46}%
 \makeatletter
@@
   {5} (\bibinfo {year} {2006}{\natexlab{a}})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Bowen}, \citenamefont {Acrivos},\ and\ \citenamefont
+  {Oppenheim}(1963)}]{bowen1963singular}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {J.}~\bibnamefont
+  {Bowen}}, \bibinfo {author} {\bibfnamefont {A.}~\bibnamefont {Acrivos}}, \
+  and\ \bibinfo {author} {\bibfnamefont {A.}~\bibnamefont {Oppenheim}},\
+  }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {Chemical
+  Engineering Science}\ }\textbf {\bibinfo {volume} {18}},\ \bibinfo {pages}
+  {177} (\bibinfo {year} {1963})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Monticelli}\ \emph {et~al.}(2008)\citenamefont
+  {Monticelli}, \citenamefont {Kandasamy}, \citenamefont {Periole},
+  \citenamefont {Larson}, \citenamefont {Tieleman},\ and\ \citenamefont
+  {Marrink}}]{monticelli2008martini}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {L.}~\bibnamefont
+  {Monticelli}}, \bibinfo {author} {\bibfnamefont {S.~K.}\ \bibnamefont
+  {Kandasamy}}, \bibinfo {author} {\bibfnamefont {X.}~\bibnamefont {Periole}},
+  \bibinfo {author} {\bibfnamefont {R.~G.}\ \bibnamefont {Larson}}, \bibinfo
+  {author} {\bibfnamefont {D.~P.}\ \bibnamefont {Tieleman}}, \ and\ \bibinfo
+  {author} {\bibfnamefont {S.-J.}\ \bibnamefont {Marrink}},\ }\href@noop {}
+  {\bibfield  {journal} {\bibinfo  {journal} {Journal of Chemical Theory and
+  Computation}\ }\textbf {\bibinfo {volume} {4}},\ \bibinfo {pages} {819}
+  (\bibinfo {year} {2008})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Spiga}\ \emph {et~al.}(2013)\citenamefont {Spiga},
+  \citenamefont {Alemani}, \citenamefont {Degiacomi}, \citenamefont
+  {Cascella},\ and\ \citenamefont {Dal~Peraro}}]{spiga2013electrostatic}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {E.}~\bibnamefont
+  {Spiga}}, \bibinfo {author} {\bibfnamefont {D.}~\bibnamefont {Alemani}},
+  \bibinfo {author} {\bibfnamefont {M.~T.}\ \bibnamefont {Degiacomi}}, \bibinfo
+  {author} {\bibfnamefont {M.}~\bibnamefont {Cascella}}, \ and\ \bibinfo
+  {author} {\bibfnamefont {M.}~\bibnamefont {Dal~Peraro}},\ }\href@noop {}
+  {\bibfield  {journal} {\bibinfo  {journal} {Journal of Chemical Theory and
+  Computation}\ } (\bibinfo {year} {2013})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Izvekov}, \citenamefont {Violi},\ and\ \citenamefont
+  {Voth}(2005)}]{izvekov2005systematic}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {S.}~\bibnamefont
+  {Izvekov}}, \bibinfo {author} {\bibfnamefont {A.}~\bibnamefont {Violi}}, \
+  and\ \bibinfo {author} {\bibfnamefont {G.~A.}\ \bibnamefont {Voth}},\
+  }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {The Journal of
+  Physical Chemistry B}\ }\textbf {\bibinfo {volume} {109}},\ \bibinfo {pages}
+  {17019} (\bibinfo {year} {2005})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Saunders}\ and\ \citenamefont
+  {Voth}(2013)}]{saunders2013coarse}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {M.~G.}\ \bibnamefont
+  {Saunders}}\ and\ \bibinfo {author} {\bibfnamefont {G.~A.}\ \bibnamefont
+  {Voth}},\ }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {Annual
+  review of biophysics}\ }\textbf {\bibinfo {volume} {42}},\ \bibinfo {pages}
+  {73} (\bibinfo {year} {2013})}\BibitemShut {NoStop}%
 \bibitem [{\citenamefont {Jolliffe}(2005)}]{jolliffe2005principal}%
@@
   Bookstore},\ \bibinfo {year} {1997})\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Nadler}\ \emph {et~al.}(2006)\citenamefont {Nadler},
+  \citenamefont {Lafon}, \citenamefont {Coifman},\ and\ \citenamefont
+  {Kevrekidis}}]{nadler2006diffusion}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {B.}~\bibnamefont
+  {Nadler}}, \bibinfo {author} {\bibfnamefont {S.}~\bibnamefont {Lafon}},
+  \bibinfo {author} {\bibfnamefont {R.~R.}\ \bibnamefont {Coifman}}, \ and\
+  \bibinfo {author} {\bibfnamefont {I.~G.}\ \bibnamefont {Kevrekidis}},\
+  }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal} {Applied and
+  Computational Harmonic Analysis}\ }\textbf {\bibinfo {volume} {21}},\
+  \bibinfo {pages} {113} (\bibinfo {year} {2006})}\BibitemShut {NoStop}%
+\bibitem [{\citenamefont {Ferrenberg}\ and\ \citenamefont
+  {Swendsen}(1988)}]{ferrenberg1988new}%
+  \BibitemOpen
+  \bibfield  {author} {\bibinfo {author} {\bibfnamefont {A.~M.}\ \bibnamefont
+  {Ferrenberg}}\ and\ \bibinfo {author} {\bibfnamefont {R.~H.}\ \bibnamefont
+  {Swendsen}},\ }\href@noop {} {\bibfield  {journal} {\bibinfo  {journal}
+  {Physical review letters}\ }\textbf {\bibinfo {volume} {61}},\ \bibinfo
+  {pages} {2635} (\bibinfo {year} {1988})}\BibitemShut {NoStop}%
 \bibitem [{\citenamefont {Burt}\ and\ \citenamefont
@@
   {Zhang}, \citenamefont {Merz}, \citenamefont {Wang}, \citenamefont {Hayik},
-  \citenamefont {Roitberg}, \citenamefont {Seabra}, \citenamefont {Kolossváry},
-  \citenamefont {Wong}, \citenamefont {Paesani}, \citenamefont {Vanicek},
-  \citenamefont {Wu}, \citenamefont {Brozell}, \citenamefont {Steinbrecher},
-  \citenamefont {Gohlke}, \citenamefont {Yang}, \citenamefont {Tan},
-  \citenamefont {Mongan}, \citenamefont {Hornak}, \citenamefont {Cui},
-  \citenamefont {Mathews}, \citenamefont {Seetin}, \citenamefont {Sagui},
-  \citenamefont {Babin},\ and\ \citenamefont {Kollman}}]{case2008Amber}%
+  \citenamefont {Roitberg}, \citenamefont {Seabra}, \citenamefont
+  {Kolossv\'{a}ry}, \citenamefont {Wong}, \citenamefont {Paesani},
+  \citenamefont {Vanicek}, \citenamefont {Wu}, \citenamefont {Brozell},
+  \citenamefont {Steinbrecher}, \citenamefont {Gohlke}, \citenamefont {Yang},
+  \citenamefont {Tan}, \citenamefont {Mongan}, \citenamefont {Hornak},
+  \citenamefont {Cui}, \citenamefont {Mathews}, \citenamefont {Seetin},
+  \citenamefont {Sagui}, \citenamefont {Babin},\ and\ \citenamefont
+  {Kollman}}]{case2008Amber}%
   \BibitemOpen
@@
   {G.}~\bibnamefont {Seabra}}, \bibinfo {author} {\bibfnamefont
-  {I.}~\bibnamefont {Kolossváry}}, \bibinfo {author} {\bibfnamefont {K.~F.}\
-  \bibnamefont {Wong}}, \bibinfo {author} {\bibfnamefont {F.}~\bibnamefont
-  {Paesani}}, \bibinfo {author} {\bibfnamefont {J.}~\bibnamefont {Vanicek}},
-  \bibinfo {author} {\bibfnamefont {X.}~\bibnamefont {Wu}}, \bibinfo {author}
-  {\bibfnamefont {S.~R.}\ \bibnamefont {Brozell}}, \bibinfo {author}
-  {\bibfnamefont {T.}~\bibnamefont {Steinbrecher}}, \bibinfo {author}
-  {\bibfnamefont {H.}~\bibnamefont {Gohlke}}, \bibinfo {author} {\bibfnamefont
-  {L.}~\bibnamefont {Yang}}, \bibinfo {author} {\bibfnamefont {C.}~\bibnamefont
-  {Tan}}, \bibinfo {author} {\bibfnamefont {J.}~\bibnamefont {Mongan}},
-  \bibinfo {author} {\bibfnamefont {V.}~\bibnamefont {Hornak}}, \bibinfo
-  {author} {\bibfnamefont {G.}~\bibnamefont {Cui}}, \bibinfo {author}
-  {\bibfnamefont {D.~H.}\ \bibnamefont {Mathews}}, \bibinfo {author}
-  {\bibfnamefont {M.~G.}\ \bibnamefont {Seetin}}, \bibinfo {author}
-  {\bibfnamefont {C.}~\bibnamefont {Sagui}}, \bibinfo {author} {\bibfnamefont
-  {V.}~\bibnamefont {Babin}}, \ and\ \bibinfo {author} {\bibfnamefont
-  {P.}~\bibnamefont {Kollman}},\ }\href@noop {} {\enquote {\bibinfo {title}
-  {{AMBER} 10},}\ }\bibinfo {howpublished} {University of California, San
-  Francisco} (\bibinfo {year} {2008})\BibitemShut {NoStop}%
+  {I.}~\bibnamefont {Kolossv\'{a}ry}}, \bibinfo {author} {\bibfnamefont
+  {K.~F.}\ \bibnamefont {Wong}}, \bibinfo {author} {\bibfnamefont
+  {F.}~\bibnamefont {Paesani}}, \bibinfo {author} {\bibfnamefont
+  {J.}~\bibnamefont {Vanicek}}, \bibinfo {author} {\bibfnamefont
+  {X.}~\bibnamefont {Wu}}, \bibinfo {author} {\bibfnamefont {S.~R.}\
+  \bibnamefont {Brozell}}, \bibinfo {author} {\bibfnamefont {T.}~\bibnamefont
+  {Steinbrecher}}, \bibinfo {author} {\bibfnamefont {H.}~\bibnamefont
+  {Gohlke}}, \bibinfo {author} {\bibfnamefont {L.}~\bibnamefont {Yang}},
+  \bibinfo {author} {\bibfnamefont {C.}~\bibnamefont {Tan}}, \bibinfo {author}
+  {\bibfnamefont {J.}~\bibnamefont {Mongan}}, \bibinfo {author} {\bibfnamefont
+  {V.}~\bibnamefont {Hornak}}, \bibinfo {author} {\bibfnamefont
+  {G.}~\bibnamefont {Cui}}, \bibinfo {author} {\bibfnamefont {D.~H.}\
+  \bibnamefont {Mathews}}, \bibinfo {author} {\bibfnamefont {M.~G.}\
+  \bibnamefont {Seetin}}, \bibinfo {author} {\bibfnamefont {C.}~\bibnamefont
+  {Sagui}}, \bibinfo {author} {\bibfnamefont {V.}~\bibnamefont {Babin}}, \ and\
+  \bibinfo {author} {\bibfnamefont {P.}~\bibnamefont {Kollman}},\ }\href@noop
+  {} {\enquote {\bibinfo {title} {{AMBER} 10},}\ }\bibinfo {howpublished}
+  {University of California, San Francisco} (\bibinfo {year}
+  {2008})\BibitemShut {NoStop}%
 \bibitem [{\citenamefont {Best}\ and\ \citenamefont

### 2013-12-27 paper/make_figures10.tex
Added files related to paper

    \documentclass[]{article}
\usepackage{graphicx}
\usepackage[paperwidth=4in, paperheight=6in, top=0in, bottom=0in, left=0in, right=0in]{geometry}

\thispagestyle{empty}

\begin{document}
\centering

\centering
\includegraphics[width=0.4\textwidth,trim=0in 0.7in 0in 0in, clip]{molec_figures/molecule300_balls}
    \includegraphics[width=0.4\textwidth,trim=0in 0.7in 0in 0in, clip]{molec_figures/molecule300_wire}\\
    (a)

\centering
  \includegraphics[width=0.4\textwidth,trim=0in 0.7in 0in 0in, clip]{molec_figures/molecule600_balls}
    \includegraphics[width=0.4\textwidth,trim=0in 0.7in 0in 0in, clip]{molec_figures/molecule600_wire}\\

(b)

\centering
  \includegraphics[width=0.4\textwidth,trim=0in 0.7in 0in 0in, clip]{molec_figures/molecule900_balls}
    \includegraphics[width=0.4\textwidth,trim=0in 0.7in 0in 0in, clip]{molec_figures/molecule900_wire}\\
(c)

\end{document}

### 2013-12-27 paper/make_figures3.tex
Added files related to paper

\documentclass[]{article}
\usepackage{graphicx}
\usepackage[paperwidth=6in, paperheight=6in, top=0in, bottom=0in, left=0in, right=0in]{geometry}

\thispagestyle{empty}

\begin{document}
\centering
\begin{minipage}{0.45\textwidth}
\centering
\includegraphics[height=6cm]{rxn_figures/rxn_manifold1}\llap{\raisebox{4cm}{\includegraphics[height=2.25cm]{rxn_figures/rxn_manifold1_2}}}\\
(a)
\end{minipage}
\begin{minipage}{0.45\textwidth}
\centering
  \includegraphics[height=6cm]{rxn_figures/rxn_manifold2}\llap{\raisebox{4cm}{\includegraphics[height=2.25cm]{rxn_figures/rxn_manifold2_2}}} \\
(b)
\end{minipage}\\
\begin{minipage}{0.45\textwidth}
\centering
  \includegraphics[height=6cm]{rxn_figures/rxn_manifold3}\llap{\raisebox{4cm}{\includegraphics[height=2.25cm]{rxn_figures/rxn_manifold3_2}}} \\
(c)
\end{minipage}
\begin{minipage}{0.45\textwidth}
\centering
  \includegraphics[height=6cm]{rxn_figures/rxn_manifold4}\llap{\raisebox{4cm}{\includegraphics[height=2.25cm]{rxn_figures/rxn_manifold4_2}}} \\
(d)
\end{minipage}
\end{document}

### 2013-12-27 paper/make_figures4.tex
Added files related to paper

\documentclass[]{article}
\usepackage{graphicx}
\usepackage[paperwidth=6in, paperheight=5in, top=0in, bottom=0in, left=0in, right=0in]{geometry}

\thispagestyle{empty}

\begin{document}
\centering
\begin{minipage}{0.4\textwidth}
\centering
    \includegraphics[width=\textwidth]{rxn_figures/rxn_NLICA1}
(a)
\end{minipage}
\begin{minipage}{0.4\textwidth}
\centering
      \includegraphics[width=\textwidth]{rxn_figures/rxn_NLICA2}
(b)
\end{minipage}\\
\begin{minipage}{0.4\textwidth}
\centering
\includegraphics[width=\textwidth]{rxn_figures/rxn_NLICA_corr1}
(c)
\end{minipage}
\begin{minipage}{0.4\textwidth}
\centering
  \includegraphics[width=\textwidth]{rxn_figures/rxn_NLICA_corr2}
(d)
\end{minipage}
\end{document}

### 2013-12-27 paper/make_figures4b.tex
Added files related to paper

\documentclass[]{article}
\usepackage{graphicx}
\usepackage[paperwidth=6in, paperheight=2.75in, top=0in, bottom=0in, left=0in, right=0in]{geometry}

\thispagestyle{empty}

\begin{document}
\centering
\begin{minipage}{0.22\textwidth}
\centering
\includegraphics[width=\textwidth]{rxn_figures/rxn_npoints_embed1}
\includegraphics[width=\textwidth]{rxn_figures/rxn_npoints_spectrum1}
(a)
\end{minipage}
\hfill
\begin{minipage}{0.22\textwidth}
\centering
\includegraphics[width=\textwidth]{rxn_figures/rxn_npoints_embed2}\\
\includegraphics[width=\textwidth]{rxn_figures/rxn_npoints_spectrum2}
(b)
\end{minipage}
\hfill
\begin{minipage}{0.22\textwidth}
\centering
\includegraphics[width=\textwidth]{rxn_figures/rxn_npoints_embed4}\\
\includegraphics[width=\textwidth]{rxn_figures/rxn_npoints_spectrum4}
(c)
\end{minipage}
\hfill
\begin{minipage}{0.22\textwidth}
\centering
\includegraphics[width=\textwidth]{rxn_figures/rxn_npoints_embed6}\\
\includegraphics[width=\textwidth]{rxn_figures/rxn_npoints_spectrum6}
(d)
\end{minipage}
\end{document}

### 2013-12-27 paper/make_figures8.tex
Added files related to paper

\documentclass[]{article}
\usepackage{graphicx}
\usepackage[paperwidth=6in, paperheight=2.5in, top=0in, bottom=0in, left=0in, right=0in]{geometry}

\thispagestyle{empty}

\begin{document}
\centering
\begin{minipage}{0.3\textwidth}
\centering
    \includegraphics[width=\textwidth]{ala_figures_allodd/ala2_embed1}
    (a)
\end{minipage}
\begin{minipage}{0.3\textwidth}
\centering
    \includegraphics[width=\textwidth]{ala_figures_allodd/ala2_embed2}

(b)
\end{minipage}
\begin{minipage}{0.3\textwidth}
\centering
    \includegraphics[width=\textwidth]{ala_figures_allodd/ala2_embed3}

(c)
\end{minipage}
\end{document}

### 2013-12-27 paper/make_figures9.tex
Added files related to paper

\documentclass[]{article}
\usepackage{graphicx}
\usepackage{tikz}
\usepackage[paperwidth=4.5in, paperheight=7in, top=0in, bottom=0in, left=0in, right=0in]{geometry}

\thispagestyle{empty}

\begin{document}

  \centering
    \foreach \atomnumber in {1,2,3,6,8} {
        \includegraphics[width=0.3\textwidth]{ala_figures_allodd/ala2_recon_corr_\atomnumber_1}
        \includegraphics[width=0.3\textwidth]{ala_figures_allodd/ala2_recon_corr_\atomnumber_2}
        \includegraphics[width=0.3\textwidth]{ala_figures_allodd/ala2_recon_corr_\atomnumber_3}\\
      }
\end{document}

### 2013-12-27 paper/revision1.tex
Added files related to paper

--- 2013-12-27 paper/draft9.tex
@@
 	d x_i(t) = a_i (x_i(t)) dt + d w_i(t), \ i=1,\ldots,d,
+	\label{eq:dynamics}
 \end{equation}
@@
 \end{algorithm}
+
+\subsection{Limiting Diffusion Operator} \label{sec:diff_limit}
+Assume that the dynamics of the underlying variables in \eqref{eq:dynamics} can be rewritten as
+\begin{equation}
+	d x_i(t) = - \nabla U_i (x_i(t)) dt + d w_i(t), \ i=1,\ldots,d
+\end{equation}
+where the drift term is expressed as a gradient of a potential field $U_i$.
+Nadler {\em et al.} \cite{nadler2006diffusion} showed that in the limit $\varepsilon \rightarrow 0$ and $T \rightarrow \infty$ the kernel $\mathbf{W}^{\mathrm{norm}}$ converges to the continuous backward Fokker-Planck operator
+\begin{equation}
+	\mathcal{L} g = \Delta g - \nabla U \cdot \nabla g
+	\label{eq:limiting_operator}
+\end{equation}
+where $g:\mathcal{M}_x \rightarrow \mathbb{R}$ is any real function on the domain of the underlying variables.
+Since we assume independent underlying variables, the potential is additive \cite{singer2008non}
+\begin{equation}
+	U(\mathbf{x}(t)) = \sum \limits _{i=1}^d U_i(x_i(t)).
+\end{equation}
+
+This converges implies that the eigenvectors of $\mathbf{W}^{\mathrm{norm}}$, which are used to represent the NIV, approximate the eigenfunctions of the limiting Fokker-Planck operator.
+%
+As such, according to \eqref{eq:limiting_operator}, they depend on the potential field.
+%
+Thus, the constructed embedding is consistent as long as the subsets of observations $\mathbf{Y}(t)$ are sampled from the same common region.
+%
+For example, when looking at multiple data sets from a molecular simulation, the embeddings will be consistent provided each simulation samples the same underling free energy surface.
+
+In order to merge observations from different regions, we need to account for the different potential fields.
+%
+ This can be achieved by a different normalization of the kernel.
+In case the normalized kernel is defined as $\widetilde{\mathbf{W}}^{\mathrm{norm}} = \mathbf{D}^{-1}\mathbf{W}\mathbf{D}^{-1}$, the limiting diffusion operator is the Laplace-Beltram operator \cite{nadler2006diffusion}, given by
+\begin{equation}
+	\widetilde{\mathcal{L}} g = \Delta g.
+\end{equation}
+%
+In other words, this diffusion captures only the geometry of the observations and invariant to the different potential fields.
+%
+Therefore, this normalization separates the geometry of the underlying manifold of intrinsic variables from the statistics of the observations.
+
+Unfortunately, in practice, this normalization is not sufficient, and one needs to consider various affine transformations to relate different sets of intrinsic variables.
+%
+This has been explored in Lafon {\em et al.} \cite{lafon2006data}, but is currently beyond the scope of this paper.
+%
+We will only consider applications where diferent simulations sample the same underlying potential field.
 
@@
 
-We generate 3000 random initial conditions $\mathbf{Y}_0(1), \dots, \mathbf{Y}_0(3000) \in \mathbb{R}^6$, enforcing that all concentrations
-must be non-negative.
+We generate 3000 initial conditions $\mathbf{Y}_0(1), \dots, \mathbf{Y}_0(3000) \in \mathbb{R}^6$ uniformly at random from the region of state space where all concentrations are non-negative.
 %
@@
 %
+We choose to sample only around minimum B, as each data set must sample the same region of the potential energy surface to obtain consistent NIV, as described in Section \ref{sec:diff_limit}.
+%
 We start many simulations at $10^{\circ}$ away from the minimum, and allow the simulations to each run for 0.1~ps, while recording the configuration of Ala2 every 1~fs (therefore, each trajectory is 100 points long).
@@
 there is qualitative agreement between the true and reconstructed configurations.
+%
+The discrepancies between the true and reconstructed positions of atoms 9 and 10 (see Figure \ref{fig:ala_fes} for atom indexing) for some of the molecular structures is not suprising.
+%
+From Figure \ref{fig:ala_embed}, one can see that the first three NIV describe the flipping of atoms 1 and 3 and the dihedral angles $\phi$ and $\psi$.
+%
+However, atoms 9 and 10 do not participate in any of these physical quantities, and so there is little information about their positions contained in the first three NIV.
+%
+We suspect that using more NIV would result in a better description of atoms 9 and 10 and lead to more accurrate structure reconstructions.
 
@@
 \bibliography{../../references/references}
+%\bibliography{references}
 

### 2013-12-27 paper/revision2.tex
Added files related to paper

--- 2013-12-27 paper/revision1.tex
@@
 %
-These developments have the potential to significantly enable the computational exploration
+Often, high-dimensional data do not uniformly fill the entire ambient space, but rather lie on a lower-dimensional manifold.
+%
+Discovering a parameterization of the manifold therefore yields a compact representation of the data, and may lead to efficient analysis and processing schemes.
+%
+In particular, these developments have the potential to significantly enable the computational exploration
 of physicochemical problems.
+%
+Traditionally, state reduction in chemical systems has been done using some {\em a priori} knowledge about separation of timescales,
+for example, using the quasi-steady state approximation to reduce a network of chemical reactions \cite{....}.
+%
+Alternatively, empirical knowledge about the system of interest has been used to develop coarse grained descriptions, 
+such as grouping atoms of a macromolecule into larger ``beads'' to accelerate simulations \cite{...}.
+%
+The use of manifold learning allows for a systematic discovery of a data-driven, low-dimensional description that circumvents the need for such {\em a priori} knowledge of the system.
 %
@@
 %
-Clearly, the ability to systematically construct a {\em unique} and {\em consistent} reduction
-coordinate set, shared by all measurement ensembles and observation modalities, is invaluable.
+
+In this paper, we propose a new method to embed data in a low-dimensional space.
+%
+This method is based on a kernel constructed from a Riemannian metric that is specially designed to exploit the variability of the data in local neighborhoods.
+%
+By solving the appropriate eigenvector problem, local affinities from the kernel are integrated into a global coordinate system for the data.
+
+We will demonstrate that this produces a {\em unique} and {\em consistent} reduction
+coordinate set, shared by all measurement ensembles and observation modalities.
 %
@@
 Assume that the dynamics of the underlying variables in \eqref{eq:dynamics} can be rewritten as
-\begin{equation}
+\begin{equation} \label{eq:Langevin}
 	d x_i(t) = - \nabla U_i (x_i(t)) dt + d w_i(t), \ i=1,\ldots,d
@@
 \end{equation}
-
-This converges implies that the eigenvectors of $\mathbf{W}^{\mathrm{norm}}$, which are used to represent the NIV, approximate the eigenfunctions of the limiting Fokker-Planck operator.
+We remark that, unlike in diffusion maps \cite{...}, when the limiting operator is in the domain of the observations $\mathbf{Y}$, here the Laplacian operator that is based on the Mahalanobis distance converges to a limiting operator in the domain of the intrinsic variables $\mathbf{x}$.
+
+The convergence to the limiting operator implies that the eigenvectors of $\mathbf{W}^{\mathrm{norm}}$, which are used to represent the NIV, approximate the eigenfunctions of the limiting Fokker-Planck operator.
 %
@@
 %
-For example, when looking at multiple data sets from a molecular simulation, the embeddings will be consistent provided each simulation samples the same underling free energy surface.
-
-In order to merge observations from different regions, we need to account for the different potential fields.
-%
- This can be achieved by a different normalization of the kernel.
-In case the normalized kernel is defined as $\widetilde{\mathbf{W}}^{\mathrm{norm}} = \mathbf{D}^{-1}\mathbf{W}\mathbf{D}^{-1}$, the limiting diffusion operator is the Laplace-Beltram operator \cite{nadler2006diffusion}, given by
+For example, when looking at multiple data sets from a molecular simulation, the embeddings will be consistent provided each simulation sufficiently samples the same underling free energy surface.
+
+In order to merge observations from different regions, we need to account for the potential field.
+%
+The fundamental question is, to what extent the density of sampling different regions affects the coordinates we obtain.
+%
+In principal, we can decouple diffusion dynamics from geometry by normalizing the diffusion kernel with sample density.
+%
+This has been explored for diffusion maps independently of NIV.
+%
+When the normalized kernel is defined as $\widetilde{\mathbf{W}}^{\mathrm{norm}} = \mathbf{D}^{-1}\mathbf{W}\mathbf{D}^{-1}$, the limiting diffusion operator is the Laplace-Beltrami operator \cite{nadler2006diffusion}, given by
 \begin{equation}
@@
 %
-In other words, this diffusion captures only the geometry of the observations and invariant to the different potential fields.
+In other words, this diffusion captures only the geometry of the observations and is invariant to the potential field.
 %
@@
 
-Unfortunately, in practice, this normalization is not sufficient, and one needs to consider various affine transformations to relate different sets of intrinsic variables.
-%
-This has been explored in Lafon {\em et al.} \cite{lafon2006data}, but is currently beyond the scope of this paper.
-%
-We will only consider applications where diferent simulations sample the same underlying potential field.
+Since local coordinates are based on the noise term in \eqref{eq:Langevin}, sets of measurements from different regions
+will essentially detect the same system of coordinates (the same variable ``axes"), yet the different sampling densities arising from the drift term in \eqref{eq:Langevin} will
+result in different scalings of these axes.
+%
+Applying the Laplace-Beltrami normalization in the NIV context will eliminate the sampling density effects.
+%
+An alternative approach would be to estimate directly the drift term for each intrinsic variable by finding the optimal transformation that would
+align the common regions (in the spirit of \cite{...}).
 
@@
     \label{fig:rxn_embedding}
+\end{figure*}
+
+To determine whether a sufficient amount of data is available to obtain an accurate embedding, we compute the NIV embeddings for different numbers of data points 
+(arising from different numbers of trajectories initialized randomly in the full concentration space).
+%
+Figure \ref{fig:rxn_convergence} shows the two-dimensional NIV embeddings and corresponding eigenvalue spectra of the kernel for the Gillespie SSA example computed from 10, 100, 500, and 1500 data points.
+%
+We observe that the eigenvalue spectra appear converged for 100 data points. 
+%
+This implies that a sufficient amount of data is available merely to discover the intrinsic variables.
+%
+However, we observe that 100 points are insufficient to construct a consistent embedding.
+%
+In order to account for the different scaling of the NIV coordinate axes, we require at least 500 data points.
+%
+This last convergence is exemplified in Figure \ref{fig:rxn_convergence} (c) and (d).
+%
+The convergence of the spectrum of the kernel and the consistency of the NIV embedding is an empirical indicator that a sufficient amount of data is available.
+\begin{figure*}[ht]
+    \includegraphics[width=6in]{make_figures4b.ps}
+    \caption{Two-dimensional NIV embeddings (top) and corresponding eigenvalue spectra (bottom) calculated from subsets of $\mathcal{Y}_1$ containing (a) 10 data points, (b) 100 data points, (c) 500 data points, and (d) 1500 data points.
+One can see that the embedding and spectrum in (a) is inconsistent with the other embeddings.
+}
+    \label{fig:rxn_convergence}
 \end{figure*}
@@
 LP arguably outperforms simple nearest neighbor search for all of the atoms.
+%
+To determine why LP outperforms nearest neighbor search, we look at the errors at different scales in the LP algorithm.
+%
+Figure \ref{fig:LapPyr_ala2_errors} shows the true and reconstructed positions, as well as the error, for the z-coordinate of atom 10 as a function of $\psi_1$ at Scale 8 in the LP algorithm.
+%
+One can see that the error is larger in some regions (when $\psi_1$ is large) and smaller in other regions (when $\psi_1$
 
@@
     \label{fig:ala_mse}
+\end{figure*}
+
+\begin{figure*}[ht]
+    \centering
+    \includegraphics[width=3in]{fig10b}
+    \caption{(a) True position (in blue) and reconstructed position (in black) of the z-coordinate of atom 10 as a function of $\psi_1$.
+(b) The error in the reconstructed position of the z-coordinate of atom 10 as a function of $\psi_1$.
+One can see that the error is larger when $\psi_1$ is large, demonstrating that our data is indeed multiscale and that Laplacian Pyramids is an appropriate interpolation algorithm to use.}
+    \label{fig:LapPyr_ala2_errors}
 \end{figure*}

### 2013-12-27 paper/revision3.tex
Added files related to paper

--- 2013-12-27 paper/draft9.tex
@@
 %
-These developments have the potential to significantly enable the computational exploration
+Often, high-dimensional data do not uniformly fill the entire ambient space, but rather lie on a lower-dimensional manifold.
+%
+Discovering a parameterizations of the manifold therefore yields a compact representation of the data, and may lead to efficient analysis and processing schemes.
+%
+In particular, these developments have the potential to significantly enable the computational exploration
 of physicochemical problems.
+%
+Traditionally, state reduction in chemical reaction networks has been done using some {\em a priori} knowledge about separation of time scales,
+for example, using the quasi-steady state approximation to reduce a network of chemical reactions\cite{bowen1963singular}.
+%
+Alternatively, empirical knowledge about the system of interest has been used to develop coarse grained descriptions,
+such as grouping atoms of a macromolecule into larger ``beads'' to accelerate simulations\cite{monticelli2008martini, spiga2013electrostatic,izvekov2005systematic, saunders2013coarse}.
+%
+The use of manifold learning allows for the systematic discovery of data-driven, low-dimensional descriptions that circumvent the need for such {\em a priori} knowledge of the system.
 %
@@
 \end{algorithm}
+
+\subsection{Limiting Diffusion Operator} \label{sec:diff_limit}
+Assume that the dynamics of the underlying variables in \eqref{eq:dynamics} can be rewritten as
+\begin{equation} \label{eq:Langevin}
+	d\mathbf{x}(t) = - \nabla U(\mathbf{x}(t)) dt + d \mathbf{w}(t),
+\end{equation}
+where the drift term is expressed as a gradient of a potential field $U$ and $\dot{\mathbf{w}}(t) \in \mathbb{R}^d$ is a vector consisting of independent white noises.
+%
+Nadler {\em et al.} \cite{nadler2006diffusion} showed that in the limit $\varepsilon \rightarrow 0$ and $T \rightarrow \infty$ the kernel $\mathbf{W}^{\mathrm{norm}}$ converges to the continuous backward Fokker-Planck operator
+\begin{equation}
+	\mathcal{L} = \Delta - \nabla U \cdot \nabla.
+	\label{eq:limiting_operator}
+\end{equation}
+%where $g:\mathcal{M}_x \rightarrow \mathbb{R}$ is any real function on the domain of the underlying variables.
+%Since we assume independent underlying variables, the potential is additive \cite{singer2008non}
+%\begin{equation}
+%	U(\mathbf{x}(t)) = \sum \limits _{i=1}^d U_i(x_i(t)).
+%\end{equation}
+We remark that, while in diffusion maps \cite{coifman2005geometric}, the limiting operator is the Laplacian on the manifold on which the observations lie,
+here the Laplacian operator on the data that is based on the Mahalanobis distance converges to a limiting Laplacian operator on the
+manifold of the intrinsic variables $\mathbf{x}$.
+
+The convergence to the limiting operator implies that the eigenvectors of $\mathbf{W}^{\mathrm{norm}}$, which are used to represent the NIV, approximate the eigenfunctions of the limiting Fokker-Planck operator on the NIV manifold.
+%
+As such, according to \eqref{eq:limiting_operator}, they depend on the potential field.
+%
+Thus, the constructed embedding is consistent as long as the subsets of observations $\mathbf{Y}(t)$ are
+adequate samples of the same (equilibrium) density.
+%
+For example, when looking at multiple data sets from a molecular simulation, the embeddings will be consistent provided each simulation sufficiently samples the same underling free energy surface.
+
+In order to merge observations from different regions,
+the fundamental question is, to what extent the respective sampling densities affect the coordinates we obtain.
+%
+In principle, we can decouple diffusion dynamics (and thus, equilibrium densities)
+from geometry by normalizing the diffusion kernel with the sample density.
+%
+This has been explored for diffusion maps independently of NIV \cite{coifman2005geometric}.
+%
+When the normalized kernel is defined as 
+\begin{equation} \label{eq:norm_kernel}
+\widetilde{\mathbf{W}}^{\mathrm{norm}} = \mathbf{D}^{-1}\mathbf{W}\mathbf{D}^{-1},
+\end{equation}
+the limiting diffusion operator is the Laplace-Beltrami operator \cite{nadler2006diffusion}, given by $\widetilde{\mathcal{L}} = \Delta $.
+%
+In other words, this diffusion captures only the geometry of the observations and is invariant to the potential field (or, alternatively,
+to the details of the sampling density).
+%
+%Therefore, this normalization separates the geometry of the underlying manifold of intrinsic variables from the statistics of the observations.
+
+Since local coordinates are based on the noise term in \eqref{eq:Langevin}, sets of measurements from different regions
+will essentially detect the same system of coordinates (the same variable ``axes"), yet different sampling densities (arising, for example,
+from the drift term in \eqref{eq:Langevin}) will result in different {\em scalings} of these axes.
+%
+Applying the Laplace-Beltrami normalization in the NIV context will thus eliminate sampling density effects.
+%
+An alternative approach would find the scaling transformations that optimally ``register"
+data sampled from common regions (in the spirit of histogram reweighting \cite{ferrenberg1988new}).
 
@@
 
+To determine whether a sufficient amount of data is available to obtain an accurate embedding, we compute the NIV embeddings for different numbers of data points 
+(arising from different numbers of trajectories initialized randomly in the full concentration space).
+%
+Figure \ref{fig:rxn_convergence} shows the two-dimensional NIV embeddings and corresponding eigenvalue spectra of the kernel for the Gillespie SSA example computed from 10, 100, 500, and 1500 data points.
+%
+We observe that the eigenvalue spectra appear converged for 100 data points and above. 
+%
+This implies that a sufficient amount of data is available merely to discover the intrinsic variables. 
+%
+However, we observe that 100 points are insufficient to construct a self-consistent embedding {\em in the two variables discovered}. 
+%
+In order to account for the different scaling of the NIV coordinate axes, we require at least 500 data points. 
+%
+This last convergence is exemplified in Figure \ref{fig:rxn_convergence} (c) and (d).
+%
+The convergence of the spectrum of the kernel and the ultimate self-consistency of the NIV embedding is our empirical indicator that a sufficient amount of data is available. 
+\begin{figure*}[ht]
+    \includegraphics[width=6in]{fig4b}
+    \caption{Two-dimensional NIV embeddings (top) and corresponding eigenvalue spectra (bottom) calculated from subsets of $\mathcal{Y}_1$ containing (a) 10 data points, (b) 100 data points, (c) 500 data points, and (d) 1500 data points.
+One can see that the embedding and spectrum in (a) is inconsistent with the other embeddings.
+}
+    \label{fig:rxn_convergence}
+\end{figure*}
+
 We then use NIV together with Laplacian Pyramids to estimate the values of $S:E$ and $F:E^{*}$ for $\mathcal{Y}_2$.
@@
 there is qualitative agreement between the true and reconstructed configurations.
+%
+The discrepancies between the true and reconstructed positions of atoms 9 and 10 (see Figure \ref{fig:ala_fes} for atom indexing) for some of the molecular structures is not surprising.
+%
+From Figure \ref{fig:ala_embed}, one can see that the first three NIV describe the flipping of atoms 1 and 3 and the dihedral angles $\phi$ and $\psi$.
+%
+However, atoms 9 and 10 do not participate in any of these physical quantities, and so there is little information about their positions contained in the first three NIV.
+%
+We suspect that using more NIV would result in a better description of atoms 9 and 10 and lead to more accurrate structure reconstructions.
 
@@
 LP arguably outperforms simple nearest neighbor search for all of the atoms.
+%
+To determine why LP outperforms nearest neighbor search, we look at the errors at different scales in the LP algorithm.
+%
+The example described in Figure \ref{fig:LP_ex} demonstrates the appropriateness of the LP algorithm to signals with an intrinsic multiscale structure. 
+%
+We performed a similar analysis for the data from the alanine dipeptide example, and showed that the signal exhibits a similar error structure, with large errors in some regions of NIV space and small errors in other regions; these results are shown in Figure \ref{fig:LapPyr_ala2_errors}.
+%
+We therefore concluded that the alanine dipeptide data does contain a measure of multiscale behavior, and, therefore, that LP is an appropriate algorithm for reconstruction.
 
@@
     \label{fig:ala_mse}
+\end{figure*}
+
+\begin{figure*}[ht]
+    \centering
+    \includegraphics[width=3in]{fig10b}
+    \caption{The error in the reconstructed position of the z-coordinate of atom 10 as a function of $\psi_1$.
+One can see that the error is larger when $\psi_1$ is large, demonstrating that our data is indeed multiscale (similar to Figure \ref{fig:LP_ex} and that Laplacian Pyramids is an appropriate interpolation algorithm to use.}
+    \label{fig:LapPyr_ala2_errors}
 \end{figure*}

### 2013-12-27 paper/LapPyr_scales/LP_for_review.m

@@ -1,0 +1,17 @@
+
+
+%%%  plots in the manner of the LP example for one of the 30 coordinates (this is the last one, z-cord of atom #10)
+
+load train_function_paper_test;
+load  Train_EstP_paper_test;
+load  deltas_paper_test;
+
+ for i=1:11
+    figure; plot(train_function_paper_test(:,30), 'b')
+    hold on;
+    t = Train_EstP_paper_test{i};
+    plot(t(:,30), 'k');
+    d = deltas_paper_test{i};
+    figure; plot(d(:,30), 'r');
+    axis([0 4500 -0.5 0.5]);
+end;

### 2013-12-27 paper/LapPyr_scales/LP_for_review_vs_NIV.m

@@ -1,0 +1,38 @@
+fmt = '-depsc';
+res = '-r600';
+set(0,'DefaultAxesFontSize',6)
+set(0,'DefaultFigurePaperUnits','inches')
+
+%%%  plots in the manner of the LP example for one of the 30 coordinates (this is the last one, z-cord of atom #10)
+
+load train_function_paper_test;
+load  Train_EstP_paper_test;
+load  deltas_paper_test;
+load train_set_paper_test;
+
+%for i=1:11
+for i=8
+    figure;
+    set(gcf,'paperposition',[0 0 3 2.5])
+    plot(train_set_paper_test(:,1),train_function_paper_test(:,30), '.b')
+    hold on;
+    t = Train_EstP_paper_test{i};
+    plot(train_set_paper_test(:,1),t(:,30), '.k');
+    xlabel('\psi_1')
+    ylabel('f(x)')
+    title(sprintf('Scale %d',i))
+    saveas(gcf,sprintf('scale_recon%d',i),'fig')
+    print(sprintf('scale_recon%d',i), fmt,res)
+    
+    
+    d = deltas_paper_test{i};
+    figure;
+    set(gcf,'paperposition',[0 0 3 2.5])
+    plot(train_set_paper_test(:,1),d(:,30), '.r');
+    axis([-300 100 -0.5 0.5]);
+    xlabel('\psi_1')
+    ylabel('f^{LP}(x)-f(x)')
+    title(sprintf('Scale %d',i))
+    saveas(gcf,sprintf('scale_error%d',i),'fig')
+    print(sprintf('scale_error%d',i), fmt,res)
+end;

### 2013-12-27 paper/ala_figures_allodd/LP.m

@@ -1,0 +1,101 @@
+
+
+
+
+function [f_app_Train_Data, f_app_Test_Data, deltas, my_err, my_err_max, correct_app] = LP(Train_Data, Test_Data, f, my_err_thresh,  my_eps)
+
+
+SzTR = size(Train_Data,1);
+SzTS = size(Test_Data,1);
+
+my_k = size(f,2);
+
+F_TRAIN = f(1:SzTR,:);
+F_TO_APP = F_TRAIN;
+F_KER = zeros(SzTR+SzTS,1);
+
+
+
+[APP_RES_TR, APP_RES_TS] = LP_iter(Train_Data, Test_Data, F_TO_APP, my_eps);
+
+
+%%
+MyErrZeroInd = zeros(SzTR,1);
+correct_app = zeros(SzTR,1);
+for ER=1:SzTR
+    MyErr(ER,1) = sum(abs(F_TO_APP(ER,:) - APP_RES_TR(ER,:)).^2)./my_k;
+    if(MyErr(ER,1) < my_err_thresh)
+        correct_app(ER) = 1;
+    end;
+end;
+
+my_err(:,1) = mean(MyErr(:,1));
+my_err_max(:,1) = max(MyErr(:,1));
+
+
+f_app_Train_Data{1} = APP_RES_TR;
+f_app_Test_Data{1} = APP_RES_TS;
+
+
+F_TO_APP = F_TO_APP-APP_RES_TR;
+F_KER = [f_app_Train_Data{1};f_app_Test_Data{1}];
+
+deltas{1} = F_TO_APP;
+
+
+
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+
+i=1;
+
+
+while(my_err_max(:,i) > my_err_thresh) 
+
+    
+ i = i+1;    
+ 
+
+my_eps = my_eps./2;
+
+[APP_RES_TR, APP_RES_TS] = LP_iter(Train_Data, Test_Data, F_TO_APP, my_eps);
+
+
+
+for ER=1:SzTR
+    MyErr(ER,i) = sum(abs(F_TO_APP(ER,:) - APP_RES_TR(ER,:)).^2)./my_k;
+    if(MyErr(ER,i) < my_err_thresh)
+        if(correct_app(ER)==0)
+            correct_app(ER) = i;
+        end;
+    end;
+end;
+
[…]

### 2013-12-27 paper/ala_figures_allodd/LP_iter_chm.m

@@ -1,0 +1,42 @@
+function [f_app_Train_Data, f_app_Test_Data] =LP_iter_chm(Train_Data, Test_Data, f, my_eps)
+
+
+Sz_Train = size(Train_Data,1);
+Sz_Test = size(Test_Data,1);
+f_SZ = size(f,2);
+
+dist1 = squareform(pdist([Train_Data;Test_Data]));
+
+D = (dist1.^2)./my_eps; 
+ker = exp(-D);
+
+
+ker = ker(:,1:Sz_Train);
+
+
+fz = find(ker < 10.^-10);
+ker(fz)=0;
+
+
+
+for i=1:size(ker,1)
+    if(sum(ker(i,:)) > 0)
+      ker(i,:) = ker(i,:) / sum(ker(i,:));
+    end;
+end;
+
+
+
+
+f_app_Data = zeros(Sz_Train+Sz_Test,f_SZ);
+
+for i=1:Sz_Train+Sz_Test
+    for j=1:Sz_Train
+         f_app_Data(i,:) = f_app_Data(i,:) + ker(i,j)*f(j,:);
+    end;
+end;
+
+
+
+f_app_Train_Data = f_app_Data(1:Sz_Train,:);
+f_app_Test_Data = f_app_Data(Sz_Train+1:end,:);

### 2013-12-27 paper/ala_figures_allodd/ala_figures_allodd.m

@@ -1,0 +1,240 @@
+clear all 
+close all
+
+fmt = '-djpeg';
+res = '-r600';
+set(0,'DefaultAxesFontSize',6)
+set(0,'DefaultFigurePaperUnits','inches')
+
+%% Load Data
+%load '~/resesarch/Alanine/Short_Trajs_Cartesian_Coord.dat'
+%load '~/resesarch/Alanine/Short_Trajs_dihedrals.dat'
+load '../../Ronen/Short_Trajs_Cartesian_Coord.dat'
+load '../../Ronen/Short_Trajs_dihedrals.dat'
+
+time_indices1 = 50001:60000;
+time_indices2 = 58001:70000;
+num_overlap = 2000;
+
+coor_indices1 = 2:31; % all
+coor_indices2 = [2,3,4,8,9,10,14,15,16,20,21,22,26,27,28]; % only odd
+
+%% Odd Coordinates
+data = Short_Trajs_Cartesian_Coord(time_indices1, coor_indices1);
+    
+% Compute local covariances
+covariances;
+    
+% NLICA
+NLICA;
+psi_mat_nlica1 = psi_mat;
+
+% DM
+DM;   
+psi_mat_dm1 = psi_mat;
+
+%% Even Coordinates  
+data = Short_Trajs_Cartesian_Coord(time_indices2, coor_indices2);
+    
+% Compute local covariances
+covariances;
+    
+% NLICA
+NLICA;
+psi_mat_nlica2 = psi_mat;
+
+% DM
+DM;   
+psi_mat_dm2 = psi_mat;
+
+%% align two embeddings
+for i=1:3
+    psi_mat_nlica1(:,i) = psi_mat_nlica1(:,i) / norm(psi_mat_nlica1(:,i)) * size(psi_mat_nlica1,1);
+    psi_mat_nlica2(:,i) = psi_mat_nlica2(:,i) / norm(psi_mat_nlica2(:,i)) * size(psi_mat_nlica2,1);
+    corr1 = sum(psi_mat_nlica1(end-num_overlap+1:end,i).*psi_mat_nlica2(1:num_overlap,i)) / sum(psi_mat_nlica1(end-num_overlap+1:end,i).^2);
+    if corr1 < 0
+        psi_mat_nlica2(:,i) = -psi_mat_nlica2(:,i);
+    end
+    psi_mat_dm1(:,i) = psi_mat_dm1(:,i) / norm(psi_mat_dm1(:,i)) * size(psi_mat_dm1,1);
[…]

### 2013-12-27 paper/ala_figures_allodd/covariances.m

@@ -1,0 +1,25 @@
+
+
+% Find neighborhoods
+M = size(data, 1);
+N = size(data, 2);
+knn = 15;
+
+% compute covariances
+inv_c_noisy = zeros(N, N, M);
+inv_c = zeros(N, N, M);
+D=2;
+
+for i = 1:M
+    lind = max(i-knn, 1+100*floor((i-1)/100));
+    rind = min(i+knn, 100*ceil(i/100));
+    neighbors = data(lind:rind,:);
+
+    c = cov(neighbors);
+    inv_c_noisy(:,:,i) = pinv(c);
+
+    [U S V] = svd(c);
+    c_hat = U(:,1:D) * S(1:D,1:D) * V(:,1:D)';
+    inv_c(:,:,i) = pinv(c_hat);
+end
+

### 2013-12-27 paper/ala_figures_evenodd/NLICA.m

@@ -1,0 +1,81 @@
+
+
+%inv_c = inv_c_noisy;
+
+%
+% Choose subset of examples
+%
+M = size(data, 1);
+m = 5000;
+subidx = 1:m;
+dataref = data(subidx, :);
+
+% 
+% Anisotropic Kernel
+% 
+% Dis = zeros(M, m);
+% h = waitbar(0, 'Please wait');
+% for i = 1:M
+%     waitbar(i/M, h);
+%     for j = 1:m
+%          Dis(i,j) = [data(i,:) - dataref(j,:)] * inv_c(:,:,subidx(j)) * [data(i,:) - dataref(j,:)]';
+%     end
+% end
+% close(h);
+
+Dis = zeros(M, m);
+h = waitbar(0, 'Please wait');
+for j = 1:m
+    waitbar(j/m, h);
+    tmp1 = inv_c(:,:,subidx(j)) * dataref(j,:)';
+
+    a2 = dataref(j,:) * tmp1;
+    b2 = sum(data .* (inv_c(:,:,subidx(j)) * data')',2);
+    ab = data * tmp1;
+    Dis(:,j) = repmat(a2, M, 1) + b2 - 2*ab;
+end
+close(h);
+
+%%
+%ep = 1e2;
+ep = median(median(Dis))
+
+A = exp(-Dis/(4*ep));
+W_sml=A'*A;    
+d1=sum(W_sml,1);
+A1=A./repmat(sqrt(d1),M,1);
+W1=A1'*A1;
+
+d2=sum(W1,1);
+A2=A1./repmat(sqrt(d2),M,1);
+W2=A2'*A2;
+
+D=diag(sqrt(1./d2));
+
+% computing eigenvectors:
+[V,E] = eigs(W2,10);
+diag(E(2:end,2:end))
+[srtdE,IE] = sort(sum(E),'descend');
+V_srt_clds = D*V(:,IE(1,2:10));
+
+Phi_1 = V_srt_clds(:, 1);
+Phi_2 = V_srt_clds(:, 2);
+
+% NOW THE EXTENSION:
+psi_mat=[];
+
+omega=sum(A2,2);
+A2_nrm=A2./repmat(omega,1,m);
+
+for i=1:size(V_srt_clds,2)
+    psi=A2_nrm*V_srt_clds(:,i)./sqrt((srtdE(i+1)));
+    psi_mat=[psi_mat,psi];
+end
+
+%Psi_1 = psi_mat(:,1);
[…]

### 2013-12-27 paper/ala_figures_evenodd/ala_figures_evenodd.m

@@ -1,0 +1,99 @@
+clear all 
+close all
+
+fmt = '-djpeg';
+res = '-r600';
+set(0,'DefaultAxesFontSize',6)
+set(0,'DefaultFigurePaperUnits','inches')
+
+%% Load Data
+%load '~/resesarch/Alanine/Short_Trajs_Cartesian_Coord.dat'
+%load '~/resesarch/Alanine/Short_Trajs_dihedrals.dat'
+load '../../Ronen/Short_Trajs_Cartesian_Coord.dat'
+load '../../Ronen/Short_Trajs_dihedrals.dat'
+
+time_indices = 10001:20000;
+
+coor_indices2 = [14,15,16,20,21,22,26,27,28]; % only odd
+coor_indices3 = [5,6,7,11,12,13,17,18,19,23,24,25,29,30,31]; % only even
+
+%% Odd Coordinates
+data = Short_Trajs_Cartesian_Coord(time_indices, coor_indices2);
+    
+% Compute local covariances
+covariances;
+    
+% NLICA
+NLICA;
+for i=1:size(psi_mat,2)
+    psi_mat(:,i) = sign(mean(psi_mat(:,i))) * psi_mat(:,i) / norm(psi_mat(:,i)) * size(psi_mat,1);
+end
+psi_mat_nlica_odd = psi_mat;
+Psi1_nlica_odd = Psi_1;
+Psi2_nlica_odd = Psi_2;
+
+% DM
+DM;   
+for i=1:size(psi_mat,2)
+    psi_mat(:,i) = sign(mean(psi_mat(:,i))) * psi_mat(:,i) / norm(psi_mat(:,i)) * size(psi_mat,1);
+end
+psi_mat_dm_odd = psi_mat;
+Psi1_dm_odd = Psi_1;
+Psi2_dm_odd = Psi_2;
+
+%% Even Coordinates  
+data = Short_Trajs_Cartesian_Coord(time_indices, coor_indices3);
+    
+% Compute local covariances
+covariances;
+    
+% NLICA
+NLICA;
+for i=1:size(psi_mat,2)
+    psi_mat(:,i) = sign(mean(psi_mat(:,i))) * psi_mat(:,i) / norm(psi_mat(:,i)) * size(psi_mat,1);
+end
+psi_mat_nlica_even = psi_mat;
+Psi1_nlica_even = Psi_1;
[…]

### 2013-12-27 paper/molec_figures/make_molec_pictures.m

@@ -1,0 +1,44 @@
+clear all
+close all
+
+%fmt = '-djpeg';
+fmt = '-depsc';
+res = '';
+%res = '-r600';
+set(0,'DefaultAxesFontSize',6)
+set(0,'DefaultFigurePaperUnits','inches')
+
+load ../ala_figures_allodd/LapPyr_recon.mat;
+
+ala2 = pdbread('ala2_noH.pdb');
+
+test_ind = 900;
+
+true_data = reshape(true_pos(test_ind,:),3,[])';
+recon_data = reshape(recon_pos(test_ind,:),3,[])';
+
+for i=1:10
+    ala2.Model.Atom(1,i).X = true_data(i,1);
+    ala2.Model.Atom(1,i).Y = true_data(i,2);
+    ala2.Model.Atom(1,i).Z = true_data(i,3);
+    ala2.Model.Atom(1,i+10).X = recon_data(i,1);
+    ala2.Model.Atom(1,i+10).Y = recon_data(i,2);
+    ala2.Model.Atom(1,i+10).Z = recon_data(i,3);
+end
+
+h = molviewer(ala2);
+evalrasmolscript(h,'rotate Y 135');
+evalrasmolscript(h,'background white');
+
+evalrasmolscript(h,'select *U; color black');
+evalrasmolscript(h,'select *X; color red');
+evalrasmolscript(h,'select all');
+
+%saveas(gcf,sprintf('molecule%d_balls',test_ind),'fig')
+print(sprintf('molecule%d_balls',test_ind),fmt,res)
+
+%set(gcf,'paperposition',[0 0 2 2])
+evalrasmolscript(h,'wireframe 10');
+
+%saveas(gcf,sprintf('molecule%d_wire',test_ind),'fig')
+print(sprintf('molecule%d_wire',test_ind),fmt,res)

### 2013-12-27 paper/rxn_figures/NL_ICA2.m

@@ -1,0 +1,45 @@
+function [psi_mat, E] = NL_ICA2(data, inv_c, ep)
+
+[m, n] = size(data);
+
+Dis = zeros(m);
+h = waitbar(0, 'Please wait');
+for i = 1:m
+    waitbar(i/m, h);
+    for j = 1:m
+        Dis(i,j) = (data(i,:) - data(j,:)) * inv_c(:,:,j) * (data(i,:) - data(j,:))';
+    end
+end
+close(h);
+
+%ep = median(median(Dis));
+
+%%
+A = exp(-Dis/(4*ep));
+W_sml=A'*A;    
+d1=sum(W_sml,1);
+A1=A./repmat(sqrt(d1),m,1);
+W1=A1'*A1;
+
+d2=sum(W1,1);
+A2=A1./repmat(sqrt(d2),m,1);
+W2=A2'*A2;
+
+D=diag(sqrt(1./d2));
+
+% computing eigenvectors:
+[V,E] = eigs(W2,10);
+E(2:end,2:end)
+[srtdE,IE] = sort(sum(E),'descend');
+V_srt_clds = D*V(:,IE(1,2:10));
+
+% NOW THE EXTENSION:
+psi_mat=[];
+
+omega=sum(A2,2);
+A2_nrm=A2./repmat(omega,1,m);
+
+for i=1:size(V_srt_clds,2)
+    psi=A2_nrm*V_srt_clds(:,i)./sqrt((srtdE(i+1)));
+    psi_mat=[psi_mat,psi];
+end

### 2013-12-27 paper/rxn_figures/calc_covar.m

@@ -1,0 +1,22 @@
+function inv_c = calc_covar(data_bursts, D)
+
+% Use end of snippets
+[~, N, M] = size(data_bursts);
+
+% compute covariances
+%inv_c_noisy = zeros(N, N, M);
+inv_c = zeros(N, N, M);
+
+%spec = zeros(N,M);
+for i = 1:M
+
+    c = cov(data_bursts(:,:,i));
+    %inv_c_noisy(:,:,i) = pinv(c);
+
+    [U S V] = svd(c);
+    c_hat = U(:,1:D) * S(1:D,1:D) * V(:,1:D)';
+    inv_c(:,:,i) = pinv(c_hat);
+    
+    %spec(:,i) = diag(S);
+end
+

### 2013-12-27 paper/rxn_figures/calc_covar2.m

@@ -1,0 +1,22 @@
+function [inv_c, spec] = calc_covar2(data_bursts, D)
+
+% Use end of snippets
+[~, N, M] = size(data_bursts);
+
+% compute covariances
+%inv_c_noisy = zeros(N, N, M);
+inv_c = zeros(N, N, M);
+
+spec = zeros(N,M);
+for i = 1:M
+
+    c = cov(data_bursts(:,:,i));
+    %inv_c_noisy(:,:,i) = pinv(c);
+
+    [U S V] = svd(c);
+    c_hat = U(:,1:D) * S(1:D,1:D) * V(:,1:D)';
+    inv_c(:,:,i) = pinv(c_hat);
+    
+    spec(:,i) = diag(S);
+end
+

### 2013-12-27 paper/rxn_figures/rxn_figures.m

@@ -1,0 +1,154 @@
+clear all
+close all
+
+fmt = '-djpeg';
+res = '-r600';
+set(0,'DefaultAxesFontSize',6)
+set(0,'DefaultFigurePaperUnits','inches')
+
+%% make figures for rxn example
+
+species = {'S','E','E:S','S:E','D"S*','F:E*'};
+
+% load data
+load('../../rxn_example/snippet_data.mat','n', 'data');
+
+% define data
+N = size(data, 1);
+train_ind = 1:1500;
+test_ind = 1000:N;
+[test_ind2, idx2] = setdiff(test_ind, train_ind);
+[~, train_overlap_ind, test_overlap_ind] = intersect(train_ind, test_ind);
+
+partial_ind = [1,2,4,6];
+partial_ind2 = [1,2,3,5];
+
+D = 2;
+eps = 1e7;
+
+% plot original data
+az = [-5+180,-25+180,85,170];
+el = [15,10,30,10];
+az2 = [-10,-15,130,150];
+el2 = [-55,-55,-35,-40];
+
+for i=3:6
+    figure;
+    set(gcf,'paperposition',[0 0 3 2.5])
+    scatter3(data(:,1),data(:,2),data(:,i),100,data(:,i),'.')
+    xlabel(species{1})
+    ylabel(species{2})
+    zlabel(species{i})
+    grid on
+    view(az(i-2),el(i-2))
+    saveas(gcf,sprintf('rxn_manifold%d',i-2),'fig')
+    print(sprintf('rxn_manifold%d',i-2), fmt,res)
+    
+    set(gcf,'paperposition',[0 0 2 2])
+    view(az2(i-2),el2(i-2))
+    saveas(gcf,sprintf('rxn_manifold%d_2',i-2),'fig')
+    print(sprintf('rxn_manifold%d_2',i-2), fmt,res)
+end
+
+% compute embeddings
+inv_c = calc_covar(n(:,partial_ind,train_ind), D);
+psi_mat = NL_ICA(data(train_ind,partial_ind), inv_c, eps);
+%psi_mat = DM(data(train_ind,partial_ind), eps);
+
+inv_c2 = calc_covar(n(:,partial_ind2,test_ind), D);
+psi_mat2 = NL_ICA(data(test_ind,partial_ind2), inv_c2, eps);
[…]

### 2013-12-27 paper/rxn_figures/rxn_figures_largelabels.m

@@ -1,0 +1,69 @@
+clear all
+close all
+
+fmt = '-djpeg';
+res = '-r600';
+set(0,'DefaultAxesFontSize',6)
+set(0,'DefaultFigurePaperUnits','inches')
+
+%% make figures for rxn example
+
+species = {'S','E','E:S','S:E','D:S*','F:E*'};
+
+% load data
+load('../../rxn_example/snippet_data.mat','n', 'data');
+
+% define data
+N = size(data, 1);
+train_ind = 1:1500;
+test_ind = 1000:N;
+[test_ind2, idx2] = setdiff(test_ind, train_ind);
+[~, train_overlap_ind, test_overlap_ind] = intersect(train_ind, test_ind);
+
+partial_ind = [1,2,4,6];
+partial_ind2 = [1,2,3,5];
+
+D = 2;
+eps = 1e7;
+
+% plot original data
+az = [-5+180,-25+180,85,170];
+el = [15,10,30,10];
+az2 = [-10,-15,130,150];
+el2 = [-55,-55,-35,-40];
+
+for i=3:6
+    figure;
+    set(gcf,'paperposition',[0 0 3 2.5])
+    scatter3(data(:,1),data(:,2),data(:,i),100,data(:,i),'.')
+    xlabel(species{1},'fontsize',14)
+    ylabel(species{2},'fontsize',14)
+    zlabel(species{i},'fontsize',14)
+    grid on
+    view(az(i-2),el(i-2))
+    %saveas(gcf,sprintf('rxn_manifold%d',i-2),'fig')
+    print(sprintf('rxn_manifold_largelabels_%d',i-2), fmt,res)
+    
+%     set(gcf,'paperposition',[0 0 2 2])
+%     view(az2(i-2),el2(i-2))
+%     saveas(gcf,sprintf('rxn_manifold%d_2',i-2),'fig')
+%     print(sprintf('rxn_manifold%d_2',i-2), fmt,res)
+end
+
+
+%% LapPyr
+load Test_EstC4
+figure;
+set(gcf,'paperposition',[0 0 3 2])
+plot(data(test_ind,4),Test_EstC4{end},'.')
+xlabel(sprintf('true %s',species{4}),'fontsize',14)
+ylabel(sprintf('reconstructed %s',species{4}),'fontsize',14)
+print('rxn_recon4_largelabels',fmt,res)
+
+load Test_EstC6
+figure;
+set(gcf,'paperposition',[0 0 3 2])
[…]

### 2013-12-27 paper/rxn_figures/rxn_figures_revisions.m

@@ -1,0 +1,75 @@
+clear all
+close all
+
+fmt = '-depsc';
+res = '-r600';
+set(0,'DefaultAxesFontSize',6)
+set(0,'DefaultFigurePaperUnits','inches')
+
+%% make figures for rxn example
+
+species = {'S','E','E:S','S:E','D"S*','F:E*'};
+
+% load data
+load('../../rxn_example/snippet_data.mat','n', 'data');
+
+% define data
+N = size(data, 1);
+train_ind = 1:1500;
+test_ind = 1:2000;
+
+partial_ind = [1,2,4,6];
+
+D = 2;
+eps = 1e7;
+
+npoints = [10 100 250 500 1000 1500];
+[inv_c, spec] = calc_covar2(n(:,partial_ind,1:npoints(end)), D);
+%f1 = figure;
+%set(gcf,'paperposition',[0 0 5 8])
+%f2 = figure;
+%set(gcf,'paperposition',[0 0 5 8])
+for j=1:length(npoints)
+    train_ind = 1:npoints(j);
+    
+    % compute embeddings
+    [psi_mat, D] = NL_ICA2(data(train_ind,partial_ind), inv_c(:,:,train_ind), eps);
+    
+    for i=1:size(psi_mat,2)
+        psi_mat(:,i) = sign(mean(psi_mat(:,i))) * psi_mat(:,i) / norm(psi_mat(:,i)) * size(psi_mat,1);
+    end
+    
+    if psi_mat(1,1) > 0
+        psi_mat(:,1) = -psi_mat(:,1);
+    end
+    if psi_mat(1,3) < 0
+        psi_mat(:,3) = -psi_mat(:,3);
+    end
+    
+    % plot embeddings
+    psi_coords = [1,3];
+    
+    figure;
+    set(gcf,'paperposition',[0 0 1.5 1.25])
+    scatter(psi_mat(:,psi_coords(1)),psi_mat(:,psi_coords(2)),50,data(train_ind,1),'.')
+    xlabel(sprintf('\\psi_%d',psi_coords(1)))
+    ylabel(sprintf('\\psi_%d',psi_coords(2)))
+    %title(sprintf('n = %d',npoints(j)))
+    saveas(gcf,sprintf('rxn_npoints_embed%d',j),'fig')
+    print(sprintf('rxn_npoints_embed%d',j), fmt,res)
+    
+    figure;
[…]

### 2014-03-03 reducible_SDE/notes1.tex
Added writeup about fast-slow brownian motion example

\documentclass[12pt]{article}

\usepackage{amsmath}
\usepackage{amssymb}

\usepackage{graphicx}

\begin{document}

Goal: Want to examine the effects of the scale of the noise on NIV

Assume we have the following underlying dynamical system (two--dimensional Brownian motion)
\begin{eqnarray}
	dx_1 & = & dW_1 \\
	dx_2 & = & dW_2
\end{eqnarray}
where $W_1, W_2$ are independent Browinain motions, $x_1, x_2 \in [0, 1]$, and we impose reflective boundary conditions.

We then apply following nonlinear transformation
%
\begin{eqnarray}
y_1 & = & x_1 + x_2^3 \\
y_2 & = & x_2 - x_1^3
\end{eqnarray}
%
The data are shown in Figure \ref{fig:data}

\begin{figure}[htb]
\includegraphics[width=0.5\textwidth]{xdata}
\includegraphics[width=0.5\textwidth]{ydata}
\caption{Left: data in ``true'' space. Right: data in ``ambient'' space.}
\label{fig:data}
\end{figure}

The results of doing DMAPS with the data $y_1, y_2$ are shown in in Figure \ref{fig:xdata_dmaps}.
%
Note that we do not recover the ``correct'' parameterization of the underlying variables $x_1, x_2$.

\begin{figure}[htb]
\includegraphics[width=0.5\textwidth]{xdata_colored_DMAPS1}
\includegraphics[width=0.5\textwidth]{xdata_colored_DMAPS2}
\caption{Data in ``true'' space ($x_1, x_2$), colored by the first (left) and second (right) nontrivial DMAPS component. The DMAPS embedding is computed from the data in the ``ambient'' space ($y_1, y_2$). Note that the parameterizationa we obtain are not the eigenfunctions that we expect for points sampled from the unit square.}
\label{fig:xdata_dmaps}
\end{figure}

If we do NIV with the data $y_1, y_2$, 
%taking our time windows so that the clouds around the points have standard deviation $1e-3$, 
we obtain the results in Figure \ref{fig:xdata_NIV}.
%
Note that we now recover the ``correct'' variables.

\begin{figure}[htb]
\includegraphics[width=0.5\textwidth]{xdata_colored_NIV1}
\includegraphics[width=0.5\textwidth]{xdata_colored_NIV2}
\caption{Data in ``true'' space ($x_1, x_2$), colored by the first (left) and second (right) nontrivial NIV. The NIV embedding is computed from the data in the ambient space ($y_1, y_2$), with clouds around the points have standard deviation $1e-3$. Note that the parameterizations we obtain are the eigenfunctions that we expect for points sampled from the unit square.}
\label{fig:xdata_NIV}
\end{figure}

We now look at the effect of adding additional noise to our system.
%
Consider adding white Gaussian noise to the $y_1, y_2$ data
\begin{eqnarray}
z_1 & = & x_1 + x_2^3 + \xi_1 \\
z_2 & = & x_2 - x_1^3 + \xi_2 \\
z_3 & = & \xi_3
\end{eqnarray}
where $\xi_1, \xi_2, \xi_3 \sim \mathcal{N}(0, \sigma^2)$

If $\sigma = 0.01$ (so the added noise is much smaller than the noise in the original clouds), we obtain a parameterization shown in Figure \ref{fig:xdata_NIV_noise1} using NIV.

\begin{figure}[htb]
\includegraphics[width=0.5\textwidth]{xdata_noise1_colored_NIV1}
\includegraphics[width=0.5\textwidth]{xdata_noise1_colored_NIV2}
\caption{Data in ``true'' space ($x_1, x_2$), colored by the first (left) and second (right) nontrivial NIV. The NIV embedding is computed from the data in the ambient space with added noise ($z_1, z_2$, with $\sigma = 0.01$). Note that the parameterizations we obtain are the eigenfunctions that we expect for points sampled from the unit square.}
\label{fig:xdata_NIV_noise1}
\end{figure}

If $\sigma = 100$ (so the added noise is much larger than the noise in the original clouds), we obtain a parameterization shown in Figure \ref{fig:xdata_NIV_noise2} using NIV.

\begin{figure}[htb]
\includegraphics[width=0.5\textwidth]{xdata_noise2_colored_NIV1}
\includegraphics[width=0.5\textwidth]{xdata_noise2_colored_NIV2}
\caption{Data in ``true'' space ($x_1, x_2$), colored by the first (left) and second (right) nontrivial NIV. The NIV embedding is computed from the data in the ambient space with added noise ($z_1, z_2$, with $\sigma = 100$). Note that the parameterizations we obtain are not the eigenfunctions that we expect for points sampled from the unit square, but are now similar to the parameterizations obtained from DMAPS shown in Figure \ref{fig:xdata_dmaps}.}
\label{fig:xdata_NIV_noise2}
\end{figure}

\end{document}

### 2014-03-03 reducible_SDE/test_Amit_NLICA_example.m

@@ -4,11 +4,24 @@
 clear all
 close all
 
+res = '-r300';
+fmt = '-djpeg';
+
+rng(321);
+
 %% define parameters
+
+% number of data points
 N = 2000;
+
+% dimension
 dim = 2;
+
+% number of points in each cloud
 Nc = 200;
-dt = 0.01;
+
+% dt for clouds
+dt1 = 0.001;
 
 %% do simulations
 
@@ -19,16 +32,21 @@ y = f(x);
 
 %% plot results
 
-% plot variables, color by time
 figure;
 plot(x(:,1),x(:,2),'.')
+xlabel('x_1')
+ylabel('x_2')
+print('xdata', fmt, res)
 
 figure;
 plot(y(:,1),y(:,2),'.')
+xlabel('y_1')
+ylabel('y_2')
+print('ydata', fmt, res)
 
 %% calculate clouds
 
-xclouds = dt * randn(Nc, dim, N);
+xclouds = dt1 * randn(Nc, dim, N);
 for i=1:N
     xclouds(:,:,i) = repmat(x(i,:), Nc, 1) + xclouds(:,:,i);
 end
@@ -55,9 +73,15 @@ neigs = 10;
 
 figure;
 scatter(x(:,1),x(:,2),200,V(:,2),'.')
+xlabel('x_1')
+ylabel('x_2')
+print('xdata_colored_NIV1', fmt, res)
 
 figure;
 scatter(x(:,1),x(:,2),200,V(:,3),'.')
+xlabel('x_1')
+ylabel('x_2')
+print('xdata_colored_NIV2', fmt, res)
 
 %% dmaps
 W = squareform(pdist(y)).^2;
@@ -65,68 +89,75 @@ W = squareform(pdist(y)).^2;
 
 figure;
 scatter(x(:,1),x(:,2),200,V(:,2),'.')
+print('xdata_colored_DMAPS1', fmt, res)
 
 figure;
 scatter(x(:,1),x(:,2),200,V(:,3),'.')
+print('xdata_colored_DMAPS2', fmt, res)
 
 %% do simulation with noise
-
 dim = 3;
+dt2 = dt1 / 100;
 
-x = zeros(N, dim);
-x(:, 1:2) = rand(N, 2);
+z = [y zeros(N, 1)];
 
-f = @(x) [x(:,1)+x(:,2).^3 x(:,2)-x(:,1).^3 x(:,3)];
-
-y = f(x);
-
[…]

### 2014-03-04 reducible_SDE/notes1.tex
added Ronen edits to writeup

@@
 
-Goal: Want to examine the effects of the scale of the noise on NIV
+Goals: 
+\begin{itemize}
+\item Want to examine the effects of the scale of the noise on NIV
+\item Will look at systems with both underlying noise in the ``intrinsic'' variables, and added noise to the variables in the transformed/ ``ambient'' space
+\item The magnitude of the noise in the ambient space will determine the embedding we compute
+\item We expect to obtain DMAPS variables if the noise in the ambient space is large (we will only ``see'' the ambient space noise, and not the , and NIV if the noise in the ambient space is small
+\end{itemize}
 
@@
 \begin{figure}[htb]
-\includegraphics[width=0.5\textwidth]{xdata_colored_DMAPS1}
-\includegraphics[width=0.5\textwidth]{xdata_colored_DMAPS2}
-\caption{Data in ``true'' space ($x_1, x_2$), colored by the first (left) and second (right) nontrivial DMAPS component.
+\includegraphics[width=0.3\textwidth]{xdata_colored_DMAPS1}
+\includegraphics[width=0.3\textwidth]{xdata_colored_DMAPS2}
+\includegraphics[width=0.3\textwidth]{embedding_dmaps}
+\caption{Data in ``true'' space ($x_1, x_2$), colored by the first (left) and second (center) nontrivial DMAPS component.
 The DMAPS embedding is computed from the data in the ``ambient'' space ($y_1, y_2$).
-Note that the parameterizationa we obtain are not the eigenfunctions that we expect for points sampled from the unit square.}
+Note that the parameterizationa we obtain are not the eigenfunctions that we expect for points sampled from the unit square.
+(Right) Data plotted in first two (nontrivial) DMAPS variables.}
 \label{fig:xdata_dmaps}
@@
 \begin{figure}[htb]
-\includegraphics[width=0.5\textwidth]{xdata_colored_NIV1}
-\includegraphics[width=0.5\textwidth]{xdata_colored_NIV2}
-\caption{Data in ``true'' space ($x_1, x_2$), colored by the first (left) and second (right) nontrivial NIV.
-The NIV embedding is computed from the data in the ambient space ($y_1, y_2$), with clouds around the points have standard deviation $1e-3$.
-Note that the parameterizations we obtain are the eigenfunctions that we expect for points sampled from the unit square.}
+\includegraphics[width=0.3\textwidth]{xdata_colored_NIV1}
+\includegraphics[width=0.3\textwidth]{xdata_colored_NIV2}
+\includegraphics[width=0.3\textwidth]{embedding_NIV}
+\caption{Data in ``true'' space ($x_1, x_2$), colored by the first (left) and second (center) nontrivial NIV.
+The NIV embedding is computed from the data in the ambient space ($y_1, y_2$), with time window $\Delta t = 0.01$.
+Note that the parameterizations we obtain are the eigenfunctions that we expect for points sampled from the unit square.
+(Right) Data plotted in first two (nontrivial) NIV.}
 \label{fig:xdata_NIV}
@@
 If $\sigma = 0.01$ (so the added noise is much smaller than the noise in the original clouds), we obtain a parameterization shown in Figure \ref{fig:xdata_NIV_noise1} using NIV.
+%
+When we do NIV, the ``ellipses'' we see are a combination of the original noises in the intrinsic variables and the noise in the ambient space.
+%
+Because the noise in the ambient space is significantly smaller than the noise in the original intrinsic variable space, we can still ``see'' the directions of the intrinsic variable noises in the ellipses, and therefore uncover the intrinsic variables.
 
 \begin{figure}[htb]
-\includegraphics[width=0.5\textwidth]{xdata_noise1_colored_NIV1}
-\includegraphics[width=0.5\textwidth]{xdata_noise1_colored_NIV2}
-\caption{Data in ``true'' space ($x_1, x_2$), colored by the first (left) and second (right) nontrivial NIV.
+\includegraphics[width=0.3\textwidth]{xdata_noise1_colored_NIV1}
+\includegraphics[width=0.3\textwidth]{xdata_noise1_colored_NIV2}
+\includegraphics[width=0.3\textwidth]{embedding_noise1}
+\caption{Data in ``true'' space ($x_1, x_2$), colored by the first (left) and second (center) nontrivial NIV.
 The NIV embedding is computed from the data in the ambient space with added noise ($z_1, z_2$, with $\sigma = 0.01$).
-Note that the parameterizations we obtain are the eigenfunctions that we expect for points sampled from the unit square.}
+Note that the parameterizations we obtain are the eigenfunctions that we expect for points sampled from the unit square.
+(right) Data plotted in first two (nontrivial) NIV.
+We (effectively) recover the unit square, i.e., the domain of the intrinsic variables.
+}
 \label{fig:xdata_NIV_noise1}
@@
 If $\sigma = 100$ (so the added noise is much larger than the noise in the original clouds), we obtain a parameterization shown in Figure \ref{fig:xdata_NIV_noise2} using NIV.
+%
+Now, the ellipses are a combination of the intrinsic varibale noise, and the ambient space noise, but the ambient space noise is {\em much} larger. 
+%
+Therefore, NIV only collapses the ambient space noise (not the intrinsic variable noise), and so we get something like the DMAPS coordinates, rather than the NIV.
 
 \begin{figure}[htb]
-\includegraphics[width=0.5\textwidth]{xdata_noise2_colored_NIV1}
-\includegraphics[width=0.5\textwidth]{xdata_noise2_colored_NIV2}
-\caption{Data in ``true'' space ($x_1, x_2$), colored by the first (left) and second (right) nontrivial NIV.
+\includegraphics[width=0.3\textwidth]{xdata_noise2_colored_NIV1}
+\includegraphics[width=0.3\textwidth]{xdata_noise2_colored_NIV2}
+\includegraphics[width=0.3\textwidth]{embedding_noise2}
+\caption{Data in ``true'' space ($x_1, x_2$), colored by the first (left) and second (center) nontrivial NIV.
 The NIV embedding is computed from the data in the ambient space with added noise ($z_1, z_2$, with $\sigma = 100$).
-Note that the parameterizations we obtain are not the eigenfunctions that we expect for points sampled from the unit square, but are now similar to the parameterizations obtained from DMAPS shown in Figure \ref{fig:xdata_dmaps}.}
+Note that the parameterizations we obtain are not the eigenfunctions that we expect for points sampled from the unit square, but are now similar to the parameterizations obtained from DMAPS shown in Figure \ref{fig:xdata_dmaps}.
+(right) Data plotted in first two (nontrivial) NIV.
+We do {\em not} recover the unit square.}
 \label{fig:xdata_NIV_noise2}

### 2014-03-04 reducible_SDE/decouple_dmaps.m

@@ -1,0 +1,8 @@
+function V_new = decouple_dmaps(V)
+
+V_new = V;
+
+theta_psi = atan(0.5 * (V(:, 4).' * (V(:,2).*V(:,2) - V(:,3).*V(:,3)) ) / ( V(:,4).' * (V(:,2) .* V(:,3)))) / 2;
+V_new(:, 2) = V(:, 2) * cos(theta_psi) - V(:,3) * sin(theta_psi);
+V_new(:, 3) = V(:, 2) * sin(theta_psi) + V(:,3) * cos(theta_psi);
+

### 2014-03-04 reducible_SDE/test_Amit_NLICA_example.m

@@ -21,7 +21,7 @@ dim = 2;
 Nc = 200;
 
 % dt for clouds
-dt1 = 0.001;
+dt1 = 0.01;
 
 %% do simulations
 
@@ -46,7 +46,7 @@ print('ydata', fmt, res)
 
 %% calculate clouds
 
-xclouds = dt1 * randn(Nc, dim, N);
+xclouds = sqrt(dt1) * randn(Nc, dim, N);
 for i=1:N
     xclouds(:,:,i) = repmat(x(i,:), Nc, 1) + xclouds(:,:,i);
 end
@@ -62,6 +62,7 @@ end
 
 %% calculate inverse covariances
 [inv_c, ~, ranks] = covariances2(yclouds, dim);
+inv_c = inv_c * dt1;
 
 %% NIV
 if size(y,1) > 4000
@@ -69,7 +70,10 @@ if size(y,1) > 4000
     return
 end
 neigs = 10;
-[V, D] = NIV(y, inv_c, 0, neigs, 0);
+eps = 2*0.005;
+%eps = 0;
+[V, D, ~] = NIV(y, inv_c, eps, neigs, 0);
+V = decouple_dmaps(V);
 
 figure;
 scatter(x(:,1),x(:,2),200,V(:,2),'.')
@@ -83,9 +87,20 @@ xlabel('x_1')
 ylabel('x_2')
 print('xdata_colored_NIV2', fmt, res)
 
+figure;
+plot(V(:,2),V(:,3),'.')
+xlabel('\phi_2')
+ylabel('\phi_3')
+print('embedding_niv', fmt, res)
+
+figure;
+bar(-2*log(diag(D))/(pi^2 * eps/2))
+
+ 
 %% dmaps
 W = squareform(pdist(y)).^2;
 [V, D] = dmaps(W, median(W(:)), neigs);
+V = decouple_dmaps(V);
 
 figure;
 scatter(x(:,1),x(:,2),200,V(:,2),'.')
@@ -94,6 +109,15 @@ print('xdata_colored_DMAPS1', fmt, res)
 figure;
 scatter(x(:,1),x(:,2),200,V(:,3),'.')
 print('xdata_colored_DMAPS2', fmt, res)
+
+figure;
+plot(V(:,2),V(:,3),'.')
+xlabel('\phi_2')
+ylabel('\phi_3')
+print('embedding_dmaps', fmt, res)
+
+figure;
+bar(-2*log(diag(D))/(pi^2 * eps/2))
 
[…]

### 2014-03-05 reducible_SDE/notes1.tex
Updated notes

@@
 \usepackage{graphicx}
-
+\usepackage{float}
 \begin{document}
@@
 \begin{itemize}
-\item Want to examine the effects of the scale of the noise on NIV
-\item Will look at systems with both underlying noise in the ``intrinsic'' variables, and added noise to the variables in the transformed/ ``ambient'' space
-\item The magnitude of the noise in the ambient space will determine the embedding we compute
-\item We expect to obtain DMAPS variables if the noise in the ambient space is large (we will only ``see'' the ambient space noise, and not the , and NIV if the noise in the ambient space is small
+\item Want to study how the scale of the noise in a system affects NIV
+\item Will look at systems with both underlying noise in the ``intrinsic'' variables, and added noise to the ``ambient'' space
+\item The magnitude of the noise in the ambient space relative to the magnitude of the noise in the intrinsic space will determine the embedding we compute
+\item We expect to obtain DMAPS variables if the noise in the ambient space is large (we will only ``see'' the ambient space noise, and not the intrinsic space noise), and NIV if the noise in the ambient space is small (so that we can still ``see'' the intrinsic space noise)
 \end{itemize}
@@
 %
-The data are shown in Figure \ref{fig:data}
+The data are shown in Figure \ref{fig:data}.
 
@@
 \includegraphics[width=0.5\textwidth]{ydata}
-\caption{Left: data in ``true'' space. Right: data in ``ambient'' space.}
+\caption{Left: data in ``intrinsic variable'' space ($x_1, x_2$). Right: data in ``ambient'' space ($y_1, y_2$).}
 \label{fig:data}
@@
 
-The results of doing DMAPS with the data $y_1, y_2$ are shown in in Figure \ref{fig:xdata_dmaps}.
+The DMAPS embedding computed from $y_1, y_2$ are shown in in Figure \ref{fig:xdata_dmaps}.
 %
@@
 
-\begin{figure}[htb]
+\begin{figure}[H]
 \includegraphics[width=0.3\textwidth]{xdata_colored_DMAPS1}
@@
 \includegraphics[width=0.3\textwidth]{embedding_dmaps}
-\caption{Data in ``true'' space ($x_1, x_2$), colored by the first (left) and second (center) nontrivial DMAPS component.
+\caption{Data in ``intrinsic variable'' space ($x_1, x_2$), colored by the first (left) and second (center) nontrivial DMAPS component.
 The DMAPS embedding is computed from the data in the ``ambient'' space ($y_1, y_2$).
@@
 
-If we do NIV with the data $y_1, y_2$, 
-%taking our time windows so that the clouds around the points have standard deviation $1e-3$, 
-we obtain the results in Figure \ref{fig:xdata_NIV}.
+The NIV embedding computed from $y_1, y_2$ are shown in in Figure  \ref{fig:xdata_NIV}.
 %
@@
 
-\begin{figure}[htb]
+\begin{figure}[H]
 \includegraphics[width=0.3\textwidth]{xdata_colored_NIV1}
@@
 \includegraphics[width=0.3\textwidth]{embedding_NIV}
-\caption{Data in ``true'' space ($x_1, x_2$), colored by the first (left) and second (center) nontrivial NIV.
+\caption{Data in ``intrinsic variable'' space ($x_1, x_2$), colored by the first (left) and second (center) nontrivial NIV.
 The NIV embedding is computed from the data in the ambient space ($y_1, y_2$), with time window $\Delta t = 0.01$.
@@
 \end{eqnarray}
-where $\xi_1, \xi_2, \xi_3 \sim \mathcal{N}(0, \sigma^2)$
+where $\xi_1, \xi_2, \xi_3 \sim \mathcal{N}(0, \sigma^2)$.
 
@@
 %
-Because the noise in the ambient space is significantly smaller than the noise in the original intrinsic variable space, we can still ``see'' the directions of the intrinsic variable noises in the ellipses, and therefore uncover the intrinsic variables.
+Because the noise in the ambient space is significantly smaller than the noise in the intrinsic variable space, we can still ``see'' the directions of the intrinsic variable noises in the ellipses, and therefore uncover the intrinsic variables.
 
-\begin{figure}[htb]
+\begin{figure}[H]
 \includegraphics[width=0.3\textwidth]{xdata_noise1_colored_NIV1}
@@
 \includegraphics[width=0.3\textwidth]{embedding_noise1}
-\caption{Data in ``true'' space ($x_1, x_2$), colored by the first (left) and second (center) nontrivial NIV.
+\caption{Data in ``intrinsic variable'' space ($x_1, x_2$), colored by the first (left) and second (center) nontrivial NIV.
 The NIV embedding is computed from the data in the ambient space with added noise ($z_1, z_2$, with $\sigma = 0.01$).
@@
 %
-Now, the ellipses are a combination of the intrinsic varibale noise, and the ambient space noise, but the ambient space noise is {\em much} larger. 
+Now, the ellipses are a combination of the intrinsic variable noise and the ambient space noise, but the ambient space noise is {\em much} larger. 
 %
@@
 
-\begin{figure}[htb]
+\begin{figure}[H]
 \includegraphics[width=0.3\textwidth]{xdata_noise2_colored_NIV1}
@@
 \includegraphics[width=0.3\textwidth]{embedding_noise2}
-\caption{Data in ``true'' space ($x_1, x_2$), colored by the first (left) and second (center) nontrivial NIV.
+\caption{Data in ``intrinsic variable'' space ($x_1, x_2$), colored by the first (left) and second (center) nontrivial NIV.
 The NIV embedding is computed from the data in the ambient space with added noise ($z_1, z_2$, with $\sigma = 100$).

### 2014-03-13 chemotaxis/draft1.tex
Added stuff for chemotaxis paper

\documentclass[12pt]{article}

\usepackage{amsmath}
\usepackage{amsfonts}

\usepackage{graphicx}
\usepackage{subcaption}

\begin{document}

The problem setting:\\
 We have a simulation of the dynamical system that is governed basically by a single parameter,­ the rate of switching.
 %
 Depending of the value of this parameter, this system dynamics range between heat equation dynamics and wave equation dynamics. 
 %
 In the heat equation dynamics, the initial conditions play very little role and the system dynamics over time are dominant. 
 %
 In the wave equation dynamics, the initial conditions play a significant role and the dynamics over time are less important.

Contribution: 
\begin{enumerate}
\item We demonstrate that an essential component of our analysis is working with the right observers (histograms/ensumble rather than a single position of a particle) and with the right metric between those observers (EMD).
%
\item We demonstrate that we recover the right factors that control the dynamical system, depending on the mode (In heat eq. mode ­ first recover the temporal dynamics, second the initial conditions; In wave eq. mode ­ first recover the initial conditions, second the temporal dynamics). In particular, these two components have a completely different ``nature” ­ for example, in this case, one parameter (initial condition) is governing the entire trajectory and the other parameter (dynamics) is governing the ``step­--by­--step" time evolution (within each trajectory).
\end{enumerate}

\section{Introduction}

\section{Problem formualtion}

\subsection{Chemotaxis ``story''} 
We look at a model of a stochastic velocity jump process that arises in models of chemotaxis \cite{...}.

\subsection{Stochastic particles}

We have a collection of $N$ particles, each of which has a position and velocity. 
%
Let $x_i(t)$ and $v_i(t)$ denote the position and velocity, respectively, of particle $i$ at time $t$.
%
The velocity of each particle is either $\pm s$, where $s$ is the speed. 
%
We initialize the particles such that
\begin{eqnarray}
x_i(0) & = 0 \\
\mathbb{P} \{ v_i(0) = +s \} & = p
\end{eqnarray}
where $p$ is some probability.
%
The velocity of each particle randomly switches between $\pm s$ following an (independent) Poisson process with rate $\lambda$.

We simulate this stochastic velocity jump process for $N=1000$ particles, and $p \in [0.1, 0.9]$.
%
We simulate the process for $t \in [0, 10]$, and record the positions of each particle at regular time intervals.
%
Let $X_{p, \lambda, s}(t)$ denote the vector of positions of the particles at time $t$ with initial right probability $p$, switching rate $\lambda$, and speed $s$.
%
We will consider two sets of simulations.
%
In one set, $\lambda = 1$ and $s=1$, and in the other set, $\lambda = 400$, and $s=20$.

\subsection{Partial differential equations}

Let $\rho(x, t)$ denote the probability density of the particles, and let $\rho^-(x, t)$ and $\rho^+(x, t)$ denote the probability density of the left and right moving particles, respectively.
%
It can be shown that, as $n \rightarrow \infty$, $\rho(x, t)$ obeys the following set of partial differential equations (PDEs)
\begin{eqnarray} \label{eqn:coupled_pdes}
\frac{\partial \rho^+}{\partial t} + s \frac{\partial \rho^+}{\partial x} & = -\lambda \rho^+ +\lambda \rho^- \\
\frac{\partial \rho^-}{\partial t} - s \frac{\partial \rho^-}{\partial x} & = \lambda \rho^+ -\lambda \rho^- 
\end{eqnarray}
%
Alternatively, \eqref{eqn:coupled_pdes} can be written as one, second--order PDE
\begin{equation} \label{eq:second_order_pde}
\frac{\partial^2 \rho}{\partial t^2} + 2 \lambda \frac{\partial \rho}{\partial t} = s^2 \frac{\partial ^2 \rho}{\partial x^2}
\end{equation}
%
We assume that $s^2/\lambda = D$ is constant.

\subsection{Asymptotic mode analysis}  \label{subsec:mode_analysis}

We will consider two regimes of simulation.
%
When $\lambda \rightarrow 0$ is small, one can see that the right-hand side of \eqref{eqn:coupled_pdes} tends to 0. 
%
Therefore, \eqref{eqn:coupled_pdes} becomes two uncoupled wave equations.
%
When $\lambda \rightarrow \infty$, $\frac{\partial^2 p}{\partial t^2} \rightarrow 0$ and \eqref{eq:second_order_pde} approaches the heat equation.

\section{Methods and Analysis}

\subsection{Diffusion maps}

\subsection{Earth mover's distance}

\section{Results}

\begin{figure}[htb]
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=\textwidth]{rawhist_p_1}
\caption{}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=\textwidth]{rawhist_t_1}
\caption{}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=\textwidth]{rawhist_p_400}
\caption{}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=\textwidth]{rawhist_t_400}
\caption{}
\end{subfigure}
\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with (a,b) $\lambda=1$, $s=1$, and (c,d) $\lambda=400$, $s=20$. The distances used in the diffusion maps kernel are the Euclidean distances between the histograms of particle positions. The data are colored by (a, c) $p$, the initial probability of left--moving particles, and (b, d) $t$, time. The first two diffusion maps modes do not capture any of the important parameters in our simulations.}
\label{fig:dmaps_embed}
\end{figure}

\begin{figure}[htb]
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=\textwidth]{EMD_p_1}
\caption{}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=\textwidth]{EMD_t_1}
\caption{}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=\textwidth]{EMD_p_400}
\caption{}
\end{subfigure}
\begin{subfigure}{0.5\textwidth}
\includegraphics[width=\textwidth]{EMD_t_400}
\caption{}
\end{subfigure}
\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with (a,b) $\lambda=1$, $s=1$, and (c,d) $\lambda=400$, $s=20$. The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions. The data are colored by (a, c) $p$, the initial probability of left--moving particles, and (b, d) $t$, time. One can see that these two parameters, $p$ and $t$, are well--correlated with the first two diffusion maps modes, $\phi_2$ and $\phi_3$. However, the ``more important'' mode, $\phi_2$, is correlated with $p$ for the small $\lambda$ case, and correalted with $t$ for the large $\lambda$ case. This is consistent with the asymtotic analysis discussed in Section \ref{subsec:mode_analysis}}
\label{fig:dmaps_embed}
\end{figure}

\end{document}

### 2014-03-13 chemotaxis/prl_paper.tex
Added stuff for chemotaxis paper

--- 2014-03-13 chemotaxis/draft1.tex
@@
-\documentclass[12pt]{article}
+\documentclass[prl,preprint]{revtex4-1}
+%\documentclass[12pt]{article}
 
@@
 mode ­ first recover the initial conditions, second the temporal dynamics).
-In particular, these two components have a completely different ``nature” ­ for example, in this case, one parameter (initial condition) is governing the entire trajectory and the other parameter (dynamics) is governing the ``step­--by­--step" time evolution (within each trajectory).
+In particular, these two components have a completely different ``nature".
+For example, in this case, one parameter (initial condition) is governing the entire trajectory and the other parameter (dynamics) is governing the ``step­--by­--step" time evolution (within each trajectory).
 \end{enumerate}
@@
 
+In dynamical systems, it is often essential to detect changes in the characteristic behavior of a system.
+%
+Methods such as bifurcation analysis, etc., have been developed with such problems in mind.
+
+We will demonstrate how we can use data mining techniques (diffusion maps \cite{...}) to {\em automatically} detect changes in a system's behavior.
+%
+We will illustrate our dynamics through a model problem arising in studies of cellular chemotaxis \cite{...}.
+%
+In our methods, it will be esseintial
+
 \section{Problem formualtion}
@@
 \subsection{Chemotaxis ``story''}
-We look at a model of a stochastic velocity jump process that arises in models of chemotaxis \cite{...}.
+
+We look at a stochastic velocity jump process that arises in models of chemotaxis \cite{othmer2000diffusion}.
+%
+Chemotaxis is movement in cells (such as bacteria) that is regulated by extracellular censed signals.
+%
+Cells must move to accomplish tasks such as finding food and navigating away from toxins, and this movement is thought to be directed by such extracellular signals.
+%
+Biologists are interested in the dynamics of dispersal of cells, and several models have been proposed \cite{othmer1988models, codling2008random}.
+%
+We will look at a velocty jump process as a model of these dynamics.
+%
+In such models, each cell is initialized with a position and a velocity, and the dynamics of each cell are goverened by a stochastic process.
+%
+In a velocity jump process, at random times, cells will ``turn around'' and switch their velocity (in biology, this turning is thought to be driven by extracellular signals such as food or toxins). 
+%
+We will analyze the dynamics of collections of such cells/particles.
 
@@
 The velocity of each particle randomly switches between $\pm s$ following an (independent) Poisson process with rate $\lambda$.
-
-We simulate this stochastic velocity jump process for $N=1000$ particles, and $p \in [0.1, 0.9]$.
 %
-We simulate the process for $t \in [0, 10]$, and record the positions of each particle at regular time intervals.
-%
-Let $X_{p, \lambda, s}(t)$ denote the vector of positions of the particles at time $t$ with initial right probability $p$, switching rate $\lambda$, and speed $s$.
-%
-We will consider two sets of simulations.
-%
-In one set, $\lambda = 1$ and $s=1$, and in the other set, $\lambda = 400$, and $s=20$.
+Note that each particle has its own ``clock" (Poisson process).
 
 \subsection{Partial differential equations}
+
+Let $X_{p, \lambda, s}(t)$ denote the vector of positions of the particles at time $t$ with initial right probability $p$, switching rate $\lambda$, and speed $s$.
 
@@
 We assume that $s^2/\lambda = D$ is constant.
+%
+Therefore, the dynamics are governed {\em only} by a single parameter $\lambda$.
 
@@
 %
-When $\lambda \rightarrow \infty$, $\frac{\partial^2 p}{\partial t^2} \rightarrow 0$ and \eqref{eq:second_order_pde} approaches the heat equation.
+Alternatively, in this regime, in the limit the second order equation \eqref{eq:second_order_pde} becomes 
+\[
+\frac{\partial^2 \rho}{\partial t^2} = s^2 \frac{\partial ^2 \rho}{\partial x^2},
+\]
+i.e., the second order wave equation.
+
+Dividing \eqref{eq:second_order_pde} by $\lambda > 0$ yields
+\[
+\frac{1}{\lambda} \frac{\partial^2 \rho}{\partial t^2} + 2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2}
+\]
+When $\lambda \rightarrow \infty$, \eqref{eq:second_order_pde} approaches the heat equation
+\[
+2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2}
+\]
 
@@
 
+TODO: we need to pay a special attention to the implementation of the EMD.
+We need to make it cleat that this distance can be very efficiently implemented  and therefore it is suitable for the practical uses and algorithm that we propose here as a viable tool for dynamical system analysis.
+
 \section{Results}
+We simulate this stochastic velocity jump process for $N=1000$ particles, and $p \in [0.1, 0.9]$.
+%
+We simulate the process for $t \in [0, 10]$, and record the positions of each particle at uniform/equal time intervals.
+%
+We will consider two sets of simulations.
+%
+In one set, $\lambda = 1$ and $s=1$, and in the other set, $\lambda = 400$, and $s=20$.
+
+TODO: In Fig. \ref{fig:dmaps_embed}, add representative (non degenerate and smooth) histograms in strategic embedded points.
 
@@
 The data are colored by (a, c) $p$, the initial probability of left--moving particles, and (b, d) $t$, time.
-The first two diffusion maps modes do not capture any of the important parameters in our simulations.}
+The first two diffusion maps modes do not accurately capture the important parameters in our simulations.}
 \label{fig:dmaps_embed}
@@
 One can see that these two parameters, $p$ and $t$, are well--correlated with the first two diffusion maps modes, $\phi_2$ and $\phi_3$.
-However, the ``more important'' mode, $\phi_2$, is correlated with $p$ for the small $\lambda$ case, and correalted with $t$ for the large $\lambda$ case.
-This is consistent with the asymtotic analysis discussed in Section \ref{subsec:mode_analysis}}
+However, the ``more important'' mode, $\phi_2$, is correlated with $p$ for the small $\lambda$ case (a), and correalted with $t$ for the large $\lambda$ case (d).
+This is consistent with the asymptotic analysis discussed in Section \ref{subsec:mode_analysis}.
+In the small $\lambda$ regime (wave equation) in (a,b) we observe that for small time, the initial condition of the particle velocity does not play a significant role -- all the particles start at position $0$, and therefore, for small time, the particles are more condensed and it is more difficult to distinguish the particles moving to the left from the particles moving to the right.
+On the other hand, for large time, once the particles evolve from the origin, this separation is clear.
+For the large $\lambda$ case (heat equation) in (c,d), we observe that for small time the initial distribution $p$ is well organized in the embedding (c), since for small time the distribution of the particles is skewed and the initial velocity plays a role in this regime.
+On the other hand, for large time, we observe that the initial distribution $p$ is not organized as well in the embedding (c), since the velocities have equilibrated and so the initial distribution is less detectable in the particle distribution.}
 \label{fig:dmaps_embed}
@@
 
+\section{Discussion}
+
+\bibliography{references}
+
 \end{document}

### 2014-03-17 chemotaxis/prl_paper.tex
Edited chemotaxis paper; made figures with histograms

@@
-\documentclass[prl,preprint]{revtex4-1}
+\documentclass[prl,reprint]{revtex4-1}
 %\documentclass[12pt]{article}
@@
 
-The problem setting:\\
+\title{Choosing Appropriate Metrics for Diffusion Maps Analysis: A Chemotaxis Case Study}
+
+\author{Carmeline J. Dsilva}
+\email{cdsilva@princeton.edu}
+\affiliation{Department of Chemical and Biological Engineering, Princeton University, Princeton, NJ, 08544}
+
+\author{Ronen Talmon}
+\email{ronen.talmon@yale.edu}
+\affiliation{Department of Mathematics, Yale University, New Haven, CT, 06520}
+
+\author{Ronald R. Coifman}
+\email{coifman@math.yale.edu}
+\affiliation{Department of Mathematics, Yale University, New Haven, CT, 06520}
+
+\author{Ioannis G. Kevrekidis}
+\email{yannis@princeton.edu}
+\affiliation{Department of Chemical and Biological Engineering, Princeton University, Princeton, NJ, 08544}
+\affiliation{Program in Applied and Computational Mathematics, Princeton University, Princeton, NJ, 08544}
+
+\date{\today}
+
+\begin{abstract}
+
 We have a simulation of the dynamical system that is governed basically by a single parameter,­ the rate of switching.
@@
  In the wave equation dynamics, the initial conditions play a significant role and the dynamics over time are less important.
-
-Contribution: 
-\begin{enumerate}
-\item We demonstrate that an essential component of our analysis is working with the right observers (histograms/ensumble rather than a single position of a particle) and with the right metric between those observers (EMD).
  %
-\item We demonstrate that we recover the right factors that control the dynamical system, depending on the mode (In heat eq.
+We demonstrate that an essential component of our analysis is working with the right observers (histograms/ensumble rather than a single position of a particle) and with the right metric between those observers (EMD).
+%
+We demonstrate that we recover the right factors that control the dynamical system, depending on the mode (In heat eq.
 mode ­ first recover the temporal dynamics, second the initial conditions; In wave eq.
@@
 For example, in this case, one parameter (initial condition) is governing the entire trajectory and the other parameter (dynamics) is governing the ``step­--by­--step" time evolution (within each trajectory).
-\end{enumerate}
+\end{abstract}
+
+\keywords{diffusion maps, earth mover's distance, histograms}
+
+\maketitle
 
@@
 
-We will demonstrate how we can use data mining techniques (diffusion maps \cite{...}) to {\em automatically} detect changes in a system's behavior.
+We will demonstrate how we can use data mining techniques (diffusion maps \cite{coifman2005geometric}) to {\em automatically} detect changes in a system's behavior.
 %
-We will illustrate our dynamics through a model problem arising in studies of cellular chemotaxis \cite{...}.
+We will illustrate our dynamics through a model problem arising in studies of cellular chemotaxis \cite{othmer2000diffusion}.
 %
@@
 %
-Chemotaxis is movement in cells (such as bacteria) that is regulated by extracellular censed signals.
+Chemotaxis is the movement of cells (such as bacteria) that is regulated by extracellular sensed signals.
 %
-Cells must move to accomplish tasks such as finding food and navigating away from toxins, and this movement is thought to be directed by such extracellular signals.
+Such signals allow cells to accomplish tasks such as finding food and navigating away from toxins.
 %
-Biologists are interested in the dynamics of dispersal of cells, and several models have been proposed \cite{othmer1988models, codling2008random}.
+Biologists are interested in the dynamics of cellular movement and dispersal, and several models have been proposed \cite{othmer1988models, codling2008random}.
 %
-We will look at a velocty jump process as a model of these dynamics.
+We will look at a velocty jump process model of these dynamics.
 %
-In such models, each cell is initialized with a position and a velocity, and the dynamics of each cell are goverened by a stochastic process.
+In such models, each cell is initialized with a position and a velocity, and the dynamics of each cell are governed by a stochastic process.
 %
-In a velocity jump process, at random times, cells will ``turn around'' and switch their velocity (in biology, this turning is thought to be driven by extracellular signals such as food or toxins). 
+In a velocity jump process, at random times, cells will ``turn around'' and switch their velocity (this turning is thought to be driven by extracellular signals from food or toxins). 
 %
@@
 
-We have a collection of $N$ particles, each of which has a position and velocity. 
+Assume we have a collection of $N$ particles, each of which has a position and velocity. 
 %
@@
 We initialize the particles such that
-\begin{eqnarray}
+\begin{equation}
+\begin{aligned}
 x_i(0) & = 0 \\
 \mathbb{P} \{ v_i(0) = +s \} & = p
-\end{eqnarray}
+\end{aligned}
+\end{equation}
 where $p$ is some probability.
@@
 It can be shown that, as $n \rightarrow \infty$, $\rho(x, t)$ obeys the following set of partial differential equations (PDEs)
-\begin{eqnarray} \label{eqn:coupled_pdes}
+\begin{equation} \label{eqn:coupled_pdes}
+\begin{aligned}
 \frac{\partial \rho^+}{\partial t} + s \frac{\partial \rho^+}{\partial x} & = -\lambda \rho^+ +\lambda \rho^- \\
 \frac{\partial \rho^-}{\partial t} - s \frac{\partial \rho^-}{\partial x} & = \lambda \rho^+ -\lambda \rho^- 
-\end{eqnarray}
+\end{aligned}
+\end{equation}
 %
@@
 %
-When $\lambda \rightarrow 0$ is small, one can see that the right-hand side of \eqref{eqn:coupled_pdes} tends to 0. 
+When $\lambda \rightarrow 0$, one can see that the right-hand side of \eqref{eqn:coupled_pdes} tends to 0. 
 %
@@
 
+We will use diffusion maps \cite{coifman2005geometric} to analyze the data from our stichastic simulations.
+%
+Diffusion maps is a nonlinear dimensionality reduction technique.
+%
+In our simulations, although the data is very high--dimensional (each data point contains the positions of all $N$ particles), there are only two important parameters ($p$ and $t$) that determine the current state of the system.
+
+Assume we have data points $z_1, z_2, \dots, z_m \in \mathbb{R}^p$, but that these data points lie on a $d$-dimensional, nonlinear manifold, with $d \ll p$. 
+%
+We want to uncover a $d$-dimensional parameterization of the points $x_1, \dots, x_m$.
+%
+We first construct the matrix $W \in \mathbb{R}^{m \times m}$, with
+\begin{equation} \label{eq:W}
+W_{ij} = \exp \left( -\frac{\|z_i - z_j \|^2}{\epsilon^2} \right)
+\end{equation}
+where $\| \cdot \|$ denotes the appropriate norm for the data (we will discuss the choice of norm in the following section), and $\epsilon$ is a characteristic distance between the data points.
+%
+We then construct the diaganol matrix $D \in \mathbb{R}^{m \times m}$, with $D_{ii} = \sum_j W_{ij}$, and then the matrix $A$, defined as
+\begin{equation}
+A = D^{-1} W.
+\end{equation}
+%
+We then compute the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_m$ and eignevalues $\phi_1, \phi_2, \dots, \phi_m$ and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_m|$.
+%
+Note that the matrix $A$ is row--stochastic ($\sum_j A_{ij} = 1$), which implies $\lambda_1 = 1$ and $\phi_1$ is a constant vector.
+%
+The next few eigenvectors $\phi_2, \phi_3, \dots$ give us the embedding coordinates for our data, such that $\phi_j(i)$ gives us the $j^{th}$ embedding coordinate for $x_i$.
+
+As we can see from the analysis in \ref{subsec:mode_analysis}, the ``important'' quantity to describe the dynamics is not the positoin of each particle, but rather the probability density of the particles.
+%
+We will empirically estimate the probability density for each timepoint by computing the histogram of the particle positions. 
+%
+Therefore, $z_i \in \mathbb{R}^p$ will be the histogram of $x_1, x_2, \dots, x_N$ into $p$ bins.
+
 \subsection{Earth mover's distance}
 
-TODO: we need to pay a special attention to the implementation of the EMD.
+The {\em essential} requirement for diffusion maps is having an appropriate distance metric between data points to compute $W$ in \ref{eq:W}.
+%
+Often, one uses the Euclidean distance between data points;
+however, for histograms which approximate a probability density, the Euclidean distance can often be misleading.
+%
+For example, the distance between two uniform distributions with finite, non-overlapping support will always be the same, {\em independent} of the distances between the support intervals.
+%
+Intuitively, we would want/expect two distributions whose supports are close to be closer (in norm) than two distributions whose supports are very far from each other.
+
+We therefore use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between our histograms.
+%
+Conceptually, EMD measures how much ``work'' it takes to transform one probability density into another.
+%
+It therefore not only considers where the densities are incosistent, but how far apart they are.
+
+Although the brute--force computation of the EMD between two histograms is computationally expensive, there has been a plethora of work in developing efficient algorithms for the computation of EMD \cite{...}.
+%
+%TODO: we need to pay a special attention to the implementation of the EMD.
 We need to make it cleat that this distance can be very efficiently implemented  and therefore it is suitable for the practical uses and algorithm that we propose here as a viable tool for dynamical system analysis.
+%
+We use the EMD implementation from \cite{Pele-eccv2008, Pele-iccv2009} to compute the distances between histograms.
+%
+The implementation is available at \url{http://www.cs.huji.ac.il/~ofirpele/FastEMD/code}.
 
 \section{Results}
+
 We simulate this stochastic velocity jump process for $N=1000$ particles, and $p \in [0.1, 0.9]$.
@@
 
-TODO: In Fig. \ref{fig:dmaps_embed}, add representative (non degenerate and smooth) histograms in strategic embedded points.
+For each data point, we compute the histogram of the particle positions into 128 evenly spaced bins. 
+%
+In Figure \ref{fig:dmaps_embed_emd}, we have used the earth mover's distance to compute the distances between histograms.
+%
+One can see that these two parameters, $p$ and $t$, are well--correlated with the first two diffusion maps modes, $\phi_2$ and $\phi_3$. 
+%
+However, the ``more important'' mode, $\phi_2$, is correlated with $p$ for the small $\lambda$ case (a), and correalted with $t$ for the large $\lambda$ case (d).
+This is consistent with the asymptotic analysis discussed in Section \ref{subsec:mode_analysis}.
+%
+In the small $\lambda$ regime (wave equation) in (a,b) we observe that for small time, the initial condition of the particle velocity does not play a significant role -- all the particles start at position $0$, and therefore, for small time, the particles are more condensed and it is more difficult to distinguish the particles moving to the left from the particles moving to the right.
+On the other hand, for large time, once the particles evolve from the origin, this separation is clear.
+%
+For the large $\lambda$ case (heat equation) in (c,d), we observe that for small time the initial distribution $p$ is well organized in the embedding (c), since for small time the distribution of the particles is skewed and the initial velocity plays a role in this regime.
+%
+On the other hand, for large time, we observe that the initial distribution $p$ is not organized as well in the embedding (c), since the velocities have equilibrated and so the initial distribution is less detectable in the particle distribution.
 
-\begin{figure}[htb]
-\begin{subfigure}{0.5\textwidth}
+\begin{figure*}[htb]
+\begin{subfigure}{0.2\textwidth}
 \includegraphics[width=\textwidth]{rawhist_p_1}
@@
 \end{subfigure}
-\begin{subfigure}{0.5\textwidth}
+\begin{subfigure}{0.2\textwidth}
 \includegraphics[width=\textwidth]{rawhist_t_1}
@@
 \end{subfigure}
-\begin{subfigure}{0.5\textwidth}
+\begin{subfigure}{0.2\textwidth}
 \includegraphics[width=\textwidth]{rawhist_p_400}
@@
 \end{subfigure}
-\begin{subfigure}{0.5\textwidth}
+\begin{subfigure}{0.2\textwidth}
 \includegraphics[width=\textwidth]{rawhist_t_400}
@@
 The first two diffusion maps modes do not accurately capture the important parameters in our simulations.}
-\label{fig:dmaps_embed}
-\end{figure}
+\label{fig:dmaps_embed_noemd}
+\end{figure*}
 
-\begin{figure}[htb]
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{EMD_p_1}
+\begin{figure*}[htb]
+\begin{subfigure}{0.4\textwidth}
+\includegraphics[width=\textwidth]{EMD_withhist_p_1}
 \caption{}
 \end{subfigure}
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{EMD_t_1}
+\begin{subfigure}{0.4\textwidth}
+\includegraphics[width=\textwidth]{EMD_withhist_t_1}
 \caption{}
 \end{subfigure}
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{EMD_p_400}
+\begin{subfigure}{0.4\textwidth}
+\includegraphics[width=\textwidth]{EMD_withhist_p_400}
 \caption{}
 \end{subfigure}
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{EMD_t_400}
+\begin{subfigure}{0.4\textwidth}
+\includegraphics[width=\textwidth]{EMD_withhist_t_400}
 \caption{}
@@
 The data are colored by (a, c) $p$, the initial probability of left--moving particles, and (b, d) $t$, time.
-One can see that these two parameters, $p$ and $t$, are well--correlated with the first two diffusion maps modes, $\phi_2$ and $\phi_3$.
-However, the ``more important'' mode, $\phi_2$, is correlated with $p$ for the small $\lambda$ case (a), and correalted with $t$ for the large $\lambda$ case (d).
-This is consistent with the asymptotic analysis discussed in Section \ref{subsec:mode_analysis}.
-In the small $\lambda$ regime (wave equation) in (a,b) we observe that for small time, the initial condition of the particle velocity does not play a significant role -- all the particles start at position $0$, and therefore, for small time, the particles are more condensed and it is more difficult to distinguish the particles moving to the left from the particles moving to the right.
-On the other hand, for large time, once the particles evolve from the origin, this separation is clear.
-For the large $\lambda$ case (heat equation) in (c,d), we observe that for small time the initial distribution $p$ is well organized in the embedding (c), since for small time the distribution of the particles is skewed and the initial velocity plays a role in this regime.
-On the other hand, for large time, we observe that the initial distribution $p$ is not organized as well in the embedding (c), since the velocities have equilibrated and so the initial distribution is less detectable in the particle distribution.}
-\label{fig:dmaps_embed}
-\end{figure}
+}
+\label{fig:dmaps_embed_emd}
+\end{figure*}
 
@@
 
-\bibliography{references}
+\bibliography{../../references/references}
 

### 2014-03-17 chemotaxis/prl_paper.tex
updated prl paper

@@
 
-We have a simulation of the dynamical system that is governed basically by a single parameter,­ the rate of switching.
+We simulate a stochastic system whose dynamics are governed by a single parameter (the rate of switching).
 %
@@
 %
-We demonstrate that an essential component of our analysis is working with the right observers (histograms/ensumble rather than a single position of a particle) and with the right metric between those observers (EMD).
+We use dimensionality reduction techniques to analyze the results of such simulations.
 %
-We demonstrate that we recover the right factors that control the dynamical system, depending on the mode (In heat eq.
-mode ­ first recover the temporal dynamics, second the initial conditions; In wave eq.
-mode ­ first recover the initial conditions, second the temporal dynamics).
-In particular, these two components have a completely different ``nature".
-For example, in this case, one parameter (initial condition) is governing the entire trajectory and the other parameter (dynamics) is governing the ``step­--by­--step" time evolution (within each trajectory).
+We demonstrate that an essential component of our analysis is working with the right observers (histograms/ensembles rather than the single position of a particle) and with the right metric between those observers (earth mover's distance rather than Euclidean distance).
+%
+We demonstrate that, depending on the ``mode'' of the dynamical system, we recover the right factors that control the dynamics.
+%
+In heat equation mode, we­ first recover the temporal dynamics, and second the initial conditions;
+in wave equation mode, we­ first recover the initial conditions, and second recover the temporal dynamics. 
+%
+Furthermore, these two components have a completely different ``nature". 
+%
+For example, in this case, one parameter (the initial condition) is governing the entire trajectory and the other parameter (time) is governing the ``step­--by­--step" evolution within each trajectory.
+
 \end{abstract}

### 2014-03-21 chemotaxis/prl_paper.tex
Updated chemotaxis paper; added more figures

@@
-\documentclass[prl,reprint]{revtex4-1}
+\documentclass[prl, reprint, final]{revtex4-1}
 %\documentclass[12pt]{article}
@@
 %
-Depending of the value of this parameter, this system dynamics range between heat equation dynamics and wave equation dynamics. 
+Depending of the value of this parameter, the system dynamics range between heat equation dynamics and wave equation dynamics. 
 %
-In the heat equation dynamics, the initial conditions play very little role and the system dynamics over time are dominant. 
+In the heat equation dynamics, the system quickly equilibrates and the initial conditions play very little role. 
 %
-In the wave equation dynamics, the initial conditions play a significant role and the dynamics over time are less important.
+In the wave equation dynamics, the initial conditions play a significant role in the trajectory.
 %
@@
 %
-We demonstrate that an essential component of our analysis is working with the right observers (histograms/ensembles rather than the single position of a particle) and with the right metric between those observers (earth mover's distance rather than Euclidean distance).
+We demonstrate that an essential component of our analysis is working with the right observers (histograms/ensembles, rather than the single position of a particle) and with the right metric between those observers (earth mover's distance, rather than Euclidean distance).
 %
-We demonstrate that, depending on the ``mode'' of the dynamical system, we recover the right factors that control the dynamics.
+We show that, depending on the ``mode'' of the dynamical system, we recover the right factors that control the dynamics.
 %
-In heat equation mode, we­ first recover the temporal dynamics, and second the initial conditions;
-in wave equation mode, we­ first recover the initial conditions, and second recover the temporal dynamics. 
+In heat equation mode, the first factor we uncover parameterizes the temporal dynamics, and the second factor parameterizes the initial conditions;
+in wave equation mode, the first factor we uncover parameterizes the initial conditions, and the second factor parameterizes the temporal dynamics; 
 %
@@
 %
-For example, in this case, one parameter (the initial condition) is governing the entire trajectory and the other parameter (time) is governing the ``step­--by­--step" evolution within each trajectory.
+One parameter (the initial condition) is governing the entire trajectory and the other parameter (time) is governing the ``step­--by­--step" evolution within each trajectory.
 
@@
 
-We will demonstrate how we can use data mining techniques (diffusion maps \cite{coifman2005geometric}) to {\em automatically} detect changes in a system's behavior.
+We will demonstrate how we can use data--driven techniques (diffusion maps \cite{coifman2005geometric}) to {\em automatically} detect changes in a system's behavior.
 %
@@
 %
-In our methods, it will be esseintial
+In our methods, it will be essential to use the appropriate observers for our system in order to uncover the underlying dynamics.
+%
+Furthermore, we will need to use the appropriate metric between observers.
 
-\section{Problem formualtion}
+\section{Problem formulation}
 
@@
 %
-We will look at a velocty jump process model of these dynamics.
+We will look at a velocity jump process model of these dynamics.
 %
@@
 
-We will use diffusion maps \cite{coifman2005geometric} to analyze the data from our stichastic simulations.
+We will use diffusion maps \cite{coifman2005geometric} to analyze the data from our stochastic simulations.
 %
@@
 %
-In our simulations, although the data is very high--dimensional (each data point contains the positions of all $N$ particles), there are only two important parameters ($p$ and $t$) that determine the current state of the system.
+In our simulations, although the data is very high--dimensional, there are only two important parameters ($p$ and $t$) that determine the current state of the system.
 
@@
 %
-We want to uncover a $d$-dimensional parameterization of the points $x_1, \dots, x_m$.
+Our goal is to uncover a $d$-dimensional parameterization of the points $z_1, \dots, z_m$ that respects the underlying manifold geometry.
 %
@@
 %
-We then construct the diaganol matrix $D \in \mathbb{R}^{m \times m}$, with $D_{ii} = \sum_j W_{ij}$, and then the matrix $A$, defined as
+We then construct the diagonal matrix $D \in \mathbb{R}^{m \times m}$, with $D_{ii} = \sum_j W_{ij}$, and the matrix $A$, defined as
 \begin{equation}
@@
 %
-We then compute the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_m$ and eignevalues $\phi_1, \phi_2, \dots, \phi_m$ and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_m|$.
+We compute the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_m$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_m$ of $A$, and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_m|$.
 %
@@
 %
-The next few eigenvectors $\phi_2, \phi_3, \dots$ give us the embedding coordinates for our data, such that $\phi_j(i)$ gives us the $j^{th}$ embedding coordinate for $x_i$.
+The next few eigenvectors $\phi_2, \phi_3, \dots$ give us the embedding coordinates for our data, such that $\phi_{j+1}(i)$ gives us the $j^{th}$ embedding coordinate for $z_i$.
 
-As we can see from the analysis in \ref{subsec:mode_analysis}, the ``important'' quantity to describe the dynamics is not the positoin of each particle, but rather the probability density of the particles.
+As we can see from the analysis in the preceeding section, the ``important'' quantity to describe the dynamics is not the position of each particle, but rather the probability density of the particles.
 %
-We will empirically estimate the probability density for each timepoint by computing the histogram of the particle positions. 
+We will empirically estimate the probability density at each time point in our simulations by computing histograms of the particle positions. 
 %
@@
 
-The {\em essential} requirement for diffusion maps is having an appropriate distance metric between data points to compute $W$ in \ref{eq:W}.
+The {\em essential} requirement for diffusion maps is having an appropriate distance metric between data points.
 %
@@
 %
-For example, the distance between two uniform distributions with finite, non-overlapping support will always be the same, {\em independent} of the distances between the support intervals.
+For example, the distance between two uniform distributions with finite, non-overlapping support will always be the same, {\em independent} of the distance between the support intervals.
 %
-Intuitively, we would want/expect two distributions whose supports are close to be closer (in norm) than two distributions whose supports are very far from each other.
+Intuitively, we would want/expect two distributions whose supports are a short distance apart to be closer (in norm) than two distributions whose supports are very far from each other.
 
-We therefore use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between our histograms.
+We use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between histograms.
 %
@@
 It therefore not only considers where the densities are incosistent, but how far apart they are.
-
+%
 Although the brute--force computation of the EMD between two histograms is computationally expensive, there has been a plethora of work in developing efficient algorithms for the computation of EMD \cite{...}.
@@
 %
-We use the EMD implementation from \cite{Pele-eccv2008, Pele-iccv2009} to compute the distances between histograms.
+We use the EMD algorithm from \cite{Pele-eccv2008, Pele-iccv2009} to compute the distances between histograms.
 %
@@
 %
+We discard the initial portion of the trajectory corresponding to $t < 1$, as this is very noisy and corresponds to initial relaxation of the system.
+%
 We will consider two sets of simulations.
@@
 
-For each data point, we compute the histogram of the particle positions into 128 evenly spaced bins. 
+For each data point, we compute the histogram of the particle positions into $p=32$ evenly spaced bins. 
 %
+In Figure \ref{fig:dmaps_embed_noemd}, we have used the standard Euclidean distance between the histograms to compute the distances for \eqref{eq:W}.
+%
+The embeddings obtained are not particularly informative, and do not accurrately capture the underlying system dynamics.
+%
+This is due to our poor choice of metric, which does not accurrately describe the distances between histograms.
+
 In Figure \ref{fig:dmaps_embed_emd}, we have used the earth mover's distance to compute the distances between histograms.
 %
-One can see that these two parameters, $p$ and $t$, are well--correlated with the first two diffusion maps modes, $\phi_2$ and $\phi_3$. 
+One can see that the two important factors, $p$ and $t$, are well--correlated with the first two diffusion maps modes, $\phi_2$ and $\phi_3$. 
 %
-However, the ``more important'' mode, $\phi_2$, is correlated with $p$ for the small $\lambda$ case (a), and correalted with $t$ for the large $\lambda$ case (d).
-This is consistent with the asymptotic analysis discussed in Section \ref{subsec:mode_analysis}.
+However, the ``more important'' mode, $\phi_2$, is correlated with $p$ for the small $\lambda$ case (a), and correlated with $t$ for the large $\lambda$ case (d).
+This is consistent with the asymptotic analysis discussed previously.
 %
-In the small $\lambda$ regime (wave equation) in (a,b) we observe that for small time, the initial condition of the particle velocity does not play a significant role -- all the particles start at position $0$, and therefore, for small time, the particles are more condensed and it is more difficult to distinguish the particles moving to the left from the particles moving to the right.
+In the small $\lambda$ regime (wave equation), shown in Figures \ref{subfig:small_lambda_p} and \ref{subfig:small_lambda_t}, we observe that for small time, the initial condition of the particle velocity does not play a significant role -- all the particles start at position $0$, and therefore, for small time, the particles are more condensed and it is more difficult to distinguish the particles moving to the left from the particles moving to the right.
 On the other hand, for large time, once the particles evolve from the origin, this separation is clear.
 %
-For the large $\lambda$ case (heat equation) in (c,d), we observe that for small time the initial distribution $p$ is well organized in the embedding (c), since for small time the distribution of the particles is skewed and the initial velocity plays a role in this regime.
+For the large $\lambda$ case (heat equation), shown in Figures \ref{subfig:large_lambda_p} and \ref{subfig:large_lambda_t}, we observe that for small time the initial distribution $p$ is well organized in the embedding in Figure \ref{subfig:large_lambda_p}, since for small time the distribution of the particles is skewed and the initial velocity plays a role in this regime.
 %
-On the other hand, for large time, we observe that the initial distribution $p$ is not organized as well in the embedding (c), since the velocities have equilibrated and so the initial distribution is less detectable in the particle distribution.
+On the other hand, for large time, we observe that the initial distribution $p$ is not organized as well in the embedding in Figure \ref{subfig:large_lambda_p}, since the velocities have equilibrated and so the initial distribution is less detectable in the particle distribution.
+
+We would like to note that the crossover between the wave equation regime, where $p$ is the more important parameter, and the heat equation regime, where $t$ is the more important parameter, depends on the time scale of observation.
+%
+To illustate this, we took a fixed value of $\lambda$ and $s$ ($\lambda=100$ and $s=10$), and allowed the system to evolve for 10 and 20 time units. 
+%
+The results are shown in Figure \ref{fig:dmaps_embed_varyt}.
+%
+When the system only evolves for 10 time units, the majority of the data still appears ``wave--like'', and so the dominant parameter, $\phi_2$, is correlated with $p$, the initial condition.
+%
+However, when the system evolves for 20 time units, the behavior is more diffusive and $\phi_2$ is now correlated with $t$.
+
+TODO: add histograms to this figure
 
@@
 The distances used in the diffusion maps kernel are the Euclidean distances between the histograms of particle positions.
-The data are colored by (a, c) $p$, the initial probability of left--moving particles, and (b, d) $t$, time.
-The first two diffusion maps modes do not accurately capture the important parameters in our simulations.}
+The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time.
+The first two diffusion maps modes do not accurately capture the important factors in our simulations.}
 \label{fig:dmaps_embed_noemd}
@@
 \caption{}
+\label{subfig:small_lambda_p}
 \end{subfigure}
@@
 \caption{}
+\label{subfig:small_lambda_t}
 \end{subfigure}
@@
 \caption{}
+\label{subfig:large_lambda_p}
 \end{subfigure}
@@
 \caption{}
+\label{subfig:large_lambda_t}
 \end{subfigure}
@@
 The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions.
-The data are colored by (a, c) $p$, the initial probability of left--moving particles, and (b, d) $t$, time.
+The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time.
 }
 \label{fig:dmaps_embed_emd}
+\end{figure*}
+
+\begin{figure*}
+\begin{subfigure}{0.4\textwidth}
+\includegraphics[width=\textwidth]{EMD2_p_10}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.4\textwidth}
+\includegraphics[width=\textwidth]{EMD2_t_10}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.4\textwidth}
+\includegraphics[width=\textwidth]{EMD2_p_20}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.4\textwidth}
+\includegraphics[width=\textwidth]{EMD2_t_20}
+\caption{}
+\end{subfigure}
+\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with $\lambda=100$, $s=10$.
+The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions.
+The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time.
+The particles are allowed to evolve for (a, b) 10 time units, and (c, d) 20 time units.
+}
+\label{fig:dmaps_embed_varyt}
 \end{figure*}

### 2014-03-21 chemotaxis/simulate_jump_process_EMD.m

@@ -85,20 +85,20 @@ eps = median(W(:));
 
 figure;
 scatter(V(:,2),V(:,3),200,all_time(idx), '.')
-xlabel('$\phi_2$', 'interpreter','latex')
-ylabel('$\phi_3$', 'interpreter','latex')
+xlabel('$\phi_2$', 'interpreter','latex', 'fontsize', 20)
+ylabel('$\phi_3$', 'interpreter','latex', 'fontsize', 20)
 %title('histograms: colored by time')
 h = colorbar;
-set(get(h,'xlabel'),'String', 't');
+set(get(h,'xlabel'),'String', 't', 'fontsize', 20);
 print(sprintf('rawhist_t_%d', lambda), '-r300','-djpeg')
 
 figure;
 scatter(V(:,2),V(:,3),200,all_p(idx), '.')
-xlabel('$\phi_2$', 'interpreter','latex')
-ylabel('$\phi_3$', 'interpreter','latex')
+xlabel('$\phi_2$', 'interpreter','latex', 'fontsize', 20)
+ylabel('$\phi_3$', 'interpreter','latex', 'fontsize', 20)
 %title('histograms: colored by p')
 h = colorbar;
-set(get(h,'xlabel'),'String', 'p');
+set(get(h,'xlabel'),'String', 'p', 'fontsize', 20);
 print(sprintf('rawhist_p_%d', lambda), '-r300','-djpeg')
 
 %% compute EMD between histograms
@@ -174,10 +174,10 @@ scatter(V2(:,2),V2(:,3),200,all_time(idx), '.');
 ax = gca;
 axis_lim = [-0.4 0.4 -0.25 0.3];
 axis(axis_lim)
-xlabel('$\phi_2$', 'interpreter','latex')
-ylabel('$\phi_3$', 'interpreter','latex')
+xlabel('$\phi_2$', 'interpreter','latex', 'fontsize', 20)
+ylabel('$\phi_3$', 'interpreter','latex', 'fontsize', 20)
 h = colorbar;
-set(get(h,'xlabel'),'String', 't');
+set(get(h,'xlabel'),'String', 't', 'fontsize', 20);
 curr_ax = get(ax, 'position');
 hist_all2 = hist_all(:, idx);
 hold on
@@ -214,10 +214,10 @@ figure;
 set(gcf,'PaperPositionMode','auto');
 scatter(V2(:,2),V2(:,3),200,all_p(idx), '.');
[…]

### 2014-03-21 chemotaxis/simulate_jump_process_EMD_varyt.m

@@ -1,0 +1,178 @@
+clear all
+close all
+
+rng(321);
+
+%% simulations
+
+% rate of switching velocity
+lambda = 100;
+% speed of particles
+s = 10;
+
+% % rate of switching velocity
+% lambda = 400;
+% % speed of particles
+% s = 20;
+
+% number of particles
+N = 1000;
+
+
+% % time step
+% dt = 2;
+% % maximum simulation time
+% tmax = 20;
+
+% time step
+dt = 1;
+% maximum simulation time
+tmax = 10;
+
+
+nsteps = floor(tmax / dt) + 1;
+
+%initial probility of a particle moving to the right (vel=+1)
+nsim = 10;
+pinit = linspace(0.1, 0.9, nsim);
+
+min_start = -10;
+max_start = 10;
+
+% histogram parameters
+nbins = 32;
+x_hist = linspace(min_start-s*tmax, max_start+s*tmax, nbins);
+hist_all = zeros(nbins, nsteps*nsim);
+hist_left = zeros(nbins, nsteps*nsim);
+hist_right = zeros(nbins, nsteps*nsim);
+all_time = zeros(1, nsteps*nsim);
+all_p = zeros(1, nsteps*nsim);
+
+%%
+subplot_dim1 = floor(sqrt(nsim));
+subplot_dim2 = ceil(nsim / subplot_dim1);
+
+figure;
+for j=1:nsim
+    % simulate velocity jump process
+    [pos, vel, time] = simulate_jumps(N, tmax, dt, lambda, s, pinit(j));
+    %[pos, vel, time] = simulate_jumps_unifdist(N, tmax, dt, lambda, s, pinit(j), min_start, max_start);
+    
+    % histogram of all particle positions as a function of time
+    for i=1:nsteps
+        
+        [hist_all(:,(j-1)*nsteps+i), ~] = hist(pos(:,i),x_hist);
+        
+        idx = (vel(:,i) == -1);
+        [hist_left(:,(j-1)*nsteps+i), ~] = hist(pos(idx,i),x_hist);
+        
+        idx = (vel(:,i) == 1);
+        [hist_right(:,(j-1)*nsteps+i), ~] = hist(pos(idx,i),x_hist);
+        
+    end
+    
+    % store times and probabilities
[…]

### 2014-03-21 reducible_SDE/notes1.tex
Added stuff from Ronen visit (multiscale SDE)

@@
 
-Goals: 
+\section{Outline/Goals}
+
+\subsection{Goals}
 \begin{itemize}
-\item Want to study how the scale of the noise in a system affects NIV
-\item Will look at systems with both underlying noise in the ``intrinsic'' variables, and added noise to the ``ambient'' space
-\item The magnitude of the noise in the ambient space relative to the magnitude of the noise in the intrinsic space will determine the embedding we compute
-\item We expect to obtain DMAPS variables if the noise in the ambient space is large (we will only ``see'' the ambient space noise, and not the intrinsic space noise), and NIV if the noise in the ambient space is small (so that we can still ``see'' the intrinsic space noise)
+\item We have a complex dynamical system: high--dimensional, many scales (intrinsic and ambient), noisy (intrinsic and ambient), nonlinear (dynamics and observation function)
+\item Our goal is to twofold
+\begin{itemize}
+	\item Recover the slow variables of the system
+	\item Recover NIVs: the recovered slow variables should be in a ``canonical'' coordinate system, invariant to the obervation/measurement function and ambient noise
 \end{itemize}
+\item We do that through the solution of an eigenproblem of the kernel based on the Mahalanobis distance.
+\item We want to study the effects of the dynamical system characteristics (scales, noises, nonlinearities) on the performance of NLICA and our ability to recover NIV.
+Furthermore, we want to layout the mathematical conditions reflecting this.
+\item When the mathematical conditions to recover NIV are not satisfied, we will look at constructing intermediate observers (i.e., histograms) to facilitate the recovery of NIVs through filtering out the fastest timescale by averaging/expected values.
+\end{itemize}
+
+\subsection{Outline}
+
+\begin{itemize}
+
+\item Start with a simple example where we don't have NIV-- only do reduction; this emphasixes one of the effects of Mahalanobis-- filtering out the fast variables in order to recover {\em only} the slow variables.
+This is the 1d-1d orthoganol example, and the 1d-1d half moon example (and perhaps the chemical network example).
+In these cases, we do not care about estimating the covariances accurrately-- we only care about collapsing the fast direction(s).
+\item Another simple example for NIV-- the 2d example from Amit's NLICA paper.
+Here, the first thing we want to do is detemine how we can accurrately estimate the covariances when we have a time series, rather than short bursts.
+We will do this by comparing our estimates to the analytical covariance, and examining both the drift and curvature effects.
+\item An example that combines the first two examples: we now take Amit's NLICA example, and add fast noise on top of it.
+We now expect the covariance estimation to be ``corrupted'' by the fast ambient noise.
+Based on the time and length scales of this noise, we expect to recover some combination of fast and slow variables, and some combinations of NIVs and DMAPs.
+\item Also, math. We need to analytically show the effects of curvature and noise on the estimation of the covariances. 
+\end{itemize}
+
+\section{Problem Formulation}
+
+Our standard example will be
+
+\begin{eqnarray}
+dx & = & f(x, y) dt + dW_1 \\
+dy & = & \frac{1}{\epsilon} g(x, y) dt + \frac{1}{\sqrt{\epsilon}} dW_2
+\end{eqnarray}
+
+We can then apply a measurement function $M(x, y)$ and also some noise, $M(x, y) + \xi$.
+
+Our goal is to find the {\em slow variables} $x$.
+
+We are interested in what happens when we do Mahalanobis/NLICA/NIV over
+\begin{itemize}
+\item very short times
+\item medium times (fast variable $y$ saturates)
+\end{itemize}
+
+For simplicity, we will first focus on the following system
+\begin{equation} \label{eq:NIV_model}
+d\vec{x} = f(\vec{x}) dt + dW
+\end{equation}
+%
+Our observations will be $M(x) + \xi$, where $M$ is a (nonlinear) measurement function, and $\xi$ is some noise.
+%
+Again, our goal is to recover $x$ using NLICA.
+
+When we take observation windows, we see a combination of 
+\begin{itemize}
+\item The intrinsic noise in $x$: this is what we are interested in/want to see.
+This allows us to invert the measurement function through the covariance (which approximates $JJ^T$) of the samples in the window.
+Note: when we have trajectories, to improve the estimation of the covariance (as if we have short busts), we compute the covariance of the {\em differences} between successive samples in the trajectory.
+\item The ambient space noise $\xi$: this is what we do not want to see, because it distorts/hides the shape of $JJ^T$, which is the intrinsic noise translated into the ambient space.
+\item The distortion of $M$: we do not want to see any curvature effects of $M$, we only want to see the local linear distortion of the ball of samples to an ellipse.
+This is because we approximate the effects of $M$ as $JJ^T$.
+Higher--order terms (such as the curvature) make the approximation of $JJ^T$ via the covariance estimation less accurate.
+\end{itemize}
+
+So, for a given time window $T$, we see
+\begin{equation}
+M(B_{intrinsic} (x, T)) + B_{ambient} (M(x), T)
+\end{equation}
+where $B_{intrinsic} (x, T)$ is the local cloud of samples in the intrinsic space, and $B_{ambient} (M(x), T)$ is the local cloud of ambient space noise.
+%
+For NLICA to recover $x$, we want
+\begin{equation}
+M(B_{intrinsic} (x, T)) + B_{ambient} (M(x), T) \approx J_{x} B_{intrinsic} (x, T)
+\end{equation}
+where $J_{x}$ is the Jacobian of $M$ at $x$.
+%
+If this is not possible, then we will have to take intermediate observers (i.e., histograms).
+
+TODO: Convert this to the notations used in Amit NLICA paper, section 6.1
+
+\section{1d+1d example: recovering the slow variables}
+
+In this example, we have a system of one slow + one fast SDEs.
+%
+We show that DMAPS picks out the slow {\em and} the fast variables, and the order in which they appear in the spectrum depends on the timescale of sampling.
+%
+However, NLICA collapses the fast direction, and so the first component captures the slow variable; the fast variable typically appears very far down in the eigenvectors.
+%
+See Figures \ref{fig:1d1drxnnetwork}, \ref{fig:1d1dexample}, and \ref{fig:halfmoons}.
+
+The scale of the time windows for Mahalanobis affects which eigenvector is correlated with the fast variable.
+%
+If the time window is smaller than (or equal to) the saturation time of the fast variable, then the fast variable will appear far down in the eigenvectors.
+%
+However, when the time windows are larger than the saturation time, the fast noise saturates, and the ellipses we sample become ``fat''.
+%
+Therefore, the slow direction is collapsed more (relative to the fast direction), and the fast variable appears higher up in the eigenvectors from NLICA.
+
+Note: This example includes only one fast and one slow variable, and therefore only requires reduction from Mahalanobis.
+%
+The question therefore arises, when we have several (perhaps coupled) slow variables, and some fast variables, when can we retrieve the (slow) NIV using Mahalanobis/NLICA?
+
+Mahalanobis for reduction requires time windows that mainly see the fast variable without the drift of the slow variable (as we see in the figures....).
+%
+However, Mahalanobis for NIV requires time windows that mainly see the translation of the intrinsic/NIV noise to the ambient space {\em without} seeing the drift of the slow or the curvature of the transformation (the local linearization assumption should hold, as we will show using a Taylor expansion).
+
+\begin{figure}
+\includegraphics[width=\textwidth]{vary_window_sizes_sde2}
+\caption{This is a completely different system (chemical reaction network example)}
+\label{fig:1d1drxnnetwork}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=0.5\textwidth]{raw_data} \\
+\includegraphics[width=0.5\textwidth]{NIV_data}
+\includegraphics[width=0.5\textwidth]{DMAPS_data}
+%\includegraphics[width=0.5\textwidth]{NIV_corr}
+%\includegraphics[width=0.5\textwidth]{DMAPS_corr}
+%\includegraphics[width=0.5\textwidth]{NIV_embed}
+%\includegraphics[width=0.5\textwidth]{DMAPS_embed}
+\caption{Top: raw data, colored by time.
+Bottom, left: Raw data, colored by first NIV.
+The first NIV captures the slow variable.
+We use $\epsilon = 0.0001$, $dt = 0.00001$, and time window $0.001$.
+Bottom, right: Raw data, colored by first DMAPS.
+The first DMAPS captures the fast variable.
+test\_scale\_separation.m}
+\label{fig:1d1dexample}
+\end{figure}
+
+\begin{figure}
+\includegraphics[width=\textwidth]{vary_windows_halfmoons}
+\caption{Varying sampling windows with half moons.
+We can always recover the slow variable; only when the windows are smaller than the equilibrium measure of the fast do we also recover the fast variable.
+test\_half\_moons.m}
+\label{fig:halfmoons}
+\end{figure}
+
+\section{Recovering NIVs}
 
@@
 \begin{eqnarray}
-	dx_1 & = & dW_1 \\
-	dx_2 & = & dW_2
+	dx_1 & = & a(x_1) dt + dW_1 \\
+	dx_2 & = & a(x_2) dt + dW_2
 \end{eqnarray}
-where $W_1, W_2$ are independent Browinain motions, $x_1, x_2 \in [0, 1]$, and we impose reflective boundary conditions.
+where $a(x) = -100x^{99}$, $W_1, W_2$ are independent Browinain motions.
+%
+INstead of imposing reflective boundary conditons (like in the NLICA paper), $a$ is a potential that is very shallow in $[-1, 1]$, and very steep outside of this interval.
+%
+Therefore, we (effective) sample a square when simulating this SDE.
+%
+We then shift $x$
+\begin{equation}
+x \mapsto (x-1)/2
+\end{equation}
 
@@
 The data are shown in Figure \ref{fig:data}.
+%
+TODO: change the figure so that it has the new data (SDE with shallow potential rather than Brownian motion with reflective bc)
 
@@
 Note that we now recover the ``correct'' variables.
+%
+TODO: we need to do this with covariance estimates from time series, rather than short busts.
+We need to also add the comparison between the analytical covariacne and the estimated covariances (based on differences from a single trajectory and also on bursts).
 
@@
 %
-Consider adding white Gaussian noise to the $y_1, y_2$ data
+Consider adding noise which saturates at some time scale
 \begin{eqnarray}
-z_1 & = & x_1 + x_2^3 + \xi_1 \\
-z_2 & = & x_2 - x_1^3 + \xi_2 \\
-z_3 & = & \xi_3
+z_1(t) & = & r\frac{x_1(t) + x_2(t)^3}{\sqrt{(x_1(t)+x_2(t)^3)^2 + (x_2(t)-x_1(t)^3)^2+1}} + \xi_1(t) \\
+z_2(t) & = & r\frac{x_2(t) - x_1(t)^3}{\sqrt{(x_1(t)+x_2(t)^3)^2 + (x_2(t)-x_1(t)^3)^2+1}} + \xi_2(t) \\
+z_3(t) & = & r\frac{1}{\sqrt{(x_1(t)+x_2(t)^3)^2 + (x_2(t)-x_1(t)^3)^2+1}} + \xi_3(t)
 \end{eqnarray}
-where $\xi_1, \xi_2, \xi_3 \sim \mathcal{N}(0, \sigma^2)$.
+%
+where $r$ is the radius of the sphere onto which we project, and
+\begin{equation}
+d \xi_i(\tau) = L a(\xi_i) d\tau + \frac{1}{\sqrt{\epsilon}} d\tilde{W}_i
+\end{equation}
+TODO: fix this to match the scaling in the code (scaled time and length).
 
-If $\sigma = 0.01$ (so the added noise is much smaller than the noise in the original clouds), we obtain a parameterization shown in Figure \ref{fig:xdata_NIV_noise1} using NIV.
+In conclusion, when this no longer works (when we cannot reover NIVs from a single application of Mahalanobis/NLICA), we need to introduce intermediate representations for our data that averages out the fast noise.
+
+\begin{figure}[h!]
+\includegraphics[width=0.5\textwidth]{embeddings_vary_with_noise_r1}
+\includegraphics[width=0.5\textwidth]{error_with_noise}
+\includegraphics[width=0.5\textwidth]{embeddings_vary_with_noise_r01}
+\includegraphics[width=0.5\textwidth]{error_with_noise_01}
+\caption{Example from Amit's NLICA paper, section 6.2 (Mushroom on sphere).
+Top, left: sphere of radius $r=1$.
+Change scale of ambient noise.
+As ambient noise increases, the embedding becomes poorer, since the ambient noise hides the shape of the mushroom originated by the intrinsic noise.
+The scale of the intrinsic noise is 0.1.
+Top, right: Error in inverse covariances relative to the inverse covariances when there is no noise.
+We see where the embedding breaks down (corresponds to increase in the error).
+Bottom, left: same thing, for $r=0.1$.
+The embedding breaks down at a smaller size of noise, due to (1) the nonlinear transformation is more curved and the covariance estimation is less accurrate, and (2) the effective scale of the intrinsic noise is smaller.
+Bottom, right: Error in the inverse covariances.
+The embedding breaks down when the error increases substantially.
+TODO: Compare with the analytic covariance (maybe through the eigenvectors/singular vectors).
+test\_2d\_1d\_example.m}
+\end{figure}
+
+%If $\sigma = 0.01$ (so the added noise is much smaller than the noise in the original clouds), we obtain a parameterization shown in Figure \ref{fig:xdata_NIV_noise1} using NIV.
+%%
+%When we do NIV, the ``ellipses'' we see are a combination of the original noises in the intrinsic variables and the noise in the ambient space.
+%%
+%Because the noise in the ambient space is significantly smaller than the noise in the intrinsic variable space, we can still ``see'' the directions of the intrinsic variable noises in the ellipses, and therefore uncover the intrinsic variables.
 %
-When we do NIV, the ``ellipses'' we see are a combination of the original noises in the intrinsic variables and the noise in the ambient space.
-%
-Because the noise in the ambient space is significantly smaller than the noise in the intrinsic variable space, we can still ``see'' the directions of the intrinsic variable noises in the ellipses, and therefore uncover the intrinsic variables.
-
-\begin{figure}[H]
-\includegraphics[width=0.3\textwidth]{xdata_noise1_colored_NIV1}
-\includegraphics[width=0.3\textwidth]{xdata_noise1_colored_NIV2}
-\includegraphics[width=0.3\textwidth]{embedding_noise1}
-\caption{Data in ``intrinsic variable'' space ($x_1, x_2$), colored by the first (left) and second (center) nontrivial NIV.
+%\begin{figure}[H]
+%\includegraphics[width=0.3\textwidth]{xdata_noise1_colored_NIV1}
+%\includegraphics[width=0.3\textwidth]{xdata_noise1_colored_NIV2}
+%\includegraphics[width=0.3\textwidth]{embedding_noise1}
+%\caption{Data in ``intrinsic variable'' space ($x_1, x_2$), colored by the first (left) and second (center) nontrivial NIV.
 The NIV embedding is computed from the data in the ambient space with added noise ($z_1, z_2$, with $\sigma = 0.01$).
@@
 }
-\label{fig:xdata_NIV_noise1}
-\end{figure}
-
-If $\sigma = 100$ (so the added noise is much larger than the noise in the original clouds), we obtain a parameterization shown in Figure \ref{fig:xdata_NIV_noise2} using NIV.
+%\label{fig:xdata_NIV_noise1}
+%\end{figure}
 %
-Now, the ellipses are a combination of the intrinsic variable noise and the ambient space noise, but the ambient space noise is {\em much} larger. 
+%If $\sigma = 100$ (so the added noise is much larger than the noise in the original clouds), we obtain a parameterization shown in Figure \ref{fig:xdata_NIV_noise2} using NIV.
+%%
+%Now, the ellipses are a combination of the intrinsic variable noise and the ambient space noise, but the ambient space noise is {\em much} larger. 
+%%
+%Therefore, NIV only collapses the ambient space noise (not the intrinsic variable noise), and so we get something like the DMAPS coordinates, rather than the NIV. 
 %
-Therefore, NIV only collapses the ambient space noise (not the intrinsic variable noise), and so we get something like the DMAPS coordinates, rather than the NIV.
-
-\begin{figure}[H]
-\includegraphics[width=0.3\textwidth]{xdata_noise2_colored_NIV1}
-\includegraphics[width=0.3\textwidth]{xdata_noise2_colored_NIV2}
-\includegraphics[width=0.3\textwidth]{embedding_noise2}
-\caption{Data in ``intrinsic variable'' space ($x_1, x_2$), colored by the first (left) and second (center) nontrivial NIV.
+%\begin{figure}[H]
+%\includegraphics[width=0.3\textwidth]{xdata_noise2_colored_NIV1}
+%\includegraphics[width=0.3\textwidth]{xdata_noise2_colored_NIV2}
+%\includegraphics[width=0.3\textwidth]{embedding_noise2}
+%\caption{Data in ``intrinsic variable'' space ($x_1, x_2$), colored by the first (left) and second (center) nontrivial NIV.
 The NIV embedding is computed from the data in the ambient space with added noise ($z_1, z_2$, with $\sigma = 100$).
@@
 We do {\em not} recover the unit square.}
-\label{fig:xdata_NIV_noise2}
-\end{figure}
+%\label{fig:xdata_NIV_noise2}
+%\end{figure}
+
+\section{Analysis using Stochastic Calculus}
+
+We will use the model described in \eqref{eq:NIV_model}, where $\theta$ are our intrinsic variables, and our observations will be $x = f(\theta) + \xi$.  
+%
+We assume that the timescale of the noise $\xi$ is faster than the timescale of the intrinsic variables $\theta$.
+
+%We have a 2-scale problem (assuming $s_2 > s_1$). This problem can be extended to multiscale problems naturally.
+
+%The problem: the NLICA analysis is based on the following linearization of the measurement function (for simplicity around $\theta = 0$):
+%\[
+%	x_t = \nabla f \theta _t + \frac{1}{2} \theta_t ^T \nabla ^2 f \theta_t + \epsilon _t
+%\]
+%and assume $\epsilon _t$ is small.
+
+Applying the Ito lemma gives us
+\begin{equation}
+dx_t^j =  \sum_{i=1}^d \left( \frac{1}{2}  \frac{\partial ^2 f^j}{\partial \theta_i^2} + a_t^i \frac{\partial f^j}{\partial \theta_i} \right) dt + \sum_{i=1}^{d} \frac{\partial f^j}{\partial \theta_i} dW_t^i + d \xi_i(t)
+\end{equation}
+
+Assuming that the derivatives are constant in the time interval $\delta t$,
+\begin{eqnarray}
+\hat{\mu}_t^j = \frac{1}{\delta t} \int_{t}^{t + \delta t} dx_t^j &=& 
+\frac{1}{\delta t} \int_{t}^{t + \delta t} \sum_{i=1}^d \left( \frac{1}{2}  \frac{\partial ^2 f^j}{\partial \theta_i^2} + a_t^i \frac{\partial f^j}{\partial \theta_i} \right) dt + \sum_{i=1}^{d} \frac{\partial f^j}{\partial \theta_i} dW_t^i + \frac{1}{\epsilon} d \hat{W}^j_t \\
+&=& \frac{1}{\delta t} \sum_{i=1}^d \left( \frac{1}{2}  \frac{\partial ^2 f^j}{\partial \theta_i^2} + a_t^i \frac{\partial f^j}{\partial \theta_i} \right) \delta t + \frac{1}{\delta t} \sum_{i=1}^{d} \frac{\partial f^j}{\partial \theta_i} V_i + \frac{1}{\delta t \epsilon} V_{d+1}\\
+&=& \sum_{i=1}^d \left( \frac{1}{2}  \frac{\partial ^2 f^j}{\partial \theta_i^2} + a_t^i \frac{\partial f^j}{\partial \theta_i} \right) + \mathcal{N}\left( 0, \frac{1}{\delta t} \sum_{i=1}^d \left( \frac{\partial f^j}{\partial \theta_i} \right)^2 + \frac{1}{\delta t \epsilon^2} \right)\\
+&=& \sum_{i=1}^d \left( \frac{1}{2}  \frac{\partial ^2 f^j}{\partial \theta_i^2} + a_t^i \frac{\partial f^j}{\partial \theta_i} \right) + U^j
+\end{eqnarray}
+where $V_i \sim \mathcal{N}(0, \delta t)$ and $U^j \sim \mathcal{N}\left( 0, \frac{1}{\delta t} \sum_{i=1}^d \left( \frac{\partial f^j}{\partial \theta_i} \right)^2 + \frac{1}{\delta t \epsilon^2} \right)$.
+
+\begin{eqnarray}
+dx_t^j dx_t^k &=& 
+\left( \sum_{i=1}^d \left( \frac{1}{2}  \frac{\partial ^2 f^j}{\partial \theta_i^2} + a_t^i \frac{\partial f^j}{\partial \theta_i} \right) dt + \sum_{i=1}^{d} \frac{\partial f^j}{\partial \theta_i} dW_t^i + \frac{1}{\epsilon} d \hat{W}^j_t \right) \\
+&& \left( \sum_{i=1}^d \left( \frac{1}{2}  \frac{\partial ^2 f^k}{\partial \theta_i^2} + a_t^i \frac{\partial f^k}{\partial \theta_i} \right) dt + \sum_{i=1}^{d} \frac{\partial f^k}{\partial \theta_i} dW_t^i + \frac{1}{\epsilon} d \hat{W}^k_t \right) \\
+&=& \sum_{i=1}^d \frac{\partial f^j}{\partial \theta_i} \frac{\partial f^k}{\partial \theta_i} dW_t^i dW_t^i + \frac{\delta_{jk}}{\epsilon^2}  d \hat{W}^j_t d \hat{W}^k_t  \\
+&=& \sum_{i=1}^d \frac{\partial f^j}{\partial \theta_i} \frac{\partial f^k}{\partial \theta_i} dt + \frac{\delta_{jk}}{\epsilon^2}  d t
+\end{eqnarray}
+
+\begin{eqnarray}
+\frac{1}{\delta t} \int_{t}^{t + \delta t} dx_t^j dx_t^k &=& \frac{1}{\delta t} \int_{t}^{t + \delta t} \sum_{i=1}^d \frac{\partial f^j}{\partial \theta_i} \frac{\partial f^k}{\partial \theta_i} dt + \frac{\delta_{jk}}{\epsilon^2}  d t  \\
+&=& \sum_{i=1}^d \frac{\partial f^j}{\partial \theta_i} \frac{\partial f^k}{\partial \theta_i}+ \frac{\delta_{jk}}{\epsilon^2}
+\end{eqnarray}
+
+\begin{eqnarray}
+\hat{C}_t^{jk} &=& \frac{1}{\delta t} \int_{t}^{t + \delta t} dx_t^j dx_t^k -  \hat{\mu}_t^j \hat{\mu}_t^k \\
+&=& \sum_{i=1}^d \frac{\partial f^j}{\partial \theta_i} \frac{\partial f^k}{\partial \theta_i}+ \frac{\delta_{jk}}{\epsilon^2} - \\
+&& \left(\sum_{i=1}^d \left( \frac{1}{2}  \frac{\partial ^2 f^j}{\partial \theta_i^2} + a_t^i \frac{\partial f^j}{\partial \theta_i} \right) + U^j \right) \left(\sum_{i=1}^d \left( \frac{1}{2}  \frac{\partial ^2 f^k}{\partial \theta_i^2} + a_t^i \frac{\partial f^k}{\partial \theta_i} \right) + U^k \right)
+\end{eqnarray}
+
+Define 
+\begin{equation}
+J = \left[ \frac{\partial f^j}{\partial \theta_i} \right]_{ji}
+\end{equation}
+
+\begin{equation}
+H = \left[ \frac{\partial^2 f^j}{\partial \theta_i^2} \right]_{ji}
+\end{equation}
+
+Then
+\begin{eqnarray}
+\hat{C}_t &=& J J^T + \frac{1}{\epsilon^2} I - \left( \frac{1}{2} H \mathbf{1} + J a + U \right) \left ( \frac{1}{2} H \mathbf{1} + J a + U \right)^T \\
+&=& J J^T + \frac{1}{\epsilon^2} I - \left[ \frac{1}{4} H \mathbf{1} \mathbf{1}^T H^T + J a a^T J^T + U U^T \right. \\
+&+& \left. U \left ( \frac{1}{2} H \mathbf{1} + J a  \right)^T + \left( \frac{1}{2} H \mathbf{1} + J a \right) U^T  + \frac{1}{2} H \mathbf{1} a^T J^T + \frac{1}{2} J a \mathbf{1}^T H^T \right]
+\end{eqnarray}
+
+TODO: fix the above analysis (1) make sure that the calculus is right, (2) are we calculating the covariance of the differences?
+(3) verify/figure out why the trends are not what we expect (dependency on curvature, drift, scale of noise, $\delta t$, etc.)
+
+For NLICA to work properly, we need to define clouds of points that respect the shape of the distortion of $f$ applied to the low dimensional Gaussian of $O(1)$ in the NIVs domain.
+Obviously, we cannot take clouds with scales smaller than $1/\epsilon$, because then we only see the Gaussian in the ambient domain.
+Thus, we must take larger scales.
+On the other hand, large scales could include the intrinsic drift (and possibly the nonlinear parts of the measurement function), and therefore, we would want to take a small as possible scale (but greater than $1/\epsilon$).
+%Consider the smallest possible choice -- clouds of scale $s_2$ and observe:
+%\begin{eqnarray}
+%	x_{t+s_2} &=&  \nabla f \theta _{t+s_2} + \frac{1}{2} \theta_{t+s_2} ^T \nabla ^2 f \theta_{t+s_2} \\
+%	&=& \nabla f (\theta _t + s_2 a _t + s_2 s_1 \tilde{\xi}) \\
+%	&+& \frac{1}{2} (\theta _t + s_2 a _t + s_2 s_1 \tilde{\xi}) ^T \nabla ^2 f (\theta _t + s_2 a _t + s_2 s_1 \tilde{\xi}), \ \tilde{\xi} \sim \mathcal{N}(0,I)
+%\end{eqnarray}
+%TODO: replace the last equation with the ito lemma application from the NLICA paper (eq. (4)). Then, fix the following, and also make it scalar (one coordinate of $x_t$).
+%The empirical mean is (for each coordinate):
+%\begin{eqnarray}
+%	\hat{\mu}_t &=& \frac{1}{N} \int \limits _{t+\tau, \tau \in [-s_2,s_2]} x_{t+\tau}, \ \ N = |[-s_2,s_2]| \\
+%		&=& \frac{1}{N} \int \limits _{t+\tau, \tau \in [-s_2,s_2]} ( \nabla f \theta _{t} + \frac{1}{2} \theta_{t} ^T \nabla ^2 f \theta_{t}) \\
+%		&+& \frac{1}{N} \int \limits _{t+\tau, \tau \in [-s_2,s_2]}  (\nabla f(0) \tau a _t + \tau^2 a_t^T \nabla ^2 f a_t ) \\
+%		&+& \frac{1}{N} \int \limits _{t+\tau, \tau \in [-s_2,s_2]}  (\nabla f(0) \tau s_1 \tilde{\xi} + \tau^2 s_1^2 \tilde{\xi}^T \nabla ^2 f \tilde{\xi}) \\
+%		&=& x_t + a_t^T \nabla ^2 f a_t \frac{2 s_2^3}{3} + s_1^2 \mathrm{tr} (\nabla ^2 f) \frac{2 s_2 ^3} {3}
+%\end{eqnarray}
+%We need to repeat that for the empirical covariance.
+However, at this point we can observe that: (1) the intrinsic drift, the intrinsic scale, the Hessian of the measurement function, and the chosen/fast scale, play a role and distort the cloud.
+(2) In order for NLICA to work (again, we need to verify for the covariance and write with the all the details) we require that $1 / s_2 ^3 \sim \nabla^2 f s_1^2, \nabla^2 f a_t^2$.
+
+In conclusion, given a problem in which these conditions/relations between the scales and the measurement function and the drift do not hold, we cannot recover NIVs from the measurements in a single application of NLICA/Mahalanobis graph.
+
+\section{Intermediate Variables and Iterative Algorithm}
+
+Basically, in the ambient space, we get such clouds of data:
+\[
+	e^{-z^T(C_y(t) + C_\xi(t))^{-1} z}
+\]
+where $C_y(t)$ is the covariance of the intrinsic noise translated into the ambient space (including the ``curvature" of the nonlinear measurement function), and $C_\xi(t)$ is the covariance of the ambient noise.
+%
+The point is, for small $t$ ($t < 1/\epsilon$), the size of the ambient noise could be much larger than the intrinsic noise (and thereofre, $C_\xi \gg C_y$). 
+%
+In this regime, we would not get good estimates of $C_y$, which means we will not see the intrinsic noise or recover NIV.
+%
+However, if we take a larger $t$ ($t > 1/\epsilon$), we may begin to see the curvature of the measurement function, and therefore cannot obtain good estimates of the {\em local} covariance $C_y$.
+
+The signal $y$ is corrupted with noise $\xi$.
+%
+We assume that the noise is faster than the signal. 
+%
+We also assume that the noise saturates/have a fixed length scale.
+%
+We use these two assumptions to separate between the signal and the noise by averaging/computing expected values and slowing down the time scale of the measurements.
+
+Since the measurements of the system are high--dimensional, we cannot compute accurate expected values using the (relatively) small number of samples.
+%
+So, we first project the data into a lower--dimensional space, where the averaging will be more robust/efficient/better.
+%
+
+For example, we could project the data onto random ``histogram'' boxes such that the projected data in those boxes such that the intrinsic noise is fixed, and we see only the ambient noise (and the ambient noise has saturated).
+%
+We would then use the projections as our new measurments/observations, and recover their NIVs, which would be (by definition) the NIVs of the original (slow) data.
+
+An empirical way to do it: 1.
+Choose random boxes.
+2.
+pick the time scale according to Mauro's method - (local PCA of the boxes) - start with a large scale (again, when the data is restricted to the low d boxes), and then start decreasing the scale, until we get a fixed number of dimensions with a fixed spectrum.
+At large scales - the spectrum would ``see" the shape of the manifold (depending on the curvature of the nonlinear measurement function embedded/restricted in the box) and at smaller scales, the restriction of the manifold to the box will be flat and then we would see just the restricted ambient noise (giving a fixed spectrum).
+All we need to do is to choose enough boxes, in different scales, such that in just few, the description above would hold.
+
+Note: we could also do this by starting with a small time scale, and then increasing until we see the saturation of the fast process/fast noise. 
+%
+However, this requires us to have a sufficient number of samples to accurately estimate the covariance at this small time scale.
+
+\subsection{Algorithm}
+
+Input: $x_t$.
+
+\begin{enumerate}
+
+	\item
+	Draw filters $\psi _k$, k = 1, \ldots, K.
+
+	\item
+	\label{algo:iter_begin}
+	Pick (and peal) the ``finest time scale":
+
+	Pick a large scale $\delta t$, aiming for a scale $\delta t > 1/\epsilon$.
+
+	BEGIN WHILE $\| C_j - C_{j-1} \| > thr$
+
+	\begin{enumerate}
+		\item
+		Ideally, if $\delta t < 1/\epsilon$, then NLICA failed and we probably need better observables.
+In practice, $1/\epsilon$ is unknown, and therefore, we take enough random observables such that at least few will be good.
+
+		\item
+		Project the data onto $\psi _k$ and get the ``random projections": $o_{t,k} = \langle x_t, \psi _k \rangle$.
+
+		\item
+		Compute the (empirical) covariance $C_j$ of $o_{t,k}$ in a time window of size $\delta t$. This is the variance+bias of the estimator of $\mathbb{E}[o_{t,k}]$ in this window of scale $\delta t$.
+
+		\item
+		Decrease $\delta t$.
+
+	\end{enumerate}	
+	END WHILE
+
+	Note: instead of looking for convergence of $C$, we could look for convergence of the eigenvalues of $C$ to examine when $C$ is (approximately) an $n$-dimensional Gaussian.
+
+	Explanation: Note, clouds of $x$ samples with scale $1/\epsilon$ see the ambient noise more than the intrinsic noise (after $1/\epsilon$ time units, the ambient noise is saturated, BUT we might begin to see the intrinsic drift or curvature of $f$.
+This is the reason why NLICA fails.).
+	%
+Our goal is to find the timescale at which the fast noise is saturated (so that different clouds only change due to effects from the slow/intrinsic process). 
+%
+When the fast noise has saturated, changing the timescale $\delta t$ of the clouds should not (significantly) change the covariance of the clouds (this is the termination condition for our algorithm).
+
+We start with a large scale ($\delta t > 1/\epsilon$). 
+%
+The goal of this step in the algorithm is to find $1/\epsilon$, the saturation time of the fast noise.
+%
+We assume that, when $\delta t > 1/\epsilon$, the clouds/covariances will see the drift of the intrinsic variables and the curvature of the measurment function, and so the covariance will change from iteration to iteration.
+%
+When our assumptions hold within the timescale $\delta t$ (there is little drift and the measurement function is relatively flat), and when the fast noise $\xi$ is saturated, the covariances should not change significantly from iteration to iteration.
+ %If at any point, we discover the clouds of our new observables $o_{t,k}$ with scale smaller or equal $s_2$ still see the intrinsic drift (or curvature), then, these observable are not useful. 
+ %In practice, $s_2$ is unknown and therefore we cannot check this case. 
+% 
+ The idea of the iterative procedure is to find the largest possible scale of clouds of observables that does not see the intrinsic drift and the curvature. 
+We examine that by searching a cloud that is relatively constant (with respect to the time window) in the dimension of the ambient space. 
+%
+In practice, this is done by looking at the change in the covariances as we change the time window.
+
+	\item
+	Compute the estimators of the observables $\hat{o}_{t,k} = \hat{\mathbb{E}}[o_{t,k}]$ in the chosen scale $\delta t^*$ by integrating in windows of this size.
+
+     \item
+	Apply the Mahalanobis graph to the new observables $\hat{o}_{t,k}$ and examine whether the eigenvectors recover NIVs.
+Note: (1) The observables are at a slower rate than the raw measurements $x$.
+(2) This step cannot be verified in practice (?) since the NIVs are unknown.
+
+	\item
+	If yes -- stop, If no (then, we are dealing with the case described above - conditions are not met) -- proceed
+
+	\item
+	Assign $\hat{o}_{t,k} \mapsto x$ and jump to step \ref{algo:iter_begin}.
+
+\end{enumerate}
+
+TODO: implement algorithm, verify that it works
+
+\subsection{Histograms}
+
+Just a connection to previous work. Note:
+\begin{itemize}
+
+	\item
+	The above procedure does not generalize the introduction of histograms as intermediate observables. For histograms, we need to replace (or generalize) the linear project onto $\psi_k$ by
+	\[
+		\hat{o}_{t,k} = \int \langle x _t , \psi _k \rangle p(x) dx \mapsto \int \psi_k( x _t) p(x) dx
+	\]
+	Then, by setting $\psi _k (x) = \mathbf{1}_{b_k} (x) $, where $b_k$ is some histogram bin (box) in the ambient space, we get
+	\[
+		\hat{o}_{t,k} = \int \mathbf{1} _{b_k} ( x _t) p(x) dx = \int \limits _{b_k} p(x) dx = \mathcal{L} (p(x))
+	\]
+	where $\mathcal{L}$ is a linear operator.
+
+	\item
+	However, the histograms do have a nice property (which we do not loose by replacing it with linear projection onto random vectors, but we do loose it if we generalize and use nonlinear transforms) -- the linearity of ambient noises.
+For example
+	\[
+		x_t = f(\theta _t ) + \xi _t = s_t + \xi_t \rightarrow p(x) = p(s)*p(\xi) = \mathcal{L}_{\xi} (p(s))
+	\]
+	Assuming $\xi$ is stationary (at least locally), we get
+	\[
+		\hat{o}_{t,k} = \mathcal{L} (p(x)) = \mathcal{L} ( \mathcal{L}_{\xi} (p(s)))
+	\]
+	Meaning, the observables are linear transformations of the statistics of the signal (and the ambient noise is in the linear transformation).
+Then, by using the Mahalanobis distance, which is invariant to linear transformations, we get a procedure that is independent of the ambient noise.
+
+	\item
+	We do not have to use all the histogram bins (i.e., to tile the entire ambient  space) -- all we need is a coverage that ``sees enough".
+In addition, the ``bins" can be just random directions of any dimension.
+
+\end{itemize}
 

### 2014-03-21 reducible_SDE/covariances_diff.m

@@ -1,0 +1,32 @@
+function [inv_c, new_data, ranks] = covariances_diff(data, knn, D, stride)
+
+% Find neighborhoods
+[M, N] = size(data);
+
+m = floor(M / stride);
+
+% compute covariances
+inv_c = zeros(N, N, m);
+
+new_data = zeros(m, N);
+ranks = zeros(m, 1);
+
+for i = 1:stride:M
+    idx = ceil(i / stride);
+    
+    lind = max(i-knn, 1);
+    rind = min(i+knn, M);
+    
+    neighbors = data(lind+1:rind,:)-data(lind:rind-1,:);
+    
+    c = cov(neighbors);
+    
+    [U, S, V] = svd(c);
+        inv_c(:,:,idx) = U(:,1:D) * diag(1./diag(S(1:D,1:D))) * V(:,1:D)';
+    new_data(idx,:) = data(i,:);
+    ranks(idx) = rank(c);
+    
+    
+    
+end
+

### 2014-03-21 reducible_SDE/test_2d_1d_example.m

@@ -1,0 +1,153 @@
+clear all
+close all
+
+rng(123);
+
+b = 1;
+a = @(x) -100*b*x^99;
+
+drift = @(t, x) [a(x(1)); a(x(2))];
+diffn = @(t, x) [1 0; 0 1];
+
+theta0 = [0; 0];
+tmax = 10;
+
+dim = 2;
+subsample_step = 80;
+
+dt = 0.0001;
+nsteps_per_step = 10;
+
+SDE = sde(drift, diffn, 'StartState', theta0);
+nPeriods = ceil(tmax / dt);
+[theta, t, Z] = SDE.simulate(nPeriods, 'DeltaTime', 4*dt, 'NSTEPS', nsteps_per_step);
+
+theta = (theta+1)/2;
+
+figure; scatter(theta(:,1),theta(:,2), 50, t,'.')
+xlabel('x_1')
+ylabel('x_2')
+title('Colored by time')
+
+%%
+window_time = 0.1;
+window_steps = round(window_time/2 / dt);
+
+start_idx = 60;
+end_idx = start_idx+window_steps;
+
+figure; 
+hist(theta(start_idx+1:end_idx,1)-theta(start_idx:end_idx-1,1),20)
+
+figure; plot(theta(:,1),theta(:,2),'.')
+hold on
+plot(theta(start_idx:end_idx,1),theta(start_idx:end_idx,2),'.k', 'markersize', 10)
+
+
+
+%%
+%f = @(x) [(x(:,1)+1)/2+((x(:,2)+1)/2).^3 (x(:,2)+1)/2-((x(:,1)+1)/2).^3 zeros(size(x,1), 1)];
+
+r = 0.1;
+f = @(x) [r*(x(:,1)+x(:,2).^3)./sqrt((x(:,1)+x(:,2).^3).^2+((x(:,2)-x(:,1).^3).^2)+1) r*(x(:,2)-x(:,1).^3)./sqrt((x(:,1)+x(:,2).^3).^2+((x(:,2)-x(:,1).^3).^2)+1) r./sqrt((x(:,1)+x(:,2).^3).^2+((x(:,2)-x(:,1).^3).^2)+1)];
+
+
[…]

### 2014-03-21 reducible_SDE/test_half_moons2.m

@@ -1,0 +1,158 @@
+% this code simulates the "half-moons" from "Detecting intrinsic slow variables in stochastic
+% dynamical systems by anisotropic diffusion maps", Singer, Erban,
+% Kevrekidis, and Coifman, PNAS
+
+clear all
+close all
+
+%% define parameters
+a1 = 1e-3;
+a2 = 1e-3;
+a3 = 1e-1;
+a4 = 1e-1;
+
+% initial conditions
+data0 = [0; 1];
+
+% time to integrate
+tmax = 2e3;
+
+%% do simulations
+
+drift = @(t, x) [a1; a3*((0.9-x(2))*(1.1-x(2))^2+(1.1-x(2))^2*(0.9-x(2)))];
+diffn = @(t, x) [a2 0; 0 a4];
+    
+dt = 1;
+nsteps_per_step = 1000;
+
+SDE = sde(drift, diffn, 'StartState', data0);
+nPeriods = ceil(tmax / dt);
+[data, t, Z] = SDE.simulate(nPeriods, 'DeltaTime', dt, 'NSTEPS', nsteps_per_step);
+
+u = data(:,1);
+v = data(:,2);
+
+%f = @(x) [x(:,2).*cos(x(:,1)+x(:,2)-1) x(:,2).*sin(x(:,1)+x(:,2)-1)];
+f = @(x) x;
+
+fdata = f(data);
+x = fdata(:,1);
+y = fdata(:,2);
+
+%% plot results
+
+figure;
+plot(v)
+
+% plot variables, color by time
+figure;
+scatter(x, y ,50,t,'.')
+xlabel('x')
+ylabel('y')
+title('colored by time')
+
+%% calculate NIV
+
+neigs = 60;
+dim = 2;
+
+dt2_range = [1e-3 1e-1 1];
+
+figure;
+plot_idx = 1;
+m_subfig = length(dt2_range);
+n_subfig = 3;
+
+fast_coord = [];
+lambda_ratio = [];
+
+for dt2=dt2_range
+    knn_NIV = 50;
+    nsteps_per_step2 = max([1 round(dt2/1e-3)]);
+
+    data_snippets = zeros(2*knn_NIV+1, size(data, 2), size(data, 1));
+    
+    if dt2 < dt
+        for i=1:size(data, 1) 
+            SDE = sde(drift, diffn, 'StartState', data(i,:)');
[…]

### 2014-04-03 chemotaxis/prl_paper.tex
updated chemotaxis paper; made new figures (eps), wrote more, etc.

@@
-\documentclass[prl, reprint, final]{revtex4-1}
-%\documentclass[12pt]{article}
+\documentclass[prl, reprint, final, showkeys]{revtex4-1}
 
@@
 
+\usepackage{epstopdf}
+
 \begin{document}
 
-\title{Choosing Appropriate Metrics for Diffusion Maps Analysis: A Chemotaxis Case Study}
+\title{Recovering System Dynamics using Diffusion Maps: A Chemotaxis Case Study}
 
@@
 
-We simulate a stochastic system whose dynamics are governed by a single parameter (the rate of switching).
+Using a stochastic model which arises in cellular chemotaxis, we illustrate how dimensionality reduction techniques (diffusion maps) can be used characterize the behavior of a dynamical system.
 %
-Depending of the value of this parameter, the system dynamics range between heat equation dynamics and wave equation dynamics. 
+Depending on the value of a single parameter, the average system dynamics range between heat equation dynamics (where particles diffuse uniformly outwards) and wave equation dynamics (where particle propagate with a fixed speed).
 %
-In the heat equation dynamics, the system quickly equilibrates and the initial conditions play very little role. 
+We demonstrate that diffusion maps with the appropriate observers and metric can recover the two variables (the initial condition and the time) that determine the system's state. 
 %
-In the wave equation dynamics, the initial conditions play a significant role in the trajectory.
+Furthermore, the order in which we recover these variables is indicative of the system dynamics (heat equation versus wave equation) over the timescales of interest.
 %
-We use dimensionality reduction techniques to analyze the results of such simulations.
+%We demonstrate that an essential component of our analysis is working with the right observers (histograms/ensembles, rather than the single position of a particle) and with the right metric between those observers (earth mover's distance, rather than Euclidean distance).
 %
-We demonstrate that an essential component of our analysis is working with the right observers (histograms/ensembles, rather than the single position of a particle) and with the right metric between those observers (earth mover's distance, rather than Euclidean distance).
+%In heat equation mode, we find that time is more important than the initial conditions, as the particles quickly equilibrate.
 %
-We show that, depending on the ``mode'' of the dynamical system, we recover the right factors that control the dynamics.
+%In wave equation mode, we find that the initial conditions are more important, as the initial distribution simply propagates in time.
 %	
-In heat equation mode, the first factor we uncover parameterizes the temporal dynamics, and the second factor parameterizes the initial conditions;
-in wave equation mode, the first factor we uncover parameterizes the initial conditions, and the second factor parameterizes the temporal dynamics; 
+%Furthermore, these two components have a completely different ``nature". 
 %
-Furthermore, these two components have a completely different ``nature". 
-%
-One parameter (the initial condition) is governing the entire trajectory and the other parameter (time) is governing the ``step­--by­--step" evolution within each trajectory.
+%One parameter (the initial condition) is governing the entire trajectory and the other parameter (time) is governing the ``step­--by­--step" evolution within each trajectory.
 
@@
 
-\section{Introduction}
+%\section{Introduction}
 
-In dynamical systems, it is often essential to detect changes in the characteristic behavior of a system.
+Often, systems which are stochastic and noisy on the microscale exhibit coherent dynamics on the macroscale which are smooth and governed by only a few parameters.
 %
-Methods such as bifurcation analysis, etc., have been developed with such problems in mind.
+For example, fluid dynamics, which can be described via the position and velocity of the fluid particles, is modeled using the Navier-Stokes equations which govern the macroscale behavior. 
+%
+However, in general, this mapping from microscale to macroscale is not always obvious.
+%
+We would like to {\em automatically} detect the macroscale behavior from data collected at the microscale using data-driven techniques.
 
-We will demonstrate how we can use data--driven techniques (diffusion maps \cite{coifman2005geometric}) to {\em automatically} detect changes in a system's behavior.
+TODO: write where this has been done in other places, context, etc.
+
+We will demonstrate how we can use data-driven techniques (diffusion maps \cite{coifman2005geometric}) to {\em automatically} detect changes in a system's macroscopic dynamics from microscale data.
 %
-We will illustrate our dynamics through a model problem arising in studies of cellular chemotaxis \cite{othmer2000diffusion}.
+We will illustrate our methodology through a model problem arising in studies of cellular chemotaxis \cite{othmer2000diffusion}.
 %
-In our methods, it will be essential to use the appropriate observers for our system in order to uncover the underlying dynamics.
+In this example, the macroscopic dynamics vary depending on the value of a single system parameter.
+%
+We will demonstrate that diffusion maps uncovers a parameterization of the microscopic data which is consistent with the macroscopic description.
+%
+Furthermore, diffusion maps shows changes in dynamical behavior as we vary the system parameter. 
+%
+For our methods, it will be essential to use the appropriate observers for our system in order to uncover the underlying dynamics.
 %
@@
 
-\section{Problem formulation}
+In this Letter, we will begin by describing the chemotaxis problem.
+%
+We will present the stochastic microscopic model, and show how, for this example, the microscopic model gives rise to a compact macroscopic description governed by a single parameter.
+%
+We will then use diffusion maps to analyze the microscopic simulation results, and show that using the appropriate metric for our data is essential to uncovering a meaningful description of our data.
+%
+Furthermore, we will show that diffusion maps can inform us as to the relative importance of different system variables in different parameter regimes.
 
-\subsection{Chemotaxis ``story''}
+%\section{Problem formulation}
 
-We look at a stochastic velocity jump process that arises in models of chemotaxis \cite{othmer2000diffusion}.
+%\subsection{Chemotaxis ``story''}
+
+We work with a stochastic velocity jump process that arises in models of chemotaxis \cite{othmer2000diffusion}.
 %
-Chemotaxis is the movement of cells (such as bacteria) that is regulated by extracellular sensed signals.
+Chemotaxis is the movement of cells (such as bacteria) which is regulated by extracellular sensed signals.
 %
-Such signals allow cells to accomplish tasks such as finding food and navigating away from toxins.
+These signals allow cells to accomplish tasks such as finding food and navigating away from toxins.
 %
-Biologists are interested in the dynamics of cellular movement and dispersal, and several models have been proposed \cite{othmer1988models, codling2008random}.
+Several models have been proposed to describe the dynamics of cellular movement and dispersal \cite{othmer1988models, codling2008random}.
 %
-We will look at a velocity jump process model of these dynamics.
+We will look at one such model described by a velocity jump process.
 %
-In such models, each cell is initialized with a position and a velocity, and the dynamics of each cell are governed by a stochastic process.
+In this model, each cell is initialized with a position and a velocity, and the dynamics of each cell are governed by a stochastic process.
 %
-In a velocity jump process, at random times, cells will ``turn around'' and switch their velocity (this turning is thought to be driven by extracellular signals from food or toxins). 
+At random times, a cell will ``turn around'' and switch its velocity (this turning is presumed to be driven by extracellular signals). 
 %
-We will analyze the dynamics of collections of such cells/particles.
+%We will analyze the dynamics of collections of such cells/particles.
 
-\subsection{Stochastic particles}
+%\subsection{Stochastic particles}
 
-Assume we have a collection of $N$ particles, each of which has a position and velocity. 
+We have a collection of $N$ particles whose states are defined by their positions and velocities. 
 %
@@
 %
-The velocity of each particle is either $\pm s$, where $s$ is the speed. 
+The velocity of each particle is either $\pm s$, where $s$ is a (fixed) speed. 
 %
@@
 %
-Note that each particle has its own ``clock" (Poisson process).
+%Note that each particle has its own ``clock" (Poisson process).
+We can simulate the dynamics of this stochastic system,
+to obtain our microscopic data from which we will do our analysis.
+%\subsection{Partial differential equations}
+%
+We simulate this stochastic velocity jump process for $N=1000$ particles, and $p \in [0.1, 0.9]$.
+%
+We simulate the process for $t \in [0, 10]$, and record the positions of each particle at uniform/equal time intervals.
+%
+We discard the initial portion of the trajectory corresponding to $t < 1$, as this is very noisy and corresponds to initial relaxation of the system.
+%
+We will consider two sets of simulations.
+%
+In one set, $\lambda = 1$ and $s=1$, and in the other set, $\lambda = 400$, and $s=20$.
 
-\subsection{Partial differential equations}
+%Let $X_{p, \lambda, s}(t)$ denote the vector of positions of the particles at time $t$ with initial right probability $p$, switching rate $\lambda$, and speed $s$.
 
-Let $X_{p, \lambda, s}(t)$ denote the vector of positions of the particles at time $t$ with initial right probability $p$, switching rate $\lambda$, and speed $s$.
-
-Let $\rho(x, t)$ denote the probability density of the particles, and let $\rho^-(x, t)$ and $\rho^+(x, t)$ denote the probability density of the left and right moving particles, respectively.
+For this specific example, we can write down explicitly the macroscopic equation that governs the overall system behavior.
 %
-It can be shown that, as $n \rightarrow \infty$, $\rho(x, t)$ obeys the following set of partial differential equations (PDEs)
+For a large collection of particles ($N \rightarrow \infty$), the system can be described by the probability distribution of the particles.
+%
+Let $\rho(x, t)$ denote the probability density of the particles, and let $\rho^-(x, t)$ and $\rho^+(x, t)$ denote the probability density of left and right moving particles, respectively.
+%
+It can be shown that, as $N \rightarrow \infty$, $\rho(x, t)$ obeys the following set of partial differential equations (PDEs):
 \begin{equation} \label{eqn:coupled_pdes}
@@
 %
-Alternatively, \eqref{eqn:coupled_pdes} can be written as one, second--order PDE
+Alternatively, \eqref{eqn:coupled_pdes} can be written as one, second--order PDE:
 \begin{equation} \label{eq:second_order_pde}
@@
 %
-Therefore, the dynamics are governed {\em only} by a single parameter $\lambda$.
+Therefore, the dynamics of the probability density are governed by a {\em single} parameter $\lambda$.
 
-\subsection{Asymptotic mode analysis}  \label{subsec:mode_analysis}
+%\subsection{Asymptotic mode analysis}  \label{subsec:mode_analysis}
 
@@
 %
-When $\lambda \rightarrow 0$, one can see that the right-hand side of \eqref{eqn:coupled_pdes} tends to 0. 
+When $\lambda \rightarrow 0$, the right-hand side of \eqref{eqn:coupled_pdes} tends to 0, and \eqref{eqn:coupled_pdes} becomes two uncoupled wave equations,
+\begin{equation}
+\begin{aligned}
+\frac{\partial \rho^+}{\partial t} + s \frac{\partial \rho^+}{\partial x} & = 0 \\
+\frac{\partial \rho^-}{\partial t} - s \frac{\partial \rho^-}{\partial x} & = 0
+\end{aligned}
+\end{equation}
+%alternatively, \eqref{eq:second_order_pde} becomes the second order wave equation,
+%\begin{equation}
+%\frac{\partial^2 \rho}{\partial t^2} = s^2 \frac{\partial ^2 \rho}{\partial x^2}.
+%\end{equation}
+
+%Dividing \eqref{eq:second_order_pde} by $\lambda > 0$ yields
+%\[
+%\frac{1}{\lambda} \frac{\partial^2 \rho}{\partial t^2} + 2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2}
+%\]
+When $\lambda \rightarrow \infty$, \eqref{eq:second_order_pde} approaches the heat equation,
+\begin{equation}
+2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2}.
+\end{equation}
 %
-Therefore, \eqref{eqn:coupled_pdes} becomes two uncoupled wave equations.
+From this analysis, we can see that the initial distribution of left- and right-moving particles (characterized by $p$ in our microscopic simulations) plays a very different role depending on the value of $\lambda$.
 %
-Alternatively, in this regime, in the limit the second order equation \eqref{eq:second_order_pde} becomes 
-\[
-\frac{\partial^2 \rho}{\partial t^2} = s^2 \frac{\partial ^2 \rho}{\partial x^2},
-\]
-i.e., the second order wave equation.
+When $\lambda \rightarrow 0$, the dynamics are described by two wave equations, and so the initial distributions persist throughout the trajectory.
+%
+When $\lambda \rightarrow \infty$, the dynamics are described by one heat equation, and so the initial conditions are not important; the velocity distribution quickly equilibrates and we see purely diffusive behavior.
+%
+For this particular system, we can deduce this from the macroscopic PDEs which describe the system dynamics.
+%
+However, we would like to demonstrate that, in general, these PDEs are not necessary for us to uncover which system variables are important to describe the macroscopic dynamics.
 
-Dividing \eqref{eq:second_order_pde} by $\lambda > 0$ yields
-\[
-\frac{1}{\lambda} \frac{\partial^2 \rho}{\partial t^2} + 2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2}
-\]
-When $\lambda \rightarrow \infty$, \eqref{eq:second_order_pde} approaches the heat equation
-\[
-2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2}
-\]
+%\section{Methods and Analysis}
 
-\section{Methods and Analysis}
+%\subsection{Diffusion maps}
 
-\subsection{Diffusion maps}
-
-We will use diffusion maps \cite{coifman2005geometric} to analyze the data from our stochastic simulations.
+We will use diffusion maps \cite{coifman2005geometric} to analyze the data from our microscopic, stochastic simulations.
+%
+We will show that diffusion maps can parameterize the data obtained from our microscopic simulations, and that this parameterization will describe the two system variables $p$ and $t$.
 %
@@
 %
-In our simulations, although the data is very high--dimensional, there are only two important parameters ($p$ and $t$) that determine the current state of the system.
-
-Assume we have data points $z_1, z_2, \dots, z_m \in \mathbb{R}^p$, but that these data points lie on a $d$-dimensional, nonlinear manifold, with $d \ll p$. 
+In our simulations, although the data is very high-dimensional (we have the positions of all $N$ particles), for a fixed value of $\lambda$, there are only two important variables ($p$ and $t$) that determine the current (macroscopic) state of the system; thus, our microscopic data {\em should} be two-dimensional, parameterized by $p$ and $t$.
+%
+For diffusion maps, we assume we have data points $z_1, z_2, \dots, z_m \in \mathbb{R}^n$ which lie on a $d$-dimensional, nonlinear manifold, with $d \ll n$. 
 %
@@
 where $\| \cdot \|$ denotes the appropriate norm for the data (we will discuss the choice of norm in the following section), and $\epsilon$ is a characteristic distance between the data points.
+%
+We take $\epsilon$ to be the median of the pairwise distances between the data points.
 %
@@
 The next few eigenvectors $\phi_2, \phi_3, \dots$ give us the embedding coordinates for our data, such that $\phi_{j+1}(i)$ gives us the $j^{th}$ embedding coordinate for $z_i$.
+%
+$\lambda_i$ gives a measure of the ``importance'' of coordinate $\phi_i$.
 
-As we can see from the analysis in the preceeding section, the ``important'' quantity to describe the dynamics is not the position of each particle, but rather the probability density of the particles.
+We first need to address what the appropriate observers are for our system.
 %
-We will empirically estimate the probability density at each time point in our simulations by computing histograms of the particle positions. 
+Because the particles are indistinguishable, vectors containing the particle positions are not the appropriate observers to analyze, as swapping two particles would change the data.
 %
-Therefore, $z_i \in \mathbb{R}^p$ will be the histogram of $x_1, x_2, \dots, x_N$ into $p$ bins.
+Instead, we use histograms of the particle positions as our data for our analysis, so that $z_i \in \mathbb{R}^n$ will be the histogram of $x_1, x_2, \dots, x_N$ into $n$ bins (for our simulations, we take $n=32$).
+%
+Clearly, histograms are invariant to the labeling or indexing of the particles.
+%
+For this particular example, using histograms is consistent with our macroscopic PDE description, which are written in terms of the probability density of the particles.
+%
+The histograms serve as an empirical estimate of the probability density at each time point in our simulations.
 
-\subsection{Earth mover's distance}
+%\subsection{Earth mover's distance}
 
-The {\em essential} requirement for diffusion maps is having an appropriate distance metric between data points.
-%
-Often, one uses the Euclidean distance between data points;
-however, for histograms which approximate a probability density, the Euclidean distance can often be misleading.
-%
-For example, the distance between two uniform distributions with finite, non-overlapping support will always be the same, {\em independent} of the distance between the support intervals.
-%
-Intuitively, we would want/expect two distributions whose supports are a short distance apart to be closer (in norm) than two distributions whose supports are very far from each other.
-
-We use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between histograms.
-%
-Conceptually, EMD measures how much ``work'' it takes to transform one probability density into another.
-%
-It therefore not only considers where the densities are incosistent, but how far apart they are.
-%
-Although the brute--force computation of the EMD between two histograms is computationally expensive, there has been a plethora of work in developing efficient algorithms for the computation of EMD \cite{...}.
-%
-%TODO: we need to pay a special attention to the implementation of the EMD.
-We need to make it cleat that this distance can be very efficiently implemented  and therefore it is suitable for the practical uses and algorithm that we propose here as a viable tool for dynamical system analysis.
-%
-We use the EMD algorithm from \cite{Pele-eccv2008, Pele-iccv2009} to compute the distances between histograms.
-%
-The implementation is available at \url{http://www.cs.huji.ac.il/~ofirpele/FastEMD/code}.
-
-\section{Results}
-
-We simulate this stochastic velocity jump process for $N=1000$ particles, and $p \in [0.1, 0.9]$.
-%
-We simulate the process for $t \in [0, 10]$, and record the positions of each particle at uniform/equal time intervals.
-%
-We discard the initial portion of the trajectory corresponding to $t < 1$, as this is very noisy and corresponds to initial relaxation of the system.
-%
-We will consider two sets of simulations.
-%
-In one set, $\lambda = 1$ and $s=1$, and in the other set, $\lambda = 400$, and $s=20$.
-
-For each data point, we compute the histogram of the particle positions into $p=32$ evenly spaced bins. 
-%
-In Figure \ref{fig:dmaps_embed_noemd}, we have used the standard Euclidean distance between the histograms to compute the distances for \eqref{eq:W}.
-%
-The embeddings obtained are not particularly informative, and do not accurrately capture the underlying system dynamics.
-%
-This is due to our poor choice of metric, which does not accurrately describe the distances between histograms.
-
-In Figure \ref{fig:dmaps_embed_emd}, we have used the earth mover's distance to compute the distances between histograms.
-%
-One can see that the two important factors, $p$ and $t$, are well--correlated with the first two diffusion maps modes, $\phi_2$ and $\phi_3$. 
-%
-However, the ``more important'' mode, $\phi_2$, is correlated with $p$ for the small $\lambda$ case (a), and correlated with $t$ for the large $\lambda$ case (d).
-This is consistent with the asymptotic analysis discussed previously.
-%
-In the small $\lambda$ regime (wave equation), shown in Figures \ref{subfig:small_lambda_p} and \ref{subfig:small_lambda_t}, we observe that for small time, the initial condition of the particle velocity does not play a significant role -- all the particles start at position $0$, and therefore, for small time, the particles are more condensed and it is more difficult to distinguish the particles moving to the left from the particles moving to the right.
-On the other hand, for large time, once the particles evolve from the origin, this separation is clear.
-%
-For the large $\lambda$ case (heat equation), shown in Figures \ref{subfig:large_lambda_p} and \ref{subfig:large_lambda_t}, we observe that for small time the initial distribution $p$ is well organized in the embedding in Figure \ref{subfig:large_lambda_p}, since for small time the distribution of the particles is skewed and the initial velocity plays a role in this regime.
-%
-On the other hand, for large time, we observe that the initial distribution $p$ is not organized as well in the embedding in Figure \ref{subfig:large_lambda_p}, since the velocities have equilibrated and so the initial distribution is less detectable in the particle distribution.
-
-We would like to note that the crossover between the wave equation regime, where $p$ is the more important parameter, and the heat equation regime, where $t$ is the more important parameter, depends on the time scale of observation.
-%
-To illustate this, we took a fixed value of $\lambda$ and $s$ ($\lambda=100$ and $s=10$), and allowed the system to evolve for 10 and 20 time units. 
-%
-The results are shown in Figure \ref{fig:dmaps_embed_varyt}.
-%
-When the system only evolves for 10 time units, the majority of the data still appears ``wave--like'', and so the dominant parameter, $\phi_2$, is correlated with $p$, the initial condition.
-%
-However, when the system evolves for 20 time units, the behavior is more diffusive and $\phi_2$ is now correlated with $t$.
-
-TODO: add histograms to this figure
-
-\begin{figure*}[htb]
+\begin{figure}[t]
 \begin{subfigure}{0.2\textwidth}
@@
 \label{fig:dmaps_embed_noemd}
-\end{figure*}
+\end{figure}
 
-\begin{figure*}[htb]
-\begin{subfigure}{0.4\textwidth}
+The second point we need to address is choosing appropriate distance metric between data points in \eqref{eq:W}.
+%
+Often, one uses the Euclidean distance between data points.
+%
+In Figure \ref{fig:dmaps_embed_noemd}, we have used the standard Euclidean distance between the histograms to compute the distances for \eqref{eq:W}.
+%
+The embeddings obtained are not particularly informative, and do not capture the underlying system dynamics; there is no strong correlation of the embedding coordinates with $p$ or $t$. 
+%
+This is due to our poor choice of metric, which does not accurately describe the distances between histograms.
+%
+We would prefer a metric which not only considers the density at each bin, but also takes into account the distance {\em between} bins,
+such that two distributions whose supports are a short distance apart are closer (in norm) than two distributions whose supports are very far from each other.
+%
+%however, for histograms which approximate a probability density, the Euclidean distance can often be misleading.
+%
+%For example, the distance between two uniform distributions with finite, non-overlapping support will always be the same, {\em independent} of the distance between the support intervals.
+
+We choose to use the earth mover's distance (EMD) \cite{rubner2000earth} as our metric between histograms.
+%
+Conceptually, EMD measures how much ``work'' it takes to transform one probability density into another.
+%
+It therefore not only considers where the densities are inconsistent, but how far apart the inconsistencies are.
+%
+Although the brute-force computation of the EMD between two histograms is computationally expensive, there has been a plethora of work in developing efficient algorithms for the computation of EMD \cite{...}.
+%
+%TODO: we need to pay a special attention to the implementation of the EMD.
+We need to make it cleat that this distance can be very efficiently implemented  and therefore it is suitable for the practical uses and algorithm that we propose here as a viable tool for dynamical system analysis.
+%
+%For example, the EMD algorithm from \cite{Pele-eccv2008, Pele-iccv2009}....
+%
+%The implementation is available at \url{http://www.cs.huji.ac.il/~ofirpele/FastEMD/code}. 
+For the special case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions \cite{...},
+\begin{equation}
+\| \rho_1  - \rho_2 \|_{EMD} = \int_{-\infty}^{\infty} \left| \int_{-\infty}^x \rho_1(y) dt - \int_{-\infty}^x \rho_2(y) dy \right| dx
+\end{equation}
+where $\rho_1(x)$ and $\rho_2(x)$ are two probability distributions defined on $x \in \mathbb{R}$. 
+%
+This can be estimated using histograms, 
+\begin{equation}
+\| z_1 - z_2 \|_{EMD} = \sum_{i=1}^{n} \left| \sum_{j=1}^i z_1(j) - \sum_{j=1}^i z_2(j) \right|
+\end{equation}
+where $z_1, z_2 \in \mathbb{R}^n$ are two histograms defined on (equally-spaced) bins in $\mathbb{R}$. 
+%\section{Results}
+
+%We simulate this stochastic velocity jump process for $N=1000$ particles, and $p \in [0.1, 0.9]$.
+%%
+%We simulate the process for $t \in [0, 10]$, and record the positions of each particle at uniform/equal time intervals.
+%%
+%We discard the initial portion of the trajectory corresponding to $t < 1$, as this is very noisy and corresponds to initial relaxation of the system.
+%%
+%We will consider two sets of simulations.
+%%
+%In one set, $\lambda = 1$ and $s=1$, and in the other set, $\lambda = 400$, and $s=20$.
+
+%In Figure \ref{fig:dmaps_embed_noemd}, we have used the standard Euclidean distance between the histograms to compute the distances for \eqref{eq:W}.
+%%
+%The embeddings obtained are not particularly informative, and do not accurrately capture the underlying system dynamics.
+%%
+%This is due to our poor choice of metric, which does not accurrately describe the distances between histograms.
+
+\begin{figure}[t]
+\begin{subfigure}{0.45\columnwidth}
 \includegraphics[width=\textwidth]{EMD_withhist_p_1}
@@
 \end{subfigure}
-\begin{subfigure}{0.4\textwidth}
+\begin{subfigure}{0.45\columnwidth}
 \includegraphics[width=\textwidth]{EMD_withhist_t_1}
@@
 \end{subfigure}
-\begin{subfigure}{0.4\textwidth}
+\begin{subfigure}{0.45\columnwidth}
 \includegraphics[width=\textwidth]{EMD_withhist_p_400}
@@
 \end{subfigure}
-\begin{subfigure}{0.4\textwidth}
+\begin{subfigure}{0.45\columnwidth}
 \includegraphics[width=\textwidth]{EMD_withhist_t_400}
@@
 \label{fig:dmaps_embed_emd}
-\end{figure*}
+\end{figure}
 
-\begin{figure*}
-\begin{subfigure}{0.4\textwidth}
-\includegraphics[width=\textwidth]{EMD2_p_10}
+In Figure \ref{fig:dmaps_embed_emd}, we have used the earth mover's distance to compute the distances between histograms.
+%
+One can see that the two important factors, $p$ and $t$, are well-correlated with the first two diffusion maps coordinates, $\phi_2$ and $\phi_3$. 
+%
+However, the ``more important'' coordinate, $\phi_2$, is correlated with $p$ for the small $\lambda$ case (a), and correlated with $t$ for the large $\lambda$ case (d); 
+this is consistent with the macroscopic PDE description.
+%
+Furthermore, in the small $\lambda$ regime (wave equation), shown in Figures \ref{subfig:small_lambda_p} and \ref{subfig:small_lambda_t},
+we can see that the points corresponding to small times are more clustered than the points corresponding to large times.
+%
+This is not surprising; at small times, the particles are more condensed around $x=0$, and it is more difficult to distinguish the particles moving to the left from the particles moving to the right. 
+%
+On the other hand, for large time, once the particles evolve from the origin, this separation is clear.  
+%
+For the large $\lambda$ case (heat equation), shown in Figures \ref{subfig:large_lambda_p} and \ref{subfig:large_lambda_t}, we observe that for small time the initial distribution $p$ is well organized in the embedding in Figure \ref{subfig:large_lambda_p}, since for small time the distribution of the particles is skewed and the initial velocity plays a role in this regime.
+%
+On the other hand, for large time, we observe that the initial distribution $p$ is not organized as well in the embedding in Figure \ref{subfig:large_lambda_p}, since the velocities have equilibrated and so the initial distribution is less detectable in the particle distribution.
+
+\begin{figure}[t]
+\begin{subfigure}{0.45\columnwidth}
+\includegraphics[width=\textwidth]{EMD2_withhist_p_10}
 \caption{}
 \end{subfigure}
-\begin{subfigure}{0.4\textwidth}
-\includegraphics[width=\textwidth]{EMD2_t_10}
+\begin{subfigure}{0.45\columnwidth}
+\includegraphics[width=\textwidth]{EMD2_withhist_t_10}
 \caption{}
 \end{subfigure}
-\begin{subfigure}{0.4\textwidth}
-\includegraphics[width=\textwidth]{EMD2_p_20}
+\begin{subfigure}{0.45\columnwidth}
+\includegraphics[width=\textwidth]{EMD2_withhist_p_20}
 \caption{}
 \end{subfigure}
-\begin{subfigure}{0.4\textwidth}
-\includegraphics[width=\textwidth]{EMD2_t_20}
+\begin{subfigure}{0.45\columnwidth}
+\includegraphics[width=\textwidth]{EMD2_withhist_t_20}
 \caption{}
@@
 The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time.
-The particles are allowed to evolve for (a, b) 10 time units, and (c, d) 20 time units.
-}
+The particles are allowed to evolve for (a, b) 10 time units, and (c, d) 20 time units.}
 \label{fig:dmaps_embed_varyt}
-\end{figure*}
+\end{figure}
 
-\section{Discussion}
+We have illustrated that diffusion maps can elucidate the relative importance of different system variables as a function of $\lambda$.
+%
+For the plots shown in Figure~\ref{fig:dmaps_embed_emd}, we can easily see the crossover from the wave-equation regime, where $p$ is more important, to the heat-equation regime, where $t$ is more important.
+%
+However, we would like to note that this crossover point depends not only on $\lambda$, but also on the timescale of observation.
+%
+We can see this from our PDE description in \eqref{eq:second_order_pde};
+clearly, $\lambda$ is defined relative to $t$, and so we should not only consider $\lambda$ in our analysis, but also the timescale of observation.
+%
+To illustrate this in our simulations, we chose an intermediate value of $\lambda$ ($\lambda=100$ and $s=10$), and allowed the system to evolve for 10 and 20 time units. 
+%
+The results are shown in Figure \ref{fig:dmaps_embed_varyt}.
+%
+When the system only evolves for 10 time units, the majority of the data still appears ``wave-like'', and so the dominant parameter, $\phi_2$, is correlated with $p$, the initial condition.
+%
+However, when the system evolves for 20 time units, the initial velocity distribution has had more time to equilibrate; therefore, the behavior is more diffusive and $\phi_2$ is now correlated with $t$.
+%
+Therefore, we can also see the crossover from wave-like behavior to heat-like behavior as a function of the observation timescale using diffusion maps.
+
+We have presented algorithms to automatically uncover the variables which govern the macroscopic behavior of a system from data sampled at the microscopic level.
+%
+The essential components for our algorithms are choosing the correct observers and selecting the appropriate metric between these observers.
+%
+In our specific example, we used histograms of particle positions and the earth mover's distance between pairs of histograms.
+%
+With these two considerations, we used diffusion maps to uncover the two variables which characterize the system dynamics.
+%
+Furthermore, these two variables have a completely different nature;
+one variable, $p$, characterizes the initial conditions of the system, whereas the other variable, $t$, describes the ``step-by-step'' evolution of the particles. 
+%
+This is reassuring, and promotes the broad applicability of our methods to a wide variety of dynamical systems.
+
+%\section{Discussion}
 

### 2014-04-03 chemotaxis/simulate_jump_process_EMD_varyt2.m

@@ -19,9 +19,10 @@ s = 10;
 N = 1000;
 
 %time step
-dt = 1;
+dt = 2;
 % maximum simulation time
-tmax = 10;
+tmax = 20;
+
 
 nsteps = floor(tmax / dt) + 1;
 
@@ -130,10 +131,10 @@ eps2 = median(W2(:));
 
 [V2, D2] = dmaps(W2, eps2, 10);
 
-if corr(V2(:,2), all_p(idx)') < 0
+if corr(V2(:,2), all_time(idx)') < 0
     V2(:,2) = -V2(:,2);
 end
-if corr(V2(:,3), all_time(idx)') < 0
+if corr(V2(:,3), all_p(idx)') < 0
     V2(:,3) = -V2(:,3);
 end
 
@@ -161,7 +162,7 @@ figure;
 set(gcf,'PaperPositionMode','auto');
 scatter(V2(:,2),V2(:,3),200,all_time(idx), '.');
 ax = gca;
-axis_lim = [-0.4 0.4 -0.25 0.3];
+axis_lim = [-0.25 0.2 -0.35 0.35];
 axis(axis_lim)
 xlabel('$\phi_2$', 'interpreter','latex', 'fontsize', 20)
 ylabel('$\phi_3$', 'interpreter','latex', 'fontsize', 20)
@@ -172,20 +173,20 @@ hist_all2 = hist_all(:, idx);
 hold on
 i = zeros(4, 1);
 i(1) = find(all_time(idx) == min(all_time(idx)) & all_p(idx) == min(all_p(idx)));
-i(2) = find(all_time(idx) == min(all_time(idx)) & all_p(idx) == max(all_p(idx)));
-i(3) = find(all_time(idx) == max(all_time(idx)) & all_p(idx) == min(all_p(idx)));
+i(2) = find(all_time(idx) == max(all_time(idx)) & all_p(idx) == min(all_p(idx)));
+i(3) = find(all_time(idx) == min(all_time(idx)) & all_p(idx) == max(all_p(idx)));
 i(4) = find(all_time(idx) == max(all_time(idx)) & all_p(idx) == max(all_p(idx)));
 
 points_x = curr_ax(1) + (V2(i, 2) - axis_lim(1)) / (axis_lim(2)-axis_lim(1)) * curr_ax(3);
[…]

### 2014-04-04 chemotaxis/prl_paper.tex
updated chemotaxis paper

@@
 %
-Depending on the value of a single parameter, the average system dynamics range between heat equation dynamics (where particles diffuse uniformly outwards) and wave equation dynamics (where particle propagate with a fixed speed).
+Depending on the value of a single parameter, the system dynamics at the macroscale range from heat equation dynamics (where particles diffuse uniformly outwards) to wave equation dynamics (where particle propagate with a fixed speed).
 %
-We demonstrate that diffusion maps with the appropriate observers and metric can recover the two variables (the initial condition and the time) that determine the system's state. 
+We demonstrate that diffusion maps with the appropriate observers and metric can recover the two system variables (the initial condition and the time) from data collected at the microscale. 
 %
@@
 
-Often, systems which are stochastic and noisy on the microscale exhibit coherent dynamics on the macroscale which are smooth and governed by only a few parameters.
+Often, systems which are stochastic and noisy on the microscale exhibit smooth, coherent dynamics on the macroscale which are governed by only a few parameters.
 %
-For example, fluid dynamics, which can be described via the position and velocity of the fluid particles, is modeled using the Navier-Stokes equations which govern the macroscale behavior. 
+For example, fluid dynamics, which can be described via the position and velocity of the fluid particles, is modeled at the macroscale using the Navier-Stokes equations.
 %
@@
 %
-We would like to {\em automatically} detect the macroscale behavior from data collected at the microscale using data-driven techniques.
+We would like to {\em automatically} uncover the macroscale behavior from data collected at the microscale using data-driven techniques.
 
@@
 %
-We will demonstrate that diffusion maps uncovers a parameterization of the microscopic data which is consistent with the macroscopic description.
+We will demonstrate that diffusion maps uncovers a description of the microscopic data which is consistent with the macroscopic model.
 %
@@
 %
-For our methods, it will be essential to use the appropriate observers for our system in order to uncover the underlying dynamics.
+For our techniques to successfully uncover the underlying dynamics, it will be essential to use the appropriate observers for our system.
 %
-Furthermore, we will need to use the appropriate metric between observers.
+Furthermore, we will need to use the appropriate metric between the observers.
 
@@
 %
-Furthermore, we will show that diffusion maps can inform us as to the relative importance of different system variables in different parameter regimes.
+Furthermore, we will show that diffusion maps can inform us as to the relative importance of system variables in different parameter regimes.
 
@@
 %
-At random times, a cell will ``turn around'' and switch its velocity (this turning is presumed to be driven by extracellular signals). 
+At random times, a cell will ``turn around'' and switch its velocity (this turning is controlled by extracellular signals). 
 %
@@
 %Note that each particle has its own ``clock" (Poisson process).
-We can simulate the dynamics of this stochastic system,
-to obtain our microscopic data from which we will do our analysis.
+We simulate the dynamics of this stochastic system to obtain our microscopic data from which we will do our analysis.
 %\subsection{Partial differential equations}
 %
-We simulate this stochastic velocity jump process for $N=1000$ particles, and $p \in [0.1, 0.9]$.
+We will consider two sets of simulations;
+in one set (small $\lambda$), $\lambda = 1$ and $s=1$, and in the other set (large $\lambda$), $\lambda = 400$, and $s=20$.
 %
-We simulate the process for $t \in [0, 10]$, and record the positions of each particle at uniform/equal time intervals.
+For each set, we run many simulations, each with $N=1000$ particles.
 %
-We discard the initial portion of the trajectory corresponding to $t < 1$, as this is very noisy and corresponds to initial relaxation of the system.
+We vary the initial condition of each simulation such that $0.1 \le p  \le 0.9$, and allow each simulation to evolve for $10$ time units.
 %
-We will consider two sets of simulations.
-%
-In one set, $\lambda = 1$ and $s=1$, and in the other set, $\lambda = 400$, and $s=20$.
+We discard the initial portion of each simulation trajectory corresponding to $t < 1$, as this is very noisy and corresponds to initial relaxation of the system.
 
@@
 %
-Let $\rho(x, t)$ denote the probability density of the particles, and let $\rho^-(x, t)$ and $\rho^+(x, t)$ denote the probability density of left and right moving particles, respectively.
+Let $\rho(x, t)$ denote the probability distribution of the particles, and let $\rho^-(x, t)$ and $\rho^+(x, t)$ denote the distributions of left and right moving particles, respectively.
 %
-It can be shown that, as $N \rightarrow \infty$, $\rho(x, t)$ obeys the following set of partial differential equations (PDEs):
+It can be shown that, as $N \rightarrow \infty$, $\rho(x, t)$ obeys the following set of partial differential equations (PDEs) \cite{othmer2000diffusion}:
 \begin{equation} \label{eqn:coupled_pdes}
@@
 %
-We assume that $s^2/\lambda = D$ is constant.
-%
-Therefore, the dynamics of the probability density are governed by a {\em single} parameter $\lambda$.
+We assume that $s^2/\lambda = D$ is constant, so that the dynamics of the probability density are governed by a {\em single} parameter $\lambda$.
 
@@
 \frac{\partial \rho^+}{\partial t} + s \frac{\partial \rho^+}{\partial x} & = 0 \\
-\frac{\partial \rho^-}{\partial t} - s \frac{\partial \rho^-}{\partial x} & = 0
+\frac{\partial \rho^-}{\partial t} - s \frac{\partial \rho^-}{\partial x} & = 0.
 \end{aligned}
@@
 %
-From this analysis, we can see that the initial distribution of left- and right-moving particles (characterized by $p$ in our microscopic simulations) plays a very different role depending on the value of $\lambda$.
+From this analysis, we can see that the initial distribution of left- and right-moving particles (given by $p$ in our microscopic simulations) plays a very different role depending on the value of $\lambda$.
 %
-When $\lambda \rightarrow 0$, the dynamics are described by two wave equations, and so the initial distributions persist throughout the trajectory.
+When $\lambda \rightarrow 0$, the dynamics are described by two wave equations, and the initial distribution persists throughout the trajectory.
 %
-When $\lambda \rightarrow \infty$, the dynamics are described by one heat equation, and so the initial conditions are not important; the velocity distribution quickly equilibrates and we see purely diffusive behavior.
+When $\lambda \rightarrow \infty$, the dynamics are described by one heat equation, and the initial conditions are not important - the velocity distribution quickly equilibrates and we see purely diffusive behavior.
 %
-For this particular system, we can deduce this from the macroscopic PDEs which describe the system dynamics.
+For this particular system, we can deduce this behavior from the macroscopic PDEs which describe the system dynamics.
 %
@@
 %
-We will show that diffusion maps can parameterize the data obtained from our microscopic simulations, and that this parameterization will describe the two system variables $p$ and $t$.
+We will show that diffusion maps can organize the data obtained from our microscopic simulations, and that this organization will be consistent with the two system variables, $p$ and $t$, that we expect from our PDE analysis.
 %
-Diffusion maps is a nonlinear dimensionality reduction technique.
+Diffusion maps, in general, is a nonlinear dimensionality reduction technique.
 %
-In our simulations, although the data is very high-dimensional (we have the positions of all $N$ particles), for a fixed value of $\lambda$, there are only two important variables ($p$ and $t$) that determine the current (macroscopic) state of the system; thus, our microscopic data {\em should} be two-dimensional, parameterized by $p$ and $t$.
+In our simulations, although the data is very high-dimensional (e.g., the positions of all $N$ particles), for a fixed value of $\lambda$, there are only two important variables ($p$ and $t$) that determine the current (macroscopic) state of the system; thus, our microscopic data {\em should} be two-dimensional and described by $p$ and $t$.
 %
@@
 \end{equation}
-where $\| \cdot \|$ denotes the appropriate norm for the data (we will discuss the choice of norm in the following section), and $\epsilon$ is a characteristic distance between the data points.
-%
-We take $\epsilon$ to be the median of the pairwise distances between the data points.
+where $\| \cdot \|$ denotes the appropriate norm for the data (we will discuss the choice of norm in the following section), and $\epsilon$ is a characteristic distance between the data points (we take $\epsilon$ to be the median of the pairwise distances between the data points).
 %
@@
 %
-Note that the matrix $A$ is row--stochastic ($\sum_j A_{ij} = 1$), which implies $\lambda_1 = 1$ and $\phi_1$ is a constant vector.
+Note that the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$), which implies $\lambda_1 = 1$ and $\phi_1$ is a constant vector.
 %
-The next few eigenvectors $\phi_2, \phi_3, \dots$ give us the embedding coordinates for our data, such that $\phi_{j+1}(i)$ gives us the $j^{th}$ embedding coordinate for $z_i$.
+The next few eigenvectors give us the embedding coordinates for our data;
+$\phi_{j+1}(i)$ gives us the $j^{th}$ embedding coordinate for $z_i$ and
+$\lambda_j$ gives a measure of the ``importance'' of coordinate $\phi_j$.
+
+We first need to address what are the appropriate observers or measurements for our system.
 %
-$\lambda_i$ gives a measure of the ``importance'' of coordinate $\phi_i$.
-
-We first need to address what the appropriate observers are for our system.
+Because the particles are indistinguishable, vectors containing the particle positions are not appropriate to analyze, as we would then be assigning indices or labels to the particles and swapping two particles would change the data.
 %
-Because the particles are indistinguishable, vectors containing the particle positions are not the appropriate observers to analyze, as swapping two particles would change the data.
-%
-Instead, we use histograms of the particle positions as our data for our analysis, so that $z_i \in \mathbb{R}^n$ will be the histogram of $x_1, x_2, \dots, x_N$ into $n$ bins (for our simulations, we take $n=32$).
+Instead, we use histograms of the particle positions as our observables, so that $z_i \in \mathbb{R}^n$ will be the histogram of $x_1, x_2, \dots, x_N$ into $n$ bins (for our simulations, we take $n=32$).
 %
@@
 %
-For this particular example, using histograms is consistent with our macroscopic PDE description, which are written in terms of the probability density of the particles.
+For this particular example, using histograms is consistent with our macroscopic PDE description, which is written in terms of the probability density of the particles.
 %
@@
 %
-In Figure \ref{fig:dmaps_embed_noemd}, we have used the standard Euclidean distance between the histograms to compute the distances for \eqref{eq:W}.
+In Figure \ref{fig:dmaps_embed_noemd}, we have computed the diffusion maps embeddings of our two sets of simulation data, used the standard Euclidean distance between the histograms to compute the distances in \eqref{eq:W}.
 %
-The embeddings obtained are not particularly informative, and do not capture the underlying system dynamics; there is no strong correlation of the embedding coordinates with $p$ or $t$. 
+The embeddings are not particularly informative and do not capture the underlying system dynamics; there is no strong correlation of the embedding coordinates with $p$ or $t$. 
 %
@@
 %
-We would prefer a metric which not only considers the density at each bin, but also takes into account the distance {\em between} bins,
-such that two distributions whose supports are a short distance apart are closer (in norm) than two distributions whose supports are very far from each other.
+We would prefer a metric which not only considers the density at each bin, but also takes into account the distance {\em between} bins.
+%,such that two distributions whose supports are a short distance apart are closer (in norm) than two distributions whose supports are very far from each other.
 %
@@
 %
-Conceptually, EMD measures how much ``work'' it takes to transform one probability density into another.
+Conceptually, EMD measures how much ``work'' it takes to transform one probability distribution into another.
 %
-It therefore not only considers where the densities are inconsistent, but how far apart the inconsistencies are.
+It therefore not only considers where the distributions are inconsistent, but also how far apart the inconsistencies are.
 %
-Although the brute-force computation of the EMD between two histograms is computationally expensive, there has been a plethora of work in developing efficient algorithms for the computation of EMD \cite{...}.
+Although the brute-force computation of the EMD is computationally expensive, there has been a plethora of work in developing efficient algorithms for the computation of EMD \cite{Pele-eccv2008, Pele-iccv2009}.
 %
@@
 %The implementation is available at \url{http://www.cs.huji.ac.il/~ofirpele/FastEMD/code}. 
-For the special case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions \cite{...},
-\begin{equation}
-\| \rho_1  - \rho_2 \|_{EMD} = \int_{-\infty}^{\infty} \left| \int_{-\infty}^x \rho_1(y) dt - \int_{-\infty}^x \rho_2(y) dy \right| dx
-\end{equation}
-where $\rho_1(x)$ and $\rho_2(x)$ are two probability distributions defined on $x \in \mathbb{R}$. 
+For the special case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions \cite{rubner2000perceptual}.
+%\begin{equation}
+%\| \rho_1  - \rho_2 \|_{EMD} = \int_{-\infty}^{\infty} \left| \int_{-\infty}^x \rho_1(y) dt - \int_{-\infty}^x \rho_2(y) dy \right| dx
+%\end{equation}
+%where $\rho_1(x)$ and $\rho_2(x)$ are two probability distributions defined on $x \in \mathbb{R}$. 
 %
@@
 %
-One can see that the two important factors, $p$ and $t$, are well-correlated with the first two diffusion maps coordinates, $\phi_2$ and $\phi_3$. 
+The two system variables, $p$ and $t$, are well-correlated with the first two diffusion maps coordinates, $\phi_2$ and $\phi_3$. 
 %
-However, the ``more important'' coordinate, $\phi_2$, is correlated with $p$ for the small $\lambda$ case (a), and correlated with $t$ for the large $\lambda$ case (d); 
+However, the ``more important'' coordinate, $\phi_2$, is correlated with $p$ for the small $\lambda$ case (Figure~\ref{subfig:small_lambda_p}), and correlated with $t$ for the large $\lambda$ case (Figure~\ref{subfig:large_lambda_t});
 this is consistent with the macroscopic PDE description.
 %
-Furthermore, in the small $\lambda$ regime (wave equation), shown in Figures \ref{subfig:small_lambda_p} and \ref{subfig:small_lambda_t},
-we can see that the points corresponding to small times are more clustered than the points corresponding to large times.
+Furthermore, in the small $\lambda$ regime (wave equation), shown in Figures \ref{subfig:small_lambda_p} and \ref{subfig:small_lambda_t}, the points corresponding to small times are more clustered than the points corresponding to large times.
 %
@@
 %
-On the other hand, for large time, we observe that the initial distribution $p$ is not organized as well in the embedding in Figure \ref{subfig:large_lambda_p}, since the velocities have equilibrated and so the initial distribution is less detectable in the particle distribution.
+On the other hand, for large time, we observe that the initial distribution $p$ is less organized in Figure \ref{subfig:large_lambda_p}, since the velocities have equilibrated and the initial distribution is less detectable in the particle distribution.
 
@@
 %
-To illustrate this in our simulations, we chose an intermediate value of $\lambda$ ($\lambda=100$ and $s=10$), and allowed the system to evolve for 10 and 20 time units. 
+We can also detect this using diffusion maps to analyze our microscopic simulations.
+%
+We chose an intermediate value of $\lambda$ ($\lambda=100$ and $s=10$), and allowed the system to evolve for 10 and 20 time units. 
 %
@@
 %
-However, when the system evolves for 20 time units, the initial velocity distribution has had more time to equilibrate; therefore, the behavior is more diffusive and $\phi_2$ is now correlated with $t$.
+However, when the system evolves for 20 time units, the initial velocity distribution has had more time to equilibrate - the behavior is more diffusive and $\phi_2$ is now correlated with $t$. 
 %
@@
 %
-This is reassuring, and promotes the broad applicability of our methods to a wide variety of dynamical systems.
+This promotes the broad applicability of our methods to a wide variety of dynamical systems.
 

### 2014-04-04 chemotaxis/simulate_jump_process_EMD.m

@@ -3,6 +3,8 @@ close all
 
 rng(321);
 
+markersize = 500;
+
 %% simulations
 
 % rate of switching velocity
@@ -84,7 +86,7 @@ eps = median(W(:));
 [V, D] = dmaps(W, eps, 10);
 
 figure;
-scatter(V(:,2),V(:,3),200,all_time(idx), '.')
+scatter(V(:,2),V(:,3),markersize,all_time(idx), '.')
 xlabel('$\phi_2$', 'interpreter','latex', 'fontsize', 20)
 ylabel('$\phi_3$', 'interpreter','latex', 'fontsize', 20)
 %title('histograms: colored by time')
@@ -94,7 +96,7 @@ set(get(h,'xlabel'),'String', 't', 'fontsize', 20);
 saveas(gcf, sprintf('rawhist_t_%d', lambda), 'epsc')
 
 figure;
-scatter(V(:,2),V(:,3),200,all_p(idx), '.')
+scatter(V(:,2),V(:,3),markersize,all_p(idx), '.')
 xlabel('$\phi_2$', 'interpreter','latex', 'fontsize', 20)
 ylabel('$\phi_3$', 'interpreter','latex', 'fontsize', 20)
 %title('histograms: colored by p')
@@ -140,10 +142,9 @@ if corr(V2(:,3), all_time(idx)') < 0
 end
 
 
-
 %%
 figure;
-scatter(V2(:,2),V2(:,3),200,all_time(idx), '.')
+scatter(V2(:,2),V2(:,3),markersize,all_time(idx), '.')
 xlabel('$\phi_2$', 'interpreter','latex')
 ylabel('$\phi_3$', 'interpreter','latex')
 h = colorbar;
@@ -152,7 +153,7 @@ set(get(h,'xlabel'),'String', 't');
 %print(sprintf('EMD_t_%d', lambda), '-r300','-djpeg')
 
 figure;
-scatter(V2(:,2),V2(:,3),200,all_p(idx), '.')
+scatter(V2(:,2),V2(:,3),markersize,all_p(idx), '.')
 xlabel('$\phi_2$', 'interpreter','latex')
 ylabel('$\phi_3$', 'interpreter','latex')
 h = colorbar;
@@ -163,7 +164,7 @@ set(get(h,'xlabel'),'String', 'p');
 %%
 figure;
[…]

### 2014-04-04 chemotaxis/simulate_jump_process_EMD2.m

@@ -2,6 +2,8 @@ clear all
 close all
 
 rng(123);
+markersize = 500;
+
 
 %% simulations
 
@@ -85,7 +87,7 @@ eps = median(W(:));
 [V, D] = dmaps(W, eps, 10);
 
 figure;
-scatter(V(:,2),V(:,3),200,all_time(idx), '.')
+scatter(V(:,2),V(:,3),markersize,all_time(idx), '.')
 xlabel('$\phi_2$', 'interpreter','latex', 'fontsize', 20)
 ylabel('$\phi_3$', 'interpreter','latex', 'fontsize', 20)
 %title('histograms: colored by time')
@@ -95,7 +97,7 @@ set(get(h,'xlabel'),'String', 't', 'fontsize', 20);
 saveas(gcf, sprintf('rawhist_t_%d', lambda), 'epsc')
 
 figure;
-scatter(V(:,2),V(:,3),200,all_p(idx), '.')
+scatter(V(:,2),V(:,3),markersize,all_p(idx), '.')
 xlabel('$\phi_2$', 'interpreter','latex', 'fontsize', 20)
 ylabel('$\phi_3$', 'interpreter','latex', 'fontsize', 20)
 %title('histograms: colored by p')
@@ -143,7 +145,7 @@ eps2 = median(W2(:));
 
 %%
 figure;
-scatter(V2(:,2),V2(:,3),200,all_time(idx), '.')
+scatter(V2(:,2),V2(:,3),markersize,all_time(idx), '.')
 xlabel('$\phi_2$', 'interpreter','latex')
 ylabel('$\phi_3$', 'interpreter','latex')
 h = colorbar;
@@ -152,7 +154,7 @@ set(get(h,'xlabel'),'String', 't');
 %print(sprintf('EMD_t_%d', lambda), '-r300','-djpeg')
 
 figure;
-scatter(V2(:,2),V2(:,3),200,all_p(idx), '.')
+scatter(V2(:,2),V2(:,3),markersize,all_p(idx), '.')
 xlabel('$\phi_2$', 'interpreter','latex')
 ylabel('$\phi_3$', 'interpreter','latex')
 h = colorbar;
@@ -164,9 +166,9 @@ set(get(h,'xlabel'),'String', 'p');
 
 figure;
 set(gcf,'PaperPositionMode','auto');
[…]

### 2014-04-04 chemotaxis/simulate_jump_process_EMD_varyt.m

@@ -3,6 +3,7 @@ close all
 
 rng(321);
 
+markersize = 500;
 %% simulations
 
 % rate of switching velocity
@@ -86,7 +87,7 @@ idx = (all_time > 1);
 % [V, D] = dmaps(W, eps, 10);
 %
 % figure;
-% scatter(V(:,2),V(:,3),200,all_time(idx), '.')
+% scatter(V(:,2),V(:,3),markersize,all_time(idx), '.')
 % xlabel('$\phi_2$', 'interpreter','latex', 'fontsize', 20)
 % ylabel('$\phi_3$', 'interpreter','latex', 'fontsize', 20)
 % %title('histograms: colored by time')
@@ -94,7 +95,7 @@ idx = (all_time > 1);
 % set(get(h,'xlabel'),'String', 't', 'fontsize', 20);
 %
 % figure;
-% scatter(V(:,2),V(:,3),200,all_p(idx), '.')
+% scatter(V(:,2),V(:,3),markersize,all_p(idx), '.')
 % xlabel('$\phi_2$', 'interpreter','latex', 'fontsize', 20)
 % ylabel('$\phi_3$', 'interpreter','latex', 'fontsize', 20)
 % %title('histograms: colored by p')
@@ -139,7 +140,7 @@ end
 
 %%
 figure;
-scatter(V2(:,2),V2(:,3),200,all_time(idx), '.')
+scatter(V2(:,2),V2(:,3),markersize,all_time(idx), '.')
 xlabel('$\phi_2$', 'interpreter','latex')
 ylabel('$\phi_3$', 'interpreter','latex')
 h = colorbar;
@@ -148,7 +149,7 @@ set(get(h,'xlabel'),'String', 't');
 %print(sprintf('EMD2_t_%d', tmax), '-r300','-djpeg')
 
 figure;
-scatter(V2(:,2),V2(:,3),200,all_p(idx), '.')
+scatter(V2(:,2),V2(:,3),markersize,all_p(idx), '.')
 xlabel('$\phi_2$', 'interpreter','latex')
 ylabel('$\phi_3$', 'interpreter','latex')
 h = colorbar;
@@ -159,9 +160,9 @@ set(get(h,'xlabel'),'String', 'p');
 %%
 figure;
 set(gcf,'PaperPositionMode','auto');
[…]

### 2014-04-04 chemotaxis/simulate_jump_process_EMD_varyt2.m

@@ -3,6 +3,8 @@ close all
 
 rng(321);
 
+markersize = 500;
+
 %% simulations
 
 % rate of switching velocity
@@ -87,7 +89,7 @@ idx = (all_time > 1);
 % [V, D] = dmaps(W, eps, 10);
 %
 % figure;
-% scatter(V(:,2),V(:,3),200,all_time(idx), '.')
+% scatter(V(:,2),V(:,3),markersize,all_time(idx), '.')
 % xlabel('$\phi_2$', 'interpreter','latex', 'fontsize', 20)
 % ylabel('$\phi_3$', 'interpreter','latex', 'fontsize', 20)
 % %title('histograms: colored by time')
@@ -95,7 +97,7 @@ idx = (all_time > 1);
 % set(get(h,'xlabel'),'String', 't', 'fontsize', 20);
 %
 % figure;
-% scatter(V(:,2),V(:,3),200,all_p(idx), '.')
+% scatter(V(:,2),V(:,3),markersize,all_p(idx), '.')
 % xlabel('$\phi_2$', 'interpreter','latex', 'fontsize', 20)
 % ylabel('$\phi_3$', 'interpreter','latex', 'fontsize', 20)
 % %title('histograms: colored by p')
@@ -140,7 +142,7 @@ end
 
 %%
 figure;
-scatter(V2(:,2),V2(:,3),200,all_time(idx), '.')
+scatter(V2(:,2),V2(:,3),markersize,all_time(idx), '.')
 xlabel('$\phi_2$', 'interpreter','latex')
 ylabel('$\phi_3$', 'interpreter','latex')
 h = colorbar;
@@ -149,7 +151,7 @@ set(get(h,'xlabel'),'String', 't');
 %print(sprintf('EMD2_t_%d', tmax), '-r300','-djpeg')
 
 figure;
-scatter(V2(:,2),V2(:,3),200,all_p(idx), '.')
+scatter(V2(:,2),V2(:,3),markersize,all_p(idx), '.')
 xlabel('$\phi_2$', 'interpreter','latex')
 ylabel('$\phi_3$', 'interpreter','latex')
 h = colorbar;
@@ -160,9 +162,9 @@ set(get(h,'xlabel'),'String', 'p');
 %%
 figure;
 set(gcf,'PaperPositionMode','auto');
[…]

### 2014-04-29 chemotaxis/prl_paper.tex
Added ronen edits to chemotaxis paper

@@
 
-Using a stochastic model which arises in cellular chemotaxis, we illustrate how dimensionality reduction techniques (diffusion maps) can be used characterize the behavior of a dynamical system.
+Using a stochastic model which arises in cellular chemotaxis, we illustrate how manifold learning techniques, such as diffusion maps, can be used to characterize the behavior of a dynamical system.
 %
-Depending on the value of a single parameter, the system dynamics at the macroscale range from heat equation dynamics (where particles diffuse uniformly outwards) to wave equation dynamics (where particle propagate with a fixed speed).
+In our model, depending on the value of a single parameter, the system dynamics at the macroscale range from heat equation dynamics, where particles diffuse uniformly outwards, to wave equation dynamics, where particle propagate with a fixed speed.
 %
-We demonstrate that diffusion maps with the appropriate observers and metric can recover the two system variables (the initial condition and the time) from data collected at the microscale. 
+We demonstrate that diffusion maps with the appropriate observers and metric can recover the two macroscopic system variables (initial condition and time) from data collected at the microscale in a variable- and equation-free manner.
 %
-Furthermore, the order in which we recover these variables is indicative of the system dynamics (heat equation versus wave equation) over the timescales of interest.
+Furthermore, the order in which these variables are recovered is indicative of the system dynamics over the timescales of interest.
 %
@@
 %
-For example, fluid dynamics, which can be described via the position and velocity of the fluid particles, is modeled at the macroscale using the Navier-Stokes equations.
+For example, fluid flow, which can be described via the position and velocity of the fluid particles, is modeled at the macroscale using the Navier-Stokes equations.
 %
@@
 %
-We would like to {\em automatically} uncover the macroscale behavior from data collected at the microscale using data-driven techniques.
-
+We would like to {\em automatically} uncover the macroscale behavior from data collected at the microscale without knowing or assuming any microscopic or macroscopic model a priori.
+%
 TODO: write where this has been done in other places, context, etc.
 
-We will demonstrate how we can use data-driven techniques (diffusion maps \cite{coifman2005geometric}) to {\em automatically} detect changes in a system's macroscopic dynamics from microscale data.
-%
+In this letter, we will present how this goal can be achieved through a data-driven method based on geometric analysis and manifold learning.
+Specifically, we aim to emphasize two particular aspects of manifold learning from a dynamical systems' point of view: (a) the significance of the observers, especially in presence of noise and random dynamics, and (b) finding an appropriate metric of comparison.
+We will show that both are essential for our technique to successfully uncover the underlying dynamics of a given system.
+
 We will illustrate our methodology through a model problem arising in studies of cellular chemotaxis \cite{othmer2000diffusion}.
@@
 %
-We will demonstrate that diffusion maps uncovers a description of the microscopic data which is consistent with the macroscopic model.
+We will demonstrate how we can use diffusion maps \cite{coifman2005geometric}, a manifold learning technique, to {\em automatically} detect changes in a system's macroscopic dynamics from microscale data.
 %
-Furthermore, diffusion maps shows changes in dynamical behavior as we vary the system parameter. 
+This case study exhibits several key points which allow us to demonstrate the strength of our approach.
 %
-For our techniques to successfully uncover the underlying dynamics, it will be essential to use the appropriate observers for our system.
+First, this system has random microscopic behavior with smooth macroscopic dynamics governed by few parameters, which determine the regime/mode of the system. 
 %
-Furthermore, we will need to use the appropriate metric between the observers.
+Second, it consists of an ensemble of individual particles. 
+%
+Third, it has an analytic macroscopic description, which serves as a ground truth and will be used to verify our results, which will be obtained in an unsupervised manner.
+%
+We will show that diffusion maps based on statistical observers and the appropriate metric between them uncovers a description of the microscopic data which is consistent with the macroscopic model.
+%
+Furthermore, diffusion maps will uncover changes in dynamical behavior as we vary system parameters.
 
-In this Letter, we will begin by describing the chemotaxis problem.
+The remainder of this letter is organized as follows. We will begin by describing the chemotaxis problem.
 %
@@
 
-We work with a stochastic velocity jump process that arises in models of chemotaxis \cite{othmer2000diffusion}.
-%
 Chemotaxis is the movement of cells (such as bacteria) which is regulated by extracellular sensed signals.
@@
 %
-We will look at one such model described by a velocity jump process.
+We will examine one such model described by a velocity jump process \cite{othmer2000diffusion}.
 %
@@
 %Note that each particle has its own ``clock" (Poisson process).
-We simulate the dynamics of this stochastic system to obtain our microscopic data from which we will do our analysis.
-%\subsection{Partial differential equations}
-%
-We will consider two sets of simulations;
-in one set (small $\lambda$), $\lambda = 1$ and $s=1$, and in the other set (large $\lambda$), $\lambda = 400$, and $s=20$.
-%
-For each set, we run many simulations, each with $N=1000$ particles.
-%
-We vary the initial condition of each simulation such that $0.1 \le p  \le 0.9$, and allow each simulation to evolve for $10$ time units.
-%
-We discard the initial portion of each simulation trajectory corresponding to $t < 1$, as this is very noisy and corresponds to initial relaxation of the system.
 
@@
 %
-For a large collection of particles ($N \rightarrow \infty$), the system can be described by the probability distribution of the particles.
+For a large collection of particles ($N \rightarrow \infty$), the system can be described by the probability density of the particles.
 %
-Let $\rho(x, t)$ denote the probability distribution of the particles, and let $\rho^-(x, t)$ and $\rho^+(x, t)$ denote the distributions of left and right moving particles, respectively.
+Let $\rho(x, t)$ denote the probability density of the particles, and let $\rho^-(x, t)$ and $\rho^+(x, t)$ denote the densities of the particles moving towards the positive and negative axis directions, respectively.
 %
-It can be shown that, as $N \rightarrow \infty$, $\rho(x, t)$ obeys the following set of partial differential equations (PDEs) \cite{othmer2000diffusion}:
+It can be shown that, as $N \rightarrow \infty$, the densities obey the following set of partial differential equations (PDEs) \cite{othmer2000diffusion}:
 \begin{equation} \label{eqn:coupled_pdes}
@@
 %
-Alternatively, \eqref{eqn:coupled_pdes} can be written as one, second--order PDE:
+Alternatively, \eqref{eqn:coupled_pdes} can be rewritten as one, second--order PDE:
 \begin{equation} \label{eq:second_order_pde}
@@
 %
-From this analysis, we can see that the initial distribution of left- and right-moving particles (given by $p$ in our microscopic simulations) plays a very different role depending on the value of $\lambda$.
+This analysis implies that the initial distribution of the moving direction of the particles (given by $p$ in our microscopic simulations) plays a very different role depending on the value of $\lambda$.
 %
@@
 %
-When $\lambda \rightarrow \infty$, the dynamics are described by one heat equation, and the initial conditions are not important - the velocity distribution quickly equilibrates and we see purely diffusive behavior.
+When $\lambda \rightarrow \infty$, the dynamics are described by one heat equation, and the initial conditions are insignificant -- the velocity distribution quickly equilibrates and we see purely diffusive behavior.
 %
@@
 
-We will use diffusion maps \cite{coifman2005geometric} to analyze the data from our microscopic, stochastic simulations.
+We use diffusion maps \cite{coifman2005geometric} to analyze the data from simulations of the described microscopic stochastic model.
 %
-We will show that diffusion maps can organize the data obtained from our microscopic simulations, and that this organization will be consistent with the two system variables, $p$ and $t$, that we expect from our PDE analysis.
+Our goal is to show that diffusion maps can organize the data obtained from the microscopic simulations, and that this organization is consistent with the two system variables, $p$ and $t$, that emerge from the PDE analysis.
 %
-Diffusion maps, in general, is a nonlinear dimensionality reduction technique.
+Diffusion maps, in general, is a nonlinear manifold learning technique that embeds (maps) the data into a lower dimensional space.
+Ideally, the dimension of the embedded samples is equal to the the intrinsic (true) dimension of the data, and the variables/coordinates of the embedded samples represent the governing variables of the system.
 %
-In our simulations, although the data is very high-dimensional (e.g., the positions of all $N$ particles), for a fixed value of $\lambda$, there are only two important variables ($p$ and $t$) that determine the current (macroscopic) state of the system; thus, our microscopic data {\em should} be two-dimensional and described by $p$ and $t$.
+In our case, although the data is very high-dimensional (e.g., the positions of all $N$ particles), for a fixed value of $\lambda$, there are only two important variables ($p$ and $t$) that determine the current (macroscopic) mode of the system; thus, our microscopic data {\em should} be two-dimensional and described by $p$ and $t$.
 %
-For diffusion maps, we assume we have data points $z_1, z_2, \dots, z_m \in \mathbb{R}^n$ which lie on a $d$-dimensional, nonlinear manifold, with $d \ll n$. 
-%
-Our goal is to uncover a $d$-dimensional parameterization of the points $z_1, \dots, z_m$ that respects the underlying manifold geometry.
+Let $z(t) \in \mathbb{R}^n$ denote an $n$-dimensional observation of the system at time $t$.
+Assume we are given $m$ observations $z(1), \dots, z(m)$, which are assumed to lie on a $d$-dimensional, nonlinear manifold, with $d \ll n$.
+Our goal is to uncover a $d$-dimensional parameterization of the $m$ observations that respects the underlying manifold geometry.
 %
@@
 \begin{equation} \label{eq:W}
-W_{ij} = \exp \left( -\frac{\|z_i - z_j \|^2}{\epsilon^2} \right)
+W_{ij} = \exp \left( -\frac{\|z(i) - z(j) \|^2}{\epsilon^2} \right), \ i,j=1,\ldots,m
 \end{equation}
-where $\| \cdot \|$ denotes the appropriate norm for the data (we will discuss the choice of norm in the following section), and $\epsilon$ is a characteristic distance between the data points (we take $\epsilon$ to be the median of the pairwise distances between the data points).
+where $\| \cdot \|$ denotes the appropriate norm for the data (which we will discuss in the following section), and $\epsilon$ is a characteristic distance between the observations (we take $\epsilon$ to be the median of the pairwise distances between the data points).
 %
@@
 %
-We compute the eigenvalues $\lambda_1, \lambda_2, \dots, \lambda_m$ and eigenvectors $\phi_1, \phi_2, \dots, \phi_m$ of $A$, and order them such that $|\lambda_1| \ge |\lambda_2| \ge \dots \ge |\lambda_m|$.
+We compute the eigenvalues $\mu_0, \mu_1, \dots, \mu_{m-1}$ and eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$, and order them such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$. 
 %
-Note that the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$), which implies $\lambda_1 = 1$ and $\phi_1$ is a constant vector.
+Note that the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$), which implies $\mu_0 = 1$ and $\phi_0$ is a constant vector.
 %
 The next few eigenvectors give us the embedding coordinates for our data;
-$\phi_{j+1}(i)$ gives us the $j^{th}$ embedding coordinate for $z_i$ and
-$\lambda_j$ gives a measure of the ``importance'' of coordinate $\phi_j$.
+$\phi_{j}(i)$ gives us the $j^{th}$ embedding coordinate for $z(i)$ and
+$\mu_j$ gives a measure of the ``importance'' of coordinate $\phi_j$.
 
@@
 %
-Instead, we use histograms of the particle positions as our observables, so that $z_i \in \mathbb{R}^n$ will be the histogram of $x_1, x_2, \dots, x_N$ into $n$ bins (for our simulations, we take $n=32$).
+Instead, we use histograms of the particle positions as our observables, so that $z(i) \in \mathbb{R}^n$ will be the histogram of $x_1(i), x_2(i), \dots, x_N(i)$ into $n$ bins (for our simulations, we take $n=32$).
 %
@@
 %
-The histograms serve as an empirical estimate of the probability density at each time point in our simulations.
+The histograms serve as an empirical estimate of the probability density at each time point.
 
@@
 The distances used in the diffusion maps kernel are the Euclidean distances between the histograms of particle positions.
-The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time.
-The first two diffusion maps modes do not accurately capture the important factors in our simulations.}
+The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time.}
 \label{fig:dmaps_embed_noemd}
 \end{figure}
+%TODO: for all figures - fix the labels of the axes - phi_2 vs. phi_1.
 
@@
 %
-In Figure \ref{fig:dmaps_embed_noemd}, we have computed the diffusion maps embeddings of our two sets of simulation data, used the standard Euclidean distance between the histograms to compute the distances in \eqref{eq:W}.
-%
-The embeddings are not particularly informative and do not capture the underlying system dynamics; there is no strong correlation of the embedding coordinates with $p$ or $t$. 
-%
-This is due to our poor choice of metric, which does not accurately describe the distances between histograms.
-%
-We would prefer a metric which not only considers the density at each bin, but also takes into account the distance {\em between} bins.
+However, the Euclidean distance only considers the density at each bin and does not take into account the distance {\em between} bins.
 %,such that two distributions whose supports are a short distance apart are closer (in norm) than two distributions whose supports are very far from each other.
@@
 
-We choose to use the earth mover's distance (EMD) \cite{rubner2000earth} as our metric between histograms.
+To overcome this critical shortcoming, we propose to use the earth mover's distance (EMD) \cite{rubner2000earth} as our metric between histograms.
 %
-Conceptually, EMD measures how much ``work'' it takes to transform one probability distribution into another.
+Conceptually, EMD measures how much ``work'' it takes to transform one probability density into another.
 %
-It therefore not only considers where the distributions are inconsistent, but also how far apart the inconsistencies are.
+It therefore not only considers where the densities are inconsistent, but also how far apart the inconsistencies are.
 %
@@
 \begin{equation}
-\| z_1 - z_2 \|_{EMD} = \sum_{i=1}^{n} \left| \sum_{j=1}^i z_1(j) - \sum_{j=1}^i z_2(j) \right|
+\| z(i) - z(j) \|_{EMD} = \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|
 \end{equation}
-where $z_1, z_2 \in \mathbb{R}^n$ are two histograms defined on (equally-spaced) bins in $\mathbb{R}$. 
+where $z(i), z(j) \in \mathbb{R}^n$ are two histograms defined on (equally-spaced) bins in $\mathbb{R}$. 
 %\section{Results}
@@
 The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time.
-}
+Representative histograms are shown for selected data points.}
 \label{fig:dmaps_embed_emd}
@@
 
+We simulate the dynamics of this stochastic system to obtain our microscopic data to which we apply diffusion maps based on histograms and the EMD.
+%
+We consider two sets of simulations;
+in one set (small $\lambda$), $\lambda = 1$ and $s=1$, and in the other set (large $\lambda$), $\lambda = 400$, and $s=20$.
+%
+For each set, we run many simulations, each with $N=1000$ particles.
+%
+We vary the initial condition of each simulation such that $0.1 \le p  \le 0.9$, and allow each simulation to evolve for $10$ time units.
+%
+We discard the initial portion of each simulation trajectory corresponding to $t < 1$, as this is very noisy and corresponds to initial relaxation of the system.
+
+In Figure \ref{fig:dmaps_embed_noemd}, we have computed the diffusion maps embeddings of our two sets of simulation data, used the standard Euclidean distance between the histograms to compute the distances in \eqref{eq:W}.
+%
+The embeddings are not particularly informative and do not capture the underlying system dynamics; there is no strong correlation of the embedding coordinates with $p$ or $t$. 
+%
+This is due to our poor choice of metric, which does not accurately describe the distances between histograms.
+
 In Figure \ref{fig:dmaps_embed_emd}, we have used the earth mover's distance to compute the distances between histograms.
 %
-The two system variables, $p$ and $t$, are well-correlated with the first two diffusion maps coordinates, $\phi_2$ and $\phi_3$. 
+The two system variables, $p$ and $t$, are well-correlated with the first two diffusion maps coordinates, $\phi_1$ and $\phi_2$. 
 %
-However, the ``more important'' coordinate, $\phi_2$, is correlated with $p$ for the small $\lambda$ case (Figure~\ref{subfig:small_lambda_p}), and correlated with $t$ for the large $\lambda$ case (Figure~\ref{subfig:large_lambda_t});
+However, the ``more important'' coordinate, $\phi_1$, is correlated with $p$ for the small $\lambda$ case (Figure~\ref{subfig:small_lambda_p}), and correlated with $t$ for the large $\lambda$ case (Figure~\ref{subfig:large_lambda_t});
 this is consistent with the macroscopic PDE description.
@@
 %
-On the other hand, for large time, we observe that the initial distribution $p$ is less organized in Figure \ref{subfig:large_lambda_p}, since the velocities have equilibrated and the initial distribution is less detectable in the particle distribution.
+On the other hand, for large time, we observe that the initial distribution $p$ is less organized in Figure \ref{subfig:large_lambda_p}, since the velocities have equilibrated and the initial distribution is less detectable in the particle density.
 
@@
 The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time.
-The particles are allowed to evolve for (a, b) 10 time units, and (c, d) 20 time units.}
+The particles are allowed to evolve for (a, b) 10 time units, and (c, d) 20 time units.
+Representative histograms are shown for selected data points.}
 \label{fig:dmaps_embed_varyt}
@@
 %
-When the system only evolves for 10 time units, the majority of the data still appears ``wave-like'', and so the dominant parameter, $\phi_2$, is correlated with $p$, the initial condition.
+When the system only evolves for 10 time units, the majority of the data still appears ``wave-like'', and so the dominant parameter, $\phi_1$, is correlated with $p$, the initial condition.
 %
-However, when the system evolves for 20 time units, the initial velocity distribution has had more time to equilibrate - the behavior is more diffusive and $\phi_2$ is now correlated with $t$. 
+However, when the system evolves for 20 time units, the initial velocity distribution has had more time to equilibrate - the behavior is more diffusive and $\phi_1$ is now correlated with $t$. 
 %
@@
 
-\bibliography{../../references/references}
+%\bibliography{../../references/references}
 

### 2014-04-29 chemotaxis/simulate_jump_process_EMD2.m

@@ -88,8 +88,8 @@ eps = median(W(:));
 
 figure;
 scatter(V(:,2),V(:,3),markersize,all_time(idx), '.')
-xlabel('$\phi_2$', 'interpreter','latex', 'fontsize', 20)
-ylabel('$\phi_3$', 'interpreter','latex', 'fontsize', 20)
+xlabel('$\phi_1$', 'interpreter','latex', 'fontsize', 20)
+ylabel('$\phi_2$', 'interpreter','latex', 'fontsize', 20)
 %title('histograms: colored by time')
 h = colorbar;
 set(get(h,'xlabel'),'String', 't', 'fontsize', 20);
@@ -98,8 +98,8 @@ saveas(gcf, sprintf('rawhist_t_%d', lambda), 'epsc')
 
 figure;
 scatter(V(:,2),V(:,3),markersize,all_p(idx), '.')
-xlabel('$\phi_2$', 'interpreter','latex', 'fontsize', 20)
-ylabel('$\phi_3$', 'interpreter','latex', 'fontsize', 20)
+xlabel('$\phi_1$', 'interpreter','latex', 'fontsize', 20)
+ylabel('$\phi_2$', 'interpreter','latex', 'fontsize', 20)
 %title('histograms: colored by p')
 h = colorbar;
 set(get(h,'xlabel'),'String', 'p', 'fontsize', 20);
@@ -146,8 +146,8 @@ eps2 = median(W2(:));
 %%
 figure;
 scatter(V2(:,2),V2(:,3),markersize,all_time(idx), '.')
-xlabel('$\phi_2$', 'interpreter','latex')
-ylabel('$\phi_3$', 'interpreter','latex')
+xlabel('$\phi_1$', 'interpreter','latex')
+ylabel('$\phi_2$', 'interpreter','latex')
 h = colorbar;
 set(get(h,'xlabel'),'String', 't');
 %title(sprintf('EMD: colored by time, \\lambda= %d', lambda))
@@ -155,8 +155,8 @@ set(get(h,'xlabel'),'String', 't');
 
 figure;
 scatter(V2(:,2),V2(:,3),markersize,all_p(idx), '.')
-xlabel('$\phi_2$', 'interpreter','latex')
-ylabel('$\phi_3$', 'interpreter','latex')
+xlabel('$\phi_1$', 'interpreter','latex')
+ylabel('$\phi_2$', 'interpreter','latex')
 h = colorbar;
[…]

### 2014-05-09 chemotaxis/prl_paper.tex
Added some edits to chemotaxis paper

@@
 %
-Furthermore, the order in which these variables are recovered is indicative of the system dynamics over the timescales of interest.
+Furthermore, the order in which these variables are recovered is indicative of the system dynamics over the timescales of interest, 
+and the eigenvalue spectrum indicates where the switch in system dynamics occurs.
 %
@@
 %
-We would like to {\em automatically} uncover the macroscale behavior from data collected at the microscale without knowing or assuming any microscopic or macroscopic model a priori.
+We would like to {\em automatically} uncover the macroscale behavior from data collected at the microscale without knowing or assuming any microscopic or macroscopic model {\em a priori}.
 %
@@
 %
-Furthermore, we will show that diffusion maps can inform us as to the relative importance of system variables in different parameter regimes.
+Furthermore, we will show that diffusion maps can inform us as to the relative importance of system variables in different parameter regimes, and allow us to detect changes in macroscopic dynamical behavior.
 
@@
 %
-This analysis implies that the initial distribution of the moving direction of the particles (given by $p$ in our microscopic simulations) plays a very different role depending on the value of $\lambda$.
+This analysis shows that the initial distribution of the moving direction of the particles (given by $p$ in our microscopic simulations) plays a very different role depending on the value of $\lambda$.
 %
@@
 However, we would like to demonstrate that, in general, these PDEs are not necessary for us to uncover which system variables are important to describe the macroscopic dynamics.
+
+\begin{figure*}[t!]
+\def\figwidth{6.4cm}
+\begin{subfigure}{\figwidth}
+\includegraphics[width=\textwidth]{EMD_withhist_p_1}
+\caption{}
+\label{subfig:small_lambda_p}
+\end{subfigure}
+\begin{subfigure}{\figwidth}
+\includegraphics[width=\textwidth]{EMD_withhist_t_1}
+\caption{}
+\label{subfig:small_lambda_t}
+\end{subfigure}
+\begin{subfigure}{\figwidth}
+\includegraphics[width=\textwidth]{EMD_withhist_p_400}
+\caption{}
+\label{subfig:large_lambda_p}
+\end{subfigure}
+\begin{subfigure}{\figwidth}
+\includegraphics[width=\textwidth]{EMD_withhist_t_400}
+\caption{}
+\label{subfig:large_lambda_t}
+\end{subfigure}
+\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with (a,b) $\lambda=1$, $s=1$, and (c,d) $\lambda=400$, $s=20$.
+The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions.
+The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time.
+Representative histograms are shown for selected data points.}
+\label{fig:dmaps_embed_emd}
+\end{figure*}
 
@@
 %
-The next few eigenvectors give us the embedding coordinates for our data;
-$\phi_{j}(i)$ gives us the $j^{th}$ embedding coordinate for $z(i)$ and
-$\mu_j$ gives a measure of the ``importance'' of coordinate $\phi_j$.
+The next few eigenvectors give us the embedding coordinates for our data, such that
+$\phi_{j}(i)$ gives us the $j^{th}$ embedding coordinate for $z(i)$.
+%
+$\mu_j$ gives a measure of the ``importance'' of coordinate $\phi_j$,
+and we can quantify how important $\phi_1$ is relative to $\phi_2$ by comparing $\mu_1$ and $\mu_2$.
 
-We first need to address what are the appropriate observers or measurements for our system.
+In order to use DMAPS to analyze our simulation results, we first need to address what are the appropriate observers or measurements for our system.
 %
@@
 
-\begin{figure}[t]
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{rawhist_p_1}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{rawhist_t_1}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{rawhist_p_400}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.2\textwidth}
-\includegraphics[width=\textwidth]{rawhist_t_400}
-\caption{}
-\end{subfigure}
-\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with (a,b) $\lambda=1$, $s=1$, and (c,d) $\lambda=400$, $s=20$.
-The distances used in the diffusion maps kernel are the Euclidean distances between the histograms of particle positions.
-The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time.}
-\label{fig:dmaps_embed_noemd}
-\end{figure}
-%TODO: for all figures - fix the labels of the axes - phi_2 vs. phi_1.
-
 The second point we need to address is choosing appropriate distance metric between data points in \eqref{eq:W}.
@@
 %For example, the distance between two uniform distributions with finite, non-overlapping support will always be the same, {\em independent} of the distance between the support intervals.
-
+%
 To overcome this critical shortcoming, we propose to use the earth mover's distance (EMD) \cite{rubner2000earth} as our metric between histograms.
@@
 
-\begin{figure}[t]
-\begin{subfigure}{0.45\columnwidth}
-\includegraphics[width=\textwidth]{EMD_withhist_p_1}
-\caption{}
-\label{subfig:small_lambda_p}
-\end{subfigure}
-\begin{subfigure}{0.45\columnwidth}
-\includegraphics[width=\textwidth]{EMD_withhist_t_1}
-\caption{}
-\label{subfig:small_lambda_t}
-\end{subfigure}
-\begin{subfigure}{0.45\columnwidth}
-\includegraphics[width=\textwidth]{EMD_withhist_p_400}
-\caption{}
-\label{subfig:large_lambda_p}
-\end{subfigure}
-\begin{subfigure}{0.45\columnwidth}
-\includegraphics[width=\textwidth]{EMD_withhist_t_400}
-\caption{}
-\label{subfig:large_lambda_t}
-\end{subfigure}
-\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with (a,b) $\lambda=1$, $s=1$, and (c,d) $\lambda=400$, $s=20$.
-The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions.
-The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time.
-Representative histograms are shown for selected data points.}
-\label{fig:dmaps_embed_emd}
-\end{figure}
-
-We simulate the dynamics of this stochastic system to obtain our microscopic data to which we apply diffusion maps based on histograms and the EMD.
+We simulate the dynamics of this stochastic system to obtain microscopic data to which we apply diffusion maps based on histograms and the EMD.
 %
-We consider two sets of simulations;
+We consider two sets of simulations:
 in one set (small $\lambda$), $\lambda = 1$ and $s=1$, and in the other set (large $\lambda$), $\lambda = 400$, and $s=20$.
@@
 %
-We vary the initial condition of each simulation such that $0.1 \le p  \le 0.9$, and allow each simulation to evolve for $10$ time units.
+We vary the initial conditions such that $0.1 \le p  \le 0.9$, and allow each simulation to evolve for $10$ time units.
 %
@@
 
-In Figure \ref{fig:dmaps_embed_noemd}, we have computed the diffusion maps embeddings of our two sets of simulation data, used the standard Euclidean distance between the histograms to compute the distances in \eqref{eq:W}.
-%
-The embeddings are not particularly informative and do not capture the underlying system dynamics; there is no strong correlation of the embedding coordinates with $p$ or $t$. 
-%
-This is due to our poor choice of metric, which does not accurately describe the distances between histograms.
+\begin{figure}[t] 
+\includegraphics[width=8cm]{detect_change_eigenvalues}
+\caption{The first two eigenvalues from DMAPS as a function of $\lambda$. Note that the convergence of the eigenvalues indicates switching of the dynamical ``mode'' of the system.}
+\label{fig:detect_change}
+\end{figure}
 
-In Figure \ref{fig:dmaps_embed_emd}, we have used the earth mover's distance to compute the distances between histograms.
+Figure \ref{fig:dmaps_embed_emd} shows the diffusion maps embedding of our data.
 %
@@
 %
-Furthermore, in the small $\lambda$ regime (wave equation), shown in Figures \ref{subfig:small_lambda_p} and \ref{subfig:small_lambda_t}, the points corresponding to small times are more clustered than the points corresponding to large times.
+Furthermore, in the small $\lambda$ regime (wave equation), shown in Figures \ref{subfig:small_lambda_p} and \ref{subfig:small_lambda_t}, the points corresponding to small times are more tightly clustered than the points corresponding to large times.
 %
-This is not surprising; at small times, the particles are more condensed around $x=0$, and it is more difficult to distinguish the particles moving to the left from the particles moving to the right. 
+This is not surprising: at small times, the particles are more condensed around $x=0$, and it is more difficult to distinguish the particles moving to the left from the particles moving to the right. 
 %
-On the other hand, for large time, once the particles evolve from the origin, this separation is clear.  
+On the other hand, at large times, once the particles evolve from the origin, this separation is clear.  
 %
-For the large $\lambda$ case (heat equation), shown in Figures \ref{subfig:large_lambda_p} and \ref{subfig:large_lambda_t}, we observe that for small time the initial distribution $p$ is well organized in the embedding in Figure \ref{subfig:large_lambda_p}, since for small time the distribution of the particles is skewed and the initial velocity plays a role in this regime.
+For the large $\lambda$ case (heat equation), shown in Figures \ref{subfig:large_lambda_p} and \ref{subfig:large_lambda_t}, we observe that for small time the initial distribution $p$ is well organized in the embedding in Figure \ref{subfig:large_lambda_p}, the distribution of the particles is initially skewed and the initial velocity plays a role.
 %
-On the other hand, for large time, we observe that the initial distribution $p$ is less organized in Figure \ref{subfig:large_lambda_p}, since the velocities have equilibrated and the initial distribution is less detectable in the particle density.
+On the other hand, for large times, we observe that the initial distribution $p$ is less organized in Figure \ref{subfig:large_lambda_p}, as the velocities have equilibrated and the initial distribution is less detectable in the particle density.
 
-\begin{figure}[t]
-\begin{subfigure}{0.45\columnwidth}
-\includegraphics[width=\textwidth]{EMD2_withhist_p_10}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.45\columnwidth}
-\includegraphics[width=\textwidth]{EMD2_withhist_t_10}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.45\columnwidth}
-\includegraphics[width=\textwidth]{EMD2_withhist_p_20}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.45\columnwidth}
-\includegraphics[width=\textwidth]{EMD2_withhist_t_20}
-\caption{}
-\end{subfigure}
-\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with $\lambda=100$, $s=10$.
+%\begin{figure}[t]
+%\begin{subfigure}{0.45\columnwidth}
+%\includegraphics[width=\textwidth]{EMD2_withhist_p_10}
+%\caption{}
+%\end{subfigure}
+%\begin{subfigure}{0.45\columnwidth}
+%\includegraphics[width=\textwidth]{EMD2_withhist_t_10}
+%\caption{}
+%\end{subfigure}
+%\begin{subfigure}{0.45\columnwidth}
+%\includegraphics[width=\textwidth]{EMD2_withhist_p_20}
+%\caption{}
+%\end{subfigure}
+%\begin{subfigure}{0.45\columnwidth}
+%\includegraphics[width=\textwidth]{EMD2_withhist_t_20}
+%\caption{}
+%\end{subfigure}
+%\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with $\lambda=100$, $s=10$.
 The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions.
@@
 Representative histograms are shown for selected data points.}
-\label{fig:dmaps_embed_varyt}
+%\label{fig:dmaps_embed_varyt}
+%\end{figure}
+
+The crossover from ``wave-like'' to ``heat-like'' behavior can be seen in the DMAPS eigenvalues: a large gap between the first two eigenvalues indicates that one system variable is significantly more important and that we are in one of the asymptotic regimes.
+%
+Figure~\ref{fig:detect_change} shows the top two DMAPS eigenvalues, $\mu_1$ and $\mu_2$, as a function of $\lambda$. 
+%
+Each data point is averaged over 10 data sets, with error bars indicating one standard deviation. 
+%
+For small $\lambda$ (the wave-equation regime), the two eigenvalues are well-separated, as the first mode (which parameterizes $p$) is significantly more important than the second mode (which parameterizes $t$).
+%
+At intermediate $\lambda$ values, the eigenvalues come together as the system dynamics change and both $p$ and $t$ are of similar importance.
+%
+The eigenvalues then separate again at large $\lambda$ (the heat equation regime), where the $t$ effects dominate the dynamics.
+
+%We have illustrated that diffusion maps can elucidate the relative importance of different system variables as a function of $\lambda$.
+%%
+%For the plots shown in Figure~\ref{fig:dmaps_embed_emd}, we can easily see the crossover from the wave-equation regime, where $p$ is more important, to the heat-equation regime, where $t$ is more important.
+%%
+%However, we would like to note that this crossover point depends not only on $\lambda$, but also on the timescale of observation.
+%%
+%We can see this from our PDE description in \eqref{eq:second_order_pde};
+%clearly, $\lambda$ is defined relative to $t$, and so we should not only consider $\lambda$ in our analysis, but also the timescale of observation.
+%%
+%We can also detect this using diffusion maps to analyze our microscopic simulations.
+%%
+%We chose an intermediate value of $\lambda$ ($\lambda=100$ and $s=10$), and allowed the system to evolve for 10 and 20 time units. 
+%%
+%The results are shown in Figure \ref{fig:dmaps_embed_varyt}.
+%%
+%When the system only evolves for 10 time units, the majority of the data still appears ``wave-like'', and so the dominant parameter, $\phi_1$, is correlated with $p$, the initial condition.
+%%
+%However, when the system evolves for 20 time units, the initial velocity distribution has had more time to equilibrate - the behavior is more diffusive and $\phi_1$ is now correlated with $t$. 
+%%
+%Therefore, we can also see the crossover from wave-like behavior to heat-like behavior as a function of the observation timescale using diffusion maps.
+
+\begin{figure}[t]
+\def\figwidth{4.2cm}
+\begin{subfigure}{\figwidth}
+\includegraphics[width=\textwidth]{rawhist_p_1}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{\figwidth}
+\includegraphics[width=\textwidth]{rawhist_t_1}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{\figwidth}
+\includegraphics[width=\textwidth]{rawhist_p_400}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{\figwidth}
+\includegraphics[width=\textwidth]{rawhist_t_400}
+\caption{}
+\end{subfigure}
+\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with (a,b) $\lambda=1$, $s=1$, and (c,d) $\lambda=400$, $s=20$.
+The distances used in the diffusion maps kernel are the Euclidean distances between the histograms of particle positions.
+The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time.}
+\label{fig:dmaps_embed_noemd}
 \end{figure}
+%TODO: for all figures - fix the labels of the axes - phi_2 vs. phi_1.
 
-We have illustrated that diffusion maps can elucidate the relative importance of different system variables as a function of $\lambda$.
+To emphasize the importance of using the correct distance metric, we also computed the diffusion maps embeddings of our two sets of simulation data, used the standard Euclidean distance between the histograms to compute the distances in \eqref{eq:W}; the results are shown in Figure \ref{fig:dmaps_embed_noemd}.
 %
-For the plots shown in Figure~\ref{fig:dmaps_embed_emd}, we can easily see the crossover from the wave-equation regime, where $p$ is more important, to the heat-equation regime, where $t$ is more important.
+The embeddings are not particularly informative and do not capture the underlying system dynamics; there is no strong correlation of the embedding coordinates with $p$ or $t$. 
 %
-However, we would like to note that this crossover point depends not only on $\lambda$, but also on the timescale of observation.
+This is due to our poor choice of metric, which does not accurately describe the distances between histograms.
+
+We have demonstrated how diffusion maps can automatically uncover the variables which govern the macroscopic behavior of a system from data sampled at the microscopic level.
 %
-We can see this from our PDE description in \eqref{eq:second_order_pde};
-clearly, $\lambda$ is defined relative to $t$, and so we should not only consider $\lambda$ in our analysis, but also the timescale of observation.
-%
-We can also detect this using diffusion maps to analyze our microscopic simulations.
-%
-We chose an intermediate value of $\lambda$ ($\lambda=100$ and $s=10$), and allowed the system to evolve for 10 and 20 time units. 
-%
-The results are shown in Figure \ref{fig:dmaps_embed_varyt}.
-%
-When the system only evolves for 10 time units, the majority of the data still appears ``wave-like'', and so the dominant parameter, $\phi_1$, is correlated with $p$, the initial condition.
-%
-However, when the system evolves for 20 time units, the initial velocity distribution has had more time to equilibrate - the behavior is more diffusive and $\phi_1$ is now correlated with $t$. 
-%
-Therefore, we can also see the crossover from wave-like behavior to heat-like behavior as a function of the observation timescale using diffusion maps.
-
-We have presented algorithms to automatically uncover the variables which govern the macroscopic behavior of a system from data sampled at the microscopic level.
+Furthermore, the eigenvalue spectrum allows us to detect changes in the macroscopic system dynamics.
 %
@@
 With these two considerations, we used diffusion maps to uncover the two variables which characterize the system dynamics.
+%
+The eigenvalue spectrum can help us detect changes in the dynamical behavior.
 %

### 2014-05-09 chemotaxis/simulate_jump_process_EMD.m

@@ -162,40 +162,53 @@ set(get(h,'xlabel'),'String', 'p');
 %print(sprintf('EMD_p_%d', lambda), '-r300','-djpeg')
 
 %%
+
+num_hist = 6;
+
 figure;
 set(gcf,'PaperPositionMode','auto');
 scatter(V2(:,2),V2(:,3),markersize,all_time(idx), '.');
 ax = gca;
-axis_lim = [-0.4 0.4 -0.25 0.3];
+axis_lim = [-0.45 0.45 -0.25 0.35];
 axis(axis_lim)
 xlabel('$\phi_1$', 'interpreter','latex', 'fontsize', 20)
 ylabel('$\phi_2$', 'interpreter','latex', 'fontsize', 20)
 h = colorbar;
 set(get(h,'xlabel'),'String', 't', 'fontsize', 20);
 curr_ax = get(ax, 'position');
+curr_ax(1) = curr_ax(1) + 0.02;
+set(ax, 'position', curr_ax);
 hist_all2 = hist_all(:, idx);
 hold on
-i = zeros(4, 1);
+i = zeros(num_hist, 1);
 i(1) = find(all_time(idx) == min(all_time(idx)) & all_p(idx) == min(all_p(idx)));
 i(2) = find(all_time(idx) == min(all_time(idx)) & all_p(idx) == max(all_p(idx)));
 i(3) = find(all_time(idx) == max(all_time(idx)) & all_p(idx) == min(all_p(idx)));
 i(4) = find(all_time(idx) == max(all_time(idx)) & all_p(idx) == max(all_p(idx)));
+i(5) = find(all_time(idx) == round(mean(all_time(idx)))+1 & all_p(idx) == min(all_p(idx)));
+i(6) = find(all_time(idx) == round(mean(all_time(idx)))+1 & all_p(idx) == max(all_p(idx)));
 
 points_x = curr_ax(1) + (V2(i, 2) - axis_lim(1)) / (axis_lim(2)-axis_lim(1)) * curr_ax(3);
 points_y = curr_ax(2) + (V2(i, 3) - axis_lim(3)) / (axis_lim(4)-axis_lim(3)) * curr_ax(4);
 
 ax_size = 0.15;
-ax_pos = [.15 .15 ax_size ax_size;
-    .6 .15 ax_size ax_size;
-    .15 .75 ax_size ax_size;
-    .6 .75 ax_size ax_size];
[…]

### 2014-05-09 chemotaxis/simulate_jump_process_EMD2.m

@@ -164,40 +164,51 @@ set(get(h,'xlabel'),'String', 'p');
 
 %%
 
+num_hist = 6;
+
 figure;
 set(gcf,'PaperPositionMode','auto');
 scatter(V2(:,2),V2(:,3),markersize,all_time(idx), '.');
 ax = gca;
-axis_lim = [-0.4 0.25 -0.35 0.35];
+axis_lim = [-0.4 0.25 -0.35 0.4];
 axis(axis_lim)
 xlabel('$\phi_1$', 'interpreter','latex', 'fontsize', 20)
 ylabel('$\phi_2$', 'interpreter','latex', 'fontsize', 20)
 h = colorbar;
 set(get(h,'xlabel'),'String', 't', 'fontsize', 20);
 curr_ax = get(ax, 'position');
+curr_ax(1) = curr_ax(1) + 0.02;
+set(ax, 'position', curr_ax);
 hist_all2 = hist_all(:, idx);
 hold on
-i = zeros(4, 1);
+i = zeros(num_hist, 1);
 i(1) = find(all_time(idx) == min(all_time(idx)) & all_p(idx) == min(all_p(idx)));
 i(2) = find(all_time(idx) == max(all_time(idx)) & all_p(idx) == min(all_p(idx)));
 i(3) = find(all_time(idx) == min(all_time(idx)) & all_p(idx) == max(all_p(idx)));
 i(4) = find(all_time(idx) == max(all_time(idx)) & all_p(idx) == max(all_p(idx)));
+i(5) = find(all_time(idx) == floor(mean(all_time(idx))-2) & all_p(idx) == min(all_p(idx)));
+i(6) = find(all_time(idx) == floor(mean(all_time(idx))-2) & all_p(idx) == max(all_p(idx)));
 
 points_x = curr_ax(1) + (V2(i, 2) - axis_lim(1)) / (axis_lim(2)-axis_lim(1)) * curr_ax(3);
 points_y = curr_ax(2) + (V2(i, 3) - axis_lim(3)) / (axis_lim(4)-axis_lim(3)) * curr_ax(4);
 
 ax_size = 0.15;
-ax_pos = [.15 .15 ax_size ax_size;
-    .6 .15 ax_size ax_size;
-    .15 .75 ax_size ax_size;
-    .6 .75 ax_size ax_size];
+ax_pos = [.17 .16 ax_size ax_size;
+    .6 .16 ax_size ax_size;
[…]

### 2014-06-03 chemotaxis/prl_paper.tex
minor edits to chemotaxi spaper

@@
 
-Using a stochastic model which arises in cellular chemotaxis, we illustrate how manifold learning techniques, such as diffusion maps, can be used to characterize the behavior of a dynamical system.
+Using a stochastic model of particle dynamics which arises in cellular chemotaxis, we illustrate how diffusion maps, a manifold learning technique, can be used to characterize the behavior of a dynamical system.
 %
-In our model, depending on the value of a single parameter, the system dynamics at the macroscale range from heat equation dynamics, where particles diffuse uniformly outwards, to wave equation dynamics, where particle propagate with a fixed speed.
+In our model, depending on the value of a single parameter, the macroscopic system dynamics range from heat equation dynamics, where particles diffuse uniformly outwards, to wave equation dynamics, where particle propagate with a fixed speed.
 %
-We demonstrate that diffusion maps with the appropriate observers and metric can recover the two macroscopic system variables (initial condition and time) from data collected at the microscale in a variable- and equation-free manner.
+We demonstrate that diffusion maps with the appropriate observers and metric can recover the two governing macroscopic system variables (initial condition and time) from data collected at the microscale.
+% in a variable- and equation-free manner.
 %
 Furthermore, the order in which these variables are recovered is indicative of the system dynamics over the timescales of interest, 
-and the eigenvalue spectrum indicates where the switch in system dynamics occurs.
+and the eigenvalue spectrum indicates where this switch in system dynamics occurs.
 %
@@
 %
-For example, fluid flow, which can be described via the position and velocity of the fluid particles, is modeled at the macroscale using the Navier-Stokes equations.
+For example, fluid flow, which can be described via the position and velocity of every fluid particle, is modeled at the macroscale using the Navier-Stokes equations.
 %
@@
 %
-We would like to {\em automatically} uncover the macroscale behavior from data collected at the microscale without knowing or assuming any microscopic or macroscopic model {\em a priori}.
+We would like to {\em automatically} uncover the macroscale behavior from data collected at the microscale without any {\em a priori} knowledge of the appropriate microscopic or macroscopic model.
 %
@@
 In this letter, we will present how this goal can be achieved through a data-driven method based on geometric analysis and manifold learning.
-Specifically, we aim to emphasize two particular aspects of manifold learning from a dynamical systems' point of view: (a) the significance of the observers, especially in presence of noise and random dynamics, and (b) finding an appropriate metric of comparison.
-We will show that both are essential for our technique to successfully uncover the underlying dynamics of a given system.
+Specifically, we aim to emphasize two particular aspects of manifold learning from a dynamical systems' point of view: (a) choosing the appropriate observers, especially in presence of stochastic dynamics, and (b) selecting an appropriate metric for comparison.
+%
+We will show that both are essential for the algorithm to successfully uncover the underlying system dynamics.
 
-We will illustrate our methodology through a model problem arising in studies of cellular chemotaxis \cite{othmer2000diffusion}.
+We will illustrate our methodology using a model problem which arises in studies of cellular chemotaxis \cite{othmer2000diffusion}.
 %
@@
 %
-We will demonstrate how we can use diffusion maps \cite{coifman2005geometric}, a manifold learning technique, to {\em automatically} detect changes in a system's macroscopic dynamics from microscale data.
+We will demonstrate how we can use diffusion maps \cite{coifman2005geometric}, a manifold learning technique, to {\em automatically} detect changes in the system's macroscopic dynamics from microscale data.
 %
@@
 %
-Third, it has an analytic macroscopic description, which serves as a ground truth and will be used to verify our results, which will be obtained in an unsupervised manner.
+Third, it has an analytic macroscopic description, which serves as a ground truth and will allow us to verify our results which are obtained in an unsupervised manner.
 %
-We will show that diffusion maps based on statistical observers and the appropriate metric between them uncovers a description of the microscopic data which is consistent with the macroscopic model.
+We will show that diffusion maps based on statistical observers and the appropriate metric uncovers a description of the microscopic data which is consistent with the macroscopic model.
 %
-Furthermore, diffusion maps will uncover changes in dynamical behavior as we vary system parameters.
+Furthermore, diffusion maps will allow us to detect changes in dynamical behavior as we vary system parameters.
 
@@
 %
-Furthermore, we will show that diffusion maps can inform us as to the relative importance of system variables in different parameter regimes, and allow us to detect changes in macroscopic dynamical behavior.
+We will demonstrate that diffusion maps can inform us as to the relative importance of system variables in different parameter regimes, and allow us to detect changes in macroscopic dynamical behavior.
 
@@
 %
-We will examine one such model described by a velocity jump process \cite{othmer2000diffusion}.
+We will analyze one such model described by a velocity jump process \cite{othmer2000diffusion}.
 %
@@
 %
-This analysis shows that the initial distribution of the moving direction of the particles (given by $p$ in our microscopic simulations) plays a very different role depending on the value of $\lambda$.
+Two variables determine the macroscopic state of the system: the initial distribution, which is controlled by a single parameter $p$, and the time, $t$.
+%
+The above analysis shows that the initial distribution of the moving direction of the particles plays a very different role depending on the value of $\lambda$.
 %
@@
 %
-However, we would like to demonstrate that, in general, these PDEs are not necessary for us to uncover which system variables are important to describe the macroscopic dynamics.
+However, we will demonstrate that, in general, these PDEs are not necessary for us to uncover which system variables are important to describe the macroscopic dynamics.
 
@@
 
-We use diffusion maps \cite{coifman2005geometric} to analyze the data from simulations of the described microscopic stochastic model.
+We will use diffusion maps \cite{coifman2005geometric} to analyze the data from simulations of the described microscopic stochastic model.
 %

### 2014-06-04 chemotaxis/prl_paper.tex
small edits to chemotais paper

@@
 
-\title{Recovering System Dynamics using Diffusion Maps: A Chemotaxis Case Study}
+\title{Characterizing System Dynamics using Diffusion Maps: A Chemotaxis Case Study}
 
@@
 Furthermore, the order in which these variables are recovered is indicative of the system dynamics over the timescales of interest, 
-and the eigenvalue spectrum indicates where this switch in system dynamics occurs.
+and the eigenvalue spectrum indicates where this switch in dynamical behavior occurs.
 %
@@
 %
-Our goal is to show that diffusion maps can organize the data obtained from the microscopic simulations, and that this organization is consistent with the two system variables, $p$ and $t$, that emerge from the PDE analysis.
+Our goal is to show that diffusion maps can organize the data obtained from the microscopic simulations, and that this organization is consistent with the two system variables, $p$ and $t$, that emerge from the macroscopic PDE analysis.
 %
@@
 $\mu_j$ gives a measure of the ``importance'' of coordinate $\phi_j$,
-and we can quantify how important $\phi_1$ is relative to $\phi_2$ by comparing $\mu_1$ and $\mu_2$.
+and we can quantify the relative importance of $\phi_1$ versus $\phi_2$ by comparing $\mu_1$ and $\mu_2$.
 
@@
 %
-However, the Euclidean distance only considers the density at each bin and does not take into account the distance {\em between} bins.
+However, the Euclidean distance between histograms only considers the density at each bin and does not take into account the distance {\em between} bins.
 %,such that two distributions whose supports are a short distance apart are closer (in norm) than two distributions whose supports are very far from each other.
@@
 \includegraphics[width=8cm]{detect_change_eigenvalues}
-\caption{The first two eigenvalues from DMAPS as a function of $\lambda$. Note that the convergence of the eigenvalues indicates switching of the dynamical ``mode'' of the system.}
+\caption{The first two eigenvalues from DMAPS as a function of $\lambda$. The convergence of the eigenvalues at $\lambda \approx 125$ indicates switching of the dynamical ``mode'' of the system.}
 \label{fig:detect_change}
@@
 
-The crossover from ``wave-like'' to ``heat-like'' behavior can be seen in the DMAPS eigenvalues: a large gap between the first two eigenvalues indicates that one system variable is significantly more important and that we are in one of the asymptotic regimes.
+The crossover from ``wave-like'' to ``heat-like'' behavior can be seen in the DMAPS eigenvalues: a large gap between the first two eigenvalues indicates that one DMAPS variable is significantly more important and that we are in one of the two asymptotic regimes.
 %
@@
 
-%\bibliography{../../references/references}
+\bibliography{../../references/references}
 

### 2014-06-05 chemotaxis/prl_paper.tex
Ronen added edits to chemotaxis

@@
 
+TODO: rewrite according to the first 3 paragraphs of the introduction
+%
 Using a stochastic model of particle dynamics which arises in cellular chemotaxis, we illustrate how diffusion maps, a manifold learning technique, can be used to characterize the behavior of a dynamical system.
@@
 %
-We will show that both are essential for the algorithm to successfully uncover the underlying system dynamics.
+We will show that both are essential for the algorithm to successfully uncover the underlying system dynamics TODO:refMallat.
 
@@
 %
-We will show that diffusion maps based on statistical observers and the appropriate metric uncovers a description of the microscopic data which is consistent with the macroscopic model.
+We will show that diffusion maps based on statistical observers and with the appropriate affinity metric between these observers uncovers a description of the microscopic data which is consistent with the macroscopic model.
 %
-Furthermore, diffusion maps will allow us to detect changes in dynamical behavior as we vary system parameters.
+Furthermore, diffusion maps will allow us to detect changes in dynamical behavior arising from the variation of the system parameters.
 
@@
 %
-We will then use diffusion maps to analyze the microscopic simulation results, and show that using the appropriate metric for our data is essential to uncovering a meaningful description of our data.
+We will then use diffusion maps to analyze the microscopic simulation results, and show that using the appropriate metric for the data is essential to uncovering a meaningful description of our data.
 %
-We will demonstrate that diffusion maps can inform us as to the relative importance of system variables in different parameter regimes, and allow us to detect changes in macroscopic dynamical behavior.
+We will demonstrate that diffusion maps can identify the relative importance of system variables in different parameter regimes, and allow us to detect changes in macroscopic dynamical behavior.
 
@@
 
-We have a collection of $N$ particles whose states are defined by their positions and velocities. 
+Consider a collection of $N$ particles whose states are defined by their positions and velocities. 
 %
-Let $x_i(t)$ and $v_i(t)$ denote the position and velocity, respectively, of particle $i$ at time $t$.
+Let $x_i(t)$ and $v_i(t)$ denote the position and velocity of particle $i$ at time $t$, respectively.
 %
@@
 We initialize the particles such that
-\begin{equation}
+\begin{equation}\label{eqn:system}
 \begin{aligned}
@@
 \end{equation}
-where $p$ is some probability.
+where $p$ is an initial probability.
 %
@@
 
-For this specific example, we can write down explicitly the macroscopic equation that governs the overall system behavior.
+For this specific example, we can explicitly write the macroscopic equation that governs the overall system behavior.
 %
@@
 
-We will consider two regimes of simulation.
+We consider two asymptotic regimes of simulation.
 %
@@
 %
-However, we will demonstrate that, in general, these PDEs are not necessary for us to uncover which system variables are important to describe the macroscopic dynamics.
+However, in general, these PDEs are not necessarily known and there is a need to uncover the system variables, which govern the macroscopic dynamics.
 
@@
 
-We will use diffusion maps \cite{coifman2005geometric} to analyze the data from simulations of the described microscopic stochastic model.
+We use diffusion maps \cite{coifman2005geometric} to analyze the simulation data of the described microscopic stochastic model.
 %
-Our goal is to show that diffusion maps can organize the data obtained from the microscopic simulations, and that this organization is consistent with the two system variables, $p$ and $t$, that emerge from the macroscopic PDE analysis.
+Our goal is to organize the data obtained from the microscopic simulations consistently with the two system variables, $p$ and $t$, that emerge from the macroscopic PDE analysis.
 %
@@
 %
-In our case, although the data is very high-dimensional (e.g., the positions of all $N$ particles), for a fixed value of $\lambda$, there are only two important variables ($p$ and $t$) that determine the current (macroscopic) mode of the system; thus, our microscopic data {\em should} be two-dimensional and described by $p$ and $t$.
+In our case, although the data is very high-dimensional (e.g., the positions of all $N$ particles), for a fixed value of $\lambda$, there are only two important variables ($p$ and $t$) that determine the current (macroscopic) mode of the system; thus, the microscopic description of the data {\em should} be two-dimensional and described by $p$ and $t$.
 %
@@
 Assume we are given $m$ observations $z(1), \dots, z(m)$, which are assumed to lie on a $d$-dimensional, nonlinear manifold, with $d \ll n$.
+%
+TODO: add a couple of sentences to relate the notion of manifolds and parameterization to the dynamical system setup. Possibly use PCA reference.
 Our goal is to uncover a $d$-dimensional parameterization of the $m$ observations that respects the underlying manifold geometry.
@@
 %
-The next few eigenvectors give us the embedding coordinates for our data, such that
-$\phi_{j}(i)$ gives us the $j^{th}$ embedding coordinate for $z(i)$.
+The next few eigenvectors are viewed as the embedding coordinates for the data, such that
+$\phi_{j}(i)$ is the $j^{th}$ embedding coordinate for $z(i)$.
 %
-$\mu_j$ gives a measure of the ``importance'' of coordinate $\phi_j$,
-and we can quantify the relative importance of $\phi_1$ versus $\phi_2$ by comparing $\mu_1$ and $\mu_2$.
+The eigenvalue $\mu_j$ represents a measure of the ``importance'' of coordinate $\phi_j$,
+and hence, the relative importance of $\phi_1$ versus $\phi_2$ can be quantified by comparing $\mu_1$ and $\mu_2$.
 
-In order to use DMAPS to analyze our simulation results, we first need to address what are the appropriate observers or measurements for our system.
+TODO: rewrite - emphasize more the idea and what we propose (right from the beginning) instead of writing as a procedure.
+
+In order to properly analyze the simulation results, we first need to identify and compute the appropriate observers for the system.
 %
@@
 %
-Instead, we use histograms of the particle positions as our observables, so that $z(i) \in \mathbb{R}^n$ will be the histogram of $x_1(i), x_2(i), \dots, x_N(i)$ into $n$ bins (for our simulations, we take $n=32$).
+Instead, we propose to use histograms of the particle positions as the observables TODO:refPNAS.
+Let $z(i) \in \mathbb{R}^n$ be the histogram of $x_1(i), x_2(i), \dots, x_N(i)$ into $n$ bins (for our simulations, we set $n=32$).
 %
@@
 %
-For this particular example, using histograms is consistent with our macroscopic PDE description, which is written in terms of the probability density of the particles.
+For this particular example, using histograms is consistent with the macroscopic PDE description, which is written in terms of the probability density of the particles.
 %
@@
 
-The second point we need to address is choosing appropriate distance metric between data points in \eqref{eq:W}.
+Once the proper observers are determined, there is a need to define an appropriate distance metric between observations in \eqref{eq:W}.
 %
-Often, one uses the Euclidean distance between data points.
+Traditionally, the Euclidean distance between data samples is used.
 %
-However, the Euclidean distance between histograms only considers the density at each bin and does not take into account the distance {\em between} bins.
+However, the Euclidean distance between histograms only accounts for the density at each bin and does not take into account the distance {\em between} bins.
 %,such that two distributions whose supports are a short distance apart are closer (in norm) than two distributions whose supports are very far from each other.
@@
 
-We simulate the dynamics of this stochastic system to obtain microscopic data to which we apply diffusion maps based on histograms and the EMD.
+We simulate the dynamics of the studies stochastic system \eqref{eqn:system} to obtain microscopic data to which we apply diffusion maps based on histograms and the EMD.
 %
-We consider two sets of simulations:
+We consider two sets of simulations: TODO: associate the sets to the asymptotic regimes.
 in one set (small $\lambda$), $\lambda = 1$ and $s=1$, and in the other set (large $\lambda$), $\lambda = 400$, and $s=20$.
@@
 %
-We discard the initial portion of each simulation trajectory corresponding to $t < 1$, as this is very noisy and corresponds to initial relaxation of the system.
+We discard the initial portion of each simulation trajectory corresponding to $t < 1$, as it tends to be very noisy and corresponds to initial relaxation of the system.
 
-\begin{figure}[t] 
-\includegraphics[width=8cm]{detect_change_eigenvalues}
-\caption{The first two eigenvalues from DMAPS as a function of $\lambda$. The convergence of the eigenvalues at $\lambda \approx 125$ indicates switching of the dynamical ``mode'' of the system.}
-\label{fig:detect_change}
-\end{figure}
+%\begin{figure}[t] 
+%\includegraphics[width=8cm]{detect_change_eigenvalues}
+%\caption{The first two eigenvalues from DMAPS as a function of $\lambda$. The convergence of the eigenvalues at $\lambda \approx 125$ indicates switching of the dynamical ``mode'' of the system.}
+%\label{fig:detect_change}
+%\end{figure}
 
-Figure \ref{fig:dmaps_embed_emd} shows the diffusion maps embedding of our data.
+Figure \ref{fig:dmaps_embed_emd} shows the diffusion maps embedding of the data.
 %
@@
 %
+TODO: give the values of the corresponding eigenvalues.
 However, the ``more important'' coordinate, $\phi_1$, is correlated with $p$ for the small $\lambda$ case (Figure~\ref{subfig:small_lambda_p}), and correlated with $t$ for the large $\lambda$ case (Figure~\ref{subfig:large_lambda_t});
@@
 %
-This is not surprising: at small times, the particles are more condensed around $x=0$, and it is more difficult to distinguish the particles moving to the left from the particles moving to the right. 
+This is consistent with the model: at small times, the particles are more condensed around $x=0$, and it is more difficult to distinguish the particles moving to the left from the particles moving to the right.
 %
@@
 
+TODO: describe as a new experiment with a new data set. The goal of this experiment is to show implicit data-driven phase shift detection.
 The crossover from ``wave-like'' to ``heat-like'' behavior can be seen in the DMAPS eigenvalues: a large gap between the first two eigenvalues indicates that one DMAPS variable is significantly more important and that we are in one of the two asymptotic regimes.
@@
 
+TODO: just report the results of figure 3 + replace figure 3 with two scatter plots (for example, phi1 vs. p with and without EMD).
 To emphasize the importance of using the correct distance metric, we also computed the diffusion maps embeddings of our two sets of simulation data, used the standard Euclidean distance between the histograms to compute the distances in \eqref{eq:W}; the results are shown in Figure \ref{fig:dmaps_embed_noemd}.
@@
 %
-This is due to our poor choice of metric, which does not accurately describe the distances between histograms.
+This is due to the inappropriate metric, which does not accurately describe the distances between histograms.
 

### 2014-06-06 chemotaxis/prl_paper.tex
started workong on prl edits
edited prl paper some more

@@
 
-TODO: rewrite according to the first 3 paragraphs of the introduction
+Many dynamical systems are governed by stochastic, noisy motions at the microscale which give rise to smooth, coherent dynamics at the macroscale.
 %
-Using a stochastic model of particle dynamics which arises in cellular chemotaxis, we illustrate how diffusion maps, a manifold learning technique, can be used to characterize the behavior of a dynamical system.
+The relationship between the microscopic and macroscopic dynamics is often not immediately obvious and either requires some modeling assumptions or extensive observation of the system.
 %
-In our model, depending on the value of a single parameter, the macroscopic system dynamics range from heat equation dynamics, where particles diffuse uniformly outwards, to wave equation dynamics, where particle propagate with a fixed speed.
+We demonstrate how extracting macroscopic dynamics from microscale data can be accomplished automatically using data-driven algorithms. 
 %
-We demonstrate that diffusion maps with the appropriate observers and metric can recover the two governing macroscopic system variables (initial condition and time) from data collected at the microscale.
-% in a variable- and equation-free manner.
+These algorithms need no {\em a priori} knowledge about the appropriate macroscopic model, but instead, only require the definition of the appropriate observers and metric in order to recover the governing variables.
 %
-Furthermore, the order in which these variables are recovered is indicative of the system dynamics over the timescales of interest, 
-and the eigenvalue spectrum indicates where this switch in dynamical behavior occurs.
-%
-%We demonstrate that an essential component of our analysis is working with the right observers (histograms/ensembles, rather than the single position of a particle) and with the right metric between those observers (earth mover's distance, rather than Euclidean distance).
-%
-%In heat equation mode, we find that time is more important than the initial conditions, as the particles quickly equilibrate.
-%
-%In wave equation mode, we find that the initial conditions are more important, as the initial distribution simply propagates in time.
-%	
-%Furthermore, these two components have a completely different ``nature". 
-%
-%One parameter (the initial condition) is governing the entire trajectory and the other parameter (time) is governing the ``step­--by­--step" evolution within each trajectory.
+We illustrate these ideas using a model which arises in studies of cellular chemotaxis.
 

### 2014-06-08 chemotaxis/prl_paper.tex
updated draft with ronen edits

@@
 %
-We demonstrate how extracting macroscopic dynamics from microscale data can be accomplished automatically using data-driven algorithms. 
+We demonstrate how extracting macroscopic dynamics from microscale data can be accomplished automatically using manifold learning techniques. 
 %
-These algorithms need no {\em a priori} knowledge about the appropriate macroscopic model, but instead, only require the definition of the appropriate observers and metric in order to recover the governing variables.
+These data-driven algorithms need no {\em a priori} knowledge about the appropriate macroscopic model, but instead, only require the appropriate observers and pairwise metric between observations in order to recover the governing macroscopic variables.
+%
+We also show how these algorithms can detect changes in the macroscopic dynamical behavior as a function of microscopic system parameters. 
 %
@@
 In this letter, we will present how this goal can be achieved through a data-driven method based on geometric analysis and manifold learning. 
-Specifically, we aim to emphasize two particular aspects of manifold learning from a dynamical systems' point of view: (a) choosing the appropriate observers, especially in presence of stochastic dynamics, and (b) selecting an appropriate metric for comparison.
 %
-We will show that both are essential for the algorithm to successfully uncover the underlying system dynamics TODO:refMallat.
+Specifically, we aim to emphasize two particular aspects of manifold learning from a dynamical systems' point of view: (a) choosing the appropriate observers, especially in presence of stochastic dynamics, and (b) selecting an appropriate metric for comparison \cite{mallat2012group}.
+%
+We will show that both are essential for data-driven algorithms to successfully uncover the underlying system dynamics.
 
-We will illustrate our methodology using a model problem which arises in studies of cellular chemotaxis \cite{othmer2000diffusion}.
+We will illustrate these points using a model problem which arises in studies of cellular chemotaxis \cite{othmer2000diffusion}.
 %
@@
 %
-We will show that diffusion maps based on statistical observers and with the appropriate affinity metric between these observers uncovers a description of the microscopic data which is consistent with the macroscopic model.
+We will show that manifold learning algorithms based on statistical observers and with the appropriate affinity metric between these observers uncovers a description of the microscopic data which is consistent with the macroscopic model.
 %
-Furthermore, diffusion maps will allow us to detect changes in dynamical behavior arising from the variation of the system parameters.
+Furthermore, these data-driven methods also allow us to detect changes in dynamical behavior arising from the variation of the system parameters.
 
-The remainder of this letter is organized as follows. We will begin by describing the chemotaxis problem.
+The remainder of this letter is organized as follows. 
 %
-We will present the stochastic microscopic model, and show how, for this example, the microscopic model gives rise to a compact macroscopic description governed by a single parameter.
+We will begin by describing the chemotaxis problem, the illustrative example which we will use to demonstrate the methodologies.
 %
-We will then use diffusion maps to analyze the microscopic simulation results, and show that using the appropriate metric for the data is essential to uncovering a meaningful description of our data.
+We will present the stochastic microscopic model, and use diffusion maps, a nonlinear dimensionality reduction technique, to analyze microscopic simulation results.
 %
-We will demonstrate that diffusion maps can identify the relative importance of system variables in different parameter regimes, and allow us to detect changes in macroscopic dynamical behavior.
+We will show that we can recover the governing macroscopic variables, provided that we use the appropriate observers and distance metric.
+%
+We will then describe how, for this example, the microscopic model gives rise to a compact macroscopic description governed by a single parameter, and show that the diffusion maps results are consistent with this macroscopic description.
+%
+Finally, we will demonstrate that diffusion maps can also inform us as to the relative importance of system variables in different parameter regimes, and allow us to detect changes in macroscopic dynamical behavior.
 
@@
 
-Chemotaxis is the movement of cells (such as bacteria) which is regulated by extracellular sensed signals.
+Biological cells are one common example of a system where microscopic entities must work together to exhibit coherent macroscopic dynamics, to accomplish tasks such as finding food or navigating away from toxins.
 %
-These signals allow cells to accomplish tasks such as finding food and navigating away from toxins.
+The cellular dynamics are often governed by extracellular sensed signals; this movement is known as chemotaxis. 
 %
-Several models have been proposed to describe the dynamics of cellular movement and dispersal \cite{othmer1988models, codling2008random}.
+Several microscopic models have been proposed to describe chemotaxis dynamics \cite{othmer1988models, codling2008random}.
 %
@@
 %
-At random times, a cell will ``turn around'' and switch its velocity (this turning is controlled by extracellular signals). 
+At random times, a cell will ``turn around'' and switch the direction of its velocity (this turning is controlled by extracellular signals). 
 %
@@
 
-Consider a collection of $N$ particles whose states are defined by their positions and velocities. 
+The microscopic model consists of a collection of $N$ particles whose states are defined by their positions and velocities. 
 %
@@
 %Let $X_{p, \lambda, s}(t)$ denote the vector of positions of the particles at time $t$ with initial right probability $p$, switching rate $\lambda$, and speed $s$.
+
+For fixed values of $\lambda$ and $s$, the macroscopic state of the system can be described by two independent variables: $p$, which controls the initial distribution of the particles, and $t$, the time.
+%
+However, for more complex microscopic models, the governing macroscopic variables may not be immediately obvious.
+%
+In general, we expect the simulation data to lie on a low-dimensional manifold which is parameterized by the macroscopic variables.
+%
+For our particular example, this implies that our data should lie on a two-dimensional manifold  parameterized by $p$ and $t$, 
+and uncovering the low-dimensional structure in our data set will inform us as to the governing macroscopic variables.
+
+We propose using data driven techniques to uncover such low-dimensional structure.
+%
+Principal components analysis (PCA) \cite{...} is a traditional dimensionality reduction algorithm which projects the data onto the optimal low-dimensional linear subspace. 
+%
+Here, we use diffusion maps \cite{coifman2005geometric}, a manifold learning technique, to analyze the results of our microscopic chemotaxis simulations.
+%
+Diffusion maps is more general than PCA, in that it can find a parameterization of data which lie on a low-dimensional, {\em nonlinear} structure in high-dimensional space.
+%
+For our example, although the data lies in a high-dimensional ambient space (e.g., the positions of all $N$ particles), for a fixed value of $\lambda$, we expect the data to lie on a two-dimensional manifold described by $p$ and $t$.
+%
+
+\begin{figure*}[t!]
+\def\figwidth{6.4cm}
+\begin{subfigure}{\figwidth}
+\includegraphics[width=\textwidth]{EMD_withhist_p_1}
+\caption{}
+\label{subfig:small_lambda_p}
+\end{subfigure}
+\begin{subfigure}{\figwidth}
+\includegraphics[width=\textwidth]{EMD_withhist_t_1}
+\caption{}
+\label{subfig:small_lambda_t}
+\end{subfigure}
+\begin{subfigure}{\figwidth}
+\includegraphics[width=\textwidth]{EMD_withhist_p_400}
+\caption{}
+\label{subfig:large_lambda_p}
+\end{subfigure}
+\begin{subfigure}{\figwidth}
+\includegraphics[width=\textwidth]{EMD_withhist_t_400}
+\caption{}
+\label{subfig:large_lambda_t}
+\end{subfigure}
+\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with (a,b) $\lambda=1$, $s=1$, and (c,d) $\lambda=400$, $s=20$.
+For each set, we run many simulations, each with $N=1000$ particles.
+%
+We vary the initial conditions such that $0.1 \le p  \le 0.9$, and allow each simulation to evolve for $10$ time units.
+%
+We discard the initial portion of each simulation trajectory corresponding to $t < 1$, as it tends to be very noisy and corresponds to initial relaxation of the system.
+%
+The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions.
+The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time.
+Representative histograms are shown for selected data points.}
+\label{fig:dmaps_embed_emd}
+\end{figure*}
+
+In general, manifold learning techniques require two essential components.
+%
+One, they require definition of the appropriate features of {\em observers} to describe the system. 
+%
+These observers should be informative as to the state of the system, as well as be (somewhat) invariant to noise in the system. 
+%
+Two, these algorithms require an informative {\em distance metric} between the observers.
+%
+This metric should accurately capture our idea of locality: observations which we consider to be similar should have a small distance. 
+%
+For this particular example, we use histograms of the particle positions as our observers TODO:refPNAS. 
+%
+Histograms are invariant to the indexing of the particles, while retaining the spatial locations of the particles.
+%
+We denote the histogram of $x_1(i), x_2(i), \dots, x_N(i)$ into $n$ bins (for our simulations, we set $n=32$) as $z(i) \in \mathbb{R}^n$.
+%
+Instead of the standard Euclidean metric, we use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between pairs of histograms. 
+%
+Conceptually, EMD measures how much ``work'' it takes to transform one probability density into another.
+%
+It therefore not only considers where the densities are inconsistent, but also how far apart the inconsistencies are.
+%
+Although the brute-force computation of the EMD is computationally expensive, there has been a plethora of work in developing efficient algorithms for the computation of EMD \cite{Pele-eccv2008, Pele-iccv2009}.
+%
+%TODO: we need to pay a special attention to the implementation of the EMD.
+We need to make it cleat that this distance can be very efficiently implemented  and therefore it is suitable for the practical uses and algorithm that we propose here as a viable tool for dynamical system analysis.
+%
+For the special case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions \cite{rubner2000perceptual}.
+%
+This can be estimated from histograms, 
+\begin{equation}
+\| z(i) - z(j) \|_{EMD} = \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|
+\end{equation}
+where $z(i), z(j) \in \mathbb{R}^n$ are two histograms defined on (equally-spaced) bins in $\mathbb{R}$.
+
+Given a set of observations with the appropriate metric, diffusion maps can then uncover a parameterization of these observations. 
+%
+Let $z(t) \in \mathbb{R}^n$ denote an $n$-dimensional observation of the system at time $t$. 
+%
+Assume we are given $m$ observations $z(1), \dots, z(m)$, which are assumed to lie on a $d$-dimensional, nonlinear manifold, with $d \ll n$. 
+%
+%TODO: add a couple of sentences to relate the notion of manifolds and parameterization to the dynamical system setup. Possibly use PCA reference.
+Our goal is to uncover a $d$-dimensional parameterization of the $m$ observations that respects the underlying manifold geometry.
+%
+We first construct the matrix $W \in \mathbb{R}^{m \times m}$, with
+\begin{equation} \label{eq:W}
+W_{ij} = \exp \left( -\frac{\|z(i) - z(j) \|^2}{\epsilon^2} \right), \ i,j=1,\ldots,m
+\end{equation}
+where $\| \cdot \|$ denotes the appropriate norm for the data, and $\epsilon$ is a characteristic distance between the observations (we take $\epsilon$ to be the median of the pairwise distances between the data points).
+%
+We then construct the diagonal matrix $D \in \mathbb{R}^{m \times m}$, with $D_{ii} = \sum_j W_{ij}$, and the matrix $A  = D^{-1} W.$
+%
+The eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ then give the embedding coordinates for the data, such that
+$\phi_{j}(i)$ is the $j^{th}$ embedding coordinate for $z(i)$.
+%
+The ``importance'' of coordinate $\phi_j$ is quantified by the corresponding eigenvalue $\mu_j$.
+%
+We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$, and if the data are low-dimensional, we only need to retain a few of the eigenvectors to describe the data.
+%
+The relative importance of $\phi_1$ versus $\phi_2$ can be quantified by comparing $\mu_1$ and $\mu_2$.
+%
+Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$), which implies $\mu_0 = 1$ and $\phi_0$ is a constant vector which is never used as an embedding coordinate.
+
+Figure~\ref{fig:dmaps_embed_emd} shows the results of analyzing two sets of chemotaxis simulations using diffusion maps. 
+%
+One set of simulations is for a small value of $\lambda$, and the other set is for the case of large $\lambda$. 
+%
+In both cases, the two macroscopic variables, $p$ and $t$, are well-correlated with the first two diffusion maps coordinates, $\phi_1$ and $\phi_2$. 
+%
+In both cases, $\phi_1$ is significantly more important than $\phi_2$, seen by the separation between the first two eigenvalues (for the small $\lambda$ case, $\mu_1 = 0.5138$, and $\mu_2 = 0.3943$, and for the large $\lambda$ case, $\mu_1 = 0.5543$, and $\mu_2 = 0.3807$).
+%
+$\phi_1$, the ``more important'' coordinate, is correlated with $p$ for the small $\lambda$ case (Figure~\ref{subfig:small_lambda_p}), and correlated with $t$ for the large $\lambda$ case (Figure~\ref{subfig:large_lambda_t}), indicating that the relative importance of $p$ and $t$ changes in the two simulations.
+
+%correlations
+%small lambda
+%Euclidean: 3.3848e-04, -0.1861
+%EMD: 0.9000, 0.9807
+ %   
+%large lambda
+%Euclidean: -0.5466, 0.3329
+%EMD: 0.9329, 0.9832
+
+%TODO: just report the results of figure 3 + replace figure 3 with two scatter plots (for example, phi1 vs. p with and without EMD).
+To emphasize the importance of using the correct distance metric, we also computed the diffusion maps embeddings of our two sets of simulation data, used the standard Euclidean distance between the histograms to compute the distances in \eqref{eq:W}.
+%
+There is no appreciable correlation between the embedding coordinates and the macroscopic variables $p$ and $t$. 
+%
+The correlations between the embedding coordinates and the macroscopic variables are all $< 0.6$.
+%
+This is due to the inappropriate metric, which does not accurately describe the distances between histograms.
+%
+In contrast, the correlations between the embedding coordinates and the macroscopic variables when using EMD in the diffusion maps calculation are all $> 0.9$.
+
+%We simulate the dynamics of the studies stochastic system \eqref{eqn:system} to obtain microscopic data to which we apply diffusion maps based on histograms and the EMD.
+%
+%We consider two sets of simulations: TODO: associate the sets to the asymptotic regimes.
+%in one set (small $\lambda$), $\lambda = 1$ and $s=1$, and in the other set (large $\lambda$), $\lambda = 400$, and $s=20$.
+%
+%For each set, we run many simulations, each with $N=1000$ particles.
+%
+%We vary the initial conditions such that $0.1 \le p  \le 0.9$, and allow each simulation to evolve for $10$ time units.
+%
+%We discard the initial portion of each simulation trajectory corresponding to $t < 1$, as it tends to be very noisy and corresponds to initial relaxation of the system.
+
+%Figure \ref{fig:dmaps_embed_emd} shows the diffusion maps embedding of the data.
+%
+%The two system variables, $p$ and $t$, are well-correlated with the first two diffusion maps coordinates, $\phi_1$ and $\phi_2$. 
+%
+%TODO: give the values of the corresponding eigenvalues.
+%However, the ``more important'' coordinate, $\phi_1$, is correlated with $p$ for the small $\lambda$ case (Figure~\ref{subfig:small_lambda_p}), and correlated with $t$ for the large $\lambda$ case (Figure~\ref{subfig:large_lambda_t}).
 
@@
 %
-Two variables determine the macroscopic state of the system: the initial distribution, which is controlled by a single parameter $p$, and the time, $t$.
+%Two variables determine the macroscopic state of the system: the initial distribution, which is controlled by a single parameter $p$, and the time, $t$.
 %
-The above analysis shows that the initial distribution of the moving direction of the particles plays a very different role depending on the value of $\lambda$.
+The above analysis shows that the initial distribution of the moving direction of the particles (determined by $p$ in the microscopic simulations) plays a very different role depending on the value of $\lambda$.
 %
@@
 When $\lambda \rightarrow \infty$, the dynamics are described by one heat equation, and the initial conditions are insignificant -- the velocity distribution quickly equilibrates and we see purely diffusive behavior.
+
+We can now easily justify the features that we see in the low-dimensional embeddings in Figure~\ref{fig:dmaps_embed_emd}.
 %
-For this particular system, we can deduce this behavior from the macroscopic PDEs which describe the system dynamics.
+Firstly, the observation that in the small $\lambda$ case, $p$, which governs the initial condition, is correlated with $\phi_1$ and therefore more important than $t$, which is correlated with $\phi_2$, is consistent with the analytic macroscopic description.
 %
-However, in general, these PDEs are not necessarily known and there is a need to uncover the system variables, which govern the macroscopic dynamics.
-
-\begin{figure*}[t!]
-\def\figwidth{6.4cm}
-\begin{subfigure}{\figwidth}
-\includegraphics[width=\textwidth]{EMD_withhist_p_1}
-\caption{}
-\label{subfig:small_lambda_p}
-\end{subfigure}
-\begin{subfigure}{\figwidth}
-\includegraphics[width=\textwidth]{EMD_withhist_t_1}
-\caption{}
-\label{subfig:small_lambda_t}
-\end{subfigure}
-\begin{subfigure}{\figwidth}
-\includegraphics[width=\textwidth]{EMD_withhist_p_400}
-\caption{}
-\label{subfig:large_lambda_p}
-\end{subfigure}
-\begin{subfigure}{\figwidth}
-\includegraphics[width=\textwidth]{EMD_withhist_t_400}
-\caption{}
-\label{subfig:large_lambda_t}
-\end{subfigure}
-\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with (a,b) $\lambda=1$, $s=1$, and (c,d) $\lambda=400$, $s=20$.
-The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions.
-The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time.
-Representative histograms are shown for selected data points.}
-\label{fig:dmaps_embed_emd}
-\end{figure*}
-
-%\section{Methods and Analysis}
-
-%\subsection{Diffusion maps}
-
-We use diffusion maps \cite{coifman2005geometric} to analyze the simulation data of the described microscopic stochastic model.
+Furthermore, in the small $\lambda$ regime (wave equation), shown in Figures \ref{subfig:small_lambda_p} and \ref{subfig:small_lambda_t}, the points corresponding to small times are more tightly clustered than the points corresponding to large times.
 %
-Our goal is to organize the data obtained from the microscopic simulations consistently with the two system variables, $p$ and $t$, that emerge from the macroscopic PDE analysis.
+This is consistent with the macroscopic model: at small times, the particles are more condensed around $x=0$, and it is more difficult to distinguish the particles moving to the left from the particles moving to the right.
 %
-Diffusion maps, in general, is a nonlinear manifold learning technique that embeds (maps) the data into a lower dimensional space.
-Ideally, the dimension of the embedded samples is equal to the the intrinsic (true) dimension of the data, and the variables/coordinates of the embedded samples represent the governing variables of the system.
+On the other hand, at large times, once the particles evolve from the origin, this separation is clear.  
 %
-In our case, although the data is very high-dimensional (e.g., the positions of all $N$ particles), for a fixed value of $\lambda$, there are only two important variables ($p$ and $t$) that determine the current (macroscopic) mode of the system; thus, the microscopic description of the data {\em should} be two-dimensional and described by $p$ and $t$.
+For the large $\lambda$ case (heat equation), shown in Figures \ref{subfig:large_lambda_p} and \ref{subfig:large_lambda_t}, we observe that for small time the initial distribution $p$ is well organized in the embedding in Figure \ref{subfig:large_lambda_p}, the distribution of the particles is initially skewed and the initial velocity plays a role.
 %
-Let $z(t) \in \mathbb{R}^n$ denote an $n$-dimensional observation of the system at time $t$.
-Assume we are given $m$ observations $z(1), \dots, z(m)$, which are assumed to lie on a $d$-dimensional, nonlinear manifold, with $d \ll n$.
-%
-TODO: add a couple of sentences to relate the notion of manifolds and parameterization to the dynamical system setup. Possibly use PCA reference.
-Our goal is to uncover a $d$-dimensional parameterization of the $m$ observations that respects the underlying manifold geometry.
-%
-We first construct the matrix $W \in \mathbb{R}^{m \times m}$, with
-\begin{equation} \label{eq:W}
-W_{ij} = \exp \left( -\frac{\|z(i) - z(j) \|^2}{\epsilon^2} \right), \ i,j=1,\ldots,m
-\end{equation}
-where $\| \cdot \|$ denotes the appropriate norm for the data (which we will discuss in the following section), and $\epsilon$ is a characteristic distance between the observations (we take $\epsilon$ to be the median of the pairwise distances between the data points).
-%
-We then construct the diagonal matrix $D \in \mathbb{R}^{m \times m}$, with $D_{ii} = \sum_j W_{ij}$, and the matrix $A$, defined as
-\begin{equation}
-A = D^{-1} W.
-\end{equation}
-%
-We compute the eigenvalues $\mu_0, \mu_1, \dots, \mu_{m-1}$ and eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$, and order them such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$. 
-%
-Note that the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$), which implies $\mu_0 = 1$ and $\phi_0$ is a constant vector.
-%
-The next few eigenvectors are viewed as the embedding coordinates for the data, such that
-$\phi_{j}(i)$ is the $j^{th}$ embedding coordinate for $z(i)$.
-%
-The eigenvalue $\mu_j$ represents a measure of the ``importance'' of coordinate $\phi_j$,
-and hence, the relative importance of $\phi_1$ versus $\phi_2$ can be quantified by comparing $\mu_1$ and $\mu_2$.
-
-TODO: rewrite - emphasize more the idea and what we propose (right from the beginning) instead of writing as a procedure.
-
-In order to properly analyze the simulation results, we first need to identify and compute the appropriate observers for the system.
-%
-Because the particles are indistinguishable, vectors containing the particle positions are not appropriate to analyze, as we would then be assigning indices or labels to the particles and swapping two particles would change the data.
-%
-Instead, we propose to use histograms of the particle positions as the observables TODO:refPNAS.
-Let $z(i) \in \mathbb{R}^n$ be the histogram of $x_1(i), x_2(i), \dots, x_N(i)$ into $n$ bins (for our simulations, we set $n=32$).
-%
-Clearly, histograms are invariant to the labeling or indexing of the particles.
-%
-For this particular example, using histograms is consistent with the macroscopic PDE description, which is written in terms of the probability density of the particles.
-%
-The histograms serve as an empirical estimate of the probability density at each time point.
-
-%\subsection{Earth mover's distance}
-
-Once the proper observers are determined, there is a need to define an appropriate distance metric between observations in \eqref{eq:W}.
-%
-Traditionally, the Euclidean distance between data samples is used.
-%
-However, the Euclidean distance between histograms only accounts for the density at each bin and does not take into account the distance {\em between} bins.
-%,such that two distributions whose supports are a short distance apart are closer (in norm) than two distributions whose supports are very far from each other.
-%
-%however, for histograms which approximate a probability density, the Euclidean distance can often be misleading.
-%
-%For example, the distance between two uniform distributions with finite, non-overlapping support will always be the same, {\em independent} of the distance between the support intervals.
-%
-To overcome this critical shortcoming, we propose to use the earth mover's distance (EMD) \cite{rubner2000earth} as our metric between histograms.
-%
-Conceptually, EMD measures how much ``work'' it takes to transform one probability density into another.
-%
-It therefore not only considers where the densities are inconsistent, but also how far apart the inconsistencies are.
-%
-Although the brute-force computation of the EMD is computationally expensive, there has been a plethora of work in developing efficient algorithms for the computation of EMD \cite{Pele-eccv2008, Pele-iccv2009}.
-%
-%TODO: we need to pay a special attention to the implementation of the EMD.
-We need to make it cleat that this distance can be very efficiently implemented  and therefore it is suitable for the practical uses and algorithm that we propose here as a viable tool for dynamical system analysis.
-%
-%For example, the EMD algorithm from \cite{Pele-eccv2008, Pele-iccv2009}....
-%
-%The implementation is available at \url{http://www.cs.huji.ac.il/~ofirpele/FastEMD/code}. 
-For the special case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions \cite{rubner2000perceptual}.
-%\begin{equation}
-%\| \rho_1  - \rho_2 \|_{EMD} = \int_{-\infty}^{\infty} \left| \int_{-\infty}^x \rho_1(y) dt - \int_{-\infty}^x \rho_2(y) dy \right| dx
-%\end{equation}
-%where $\rho_1(x)$ and $\rho_2(x)$ are two probability distributions defined on $x \in \mathbb{R}$. 
-%
-This can be estimated using histograms, 
-\begin{equation}
-\| z(i) - z(j) \|_{EMD} = \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|
-\end{equation}
-where $z(i), z(j) \in \mathbb{R}^n$ are two histograms defined on (equally-spaced) bins in $\mathbb{R}$. 
-%\section{Results}
+On the other hand, for large times, we observe that the initial distribution $p$ is less organized in Figure \ref{subfig:large_lambda_p}, as the velocities have equilibrated and the initial distribution is less detectable in the particle density.
 
@@
 
-We simulate the dynamics of the studies stochastic system \eqref{eqn:system} to obtain microscopic data to which we apply diffusion maps based on histograms and the EMD.
-%
-We consider two sets of simulations: TODO: associate the sets to the asymptotic regimes.
-in one set (small $\lambda$), $\lambda = 1$ and $s=1$, and in the other set (large $\lambda$), $\lambda = 400$, and $s=20$.
-%
-For each set, we run many simulations, each with $N=1000$ particles.
-%
-We vary the initial conditions such that $0.1 \le p  \le 0.9$, and allow each simulation to evolve for $10$ time units.
-%
-We discard the initial portion of each simulation trajectory corresponding to $t < 1$, as it tends to be very noisy and corresponds to initial relaxation of the system.
-
-%\begin{figure}[t] 
-%\includegraphics[width=8cm]{detect_change_eigenvalues}
-%\caption{The first two eigenvalues from DMAPS as a function of $\lambda$. The convergence of the eigenvalues at $\lambda \approx 125$ indicates switching of the dynamical ``mode'' of the system.}
-%\label{fig:detect_change}
-%\end{figure}
-
-Figure \ref{fig:dmaps_embed_emd} shows the diffusion maps embedding of the data.
-%
-The two system variables, $p$ and $t$, are well-correlated with the first two diffusion maps coordinates, $\phi_1$ and $\phi_2$. 
-%
-TODO: give the values of the corresponding eigenvalues.
-However, the ``more important'' coordinate, $\phi_1$, is correlated with $p$ for the small $\lambda$ case (Figure~\ref{subfig:small_lambda_p}), and correlated with $t$ for the large $\lambda$ case (Figure~\ref{subfig:large_lambda_t});
-this is consistent with the macroscopic PDE description.
-%
-Furthermore, in the small $\lambda$ regime (wave equation), shown in Figures \ref{subfig:small_lambda_p} and \ref{subfig:small_lambda_t}, the points corresponding to small times are more tightly clustered than the points corresponding to large times.
-%
-This is consistent with the model: at small times, the particles are more condensed around $x=0$, and it is more difficult to distinguish the particles moving to the left from the particles moving to the right.
-%
-On the other hand, at large times, once the particles evolve from the origin, this separation is clear.  
-%
-For the large $\lambda$ case (heat equation), shown in Figures \ref{subfig:large_lambda_p} and \ref{subfig:large_lambda_t}, we observe that for small time the initial distribution $p$ is well organized in the embedding in Figure \ref{subfig:large_lambda_p}, the distribution of the particles is initially skewed and the initial velocity plays a role.
-%
-On the other hand, for large times, we observe that the initial distribution $p$ is less organized in Figure \ref{subfig:large_lambda_p}, as the velocities have equilibrated and the initial distribution is less detectable in the particle density.
+\begin{figure}[t] 
+\includegraphics[width=8cm]{detect_change_eigenvalues}
+\caption{The first two eigenvalues from DMAPS as a function of $\lambda$. Each point is averaged over 10 simulation experiments, with error bars indicating one standard deviation. 
+The convergence of the eigenvalues at $\lambda \approx 125$ indicates switching of the dynamical ``mode'' of the system.}
+\label{fig:detect_change}
+\end{figure}
 
@@
 
-TODO: describe as a new experiment with a new data set. The goal of this experiment is to show implicit data-driven phase shift detection.
-The crossover from ``wave-like'' to ``heat-like'' behavior can be seen in the DMAPS eigenvalues: a large gap between the first two eigenvalues indicates that one DMAPS variable is significantly more important and that we are in one of the two asymptotic regimes.
+We have seen that the eigenvalues from diffusion maps, which quantify the importance of each embedding coordinate, can inform us as to the relative importance of the variables which govern the macroscopic dynamics.
 %
-Figure~\ref{fig:detect_change} shows the top two DMAPS eigenvalues, $\mu_1$ and $\mu_2$, as a function of $\lambda$. 
+We can also use these eigenvalues to detect the change in dynamical behavior.
 %
-Each data point is averaged over 10 data sets, with error bars indicating one standard deviation. 
+We have seen that at the two asymptotic regimes ($\lambda \rightarrow 0$ and $\lambda \rightarrow \infty$), there will be a gap between the two eigenvalues, as one of the macroscopic variables is significantly more important to the dynamical behavior.
+%
+For intermediate $\lambda$ values, we expect the gap between $|\mu_1|$ and $|\mu_2|$ to narrow, and that $|\mu_1| \approx |\mu_2|$ corresponds to the transition between ``wave-like'' and ``heat-like'' behavior.
+
+Figure~\ref{fig:detect_change} shows the first two eigenvalues, $\mu_1$ and $\mu_2$, computed from simulation data, as a function of $\lambda$.
+% 
+%TODO: describe as a new experiment with a new data set. The goal of this experiment is to show implicit data-driven phase shift detection.
 %
@@
 The eigenvalues then separate again at large $\lambda$ (the heat equation regime), where the $t$ effects dominate the dynamics.
+%
+We can therefore detect where the change in dynamical behavior occurs using data-driven techniques.
 
@@
 
-\begin{figure}[t]
-\def\figwidth{4.2cm}
-\begin{subfigure}{\figwidth}
-\includegraphics[width=\textwidth]{rawhist_p_1}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{\figwidth}
-\includegraphics[width=\textwidth]{rawhist_t_1}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{\figwidth}
-\includegraphics[width=\textwidth]{rawhist_p_400}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{\figwidth}
-\includegraphics[width=\textwidth]{rawhist_t_400}
-\caption{}
-\end{subfigure}
-\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with (a,b) $\lambda=1$, $s=1$, and (c,d) $\lambda=400$, $s=20$.
+%\begin{figure}[t]
+%\def\figwidth{4.2cm}
+%\begin{subfigure}{\figwidth}
+%\includegraphics[width=\textwidth]{rawhist_p_1}
+%\caption{}
+%\end{subfigure}
+%\begin{subfigure}{\figwidth}
+%\includegraphics[width=\textwidth]{rawhist_t_1}
+%\caption{}
+%\end{subfigure}
+%\begin{subfigure}{\figwidth}
+%\includegraphics[width=\textwidth]{rawhist_p_400}
+%\caption{}
+%\end{subfigure}
+%\begin{subfigure}{\figwidth}
+%\includegraphics[width=\textwidth]{rawhist_t_400}
+%\caption{}
+%\end{subfigure}
+%\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with (a,b) $\lambda=1$, $s=1$, and (c,d) $\lambda=400$, $s=20$.
 The distances used in the diffusion maps kernel are the Euclidean distances between the histograms of particle positions.
 The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time.}
-\label{fig:dmaps_embed_noemd}
-\end{figure}
-%TODO: for all figures - fix the labels of the axes - phi_2 vs. phi_1.
-
-TODO: just report the results of figure 3 + replace figure 3 with two scatter plots (for example, phi1 vs. p with and without EMD).
-To emphasize the importance of using the correct distance metric, we also computed the diffusion maps embeddings of our two sets of simulation data, used the standard Euclidean distance between the histograms to compute the distances in \eqref{eq:W}; the results are shown in Figure \ref{fig:dmaps_embed_noemd}.
-%
-The embeddings are not particularly informative and do not capture the underlying system dynamics; there is no strong correlation of the embedding coordinates with $p$ or $t$. 
-%
-This is due to the inappropriate metric, which does not accurately describe the distances between histograms.
+%\label{fig:dmaps_embed_noemd}
+%\end{figure}
 

### 2014-06-08 chemotaxis/simulate_jump_process_EMD.m

@@ -93,7 +93,7 @@ ylabel('$\phi_2$', 'interpreter','latex', 'fontsize', 20)
 h = colorbar;
 set(get(h,'xlabel'),'String', 't', 'fontsize', 20);
 %print(sprintf('rawhist_t_%d', lambda), '-r300','-djpeg')
-saveas(gcf, sprintf('rawhist_t_%d', lambda), 'epsc')
+% saveas(gcf, sprintf('rawhist_t_%d', lambda), 'epsc')
 
 figure;
 scatter(V(:,2),V(:,3),markersize,all_p(idx), '.')
@@ -103,7 +103,11 @@ ylabel('$\phi_2$', 'interpreter','latex', 'fontsize', 20)
 h = colorbar;
 set(get(h,'xlabel'),'String', 'p', 'fontsize', 20);
 %print(sprintf('rawhist_p_%d', lambda), '-r300','-djpeg')
-saveas(gcf, sprintf('rawhist_p_%d', lambda), 'epsc')
+% saveas(gcf, sprintf('rawhist_p_%d', lambda), 'epsc')
+
+corr(V(:,2),all_p(idx)')
+
+corr(V(:,3),all_time(idx)')
 
 %% compute EMD between histograms
 
@@ -161,6 +165,12 @@ set(get(h,'xlabel'),'String', 'p');
 %title(sprintf('EMD: colored by p, \\lambda= %d', lambda))
 %print(sprintf('EMD_p_%d', lambda), '-r300','-djpeg')
 
+corr(V2(:,2),all_p(idx)')
+
+corr(V2(:,3),all_time(idx)')
+
+
+return
 %%
 
 num_hist = 6;

### 2014-06-08 chemotaxis/simulate_jump_process_EMD2.m

@@ -94,7 +94,7 @@ ylabel('$\phi_2$', 'interpreter','latex', 'fontsize', 20)
 h = colorbar;
 set(get(h,'xlabel'),'String', 't', 'fontsize', 20);
 %print(sprintf('rawhist_t_%d', lambda), '-r300','-djpeg')
-saveas(gcf, sprintf('rawhist_t_%d', lambda), 'epsc')
+% saveas(gcf, sprintf('rawhist_t_%d', lambda), 'epsc')
 
 figure;
 scatter(V(:,2),V(:,3),markersize,all_p(idx), '.')
@@ -104,7 +104,11 @@ ylabel('$\phi_2$', 'interpreter','latex', 'fontsize', 20)
 h = colorbar;
 set(get(h,'xlabel'),'String', 'p', 'fontsize', 20);
 %print(sprintf('rawhist_p_%d', lambda), '-r300','-djpeg')
-saveas(gcf, sprintf('rawhist_p_%d', lambda), 'epsc')
+% saveas(gcf, sprintf('rawhist_p_%d', lambda), 'epsc')
+
+corr(V(:,2),all_time(idx)')
+
+corr(V(:,3),all_p(idx)')
 
 %% compute EMD between histograms
 
@@ -162,6 +166,11 @@ set(get(h,'xlabel'),'String', 'p');
 %title(sprintf('EMD: colored by p, \\lambda= %d', lambda))
 %print(sprintf('EMD_p_%d', lambda), '-r300','-djpeg')
 
+corr(V2(:,2),all_time(idx)')
+
+corr(V2(:,3),all_p(idx)')
+
+return
 %%
 
 num_hist = 6;

### 2014-06-20 chemotaxis/prl_paper.tex
added ronen edits to chemotaxis

@@
 %
-We will begin by describing the chemotaxis problem, the illustrative example which we will use to demonstrate the methodologies.
+We begin by describing the chemotaxis problem, the illustrative example used to demonstrate the methodologies.
 %
-We will present the stochastic microscopic model, and use diffusion maps, a nonlinear dimensionality reduction technique, to analyze microscopic simulation results.
+We present the stochastic microscopic model, and use diffusion maps, a nonlinear dimensionality reduction technique, to analyze microscopic simulation results.
 %
-We will show that we can recover the governing macroscopic variables, provided that we use the appropriate observers and distance metric.
+We show that the governing macroscopic variables can be recovered, provided that the appropriate observers and distance metric are used.
 %
-We will then describe how, for this example, the microscopic model gives rise to a compact macroscopic description governed by a single parameter, and show that the diffusion maps results are consistent with this macroscopic description.
+We then describe how, for this example, the microscopic model gives rise to a compact macroscopic description governed by a single parameter, and show that the diffusion maps results are consistent with this macroscopic description.
 %
-Finally, we will demonstrate that diffusion maps can also inform us as to the relative importance of system variables in different parameter regimes, and allow us to detect changes in macroscopic dynamical behavior.
+Finally, we demonstrate that diffusion maps can also identify the relative importance of system variables in different parameter regimes, and detect changes in macroscopic dynamical behavior.
 
@@
 %
-In this model, each cell is initialized with a position and a velocity, and the dynamics of each cell are governed by a stochastic process.
+In this model, each cell is initialized with a position and a velocity on a line, and the dynamics of each cell are governed by a stochastic process.
 %
@@
 %
-However, for more complex microscopic models, the governing macroscopic variables may not be immediately obvious.
+This implies that the data, which lie in a high-dimensional ambient space (e.g., the positions of all $N$ particles), for a fixed value of $\lambda$, should lie on a two-dimensional manifold parameterized by $p$ and $t$,
+thereby uncovering the low-dimensional structure reveals the governing macroscopic variables.
+%
+TODO: whether the equations should come here so this statement could be explained based on the analysis?
+%
+However, for more complex microscopic models, the governing macroscopic variables may not be immediately obvious, and they need to be inferred/identified from the data.
 %
 In general, we expect the simulation data to lie on a low-dimensional manifold which is parameterized by the macroscopic variables.
+
+We use diffusion maps \cite{coifman2005geometric}, which is a manifold learning technique, to uncover such low-dimensional structure and to analyze the results of our microscopic chemotaxis simulations.
 %
-For our particular example, this implies that our data should lie on a two-dimensional manifold  parameterized by $p$ and $t$, 
-and uncovering the low-dimensional structure in our data set will inform us as to the governing macroscopic variables.
-
-We propose using data driven techniques to uncover such low-dimensional structure.
+Compared to principal components analysis (PCA) \cite{...}, a standard linear technique, diffusion maps provides a parameterization of data, which lie on a low-dimensional (possibly) {\em nonlinear} structure in a high-dimensional space.
 %
-Principal components analysis (PCA) \cite{...} is a traditional dimensionality reduction algorithm which projects the data onto the optimal low-dimensional linear subspace. 
-%
-Here, we use diffusion maps \cite{coifman2005geometric}, a manifold learning technique, to analyze the results of our microscopic chemotaxis simulations.
-%
-Diffusion maps is more general than PCA, in that it can find a parameterization of data which lie on a low-dimensional, {\em nonlinear} structure in high-dimensional space.
-%
-For our example, although the data lies in a high-dimensional ambient space (e.g., the positions of all $N$ particles), for a fixed value of $\lambda$, we expect the data to lie on a two-dimensional manifold described by $p$ and $t$.
-%
+The choice of using diffusion maps is not unique, and it is one of many recently introduced manifold learning techniques that could be used to analyzed the data.
 
@@
 
-In general, manifold learning techniques require two essential components.
+In general, manifold learning techniques have two essential components.
 %
-One, they require definition of the appropriate features of {\em observers} to describe the system. 
+One is the appropriate {\em observers} of the system. 
 %
-These observers should be informative as to the state of the system, as well as be (somewhat) invariant to noise in the system. 
+These observers should be informative as to the state of the system, as well as invariant to noise in the system. 
 %
-Two, these algorithms require an informative {\em distance metric} between the observers.
+Two is a {\em distance metric} between the observations that captures a notion of locality: observations which we consider to be similar should have a small distance.
+
+Let $z(t) \in \mathbb{R}^n$ denote an $n$-dimensional observation of the system at time $t$. 
 %
-This metric should accurately capture our idea of locality: observations which we consider to be similar should have a small distance. 
-%
-For this particular example, we use histograms of the particle positions as our observers TODO:refPNAS. 
+For the chemotaxis example, we use histograms of the particle positions as observers \cite{talmon2013empirical}. 
 %
@@
 %
-We denote the histogram of $x_1(i), x_2(i), \dots, x_N(i)$ into $n$ bins (for our simulations, we set $n=32$) as $z(i) \in \mathbb{R}^n$.
+We denote the histogram of $x_1(i), x_2(i), \dots, x_N(i)$ into $n$ bins as $z(i) \in \mathbb{R}^n$, where we set $n=32$.
 %
@@
 %
-For the special case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions \cite{rubner2000perceptual}.
-%
-This can be estimated from histograms, 
+For the special case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions \cite{rubner2000perceptual}, which can be estimated from histograms as
 \begin{equation}
-\| z(i) - z(j) \|_{EMD} = \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|
+\| z(i) - z(j) \|_{EMD} \approx \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|
 \end{equation}
-where $z(i), z(j) \in \mathbb{R}^n$ are two histograms defined on (equally-spaced) bins in $\mathbb{R}$.
+where the histograms are defined on equally-spaced bins in $\mathbb{R}$.
 
-Given a set of observations with the appropriate metric, diffusion maps can then uncover a parameterization of these observations. 
-%
-Let $z(t) \in \mathbb{R}^n$ denote an $n$-dimensional observation of the system at time $t$. 
-%
 Assume we are given $m$ observations $z(1), \dots, z(m)$, which are assumed to lie on a $d$-dimensional, nonlinear manifold, with $d \ll n$. 
@@
 \end{equation}
-where $\| \cdot \|$ denotes the appropriate norm for the data, and $\epsilon$ is a characteristic distance between the observations (we take $\epsilon$ to be the median of the pairwise distances between the data points).
+where $\| \cdot \|$ denotes the appropriate norm for the data, and $\epsilon$ is a characteristic distance between the observations. 
+%
+For the chemotaxis example, we use the EMD and we take $\epsilon$ to be the median of the pairwise distances between the data points.
 %
@@
 %
-The eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ then give the embedding coordinates for the data, such that
+The eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ then provide the embedding coordinates for the data, such that
 $\phi_{j}(i)$ is the $j^{th}$ embedding coordinate for $z(i)$.
 %
-The ``importance'' of coordinate $\phi_j$ is quantified by the corresponding eigenvalue $\mu_j$.
+The ``importance'' of coordinate $\phi_j$ is quantified by the corresponding eigenvalue $\mu_j$, e.g.
+the relative importance of $\phi_1$ versus $\phi_2$ can be quantified by comparing $\mu_1$ and $\mu_2$.
 %
 We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$, and if the data are low-dimensional, we only need to retain a few of the eigenvectors to describe the data.
-%
-The relative importance of $\phi_1$ versus $\phi_2$ can be quantified by comparing $\mu_1$ and $\mu_2$.
 %
@@
 
-%TODO: just report the results of figure 3 + replace figure 3 with two scatter plots (for example, phi1 vs. p with and without EMD).
-To emphasize the importance of using the correct distance metric, we also computed the diffusion maps embeddings of our two sets of simulation data, used the standard Euclidean distance between the histograms to compute the distances in \eqref{eq:W}.
+To emphasize the importance of using the correct distance metric, we also computed the diffusion maps embeddings of the two sets of simulation data, using the standard Euclidean distance between the histograms to compute the distances in \eqref{eq:W}.
 %
@@
 
-For this specific example, we can explicitly write the macroscopic equation that governs the overall system behavior.
+This example has a macroscopic equation that governs the overall system behavior.
 %
@@
 
-We can now easily justify the features that we see in the low-dimensional embeddings in Figure~\ref{fig:dmaps_embed_emd}.
+We can now interpret the coordinates in the low-dimensional embeddings in Figure~\ref{fig:dmaps_embed_emd}.
 %
@@
 
-\begin{figure}[t] 
-\includegraphics[width=8cm]{detect_change_eigenvalues}
-\caption{The first two eigenvalues from DMAPS as a function of $\lambda$. Each point is averaged over 10 simulation experiments, with error bars indicating one standard deviation. 
-The convergence of the eigenvalues at $\lambda \approx 125$ indicates switching of the dynamical ``mode'' of the system.}
-\label{fig:detect_change}
-\end{figure}
+%\begin{figure}[t] 
+%\includegraphics[width=8cm]{detect_change_eigenvalues}
+%\caption{The first two eigenvalues from DMAPS as a function of $\lambda$. Each point is averaged over 10 simulation experiments, with error bars indicating one standard deviation. 
+%The convergence of the eigenvalues at $\lambda \approx 125$ indicates switching of the dynamical ``mode'' of the system.}
+%\label{fig:detect_change}
+%\end{figure}
 
@@
 
-We have seen that the eigenvalues from diffusion maps, which quantify the importance of each embedding coordinate, can inform us as to the relative importance of the variables which govern the macroscopic dynamics.
+We have seen that the eigenvalues $\mu_i$, which quantify the importance of each embedding coordinate, represent the relative importance of the variables which govern the macroscopic dynamics.
 %
@@
 %
-We have seen that at the two asymptotic regimes ($\lambda \rightarrow 0$ and $\lambda \rightarrow \infty$), there will be a gap between the two eigenvalues, as one of the macroscopic variables is significantly more important to the dynamical behavior.
+We showed that at the two asymptotic regimes ($\lambda \rightarrow 0$ and $\lambda \rightarrow \infty$), there is a gap between the two eigenvalues, as one of the macroscopic variables is significantly more important to the dynamical behavior.
 %
-For intermediate $\lambda$ values, we expect the gap between $|\mu_1|$ and $|\mu_2|$ to narrow, and that $|\mu_1| \approx |\mu_2|$ corresponds to the transition between ``wave-like'' and ``heat-like'' behavior.
+For intermediate $\lambda$ values, we expect the gap between $|\mu_1|$ and $|\mu_2|$ to narrow, and that the $\lambda$ value where $|\mu_1| \approx |\mu_2|$ corresponds to the transition between ``wave-like'' and ``heat-like'' behavior.
 
@@
 
-We have demonstrated how diffusion maps can automatically uncover the variables which govern the macroscopic behavior of a system from data sampled at the microscopic level.
+%We have demonstrated how diffusion maps can automatically uncover the variables which govern the macroscopic behavior of a system from data sampled at the microscopic level.
+%%
+%Furthermore, the eigenvalue spectrum allows us to detect changes in the macroscopic system dynamics.
+%%
+%The essential components for our algorithms are choosing the correct observers and selecting the appropriate metric between these observers.
+%%
+%In our specific example, we used histograms of particle positions and the earth mover's distance between pairs of histograms.
+%%
+%With these two considerations, we used diffusion maps to uncover the two variables which characterize the system dynamics.
+%%
+%The eigenvalue spectrum can help us detect changes in the dynamical behavior.
+%%
+%Furthermore, these two variables have a completely different nature;
+%one variable, $p$, characterizes the initial conditions of the system, whereas the other variable, $t$, describes the ``step-by-step'' evolution of the particles. 
+%%
+%This promotes the broad applicability of our methods to a wide variety of dynamical systems.
+
+This paper aims to bridge the fields of data mining and dynamical systems.
+For data mining methods to be informative, processing the appropriate ``statistical moments" or the ``right" observables instead of individual particles/realizations is essential.
+In addition, one must also define the appropriate distance metric between the observations.
+These two components induce the ``right" Riemannian geometry and allow for informative analysis, such as meaningful and compact parameterization, phase shift detection, and dynamical characterization, through manifold learning.
+
+In this paper, these concepts were demonstrated using a dynamical model of cellular chemotaxis with analytical macroscopic behavior.
 %
-Furthermore, the eigenvalue spectrum allows us to detect changes in the macroscopic system dynamics.
-%
-The essential components for our algorithms are choosing the correct observers and selecting the appropriate metric between these observers.
-%
-In our specific example, we used histograms of particle positions and the earth mover's distance between pairs of histograms.
-%
-With these two considerations, we used diffusion maps to uncover the two variables which characterize the system dynamics.
-%
-The eigenvalue spectrum can help us detect changes in the dynamical behavior.
-%
-Furthermore, these two variables have a completely different nature;
-one variable, $p$, characterizes the initial conditions of the system, whereas the other variable, $t$, describes the ``step-by-step'' evolution of the particles. 
-%
-This promotes the broad applicability of our methods to a wide variety of dynamical systems.
+We showed that, using histograms as observers and EMD as a distance metric between them, diffusion maps can uncover the ``right" macroscopic dynamical behavior, which is consistent with the analytical macroscopic model.
 

### 2014-06-20 chemotaxis/prl_paper.tex
made prl edits

@@
 %
-We also show how these algorithms can detect changes in the macroscopic dynamical behavior as a function of microscopic system parameters. 
+We also show how these algorithms can detect changes in the macroscopic dynamical behavior as a function of the microscopic system parameters. 
 %
@@
 
-Often, systems which are stochastic and noisy on the microscale exhibit smooth, coherent dynamics on the macroscale which are governed by only a few parameters.
+Often, systems which are stochastic and noisy on the microscale exhibit coherent dynamics on the macroscale which are governed by only a few parameters.
 %
@@
 %
-We would like to {\em automatically} uncover the macroscale behavior from data collected at the microscale without any {\em a priori} knowledge of the appropriate microscopic or macroscopic model.
+We would like to {\em automatically} uncover a system's macroscale behavior from data collected at the microscale without any {\em a priori} knowledge of the appropriate microscopic or macroscopic model.
 %
-TODO: write where this has been done in other places, context, etc.
+%TODO: write where this has been done in other places, context, etc.
 
@@
 %
-First, this system has random microscopic behavior with smooth macroscopic dynamics governed by few parameters, which determine the regime/mode of the system. 
+First, this system has random microscopic behavior with smooth macroscopic dynamics governed by a single parameter which determines the regime/mode of the system. 
 %
-Second, it consists of an ensemble of individual particles.
+Second, it consists of an ensemble of individual stochastic particles which must be described using the appropriate observers.
 %
@@
 %
-We will show that manifold learning algorithms based on statistical observers and with the appropriate affinity metric between these observers uncovers a description of the microscopic data which is consistent with the macroscopic model.
+We will show that manifold learning algorithms based on statistical observers and with the appropriate affinity metric between pairs of observations uncovers a description of the microscopic data which is consistent with the macroscopic model.
 %
-Furthermore, these data-driven methods also allow us to detect changes in dynamical behavior arising from the variation of the system parameters.
+Furthermore, these data-driven methods allow us to detect changes in dynamical behavior resulting from changes in the system parameters.
 
@@
 
-Biological cells are one common example of a system where microscopic entities must work together to exhibit coherent macroscopic dynamics, to accomplish tasks such as finding food or navigating away from toxins.
+Biological cells are one common example of a system where microscopic entities work together and exhibit coherent macroscopic dynamics, in order to to accomplish tasks such as finding food or navigating away from toxins.
 %
@@
 %
-We will analyze one such model described by a velocity jump process \cite{othmer2000diffusion}.
+We will analyze one such model described by a one-dimensional velocity jump process \cite{othmer2000diffusion}.
 %
@@
 %
-%Note that each particle has its own ``clock" (Poisson process).
 
-%Let $X_{p, \lambda, s}(t)$ denote the vector of positions of the particles at time $t$ with initial right probability $p$, switching rate $\lambda$, and speed $s$.
+This particular example has an analytic macroscopic equation that governs the overall system behavior.
+%
+For a large collection of particles ($N \rightarrow \infty$), the system can be described by the probability density of the particles.
+%
+Let $\rho(x, t)$ denote the probability density of the particles, and let $\rho^-(x, t)$ and $\rho^+(x, t)$ denote the densities of the particles moving towards the positive and negative axis directions, respectively.
+%
+It can be shown that, as $N \rightarrow \infty$, the densities obey the following set of partial differential equations (PDEs) \cite{othmer2000diffusion}:
+\begin{equation} \label{eqn:coupled_pdes}
+\begin{aligned}
+\frac{\partial \rho^+}{\partial t} + s \frac{\partial \rho^+}{\partial x} & = -\lambda \rho^+ +\lambda \rho^- \\
+\frac{\partial \rho^-}{\partial t} - s \frac{\partial \rho^-}{\partial x} & = \lambda \rho^+ -\lambda \rho^- 
+\end{aligned}
+\end{equation}
+%
+Alternatively, \eqref{eqn:coupled_pdes} can be rewritten as one, second--order PDE:
+\begin{equation} \label{eq:second_order_pde}
+\frac{\partial^2 \rho}{\partial t^2} + 2 \lambda \frac{\partial \rho}{\partial t} = s^2 \frac{\partial ^2 \rho}{\partial x^2}
+\end{equation}
+%
+%We assume that $s^2/\lambda = D$ is constant, so that the dynamics of the probability density are governed by a {\em single} parameter $\lambda$.
+%
+For fixed values of $\lambda$ and $s$, the macroscopic state of the system (the probability density of the particles) is a function of two independent variables: $p$, which controls the initial distribution of the particles, and $t$, the time.
+%
+This implies that for fixed $\lambda$ and $s$, the {\em microscopic} data which lie in a high-dimensional ambient space (e.g., the positions of all $N$ particles) should lie on a two-dimensional manifold parameterized by $p$ and $t$,
+so that uncovering the low-dimensional structure reveals the governing macroscopic variables.
 
-For fixed values of $\lambda$ and $s$, the macroscopic state of the system can be described by two independent variables: $p$, which controls the initial distribution of the particles, and $t$, the time.
+We consider two asymptotic regimes of simulation.
 %
-This implies that the data, which lie in a high-dimensional ambient space (e.g., the positions of all $N$ particles), for a fixed value of $\lambda$, should lie on a two-dimensional manifold parameterized by $p$ and $t$,
-thereby uncovering the low-dimensional structure reveals the governing macroscopic variables.
+When $\lambda \rightarrow 0$, the right-hand side of \eqref{eqn:coupled_pdes} tends to 0, and \eqref{eqn:coupled_pdes} becomes two uncoupled wave equations,
+\begin{equation}
+\begin{aligned}
+\frac{\partial \rho^+}{\partial t} + s \frac{\partial \rho^+}{\partial x} & = 0 \\
+\frac{\partial \rho^-}{\partial t} - s \frac{\partial \rho^-}{\partial x} & = 0.
+\end{aligned}
+\end{equation}
+%alternatively, \eqref{eq:second_order_pde} becomes the second order wave equation,
+%\begin{equation}
+%\frac{\partial^2 \rho}{\partial t^2} = s^2 \frac{\partial ^2 \rho}{\partial x^2}.
+%\end{equation}
+
+%Dividing \eqref{eq:second_order_pde} by $\lambda > 0$ yields
+%\[
+%\frac{1}{\lambda} \frac{\partial^2 \rho}{\partial t^2} + 2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2}
+%\]
+When $\lambda \rightarrow \infty$, \eqref{eq:second_order_pde} approaches the heat equation,
+\begin{equation}
+2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2}.
+\end{equation}
 %
-TODO: whether the equations should come here so this statement could be explained based on the analysis?
+%Two variables determine the macroscopic state of the system: the initial distribution, which is controlled by a single parameter $p$, and the time, $t$.
 %
-However, for more complex microscopic models, the governing macroscopic variables may not be immediately obvious, and they need to be inferred/identified from the data.
+The above analysis shows that the initial distribution of velocities of the particles (determined by $p$ in the microscopic simulations) plays a very different role depending on the value of $\lambda$.
 %
-In general, we expect the simulation data to lie on a low-dimensional manifold which is parameterized by the macroscopic variables.
+When $\lambda \rightarrow 0$, the dynamics are described by two wave equations, and the initial distribution persists throughout the trajectory.
+%
+When $\lambda \rightarrow \infty$, the dynamics are described by one heat equation, and the initial conditions are insignificant -- the velocity distribution quickly equilibrates and we see purely diffusive behavior.
 
+For general dynamical systems, the governing macroscopic variables and asymptotic behavior may not be immediately obvious, and instead need to be inferred/identified from the data.
+%
+We expect the simulation data to lie on a low-dimensional manifold which is parameterized by the macroscopic variables.
+%
 We use diffusion maps \cite{coifman2005geometric}, which is a manifold learning technique, to uncover such low-dimensional structure and to analyze the results of our microscopic chemotaxis simulations.
 %
-Compared to principal components analysis (PCA) \cite{...}, a standard linear technique, diffusion maps provides a parameterization of data, which lie on a low-dimensional (possibly) {\em nonlinear} structure in a high-dimensional space.
+Compared to principal components analysis (PCA) \cite{...}, a standard linear technique, diffusion maps provides a parameterization of data which lie on a low-dimensional, (possibly) {\em nonlinear} structure in a high-dimensional space.
 %
-The choice of using diffusion maps is not unique, and it is one of many recently introduced manifold learning techniques that could be used to analyzed the data.
+Diffusion maps is simply one of many recently introduced manifold learning techniques that could be used to analyzed the data \cite{...}.
 
@@
 %
+The observations $z(i) \in \mathbb{R}^n$ are the histogram of $x_1(i), x_2(i), \dots, x_N(i)$ into $n =32$ bins.
+%
 The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions.
@@
 %
-Two is a {\em distance metric} between the observations that captures a notion of locality: observations which we consider to be similar should have a small distance.
+Two is a {\em distance metric} between the observations that captures a notion of locality: observations which we percieve to be similar should have a small distance.
 
@@
 Histograms are invariant to the indexing of the particles, while retaining the spatial locations of the particles.
-%
-We denote the histogram of $x_1(i), x_2(i), \dots, x_N(i)$ into $n$ bins as $z(i) \in \mathbb{R}^n$, where we set $n=32$.
 %
@@
 \end{equation}
-where the histograms are defined on equally-spaced bins in $\mathbb{R}$.
+where the histograms $z(i), z(j)$ are defined on equally-spaced bins in $\mathbb{R}$.
 
-Assume we are given $m$ observations $z(1), \dots, z(m)$, which are assumed to lie on a $d$-dimensional, nonlinear manifold, with $d \ll n$. 
+Assume we are given $m$ observations $z(1), \dots, z(m)$, which lie on a $d$-dimensional, nonlinear manifold, with $d \ll n$. 
 %
@@
 %
-For the chemotaxis example, we use the EMD and we take $\epsilon$ to be the median of the pairwise distances between the data points.
+For the chemotaxis example, we use the EMD as the norm, and we take $\epsilon$ to be the median of the pairwise distances between the data points.
 %
@@
 The ``importance'' of coordinate $\phi_j$ is quantified by the corresponding eigenvalue $\mu_j$, e.g.
-the relative importance of $\phi_1$ versus $\phi_2$ can be quantified by comparing $\mu_1$ and $\mu_2$.
+the relative importance of $\phi_1$ versus $\phi_2$ can be measured by comparing $\mu_1$ and $\mu_2$.
 %
-We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$, and if the data are low-dimensional, we only need to retain a few of the eigenvectors to describe the data.
+We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$, and if the data are low-dimensional, we only need to retain a few of the leading eigenvectors to adequately describe the data.
 %
-Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$), which implies $\mu_0 = 1$ and $\phi_0$ is a constant vector which is never used as an embedding coordinate.
+Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a constant vector which is not used as an embedding coordinate.
 
@@
 %
-One set of simulations is for a small value of $\lambda$, and the other set is for the case of large $\lambda$. 
+One set of simulations (Figures~\ref{subfig:small_lambda_p}~and~\ref{subfig:small_lambda_t}) is for a small value of $\lambda$, and the other set (Figures~\ref{subfig:large_lambda_p}~and~\ref{subfig:large_lambda_t}) is for a large value of $\lambda$.
 %
-In both cases, the two macroscopic variables, $p$ and $t$, are well-correlated with the first two diffusion maps coordinates, $\phi_1$ and $\phi_2$. 
+For both sets of simulations, the two macroscopic variables, $p$ and $t$, are well-correlated with the first two diffusion maps coordinates, $\phi_1$ and $\phi_2$. 
 %
-In both cases, $\phi_1$ is significantly more important than $\phi_2$, seen by the separation between the first two eigenvalues (for the small $\lambda$ case, $\mu_1 = 0.5138$, and $\mu_2 = 0.3943$, and for the large $\lambda$ case, $\mu_1 = 0.5543$, and $\mu_2 = 0.3807$).
+In both cases, $\phi_1$ is significantly more important than $\phi_2$, indicated by the separation between the first two eigenvalues (for the small $\lambda$ case, $\mu_1 = 0.5138$, and $\mu_2 = 0.3943$, and for the large $\lambda$ case, $\mu_1 = 0.5543$, and $\mu_2 = 0.3807$).
 %
-$\phi_1$, the ``more important'' coordinate, is correlated with $p$ for the small $\lambda$ case (Figure~\ref{subfig:small_lambda_p}), and correlated with $t$ for the large $\lambda$ case (Figure~\ref{subfig:large_lambda_t}), indicating that the relative importance of $p$ and $t$ changes in the two simulations.
+$\phi_1$, the ``more important'' coordinate, is correlated with $p$ for the small $\lambda$ case (Figure~\ref{subfig:small_lambda_p}), and correlated with $t$ for the large $\lambda$ case (Figure~\ref{subfig:large_lambda_t}), indicating that the relative importance of $p$ and $t$ changes in the two simulations;
+this is consistent with the analytic macroscopic description. 
+%
+Furthermore, in the small $\lambda$ regime (wave equation), shown in Figures \ref{subfig:small_lambda_p} and \ref{subfig:small_lambda_t}, the points corresponding to small times are more tightly clustered than the points corresponding to large times.
+%
+This is consistent with the macroscopic model: at small times, the particles are more condensed around $x=0$, and it is more difficult to distinguish the particles moving to the left from the particles moving to the right.
+%
+On the other hand, at large times, once the particles evolve from the origin, this separation is clear.  
+%
+For the large $\lambda$ case (heat equation), shown in Figures \ref{subfig:large_lambda_p} and \ref{subfig:large_lambda_t}, we observe that at small times, the initial distribution $p$ is well organized in the embedding in Figure \ref{subfig:large_lambda_p}, as the velocity distribution of the particles is initially skewed and can be seen in the initial displacements
+%
+On the other hand, for large times, we observe that the initial distribution $p$ is less organized in Figure \ref{subfig:large_lambda_p}, as the velocities have equilibrated and the initial distribution is less detectable in the particle density.
 
@@
 %
-There is no appreciable correlation between the embedding coordinates and the macroscopic variables $p$ and $t$. 
-%
-The correlations between the embedding coordinates and the macroscopic variables are all $< 0.6$.
+There is no appreciable correlation between the embedding coordinates and the macroscopic variables $p$ and $t$ (the correlations between the embedding coordinates and the macroscopic variables are all  less than 60\%, with the correlations for the small $\lambda$ case being less than 20\%).
 %
@@
 %
-In contrast, the correlations between the embedding coordinates and the macroscopic variables when using EMD in the diffusion maps calculation are all $> 0.9$.
+In contrast, the correlations between the embedding coordinates and the macroscopic variables when using EMD in the diffusion maps calculation are all greater than 90\%.
 
-%We simulate the dynamics of the studies stochastic system \eqref{eqn:system} to obtain microscopic data to which we apply diffusion maps based on histograms and the EMD.
-%
-%We consider two sets of simulations: TODO: associate the sets to the asymptotic regimes.
-%in one set (small $\lambda$), $\lambda = 1$ and $s=1$, and in the other set (large $\lambda$), $\lambda = 400$, and $s=20$.
-%
-%For each set, we run many simulations, each with $N=1000$ particles.
-%
-%We vary the initial conditions such that $0.1 \le p  \le 0.9$, and allow each simulation to evolve for $10$ time units.
-%
-%We discard the initial portion of each simulation trajectory corresponding to $t < 1$, as it tends to be very noisy and corresponds to initial relaxation of the system.
+\begin{figure}[t] 
+\includegraphics[width=8cm]{detect_change_eigenvalues}
+\caption{The first two eigenvalues from DMAPS as a function of $\lambda$.
+Each point is averaged over 10 simulation experiments (as in Figure~\ref{fig:dmaps_embed_emd}, one experiment consists of data collected at varying $p$ and $t$, for a fixed value of $\lambda$), with error bars indicating one standard deviation.
+The convergence of the eigenvalues at $\lambda \approx 125$ indicates switching of the dynamical ``mode'' of the system.}
+\label{fig:detect_change}
+\end{figure}
 
-%Figure \ref{fig:dmaps_embed_emd} shows the diffusion maps embedding of the data.
-%
-%The two system variables, $p$ and $t$, are well-correlated with the first two diffusion maps coordinates, $\phi_1$ and $\phi_2$. 
-%
-%TODO: give the values of the corresponding eigenvalues.
-%However, the ``more important'' coordinate, $\phi_1$, is correlated with $p$ for the small $\lambda$ case (Figure~\ref{subfig:small_lambda_p}), and correlated with $t$ for the large $\lambda$ case (Figure~\ref{subfig:large_lambda_t}).
-
-This example has a macroscopic equation that governs the overall system behavior.
-%
-For a large collection of particles ($N \rightarrow \infty$), the system can be described by the probability density of the particles.
-%
-Let $\rho(x, t)$ denote the probability density of the particles, and let $\rho^-(x, t)$ and $\rho^+(x, t)$ denote the densities of the particles moving towards the positive and negative axis directions, respectively.
-%
-It can be shown that, as $N \rightarrow \infty$, the densities obey the following set of partial differential equations (PDEs) \cite{othmer2000diffusion}:
-\begin{equation} \label{eqn:coupled_pdes}
-\begin{aligned}
-\frac{\partial \rho^+}{\partial t} + s \frac{\partial \rho^+}{\partial x} & = -\lambda \rho^+ +\lambda \rho^- \\
-\frac{\partial \rho^-}{\partial t} - s \frac{\partial \rho^-}{\partial x} & = \lambda \rho^+ -\lambda \rho^- 
-\end{aligned}
-\end{equation}
-%
-Alternatively, \eqref{eqn:coupled_pdes} can be rewritten as one, second--order PDE:
-\begin{equation} \label{eq:second_order_pde}
-\frac{\partial^2 \rho}{\partial t^2} + 2 \lambda \frac{\partial \rho}{\partial t} = s^2 \frac{\partial ^2 \rho}{\partial x^2}
-\end{equation}
-%
-We assume that $s^2/\lambda = D$ is constant, so that the dynamics of the probability density are governed by a {\em single} parameter $\lambda$.
-
-%\subsection{Asymptotic mode analysis}  \label{subsec:mode_analysis}
-
-We consider two asymptotic regimes of simulation.
-%
-When $\lambda \rightarrow 0$, the right-hand side of \eqref{eqn:coupled_pdes} tends to 0, and \eqref{eqn:coupled_pdes} becomes two uncoupled wave equations,
-\begin{equation}
-\begin{aligned}
-\frac{\partial \rho^+}{\partial t} + s \frac{\partial \rho^+}{\partial x} & = 0 \\
-\frac{\partial \rho^-}{\partial t} - s \frac{\partial \rho^-}{\partial x} & = 0.
-\end{aligned}
-\end{equation}
-%alternatively, \eqref{eq:second_order_pde} becomes the second order wave equation,
-%\begin{equation}
-%\frac{\partial^2 \rho}{\partial t^2} = s^2 \frac{\partial ^2 \rho}{\partial x^2}.
-%\end{equation}
-
-%Dividing \eqref{eq:second_order_pde} by $\lambda > 0$ yields
-%\[
-%\frac{1}{\lambda} \frac{\partial^2 \rho}{\partial t^2} + 2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2}
-%\]
-When $\lambda \rightarrow \infty$, \eqref{eq:second_order_pde} approaches the heat equation,
-\begin{equation}
-2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2}.
-\end{equation}
-%
-%Two variables determine the macroscopic state of the system: the initial distribution, which is controlled by a single parameter $p$, and the time, $t$.
-%
-The above analysis shows that the initial distribution of the moving direction of the particles (determined by $p$ in the microscopic simulations) plays a very different role depending on the value of $\lambda$.
-%
-When $\lambda \rightarrow 0$, the dynamics are described by two wave equations, and the initial distribution persists throughout the trajectory.
-%
-When $\lambda \rightarrow \infty$, the dynamics are described by one heat equation, and the initial conditions are insignificant -- the velocity distribution quickly equilibrates and we see purely diffusive behavior.
-
-We can now interpret the coordinates in the low-dimensional embeddings in Figure~\ref{fig:dmaps_embed_emd}.
-%
-Firstly, the observation that in the small $\lambda$ case, $p$, which governs the initial condition, is correlated with $\phi_1$ and therefore more important than $t$, which is correlated with $\phi_2$, is consistent with the analytic macroscopic description.
-%
-Furthermore, in the small $\lambda$ regime (wave equation), shown in Figures \ref{subfig:small_lambda_p} and \ref{subfig:small_lambda_t}, the points corresponding to small times are more tightly clustered than the points corresponding to large times.
-%
-This is consistent with the macroscopic model: at small times, the particles are more condensed around $x=0$, and it is more difficult to distinguish the particles moving to the left from the particles moving to the right.
-%
-On the other hand, at large times, once the particles evolve from the origin, this separation is clear.  
-%
-For the large $\lambda$ case (heat equation), shown in Figures \ref{subfig:large_lambda_p} and \ref{subfig:large_lambda_t}, we observe that for small time the initial distribution $p$ is well organized in the embedding in Figure \ref{subfig:large_lambda_p}, the distribution of the particles is initially skewed and the initial velocity plays a role.
-%
-On the other hand, for large times, we observe that the initial distribution $p$ is less organized in Figure \ref{subfig:large_lambda_p}, as the velocities have equilibrated and the initial distribution is less detectable in the particle density.
-
-%We simulate this stochastic velocity jump process for $N=1000$ particles, and $p \in [0.1, 0.9]$.
-%%
-%We simulate the process for $t \in [0, 10]$, and record the positions of each particle at uniform/equal time intervals.
-%%
-%We discard the initial portion of the trajectory corresponding to $t < 1$, as this is very noisy and corresponds to initial relaxation of the system.
-%%
-%We will consider two sets of simulations.
-%%
-%In one set, $\lambda = 1$ and $s=1$, and in the other set, $\lambda = 400$, and $s=20$.
-
-%In Figure \ref{fig:dmaps_embed_noemd}, we have used the standard Euclidean distance between the histograms to compute the distances for \eqref{eq:W}.
-%%
-%The embeddings obtained are not particularly informative, and do not accurrately capture the underlying system dynamics.
-%%
-%This is due to our poor choice of metric, which does not accurrately describe the distances between histograms.
-
-%\begin{figure}[t] 
-%\includegraphics[width=8cm]{detect_change_eigenvalues}
-%\caption{The first two eigenvalues from DMAPS as a function of $\lambda$. Each point is averaged over 10 simulation experiments, with error bars indicating one standard deviation. 
-%The convergence of the eigenvalues at $\lambda \approx 125$ indicates switching of the dynamical ``mode'' of the system.}
-%\label{fig:detect_change}
-%\end{figure}
-
-%\begin{figure}[t]
-%\begin{subfigure}{0.45\columnwidth}
-%\includegraphics[width=\textwidth]{EMD2_withhist_p_10}
-%\caption{}
-%\end{subfigure}
-%\begin{subfigure}{0.45\columnwidth}
-%\includegraphics[width=\textwidth]{EMD2_withhist_t_10}
-%\caption{}
-%\end{subfigure}
-%\begin{subfigure}{0.45\columnwidth}
-%\includegraphics[width=\textwidth]{EMD2_withhist_p_20}
-%\caption{}
-%\end{subfigure}
-%\begin{subfigure}{0.45\columnwidth}
-%\includegraphics[width=\textwidth]{EMD2_withhist_t_20}
-%\caption{}
-%\end{subfigure}
-%\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with $\lambda=100$, $s=10$.
-The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions.
-The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time.
-The particles are allowed to evolve for (a, b) 10 time units, and (c, d) 20 time units.
-Representative histograms are shown for selected data points.}
-%\label{fig:dmaps_embed_varyt}
-%\end{figure}
-
-We have seen that the eigenvalues $\mu_i$, which quantify the importance of each embedding coordinate, represent the relative importance of the variables which govern the macroscopic dynamics.
+We have seen that the eigenvalues $\mu_i$, which quantify the importance of each embedding coordinate, reveal the relative importance of the variables which govern the macroscopic dynamics.
 %
@@
 
-Figure~\ref{fig:detect_change} shows the first two eigenvalues, $\mu_1$ and $\mu_2$, computed from simulation data, as a function of $\lambda$.
+To demonstrate this point, we performed a new set of experiments, where we simulated the dynamical system over a range of $\lambda$ values. 
+% 
+Figure~\ref{fig:detect_change} shows the first two diffusion maps eigenvalues, $\mu_1$ and $\mu_2$, computed from these simulations, as a function of $\lambda$.
 % 
@@
 %
-For small $\lambda$ (the wave-equation regime), the two eigenvalues are well-separated, as the first mode (which parameterizes $p$) is significantly more important than the second mode (which parameterizes $t$).
+For small $\lambda$ (the wave-equation regime, corresponding to Figures~\ref{subfig:small_lambda_p}~and~\ref{subfig:small_lambda_t}), the two eigenvalues are well-separated, as the first mode (which parameterizes $p$) is significantly more important than the second mode (which parameterizes $t$).
 %
@@
 %
-The eigenvalues then separate again at large $\lambda$ (the heat equation regime), where the $t$ effects dominate the dynamics.
+The eigenvalues then separate again at large $\lambda$ (the heat equation regime, corresponding to Figures~\ref{subfig:large_lambda_p}~and~\ref{subfig:large_lambda_t}), where the $t$ effects dominate the dynamics.
 %
-We can therefore detect where the change in dynamical behavior occurs using data-driven techniques.
-
-%We have illustrated that diffusion maps can elucidate the relative importance of different system variables as a function of $\lambda$.
-%%
-%For the plots shown in Figure~\ref{fig:dmaps_embed_emd}, we can easily see the crossover from the wave-equation regime, where $p$ is more important, to the heat-equation regime, where $t$ is more important.
-%%
-%However, we would like to note that this crossover point depends not only on $\lambda$, but also on the timescale of observation.
-%%
-%We can see this from our PDE description in \eqref{eq:second_order_pde};
-%clearly, $\lambda$ is defined relative to $t$, and so we should not only consider $\lambda$ in our analysis, but also the timescale of observation.
-%%
-%We can also detect this using diffusion maps to analyze our microscopic simulations.
-%%
-%We chose an intermediate value of $\lambda$ ($\lambda=100$ and $s=10$), and allowed the system to evolve for 10 and 20 time units. 
-%%
-%The results are shown in Figure \ref{fig:dmaps_embed_varyt}.
-%%
-%When the system only evolves for 10 time units, the majority of the data still appears ``wave-like'', and so the dominant parameter, $\phi_1$, is correlated with $p$, the initial condition.
-%%
-%However, when the system evolves for 20 time units, the initial velocity distribution has had more time to equilibrate - the behavior is more diffusive and $\phi_1$ is now correlated with $t$. 
-%%
-%Therefore, we can also see the crossover from wave-like behavior to heat-like behavior as a function of the observation timescale using diffusion maps.
+We can therefore detect where the change in dynamical behavior occurs using data-driven techniques, without any knowledge of the governing macroscopic model.
 
@@
 
-%We have demonstrated how diffusion maps can automatically uncover the variables which govern the macroscopic behavior of a system from data sampled at the microscopic level.
-%%
-%Furthermore, the eigenvalue spectrum allows us to detect changes in the macroscopic system dynamics.
-%%
-%The essential components for our algorithms are choosing the correct observers and selecting the appropriate metric between these observers.
-%%
-%In our specific example, we used histograms of particle positions and the earth mover's distance between pairs of histograms.
-%%
-%With these two considerations, we used diffusion maps to uncover the two variables which characterize the system dynamics.
-%%
-%The eigenvalue spectrum can help us detect changes in the dynamical behavior.
-%%
-%Furthermore, these two variables have a completely different nature;
-%one variable, $p$, characterizes the initial conditions of the system, whereas the other variable, $t$, describes the ``step-by-step'' evolution of the particles. 
-%%
-%This promotes the broad applicability of our methods to a wide variety of dynamical systems.
-
 This paper aims to bridge the fields of data mining and dynamical systems. 
-For data mining methods to be informative, processing the appropriate ``statistical moments" or the ``right" observables instead of individual particles/realizations is essential.
+%
+For data mining methods to be informative, processing the appropriate statistical moments or observables instead of individual particles/realizations is essential. 
+%
 In addition, one must also define the appropriate distance metric between the observations.
-These two components induce the ``right" Riemannian geometry and allow for informative analysis, such as meaningful and compact parameterization, phase shift detection, and dynamical characterization, through manifold learning.
-
-In this paper, these concepts were demonstrated using a dynamical model of cellular chemotaxis with analytical macroscopic behavior.
 %
-We showed that, using histograms as observers and EMD as a distance metric between them, diffusion maps can uncover the ``right" macroscopic dynamical behavior, which is consistent with the analytical macroscopic model.
+These two components induce the ``correct" Riemannian geometry and allow for informative analyses, such as meaningful and compact parameterization, phase shift detection, and dynamical characterization, through manifold learning.
+%
+In this paper, these concepts were demonstrated using a dynamical model of cellular chemotaxis with a known analytic macroscopic behavior.
+%
+We showed that, using histograms as observers and EMD as a distance metric, diffusion maps can uncover the underlying macroscopic dynamical behavior which is consistent with the analytic macroscopic model.
+%
+We are confident that such techniques can help inform modeling efforts for systems where the macroscopic dynamics and behavior are not currently known.
 

### 2014-06-23 chemotaxis/prl_paper.tex
Added Ronen small edits, and one last pass; now send to Yannis

@@
-\documentclass[prl, reprint, final, showkeys]{revtex4-1}
+\documentclass[prl, reprint]{revtex4-1}
 
@@
 
-Many dynamical systems are governed by stochastic, noisy motions at the microscale which give rise to smooth, coherent dynamics at the macroscale.
+Many dynamical systems are governed by stochastic motions at the microscale which give rise to smooth, coherent dynamics at the macroscale.
 %
@@
 We would like to {\em automatically} uncover a system's macroscale behavior from data collected at the microscale without any {\em a priori} knowledge of the appropriate microscopic or macroscopic model.
-%
-%TODO: write where this has been done in other places, context, etc.
 
@@
 
-The remainder of this letter is organized as follows. 
+%The remainder of this letter is organized as follows. 
 %
-We begin by describing the chemotaxis problem, the illustrative example used to demonstrate the methodologies.
-%
-We present the stochastic microscopic model, and use diffusion maps, a nonlinear dimensionality reduction technique, to analyze microscopic simulation results.
-%
-We show that the governing macroscopic variables can be recovered, provided that the appropriate observers and distance metric are used.
-%
-We then describe how, for this example, the microscopic model gives rise to a compact macroscopic description governed by a single parameter, and show that the diffusion maps results are consistent with this macroscopic description.
-%
-Finally, we demonstrate that diffusion maps can also identify the relative importance of system variables in different parameter regimes, and detect changes in macroscopic dynamical behavior.
+%We begin by describing the chemotaxis problem, the illustrative example used to demonstrate the methodologies.
+%%
+%We present the stochastic microscopic model, and use diffusion maps, a nonlinear dimensionality reduction technique, to analyze microscopic simulation results.
+%%
+%We show that the governing macroscopic variables can be recovered, provided that the appropriate observers and distance metric are used.
+%%
+%We then describe how, for this example, the microscopic model gives rise to a compact macroscopic description governed by a single parameter, and show that the diffusion maps results are consistent with this macroscopic description.
+%%
+%Finally, we demonstrate that diffusion maps can also identify the relative importance of system variables in different parameter regimes, and detect changes in macroscopic dynamical behavior.
 
@@
 
-Biological cells are one common example of a system where microscopic entities work together and exhibit coherent macroscopic dynamics, in order to to accomplish tasks such as finding food or navigating away from toxins.
+Biological cells are one common example of a system where microscopic entities work together and exhibit coherent macroscopic dynamics, in order to accomplish tasks such as finding food or navigating away from toxins.
 %
@@
 
-This particular example has an analytic macroscopic equation that governs the overall system behavior.
+This particular example has a known analytic macroscopic equation that governs the overall system behavior.
 %
@@
 %
-For fixed values of $\lambda$ and $s$, the macroscopic state of the system (the probability density of the particles) is a function of two independent variables: $p$, which controls the initial distribution of the particles, and $t$, the time.
+From \eqref{eq:second_order_pde}, we see that, for fixed values of $\lambda$ and $s$, the macroscopic state of the system (the probability density of the particles) is a function of two independent variables: $p$, which controls the initial distribution of the particles, and $t$, the time.
 %
-This implies that for fixed $\lambda$ and $s$, the {\em microscopic} data which lie in a high-dimensional ambient space (e.g., the positions of all $N$ particles) should lie on a two-dimensional manifold parameterized by $p$ and $t$,
-so that uncovering the low-dimensional structure reveals the governing macroscopic variables.
+This implies that for fixed $\lambda$ and $s$, the {\em microscopic} data in a high-dimensional ambient space (e.g., the positions of all $N$ particles) should lie on a two-dimensional manifold parameterized by $p$ and $t$,
+and uncovering the low-dimensional structure of the microscopic data reveals the governing macroscopic variables.
 
@@
 
-For general dynamical systems, the governing macroscopic variables and asymptotic behavior may not be immediately obvious, and instead need to be inferred/identified from the data.
+For general dynamical systems, the governing macroscopic variables and asymptotic behavior may not be immediately obvious, and instead need to be inferred from the microscopic data.
 %
-We expect the simulation data to lie on a low-dimensional manifold which is parameterized by the macroscopic variables.
+However, similarly to the chemotaxis model, we expect the simulation data to lie on a low-dimensional manifold which is parameterized by the macroscopic variables.
 %
-We use diffusion maps \cite{coifman2005geometric}, which is a manifold learning technique, to uncover such low-dimensional structure and to analyze the results of our microscopic chemotaxis simulations.
+We use diffusion maps \cite{coifman2005geometric}, which is a manifold learning technique, to uncover such low-dimensional structure and to analyze the results of the microscopic chemotaxis simulations.
 %
-Compared to principal components analysis (PCA) \cite{...}, a standard linear technique, diffusion maps provides a parameterization of data which lie on a low-dimensional, (possibly) {\em nonlinear} structure in a high-dimensional space.
+Compared to principal components analysis (PCA) \cite{shlens2005tutorial}, a standard linear technique, diffusion maps provides a parameterization of data which lie on a low-dimensional, (possibly) {\em nonlinear} structure in a high-dimensional space.
 %
-Diffusion maps is simply one of many recently introduced manifold learning techniques that could be used to analyzed the data \cite{...}.
+Diffusion maps is one of many recently introduced manifold learning techniques that could be used to analyzed this data \cite{roweis2000nonlinear, tenenbaum2000global, Belkin2003}.
 
@@
 %
-We discard the initial portion of each simulation trajectory corresponding to $t < 1$, as it tends to be very noisy and corresponds to initial relaxation of the system.
+We discard the initial portion of each simulation trajectory ($t < 1$), which corresponds to the initial relaxation of the system.
 %
-The observations $z(i) \in \mathbb{R}^n$ are the histogram of $x_1(i), x_2(i), \dots, x_N(i)$ into $n =32$ bins.
+The observation $z(i) \in \mathbb{R}^n$ is the histogram of $x_1(i), x_2(i), \dots, x_N(i)$ into $n = 32$ bins.
 %
@@
 %
-These observers should be informative as to the state of the system, as well as invariant to noise in the system. 
+These observers should be informative as to the state of the system, as well as insensitive to noise in the system. 
 %
-Two is a {\em distance metric} between the observations that captures a notion of locality: observations which we percieve to be similar should have a small distance.
+Two is a {\em distance metric} between the observations that captures a notion of locality: observations which we perceive to be similar should have a small distance.
 
@@
 %
-Instead of the standard Euclidean metric, we use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between pairs of histograms. 
+Instead of the standard Euclidean distance, we use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between pairs of histograms. 
 %
-Conceptually, EMD measures how much ``work'' it takes to transform one probability density into another.
+Conceptually, the EMD measures how much ``work'' it takes to transform one probability density into another.
 %
@@
 %
-Although the brute-force computation of the EMD is computationally expensive, there has been a plethora of work in developing efficient algorithms for the computation of EMD \cite{Pele-eccv2008, Pele-iccv2009}.
-%
-%TODO: we need to pay a special attention to the implementation of the EMD.
-We need to make it cleat that this distance can be very efficiently implemented  and therefore it is suitable for the practical uses and algorithm that we propose here as a viable tool for dynamical system analysis.
+Although the brute-force computation of the EMD is computationally expensive, there has been a plethora of work in developing efficient algorithms for its computation \cite{Pele-eccv2008, Pele-iccv2009}.
 %
@@
 %
-%TODO: add a couple of sentences to relate the notion of manifolds and parameterization to the dynamical system setup. Possibly use PCA reference.
 Our goal is to uncover a $d$-dimensional parameterization of the $m$ observations that respects the underlying manifold geometry.
@@
 \end{equation}
-where $\| \cdot \|$ denotes the appropriate norm for the data, and $\epsilon$ is a characteristic distance between the observations. 
+where $\| \cdot \|$ denotes the appropriate norm for the observations, and $\epsilon$ is a characteristic distance between the observations. 
 %
@@
 %
-The eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ then provide the embedding coordinates for the data, such that
+The eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ provide the embedding coordinates for the data, such that
 $\phi_{j}(i)$ is the $j^{th}$ embedding coordinate for $z(i)$.
 %
-The ``importance'' of coordinate $\phi_j$ is quantified by the corresponding eigenvalue $\mu_j$, e.g.
-the relative importance of $\phi_1$ versus $\phi_2$ can be measured by comparing $\mu_1$ and $\mu_2$.
+The ``importance'' of coordinate $\phi_j$ is quantified by the corresponding eigenvalue $\mu_j$, e.g., the relative importance of $\phi_1$ versus $\phi_2$ can be measured by comparing $\mu_1$ and $\mu_2$.
 %
@@
 %
-In both cases, $\phi_1$ is significantly more important than $\phi_2$, indicated by the separation between the first two eigenvalues (for the small $\lambda$ case, $\mu_1 = 0.5138$, and $\mu_2 = 0.3943$, and for the large $\lambda$ case, $\mu_1 = 0.5543$, and $\mu_2 = 0.3807$).
+In both cases, $\phi_1$ is significantly more important than $\phi_2$, as indicated by the separation between the first two eigenvalues (for the small $\lambda$ case, $\mu_1 = 0.5138$, and $\mu_2 = 0.3943$, and for the large $\lambda$ case, $\mu_1 = 0.5543$, and $\mu_2 = 0.3807$).
 %
@@
 %
-This is consistent with the macroscopic model: at small times, the particles are more condensed around $x=0$, and it is more difficult to distinguish the particles moving to the left from the particles moving to the right.
+This is in agreement with the macroscopic model: at small times, the particles are more condensed around $x=0$, and it is more difficult to distinguish the particles moving to the left from the particles moving to the right.
 %
@@
 %
-For the large $\lambda$ case (heat equation), shown in Figures \ref{subfig:large_lambda_p} and \ref{subfig:large_lambda_t}, we observe that at small times, the initial distribution $p$ is well organized in the embedding in Figure \ref{subfig:large_lambda_p}, as the velocity distribution of the particles is initially skewed and can be seen in the initial displacements
+For the large $\lambda$ case (heat equation), shown in Figures \ref{subfig:large_lambda_p} and \ref{subfig:large_lambda_t}, we observe that at small times, the initial distribution $p$ is well organized in the embedding in Figure \ref{subfig:large_lambda_p}, as the skew of the velocity distribution can be seen in the initial displacements.
 %
 On the other hand, for large times, we observe that the initial distribution $p$ is less organized in Figure \ref{subfig:large_lambda_p}, as the velocities have equilibrated and the initial distribution is less detectable in the particle density.
+%
+Overall, in both cases, we obtain, in an unsupervised data-driven manner, an accurate picture of the macroscopic variables that govern the system dynamics.
 
@@
 
-To emphasize the importance of using the correct distance metric, we also computed the diffusion maps embeddings of the two sets of simulation data, using the standard Euclidean distance between the histograms to compute the distances in \eqref{eq:W}.
+To emphasize the importance of using the correct distance metric, we applied diffusion maps to the two sets of simulation data, using the standard Euclidean distance between the histograms to compute the distances in \eqref{eq:W}.
 %
-There is no appreciable correlation between the embedding coordinates and the macroscopic variables $p$ and $t$ (the correlations between the embedding coordinates and the macroscopic variables are all  less than 60\%, with the correlations for the small $\lambda$ case being less than 20\%).
+There is no appreciable correlation between the embedding coordinates and the macroscopic variables $p$ and $t$ (the correlations between the embedding coordinates and the governing macroscopic variables are all  less than 60\%, with the correlations for the small $\lambda$ case being less than 20\%).
 %
-This is due to the inappropriate metric, which does not accurately describe the distances between histograms.
+This is due to using an inappropriate metric which does not accurately describe the distances between histograms.
 %
@@
 \includegraphics[width=8cm]{detect_change_eigenvalues}
-\caption{The first two eigenvalues from DMAPS as a function of $\lambda$.
-Each point is averaged over 10 simulation experiments (as in Figure~\ref{fig:dmaps_embed_emd}, one experiment consists of data collected at varying $p$ and $t$, for a fixed value of $\lambda$), with error bars indicating one standard deviation.
-The convergence of the eigenvalues at $\lambda \approx 125$ indicates switching of the dynamical ``mode'' of the system.}
+\caption{The first two eigenvalues from diffusion maps as a function of $\lambda$.
+Each point is averaged over $10$ simulation experiments (as in Figure~\ref{fig:dmaps_embed_emd}, one experiment consists of data collected at varying $p$ and $t$, for a fixed value of $\lambda$), with error bars denoting one standard deviation.
+The convergence of the eigenvalues at $\lambda \approx 125$ indicates switching of the dynamical ``mode''.}
 \label{fig:detect_change}
@@
 %
-We can also use these eigenvalues to detect the change in dynamical behavior.
+We can also use these eigenvalues to detect changes in dynamical behavior.
 %
-We showed that at the two asymptotic regimes ($\lambda \rightarrow 0$ and $\lambda \rightarrow \infty$), there is a gap between the two eigenvalues, as one of the macroscopic variables is significantly more important to the dynamical behavior.
+We showed that at the two asymptotic regimes ($\lambda \rightarrow 0$ and $\lambda \rightarrow \infty$), there is a gap between the two eigenvalues, as one of the two macroscopic variables is significantly more important to the dynamical behavior.
 %
-For intermediate $\lambda$ values, we expect the gap between $|\mu_1|$ and $|\mu_2|$ to narrow, and that the $\lambda$ value where $|\mu_1| \approx |\mu_2|$ corresponds to the transition between ``wave-like'' and ``heat-like'' behavior.
+For intermediate $\lambda$ values, we expect the gap between $|\mu_1|$ and $|\mu_2|$ to narrow, with the $\lambda$ value where $|\mu_1| \approx |\mu_2|$ corresponding to the transition between ``wave-like'' and ``heat-like'' behavior.
 
-To demonstrate this point, we performed a new set of experiments, where we simulated the dynamical system over a range of $\lambda$ values. 
+To demonstrate this point, we performed a new set of experiments, where we simulated the velocity jump process system over a range of $\lambda$ values. 
 % 
 Figure~\ref{fig:detect_change} shows the first two diffusion maps eigenvalues, $\mu_1$ and $\mu_2$, computed from these simulations, as a function of $\lambda$.
-% 
-%TODO: describe as a new experiment with a new data set. The goal of this experiment is to show implicit data-driven phase shift detection.
 %
@@
 %
-We can therefore detect where the change in dynamical behavior occurs using data-driven techniques, without any knowledge of the governing macroscopic model.
+We can therefore detect changes in dynamical behavior using data-driven techniques, without any knowledge of the governing macroscopic model.
 
@@
 %
-For data mining methods to be informative, processing the appropriate statistical moments or observables instead of individual particles/realizations is essential. 
+For data mining methods to be informative with regards to the underlying system dynamics, processing the appropriate statistical moments or observables instead of individual particles is essential. 
 %
@@
 %
-In this paper, these concepts were demonstrated using a dynamical model of cellular chemotaxis with a known analytic macroscopic behavior.
+In this paper, these concepts were demonstrated using a dynamical model of cellular chemotaxis with a known macroscopic behavior.
 %
@@
 
-%\section{Discussion}
+\begin{acknowledgments}
+
+The authors would like to thank...
+
+\end{acknowledgments}
 
@@
 \end{document}
+%\end{document}

### 2014-06-30 reducible_SDE/new_SDE/test1.m
added nultiscale writup and example with quadratic function

@@ -1,0 +1,261 @@
+clear all
+close all
+
+
+%%
+
+fmt = '-djpeg';
+res = '-r600';
+
+dim = 2;
+
+a = 0.1;
+epsilon = 0.01;
+
+DriftFn = @(t, x) [a; -x(2)];
+DiffnFn = @(t, x) [1 0; 0 1];
+
+dt = 0.5;
+dt_step = dt;
+nsteps = 2000;
+
+nsteps_per_step = dt / dt_step;
+if mod(nsteps_per_step, 1) ~= 0
+    disp('Step dt is not a multiple of dt');
+    return;
+end
+
+data_start = [0; 0];
+
+%%
+
+rng(123);
+
+SDE = sde(DriftFn, DiffnFn, 'StartState', data_start);
+[data_init, t, Z] = SDE.simulate(nsteps, 'DeltaTime', dt, 'NSTEPS', nsteps_per_step);
+data_init = data_init(2:end, :);
+t = t(2:end);
+
+%%
+
+make_fig;
+scatter(data_init(:,1),data_init(:,2),50,t,'.')
+xlabel('x')
+ylabel('y')
+h = colorbar('peer',gca);
+set(get(h,'xlabel'),'String', 't');
+axis equal 
+% print('raw_data', fmt, res);
+
+%%
+
+nsteps_burst = 200;
+dt_burst = dt/100;
+dt_burst_step = dt_burst;
+
+nsteps_burst_per_step = dt_burst / dt_burst_step;
+if mod(nsteps_burst_per_step, 1) ~= 0
+    disp('Step dt_burst is not a multiple of dt_burst');
+    return;
+end
+
+data_burst_init = zeros(nsteps_burst, dim, nsteps);
+for i=1:nsteps
+    SDE = sde(DriftFn, DiffnFn, 'StartState', data_init(i, :)');
+    [data_tmp, ~, Z] = SDE.simulate(1, 'DeltaTime', dt_burst, 'NSTEPS', nsteps_burst_per_step, 'ntrials', nsteps_burst);
+    data_burst_init(:, :, i) = squeeze(data_tmp(end, :, :))';
+end
+
+%%
+
+W = squareform(pdist(data_init)).^2;
+eps = median(W(:))/10;
+
+[V_init_dmaps, D_init_dmaps] = dmaps(W, eps, 10);
+
+make_fig;
[…]

### 2014-07-01 reducible_SDE/new_SDE/test_quadratic.m
did everal multiscale simulations

@@ -3,39 +3,61 @@ close all
 
 
 %% define parameters
+% 
+fig_idx = 1;
+epsilon = 1e-3;
+eps_dmaps = 0.01;
+dt_burst = 1e-8;
+% 
+% fig_idx = 2;
+% epsilon = 1e-3;
+% eps_dmaps = 10;
+% dt_burst = 1e-8;
+% 
+% fig_idx = 3;
+% epsilon = 1e-3;
+% eps_dmaps = 0.01;
+% dt_burst = 1e-5;
+% 
+% fig_idx = 4;
+% epsilon = 1e-4;
+% eps_dmaps = 0.01;
+% dt_burst = 1e-10;
+% 
+% fig_idx = 5;
+% epsilon = 1e-4;
+% eps_dmaps = 0.01;
+% dt_burst = 1e-5;
 
 dim = 2;
 
 a = 3;
-epsilon = 0.001;
-eps_dmaps = 0.02;
-
 DriftFn = @(t, x) [a; -x(2)/epsilon];
 DiffnFn = @(t, x) [1 0; 0 1];
+
 f1 = @(x) [x(:,1)+x(:,2).^2/epsilon x(:,2)/sqrt(epsilon)];
 
-dt = 1e-3;
-% dt_burst = 1e-8;
-dt_burst = 1e-5;
+dt = 1e-4;
+min_dt = 1e-5;
 
-
-%%
-
-figure;
-deltas = linspace(1e-4, 1, 100);
-semilogy(deltas, 0.5*(2 + epsilon + sqrt(9 + 2 * epsilon + epsilon^2))*deltas.^2)
-hold on
-semilogy(deltas, 10*deltas.^4, '-r')
-semilogy(deltas, (38 * dt_burst/(epsilon^2 + 2 * dt_burst))*deltas.^2, '-g')
-semilogy(sqrt(eps_dmaps/(0.5*(2 + epsilon + sqrt(9 + 2 * epsilon + epsilon^2)))), eps_dmaps, 'ok')
-legend('true distance','error from Taylor expansion','error from covariance estimation','location','best')
-
-
-
-%%
-min_dt = 1e-4;
-dt_step = min(dt, min_dt);
 nsteps = 3000;
+
+%%
+
+make_fig;
+deltas = logspace(-3, 1, 100);
+loglog(deltas, 0.5*(2 + epsilon + sqrt(9 + 2 * epsilon + epsilon^2))*deltas.^2)
[…]

### 2014-07-07 reducible_SDE/new_SDE/quadratic_example/test_quadratic_empirical.m
updated quadratic example

@@ -18,6 +18,8 @@ nsteps_burst = 200;
 
 %%
 
+dy_axis_lim = [1e-2 1e1 1e-10 1e5];
+
 make_fig;
 deltas = logspace(-2, 1, 100);
 loglog(deltas, 0.5*(2 + epsilon + sqrt(9 + 2 * epsilon + epsilon^2))*deltas.^2)
@@ -28,14 +30,17 @@ loglog(deltas, 10*deltas.^4, '-r')
 % loglog(sqrt(eps_dmaps/(0.5*(2 + epsilon + sqrt(9 + 2 * epsilon + epsilon^2)))), eps_dmaps, 'ok')
 % legend('linear approximation distance','error from covariance estimation','error from Taylor expansion','location','best')
 xlabel('$\| Y_2 - Y_1 \|$','interpreter','latex')
+ylabel('$\| X_2 - X_1 \|^2_{est}$','interpreter','latex')
+axis(dy_axis_lim)
 print(gcf, '-depsc', 'errors_function_dy');
 
 make_fig;
 loglog(deltas, 0.5*(2 + epsilon + sqrt(9 + 2 * epsilon + epsilon^2))*deltas.^2+(38 * dt_burst/(epsilon^2 + 2 * dt_burst))*deltas.^2+10*deltas.^4, '-k')
 hold on
-loglog(deltas, 10*deltas.^2, '-r')
+loglog(deltas, 2*deltas.^2, '-r')
 xlabel('$\| Y_2 - Y_1 \|$','interpreter','latex')
 ylabel('$\| X_2 - X_1 \|^2_{est}$','interpreter','latex')
+axis(dy_axis_lim)
 print(gcf, '-depsc', 'totaldist_function_dy');
 
 %% simulate SDE
@@ -76,13 +81,19 @@ inv_c1 = inv_c1 * dt_burst;
 
 Dis_Y = squareform(pdist(data1));
 
-% make_fig;
-% loglog(Dis_Y(:), Dis_data1(:), '.')
-% hold on
+idx = floor(linspace(1,numel(Dis_Y), 1000));
+make_fig;
+loglog(Dis_Y(idx), Dis_data1(idx), '.')
+hold on
[…]

### 2014-07-09 reducible_SDE/new_SDE/paper_outline.tex
updated stuff working with neta and ronen

\documentclass[12pt]{article}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\usepackage{epstopdf}
\graphicspath{ {quadratic_example/} }

\title{Data-driven Reduction of Multiscale Stochastic Dynamical Systems }

\begin{document}

\maketitle

\section{Outline}

Main message/contribution:
\begin{itemize}

\item We want to emphasize the link between multi scale SDEs and NIVs: NIVs rescales the variables so that the ``fast'' variables become small

\item

\end{itemize}

\begin{itemize}

\item Introduction

\item Multiscale SDE setting

\item Analysis of NIVs (remark that applies to more general settings)

\item Toy problem results (half-moon): illustration of the analysis + extraction of slow variable and ``suppression'' of fast variable depending on $\delta t$ and saturation time of fast

\item Conclusion:

If the cloud of samples is too big then we can observe the cloud of clouds (and those clouds can be histograms, Fourier, scattering, etc.) as a way to get smaller clouds.

Richardson extrapolation allows us to get an estimate of a second-order term in the covariance estimation, thereby locally approximating the function using a quadratic form, rather than a linear form,  which can lead to a better/improved/more accurate ``Mahalanobis'' metric.

\end{itemize}

\end{document}

### 2014-07-09 reducible_SDE/new_SDE/quadratic_example/covariances_inv.m

@@ -1,0 +1,32 @@
+function [inv_c, new_data, ranks, sing_vals] = covariances_inv(data, D)
+
+% Find neighborhoods
+[M, N, P] = size(data);
+
+% compute covariances
+inv_c = zeros(N, N, P);
+
+new_data = zeros(P, N);
+ranks = zeros(P, 1);
+sing_vals = zeros(P, D);
+
+center_idx = floor(M/2);
+
+for i = 1:P    
+    c = cov(data(:,:,i));
+    
+    [U, S, V] = svd(c);
+    
+    sing_vals(i,:) = diag(S(1:D,1:D));
+    
+    S(1,1) = S(1,1)/1000;
+    %c_hat = U(:,1:D) * S(1:D,1:D) * V(:,1:D)';
+    %inv_c(:,:,idx) = pinv(c_hat);
+    inv_c(:,:,i) = U(:,1:D) * S(1:D,1:D) * V(:,1:D)';
+%     inv_c(:,:,i) = U(:,1:D) * V(:,1:D)';
+    new_data(i,:) = data(center_idx, :, i);
+    ranks(i) = rank(c);
+    
+    
+end
+

### 2014-07-09 reducible_SDE/new_SDE/quadratic_example/simulate_quadratic2.m

@@ -1,0 +1,54 @@
+function [data_init, data_burst_init, data1, data1_burst, t] = simulate_quadratic2(nsteps, dt, nsteps_burst, dt_burst, epsilon, data_init)
+
+dim = 2;
+
+a = 3;
+DriftFn = @(t, x) [a; -x(2)/epsilon];
+DiffnFn = @(t, x) [1 0; 0 1];
+
+quad_scale = 1.0;
+f1 = @(x) [x(:,1)+quad_scale*x(:,1).*x(:,2).^2/epsilon x(:,2)/sqrt(epsilon)];
+
+min_dt = 1e-5;
+
+if nargin < 6
+    dt_step = min(dt, min_dt);
+    nsteps_per_step = dt / dt_step;
+    if mod(nsteps_per_step, 1) ~= 0
+        disp('Step dt is not a multiple of dt');
+        return;
+    end
+
+    data_start = [0; 0];
+
+    SDE = sde(DriftFn, DiffnFn, 'StartState', data_start);
+    [data_init, t, ~] = SDE.simulate(nsteps, 'DeltaTime', dt, 'NSTEPS', nsteps_per_step);
+    data_init = data_init(2:end, :);
+    t = t(2:end);
+else
+    nsteps = size(data_init, 1);
+    t = (1:nsteps)';
+end
+
+dt_burst_step = min(dt_burst, min_dt);
+nsteps_burst_per_step = round(dt_burst / dt_burst_step);
+% if mod(nsteps_burst_per_step, 1) ~= 0
+%     disp('Step dt_burst is not a multiple of dt_burst');
+%     return;
+% end
+
+data_burst_init = zeros(nsteps_burst, dim, nsteps);
+for i=1:nsteps
+    SDE = sde(DriftFn, DiffnFn, 'StartState', data_init(i, :)');
+    [data_tmp, ~, ~] = SDE.simulate(1, 'DeltaTime', dt_burst, 'NSTEPS', nsteps_burst_per_step, 'ntrials', nsteps_burst);
+    data_burst_init(:, :, i) = squeeze(data_tmp(end, :, :))';
+end
+
+data1 = f1(data_init);
+
+data1_burst = zeros(size(data_burst_init));
+for i=1:nsteps
+    data1_burst(:, :, i) = f1(data_burst_init(:, :, i));
+end
+
+

### 2014-07-09 reducible_SDE/new_SDE/quadratic_example/test_varying_quadratic.m

@@ -1,0 +1,83 @@
+clear all
+close all
+
+
+%% define parameters
+
+dim = 2;
+
+epsilon = 1e-3;
+
+
+%%
+
+eps_dmaps = 0.01;
+dt_burst = 1e-9;
+dt = 1e-4;
+
+
+nsteps = 3000;
+nsteps_burst = 200;
+
+
+rng(123);
+[data_init, data_burst_init, data1, data1_burst, t] = simulate_quadratic2(nsteps, dt, nsteps_burst, dt_burst, epsilon);
+
+
+make_fig;
+scatter(data_init(:,1),data_init(:,2),50,t,'.')
+xlabel('X(1)')
+ylabel('X(2)')
+h = colorbar('peer',gca);
+set(get(h,'xlabel'),'String', 't');
+axis equal
+
+make_fig;
+scatter(data1(:,1),data1(:,2),50,t,'.')
+xlabel('Y(1)')
+ylabel('Y(2)')
+h = colorbar('peer',gca);
+set(get(h,'xlabel'),'String', 't');
+axis equal
+
+
+%% DMAPS
+
+W = squareform(pdist(data_init)).^2;
+[V_init_dmaps, D_init_dmaps] = dmaps(W, eps_dmaps, 10);
+
+make_fig;
+scatter(data_init(:,1),data_init(:,2),50,V_init_dmaps(:,2),'.')
+xlabel('x')
+ylabel('y')
+h = colorbar('peer',gca);
+set(get(h,'xlabel'),'String', '\phi_1');
+axis equal
+
+
+%%
+
+[inv_c1, new_data1, ranks1] = covariances2(data1_burst, dim);
+inv_c1 = inv_c1 * dt_burst;
+
+%%
+
+eps_niv = eps_dmaps;
+
+[V1_niv, D1_niv, ~, Dis_data1] = NIV_return_dist(data1, inv_c1, eps_niv, 10, 0);
+
+
+make_fig;
+scatter(data_init(:,1),data_init(:,2),50,V1_niv(:,2),'.')
+xlabel('x')
+ylabel('y')
+h = colorbar('peer',gca);
+set(get(h,'xlabel'),'String', '\phi_1');
+axis equal
+
+make_fig;
+plot(V1_niv(:,2), V_init_dmaps(:,2),'.')
+xlabel('NIV from Y data')
+ylabel('DMAPS from X data')
+axis equal
+

### 2014-07-10 reducible_SDE/new_SDE/paper_outline.tex
added more to multiscale example

@@
 
-\section{Outline}
+\section{Main message/contribution}
 
-Main message/contribution:
 \begin{itemize}
 
-\item We want to emphasize the link between multi scale SDEs and NIVs: NIVs rescales the variables so that the ``fast'' variables become small
+\item Link between data mining and dynamical systems
 
-\item
+\item Doing slow variable reduction from SDEs using the Mahalanobis graph (using NIV for slow variable extraction)
+
+\item We give a full mathematical analysis of the slow variable reduction/procedure, including necessary conditions and user configuration
 
 \end{itemize}
+
+\section{Outline}
 
@@
 
-\item Multiscale SDE setting
+\item Multiscale SDE setting: fast/slow, want to recover slow variable(s) even when the fast variables are large/very noisy
 
-\item Analysis of NIVs (remark that applies to more general settings)
+\item Background on NIVs
+
+\item We want to emphasize the link between multi scale SDEs and NIVs: NIVs rescales the variables so that the ``fast'' variables become small
+
+\item Error and estimation analysis of NIVs (remark that applies to more general settings): give conditions/bounds when we can recover the intrinsic slow variable
+
+\item Empirical configuration/demonstration of the method, how to use the analysis to set the parameters in the algorithm
 
@@
 
+\item Experimental results/demonstrations:
+\begin{itemize}
+
+\item DMAPS gives the fast, NIVs gives the slow
+
+\item Illustration of different aspects (error terms) in the analysis
+
+\item Show empirically what happens as we change the neighborhood (example from notes 1) (recovery of fast variable deep in the spectrum)
+
+\end{itemize}
+
 \item Conclusion:
+
+We showed that in certain cases (when we do {\em not} have a simulator where we can change $\delta t$), the data cannot be processed as-is (we cannot find the right kernel scale given a fixed $\delta t$ such that we can accurately recover the slow variable).
 
@@
 
+saturation time of SDE has to be smaller than the timescales
+
+dmaps example: show we recover fast (or garbage), not slow variable
+
 \end{document}

### 2014-07-10 reducible_SDE/new_SDE/quadratic_example/simulate_quadratic3.m

@@ -1,0 +1,55 @@
+function [data_init, data_burst_init, data1, data1_burst, t] = simulate_quadratic3(nsteps, dt, nsteps_burst, dt_burst, epsilon, data_init)
+
+dim = 2;
+
+a = 3;
+DriftFn = @(t, x) [a; -x(2)/epsilon];
+DiffnFn = @(t, x) [1 0; 0 1];
+
+quad_scale = 0.2;
+sin_scale = 1;
+f1 = @(x) [x(:,1)+quad_scale*x(:,1).*x(:,2).^2/epsilon sin_scale*sin(x(:,1)*pi)+x(:,2)/sqrt(epsilon)];
+
+min_dt = 1e-5;
+
+if nargin < 6
+    dt_step = min(dt, min_dt);
+    nsteps_per_step = dt / dt_step;
+    if mod(nsteps_per_step, 1) ~= 0
+        disp('Step dt is not a multiple of dt');
+        return;
+    end
+
+    data_start = [0; 0];
+
+    SDE = sde(DriftFn, DiffnFn, 'StartState', data_start);
+    [data_init, t, ~] = SDE.simulate(nsteps, 'DeltaTime', dt, 'NSTEPS', nsteps_per_step);
+    data_init = data_init(2:end, :);
+    t = t(2:end);
+else
+    nsteps = size(data_init, 1);
+    t = (1:nsteps)';
+end
+
+dt_burst_step = min(dt_burst, min_dt);
+nsteps_burst_per_step = round(dt_burst / dt_burst_step);
+% if mod(nsteps_burst_per_step, 1) ~= 0
+%     disp('Step dt_burst is not a multiple of dt_burst');
+%     return;
+% end
+
+data_burst_init = zeros(nsteps_burst, dim, nsteps);
+for i=1:nsteps
+    SDE = sde(DriftFn, DiffnFn, 'StartState', data_init(i, :)');
+    [data_tmp, ~, ~] = SDE.simulate(1, 'DeltaTime', dt_burst, 'NSTEPS', nsteps_burst_per_step, 'ntrials', nsteps_burst);
+    data_burst_init(:, :, i) = squeeze(data_tmp(end, :, :))';
+end
+
+data1 = f1(data_init);
+
+data1_burst = zeros(size(data_burst_init));
+for i=1:nsteps
+    data1_burst(:, :, i) = f1(data_burst_init(:, :, i));
+end
+
+

### 2014-07-10 reducible_SDE/new_SDE/quadratic_example/test_varying_quadratic.m

@@ -22,6 +22,7 @@ nsteps_burst = 200;
 
 rng(123);
 [data_init, data_burst_init, data1, data1_burst, t] = simulate_quadratic2(nsteps, dt, nsteps_burst, dt_burst, epsilon);
+% [data_init, data_burst_init, data1, data1_burst, t] = simulate_quadratic3(nsteps, dt, nsteps_burst, dt_burst, epsilon);
 
 
 make_fig;

### 2014-07-10 reducible_SDE/paper/paper_draft.tex
added more stuff for reducible sde paper

\documentclass[1p]{elsarticle}

\usepackage{amsmath}
\usepackage{amsfonts}

\title{Data-driven Reduction of Multiscale Stochastic Dynamical Systems }

\author[PrincetonCBE]{C.J.~Dsilva}
\ead{cdsilva@princeton.edu}

\author[YaleMath]{R.~Talmon}
\ead{ronen.talmon@yale.edu}

\author[YaleMath]{R.R.~Coifman}
\ead{coifman@math.yale.edu}

\author[PrincetonCBE, PrincetonPACM]{I.G.~Kevrekidis\corref{cor1}}
\ead{yannis@princeton.edu}

\address[PrincetonCBE]{Department of Chemical and Biological Engineering, Princton University, Princeton, New Jersey, 08544, USA}
\address[YaleMath]{Department of Mathematics, Yale University, New Haven, Connecticut, 06520, USA}
\address[PrincetonPACM]{Program in Applied and Computational Mathematics, Princton University, Princeton, New Jersey, 08544, USA}

\cortext[cor1]{Corresponding author}

\begin{document}

\begin{abstract}

\begin{itemize}

\item Link between data mining and dynamical systems

\item Extracting or reducing an SDE system to its slow variables using the Mahalanobis graph and NIVs

\item Full mathematical analysis of the slow variable reduction procedure, including necessary conditions for accurate recovery of the slow variables and a framework for empirically setting the algorithm parameters

\end{itemize}

\end{abstract}

\begin{keyword}

\end{keyword}

\maketitle

\section{Introduction}

\section{Theory}

\subsection{Multiscale SDEs}

Multiscale SDE setting: fast/slow, want to recover slow variable(s) even when the fast variables are large/very noisy

\begin{equation}
\begin{aligned}
dx &= adt + dW_1\\
dy &= -\frac{y}{\epsilon} dt + \frac{1}{\sqrt{\epsilon}} dW_2
\end{aligned}
\end{equation}

So $y$ is a fast noise whose equilibrium measure is bounded and $\mathcal{O}(1)$

\subsection{Nonlinear Intrinsic Variables}

Background on NIVs

\begin{equation}
dx_i = a(X) dt + dW_i
\end{equation}
and we will let $X = \begin{bmatrix} X^1 \\ X^2 \\ \vdots \\ X^n \end{bmatrix}$

We want to emphasize the link between multi scale SDEs and NIVs: NIVs rescales the variables so that the ``fast'' variables become small

saturation time of SDE has to be smaller than the timescales

Let $y = \frac{z}{\sqrt{\epsilon}}$. Then
\begin{equation}
\begin{aligned}
dx &= adt + dW_1\\
dz &= -\frac{z}{\epsilon} dt +  dW_2
\end{aligned}
\end{equation}

Therefore, now $z$ is a stochastic variable whose diffusion is unity, and whose equilibrium measure is $\mathcal{O}(\epsilon)$. 
%
Therefore, when $\epsilon \ll 1$, we have that $z \ll x$, and so we will recover $x$ only.

\subsection{Errors and Estimation}

Error and estimation analysis of NIVs (remark that applies to more general settings): give conditions/bounds when we can recover the intrinsic slow variable

Empirical configuration/demonstration of the method, how to use the analysis to set the parameters in the algorithm

Given $X$ governed by the above SDE, we can sample ``bursts'' around a point $X_0$ in order to estimate the covariance.

These clouds will be distributed

\begin{equation}
X \sim \mathcal{N}\left( X_0, \delta t I \right)
\end{equation}

We now consider a function $\mathbf{f}: \mathbb{R}^n \mapsto \mathbb{R}^m$.
%
Let $f^i(X)$ denote the $i^{th}$ entry of $\mathbf{f}$, and let $f^i_j(X) = \frac{\partial f^i}{\partial X^j}$.
%
By Taylor expansion around $X=X_0$,
%
\begin{eqnarray}
\hat{C}(X_0) &=& \delta t J_1 J_1^T 
+ \frac{\delta t^2}{2} \left[ J_1 J_3^T + J_3 J_1^T  
+ \frac{3}{2} J_2 J_2^T 
-\frac{1 }{2} J_2 1 1^T J_2^T \right]
+ \mathcal{O} (\delta t^3) %\\
%&=& \delta t J_1 J_1^T 
%+ \frac{\delta t^2}{2} U \Sigma V
%+ \mathcal{O} (\delta t^3)
\end{eqnarray}
where
\begin{equation}
\begin{aligned}
J_1(X_0) &= \begin{bmatrix}
f_1^1 & f_2^1 & \dots & f_n^1 \\
\vdots & \vdots & \cdots & \vdots \\
f_1^m & f_2^m & \dots & f_n^m
\end{bmatrix}\\
%
J_2(X_0) &= \begin{bmatrix}
f_{11}^1 & f_{22}^1 & \dots & f_{nn}^1 \\
\vdots & \vdots & \cdots & \vdots \\
f_{11}^m & f_{22}^m & \dots & f_{nn}^m
\end{bmatrix}\\
%
J_3(X_0) &= \begin{bmatrix}
f_{111}^1 & f_{222}^1 & \dots & f_{nnn}^1 \\
\vdots & \vdots & \cdots & \vdots \\
f_{111}^m & f_{222}^m & \dots & f_{nnn}^m
\end{bmatrix}
\end{aligned}
\end{equation}
%
In this equation, we see what we expect (a linear function of the variance, due to the gradient of $f$) and a second (higher-order) term arising from the curvature and third derivative of $f$.
%
The importance of this second term increases as the curvature increases, as well as when $\delta t$ increases.

Then
\begin{equation}
\hat{C}^{-1}(X_0) \approx \frac{1}{\delta t} \left( J J^T \right)^{-1} - A
\end{equation}
%
where 
\begin{equation}
A = \frac{1}{2} (J_1 J_1^T)^{-1} U \left(\Sigma^{-1} + \frac{\delta t}{2} V \left( J_1 J_1^T \right)^{-1} U \right)^{-1} V \left( J_1 J_1^T \right)^{-1}
\end{equation}
and
\begin{equation}
U \Sigma V = J_1 J_3^T + J_3 J_1^T + \frac{3}{2} J_2 J_2^T -\frac{1 }{2} J_2 1 1^T J_2^T 
\end{equation}
$A$ is a function of the first-, second-, and third-order derivatives of $\mathbf{f}$.

We now approximate the distances $\| X_2 - X_1 \|$.
%
Let $g = f^{-1}: \mathbb{R}^m \mapsto \mathbb{R}^n$, and let $Y_1 = f(X_1)$, and $Y_2 = f(X_2)$.
%
Again, we let $g_j^i = \frac{\partial g^i}{\partial Y^j}$.
%
Then, by Taylor expansion, we obtain
\begin{equation} \label{eq:taylor_expansion}
\begin{aligned}
&\| X_2 - X_1 \|^2 \\
\le&  \left| \frac{1}{2} (Y_2 - Y_1 )^T ((J J^T)^{-1} (Y_1) + (J J^T)^{-1}(Y_2)) (Y_2 - Y_1 ) \right| \\
& + n \left( \left| K_1 K_2 \right| + \left| \frac{ K_2^2}{4} \right|  + \left| \frac{K_1 K_3}{3} \right|  \right) \| Y_2 - Y_1 \| ^4  \\
& + \mathcal{O} (\|Y_1 - Y_2 \|^6 ) 
\end{aligned}
\end{equation}
%
where
%
\begin{equation}
\begin{aligned}
K_1 &= \sup_{i,j,Y} |g_j^i(Y)|\\
K_2 &= \sup_{i,j,k,Y} |g_{jk}^i(Y)|\\
K_3 &= \sup_{i,j,k,l,Y} |g_{jkl}^i(Y)|
\end{aligned}
\end{equation}

Substituting our estimate of the covariance into \eqref{eq:taylor_expansion}, we obtain
\begin{equation}
\begin{aligned}
&\| X_2 - X_1 \|^2 \\
\approx &  \frac{1}{2} (Y_2 - Y_1)^T \left((J J^T)^{-1}(Y_1)  + (J J^T)^{-1}(Y_2) \right) (Y_2 - Y_1) \\
& - \frac{\delta t}{2} (Y_2 - Y_1)^T \left( A(Y_1) + A(Y_2) \right) (Y_2 - Y_1)\\
& + n \left( \left| K_1 K_2 \right| + \left| \frac{ K_2^2}{4} \right|  + \left| \frac{K_1 K_3}{3} \right|  \right) \| Y_2 - Y_1 \| ^4  \\
& + \mathcal{O} (\|Y_1 - Y_2 \|^6 ) 
\end{aligned}
\end{equation}
%
We want the first term in this expression to dominate.

\section{Results}

 Toy problem results (half-moon): illustration of the analysis + extraction of slow variable and ``suppression'' of fast variable depending on $\delta t$ and saturation time of fast

 Define the following two-dimensional system of SDEs
\begin{eqnarray}
dX(1) &=& 3 dt + dW_1 \\ 
dX(2) &=& \frac{-X(2)}{\epsilon} dt + dW_2 
\end{eqnarray}
where $W_1, W_2$ are independent Brownian motions, and $\epsilon \ll 1$.

Let
\begin{eqnarray}
f^1(X) &=& X(1) + \frac{ X(2)^2}{\epsilon} \\
f^2(X) &=& \frac{X(2)}{\sqrt{\epsilon}}
\end{eqnarray}
Therefore, $X(2)$ is now a ``fast'' variable, and we have further applied a nonlinear transformation to this fast-slow system.

The inverse function $g = f^{-1}$ is
\begin{eqnarray}
g^1(Y) &=& Y(1) - Y(2)^2 \\
g^2(Y) &=& Y(2) \sqrt{\epsilon}
\end{eqnarray}

\subsection{NIVs versus DMAPS}

DMAPS gives the fast, NIVs gives the slow

\subsection{Determining Appropriate Parameters}

Illustration of different aspects (error terms) in the analysis

The three relevant quantities for NIVs are 
\begin{itemize}
\item The linearized approximation to the distance,
%
\begin{equation}
(Y_2 - Y_1)^T (JJ^T)^{-1} (Y_2 - Y_1) 
\approx \mathcal{O} \left(  \frac{1}{2} \left( 2 + \epsilon + \sqrt{ 9 + 2 \epsilon + \epsilon^2}\right) \Delta Y^2\right) 
\end{equation}
where $\Delta Y =  \|Y_2 - Y_1\|$

\item The error from the {\em estimation} of the covariance.
%
This is
\begin{equation}
\frac{\delta t}{2} (Y_2 - Y_1)^T A (Y_2 - Y_1) \approx \mathcal{O} \left( \frac{38 \delta t}{\epsilon ^2 + 2 \delta t}\Delta Y^2 \right)
\end{equation}

\item The errors from the {\em truncation} of the Taylor expansion for the Mahalanobis distance, which is 
\begin{equation}
n \left( \left| K_1 K_2 \right| + \left| \frac{ K_2^2}{4} \right|  + \left| \frac{K_1 K_3}{3} \right|  \right) \| Y_2 - Y_1 \| ^4 \approx \mathcal{O} \left( 10 \Delta Y^4  \right) 
\end{equation}

\end{itemize}

\subsection{Recovery of Fast Variable}

Show empirically what happens as we change the neighborhood (example from notes 1) (recovery of fast variable deep in the spectrum)

\section{Conclusion}

We showed that in certain cases (when we do {\em not} have a simulator where we can change $\delta t$), the data cannot be processed as-is (we cannot find the right kernel scale given a fixed $\delta t$ such that we can accurately recover the slow variable).

If the cloud of samples is too big then we can observe the cloud of clouds (and those clouds can be histograms, Fourier, scattering, etc.) as a way to get smaller clouds.

Richardson extrapolation allows us to get an estimate of a second-order term in the covariance estimation, thereby locally approximating the function using a quadratic form, rather than a linear form,  which can lead to a better/improved/more accurate ``Mahalanobis'' metric.

\end{document}

### 2014-07-10 reducible_SDE/paper/paper_outline.tex
added more stuff for reducible sde paper

--- 2014-07-10 reducible_SDE/new_SDE/paper_outline.tex
@@
 
-\item Doing slow variable reduction from SDEs using the Mahalanobis graph (using NIV for slow variable extraction)
+\item Extracting or reducing an SDE system to its slow variables using the Mahalanobis graph and NIVs
 
-\item We give a full mathematical analysis of the slow variable reduction/procedure, including necessary conditions and user configuration
+\item Full mathematical analysis of the slow variable reduction procedure, including necessary conditions for accurate recovery of the slow variables and a framework for empirically setting the algorithm parameters
 

### 2014-07-11 reducible_SDE/paper/paper_draft.tex
edied multiscale paper

@@
 
-Multiscale SDE setting: fast/slow, want to recover slow variable(s) even when the fast variables are large/very noisy
+Setting
+\begin{itemize}
+\item system of SDEs with separation of timescales
+\item want to recover slow variable(s) even when the fast variables are have large variations
+\end{itemize}
 
-\begin{equation}
+Consider equations of the form
+\begin{equation} \label{eq:fast_slow_SDE}
 \begin{aligned}
-dx &= adt + dW_1\\
-dy &= -\frac{y}{\epsilon} dt + \frac{1}{\sqrt{\epsilon}} dW_2
+dx &= adt + dW^1\\
+dy &= -\frac{y}{\epsilon} dt + \frac{1}{\sqrt{\epsilon}} dW^2
 \end{aligned}
@@
 
-So $y$ is a fast noise whose equilibrium measure is bounded and $\mathcal{O}(1)$
+So $y$ is a fast noise whose equilibrium measure is bounded and $\mathcal{O}(1)$.
+%
+We assume that the saturation time of fast scale  of the SDE is smaller than the other relevant timescales.
+
+TODO: should we write the SDE in a more general form?
 
@@
 Background on NIVs 	
+\begin{itemize}
+
+\item Want to recover ``intrinsic'' variables that are independent of the measurement.observation function
+\item We define intrinsic as variables that have uncoupled noise with unit variance
+\end{itemize}
 
 \begin{equation}
-dx_i = a(X) dt + dW_i
+dX^i = a(X) dt + dW^i
 \end{equation}
@@
 
-We want to emphasize the link between multi scale SDEs and NIVs: NIVs rescales the variables so that the ``fast'' variables become small
-
-saturation time of SDE has to be smaller than the timescales
-
-Let $y = \frac{z}{\sqrt{\epsilon}}$. Then
+We want to emphasize the link between multi scale SDEs and NIVs: NIVs rescales the variables so that the ``fast'' variables become small.
+%
+Consider the SDE in \eqref{eq:fast_slow_SDE}, and let $y = \frac{z}{\sqrt{\epsilon}}$. 
+%
+Then
 \begin{equation}
@@
 \end{equation}
-
-Therefore, now $z$ is a stochastic variable whose diffusion is unity, and whose equilibrium measure is $\mathcal{O}(\epsilon)$. 
 %
-Therefore, when $\epsilon \ll 1$, we have that $z \ll x$, and so we will recover $x$ only.
+Therefore, $z$ is a stochastic variable whose diffusion is unity, and so $x$ and $z$ are the ``intrinsic variables'' for the underlying SDE.
+%
+However, the equilibrium measure of $z$ is $\mathcal{O}(\epsilon)$, and so when $\epsilon \ll 1$, $z \ll x$, and so we will only recover $x$ (the slow variable).
 

### 2014-07-14 reducible_SDE/paper/paper_draft.tex
Added paper draft

--- 2014-07-10 reducible_SDE/paper/paper_draft.tex
@@
 
-\begin{equation}
+\begin{equation} \label{eq:multiscale_SDE}
 \begin{aligned}
@@
 
-We want to emphasize the link between multi scale SDEs and NIVs: NIVs rescales the variables so that the ``fast'' variables become small
-
-saturation time of SDE has to be smaller than the timescales
-
-Let $y = \frac{z}{\sqrt{\epsilon}}$. Then
+We want to emphasize the link between multi scale SDEs and NIVs: NIVs rescales the variables so that the ``fast'' variables become small.
+%
+For example, in \eqref{eq:multiscale_SDE}, let $y = \frac{z}{\sqrt{\epsilon}}$. 
+%
+Then
 \begin{equation}
@@
 
+Note: saturation time of SDE has to be smaller than the timescales
+
 \subsection{Errors and Estimation}
@@
 Error and estimation analysis of NIVs (remark that applies to more general settings): give conditions/bounds when we can recover the intrinsic slow variable
-
-Empirical configuration/demonstration of the method, how to use the analysis to set the parameters in the algorithm
 
@@
 
+Empirical configuration/demonstration of the method, how to use the analysis to set the parameters in the algorithm.
+
+From \eqref{eq:taylor_expansion}, we can see that the squared {\em intrinsic} distance $\| X_2 - X_1 \|^2$, is a function of $\|Y_2 - Y_1\|$ and $\delta t$. 
+%
+If the first term in the expansion dominates, then $\|X_2 - X_1 \|^2$ should be independent of $\delta t$, and should vary quadratically with $\| Y_2 - Y_1\|$. 
+%
+We can find such a regime empirically (by plotting $\| X_2 - X_1 \|^2$ as a function of $\delta t$ and $\|Y_2 - Y_1\|$), and select a kernel distance where both conditions hold.
+
 \section{Results}

### 2014-07-17 reducible_SDE/paper/paper_draft.tex
edited paper'

--- 2014-07-11 reducible_SDE/paper/paper_draft.tex
@@
 \usepackage{amsfonts}
+
+\usepackage{subcaption}
 
@@
 
-Error and estimation analysis of NIVs (remark that applies to more general settings): give conditions/bounds when we can recover the intrinsic slow variable
-
-Empirical configuration/demonstration of the method, how to use the analysis to set the parameters in the algorithm
+\begin{itemize}
+\item Error and estimation analysis of NIVs (remark that applies to more general settings): give conditions/bounds when we can recover the intrinsic slow variable
+%
+\item Empirical configuration/demonstration of the method, how to use the analysis to set the parameters in the algorithm
+\end{itemize}
 
@@
  Define the following two-dimensional system of SDEs
-\begin{eqnarray}
+\begin{eqnarray} \label{eq:init_data}
 dX(1) &=& 3 dt + dW_1 \\ 
@@
 Let
-\begin{eqnarray}
+\begin{eqnarray}\label{eq:transformed_data}
 f^1(X) &=& X(1) + \frac{ X(2)^2}{\epsilon} \\
@@
 
+\begin{figure}[h]
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{data_init}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{data_transformed}
+\caption{}
+\end{subfigure}
+\caption{(a) The original data, simulated from \eqref{eq:init_data}. (b) The data from (a), transformed according to \eqref{eq:transformed_data}. }
+\label{fig:initial_data}
+\end{figure}
+
 \subsection{NIVs versus DMAPS}
@@
 DMAPS gives the fast, NIVs gives the slow
+
+\begin{figure}[h]
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{data_transformed_colored_NIV}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{data_transformed_colored_DMAPS}
+\caption{}
+\end{subfigure}
+\caption{(a) The data from Figure~\ref{fig:initial_data}(b), colored by the first NIV.
+Note that we recover the slow variable.
+(b) The data from Figure~\ref{fig:initial_data}(b), colored by the first DMAPS variable.
+Note that we do {\em not} recover the slow variable.
+}
+\label{fig:NIV_versus_DMAPS}
+\end{figure}
 
@@
 
+\begin{figure}[h]
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{errors_function_dt}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{empirical_totaldist_function_dt}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{errors_function_dy}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{empirical_totaldist_function_dy}
+\caption{}
+\end{subfigure}
+\caption{(a) Analytical estimation of errors as a function of $\delta t$. 
+%
+(b) Empirical estimation of total distance as a function of $\delta t$.  
+%
+(c) Analytical estimation of errors as a function of $\Delta Y$. 
+%
+(d) Empirical estimation of total distance as a function of $\Delta Y$}
+\end{figure}
+
 \subsection{Recovery of Fast Variable}

### 2014-07-17 reducible_SDE/paper/test_quadratic_empirical.m

@@ -26,7 +26,7 @@ loglog(deltas, 0.5*(2 + epsilon + sqrt(9 + 2 * epsilon + epsilon^2))*deltas.^2)
 hold on
 loglog(deltas, (38 * dt_burst/(epsilon^2 + 2 * dt_burst))*deltas.^2, '-g')
 loglog(deltas, 10*deltas.^4, '-r')
-% loglog(deltas, 0.5*(2 + epsilon + sqrt(9 + 2 * epsilon + epsilon^2))*deltas.^2+(38 * dt_burst/(epsilon^2 + 2 * dt_burst))*deltas.^2+10*deltas.^4, '-k')
+loglog(deltas, 0.5*(2 + epsilon + sqrt(9 + 2 * epsilon + epsilon^2))*deltas.^2+(38 * dt_burst/(epsilon^2 + 2 * dt_burst))*deltas.^2+10*deltas.^4, '-k')
 % loglog(sqrt(eps_dmaps/(0.5*(2 + epsilon + sqrt(9 + 2 * epsilon + epsilon^2)))), eps_dmaps, 'ok')
 % legend('linear approximation distance','error from covariance estimation','error from Taylor expansion','location','best')
 xlabel('$\| Y_2 - Y_1 \|$','interpreter','latex')
@@ -34,14 +34,14 @@ ylabel('$\| X_2 - X_1 \|^2_{est}$','interpreter','latex')
 axis(dy_axis_lim)
 print(gcf, '-depsc', 'errors_function_dy');
 
-make_fig;
-loglog(deltas, 0.5*(2 + epsilon + sqrt(9 + 2 * epsilon + epsilon^2))*deltas.^2+(38 * dt_burst/(epsilon^2 + 2 * dt_burst))*deltas.^2+10*deltas.^4, '-k')
-hold on
-loglog(deltas, 2*deltas.^2, '-r')
-xlabel('$\| Y_2 - Y_1 \|$','interpreter','latex')
-ylabel('$\| X_2 - X_1 \|^2_{est}$','interpreter','latex')
-axis(dy_axis_lim)
-print(gcf, '-depsc', 'totaldist_function_dy');
+% make_fig;
+% loglog(deltas, 0.5*(2 + epsilon + sqrt(9 + 2 * epsilon + epsilon^2))*deltas.^2+(38 * dt_burst/(epsilon^2 + 2 * dt_burst))*deltas.^2+10*deltas.^4, '-k')
[…]

### 2014-07-17 reducible_SDE/paper/paper_draft.tex

@@
 \usepackage{subcaption}
+
+\usepackage{epstopdf}
 
@@
 %
+
 Therefore, $z$ is a stochastic variable whose diffusion is unity, and so $x$ and $z$ are the ``intrinsic variables'' for the underlying SDE.
@@
 We want the first term in this expression to dominate.
+
+Empirical configuration/demonstration of the method, how to use the analysis to set the parameters in the algorithm.
+
+From \eqref{eq:taylor_expansion}, we can see that the squared {\em intrinsic} distance $\| X_2 - X_1 \|^2$, is a function of $\|Y_2 - Y_1\|$ and $\delta t$. 
+%
+If the first term in the expansion dominates, then $\|X_2 - X_1 \|^2$ should be independent of $\delta t$, and should vary quadratically with $\| Y_2 - Y_1\|$. 
+%
+We can find such a regime empirically (by plotting $\| X_2 - X_1 \|^2$ as a function of $\delta t$ and $\|Y_2 - Y_1\|$), and select a kernel distance where both conditions hold.
 

### 2014-07-21 reducible_SDE/paper/paper_draft.tex
added more to paper

@@
 
-\item Link between data mining and dynamical systems
+\item Link between data mining and dynamical systems analysis
 
-\item Extracting or reducing an SDE system to its slow variables using the Mahalanobis graph and NIVs
+\item Extracting/reducing an SDE system to its slow variables using the Mahalanobis graph and NIVs
 
-\item Full mathematical analysis of the slow variable reduction procedure, including necessary conditions for accurate recovery of the slow variables and a framework for empirically setting the algorithm parameters
+\item Full mathematical analysis of the slow variable reduction procedure, including necessary conditions for accurate recovery of the intrinsic variables and a framework for empirically setting the algorithm parameters
 
@@
 
-Setting
-\begin{itemize}
-\item system of SDEs with separation of timescales
-\item want to recover slow variable(s) even when the fast variables are have large variations
-\end{itemize}
+Given a system of SDEs with separation of timescales, we want to recover slow variable(s) even when the fast variables are have large variations.
 
-Consider equations of the form
+Consider the following system of SDEs:
 \begin{equation} \label{eq:fast_slow_SDE}
@@
 
-So $y$ is a fast noise whose equilibrium measure is bounded and $\mathcal{O}(1)$.
+So $x$ is the slow variable, and $y$ is a fast noise whose equilibrium measure is bounded and $\mathcal{O}(1)$.
 %
 We assume that the saturation time of fast scale  of the SDE is smaller than the other relevant timescales.
+%
+We would like to recover $x$ using data-driven techniques.
 
-TODO: should we write the SDE in a more general form?
+%TODO: should we write the SDE in a more general form?
 
@@
 
-Background on NIVs 	
-\begin{itemize}
-
-\item Want to recover ``intrinsic'' variables that are independent of the measurement.observation function
-\item We define intrinsic as variables that have uncoupled noise with unit variance
-\end{itemize}
-
-\begin{equation}
+The goal of NIVs is to recover ``intrinsic'' variables that are independent of the measurement/observation function.
+%
+We define intrinsic variables as those variables which have uncoupled noise with unit variance,
+i.e., the intrinsic variables $X^1, \dots, X^n$ are governed by the following SDEs:
+\begin{equation} \label{eq:NIV_formulation}
 dX^i = a(X) dt + dW^i
 \end{equation}
-and we will let $X = \begin{bmatrix} X^1 \\ X^2 \\ \vdots \\ X^n \end{bmatrix}$
+where $X = \begin{bmatrix} X^1 & X^2 & \cdots & X^n \end{bmatrix}^T$.
 
-We want to emphasize the link between multi scale SDEs and NIVs: NIVs rescales the variables so that the ``fast'' variables become small.
+We want to emphasize the link between multiscale SDEs and NIVs: NIVs rescales the noises, so that the ``fast'' variables become small.
 %
@@
 %
-
 Therefore, $z$ is a stochastic variable whose diffusion is unity, and so $x$ and $z$ are the ``intrinsic variables'' for the underlying SDE.
 %
-However, the equilibrium measure of $z$ is $\mathcal{O}(\epsilon)$, and so when $\epsilon \ll 1$, $z \ll x$, and so we will only recover $x$ (the slow variable).
+However, the equilibrium measure of $z$ is $\mathcal{O}(\epsilon)$, and so when $\epsilon \ll 1$, $z \ll x$, and so we will recover $x$ (the slow variable). 
+%
+In general, we want to recover the slow variable(s) even when the observations of the underlying intrinsic stochastic process are obstructed by a {\em nonlinear} measurement function.
 
@@
 
-Given $X$ governed by the above SDE, we can sample ``bursts'' around a point $X_0$ in order to estimate the covariance.
+Given $X$ governed by the SDE in \eqref{eq:NIV_formulation}, we can sample ``bursts'' around a point $X_0$ in order to estimate the local covariance of the noise.
 
@@
 
-Empirical configuration/demonstration of the method, how to use the analysis to set the parameters in the algorithm.
-
 From \eqref{eq:taylor_expansion}, we can see that the squared {\em intrinsic} distance $\| X_2 - X_1 \|^2$, is a function of $\|Y_2 - Y_1\|$ and $\delta t$. 
@@
 
- Toy problem results (half-moon): illustration of the analysis + extraction of slow variable and ``suppression'' of fast variable depending on $\delta t$ and saturation time of fast
-
 Define the following two-dimensional system of SDEs
@@
 where $W_1, W_2$ are independent Brownian motions, and $\epsilon \ll 1$.
+%
 
@@
 \end{eqnarray}
-Therefore, $X(2)$ is now a ``fast'' variable, and we have further applied a nonlinear transformation to this fast-slow system.
-
+Therefore, the two ``intrinsic variables'' $X(1), X(2)$ have been nonlinearly transformed and rescaled such that the new variables $f^1(X), f^2(X)$ are both $\mathcal{O}(1)$, even though $X(1)$ is the ``fast'' variable.
+%
 The inverse function $g = f^{-1}$ is
@@
 \end{eqnarray}
+%
+Data from a simulation is shown in Figure~\ref{fig:initial_data}.
 
@@
 \end{subfigure}
-\caption{(a) The original data, simulated from \eqref{eq:init_data}. (b) The data from (a), transformed according to \eqref{eq:transformed_data}. }
+\caption{(a) The original data, simulated from \eqref{eq:init_data} with $\epsilon = 10^{-3}$.
+We take $3000$ timesteps with $dt = 10^{-4}$ (b) The data from (a), transformed according to \eqref{eq:transformed_data}.
+}
 \label{fig:initial_data}
@@
 
-DMAPS gives the fast, NIVs gives the slow
+We first want to demonstrate the utility of NIVs over DMAPS. 
+%
+We compute the first (nontrivial) NIV and DMAPS coordinate using the data in Figure~\ref{fig:initial_data}(b). 
+%
+The results are shown in Figure~\ref{fig:NIV_versus_DMAPS}.
+%
+NIV accurately recovers the slow variable (the coloring in Figure~\ref{fig:NIV_versus_DMAPS}(a) is consistent with the coloring in Figure~\ref{fig:initial_data}(b)). 
+%
+In contrast, DMAPS does not recover the slow variable.
 
@@
 \end{subfigure}
-\caption{(a) The data from Figure~\ref{fig:initial_data}(b), colored by the first NIV.
+\caption{(a) The data from Figure~\ref{fig:initial_data}(b), colored by the first NIV (we take $\sigma_{kernel} = 0.01$, and $\delta t = 10^{-9}$).
 Note that we recover the slow variable.
-(b) The data from Figure~\ref{fig:initial_data}(b), colored by the first DMAPS variable.
-Note that we do {\em not} recover the slow variable.
-}
+(b) The data from Figure~\ref{fig:initial_data}(b), colored by the first DMAPS variable (we take $\sigma_{kernel} = 0.01$).
+Note that we do {\em not} recover the slow variable.}
 \label{fig:NIV_versus_DMAPS}
@@
 
-Illustration of different aspects (error terms) in the analysis
+We would like to demonstrate how we can {\em empirically} determine the appropriate parameters $\delta t$ and $\sigma_{kernel}$ so that we can accurately recover the slow variable(s).
 
-The three relevant quantities for NIVs are 
+For the system in \eqref{eq:init_data} and \eqref{eq:transformed_data}, the three relevant quantities for NIVs are 
 \begin{itemize}
@@
 \end{itemize}
+%
+We can plot each of these quantities analytically as a function of $\delta t$ and $\|Y_2 - Y_1\|$. 
+%
+The results are shown in Figure~\ref{fig:empirical_estimation}(a)~and~(c). 
+%
+In practice, for more complex SDEs, we cannot write down the three terms analytically.
+%
+We can only measure the sum of the three terms; this is the estimate of the intrinsic distance that we calculate, $\|X_2 - X_1 \|^2_{est}$.
+%
+This is also shown in Figure~\ref{fig:empirical_estimation}(a)~and~(c).
+%
+Clearly, when higher-order terms become important corresponds to a ``kink'' in the plots. 
+%
+We can also plot the empirical estimates of $\| X_2 - X_1 \|^2_{est}$ obtained from simulation data, as a function of $\delta t$ and $\| Y_2 - Y_1 \|$.
+%
+This is shown in Figure~\ref{fig:empirical_estimation}(b)~and~(d). 
+%
+The same ``kink'' is evident in these plots at approximately the same position as the analytical plots. 
+%
+Therefore, we choose $\delta t$ and a $\sigma_{kernel}$ which corresponds to a $\|X_2 - X_1 \|_{est}^2$ value {\em before} the kink. 
+%
+For this particular example, we take $\delta t = 10^{-9}$ and $\sigma_{kernel} = 0.01$, which allows us to accurately recover the slow variable.
 
@@
 \end{subfigure}
-\caption{(a) Analytical estimation of errors as a function of $\delta t$. 
+\caption{(a) Total estimated distance (black) and the three separate terms in the estimated distance, as a function of $\delta t$.
+The linearized approximation is in blue, the error from the covariance estimation is in green, and the error from the truncation of the distance expansion is in red.
 %
-(b) Empirical estimation of total distance as a function of $\delta t$.  
+(b) Empirical estimation of total distance as a function of $\delta t$, with $\|Y_2 - Y_1 \|^2 = 0.01$.  
 %
-(c) Analytical estimation of errors as a function of $\Delta Y$. 
+(c) Total estimated distance (black) and the three separate terms in the estimated distance, as a function of $\|Y_2 - Y_1\|$.
+The linearized approximation is in blue, the error from the covariance estimation is in green, and the error from the truncation of the distance expansion is in red.
 %
-(d) Empirical estimation of total distance as a function of $\Delta Y$}
+(d) Empirical estimation of total distance as a function of $\|Y_2 - Y_1\|$, with $\delta t =  10^{-10}$. The red line corresponds to the quadratic $\|X_2 - X_1 \|^2_{est} \propto \|Y_2 - Y_1 \|^2$.}
+\label{fig:empirical_estimation}
 \end{figure}
@@
 
-Show empirically what happens as we change the neighborhood (example from notes 1) (recovery of fast variable deep in the spectrum)
+The size of the ``burst'' we take also affects the recovery of the fast variable, even when there is no nonlinear transformation.
+%
+When the time scale of the burst is smaller than that of the equilibration time of the fast variable, the estimated covariance is constant and the fast variable is collapsed significantly relative to the slow variable.
+%
+This means that the fast variable is recovered {\em very} far down in the DMAPS eigenvectors.
+%
+However, if the time scale of the burst is {\em longer} than the saturation time of the fast variable, the estimated covariance changes: the variance in the slow direction continues to grow, while the variance in the fast direction is fixed.
+%
+This means that the collapse of the fast variable is less pronounced relative to the slow variable, and the fast variable is recovered in a higher eigenvector.
+
+We again consider the SDE system from \eqref{eq:init_data}, and consider the transformation
+\begin{eqnarray}\label{eq:transformed_data2}
+h^1(X) &=& X(1)  \\
+h^2(X) &=& \frac{X(2)}{\sqrt{\epsilon}}
+\end{eqnarray}
+%
+Unlike previously, there is no nonlinear component to the transformation, only a rescaling that makes the noise in the ``fast'' direction larger.
+
+Figure~\ref{fig:vary_burst} shows data collected from simulation of \eqref{eq:transformed_data2}.
+%
+Different timescales were used to simulate the bursts and estimate the local covariance. 
+%
+We can see that, when the timescale is less than the saturation time of the fast variable (the first two plots), the fast variable is recovered deep in the eigenvalue spectrum (eigenvector 14).
+%
+However when the timescale exceeds the saturation time of the fast variable (the third plot), the fast variable moves up in the eigenvalue spectrum (eigenvector 8).
+
+\begin{figure}[h]
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=\textwidth]{dat_varyburst}
+%\caption{}
+%\end{subfigure}
+%
+\begin{subfigure}{0.3\textwidth}
+\includegraphics[width=\textwidth]{data_withburst_1}
+\includegraphics[width=\textwidth]{fast_var_corr_1}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.3\textwidth}
+\includegraphics[width=\textwidth]{data_withburst_2}
+\includegraphics[width=\textwidth]{fast_var_corr_2}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.3\textwidth}
+\includegraphics[width=\textwidth]{data_withburst_3}
+\includegraphics[width=\textwidth]{fast_var_corr_3}
+\caption{}
+\end{subfigure}
+\caption{
+(a) Top: data in blue, simulated from \eqref{eq:init_data} and \eqref{eq:transformed_data2}, with representative burst in red.  
+%
+Bottom: correlation between NIV coordinate and fast variable. 
+%
+(b) Top: data in blue, with representative burst in red.  
+%
+Bottom: correlation between NIV coordinate and fast variable. 
+%
+(c) Top: data in blue, with representative burst in red.  
+%
+Bottom: correlation between NIV coordinate and fast variable. }
+\label{fig:vary_burst}
+\end{figure}
 
@@
 
-Richardson extrapolation allows us to get an estimate of a second-order term in the covariance estimation, thereby locally approximating the function using a quadratic form, rather than a linear form,  which can lead to a better/improved/more accurate ``Mahalanobis'' metric.
+Richardson extrapolation could allow us to get an estimate of a second-order term in the covariance estimation, thereby locally approximating the function using a quadratic form, rather than a linear form,  which can lead to a better/improved/more accurate ``Mahalanobis'' metric.
 

### 2014-07-21 reducible_SDE/paper/test_compare_NIV_DMAPS.m

@@ -25,7 +25,7 @@ rng(123);
 
 
 make_fig;
-scatter(data_init(:,1),data_init(:,2),50,t,'.')
+scatter(data_init(:,1),data_init(:,2),50,t*dt,'.')
 xlabel('X(1)')
 ylabel('X(2)')
 h = colorbar('peer',gca);
@@ -35,7 +35,7 @@ print(gcf, '-depsc', 'data_init');
 
 
 make_fig;
-scatter(data1(:,1),data1(:,2),50,t,'.')
+scatter(data1(:,1),data1(:,2),50,t*dt,'.')
 xlabel('Y(1)')
 ylabel('Y(2)')
 h = colorbar('peer',gca);

### 2014-07-21 reducible_SDE/paper/test_quadratic_empirical.m

@@ -55,7 +55,7 @@ rng(123);
 % nsteps = length(idx);
 
 make_fig;
-scatter(data_init(:,1),data_init(:,2),50,t,'.')
+scatter(data_init(:,1),data_init(:,2),50,t*dt,'.')
 xlabel('X(1)')
 ylabel('X(2)')
 h = colorbar('peer',gca);
@@ -63,7 +63,7 @@ set(get(h,'xlabel'),'String', 't');
 axis equal
 
 make_fig;
-scatter(data1(:,1),data1(:,2),50,t,'.')
+scatter(data1(:,1),data1(:,2),50,t*dt,'.')
 xlabel('Y(1)')
 ylabel('Y(2)')
 h = colorbar('peer',gca);
@@ -185,7 +185,7 @@ rng(123);
 
 
 make_fig;
-scatter(data_init(:,1),data_init(:,2),50,t,'.')
+scatter(data_init(:,1),data_init(:,2),50,t*dt,'.')
 xlabel('X(1)')
 ylabel('X(2)')
 h = colorbar('peer',gca);
@@ -194,7 +194,7 @@ axis equal
 print(gcf, '-depsc', 'original_data');
 
 make_fig;
-scatter(data1(:,1),data1(:,2),50,t,'.')
+scatter(data1(:,1),data1(:,2),50,t*dt,'.')
 xlabel('Y(1)')
 ylabel('Y(2)')
 h = colorbar('peer',gca);

### 2014-07-21 reducible_SDE/paper/test_varying_burst.m

@@ -7,98 +7,75 @@ close all
 dim = 2;
 
 epsilon = 1e-3;
+eps_dmaps = 10;
+dt = 1e-3;
 
+nsteps = 3000;
+nsteps_burst = 100;
+
+%% get initial data
+
+rng(123);
+[data_init, ~, data1, ~, t] = simulate_quadratic_linear(nsteps, dt, 1, dt, epsilon);
+
+
+make_fig;
+scatter(data1(:,1),data1(:,2),50,t*dt,'.')
+xlabel('Y(1)')
+ylabel('Y(2)')
+h = colorbar('peer',gca);
+set(get(h,'xlabel'),'String', 't');
+axis equal
+print(gcf, '-depsc', 'dat_varyburst');
 
 %%
 
-for dt_burst=[1e-9 1e-4 1e-2]
-    
-    eps_dmaps = 1;
-    % dt_burst = 1e-9;
-    dt = 1e-3;
-    
-    
-    nsteps = 3000;
-    nsteps_burst = 100;
-    
+fig_idx = 1;
+
+for dt_burst=[1e-8 1e-4 1e-2]
     
     rng(123);
-    [data_init, data_burst_init, data1, data1_burst, t] = simulate_quadratic_linear(nsteps, dt, nsteps_burst, dt_burst, epsilon);
+    [~, data_burst_init, data1, data1_burst, ~] = simulate_quadratic_linear(nsteps, dt, nsteps_burst, dt_burst, epsilon, data_init);
     
-%     
-%     make_fig;
-%     scatter(data_init(:,1),data_init(:,2),50,t,'.')
-%     xlabel('X(1)')
-%     ylabel('X(2)')
-%     h = colorbar('peer',gca);
-%     set(get(h,'xlabel'),'String', 't');
-%     axis equal
-    
-%     make_fig;
-%     scatter(data1(:,1),data1(:,2),50,t,'.')
-%     xlabel('Y(1)')
-%     ylabel('Y(2)')
-%     h = colorbar('peer',gca);
-%     set(get(h,'xlabel'),'String', 't');
-%     axis equal
-%     
     make_fig;
-%     scatter(data1(:,1),data1(:,2),50,t,'.')
-    plot(data1(:,1),data1(:,2),'.r')
+    %     scatter(data1(:,1),data1(:,2),50,t,'.')
+    plot(data1(:,1),data1(:,2),'.b')
     hold on
-    plot(data1_burst(:,1,1000),data1_burst(:,2,1000),'.k')
[…]

### 2014-07-23 chemotaxis/prl_paper.tex
added small edits to chemotaxis paper

@@
 %
-We demonstrate how extracting macroscopic dynamics from microscale data can be accomplished automatically using manifold learning techniques. 
+We demonstrate how extracting a parameterization of microscale data can be accomplished automatically using manifold learning techniques, and that this parameterization is consistent with the governing macroscopic dynamical model.
 %
@@
 %
-We would like to {\em automatically} uncover a system's macroscale behavior from data collected at the microscale without any {\em a priori} knowledge of the appropriate microscopic or macroscopic model.
+We would like to extract a parameterization of data collected at the microscale which is consistent with the macroscopic behavior, without any {\em a priori} knowledge of the appropriate microscopic or macroscopic model.
 
@@
 %
+We would like to note that we have chosen a one-parameter family of initial conditions.
+%
+Other initial conditions could easily be chosen, but the simplicity of this setup will allow us to easily illustrate our main points.
 

### 2014-07-29 chemotaxis/prl_paper.tex
added small edits to chemotaxis paper

@@
 %
-We demonstrate how extracting a parameterization of microscale data can be accomplished automatically using manifold learning techniques, and that this parameterization is consistent with the governing macroscopic dynamical model.
+We demonstrate automatically extracting a parameterization of microscale data using manifold learning techniques, and show that this parameterization is consistent with the governing macroscopic dynamical model.
 %
-These data-driven algorithms need no {\em a priori} knowledge about the appropriate macroscopic model, but instead, only require the appropriate observers and pairwise metric between observations in order to recover the governing macroscopic variables.
+These data-driven algorithms need no {\em a priori} knowledge about the appropriate macroscopic model, but instead, only require the appropriate observers and pairwise metric between observations.
 %
@@
 
-In this letter, we will present how this goal can be achieved through a data-driven method based on geometric analysis and manifold learning. 
+In this letter, we will demonstrate how this goal can be achieved through a data-driven method based on geometric analysis and manifold learning. 
 %
@@
 %
-We will show that manifold learning algorithms based on statistical observers and with the appropriate affinity metric between pairs of observations uncovers a description of the microscopic data which is consistent with the macroscopic model.
+We will show that manifold learning algorithms based on statistical observers and with the appropriate affinity metric between pairs of observations uncovers a parameterization of the microscopic data which is consistent with the macroscopic model.
 %
@@
 %
-Let $x_i(t)$ and $v_i(t)$ denote the position and velocity of particle $i$ at time $t$, respectively.
+Let $x_i(t)$ and $v_i(t)$ denote the position and velocity, respectively, of particle $i$ at time $t$.
 %
@@
 \end{equation}
-where $p$ is an initial probability.
+where $0 < p < 1$ is the probability of a particle initially moving to the right.
 %
@@
 %
-We would like to note that we have chosen a one-parameter family of initial conditions.
+We would like to note that we have chosen a very specific one-parameter family of initial conditions, which lead to simple dynamics and straightforward illustration of our main points.
 %
-Other initial conditions could easily be chosen, but the simplicity of this setup will allow us to easily illustrate our main points.
+More complex initial conditions could easily be chosen.
 
@@
 This implies that for fixed $\lambda$ and $s$, the {\em microscopic} data in a high-dimensional ambient space (e.g., the positions of all $N$ particles) should lie on a two-dimensional manifold parameterized by $p$ and $t$,
-and uncovering the low-dimensional structure of the microscopic data reveals the governing macroscopic variables.
+and uncovering the low-dimensional structure of the microscopic data reveals this manifold.
 
@@
 %
-However, similarly to the chemotaxis model, we expect the simulation data to lie on a low-dimensional manifold which is parameterized by the macroscopic variables.
+We expect the simulation data to lie on a low-dimensional manifold which is parameterized by the macroscopic variables.
 %
@@
 %
-We showed that, using histograms as observers and EMD as a distance metric, diffusion maps can uncover the underlying macroscopic dynamical behavior which is consistent with the analytic macroscopic model.
+We showed that, using histograms as observers and EMD as a distance metric, diffusion maps can uncover a parameterization of the microscale data which is consistent with the analytic macroscopic model.
 %

### 2014-08-05 reducible_SDE/paper/paper_draft.tex
edited multiscale paper

@@
 \ead{ronen.talmon@yale.edu}
+
+\author[PrincetonCBE]{C.W.~Gear}
+\ead{wgear@princeton.edu}
 
@@
 
+We would like to note that the ratio of the drift and diffusion terms in the $y$ equation is essential.
+%
+We need the square of the diffusivity to be of the same order as the drift.
+%
+If the diffusivity is too large, the equilibrium measure of $y$ will be unbounded.
+%
+Conversely, if the diffusivity is too small, the equilibrium measure of $y$ will go to 0.
+
 %TODO: should we write the SDE in a more general form?
@@
 
+These same conditions for uncoupled noises with unit variance will allow us to recover the slow variable(s) from data with multiple timescales. 
+%
 We want to emphasize the link between multiscale SDEs and NIVs: NIVs rescales the noises, so that the ``fast'' variables become small.
@@
 We can find such a regime empirically (by plotting $\| X_2 - X_1 \|^2$ as a function of $\delta t$ and $\|Y_2 - Y_1\|$), and select a kernel distance where both conditions hold.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{schematic}
+\caption{Illustration of how to choose $\delta t$ and $\sigma_{kernel}$ appropriately.
+The data shows the evolution of the ``fast'' variable at a fixed value of the ``slow'' variable.
+We must choose both parameters so that the curvature effects and other nonlinearities are negligable.
+}
+\label{fig:schematic}
+\end{figure}
 
@@
 %
-(d) Empirical estimation of total distance as a function of $\|Y_2 - Y_1\|$, with $\delta t =  10^{-10}$. The red line corresponds to the quadratic $\|X_2 - X_1 \|^2_{est} \propto \|Y_2 - Y_1 \|^2$.}
+(d) Empirical estimation of total distance as a function of $\|Y_2 - Y_1\|$, with $\delta t =  10^{-9}$. The red line corresponds to the quadratic $\|X_2 - X_1 \|^2_{est} \propto \|Y_2 - Y_1 \|^2$.}
 \label{fig:empirical_estimation}
+\end{figure}
+
+\begin{figure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{changing_parameters_error_terms_1}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{changing_parameters_NIV_corr_1}
+\caption{}
+\end{subfigure}
+
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{changing_parameters_error_terms_2}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{changing_parameters_NIV_corr_2}
+\caption{}
+\end{subfigure}
+
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{changing_parameters_error_terms_3}
+\caption{}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{changing_parameters_NIV_corr_3}
+\caption{}
+\end{subfigure}
+
+\caption{ Error analysis and recovery of the slow variable for three different simulation configurations.
+(a, b) $\sigma_{kernel} = 10^{-2}$, $\delta t = 10^{-9}$, (c, d) $\sigma_{kernel} = 10^{2}$, $\delta t = 10^{-9}$ (e, f) $\sigma_{kernel} = 10^{-2}$, $\delta t = 10^{-5}$.
+(a, c, e) The three separate terms in the estimated distance, as a function of $\|Y_2 - Y_1\|$.
+The linearized approximation is in blue, the error from the covariance estimation is in green, and the error from the truncation of the distance expansion is in red.
+The blue circle indicates the value of $\sigma_{kernel}$ used in the diffusion maps calculations.
+(b, d, f) The first NIV, obtained from the transformed data $Y$, versus the slow variable $X(1)$.
+Note that we only recover the slow variable $X(1)$ when the linearized approximation dominates the other two error terms in the distance function at the scale of $\sigma_{kernel}$.
+}
+
 \end{figure}

### 2014-08-05 reducible_SDE/paper/test_changing_parameters_compare.m

@@ -5,8 +5,8 @@ close all
 %% define parameters
 
 epsilon = 1e-3;
-eps_dmaps_all = [0.01 10 0.01 0.01 0.01];
-dt_burst_all = [1e-8 1e-8 1e-5 1e-10 1e-5];
+eps_dmaps_all = [0.01 10 0.01];
+dt_burst_all = [1e-9 1e-9 1e-5];
 
 dim = 2;
 
@@ -15,13 +15,14 @@ dt = 1e-4;
 nsteps = 3000;
 nsteps_burst = 200;
 
+dy_axis_lim = [1e-2 1e1 1e-10 1e5];
+
 
 for i = 1:length(eps_dmaps_all)
     
     fig_idx = i;
     eps_dmaps = eps_dmaps_all(i);
     dt_burst = dt_burst_all(i);
-    
     
     make_fig;
     deltas = logspace(-3, 1, 100);
@@ -31,106 +32,53 @@ for i = 1:length(eps_dmaps_all)
     loglog(deltas, 10*deltas.^4, '-r')
     loglog(sqrt(eps_dmaps/(0.5*(2 + epsilon + sqrt(9 + 2 * epsilon + epsilon^2)))), eps_dmaps, 'ok')
     % legend('linear approximation distance','error from covariance estimation','error from Taylor expansion','location','best')
-    xlabel('$\Delta Y$','interpreter','latex')
-    % print(gcf, '-depsc', sprintf('error_terms_%d', fig_idx));
-
-
-%% simulate SDE
-
-rng(123);
-[data_init, data_burst_init, data1, data1_burst, t] = simulate_quadratic(nsteps, dt, nsteps_burst, dt_burst, epsilon);
-
-% idx = find(data_init(:, 2).^2 < epsilon/2);
-% data_init = data_init(idx, :);
-% t = t(idx, :);
-% nsteps = length(idx);
-
-% make_fig;
-% scatter(data_init(:,1),data_init(:,2),50,t,'.')
-% xlabel('X(1)')
-% ylabel('X(2)')
-% h = colorbar('peer',gca);
-% set(get(h,'xlabel'),'String', 't');
-% axis equal
-% % print(gcf, '-depsc', sprintf('orig_data_%d', fig_idx));
-% 
-% make_fig;
-% scatter(data1(:,1),data1(:,2),50,t,'.')
-% xlabel('Y(1)')
-% ylabel('Y(2)')
-% h = colorbar('peer',gca);
[…]

### 2014-08-05 reducible_SDE/paper/test_quadratic_empirical.m

@@ -10,7 +10,7 @@ epsilon = 1e-3;
 
 
 dt = 1e-4;
-dt_burst = 1e-10;
+dt_burst = 1e-9;
 
 nsteps = 3000;
 nsteps_burst = 200;

### 2014-08-14 reducible_SDE/paper/SIADS_draft.tex
added siads draft

--- 2014-07-21 reducible_SDE/paper/paper_draft.tex
@@
-\documentclass[1p]{elsarticle}
+%%
+%% This is file `docultexmm.tex', 
+%% Documentation for siam multimedia macros for use with LaTeX 2e
+%% 
+%% December 19, 2013
+%%
+%% Version 1.0.1
+%% 
+%% You are not allowed to change this file. 
+%% 
+%% You are allowed to distribute this file under the condition that 
+%% it is distributed together with all of the files in the siam macro 
+%% distribution. These are:
+%%
+%%  siamltexmm.cls (this file)
+%%  siam11.clo   (required size option for 11pt papers)
+%%  subeqn.clo   (allows equation numbers with lettered subelements)
+%%  siam.bst     (bibliographic style file for BibTeX)
+%%  docultexmm.tex (documentation file)
+%%
+%% If you receive only some of these files from someone, please contact: 
+%% multimedia@siam.org  
+%% 
+%% You are not allowed to distribute this file alone. You are not 
+%% allowed to take money for the distribution or use of either this 
+%% file or a changed version, except for a nominal charge for copying 
+%% etc.
+%%
+%% \CharacterTable
+%%  {Upper-case    \A\B\C\D\E\F\G\H\I\J\K\L\M\N\O\P\Q\R\S\T\U\V\W\X\Y\Z
+%%   Lower-case    \a\b\c\d\e\f\g\h\i\j\k\l\m\n\o\p\q\r\s\t\u\v\w\x\y\z
+%%   Digits        \0\1\2\3\4\5\6\7\8\9
+%%   Exclamation   \!     Double quote  \"     Hash (number) \#
+%%   Dollar        \$     Percent       \%     Ampersand     \&
+%%   Acute accent  \'     Left paren    \(     Right paren   \)
+%%   Asterisk      \*     Plus          \+     Comma         \,
+%%   Minus         \-     Point         \.     Solidus       \/
+%%   Colon         \:     Semicolon     \;     Less than     \<
+%%   Equals        \=     Greater than  \>     Question mark \?
+%%   Commercial at \@     Left bracket  \[     Backslash     \\
+%%   Right bracket \]     Circumflex    \^     Underscore    \_
+%%   Grave accent  \`     Left brace    \{     Vertical bar  \|
+%%   Right brace   \}     Tilde         \~}
+
+\documentclass[final,leqno,onefignum,onetabnum]{siamltexmm}
 
@@
 
-\usepackage{epstopdf}
+\title{Data-driven Reduction of Multiscale Stochastic Dynamical Systems \thanks{This work was
+supported by ....}}
 
-\title{Data-driven Reduction of Multiscale Stochastic Dynamical Systems }
-
-\author[PrincetonCBE]{C.J.~Dsilva}
-\ead{cdsilva@princeton.edu}
-
-\author[YaleMath]{R.~Talmon}
-\ead{ronen.talmon@yale.edu}
-
-\author[YaleMath]{R.R.~Coifman}
-\ead{coifman@math.yale.edu}
-
-\author[PrincetonCBE, PrincetonPACM]{I.G.~Kevrekidis\corref{cor1}}
-\ead{yannis@princeton.edu}
-
-\address[PrincetonCBE]{Department of Chemical and Biological Engineering, Princton University, Princeton, New Jersey, 08544, USA}
-\address[YaleMath]{Department of Mathematics, Yale University, New Haven, Connecticut, 06520, USA}
-\address[PrincetonPACM]{Program in Applied and Computational Mathematics, Princton University, Princeton, New Jersey, 08544, USA}
-
-\cortext[cor1]{Corresponding author}
+\author{TeX Production\thanks{Society for Industrial and
+Applied Mathematics, Philadelphia, Pennsylvania. 
+(\email{multimedia@siam.org}). Questions, comments, or corrections
+to this document may be directed to that email address.}}
 
 \begin{document}
+\maketitle
+\newcommand{\slugmaster}{%
+\slugger{siads}{xxxx}{xx}{x}{x--x}}%slugger should be set to juq, siads, sifin, or siims
 
@@
 
-\begin{itemize}
-
-\item Link between data mining and dynamical systems analysis
-
-\item Extracting/reducing an SDE system to its slow variables using the Mahalanobis graph and NIVs
-
-\item Full mathematical analysis of the slow variable reduction procedure, including necessary conditions for accurate recovery of the intrinsic variables and a framework for empirically setting the algorithm parameters
-
-\end{itemize}
-
 \end{abstract}
 
-\begin{keyword}
+\begin{keywords}\end{keywords}
 
-\end{keyword}
+\begin{AMS}\end{AMS}
 
-\maketitle
+\pagestyle{myheadings}
+\thispagestyle{plain}
+\markboth{TEX PRODUCTION}{USING SIAM'S MM \LaTeX\ MACROS}
 
 \section{Introduction}
-
-\section{Theory}
 
@@
 \begin{aligned}
-dx &= adt + dW^1\\
-dy &= -\frac{y}{\epsilon} dt + \frac{1}{\sqrt{\epsilon}} dW^2
+dx &= adt + dW_1\\
+dy &= -\frac{y}{\epsilon} dt + \frac{1}{\sqrt{\epsilon}} dW_2
 \end{aligned}
 \end{equation}
+TODO: add more general formulation of the multi scale SDE
 
@@
 
-%TODO: should we write the SDE in a more general form?
+We would like to note that the ratio of the drift and diffusion terms in the $y$ equation is essential.
+%
+We need the square of the diffusivity to be of the same order as the drift.
+%
+If the diffusivity is too large, the equilibrium measure of $y$ will be unbounded.
+%
+Conversely, if the diffusivity is too small, the equilibrium measure of $y$ will go to 0.
 
-\subsection{Nonlinear Intrinsic Variables}
+\subsection{Nonlinear functions}
 
-The goal of NIVs is to recover ``intrinsic'' variables that are independent of the measurement/observation function.
+We will also consider a (nonlinear) function $\mathbf{f}: \mathbb{R}^n \mapsto \mathbb{R}^m$ of the data sampled from the SDE in \eqref{eq:fast_slow_SDE}.
+
+We will consider two specific examples. 
+\begin{enumerate}
+\item fat strip
+\item half moon
+\end{enumerate}
 %
-We define intrinsic variables as those variables which have uncoupled noise with unit variance,
-i.e., the intrinsic variables $X^1, \dots, X^n$ are governed by the following SDEs:
-\begin{equation} \label{eq:NIV_formulation}
-dX^i = a(X) dt + dW^i
-\end{equation}
-where $X = \begin{bmatrix} X^1 & X^2 & \cdots & X^n \end{bmatrix}^T$.
+We will show in both cases how we can recover the slow variables using data-driven techniques.
 
+\section{Data mining: Nonlinear Intrinsic Variables}
+
+%These same conditions for uncoupled noises with unit variance will allow us to recover the slow variable(s) from data with multiple timescales. 
+%
 We want to emphasize the link between multiscale SDEs and NIVs: NIVs rescales the noises, so that the ``fast'' variables become small.
@@
 
-\subsection{Errors and Estimation}
+We will define $X = \begin{bmatrix} x \\ z \end{bmatrix}$
 
+To get a parameterization of NIV, we require $\| X_2 - X_1\|_2$.
+%
+However, we only have access to $Y = f(X)$.
+%
+Assuming that we have a local approximation to $\| X_2 - X_1\|$ from the measurements $Y$, we can then use DMAPS to uncover a parameterization of $X$ from the estimates of the pairwise distances.
+
+Description of DMAPS........
+
+Our approximation of the pairwise distances will rely on the Taylor expansion of the measurement function $f$. 
+%
+We will then {\em empirically} estimate the first-order term in this expansion using simulation bursts to estimate the local covariance.  
+%
+Accordingly, we will address the accuracy of our method from two standpoints: (1) the accuracy of the Taylor expansion, and (2) the accuracy of the covariance estimation (see Figure...).
+%
+We will present both analytical results for the error bounds, as well as an empirical methodology to set the appropriate parameters for our method to accurately recover the intrinsic slow variable(s).
+
+TODO: change ``curvature'' in figure to ``nonlinearity'' in $\sigma_{kernel}$
+TODO: emphasize that if $\delta t$ is too large, then data does not look Gaussian
+TODO: change strip to look different than other data
+%\begin{figure}
+%\includegraphics[width=\textwidth]{schematic}
+%\caption{Illustration of how to choose $\delta t$ and $\sigma_{kernel}$ appropriately.
+The data shows the evolution of the ``fast'' variable at a fixed value of the ``slow'' variable.
+We must choose both parameters so that the curvature effects and other nonlinearities are negligable.
+}
+%\label{fig:schematic}
+%\end{figure}
+
+\subsection{Error analysis of the Mahalanobis distance}
+
+Let $g = f^{-1}: \mathbb{R}^m \mapsto \mathbb{R}^n$, and let $Y_1 = f(X_1)$, and $Y_2 = f(X_2)$.
+%
+Again, we let $g_j^i = \frac{\partial g^i}{\partial Y^j}$.
+%
+Then, by Taylor expansion, we obtain
+\begin{eqnarray}
+X_2^i &=& X_1^i + \sum_j g_j^i (Y_1) (Y^j_2 - Y^j_1 ) 
++ \frac{1}{2} \sum_{kl}  g^i_{kl} (Y_1) (Y^k_2 - Y^k_1)(Y^l_2 - Y^l_1) \\
+&&+ \frac{1}{6} \sum_{klm}  g^i_{klm} (Y_1) (Y^k_2 - Y^k_1)(Y^l_2 - Y^l_1) (Y^m_2 - Y^m_1) 
++ \mathcal{O}( \|Y_2 - Y_1\|^4 )
+\end{eqnarray}
+
+Subsitituting the Taylor expansion into the quadratic form, we obtain
+\begin{eqnarray}
+&&\| X_2 - X_1 \|^2 \\
+&=& \frac{1}{2} (Y_2 - Y_1 )^T ((J J^T)^{-1} (Y_1) + (J J^T)^{-1}(Y_2)) (Y_2 - Y_1 ) \\
+&& + \frac{1}{2} \sum_{ijkl} \left( g_j^i (Y_1) g^i_{kl} (Y_1) - g_j^i (Y_2) g^i_{kl} (Y_2) \right) (Y^j_2 - Y^j_1 )  (Y^k_2 - Y^k_1)(Y^l_2 - Y^l_1) \\
+&& + \frac{1}{8} \sum_{ijklm}  \left( g^i_{jk} (Y_1) g^i_{lm} (Y_1) + g^i_{jk} (Y_2) g^i_{lm} (Y_2)  \right) (Y^j_2 - Y^j_1) (Y^k_2 - Y^k_1) (Y^l_2 - Y^l_1) (Y^m_2 - Y^m_1) \\
+&& + \frac{1}{6} \sum_{ijklm}  \left( g^i_{j} (Y_1) g^i_{klm} (Y_1) + g^i_{j} (Y_2) g^i_{klm} (Y_2)  \right)(Y^j_2 - Y^j_1) (Y^k_2 - Y^k_1) (Y^l_2 - Y^l_1) (Y^m_2 - Y^m_1) \\
+&& + \mathcal{O} (\|Y_1 - Y_2 \|^6 ) 
+\end{eqnarray}
+
+In general, we do not have access to $f$, $g$, or any of its derivatives.
+%
+However, we note that $J J^T(Y_1)$ is the local covariance of $Y_1$, which we can empirically estimate from data (this will be discussed in a subsequent section).
+%
+Therefore, we choose to truncate the distance approximation at the first order term.
+%
+We call this distance the {\em Mahalanobis distance}, 
+\begin{equation}
+ \| Y_2 - Y_1 \|^2_M = \frac{1}{2} (Y_2 - Y_1 )^T ((J J^T)^{-1} (Y_1) + (J J^T)^{-1}(Y_2)) (Y_2 - Y_1 ) 
+\end{equation}
+%
+The error in the approximation is defined as
+\begin{equation}
+e_M(Y_1, Y_2) = \| X_2 - X_1 \|^2_2 - \| Y_2 - Y_1 \|^2_M 
+\end{equation}
+
+We can bound the error by
+\begin{equation}
+| e_M(Y_1, Y_2)  | \le n \left( \left| K_1 K_2 \right| + \left| \frac{ K_2^2}{4} \right|  + \left| \frac{K_1 K_3}{3} \right|  \right) \| Y_2 - Y_1 \| ^4  
++ \mathcal{O} (\|Y_1 - Y_2 \|^6 ) 
+\end{equation}
+%
+where
+%
+\begin{equation}
+\begin{aligned}
+K_1 &= \sup_{i,j,Y} |g_j^i(Y)|\\
+K_2 &= \sup_{i,j,k,Y} |g_{jk}^i(Y)|\\
+K_3 &= \sup_{i,j,k,l,Y} |g_{jkl}^i(Y)|
+\end{aligned}
+\end{equation}
+
+\subsection{Error analysis of the covariance estimation}
+
+To compute the Mahalanobis distance, we require $J J^T$.
+%
+By Ito formula, we have
+\begin{equation}
+dY^j = \sum_{i=1}^n \left( \frac{1}{2} f_{ii}^j + a_i f_i^j \right) dt + \sum_{i=1}^n f_i^j dW_i
+\end{equation}
+%
+TODO: define partials...
+%
+The covariance $C$ defined by 
+\begin{equation}
+[C]_{jk} = \mathrm{Cov} \left( dY^j, dY^k \right)
+\end{equation}
+equals $C = J J^T$ \cite{...}.
+
+We will empirically estimate the covariance from data.
+
+Comment: Use a more general model (not necessarily multi scale)
+
+The goal of NIVs is to recover ``intrinsic'' variables that are independent of the measurement/observation function.
+%
+We define intrinsic variables as those variables which have uncoupled noise with unit variance,
+i.e., the intrinsic variables $X^1, \dots, X^n$ are governed by the following SDEs:
+\begin{equation} \label{eq:NIV_formulation}
+dX^i = a(X) dt + dW^i
+\end{equation}
+where $X = \begin{bmatrix} X^1 & X^2 & \cdots & X^n \end{bmatrix}^T$.
+
+Restrictions:
 \begin{itemize}
-\item Error and estimation analysis of NIVs (remark that applies to more general settings): give conditions/bounds when we can recover the intrinsic slow variable
-%
-\item Empirical configuration/demonstration of the method, how to use the analysis to set the parameters in the algorithm
+\item Drift is bilipshitz wrt process $X$
+\item Access to simulator, BUT the analysis could be extended to a single trajectory (cite people....); emphasize the difference between simulation and experiment
 \end{itemize}
+
+We assume that
+\begin{equation}
+ \int_{t}^{t + \delta t} a(X_\tau) d \tau - a(X_t) \delta t  = \mathcal{O} (\delta t) \: \forall t
+\end{equation}
+TODO: for the toy problems, find the order of this term, possibly as a function of $\epsilon$-- then we can relate $\epsilon$ to $\delta t$, and place bounds on $\delta t$
+
+\begin{equation}
+\begin{aligned}
+X(t + \delta t) =& X_t + \int_{t}^{t+\delta t} a(X_\tau) d \tau + d W \\
+= & X_t + \int_{t}^{t+\delta t} \left[ a(X_t) + \left( a(X_\tau) - a(X_t) \right) \right] d \tau + d W \\
+= & X_t + a(X_t) \delta t  + \sqrt{\delta t} W + \mathcal{O} (\delta t)
+\end{aligned}
+\end{equation}
 
@@
 \end{equation}
+where 
+$$X_0 = X_t + \left( a(X_t) + \mathcal{O}(1) \right) \delta t $$
+%
+Note: we bound the random variation due to the variations in the drift, and note that this bound makes these variations smaller than the variations due to the local noise/diffusion. 
+%
+Therefore, we simply absorb these variations into an $\mathcal{O}(\delta t)$ correction of the mean of the sample distribution, turning a random term into a deterministic term.
 
-We now consider a function $\mathbf{f}: \mathbb{R}^n \mapsto \mathbb{R}^m$.
-%
 Let $f^i(X)$ denote the $i^{th}$ entry of $\mathbf{f}$, and let $f^i_j(X) = \frac{\partial f^i}{\partial X^j}$.
@@
 
-We now approximate the distances $\| X_2 - X_1 \|$.
-%
-Let $g = f^{-1}: \mathbb{R}^m \mapsto \mathbb{R}^n$, and let $Y_1 = f(X_1)$, and $Y_2 = f(X_2)$.
-%
-Again, we let $g_j^i = \frac{\partial g^i}{\partial Y^j}$.
-%
-Then, by Taylor expansion, we obtain
-\begin{equation} \label{eq:taylor_expansion}
-\begin{aligned}
-&\| X_2 - X_1 \|^2 \\
-\le&  \left| \frac{1}{2} (Y_2 - Y_1 )^T ((J J^T)^{-1} (Y_1) + (J J^T)^{-1}(Y_2)) (Y_2 - Y_1 ) \right| \\
-& + n \left( \left| K_1 K_2 \right| + \left| \frac{ K_2^2}{4} \right|  + \left| \frac{K_1 K_3}{3} \right|  \right) \| Y_2 - Y_1 \| ^4  \\
-& + \mathcal{O} (\|Y_1 - Y_2 \|^6 ) 
-\end{aligned}
-\end{equation}
-%
-where
-%
-\begin{equation}
-\begin{aligned}
-K_1 &= \sup_{i,j,Y} |g_j^i(Y)|\\
-K_2 &= \sup_{i,j,k,Y} |g_{jk}^i(Y)|\\
-K_3 &= \sup_{i,j,k,l,Y} |g_{jkl}^i(Y)|
-\end{aligned}
-\end{equation}
+TODO: define error function for covariance estimation
 
@@
 \section{Results}
+
+\begin{itemize}
+
+\item Start with strip: show magnitude of terms, but not figures for terms, because everything will be constant/zero
+\item Only error in covariance of strip will be from drift (verify order 1) 
+\item Then, show half moons; write error terms explicitly
+\item Add drift term to covariance estimation
+\end{itemize}
+
+TODO: Also add second example (just strip, no nonlinear transform)
 
@@
 
-\begin{figure}[h]
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{data_init}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{data_transformed}
-\caption{}
-\end{subfigure}
-\caption{(a) The original data, simulated from \eqref{eq:init_data} with $\epsilon = 10^{-3}$.
+%\begin{figure}[h]
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=\textwidth]{data_init}
+%\caption{}
+%\end{subfigure}
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=\textwidth]{data_transformed}
+%\caption{}
+%\end{subfigure}
+%\caption{(a) The original data, simulated from \eqref{eq:init_data} with $\epsilon = 10^{-3}$.
 We take $3000$ timesteps with $dt = 10^{-4}$ (b) The data from (a), transformed according to \eqref{eq:transformed_data}.
 }
-\label{fig:initial_data}
-\end{figure}
+%\label{fig:initial_data}
+%\end{figure}
 
@@
 
-\begin{figure}[h]
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{data_transformed_colored_NIV}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{data_transformed_colored_DMAPS}
-\caption{}
-\end{subfigure}
-\caption{(a) The data from Figure~\ref{fig:initial_data}(b), colored by the first NIV (we take $\sigma_{kernel} = 0.01$, and $\delta t = 10^{-9}$).
+%\begin{figure}[h]
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=\textwidth]{data_transformed_colored_NIV}
+%\caption{}
+%\end{subfigure}
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=\textwidth]{data_transformed_colored_DMAPS}
+%\caption{}
+%\end{subfigure}
+%\caption{(a) The data from Figure~\ref{fig:initial_data}(b), colored by the first NIV (we take $\sigma_{kernel} = 0.01$, and $\delta t = 10^{-9}$).
 Note that we recover the slow variable.
@@
 Note that we do {\em not} recover the slow variable.}
-\label{fig:NIV_versus_DMAPS}
-\end{figure}
+%\label{fig:NIV_versus_DMAPS}
+%\end{figure}
 
@@
 
-\begin{figure}[h]
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{errors_function_dt}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{empirical_totaldist_function_dt}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{errors_function_dy}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{empirical_totaldist_function_dy}
-\caption{}
-\end{subfigure}
-\caption{(a) Total estimated distance (black) and the three separate terms in the estimated distance, as a function of $\delta t$.
+%\begin{figure}[h]
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=\textwidth]{errors_function_dt}
+%\caption{}
+%\end{subfigure}
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=\textwidth]{empirical_totaldist_function_dt}
+%\caption{}
+%\end{subfigure}
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=\textwidth]{errors_function_dy}
+%\caption{}
+%\end{subfigure}
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=\textwidth]{empirical_totaldist_function_dy}
+%\caption{}
+%\end{subfigure}
+%\caption{(a) Total estimated distance (black) and the three separate terms in the estimated distance, as a function of $\delta t$.
 The linearized approximation is in blue, the error from the covariance estimation is in green, and the error from the truncation of the distance expansion is in red.
+%%
+%(b) Empirical estimation of total distance as a function of $\delta t$, with $\|Y_2 - Y_1 \|^2 = 0.01$.  
+%%
+%(c) Total estimated distance (black) and the three separate terms in the estimated distance, as a function of $\|Y_2 - Y_1\|$.
+The linearized approximation is in blue, the error from the covariance estimation is in green, and the error from the truncation of the distance expansion is in red.
+%%
+%(d) Empirical estimation of total distance as a function of $\|Y_2 - Y_1\|$, with $\delta t =  10^{-9}$. The red line corresponds to the quadratic $\|X_2 - X_1 \|^2_{est} \propto \|Y_2 - Y_1 \|^2$.}
+%\label{fig:empirical_estimation}
+%\end{figure}
 %
-(b) Empirical estimation of total distance as a function of $\delta t$, with $\|Y_2 - Y_1 \|^2 = 0.01$.  
+%\begin{figure}
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=\textwidth]{changing_parameters_error_terms_1}
+%\caption{}
+%\end{subfigure}
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=\textwidth]{changing_parameters_NIV_corr_1}
+%\caption{}
+%\end{subfigure}
 %
-(c) Total estimated distance (black) and the three separate terms in the estimated distance, as a function of $\|Y_2 - Y_1\|$.
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=\textwidth]{changing_parameters_error_terms_2}
+%\caption{}
+%\end{subfigure}
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=\textwidth]{changing_parameters_NIV_corr_2}
+%\caption{}
+%\end{subfigure}
+%
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=\textwidth]{changing_parameters_error_terms_3}
+%\caption{}
+%\end{subfigure}
+%\begin{subfigure}{0.5\textwidth}
+%\includegraphics[width=\textwidth]{changing_parameters_NIV_corr_3}
+%\caption{}
+%\end{subfigure}
+%
+%\caption{ Error analysis and recovery of the slow variable for three different simulation configurations.
+(a, b) $\sigma_{kernel} = 10^{-2}$, $\delta t = 10^{-9}$, (c, d) $\sigma_{kernel} = 10^{2}$, $\delta t = 10^{-9}$ (e, f) $\sigma_{kernel} = 10^{-2}$, $\delta t = 10^{-5}$.
+(a, c, e) The three separate terms in the estimated distance, as a function of $\|Y_2 - Y_1\|$.
 The linearized approximation is in blue, the error from the covariance estimation is in green, and the error from the truncation of the distance expansion is in red.
+The blue circle indicates the value of $\sigma_{kernel}$ used in the diffusion maps calculations.
+(b, d, f) The first NIV, obtained from the transformed data $Y$, versus the slow variable $X(1)$.
+Note that we only recover the slow variable $X(1)$ when the linearized approximation dominates the other two error terms in the distance function at the scale of $\sigma_{kernel}$.
+}
 %
-(d) Empirical estimation of total distance as a function of $\|Y_2 - Y_1\|$, with $\delta t =  10^{-10}$. The red line corresponds to the quadratic $\|X_2 - X_1 \|^2_{est} \propto \|Y_2 - Y_1 \|^2$.}
-\label{fig:empirical_estimation}
-\end{figure}
+%\end{figure}
 
@@
 
-\begin{figure}[h]
-%\begin{subfigure}{0.5\textwidth}
-%\includegraphics[width=\textwidth]{dat_varyburst}
+%\begin{figure}[h]
+%%\begin{subfigure}{0.5\textwidth}
+%%\includegraphics[width=\textwidth]{dat_varyburst}
+%%\caption{}
+%%\end{subfigure}
+%%
+%\begin{subfigure}{0.3\textwidth}
+%\includegraphics[width=\textwidth]{data_withburst_1}
+%\includegraphics[width=\textwidth]{fast_var_corr_1}
 %\caption{}
 %\end{subfigure}
-%
-\begin{subfigure}{0.3\textwidth}
-\includegraphics[width=\textwidth]{data_withburst_1}
-\includegraphics[width=\textwidth]{fast_var_corr_1}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.3\textwidth}
-\includegraphics[width=\textwidth]{data_withburst_2}
-\includegraphics[width=\textwidth]{fast_var_corr_2}
-\caption{}
-\end{subfigure}
-\begin{subfigure}{0.3\textwidth}
-\includegraphics[width=\textwidth]{data_withburst_3}
-\includegraphics[width=\textwidth]{fast_var_corr_3}
-\caption{}
-\end{subfigure}
-\caption{
-(a) Top: data in blue, simulated from \eqref{eq:init_data} and \eqref{eq:transformed_data2}, with representative burst in red.  
-%
-Bottom: correlation between NIV coordinate and fast variable. 
-%
-(b) Top: data in blue, with representative burst in red.  
-%
-Bottom: correlation between NIV coordinate and fast variable. 
-%
-(c) Top: data in blue, with representative burst in red.  
-%
-Bottom: correlation between NIV coordinate and fast variable. }
-\label{fig:vary_burst}
-\end{figure}
+%\begin{subfigure}{0.3\textwidth}
+%\includegraphics[width=\textwidth]{data_withburst_2}
+%\includegraphics[width=\textwidth]{fast_var_corr_2}
+%\caption{}
+%\end{subfigure}
+%\begin{subfigure}{0.3\textwidth}
+%\includegraphics[width=\textwidth]{data_withburst_3}
+%\includegraphics[width=\textwidth]{fast_var_corr_3}
+%\caption{}
+%\end{subfigure}
+%\caption{
+%(a) Top: data in blue, simulated from \eqref{eq:init_data} and \eqref{eq:transformed_data2}, with representative burst in red.  
+%%
+%Bottom: correlation between NIV coordinate and fast variable. 
+%%
+%(b) Top: data in blue, with representative burst in red.  
+%%
+%Bottom: correlation between NIV coordinate and fast variable. 
+%%
+%(c) Top: data in blue, with representative burst in red.  
+%%
+%Bottom: correlation between NIV coordinate and fast variable. }
+%\label{fig:vary_burst}
+%\end{figure}
 
@@
 
+\Appendix
+\section{The use of appendices}
+The \verb|\appendix| command may be used before the final sections
+of a paper to designate them as appendices. Once \verb|\appendix|
+is called, all subsequent sections will appear as
+
+\appendix
+\section{Title of appendix} Each one will be sequentially lettered
+instead of numbered. Theorem-like environments, subsections,
+and equations will also have the section number changed to a letter.
+
+If there is only {\em one} appendix, however, the \verb|\Appendix|
+(with a capital letter) should be used instead. This produces only 
+the word {\bf Appendix} in the section title, and does not add a letter.
+Equation numbers, theorem numbers and subsections of the appendix
+will have the letter ``A''  designating the section number.
+
+If you don't want to title your appendix, and just call it
+{\bf Appendix A.} for example, use \verb|\appendix\section*{}| 
+and don't include anything in the title field. This works
+opposite to the way \verb|\section*| usually works, by including the
+section number, but not using a title.
+
+Appendices should appear before the bibliography section, not after,
+and any acknowledgments should be placed after the appendices and before
+the bibliography.
+
+\begin{thebibliography}{1}
+\bibitem{GoMiSa} {\sc M. Goossens, F. Mittelbach, and A. Samarin},
+{\em The} \LaTeX\ {\em Companion}, Addison-Wesley, Reading, MA, 1994.
+
+\bibitem{Higham} {\sc N.~J. Higham}, {\em Handbook of Writing for
+the Mathematical Sciences}, Society for Industrial and Applied
+Mathematics, Philadelphia, PA, 1993.
+
+\bibitem{Lamport} {\sc L. Lamport}, \LaTeX: {\em A Document
+Preparation System}, Addison-Wesley, Reading, MA, 1986.
+
+\bibitem{SerLev} {\sc R. Seroul and S. Levy}, {\em A
+Beginner's Book of} \TeX, Springer-Verlag, Berlin, New
+York, 1991.
+\end{thebibliography}
+
 \end{document}
+%% end of file `docultexmm.tex'

### 2014-08-18 reducible_SDE/paper/SIADS_draft.tex
added small edits to reducible SDE paper

@@
 
-\usepackage{subcaption}
+\usepackage{graphicx}
+%\usepackage{subcaption}
+%\usepackage{epstopdf}
 
@@
 
-\author{TeX Production\thanks{Society for Industrial and
-Applied Mathematics, Philadelphia, Pennsylvania. 
-(\email{multimedia@siam.org}). Questions, comments, or corrections
-to this document may be directed to that email address.}}
+\author{Carmeline J. Dsilva \and Ronen Talmon \and C. William Gear \and Ronald R. Coifman \and Ioannis G. Kevrekidis}
 
@@
 TODO: change strip to look different than other data
-%\begin{figure}
-%\includegraphics[width=\textwidth]{schematic}
-%\caption{Illustration of how to choose $\delta t$ and $\sigma_{kernel}$ appropriately.
+
+\begin{figure}
+\includegraphics[width=\textwidth]{schematic.eps}
+\caption{Illustration of how to choose $\delta t$ and $\sigma_{kernel}$ appropriately.
 The data shows the evolution of the ``fast'' variable at a fixed value of the ``slow'' variable.
@@
 }
-%\label{fig:schematic}
-%\end{figure}
+\label{fig:schematic}
+\end{figure}
 

### 2014-08-30 reducible_SDE/paper/SIADS_draft.tex
added edits after meeting with Ronen

@@
 
-\usepackage{graphicx}
-%\usepackage{subcaption}
-%\usepackage{epstopdf}
+\usepackage{epsfig}
 
@@
 
-\author{Carmeline J. Dsilva \and Ronen Talmon \and C. William Gear \and Ronald R. Coifman \and Ioannis G. Kevrekidis}
+\author{Carmeline J.
+Dsilva\footnotemark[2] \and Ronen Talmon\footnotemark[3] \and C.
+William Gear\footnotemark[2] \and Ronald R.
+Coifman\footnotemark[3] \and Ioannis G.
+Kevrekidis\footnotemark[2]\ \footnotemark[4]}
 
@@
 \slugger{siads}{xxxx}{xx}{x}{x--x}}%slugger should be set to juq, siads, sifin, or siims
+
+\renewcommand{\thefootnote}{\fnsymbol{footnote}}
+
+\footnotetext[2]{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, 08544, USA}
+\footnotetext[3]{Department of Mathematics, Yale University, New Haven, Connecticut, 06520, USA}
+\footnotetext[4]{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, 08544, USA}
+
+\renewcommand{\thefootnote}{\arabic{footnote}}
 
@@
 
-\begin{AMS}\end{AMS}
+\begin{AMS}
+37M10, 62-07
+\end{AMS}
 
@@
 \thispagestyle{plain}
-\markboth{TEX PRODUCTION}{USING SIAM'S MM \LaTeX\ MACROS}
+\markboth{C.~J. DSILVA {\it ET AL}}{DATA-DRIVEN REDUCTION OF SDES}
 
@@
 
+\begin{itemize}
+
+\item
+Bridge/connect between data mining/analysis and dynamical systems
+
+\begin{itemize}
+
+\item
+Main idea and scope: analyze stochastic dynamical systems using data-driven methods.
+
+\begin{itemize}
+
+\item
+Reduction in fast-slow systems using manifold learning
+
+\end{itemize}
+
+\end{itemize}
+\item
+We cannot use off-the-shelf data-driven methods because they are typically based on metrics and geometry and are not informed (take into account) the dynamics and time.
+
+\item
+Our contribution:
+
+\begin{itemize}
+
+\item
+We show how to use data analysis methods that are typically applied to data sets for analyzing data collected as a time series from a dynamical system.
+In particular, we ``propose" a data-driven (graph-based) method for recovering the slow variable of reducible systems.
+
+\item
+The method is based mostly on local analysis that combines geometry and dynamics.
+
+\item
+We offer rigorous analysis for the method that gives the conditions (for our method to successfully recover the ``right" slow variables), intuition.
+
+\end{itemize}
+
+\end{itemize}
+
 \subsection{Multiscale SDEs}
+
+Add a general formulation of a reducible system (with a general drift, and decoupled diffusions) and Dror/Kasmiskii's conditions.
+(We assume that the saturation time of fast scale  of the SDE is smaller than the other relevant timescales.)
+\begin{equation} 
+\begin{aligned}
+dx &= a(x,y) dt + dW_1\\
+dy &= -\frac{b(x,y)}{\epsilon} dt + \frac{1}{\sqrt{\epsilon}} dW_2
+\end{aligned}
+\end{equation}
 
@@
 
-Consider the following system of SDEs:
+We consider systems that can be written as deterministic nonlinear functions $\mathbf{f}: \mathbb{R}^n \mapsto \mathbb{R}^m$ of the following system of SDEs (in the general form, in particular, this one):
 \begin{equation} \label{eq:fast_slow_SDE}
@@
 \end{equation}
-TODO: add more general formulation of the multi scale SDE
+where the diffusion terms of the fast and slow variables are decoupled.
 
 So $x$ is the slow variable, and $y$ is a fast noise whose equilibrium measure is bounded and $\mathcal{O}(1)$.
-%
-We assume that the saturation time of fast scale  of the SDE is smaller than the other relevant timescales.
 %
@@
 
-\subsection{Nonlinear functions}
-
-We will also consider a (nonlinear) function $\mathbf{f}: \mathbb{R}^n \mapsto \mathbb{R}^m$ of the data sampled from the SDE in \eqref{eq:fast_slow_SDE}.
-
 We will consider two specific examples. 
@@
 
-\section{Data mining: Nonlinear Intrinsic Variables}
+\section{Local Invariant Metrics}
 
-%These same conditions for uncoupled noises with unit variance will allow us to recover the slow variable(s) from data with multiple timescales. 
+Assume we want to locally recover the slow variables. 
 %
-We want to emphasize the link between multiscale SDEs and NIVs: NIVs rescales the noises, so that the ``fast'' variables become small.
+To do so, we need to build a metric that is invariant to the fast variables. 
 %
-Consider the SDE in \eqref{eq:fast_slow_SDE}, and let $y = \frac{z}{\sqrt{\epsilon}}$. 
+This is typically done by averaging out the fast variables. 
 %
-Then
+We propose to do this using the Mahalanobis distance, which is a local PCA-type approach that inverts the local covariance matrix of the data, thereby creating a space with normalized diffusion terms.
+%
+It can be shown that the Mahalanobis distance locally approximates: 
+\begin{equation}
+\| f(x_1, y_1) - f(x_2, y_2) \|^2_M = \| X_1 - X_2 \|^2_2 + \mathcal{O}(\| f(x_1, y_1) - f(x_2, y_2) \|^4_2)
+\end{equation}
+where $X$ comprises variables which have uncoupled noise with unit variance,
+i.e., the variables $X^1, \dots, X^n$ are governed by the following SDEs:
+\begin{equation} \label{eq:NIV_formulation}
+dX^i = a(X) dt + dW^i
+\end{equation}
+where $X = \begin{bmatrix} X^1 & X^2 & \cdots & X^n \end{bmatrix}^T$.
+%
+In particular for the case TODO, consider the change of variable $z = y\sqrt{\epsilon}$. This change of variables gives the following system:
 \begin{equation}
@@
 \end{equation}
+Therefore, $z$ is a stochastic variable whose diffusion is unity, and $X = \begin{bmatrix} x \\ z \end{bmatrix}$.
+
+Note that for this specific case
+\begin{equation}
+\| X_1 - X_2 \|^2_2 = \|x_1 - x_2 \|^2_2 + \epsilon \|y_1 - y_2 \|^2_2.
+\end{equation}
+Remarkably, it implies two features.
 %
-Therefore, $z$ is a stochastic variable whose diffusion is unity, and so $x$ and $z$ are the ``intrinsic variables'' for the underlying SDE.
+One, it rescales that data (in the sense that the variables have unit diffusion terms) so that the fast variables are collapsed and become (``epsilon") small as observed in the second term in the right hand side.
 %
-However, the equilibrium measure of $z$ is $\mathcal{O}(\epsilon)$, and so when $\epsilon \ll 1$, $z \ll x$, and so we will recover $x$ (the slow variable). 
+Two, it inverts the function $f$ (locally) and allows for the approximation of the Euclidean distance between samples of the slow variables in the ``decoupled" SDE space.
+
+%These same conditions for uncoupled noises with unit variance will allow us to recover the slow variable(s) from data with multiple timescales. 
 %
-In general, we want to recover the slow variable(s) even when the observations of the underlying intrinsic stochastic process are obstructed by a {\em nonlinear} measurement function.
+%
 
-We will define $X = \begin{bmatrix} x \\ z \end{bmatrix}$
+%To get a parameterization of NIV, we require $\| X_2 - X_1\|_2$.
+%
+%However, we only have access to $Y = f(X)$.
+%
+%Assuming that we have a local approximation to $\| X_2 - X_1\|$ from the measurements $Y$, we can then use DMAPS to uncover a parameterization of $X$ from the estimates of the pairwise distances.
 
-To get a parameterization of NIV, we require $\| X_2 - X_1\|_2$.
-%
-However, we only have access to $Y = f(X)$.
-%
-Assuming that we have a local approximation to $\| X_2 - X_1\|$ from the measurements $Y$, we can then use DMAPS to uncover a parameterization of $X$ from the estimates of the pairwise distances.
+\section{DMAPS/Graph-Based/Kernel-based method}
 
 Description of DMAPS........
+
+\section{Analysis}
+
+\begin{figure}
+\epsfig{width=\textwidth, file=schematic.eps}
+\caption{Illustration of how to choose $\delta t$ and $\sigma_{kernel}$ appropriately.
+The data shows the evolution of the ``fast'' variable at a fixed value of the ``slow'' variable.
+We must choose both parameters so that the curvature effects and other nonlinearities are negligable.
+}
+\label{fig:schematic}
+\end{figure}
+
+See schematic figure.
 
@@
 
-\begin{figure}
-\includegraphics[width=\textwidth]{schematic.eps}
-\caption{Illustration of how to choose $\delta t$ and $\sigma_{kernel}$ appropriately.
-The data shows the evolution of the ``fast'' variable at a fixed value of the ``slow'' variable.
-We must choose both parameters so that the curvature effects and other nonlinearities are negligable.
-}
-\label{fig:schematic}
-\end{figure}
+\subsection{Error analysis of the Mahalanobis distance}
 
-\subsection{Error analysis of the Mahalanobis distance}
+Given the true covariance, we can write the Euclidean distance $\|X_2 - X_1 \|$ using Taylor expansion as...
+
+Talk about fourth-order error terms, say we want to choose the kernel scale such that these error terms are negligible. 
+%
+The error terms involve curvature of $f$, etc.
 
@@
 
-To compute the Mahalanobis distance, we require $J J^T$.
+To compute the Mahalanobis distance, we require $J J^T$ (estimate the covariance matrix from the data).
 %
-By Ito formula, we have
-\begin{equation}
-dY^j = \sum_{i=1}^n \left( \frac{1}{2} f_{ii}^j + a_i f_i^j \right) dt + \sum_{i=1}^n f_i^j dW_i
-\end{equation}
+We do this using local simulation ``bursts''. 
 %
-TODO: define partials...
+The error in the covariance estimation is... (add equations).
 %
-The covariance $C$ defined by 
-\begin{equation}
-[C]_{jk} = \mathrm{Cov} \left( dY^j, dY^k \right)
-\end{equation}
-equals $C = J J^T$ \cite{...}.
-
-We will empirically estimate the covariance from data.
-
-Comment: Use a more general model (not necessarily multi scale)
-
-The goal of NIVs is to recover ``intrinsic'' variables that are independent of the measurement/observation function.
-%
-We define intrinsic variables as those variables which have uncoupled noise with unit variance,
-i.e., the intrinsic variables $X^1, \dots, X^n$ are governed by the following SDEs:
-\begin{equation} \label{eq:NIV_formulation}
-dX^i = a(X) dt + dW^i
-\end{equation}
-where $X = \begin{bmatrix} X^1 & X^2 & \cdots & X^n \end{bmatrix}^T$.
-
-Restrictions:
-\begin{itemize}
-\item Drift is bilipshitz wrt process $X$
-\item Access to simulator, BUT the analysis could be extended to a single trajectory (cite people....); emphasize the difference between simulation and experiment
-\end{itemize}
-
-We assume that
-\begin{equation}
- \int_{t}^{t + \delta t} a(X_\tau) d \tau - a(X_t) \delta t  = \mathcal{O} (\delta t) \: \forall t
-\end{equation}
-TODO: for the toy problems, find the order of this term, possibly as a function of $\epsilon$-- then we can relate $\epsilon$ to $\delta t$, and place bounds on $\delta t$
-
-\begin{equation}
-\begin{aligned}
-X(t + \delta t) =& X_t + \int_{t}^{t+\delta t} a(X_\tau) d \tau + d W \\
-= & X_t + \int_{t}^{t+\delta t} \left[ a(X_t) + \left( a(X_\tau) - a(X_t) \right) \right] d \tau + d W \\
-= & X_t + a(X_t) \delta t  + \sqrt{\delta t} W + \mathcal{O} (\delta t)
-\end{aligned}
-\end{equation}
-
-Given $X$ governed by the SDE in \eqref{eq:NIV_formulation}, we can sample ``bursts'' around a point $X_0$ in order to estimate the local covariance of the noise.
-
-These clouds will be distributed
-
-\begin{equation}
-X \sim \mathcal{N}\left( X_0, \delta t I \right)
-\end{equation}
-where 
-$$X_0 = X_t + \left( a(X_t) + \mathcal{O}(1) \right) \delta t $$
-%
-Note: we bound the random variation due to the variations in the drift, and note that this bound makes these variations smaller than the variations due to the local noise/diffusion. 
-%
-Therefore, we simply absorb these variations into an $\mathcal{O}(\delta t)$ correction of the mean of the sample distribution, turning a random term into a deterministic term.
-
-Let $f^i(X)$ denote the $i^{th}$ entry of $\mathbf{f}$, and let $f^i_j(X) = \frac{\partial f^i}{\partial X^j}$.
-%
-By Taylor expansion around $X=X_0$,
-%
-\begin{eqnarray}
-\hat{C}(X_0) &=& \delta t J_1 J_1^T 
-+ \frac{\delta t^2}{2} \left[ J_1 J_3^T + J_3 J_1^T  
-+ \frac{3}{2} J_2 J_2^T 
--\frac{1 }{2} J_2 1 1^T J_2^T \right]
-+ \mathcal{O} (\delta t^3) %\\
-%&=& \delta t J_1 J_1^T 
-%+ \frac{\delta t^2}{2} U \Sigma V
-%+ \mathcal{O} (\delta t^3)
-\end{eqnarray}
-where
-\begin{equation}
-\begin{aligned}
-J_1(X_0) &= \begin{bmatrix}
-f_1^1 & f_2^1 & \dots & f_n^1 \\
-\vdots & \vdots & \cdots & \vdots \\
-f_1^m & f_2^m & \dots & f_n^m
-\end{bmatrix}\\
-%
-J_2(X_0) &= \begin{bmatrix}
-f_{11}^1 & f_{22}^1 & \dots & f_{nn}^1 \\
-\vdots & \vdots & \cdots & \vdots \\
-f_{11}^m & f_{22}^m & \dots & f_{nn}^m
-\end{bmatrix}\\
-%
-J_3(X_0) &= \begin{bmatrix}
-f_{111}^1 & f_{222}^1 & \dots & f_{nnn}^1 \\
-\vdots & \vdots & \cdots & \vdots \\
-f_{111}^m & f_{222}^m & \dots & f_{nnn}^m
-\end{bmatrix}
-\end{aligned}
-\end{equation}
-%
-In this equation, we see what we expect (a linear function of the variance, due to the gradient of $f$) and a second (higher-order) term arising from the curvature and third derivative of $f$.
-%
-The importance of this second term increases as the curvature increases, as well as when $\delta t$ increases.
-
-Then
-\begin{equation}
-\hat{C}^{-1}(X_0) \approx \frac{1}{\delta t} \left( J J^T \right)^{-1} - A
-\end{equation}
-%
-where 
-\begin{equation}
-A = \frac{1}{2} (J_1 J_1^T)^{-1} U \left(\Sigma^{-1} + \frac{\delta t}{2} V \left( J_1 J_1^T \right)^{-1} U \right)^{-1} V \left( J_1 J_1^T \right)^{-1}
-\end{equation}
-and
-\begin{equation}
-U \Sigma V = J_1 J_3^T + J_3 J_1^T + \frac{3}{2} J_2 J_2^T -\frac{1 }{2} J_2 1 1^T J_2^T 
-\end{equation}
-$A$ is a function of the first-, second-, and third-order derivatives of $\mathbf{f}$.
+The lowest-order error term depends on the curvature of $f$.
 
@@
 
-From \eqref{eq:taylor_expansion}, we can see that the squared {\em intrinsic} distance $\| X_2 - X_1 \|^2$, is a function of $\|Y_2 - Y_1\|$ and $\delta t$. 
+TODO: interpret the different terms we get w.r.t. the curvature of f, the drift, the scale epsilon, and the window size.
+
+From \eqref{eq:taylor_expansion}, we can see that the squared distance $\| X_2 - X_1 \|^2$, is a function of $\|Y_2 - Y_1\|$ and $\delta t$. 
 %
@@
 
-\Appendix
-\section{The use of appendices}
-The \verb|\appendix| command may be used before the final sections
-of a paper to designate them as appendices. Once \verb|\appendix|
-is called, all subsequent sections will appear as
-
-\appendix
-\section{Title of appendix} Each one will be sequentially lettered
-instead of numbered. Theorem-like environments, subsections,
-and equations will also have the section number changed to a letter.
-
-If there is only {\em one} appendix, however, the \verb|\Appendix|
-(with a capital letter) should be used instead. This produces only 
-the word {\bf Appendix} in the section title, and does not add a letter.
-Equation numbers, theorem numbers and subsections of the appendix
-will have the letter ``A''  designating the section number.
-
-If you don't want to title your appendix, and just call it
-{\bf Appendix A.} for example, use \verb|\appendix\section*{}| 
-and don't include anything in the title field. This works
-opposite to the way \verb|\section*| usually works, by including the
-section number, but not using a title.
-
-Appendices should appear before the bibliography section, not after,
-and any acknowledgments should be placed after the appendices and before
-the bibliography.
-
-\begin{thebibliography}{1}
-\bibitem{GoMiSa} {\sc M. Goossens, F. Mittelbach, and A. Samarin},
-{\em The} \LaTeX\ {\em Companion}, Addison-Wesley, Reading, MA, 1994.
-
-\bibitem{Higham} {\sc N.~J. Higham}, {\em Handbook of Writing for
-the Mathematical Sciences}, Society for Industrial and Applied
-Mathematics, Philadelphia, PA, 1993.
-
-\bibitem{Lamport} {\sc L. Lamport}, \LaTeX: {\em A Document
-Preparation System}, Addison-Wesley, Reading, MA, 1986.
-
-\bibitem{SerLev} {\sc R. Seroul and S. Levy}, {\em A
-Beginner's Book of} \TeX, Springer-Verlag, Berlin, New
-York, 1991.
-\end{thebibliography}
-
 \end{document}

### 2014-08-31 reducible_SDE/paper/SIADS_draft.tex
added more it taylor calc

@@
 
-Add a general formulation of a reducible system (with a general drift, and decoupled diffusions) and Dror/Kasmiskii's conditions.
-(We assume that the saturation time of fast scale  of the SDE is smaller than the other relevant timescales.)
-\begin{equation} 
+We consider the following two-scale SDE. 
+\begin{equation} \label{eq:general_SDE}
 \begin{aligned}
-dx &= a(x,y) dt + dW_1\\
-dy &= -\frac{b(x,y)}{\epsilon} dt + \frac{1}{\sqrt{\epsilon}} dW_2
+dx_i &= a_i(x_1, \dots, x_n) dt + dW_i, \: 1 \le i \le m \\
+dx_i &= -\frac{a_i(x_1, \dots, x_n)}{\epsilon} dt + \frac{1}{\sqrt{\epsilon}} dW_i , \: m+1 \le i \le n
 \end{aligned}
 \end{equation}
-
-Given a system of SDEs with separation of timescales, we want to recover slow variable(s) even when the fast variables are have large variations.
-
-We consider systems that can be written as deterministic nonlinear functions $\mathbf{f}: \mathbb{R}^n \mapsto \mathbb{R}^m$ of the following system of SDEs (in the general form, in particular, this one):
-\begin{equation} \label{eq:fast_slow_SDE}
-\begin{aligned}
-dx &= adt + dW_1\\
-dy &= -\frac{y}{\epsilon} dt + \frac{1}{\sqrt{\epsilon}} dW_2
-\end{aligned}
-\end{equation}
-where the diffusion terms of the fast and slow variables are decoupled.
-
-So $x$ is the slow variable, and $y$ is a fast noise whose equilibrium measure is bounded and $\mathcal{O}(1)$.
+where $W_i$ are independent Brownian motions, and $\epsilon \ll 1$.
 %
-We would like to recover $x$ using data-driven techniques.
-
-We would like to note that the ratio of the drift and diffusion terms in the $y$ equation is essential.
+Therefore, \eqref{eq:general_SDE} defines an $n$-dimensional stochastic system with $m$ slow variables and $n-m$ fast variables. 
+%
+We would like to note that the ratio of the drift and diffusion terms is essential.
 %
@@
 %
-If the diffusivity is too large, the equilibrium measure of $y$ will be unbounded.
+If the diffusivity is too large, the equilibrium measure will be unbounded.
 %
-Conversely, if the diffusivity is too small, the equilibrium measure of $y$ will go to 0.
+Conversely, if the diffusivity is too small, the equilibrium measure will go to 0.
+
+Assuming that the functions $a_i$ are bounded and Lipschitz, and we can find
+\begin{equation}
+\mathbb{E} \left| \frac{1}{T} \int_t^{t + T} a_i(x_1, \dots, x_m, X_{m+1}(s), \dots, X_n(s)) ds - \overline{a} (x_1, \dots, x_m) \right| < \kappa_i (T) 
+\end{equation}
+for $1 \le i \le m$, where $\kappa_i((T) \rightarrow 0 $ as $T \rightarrow \infty$, then the averaging principle \cite{...} states that we can write an effective equation in the slow variables $x_1, \dots, x_m$.
+%
+We will consider systems that can be written as deterministic (perhaps nonlinear) bi-Lipshitz functions $\mathbf{f}: \mathbb{R}^n \mapsto \mathbb{R}^d$ of the SDE in \eqref{eq:general_SDE}.
+%
+Our goal is to recover the slow variables $x_1, \dots, x_m$ using data-driven techniques.
+
+For illustrative purposes, we will consider the following two-dimensional SDE as a specific example of \eqref{eq:general_SDE}. 
+\begin{equation} \label{eq:fast_slow_SDE}
+\begin{aligned}
+dx_1 &= adt + dW_1\\
+dx_2 &= -\frac{x_2}{\epsilon} dt + \frac{1}{\sqrt{\epsilon}} dW_2
+\end{aligned}
+\end{equation}
+%
+where $a$ is a constant of order 1. 
+%
+So $x_1$ is the slow variable, and $x_2$ is a fast noise whose equilibrium measure is bounded and $\mathcal{O}(1)$.
+%
+We would like to recover $x_1$ using data-driven techniques.
 
 We will consider two specific examples. 
-\begin{enumerate}
-\item fat strip
-\item half moon
-\end{enumerate}
+%
+In the first example, our data will form a strip where the fast and slow remain uncoupled,
+\begin{equation}
+\mathbf{f_1}(x_1, x_2) = \begin{bmatrix} x_1 \\ x_2 \end{bmatrix}
+\end{equation}
+
+In the second example, our data will be warped into half-moon shapes
+\begin{equation}
+\mathbf{f_2}(x_1, x_2) = 
+\begin{bmatrix} 
+x_1 + x_2^2 \\
+x_2
+\end{bmatrix}
+\end{equation}
 %
@@
 
-Assume we want to locally recover the slow variables. 
+To locally recover the slow variables, we need to build a metric that is invariant to the fast variables. 
 %
-To do so, we need to build a metric that is invariant to the fast variables. 
-%
-This is typically done by averaging out the fast variables. 
+Typically, this is done by averaging out the fast variables. 
 %
@@
 It can be shown that the Mahalanobis distance locally approximates: 
-\begin{equation}
-\| f(x_1, y_1) - f(x_2, y_2) \|^2_M = \| X_1 - X_2 \|^2_2 + \mathcal{O}(\| f(x_1, y_1) - f(x_2, y_2) \|^4_2)
+\begin{equation} \label{eq:mahalanobis}
+\| \mathbf{f}(Y_1) - \mathbf{f}(Y_2) \|^2_M = \| Y_1 - Y_2 \|^2_2 + \mathcal{O}(\| \mathbf{f}(Y_1) - \mathbf{f}(Y_2) \|^4_2)
 \end{equation}
-where $X$ comprises variables which have uncoupled noise with unit variance,
-i.e., the variables $X^1, \dots, X^n$ are governed by the following SDEs:
+where the components of $Y$ have uncoupled noise with unit variance,
+i.e., the variables $y_1, \dots, y_n$ are governed by the following SDEs:
 \begin{equation} \label{eq:NIV_formulation}
-dX^i = a(X) dt + dW^i
+dy_i = b(y) dt + dW_i
 \end{equation}
-where $X = \begin{bmatrix} X^1 & X^2 & \cdots & X^n \end{bmatrix}^T$.
-%
-In particular for the case TODO, consider the change of variable $z = y\sqrt{\epsilon}$. This change of variables gives the following system:
+and $Y = \begin{bmatrix} y_1 & y_2& \cdots & y_n \end{bmatrix}^T$.
+
+In particular, for SDEs of the form in \eqref{eq:general_SDE}, consider the transformation
+\begin{equation} \label{eq:general_rescale}
+\begin{aligned}
+x_i = y_i, \: &1 \le i \le m \\
+x_i = \frac{y_i}{\sqrt{\epsilon}} \: &1 \le i \le m
+\end{aligned}
+\end{equation}
+
+Then, \eqref{eq:general_SDE} becomes
 \begin{equation} 
 \begin{aligned}
-dx &= adt + dW_1\\
-dz &= -\frac{z}{\epsilon} dt +  dW_2
+dy_i &= b_i(y_1, \dots, y_n) dt + dW_i, \: 1 \le i \le m \\
+dy_i &= -\frac{b_i(y_1, \dots, y_n)}{\sqrt{\epsilon}} dt + dW_i , \: m+1 \le i \le n
 \end{aligned}
 \end{equation}
-Therefore, $z$ is a stochastic variable whose diffusion is unity, and $X = \begin{bmatrix} x \\ z \end{bmatrix}$.
+where $b_i(y_1, \dots, y_n) = a_i (y_1, \dots, y_m, \frac{y_{m+1}}{\sqrt{\epsilon}}, \dots, \frac{y_n}{\sqrt{\epsilon}})$.
+%
+From \eqref{eq:general_rescale}, we can see that if $x_i = \mathcal{O}(1)$, then $y_i = \mathcal{O}({\sqrt{\epsilon}})$ for $ m+1 \le i \le n$, and 
+\begin{equation}
+\| Y_2 - Y_1 \|^2_2 = \sum_{i=1}^m \| y_{i,2} - y_{i,1} \|^2 = \sum_{i=1}^m \|x_{i,2} - x_{i,1} \|^2 + \epsilon \sum_{i=m+1}^n \|x_{i,2} - x_{i,1}\|^2
+\end{equation}
+%
+Remarkably, this metric induces two important features. 
+%
+One, rescaling the data so that all variables have unit diffusion terms means that fast variables are collapsed and become ``epsilon" small, as observed in the second term in the right hand side. 
+%
+Two, the righthandside of \eqref{eq:mahalanobis} is invariant (to fourth order) to the function $\mathbf{f}$ and allows for the approximation of the Euclidean distance between samples of the slow variables in the ``decoupled" SDE space.
 
-Note that for this specific case
+For the specific example \eqref{eq:fast_slow_SDE}, the change of variables gives the following system:
+\begin{equation}
+\begin{aligned}
+dy_1 &= adt + dW_1\\
+dy_2 &= -\frac{y}{\epsilon} dt +  dW_2
+\end{aligned}
+\end{equation}
+Therefore, $y_2$ is a stochastic variable whose diffusion is unity, and 
 \begin{equation}
@@
 \end{equation}
-Remarkably, it implies two features.
+
+\section{Diffusion Maps for Global Parameterization}
+
+Our goal is to extract a {\em global} parameterization of the data that respects the slow variables. 
 %
-One, it rescales that data (in the sense that the variables have unit diffusion terms) so that the fast variables are collapsed and become (``epsilon") small as observed in the second term in the right hand side.
+We will use diffusion maps, a kernel-based manifold learning technique, to extract a global parameterization using the local distances that we described in the previous section. 
 %
-Two, it inverts the function $f$ (locally) and allows for the approximation of the Euclidean distance between samples of the slow variables in the ``decoupled" SDE space.
+Given data $z_1, \dots, z_N \in \mathbb{R}^d$, we first construct the matrix $W \in \mathbb{R}^{N \times N}$, where 
+\begin{equation}
+W_{ij} = \exp \left( -\frac{\|z_i - z_j \|^2}{\sigma_{kernel}^2} \right)
+\end{equation}
+where $\| \cdot \|$ denotes the appropriate norm, and $\sigma_{kernel}$ is the kernel scale and denotes a characteristic distance within the data set (we often set $\sigma_{kernel}$ to be the median of the pairwise distances).
+%
+In our applications, we will use the Mahalanobis distance \eqref{eq:mahalanobis} as the norm between points, so that we can recover the slow variables.
 
-%These same conditions for uncoupled noises with unit variance will allow us to recover the slow variable(s) from data with multiple timescales. 
+We then construct the diagonal matrix $D \in \mathbb{R}^{N \times N}$, with 
+\begin{equation}
+D_{ii} = \sum_{j=1}^N W_{ij}
+\end{equation}
 %
+We compute the eigenvalues $\lambda_0, \dots, \lambda_{N-1}$ and eigenvalues $\phi_0, \dots, \phi_{N-1}$ of the matrix $A = D^{-1}W$, and order them such that $|\lambda_0| \ge |\lambda_1| \ge \dots \ge |\lambda_{N-1}|$.
 %
-
-%To get a parameterization of NIV, we require $\| X_2 - X_1\|_2$.
-%
-%However, we only have access to $Y = f(X)$.
-%
-%Assuming that we have a local approximation to $\| X_2 - X_1\|$ from the measurements $Y$, we can then use DMAPS to uncover a parameterization of $X$ from the estimates of the pairwise distances.
-
-\section{DMAPS/Graph-Based/Kernel-based method}
-
-Description of DMAPS........
+$\phi_0$ is a constant trivial eigenvector; the next few eigenvectors give a parameterization/embedding coordinates for the data.
 
@@
 
-See schematic figure.
-
 Our approximation of the pairwise distances will rely on the Taylor expansion of the measurement function $f$. 
@@
 %
-Accordingly, we will address the accuracy of our method from two standpoints: (1) the accuracy of the Taylor expansion, and (2) the accuracy of the covariance estimation (see Figure...).
+Accordingly, we will address the accuracy of our method from two standpoints: (1) the accuracy of the Taylor expansion, and (2) the accuracy of the covariance estimation (see Figure~\ref{fig:schematic}).
 %
@@
 \subsection{Error analysis of the covariance estimation}
+
+\begin{equation}
+\begin{aligned}
+f( X_\tau) = & f( X_\rho) \\
+&+ \sum_{j=1}^d \left. \frac{\partial f}{\partial x^j} \right|_{X_\rho}  \int_{\rho}^{\tau}  dW_s^j \\ &
++  \left.
+\left( \sum_{k=1}^d a^k \frac{\partial f}{\partial x^k} + \frac{1}{2} \sum_{j=1}^d \frac{\partial^2 f}{\partial {x^j}^2} \right) \right|_{X_\rho} \int_{\rho}^{\tau} ds + \sum_{j=1}^m \sum_{k=1}^m \int_{\rho}^{\tau} \int_{\rho}^{s_2} \frac{\partial^2 f}{\partial x^j \partial x^k} dW_{s_1}^j dW_{s_2}^k \\
+& + \sum_{j=1}^d  \int_{\rho}^{\tau} \int_{\rho}^{s_2} \left( \sum_{k=1}^d a^k \frac{\partial^2 f}{\partial x^j \partial x^k} + \frac{1}{2} \sum_{k=1}^d \frac{\partial^3 f}{\partial x^j \partial {x^k}^2} \right) dW_{s_1}^j ds_2 \\
+& + \sum_{j=1}^d  \int_{\rho}^{\tau} \int_{\rho}^{s_2} \left( \sum_{k=1}^d a^k \frac{\partial^ f}{\partial x^k \partial x^j} + \frac{1}{2} \sum_{k=1}^d  \frac{\partial^3 f}{\partial {x^k}^2 \partial x^j} \right) ds_1 dW_{s_2}^j  \\
+& + \int_{\rho}^{\tau} \int_{\rho}^{s_2} \left( \sum_{j=1}^d \sum_{k=1}^d \left[ a^j \frac{\partial}{\partial x^j}\left( a^k \frac{\partial f}{\partial x^k} \right) + \frac{1}{2} \frac{\partial^2}{\partial {x^j}^2} \left( a^k \frac{\partial f}{\partial x^k} \right) + \frac{1}{2} a^j \frac{\partial^3 f}{\partial x^j \partial {x^k}^2} + \frac{1}{4} \frac{\partial^4 f}{\partial {x^j}^2 \partial {x^k}^2} \right] \right)  ds_1 ds_2
+\end{aligned}
+\end{equation}
 

### 2014-09-01 reducible_SDE/paper/SIADS_draft.tex
Added more to ito calc stuff

@@
 \begin{equation} \label{eq:mahalanobis}
-\| \mathbf{f}(Y_1) - \mathbf{f}(Y_2) \|^2_M = \| Y_1 - Y_2 \|^2_2 + \mathcal{O}(\| \mathbf{f}(Y_1) - \mathbf{f}(Y_2) \|^4_2)
+\| \mathbf{f}(\Xi_1) - \mathbf{f}(\Xi_2) \|^2_M = \| \Xi_1 - \Xi_2 \|^2_2 + \mathcal{O}(\| \mathbf{f}(\Xi_1) - \mathbf{f}(\Xi_2) \|^4_2)
 \end{equation}
-where the components of $Y$ have uncoupled noise with unit variance,
-i.e., the variables $y_1, \dots, y_n$ are governed by the following SDEs:
+where the components of $\Xi$ have uncoupled noise with unit variance,
+i.e., the variables $\xi_1, \dots, \xi_n$ are governed by the following SDEs:
 \begin{equation} \label{eq:NIV_formulation}
-dy_i = b(y) dt + dW_i
+d\xi_i = b(\xi) dt + dW_i
 \end{equation}
-and $Y = \begin{bmatrix} y_1 & y_2& \cdots & y_n \end{bmatrix}^T$.
+and $\Xi = \begin{bmatrix} \xi_1 & \xi_2& \cdots & \xi_n \end{bmatrix}^T$.
 
@@
 \begin{aligned}
-x_i = y_i, \: &1 \le i \le m \\
-x_i = \frac{y_i}{\sqrt{\epsilon}} \: &1 \le i \le m
+x_i = \xi_i, \: &1 \le i \le m \\
+x_i = \frac{\xi_i}{\sqrt{\epsilon}} \: &1 \le i \le m
 \end{aligned}
@@
 \begin{aligned}
-dy_i &= b_i(y_1, \dots, y_n) dt + dW_i, \: 1 \le i \le m \\
-dy_i &= -\frac{b_i(y_1, \dots, y_n)}{\sqrt{\epsilon}} dt + dW_i , \: m+1 \le i \le n
+d\xi_i &= b_i(\xi_1, \dots, \xi_n) dt + dW_i, \: 1 \le i \le m \\
+d\xi_i &= -\frac{b_i(\xi_1, \dots, \xi_n)}{\sqrt{\epsilon}} dt + dW_i , \: m+1 \le i \le n
 \end{aligned}
 \end{equation}
-where $b_i(y_1, \dots, y_n) = a_i (y_1, \dots, y_m, \frac{y_{m+1}}{\sqrt{\epsilon}}, \dots, \frac{y_n}{\sqrt{\epsilon}})$.
+where $b_i(\xi_1, \dots, \xi_n) = a_i (\xi_1, \dots, \xi_m, \frac{\xi_{m+1}}{\sqrt{\epsilon}}, \dots, \frac{\xi_n}{\sqrt{\epsilon}})$.
 %
-From \eqref{eq:general_rescale}, we can see that if $x_i = \mathcal{O}(1)$, then $y_i = \mathcal{O}({\sqrt{\epsilon}})$ for $ m+1 \le i \le n$, and 
+From \eqref{eq:general_rescale}, we can see that if $x_i = \mathcal{O}(1)$, then $\xi_i = \mathcal{O}({\sqrt{\epsilon}})$ for $ m+1 \le i \le n$, and 
 \begin{equation}
-\| Y_2 - Y_1 \|^2_2 = \sum_{i=1}^m \| y_{i,2} - y_{i,1} \|^2 = \sum_{i=1}^m \|x_{i,2} - x_{i,1} \|^2 + \epsilon \sum_{i=m+1}^n \|x_{i,2} - x_{i,1}\|^2
+\| \Xi_2 - \Xi_1 \|^2_2 = \sum_{i=1}^m \| \xi_{i,2} - \xi_{i,1} \|^2 = \sum_{i=1}^m \|x_{i,2} - x_{i,1} \|^2 + \epsilon \sum_{i=m+1}^n \|x_{i,2} - x_{i,1}\|^2
 \end{equation}
@@
 \begin{aligned}
-dy_1 &= adt + dW_1\\
-dy_2 &= -\frac{y}{\epsilon} dt +  dW_2
+d\xi_1 &= adt + dW_1\\
+d\xi_2 &= -\frac{\xi}{\epsilon} dt +  dW_2
 \end{aligned}
-\end{equation}
-Therefore, $y_2$ is a stochastic variable whose diffusion is unity, and 
-\begin{equation}
-\| X_1 - X_2 \|^2_2 = \|x_1 - x_2 \|^2_2 + \epsilon \|y_1 - y_2 \|^2_2.
 \end{equation}
@@
 
+For our general multi scale + function SDE formulation,
+
+For our general multi scale + function SDE formulation, 
 \begin{equation}
 \begin{aligned}
-f( X_\tau) = & f( X_\rho) \\
-&+ \sum_{j=1}^d \left. \frac{\partial f}{\partial x^j} \right|_{X_\rho}  \int_{\rho}^{\tau}  dW_s^j \\ &
-+  \left.
-\left( \sum_{k=1}^d a^k \frac{\partial f}{\partial x^k} + \frac{1}{2} \sum_{j=1}^d \frac{\partial^2 f}{\partial {x^j}^2} \right) \right|_{X_\rho} \int_{\rho}^{\tau} ds + \sum_{j=1}^m \sum_{k=1}^m \int_{\rho}^{\tau} \int_{\rho}^{s_2} \frac{\partial^2 f}{\partial x^j \partial x^k} dW_{s_1}^j dW_{s_2}^k \\
-& + \sum_{j=1}^d  \int_{\rho}^{\tau} \int_{\rho}^{s_2} \left( \sum_{k=1}^d a^k \frac{\partial^2 f}{\partial x^j \partial x^k} + \frac{1}{2} \sum_{k=1}^d \frac{\partial^3 f}{\partial x^j \partial {x^k}^2} \right) dW_{s_1}^j ds_2 \\
-& + \sum_{j=1}^d  \int_{\rho}^{\tau} \int_{\rho}^{s_2} \left( \sum_{k=1}^d a^k \frac{\partial^ f}{\partial x^k \partial x^j} + \frac{1}{2} \sum_{k=1}^d  \frac{\partial^3 f}{\partial {x^k}^2 \partial x^j} \right) ds_1 dW_{s_2}^j  \\
-& + \int_{\rho}^{\tau} \int_{\rho}^{s_2} \left( \sum_{j=1}^d \sum_{k=1}^d \left[ a^j \frac{\partial}{\partial x^j}\left( a^k \frac{\partial f}{\partial x^k} \right) + \frac{1}{2} \frac{\partial^2}{\partial {x^j}^2} \left( a^k \frac{\partial f}{\partial x^k} \right) + \frac{1}{2} a^j \frac{\partial^3 f}{\partial x^j \partial {x^k}^2} + \frac{1}{4} \frac{\partial^4 f}{\partial {x^j}^2 \partial {x^k}^2} \right] \right)  ds_1 ds_2
+&\mathbb{E} \left[ f_1(X_\tau) f_2(X_\tau) \right] - \mathbb{E}[f_1(X_\tau)]\mathbb{E}[f_2(X_\tau)] = \\
+& \sum_{i=1}^m \left. \frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \left. \frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i \right)^2 \right] \\
+& + \frac{1}{\epsilon} \sum_{i=m+1}^n \left.
+\frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \left.
+\frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i \right)^2 \right] \\
+& + \sum_{i=1}^m \left.
+\frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i  \right) \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_2}{\partial x_i^2} dW_{s_1}^i dW_{s_2}^i  \right) \right] \\
+& + \frac{1}{\epsilon^{3/2}} \sum_{i=m+1}^n \left.
+\frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i  \right) \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_2}{\partial x_i^2} dW_{s_1}^i dW_{s_2}^i  \right) \right] \\
+& + \sum_{i=1}^m \left.
+\frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i \right)  \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_1}{\partial x_i^2}  dW_{s_1}^i dW_{s_2}^i \right) \right] \\
+& + \frac{1}{\epsilon^{3/2}} \sum_{i=m+1}^n \left.
+\frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i \right)  \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_1}{\partial x_i^2}  dW_{s_1}^i dW_{s_2}^i \right) \right] \\
+&+ \mathcal{O} ((\tau - \rho)^2 ) 
 \end{aligned}
@@
 
-To compute the Mahalanobis distance, we require $J J^T$ (estimate the covariance matrix from the data).
-%
-We do this using local simulation ``bursts''. 
-%
-The error in the covariance estimation is... (add equations).
-%
-The lowest-order error term depends on the curvature of $f$.
-
-TODO: define error function for covariance estimation
-
-Substituting our estimate of the covariance into \eqref{eq:taylor_expansion}, we obtain
+For the case when $f$ is linear,
 \begin{equation}
 \begin{aligned}
-&\| X_2 - X_1 \|^2 \\
-\approx &  \frac{1}{2} (Y_2 - Y_1)^T \left((J J^T)^{-1}(Y_1)  + (J J^T)^{-1}(Y_2) \right) (Y_2 - Y_1) \\
-& - \frac{\delta t}{2} (Y_2 - Y_1)^T \left( A(Y_1) + A(Y_2) \right) (Y_2 - Y_1)\\
-& + n \left( \left| K_1 K_2 \right| + \left| \frac{ K_2^2}{4} \right|  + \left| \frac{K_1 K_3}{3} \right|  \right) \| Y_2 - Y_1 \| ^4  \\
-& + \mathcal{O} (\|Y_1 - Y_2 \|^6 ) 
+& \mathbb{E} [f_1(X_\tau) f_2(X_\tau) ] - \mathbb{E}[f_1(X_\tau)]\mathbb{E}[f_2(X_\tau)]= \\
+=  &  \sum_{i=1}^m  \left.
+\frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \left.
+\frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)^2  \right] \\
+& + \frac{1}{\epsilon} \sum_{i=m+1}^n  \left.
+\frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \left.
+\frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)^2  \right] \\
+ & + \sum_{i=1}^m \left.
+\frac{\partial f_1}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_2}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_2}{\partial x_k} \right)  dW_{s_1}^i ds_2 \right) \right] \\
+ & + \frac{1}{\epsilon} \sum_{i=m+1}^n \left.
+\frac{\partial f_1}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_2}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_2}{\partial x_k} \right)  dW_{s_1}^i ds_2 \right) \right] \\
+ & + \sum_{i=1}^m  \left.
+\frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_1}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_1}{\partial x_k} \right) dW_{s_1}^i ds_2 \right) \right] \\
+& + \frac{1}{\epsilon} \sum_{i=m+1}^n  \left.
+\frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_1}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_1}{\partial x_k} \right) dW_{s_1}^i ds_2 \right) \right] \\
+& + \mathcal{O} ((\tau - \rho)^{5/2})
+ \end{aligned}
+\end{equation}
+
+Therefore, we have
+\begin{equation}
+\begin{aligned}
+e_c = &  \sum_{i=1}^m \left.
+\frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i  \right) \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_2}{\partial x_i^2} dW_{s_1}^i dW_{s_2}^i  \right) \right] \\
+& + \frac{1}{\epsilon^{3/2}} \sum_{i=m+1}^n \left.
+\frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i  \right) \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_2}{\partial x_i^2} dW_{s_1}^i dW_{s_2}^i  \right) \right] \\
+& + \sum_{i=1}^m \left.
+\frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i \right)  \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_1}{\partial x_i^2}  dW_{s_1}^i dW_{s_2}^i \right) \right] \\
+& + \frac{1}{\epsilon^{3/2}} \sum_{i=m+1}^n \left.
+\frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i \right)  \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_1}{\partial x_i^2}  dW_{s_1}^i dW_{s_2}^i \right) \right] \\
+&+ \mathcal{O} ((\tau - \rho)^2 ) 
 \end{aligned}
@@
 %
-We want the first term in this expression to dominate.
-
-TODO: interpret the different terms we get w.r.t. the curvature of f, the drift, the scale epsilon, and the window size.
-
-From \eqref{eq:taylor_expansion}, we can see that the squared distance $\| X_2 - X_1 \|^2$, is a function of $\|Y_2 - Y_1\|$ and $\delta t$. 
+when $\frac{\partial f}{\partial x_i} \ne 0$, and 
 %
-If the first term in the expansion dominates, then $\|X_2 - X_1 \|^2$ should be independent of $\delta t$, and should vary quadratically with $\| Y_2 - Y_1\|$. 
-%
-We can find such a regime empirically (by plotting $\| X_2 - X_1 \|^2$ as a function of $\delta t$ and $\|Y_2 - Y_1\|$), and select a kernel distance where both conditions hold.
+\begin{equation}
+\begin{aligned}
+e_c =  &  \sum_{i=1}^m \left.
+\frac{\partial f_1}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_2}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_2}{\partial x_k} \right)  dW_{s_1}^i ds_2 \right) \right] \\
+ & + \frac{1}{\epsilon} \sum_{i=m+1}^n \left.
+\frac{\partial f_1}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_2}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_2}{\partial x_k} \right)  dW_{s_1}^i ds_2 \right) \right] \\
+ & + \sum_{i=1}^m  \left.
+\frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_1}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_1}{\partial x_k} \right) dW_{s_1}^i ds_2 \right) \right] \\
+& + \frac{1}{\epsilon} \sum_{i=m+1}^n  \left.
+\frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_1}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_1}{\partial x_k} \right) dW_{s_1}^i ds_2 \right) \right] \\
+& + \mathcal{O} ((\tau - \rho)^{5/2})
+\end{aligned}
+\end{equation}
 

### 2014-09-01 reducible_SDE/paper/SIADS_draft.tex
added more to paper (more math)

@@
 %
-In the first example, our data will form a strip where the fast and slow remain uncoupled,
+In the first example, our transformation $\mathbf{f}$ will be the identity function and fast and slow remain uncoupled,
 \begin{equation}
@@
 \begin{equation}
-\| \Xi_2 - \Xi_1 \|^2_2 = \sum_{i=1}^m \| \xi_{i,2} - \xi_{i,1} \|^2 = \sum_{i=1}^m \|x_{i,2} - x_{i,1} \|^2 + \epsilon \sum_{i=m+1}^n \|x_{i,2} - x_{i,1}\|^2
+\| \Xi_2 - \Xi_1 \|^2_2 = \sum_{i=1}^m \| \xi_{2,i} - \xi_{1,i} \|^2 = \sum_{i=1}^m \|x_{2,i} - x_{1,i} \|^2 + \epsilon \sum_{i=m+1}^n \|x_{2,i} - x_{1,i}\|^2
 \end{equation}
@@
 
-Let $g = f^{-1}: \mathbb{R}^m \mapsto \mathbb{R}^n$, and let $Y_1 = f(X_1)$, and $Y_2 = f(X_2)$.
+Let $\mathbf{g} = \mathbf{f}^{-1}: \mathbb{R}^d \mapsto \mathbb{R}^n$, and let $Y_1 = f(X_1)$, and $Y_2 = f(X_2)$.
 %
@@
 Then, by Taylor expansion, we obtain
-\begin{eqnarray}
-X_2^i &=& X_1^i + \sum_j g_j^i (Y_1) (Y^j_2 - Y^j_1 ) 
-+ \frac{1}{2} \sum_{kl}  g^i_{kl} (Y_1) (Y^k_2 - Y^k_1)(Y^l_2 - Y^l_1) \\
-&&+ \frac{1}{6} \sum_{klm}  g^i_{klm} (Y_1) (Y^k_2 - Y^k_1)(Y^l_2 - Y^l_1) (Y^m_2 - Y^m_1) 
+\begin{equation}
+\begin{aligned}
+x_{2,i} &=& 
+x_{1,i} 
++ \sum_{j=1}^d \left. \frac{\partial g_i}{\partial y_j} \right|_{Y_1} (y_{2,j} - y_{1,j} ) 
++ \frac{1}{2} \sum_{j=1}^d \sum_{k=1}^d \left. \frac{\partial^2 g_i}{\partial y_j \partial y_k} \right|_{Y_1} (y_{2,j} - y_{1,j}) (y_{2,k} - y_{1,k}) \\
+&&+ \frac{1}{6} \sum_{j=1}^d \sum_{k=1}^d \sum_{l=1}^d \left. \frac{\partial^3 g_i}{\partial y_j \partial y_k \partial y_l} \right|_{Y_1} (y_{2,j} - y_{1,j}) (y_{2,k} - y_{1,k}) (y_{2,l} - y_{1,l})
 + \mathcal{O}( \|Y_2 - Y_1\|^4 )
-\end{eqnarray}
+\end{aligned}
+\end{equation}
 
-Subsitituting the Taylor expansion into the quadratic form, we obtain
-\begin{eqnarray}
-&&\| X_2 - X_1 \|^2 \\
-&=& \frac{1}{2} (Y_2 - Y_1 )^T ((J J^T)^{-1} (Y_1) + (J J^T)^{-1}(Y_2)) (Y_2 - Y_1 ) \\
-&& + \frac{1}{2} \sum_{ijkl} \left( g_j^i (Y_1) g^i_{kl} (Y_1) - g_j^i (Y_2) g^i_{kl} (Y_2) \right) (Y^j_2 - Y^j_1 )  (Y^k_2 - Y^k_1)(Y^l_2 - Y^l_1) \\
-&& + \frac{1}{8} \sum_{ijklm}  \left( g^i_{jk} (Y_1) g^i_{lm} (Y_1) + g^i_{jk} (Y_2) g^i_{lm} (Y_2)  \right) (Y^j_2 - Y^j_1) (Y^k_2 - Y^k_1) (Y^l_2 - Y^l_1) (Y^m_2 - Y^m_1) \\
-&& + \frac{1}{6} \sum_{ijklm}  \left( g^i_{j} (Y_1) g^i_{klm} (Y_1) + g^i_{j} (Y_2) g^i_{klm} (Y_2)  \right)(Y^j_2 - Y^j_1) (Y^k_2 - Y^k_1) (Y^l_2 - Y^l_1) (Y^m_2 - Y^m_1) \\
-&& + \mathcal{O} (\|Y_1 - Y_2 \|^6 ) 
-\end{eqnarray}
+Substituting the Taylor expansion into the quadratic form, we obtain
+\begin{equation}
+\begin{aligned}
+\| X_2 - X_1 \|^2_2 
+=& \frac{1}{2} (Y_2 - Y_1 )^T \left(\left.(J J^T)^{-1}\right|_{Y_1} + \left.(J J^T)^{-1}\right|_{Y_2}\right) (Y_2 - Y_1 ) \\
+& + \frac{1}{2} \sum_{i=1}^n \sum_{jkl=1}^{d} \left( \left.
+\frac{\partial g_i}{\partial y_j} \right|_{Y_1} \left.
+\frac{\partial^2 g_i}{\partial y_k \partial y_l} \right|_{Y_1} - \left.
+\frac{\partial g_i}{\partial y_j} \right|_{Y_2} \left.
+\frac{\partial^2 g_i}{\partial y_k \partial y_l} \right|_{Y_2} \right) (y_{2,j} - y_{1,j})  (y_{2,k} - y_{1,k})(y_{2,l} - y_{1,l}) \\
+& + \frac{1}{8} \sum_{i=1}^n \sum_{jklm=1}^d  \left( \left.
+\frac{\partial^2 g_i}{\partial y_j \partial y_k} \right|_{Y_1} \left.
+\frac{\partial^2 g_i}{\partial y_l \partial y_m} \right|_{Y_1} + \left.
+\frac{\partial^2 g_i}{\partial y_j \partial y_k} \right|_{Y_2} \left.
+\frac{\partial^2 g_i}{\partial y_l \partial y_m} \right|_{Y_2} \right) (y_{2,j} - y_{1,j})  (y_{2,k} - y_{1,k})(y_{2,l} - y_{1,l})(y_{2,m} - y_{1,m}) \\
+& + \frac{1}{6}  \sum_{i=1}^n \sum_{jklm=1}^d  \left( \left.
+\frac{\partial g_i}{\partial y_j} \right|_{Y_1} \left.
+\frac{\partial^3 g_i}{ \partial y_k \partial y_l \partial y_m} \right|_{Y_1} + \left.
+\frac{\partial g_i}{\partial y_j} \right|_{Y_2} \left.
+\frac{\partial^3 g_i}{ \partial y_k \partial y_l \partial y_m} \right|_{Y_2} \right) (y_{2,j} - y_{1,j})  (y_{2,k} - y_{1,k})(y_{2,l} - y_{1,l})(y_{2,m} - y_{1,m}) \\
+& + \mathcal{O} (\|Y_1 - Y_2 \|^6 ) 
+\end{aligned}
+\end{equation}
 
@@
 \begin{equation}
- \| Y_2 - Y_1 \|^2_M = \frac{1}{2} (Y_2 - Y_1 )^T ((J J^T)^{-1} (Y_1) + (J J^T)^{-1}(Y_2)) (Y_2 - Y_1 ) 
+ \| Y_2 - Y_1 \|^2_M = \frac{1}{2} (Y_2 - Y_1 )^T \left(\left.(J J^T)^{-1}\right|_{Y_1} + \left.(J J^T)^{-1}\right|_{Y_2}\right) (Y_2 - Y_1 )
 \end{equation}
@@
 
-For our general multi scale + function SDE formulation,
+To compute the Mahalanobis distance, we require $J J^T$. 
+%
+From \cite{...}, we know that $JJ^T = C$, where $C$ is the local covariance of the observed stochastic process. 
+%
+Therefore, we would like to estimate the local covariance at a point $X_t$.
 
-For our general multi scale + function SDE formulation, 
+We know that, for a stochastic process following \eqref{eq:general_SDE} and observed through a function $\mathbf{f}$, the covariance $C$ is given by 
+\begin{equation}
+C_{jk}(X_t) = 
+\sum_{i=1}^m \left. \frac{\partial f_j}{\partial x_i} \right|_{X_t} \left. \frac{\partial f_k}{\partial x_i} \right|_{X_t} 
++ \frac{1}{\epsilon} \sum_{i=m+1}^n \left. \frac{\partial f_j}{\partial x_i} \right|_{X_t} \left. \frac{\partial f_k}{\partial x_i} \right|_{X_t} 
+\end{equation}
+
+We will use local simulation bursts to estimate the local covariance.
+%
+Let $X_\tau$ denote the samples at time $\tau > t$. 
+%
+For the general SDE formulation outlined in \eqref{eq:general_SDE}, we write the estimated covariance $\hat{C}$ as
 \begin{equation}
 \begin{aligned}
-&\mathbb{E} \left[ f_1(X_\tau) f_2(X_\tau) \right] - \mathbb{E}[f_1(X_\tau)]\mathbb{E}[f_2(X_\tau)] = \\
-& \sum_{i=1}^m \left. \frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \left. \frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i \right)^2 \right] \\
-& + \frac{1}{\epsilon} \sum_{i=m+1}^n \left.
-\frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \left.
-\frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i \right)^2 \right] \\
-& + \sum_{i=1}^m \left.
-\frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i  \right) \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_2}{\partial x_i^2} dW_{s_1}^i dW_{s_2}^i  \right) \right] \\
-& + \frac{1}{\epsilon^{3/2}} \sum_{i=m+1}^n \left.
-\frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i  \right) \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_2}{\partial x_i^2} dW_{s_1}^i dW_{s_2}^i  \right) \right] \\
-& + \sum_{i=1}^m \left.
-\frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i \right)  \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_1}{\partial x_i^2}  dW_{s_1}^i dW_{s_2}^i \right) \right] \\
-& + \frac{1}{\epsilon^{3/2}} \sum_{i=m+1}^n \left.
-\frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i \right)  \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_1}{\partial x_i^2}  dW_{s_1}^i dW_{s_2}^i \right) \right] \\
-&+ \mathcal{O} ((\tau - \rho)^2 ) 
-\end{aligned}
-\end{equation}
-
-For the case when $f$ is linear,
-\begin{equation}
-\begin{aligned}
-& \mathbb{E} [f_1(X_\tau) f_2(X_\tau) ] - \mathbb{E}[f_1(X_\tau)]\mathbb{E}[f_2(X_\tau)]= \\
-=  &  \sum_{i=1}^m  \left.
-\frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \left.
-\frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)^2  \right] \\
-& + \frac{1}{\epsilon} \sum_{i=m+1}^n  \left.
-\frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \left.
-\frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)^2  \right] \\
- & + \sum_{i=1}^m \left.
-\frac{\partial f_1}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_2}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_2}{\partial x_k} \right)  dW_{s_1}^i ds_2 \right) \right] \\
- & + \frac{1}{\epsilon} \sum_{i=m+1}^n \left.
-\frac{\partial f_1}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_2}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_2}{\partial x_k} \right)  dW_{s_1}^i ds_2 \right) \right] \\
- & + \sum_{i=1}^m  \left.
-\frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_1}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_1}{\partial x_k} \right) dW_{s_1}^i ds_2 \right) \right] \\
-& + \frac{1}{\epsilon} \sum_{i=m+1}^n  \left.
-\frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_1}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_1}{\partial x_k} \right) dW_{s_1}^i ds_2 \right) \right] \\
-& + \mathcal{O} ((\tau - \rho)^{5/2})
- \end{aligned}
-\end{equation}
-
-Therefore, we have
-\begin{equation}
-\begin{aligned}
-e_c = &  \sum_{i=1}^m \left.
-\frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i  \right) \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_2}{\partial x_i^2} dW_{s_1}^i dW_{s_2}^i  \right) \right] \\
-& + \frac{1}{\epsilon^{3/2}} \sum_{i=m+1}^n \left.
-\frac{\partial f_1}{\partial x_i} \right|_{X_\rho} \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i  \right) \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_2}{\partial x_i^2} dW_{s_1}^i dW_{s_2}^i  \right) \right] \\
-& + \sum_{i=1}^m \left.
-\frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i \right)  \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_1}{\partial x_i^2}  dW_{s_1}^i dW_{s_2}^i \right) \right] \\
-& + \frac{1}{\epsilon^{3/2}} \sum_{i=m+1}^n \left.
-\frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left( \int_\rho^\tau dW_s^i \right)  \left( \int_\rho^\tau \int_\rho^{s_2} \frac{\partial^2 f_1}{\partial x_i^2}  dW_{s_1}^i dW_{s_2}^i \right) \right] \\
-&+ \mathcal{O} ((\tau - \rho)^2 ) 
+\hat{C}_{jk} (X_t, \tau - t) = & 
+\frac{1}{\tau - t} \left(  \mathbb{E} \left[ f_j(X_\tau) f_k(X_\tau) \right] - \mathbb{E}[f_j(X_\tau)]\mathbb{E}[f_k(X_\tau)]   \right)\\
+= & \sum_{i=1}^m \left. \frac{\partial f_j}{\partial x_i} \right|_{X_t} \left. \frac{\partial f_k}{\partial x_i} \right|_{X_t}  \\
+& + \frac{1}{\epsilon} \sum_{i=m+1}^n \left. \frac{\partial f_j}{\partial x_i} \right|_{X_t} \left. \frac{\partial f_k}{\partial x_i} \right|_{X_t} \\
+& + \frac{1}{\tau - t} \sum_{i=1}^m \left.
+\frac{\partial f_j}{\partial x_i} \right|_{X_t} \mathbb{E} \left[ \left( \int_t^\tau dW_s^i  \right) \left( \int_t^\tau \int_t^{s_2} \frac{\partial^2 f_k}{\partial x_i^2} dW_{s_1}^i dW_{s_2}^i  \right) \right] \\
+& + \frac{1}{\epsilon^{3/2} (\tau - t)} \sum_{i=m+1}^n \left.
+\frac{\partial f_j}{\partial x_i} \right|_{X_t} \mathbb{E} \left[ \left( \int_t^\tau dW_s^i  \right) \left( \int_t^\tau \int_t^{s_2} \frac{\partial^2 f_k}{\partial x_i^2} dW_{s_1}^i dW_{s_2}^i  \right) \right] \\
+& + \frac{1}{\tau - t} \sum_{i=1}^m \left.
+\frac{\partial f_k}{\partial x_i} \right|_{X_t}  \mathbb{E} \left[ \left( \int_t^\tau dW_s^i \right)  \left( \int_t^\tau \int_t^{s_2} \frac{\partial^2 f_j}{\partial x_i^2}  dW_{s_1}^i dW_{s_2}^i \right) \right] \\
+& + \frac{1}{\epsilon^{3/2} (\tau - t)} \sum_{i=m+1}^n \left.
+\frac{\partial f_k}{\partial x_i} \right|_{X_t}  \mathbb{E} \left[ \left( \int_t^\tau dW_s^i \right)  \left( \int_t^\tau \int_t^{s_2} \frac{\partial^2 f_j}{\partial x_i^2}  dW_{s_1}^i dW_{s_2}^i \right) \right] \\
+&+ \mathcal{O} ((\tau - t) ) 
 \end{aligned}
@@
 %
-when $\frac{\partial f}{\partial x_i} \ne 0$, and 
+However, for the specific case when $\mathbf{f}$ is linear, the $\mathcal{O}((\tau - t)^{3/2})$ terms go to 0, and we must consider higher-order terms. 
+%
+In this case, 
+\begin{equation}
+\begin{aligned}
+\hat{C}_{jk} (X_t, \tau - t) = & 
+\frac{1}{\tau - t} \left(  \mathbb{E} \left[ f_j(X_\tau) f_k(X_\tau) \right] - \mathbb{E}[f_j(X_\tau)]\mathbb{E}[f_k(X_\tau)]   \right)\\
+= & \sum_{i=1}^m \left. \frac{\partial f_j}{\partial x_i} \right|_{X_t} \left. \frac{\partial f_k}{\partial x_i} \right|_{X_t}  \\
+& + \frac{1}{\epsilon} \sum_{i=m+1}^n \left. \frac{\partial f_j}{\partial x_i} \right|_{X_t} \left. \frac{\partial f_k}{\partial x_i} \right|_{X_t} \\
+ & + \frac{1}{\tau - t} \sum_{i=1}^m \left.
+\frac{\partial f_j}{\partial x_i} \right|_{X_t}  \mathbb{E} \left[ \left(\int_t^\tau dW_s^i \right)  \left(\int_t^\tau \int_{t}^{s_2} \left(  \sum_{l=1}^m \frac{\partial a_l}{\partial x_i} \frac{\partial f_k}{\partial x_l}  + \frac{1}{\epsilon} \sum_{l=m+1}^m \frac{\partial a_l}{\partial x_i} \frac{\partial f_k}{\partial x_l} \right)  dW_{s_1}^i ds_2 \right) \right] \\
+ & + \frac{1}{\epsilon (\tau - t)} \sum_{i=m+1}^n \left.
+\frac{\partial f_j}{\partial x_i} \right|_{X_t}  \mathbb{E} \left[ \left(\int_t^\tau dW_s^i \right)  \left(\int_t^\tau \int_{t}^{s_2} \left(  \sum_{l=1}^m \frac{\partial a_l}{\partial x_i} \frac{\partial f_k}{\partial x_l}  + \frac{1}{\epsilon} \sum_{l=m+1}^m \frac{\partial a_l }{\partial x_i} \frac{\partial f_k}{\partial x_l} \right)  dW_{s_1}^i ds_2 \right) \right] \\
+ & + \frac{1}{\tau - t} \sum_{i=1}^m  \left.
+\frac{\partial f_k}{\partial x_i} \right|_{X_t}  \mathbb{E} \left[ \left(\int_t^\tau dW_s^i \right)  \left(\int_t^\tau \int_{t}^{s_2} \left(  \sum_{l=1}^m \frac{\partial a_l}{\partial x_i} \frac{\partial f_j}{\partial x_l}  + \frac{1}{\epsilon} \sum_{l=m+1}^m \frac{\partial a_l }{\partial x_i} \frac{\partial f_j}{\partial x_l} \right) dW_{s_1}^i ds_2 \right) \right] \\
+& + \frac{1}{\epsilon (\tau - t)} \sum_{i=m+1}^n  \left.
+\frac{\partial f_k}{\partial x_i} \right|_{X_t}  \mathbb{E} \left[ \left(\int_t^\tau dW_s^i \right)  \left(\int_t^\tau \int_{t}^{s_2} \left(  \sum_{l=1}^m \frac{\partial a_l}{\partial x_i} \frac{\partial f_j}{\partial x_l}  + \frac{1}{\epsilon} \sum_{l=m+1}^m \frac{\partial a_l }{\partial x_i} \frac{\partial f_j}{\partial x_l} \right) dW_{s_1}^i ds_2 \right) \right] \\
+& + \mathcal{O} ((\tau - t)^{3/2})
+ \end{aligned}
+\end{equation}
+%
+For both cases,  we define the error in the estimation as
 %
@@
 \begin{aligned}
-e_c =  &  \sum_{i=1}^m \left.
-\frac{\partial f_1}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_2}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_2}{\partial x_k} \right)  dW_{s_1}^i ds_2 \right) \right] \\
- & + \frac{1}{\epsilon} \sum_{i=m+1}^n \left.
-\frac{\partial f_1}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_2}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_2}{\partial x_k} \right)  dW_{s_1}^i ds_2 \right) \right] \\
- & + \sum_{i=1}^m  \left.
-\frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_1}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_1}{\partial x_k} \right) dW_{s_1}^i ds_2 \right) \right] \\
-& + \frac{1}{\epsilon} \sum_{i=m+1}^n  \left.
-\frac{\partial f_2}{\partial x_i} \right|_{X_\rho}  \mathbb{E} \left[ \left(\int_\rho^\tau dW_s^i \right)  \left(\int_\rho^\tau \int_{\rho}^{s_2} \left(  \sum_{k=1}^m \frac{\partial a_k}{\partial x_i} \frac{\partial f_1}{\partial x_k}  + \frac{1}{\epsilon} \sum_{k=m+1}^m \frac{\partial a_k }{\partial x_i} \frac{\partial f_1}{\partial x_k} \right) dW_{s_1}^i ds_2 \right) \right] \\
-& + \mathcal{O} ((\tau - \rho)^{5/2})
+e_C (X_t, \tau - t) = \| \hat{C} (X_t, \tau - t)  - C(X_t, \tau - t) \|
 \end{aligned}
@@
 \section{Results}
+
+For illustrative purposes, we will consider the following two-dimensional SDE as a specific example of \eqref{eq:general_SDE}. 
+\begin{equation}
+\begin{aligned}
+dx_1 &= adt + dW_1\\
+dx_2 &= -\frac{x_2}{\epsilon} dt + \frac{1}{\sqrt{\epsilon}} dW_2
+\end{aligned}
+\end{equation}
+%
+where $a$ is a constant of order 1 (we will take $a=3$). 
+%
+So $x_1$ is the slow variable, and $x_2$ is a fast noise whose equilibrium measure is bounded and $\mathcal{O}(1)$.
+%
+We would like to recover $x_1$ using data-driven techniques.
+
+\subsection{Linear functions}
+
+In the first example, our transformation $\mathbf{f}$ will be the identity function and fast and slow remain uncoupled,
+\begin{equation}
+\mathbf{f}(x_1, x_2) = \begin{bmatrix} x_1 \\ x_2 \end{bmatrix}
+\end{equation}
+%
+and
+%
+\begin{equation}
+\mathbf{g}(y_1, y_2) = \mathbf{f}^{-1} (y_1, y_2) = \begin{bmatrix} y_1 \\ y_2 \end{bmatrix}
+\end{equation}
+
+In this case, $e_M = 0$, as the second- and higher-order derivatives of $\mathbf{g}$ are all 0.
+
+\begin{equation}
+\begin{aligned}
+e_{C, jk} (X_t, \tau - t) = & 
+\frac{1}{\tau - t} \sum_{i=1}^m \left.
+\frac{\partial f_j}{\partial x_i} \right|_{X_t}  \mathbb{E} \left[ \left(\int_t^\tau dW_s^i \right)  \left(\int_t^\tau \int_{t}^{s_2} \left(  \sum_{l=1}^m \frac{\partial a_l}{\partial x_i} \frac{\partial f_k}{\partial x_l}  + \frac{1}{\epsilon} \sum_{l=m+1}^m \frac{\partial a_l}{\partial x_i} \frac{\partial f_k}{\partial x_l} \right)  dW_{s_1}^i ds_2 \right) \right] \\
+ & + \frac{1}{\epsilon (\tau - t)} \sum_{i=m+1}^n \left.
+\frac{\partial f_j}{\partial x_i} \right|_{X_t}  \mathbb{E} \left[ \left(\int_t^\tau dW_s^i \right)  \left(\int_t^\tau \int_{t}^{s_2} \left(  \sum_{l=1}^m \frac{\partial a_l}{\partial x_i} \frac{\partial f_k}{\partial x_l}  + \frac{1}{\epsilon} \sum_{l=m+1}^m \frac{\partial a_l }{\partial x_i} \frac{\partial f_k}{\partial x_l} \right)  dW_{s_1}^i ds_2 \right) \right] \\
+ & + \frac{1}{\tau - t} \sum_{i=1}^m  \left.
+\frac{\partial f_k}{\partial x_i} \right|_{X_t}  \mathbb{E} \left[ \left(\int_t^\tau dW_s^i \right)  \left(\int_t^\tau \int_{t}^{s_2} \left(  \sum_{l=1}^m \frac{\partial a_l}{\partial x_i} \frac{\partial f_j}{\partial x_l}  + \frac{1}{\epsilon} \sum_{l=m+1}^m \frac{\partial a_l }{\partial x_i} \frac{\partial f_j}{\partial x_l} \right) dW_{s_1}^i ds_2 \right) \right] \\
+& + \frac{1}{\epsilon (\tau - t)} \sum_{i=m+1}^n  \left.
+\frac{\partial f_k}{\partial x_i} \right|_{X_t}  \mathbb{E} \left[ \left(\int_t^\tau dW_s^i \right)  \left(\int_t^\tau \int_{t}^{s_2} \left(  \sum_{l=1}^m \frac{\partial a_l}{\partial x_i} \frac{\partial f_j}{\partial x_l}  + \frac{1}{\epsilon} \sum_{l=m+1}^m \frac{\partial a_l }{\partial x_i} \frac{\partial f_j}{\partial x_l} \right) dW_{s_1}^i ds_2 \right) \right] \\
+& + \mathcal{O} ((\tau - t)^{3/2})
+ \end{aligned}
+\end{equation}
+
+\subsection{Nonlinear functions}
+
+In the second example, our data will be warped into half-moon shapes
+\begin{equation}
+\mathbf{f_2}(x_1, x_2) = 
+\begin{bmatrix} 
+x_1 + x_2^2 \\
+x_2
+\end{bmatrix}
+\end{equation}
 

### 2014-09-02 reducible_SDE/paper/SIADS_draft.tex
added more to paper

@@
 \begin{equation}
-\mathbb{E} \left| \frac{1}{T} \int_t^{t + T} a_i(x_1, \dots, x_m, X_{m+1}(s), \dots, X_n(s)) ds - \overline{a} (x_1, \dots, x_m) \right| < \kappa_i (T) 
+\mathbb{E} \left| \frac{1}{T} \int_t^{t + T} a_i(x_1, \dots, x_m, x_{m+1}(s), \dots, x_n(s)) ds - \overline{a} (x_1, \dots, x_m) \right| < \kappa_i (T) 
 \end{equation}
 for $1 \le i \le m$, where $\kappa_i((T) \rightarrow 0 $ as $T \rightarrow \infty$, then the averaging principle \cite{...} states that we can write an effective equation in the slow variables $x_1, \dots, x_m$.
-%
+
 We will consider systems that can be written as deterministic (perhaps nonlinear) bi-Lipshitz functions $\mathbf{f}: \mathbb{R}^n \mapsto \mathbb{R}^d$ of the SDE in \eqref{eq:general_SDE}.
 %
-Our goal is to recover the slow variables $x_1, \dots, x_m$ using data-driven techniques.
+Our goal is to recover the slow variables $x_1, \dots, x_m$  from observations of $\mathbf{f}(x_1, \dots, x_n)$ using data-driven techniques.
 
@@
 \begin{equation} \label{eq:mahalanobis}
-\| \mathbf{f}(\Xi_1) - \mathbf{f}(\Xi_2) \|^2_M = \| \Xi_1 - \Xi_2 \|^2_2 + \mathcal{O}(\| \mathbf{f}(\Xi_1) - \mathbf{f}(\Xi_2) \|^4_2)
+\| \mathbf{f}(X_1) - \mathbf{f}(X_2) \|^2_M = \| \Xi_1 - \Xi_2 \|^2_2 + \mathcal{O}(\| \mathbf{f}(X_1) - \mathbf{f}(X_2) \|^4_2)
 \end{equation}
-where the components of $\Xi$ have uncoupled noise with unit variance,
-i.e., the variables $\xi_1, \dots, \xi_n$ are governed by the following SDEs:
+where the components of $\Xi$ are sampled from a stochastic process with uncoupled noise with unit variance,
+i.e., $\Xi = \begin{bmatrix} \xi_1 & \xi_2& \cdots & \xi_n \end{bmatrix}^T$, where $\xi_1, \dots, \xi_n$ are governed by the following SDEs:
 \begin{equation} \label{eq:NIV_formulation}
@@
 \end{equation}
-and $\Xi = \begin{bmatrix} \xi_1 & \xi_2& \cdots & \xi_n \end{bmatrix}^T$.
 
@@
 x_i = \xi_i, \: &1 \le i \le m \\
-x_i = \frac{\xi_i}{\sqrt{\epsilon}} \: &1 \le i \le m
+x_i = \frac{\xi_i}{\sqrt{\epsilon}}, \: &m+1 \le i \le n
 \end{aligned}
@@
 
-\begin{figure}
+\begin{figure}[t]
 \epsfig{width=\textwidth, file=schematic.eps}
@@
 
-Our approximation of the pairwise distances will rely on the Taylor expansion of the measurement function $f$. 
+Our approximation of the pairwise distances will rely on the Taylor expansion of the measurement function $\mathbf{f}$. 
 %
@@
 
-Given the true covariance, we can write the Euclidean distance $\|X_2 - X_1 \|$ using Taylor expansion as...
-
-Talk about fourth-order error terms, say we want to choose the kernel scale such that these error terms are negligible. 
+We want to calculate the distance $\|\Xi_2 - \Xi_1\|$ in terms of the observations $\mathbf{f}(X)$. 
 %
-The error terms involve curvature of $f$, etc.
-
-Let $\mathbf{g} = \mathbf{f}^{-1}: \mathbb{R}^d \mapsto \mathbb{R}^n$, and let $Y_1 = f(X_1)$, and $Y_2 = f(X_2)$.
+Let $\mathbf{g} = \mathbf{f}^{-1}: \mathbb{R}^d \mapsto \mathbb{R}^n$.
 %
-Again, we let $g_j^i = \frac{\partial g^i}{\partial Y^j}$.
+By \eqref{eq:general_rescale}, we have
+\begin{equation}
+\begin{aligned}
+g_i(Y) = x_i = \xi_i,& \: 1 \le i \le m \\
+\sqrt{\epsilon} g_i(Y) = \sqrt{\epsilon} x_i = \xi_i,& \: m+1 \le i \le n
+\end{aligned}
+\end{equation}
+%
+Let $Y_1 = \mathbf{f}(X_1)$, and $Y_2 = \mathbf{f}(X_2)$.
 %
@@
 \begin{aligned}
-x_{2,i} &=& 
-x_{1,i} 
+\xi_{2,i} &=& 
+\xi_{1,i} 
 + \sum_{j=1}^d \left. \frac{\partial g_i}{\partial y_j} \right|_{Y_1} (y_{2,j} - y_{1,j} ) 
@@
 \end{equation}
-
-Substituting the Taylor expansion into the quadratic form, we obtain
+for $1 \le i \le m$, and 
 \begin{equation}
 \begin{aligned}
-\| X_2 - X_1 \|^2_2 
-=& \frac{1}{2} (Y_2 - Y_1 )^T \left(\left.(J J^T)^{-1}\right|_{Y_1} + \left.(J J^T)^{-1}\right|_{Y_2}\right) (Y_2 - Y_1 ) \\
-& + \frac{1}{2} \sum_{i=1}^n \sum_{jkl=1}^{d} \left( \left.
+\xi_{2,i} &=& 
+\xi_{1,i} 
++ \sqrt{\epsilon} \left[ \sum_{j=1}^d \left. \frac{\partial g_i}{\partial y_j} \right|_{Y_1} (y_{2,j} - y_{1,j} ) 
++ \frac{1}{2} \sum_{j=1}^d \sum_{k=1}^d \left. \frac{\partial^2 g_i}{\partial y_j \partial y_k} \right|_{Y_1} (y_{2,j} - y_{1,j}) (y_{2,k} - y_{1,k}) \right. \\
+&& \left.+ \frac{1}{6} \sum_{j=1}^d \sum_{k=1}^d \sum_{l=1}^d \left.
+\frac{\partial^3 g_i}{\partial y_j \partial y_k \partial y_l} \right|_{Y_1} (y_{2,j} - y_{1,j}) (y_{2,k} - y_{1,k}) (y_{2,l} - y_{1,l}) \right]
++ \mathcal{O}( \|Y_2 - Y_1\|^4 )
+\end{aligned}
+\end{equation}
+for $m+1 \le i \le n$. 
+%
+Substituting the Taylor expansion into the quadratic form, we obtain
+\begin{equation} \label{eq:distance_taylor_expansion}
+\begin{aligned}
+&\| \Xi_2 - \Xi_1 \|^2_2 = \\
+& \frac{1}{2} (Y_2 - Y_1 )^T \left(\left.(J I_\epsilon J^T)^{-1}\right|_{Y_1} + \left.( J I_\epsilon J^T)^{-1}\right|_{Y_2}\right) (Y_2 - Y_1 ) \\
+& + \frac{1}{2} \sum_{i=1}^m \sum_{jkl=1}^{d} \left( \left.
 \frac{\partial g_i}{\partial y_j} \right|_{Y_1} \left.
@@
 \frac{\partial^2 g_i}{\partial y_k \partial y_l} \right|_{Y_2} \right) (y_{2,j} - y_{1,j})  (y_{2,k} - y_{1,k})(y_{2,l} - y_{1,l}) \\
-& + \frac{1}{8} \sum_{i=1}^n \sum_{jklm=1}^d  \left( \left.
+& + \frac{\epsilon}{2} \sum_{i=m+1}^n \sum_{jkl=1}^{d} \left( \left.
+\frac{\partial g_i}{\partial y_j} \right|_{Y_1} \left.
+\frac{\partial^2 g_i}{\partial y_k \partial y_l} \right|_{Y_1} - \left.
+\frac{\partial g_i}{\partial y_j} \right|_{Y_2} \left.
+\frac{\partial^2 g_i}{\partial y_k \partial y_l} \right|_{Y_2} \right) (y_{2,j} - y_{1,j})  (y_{2,k} - y_{1,k})(y_{2,l} - y_{1,l}) \\
+& + \frac{1}{8} \sum_{i=1}^m \sum_{jklm=1}^d  \left( \left.
 \frac{\partial^2 g_i}{\partial y_j \partial y_k} \right|_{Y_1} \left.
@@
 \frac{\partial^2 g_i}{\partial y_l \partial y_m} \right|_{Y_2} \right) (y_{2,j} - y_{1,j})  (y_{2,k} - y_{1,k})(y_{2,l} - y_{1,l})(y_{2,m} - y_{1,m}) \\
-& + \frac{1}{6}  \sum_{i=1}^n \sum_{jklm=1}^d  \left( \left.
+& + \frac{\epsilon}{8} \sum_{i=m+1}^n \sum_{jklm=1}^d  \left( \left.
+\frac{\partial^2 g_i}{\partial y_j \partial y_k} \right|_{Y_1} \left.
+\frac{\partial^2 g_i}{\partial y_l \partial y_m} \right|_{Y_1} + \left.
+\frac{\partial^2 g_i}{\partial y_j \partial y_k} \right|_{Y_2} \left.
+\frac{\partial^2 g_i}{\partial y_l \partial y_m} \right|_{Y_2} \right) (y_{2,j} - y_{1,j})  (y_{2,k} - y_{1,k})(y_{2,l} - y_{1,l})(y_{2,m} - y_{1,m}) \\
+& + \frac{1}{6}  \sum_{i=1}^m \sum_{jklm=1}^d  \left( \left.
+\frac{\partial g_i}{\partial y_j} \right|_{Y_1} \left.
+\frac{\partial^3 g_i}{ \partial y_k \partial y_l \partial y_m} \right|_{Y_1} + \left.
+\frac{\partial g_i}{\partial y_j} \right|_{Y_2} \left.
+\frac{\partial^3 g_i}{ \partial y_k \partial y_l \partial y_m} \right|_{Y_2} \right) (y_{2,j} - y_{1,j})  (y_{2,k} - y_{1,k})(y_{2,l} - y_{1,l})(y_{2,m} - y_{1,m}) \\
+& + \frac{\epsilon}{6}  \sum_{i=m+1}^n \sum_{jklm=1}^d  \left( \left.
 \frac{\partial g_i}{\partial y_j} \right|_{Y_1} \left.
@@
 \end{equation}
+where $J$ is the Jacobian of $\mathbf{f}$, and $I_\epsilon$ is a diagonal matrix with $I_{\epsilon,ii} = 1$ for $1 \le i \le m$, and $I_{\epsilon,ii} = \frac{1}{\epsilon}$ for $m+1 \le i \le n$.
 
-In general, we do not have access to $f$, $g$, or any of its derivatives.
+In general, we do not have access to $\mathbf{f}$, $\mathbf{g}$, or any of its derivatives.
 %
-However, we note that $J J^T(Y_1)$ is the local covariance of $Y_1$, which we can empirically estimate from data (this will be discussed in a subsequent section).
+However, we note that $\left.
+J I_\epsilon J^T\right|_Y$ is the local covariance of the stochastic process at $Y$, which we can empirically estimate from data (this will be discussed in a subsequent section).
 %
@@
 \begin{equation}
- \| Y_2 - Y_1 \|^2_M = \frac{1}{2} (Y_2 - Y_1 )^T \left(\left.(J J^T)^{-1}\right|_{Y_1} + \left.(J J^T)^{-1}\right|_{Y_2}\right) (Y_2 - Y_1 )
+ \| Y_2 - Y_1 \|^2_M = \frac{1}{2} (Y_2 - Y_1 )^T \left(\left.(J I_\epsilon J^T)^{-1}\right|_{Y_1} + \left.(J I_\epsilon J^T)^{-1}\right|_{Y_2}\right) (Y_2 - Y_1 )
 \end{equation}
@@
 The error in the approximation is defined as
-\begin{equation}
-e_M(Y_1, Y_2) = \| X_2 - X_1 \|^2_2 - \| Y_2 - Y_1 \|^2_M 
+\begin{equation} \label{eq:mahanaobis_error}
+e_M(Y_1, Y_2) = \| \Xi_2 - \Xi_1 \|^2_2 - \| Y_2 - Y_1 \|^2_M 
 \end{equation}
 
-We can bound the error by
-\begin{equation}
-| e_M(Y_1, Y_2)  | \le n \left( \left| K_1 K_2 \right| + \left| \frac{ K_2^2}{4} \right|  + \left| \frac{K_1 K_3}{3} \right|  \right) \| Y_2 - Y_1 \| ^4  
-+ \mathcal{O} (\|Y_1 - Y_2 \|^6 ) 
-\end{equation}
-%
-where
-%
-\begin{equation}
-\begin{aligned}
-K_1 &= \sup_{i,j,Y} |g_j^i(Y)|\\
-K_2 &= \sup_{i,j,k,Y} |g_{jk}^i(Y)|\\
-K_3 &= \sup_{i,j,k,l,Y} |g_{jkl}^i(Y)|
-\end{aligned}
-\end{equation}
+%We can bound the error by
+%\begin{equation}
+%| e_M(Y_1, Y_2)  | \le n \left( \left| K_1 K_2 \right| + \left| \frac{ K_2^2}{4} \right|  + \left| \frac{K_1 K_3}{3} \right|  \right) \| Y_2 - Y_1 \| ^4  
+%+ \mathcal{O} (\|Y_1 - Y_2 \|^6 ) 
+%\end{equation}
+%%
+%where
+%%
+%\begin{equation}
+%\begin{aligned}
+%K_1 &= \sup_{i,j,Y} |g_j^i(Y)|\\
+%K_2 &= \sup_{i,j,k,Y} |g_{jk}^i(Y)|\\
+%K_3 &= \sup_{i,j,k,l,Y} |g_{jkl}^i(Y)|
+%\end{aligned}
+%\end{equation}
 
@@
 
-To compute the Mahalanobis distance, we require $J J^T$. 
+To compute the Mahalanobis distance, we require $J I_\epsilon J^T$. 
 %
-From \cite{...}, we know that $JJ^T = C$, where $C$ is the local covariance of the observed stochastic process. 
+From \cite{...}, we know that $J I_\epsilon J^T = C$, where $C$ is the local covariance of the observed stochastic process. 
 %
@@
 %
-Let $X_\tau$ denote the samples at time $\tau > t$. 
+Let $X_\tau$ denote samples at time $\tau > t$. 
 %
 For the general SDE formulation outlined in \eqref{eq:general_SDE}, we write the estimated covariance $\hat{C}$ as
-\begin{equation}
+\begin{equation} \label{eq:estimated_cov}
 \begin{aligned}
@@
 In this case, 
-\begin{equation}
+\begin{equation} \label{eq:estimated_cov_linear_f}
 \begin{aligned}
@@
 %
-\begin{equation}
+\begin{equation} \label{eq:cov_error}
 \begin{aligned}
-e_C (X_t, \tau - t) = \| \hat{C} (X_t, \tau - t)  - C(X_t, \tau - t) \|
+e_C (X_t, \tau - t) = \hat{C} (X_t, \tau - t)  - C(X_t, \tau - t) 
 \end{aligned}
 \end{equation}
+
+\section{Empirical Estimation of Parameters}
+
+Given data governed by an SDE as in \eqref{eq:general_SDE}, we would like to empirically set the simulation and analysis parameters.
+%
+There are two parameters: $\sigma_{kernel}$, which is the diffusion maps kernel scale, and $\tau - t$, which is the timescale of the ``burst'' used to estimate the covariance.
+
+$\sigma_{kernel}$ controls which distances we ``trust''.
+%
+Therefore, we want the Mahalanobis distance approximation to be accurate within a ball of radius $\sigma_{kernel}$. 
+%
+From \eqref{eq:distance_taylor_expansion}, we can see that, for $e_M$ to be small, we want $\|Y_2 - Y_1 \|^2_M$ to vary linearly with $\|Y_2 - Y_1\|^2_2$.
+
+For $e_C$ to be accurate, we want $\hat{C}$ to not vary as a function of $\tau - t$.
+
+We can use simulations to check both of these conditions empirically.
 
@@
 
+For this example, the analytical covariance is
+ \begin{equation}
+C(X_t) = 
+\begin{bmatrix}
+1 & 0 \\
+0 & \frac{1}{\epsilon}
+\end{bmatrix}
+\end{equation}
+
 \subsection{Linear functions}
@@
 %
-and
-%
 \begin{equation}
@@
 
+%\begin{equation}
+%\begin{aligned}
+%e_{C, jk} (X_t, \tau - t) = & 
+%\frac{1}{\tau - t} \left.
+\frac{\partial f_j}{\partial x_1} \right|_{X_t}  \mathbb{E} \left[ \left(\int_t^\tau dW_s^1 \right)  \left(\int_t^\tau \int_{t}^{s_2} \left(  \frac{\partial a_1}{\partial x_1} \frac{\partial f_k}{\partial x_1}  + \frac{1}{\epsilon} \frac{\partial a_2}{\partial x_1} \frac{\partial f_k}{\partial x_2} \right)  dW_{s_1}^1 ds_2 \right) \right] \\
+% & + \frac{1}{\epsilon (\tau - t)} \left.
+\frac{\partial f_j}{\partial x_2} \right|_{X_t}  \mathbb{E} \left[ \left(\int_t^\tau dW_s^2 \right)  \left(\int_t^\tau \int_{t}^{s_2} \left(\frac{\partial a_1}{\partial x_2} \frac{\partial f_k}{\partial x_1}  + \frac{1}{\epsilon} \frac{\partial a_2 }{\partial x_2} \frac{\partial f_k}{\partial x_2} \right)  dW_{s_1}^2 ds_2 \right) \right] \\
+% & + \frac{1}{\tau - t}  \left.
+\frac{\partial f_k}{\partial x_1} \right|_{X_t}  \mathbb{E} \left[ \left(\int_t^\tau dW_s^1 \right)  \left(\int_t^\tau \int_{t}^{s_2} \left(  \frac{\partial a_1}{\partial x_1} \frac{\partial f_j}{\partial x_1}  + \frac{1}{\epsilon} \frac{\partial a_2}{\partial x_1} \frac{\partial f_j}{\partial x_2} \right) dW_{s_1}^1 ds_2 \right) \right] \\
+%& + \frac{1}{\epsilon (\tau - t)}  \left.
+\frac{\partial f_k}{\partial x_2} \right|_{X_t}  \mathbb{E} \left[ \left(\int_t^\tau dW_s^2 \right)  \left(\int_t^\tau \int_{t}^{s_2} \left(  \frac{\partial a_1}{\partial x_2} \frac{\partial f_j}{\partial x_1}  + \frac{1}{\epsilon} \frac{\partial a_2}{\partial x_2} \frac{\partial f_j}{\partial x_2} \right) dW_{s_1}^2 ds_2 \right) \right] \\
+%& + \mathcal{O} ((\tau - t)^{3/2})\\
+% = & 
+%\frac{1}{\epsilon^2 (\tau - t)} \left.
+\frac{\partial f_j}{\partial x_2} \right|_{X_t}  \mathbb{E} \left[ \left(\int_t^\tau dW_s^2 \right)  \left(\int_t^\tau \int_{t}^{s_2} -\frac{\partial f_k}{\partial x_2} dW_{s_1}^2 ds_2 \right) \right] \\
+%& + \frac{1}{\epsilon^2 (\tau - t)}  \left.
+\frac{\partial f_k}{\partial x_2} \right|_{X_t}  \mathbb{E} \left[ \left(\int_t^\tau dW_s^2 \right)  \left(\int_t^\tau \int_{t}^{s_2} -\frac{\partial f_j}{\partial x_2} dW_{s_1}^2 ds_2 \right) \right] \\
+%& + \mathcal{O} ((\tau - t)^{3/2})
+% \end{aligned}
+%\end{equation}
+
+From \eqref{eq:estimated_cov_linear_f}, we find that 
 \begin{equation}
 \begin{aligned}
-e_{C, jk} (X_t, \tau - t) = & 
-\frac{1}{\tau - t} \sum_{i=1}^m \left.
-\frac{\partial f_j}{\partial x_i} \right|_{X_t}  \mathbb{E} \left[ \left(\int_t^\tau dW_s^i \right)  \left(\int_t^\tau \int_{t}^{s_2} \left(  \sum_{l=1}^m \frac{\partial a_l}{\partial x_i} \frac{\partial f_k}{\partial x_l}  + \frac{1}{\epsilon} \sum_{l=m+1}^m \frac{\partial a_l}{\partial x_i} \frac{\partial f_k}{\partial x_l} \right)  dW_{s_1}^i ds_2 \right) \right] \\
- & + \frac{1}{\epsilon (\tau - t)} \sum_{i=m+1}^n \left.
-\frac{\partial f_j}{\partial x_i} \right|_{X_t}  \mathbb{E} \left[ \left(\int_t^\tau dW_s^i \right)  \left(\int_t^\tau \int_{t}^{s_2} \left(  \sum_{l=1}^m \frac{\partial a_l}{\partial x_i} \frac{\partial f_k}{\partial x_l}  + \frac{1}{\epsilon} \sum_{l=m+1}^m \frac{\partial a_l }{\partial x_i} \frac{\partial f_k}{\partial x_l} \right)  dW_{s_1}^i ds_2 \right) \right] \\
- & + \frac{1}{\tau - t} \sum_{i=1}^m  \left.
-\frac{\partial f_k}{\partial x_i} \right|_{X_t}  \mathbb{E} \left[ \left(\int_t^\tau dW_s^i \right)  \left(\int_t^\tau \int_{t}^{s_2} \left(  \sum_{l=1}^m \frac{\partial a_l}{\partial x_i} \frac{\partial f_j}{\partial x_l}  + \frac{1}{\epsilon} \sum_{l=m+1}^m \frac{\partial a_l }{\partial x_i} \frac{\partial f_j}{\partial x_l} \right) dW_{s_1}^i ds_2 \right) \right] \\
-& + \frac{1}{\epsilon (\tau - t)} \sum_{i=m+1}^n  \left.
-\frac{\partial f_k}{\partial x_i} \right|_{X_t}  \mathbb{E} \left[ \left(\int_t^\tau dW_s^i \right)  \left(\int_t^\tau \int_{t}^{s_2} \left(  \sum_{l=1}^m \frac{\partial a_l}{\partial x_i} \frac{\partial f_j}{\partial x_l}  + \frac{1}{\epsilon} \sum_{l=m+1}^m \frac{\partial a_l }{\partial x_i} \frac{\partial f_j}{\partial x_l} \right) dW_{s_1}^i ds_2 \right) \right] \\
-& + \mathcal{O} ((\tau - t)^{3/2})
+e_{C, 11}(X_t, \tau - t) &=& 
+e_{C, 12}(X_t, \tau - t) = 
+e_{C, 21}(X_t, \tau - t) = 0 \\
+e_{C, 22}(X_t, \tau - t) &=& 
+\frac{-2}{\epsilon^2 (\tau - t)} \mathbb{E} \left[ \left(\int_t^\tau dW_s^2 \right)  \left(\int_t^\tau \int_{t}^{s_2} -dW_{s_1}^2 ds_2 \right) \right]  
++ \mathcal{O} ((\tau - t)^{3/2})
 \end{aligned}
 \end{equation}
+%
+Therefore, $\|e_C \| = \mathcal{O}\left(\frac{\tau - t}{\epsilon^2} \right)$ and so we need to choose $\tau - t < \epsilon^2$.
 
@@
 \begin{equation}
-\mathbf{f_2}(x_1, x_2) = 
+\mathbf{f}(x_1, x_2) = 
 \begin{bmatrix} 
@@
 \end{bmatrix}
+\end{equation}
+
+\begin{equation}
+\mathbf{g}(y_1, y_2) = \mathbf{f}^{-1} (y_1, y_2) = \begin{bmatrix} y_1 - y_2^2 \\ y_2 \end{bmatrix}
+\end{equation}
+
+From \eqref{eq:distance_taylor_expansion}, we have 
+\begin{equation}
+e_M(Y_1, Y_2) = (\epsilon-1) (y_{2,2} - y_{1,2})^4 
++ \mathcal{O} (\|Y_1 - Y_2 \|^6 ) 
+\end{equation}
+
+%Therefore, we have
+%\begin{equation}
+%\begin{aligned}
+%\frac{\partial f_1}{\partial x_1} &=& 1 \\
+%\frac{\partial f_1}{\partial x_2} &=& 2 x_2 \\
+%\frac{\partial f_2}{\partial x_1} &=& 0 \\
+%\frac{\partial f_2}{\partial x_2} &=& 1 \\
+%\frac{\partial^2 f_1}{\partial x_1^2} = \frac{\partial^2 f_1}{\partial x_1 \partial x_2} &=& 0 \\
+%\frac{\partial^2 f_1}{\partial x_2^2} &=& 2 \\
+%\frac{\partial^2 f_2}{\partial x_1^2} = \frac{\partial^2 f_2}{\partial x_1 \partial x_2} = \frac{\partial^2 f_2}{\partial x_2^2} &=& 0 \\
+%\end{aligned}
+%\end{equation}
+%
+%\begin{equation}
+%\begin{aligned}
+%e_{C, jk} (X_t, \tau - t) = & 
+%\frac{1}{\tau - t} \left.
+\frac{\partial f_j}{\partial x_1} \right|_{X_t} \mathbb{E} \left[ \left( \int_t^\tau dW_s^1  \right) \left( \int_t^\tau \int_t^{s_2} \frac{\partial^2 f_k}{\partial x_1^2} dW_{s_1}^1 dW_{s_2}^1  \right) \right] \\
+%& + \frac{1}{\epsilon^{3/2} (\tau - t)} \left.
+\frac{\partial f_j}{\partial x_2} \right|_{X_t} \mathbb{E} \left[ \left( \int_t^\tau dW_s^2  \right) \left( \int_t^\tau \int_t^{s_2} \frac{\partial^2 f_k}{\partial x_2^2} dW_{s_1}^2 dW_{s_2}^2  \right) \right] \\
+%& + \frac{1}{\tau - t} \left.
+\frac{\partial f_k}{\partial x_1} \right|_{X_t}  \mathbb{E} \left[ \left( \int_t^\tau dW_s^1 \right)  \left( \int_t^\tau \int_t^{s_2} \frac{\partial^2 f_j}{\partial x_1^2}  dW_{s_1}^1 dW_{s_2}^1 \right) \right] \\
+%& + \frac{1}{\epsilon^{3/2} (\tau - t)} \left.
+\frac{\partial f_k}{\partial x_2} \right|_{X_t}  \mathbb{E} \left[ \left( \int_t^\tau dW_s^2 \right)  \left( \int_t^\tau \int_t^{s_2} \frac{\partial^2 f_j}{\partial x_2^2}  dW_{s_1}^2 dW_{s_2}^2 \right) \right] \\
+%&+ \mathcal{O} ((\tau - t) ) \\
+%= & 
+%\frac{1}{\epsilon^{3/2} (\tau - t)} \left.
+\frac{\partial f_j}{\partial x_2} \right|_{X_t} \mathbb{E} \left[ \left( \int_t^\tau dW_s^2  \right) \left( \int_t^\tau \int_t^{s_2} \frac{\partial^2 f_k}{\partial x_2^2} dW_{s_1}^2 dW_{s_2}^2  \right) \right] \\
+%& + \frac{1}{\epsilon^{3/2} (\tau - t)} \left.
+\frac{\partial f_k}{\partial x_2} \right|_{X_t}  \mathbb{E} \left[ \left( \int_t^\tau dW_s^2 \right)  \left( \int_t^\tau \int_t^{s_2} \frac{\partial^2 f_j}{\partial x_2^2}  dW_{s_1}^2 dW_{s_2}^2 \right) \right] \\
+%&+ \mathcal{O} ((\tau - t) ) 
+%\end{aligned}
+%\end{equation}
+
+From \eqref{eq:estimated_cov}, we find that
+\begin{equation}
+\begin{aligned}
+e_{C,11}(X_t, \tau-t) &=& \frac{8}{\epsilon^{3/2} (\tau - t)} X_t \mathbb{E} \left[ \left( \int_t^\tau dW_s^2  \right) \left( \int_t^\tau \int_t^{s_2} dW_{s_1}^2 dW_{s_2}^2  \right) \right] 
++ \mathcal{O} ((\tau - t) ) \\
+e_{C,12}(X_t, \tau-t) &=& \frac{2}{\epsilon^{3/2} (\tau - t)}  \mathbb{E} \left[ \left( \int_t^\tau dW_s^2 \right)  \left( \int_t^\tau \int_t^{s_2}  dW_{s_1}^2 dW_{s_2}^2 \right) \right] 
++ \mathcal{O} ((\tau - t) ) \\
+e_{C,21}(X_t, \tau-t) &=& \frac{2}{\epsilon^{3/2} (\tau - t)} \mathbb{E} \left[ \left( \int_t^\tau dW_s^2  \right) \left( \int_t^\tau \int_t^{s_2} dW_{s_1}^2 dW_{s_2}^2  \right) \right] 
++ \mathcal{O} ((\tau - t) ) \\
+e_{C,22}(X_t, \tau-t) &=& \mathcal{O} ((\tau - t) ) 
+\end{aligned}
 \end{equation}

### 2014-09-03 reducible_SDE/paper/SIADS_paper/NIV_return_dist.m
SIADS draft

@@ -1,0 +1,23 @@
+function [V, D, eps, Dis] = NIV_return_dist(data, inv_c, eps, N, alpha)
+
+[m, n] = size(data);
+
+Dis = zeros(m);
+%h = waitbar(0, 'Please wait');
+for i=1:m
+    %waitbar(i/m, h);
+    Dis(:,i) = sum((data - repmat(data(i,:),m,1)) * inv_c(:,:,i) .* (data - repmat(data(i,:),m,1)), 2);
+end
+%close(h);
+Dis = (Dis + Dis')/2;
+
+if eps == 0
+    eps = median(Dis(:));
+end
+
+if alpha == 0
+    [V, D] = dmaps(Dis, eps, N);
+else
+    [V, D] = dmaps_weight(Dis, alpha, eps, N);
+end
+

### 2014-09-03 reducible_SDE/paper/SIADS_paper/covariances.m

@@ -1,0 +1,21 @@
+function [c, inv_c, ranks] = covariances(data, D, deltat)
+
+% Find neighborhoods
+[M, N, P] = size(data);
+
+% compute covariances
+c = zeros(N, N, P);
+inv_c = zeros(N, N, P);
+
+ranks = zeros(P, 1);
+
+for i = 1:P    
+    c_tmp = cov(data(:,:,i))/deltat;
+    
+    [U, S, V] = svd(c_tmp);
+
+    inv_c(:,:,i) = U(:,1:D) * diag(1./diag(S(1:D,1:D))) * V(:,1:D)';
+    ranks(i) = rank(c_tmp);
+    c(:,:,i) = c_tmp;
+end
+

### 2014-09-03 reducible_SDE/paper/SIADS_paper/make_figures_linear.m

@@ -1,0 +1,136 @@
+clear all
+close all
+
+%%
+
+epsilon = 1e-3;
+
+dt_burst = 1e-9;
+dt = 1e-4;
+
+nsteps = 3000;
+nsteps_burst = 50;
+
+rng(123);
+[data_init, data_burst_init, t] = simulate_SDE(nsteps, dt, nsteps_burst, dt_burst, epsilon);
+
+make_fig(6, 6);
+scatter(data_init(:,1),data_init(:,2),50,t, '.')
+axis tight
+axis square
+xlabel('x_1')
+ylabel('x_2')
+saveas(gcf, 'data_init.eps');
+
+
+%%
+
+W = squareform(pdist(data_init)).^2;
+sigma = median(W(:));
+[V, D] = dmaps(W, sigma, 10);
+
+make_fig(6, 6);
+scatter(data_init(:,1),data_init(:,2),50,V(:,2), '.')
+axis tight
+axis square
+xlabel('x_1')
+ylabel('x_2')
+saveas(gcf, 'data_linear_DMAPS.eps');
+
+%%
+
+dim = 2;
+[c, inv_c, ranks] = covariances(data_burst_init, dim, dt_burst);
+
+[V, D, eps, Dis] = NIV_return_dist(data_init, inv_c, 0, 10, 0);
+
+make_fig(6, 6);
+scatter(data_init(:,1),data_init(:,2),50,V(:,2), '.')
+axis tight
+axis square
+xlabel('x_1')
+ylabel('x_2')
+saveas(gcf, 'data_linear_NIV.eps');
+
+%%
+
+dt_tmp = 10.^(-8:-2);
+dt_tmp2 = logspace(-8, -2, 100);
+
+figure; 
+loglog(dt_tmp2, (1+1/epsilon)*ones(size(dt_tmp2)), '-b')
+hold on
+loglog(dt_tmp2, dt_tmp2/epsilon^2, '-r')
+
+figure; 
+loglog(dt_tmp2, 1+1/epsilon+dt_tmp2/epsilon^2)
+hold on
+loglog(dt_tmp2, 1+1/epsilon-dt_tmp2/epsilon^2)
+
+%%
+
+ntest_points = 10;
+data_init_tmp = data_init(1:10:10*ntest_points,:);
+norm_c = zeros(length(dt_tmp), ntest_points);
+
+for i=1:length(dt_tmp)
[…]

### 2014-09-03 reducible_SDE/paper/SIADS_paper/make_figures_nonlinear.m

@@ -1,0 +1,195 @@
+clear all
+close all
+
+%%
+
+epsilon = 1e-3;
+
+dt_burst = 1e-9;
+dt = 1e-4;
+
+nsteps = 3000;
+nsteps_burst = 50;
+
+rng(123);
+[data_init, data_burst_init, t] = simulate_SDE(nsteps, dt, nsteps_burst, dt_burst, epsilon);
+
+f = @(x) [x(:,1)+x(:,2).^2 x(:,2)];
+
+
+data1 = f(data_init);
+data1_burst = zeros(size(data_burst_init));
+for i=1:nsteps
+    data1_burst(:,:,i) = f(data_burst_init(:,:,i));
+end
+
+make_fig(6, 6);
+scatter(data_init(:,1),data_init(:,2),50,t, '.')
+axis tight
+axis square
+xlabel('x_1')
+ylabel('x_2')
+
+make_fig(6, 6);
+scatter(data1(:,1),data1(:,2),50,t, '.')
+axis tight
+axis square
+xlabel('x_1')
+ylabel('x_2')
+saveas(gcf, 'data_init_nonlinear.eps');
+
+%%
+
+W = squareform(pdist(data1)).^2;
+sigma = median(W(:))*100;
+[V, D] = dmaps(W, sigma, 10);
+
+make_fig(6, 6);
+scatter(data1(:,1),data1(:,2),50,V(:,2), '.')
+axis tight
+axis square
+xlabel('x_1')
+ylabel('x_2')
+saveas(gcf, 'data_nonlinear_DMAPS.eps');
+
+%%
+
+dim = 2;
+[c, inv_c, ranks] = covariances(data1_burst, dim, dt_burst);
+
+[V, D, eps, Dis] = NIV_return_dist(data1, inv_c, 0.01, 10, 0);
+
+make_fig(6, 6);
+scatter(data1(:,1),data1(:,2),50,V(:,2), '.')
+axis tight
+axis square
+xlabel('x_1')
+ylabel('x_2')
+saveas(gcf, 'data_nonlinear_NIV.eps');
+
+%%
+
+dt_tmp = 10.^(-8:-2);
+dt_tmp2 = logspace(-8, -2, 100);
+
+figure; 
+loglog(dt_tmp2, (1+1/epsilon)*ones(size(dt_tmp2)), '-b')
+hold on
+loglog(dt_tmp2, dt_tmp2/epsilon^2, '-r')
+
+figure; 
[…]

### 2014-09-05 reducible_SDE/paper/SIADS_paper/make_figures_linear.m
updated siads paper after talking with Ronen

@@ -5,7 +5,7 @@ close all
 
 epsilon = 1e-3;
 
-dt_burst = 1e-9;
+dt_burst = 1e-6;
 dt = 1e-4;
 
 nsteps = 3000;
@@ -20,8 +20,9 @@ axis tight
 axis square
 xlabel('x_1')
 ylabel('x_2')
-saveas(gcf, 'data_init.eps');
-
+h = colorbar;
+xlabel(h,'t');
+saveas(gcf, 'data_init.eps', 'epsc');
 
 %%
 
@@ -35,7 +36,9 @@ axis tight
 axis square
 xlabel('x_1')
 ylabel('x_2')
-saveas(gcf, 'data_linear_DMAPS.eps');
+h = colorbar;
+xlabel(h,'\phi_1');
+saveas(gcf, 'data_linear_DMAPS.eps','epsc');
 
 %%
 
@@ -50,17 +53,28 @@ axis tight
 axis square
 xlabel('x_1')
 ylabel('x_2')
-saveas(gcf, 'data_linear_NIV.eps');
+h = colorbar;
+xlabel(h,'\phi_1');
+saveas(gcf, 'data_linear_NIV.eps','epsc');
 
 %%
 
-dt_tmp = 10.^(-8:-2);
-dt_tmp2 = logspace(-8, -2, 100);
+dt_tmp = 10.^(-6:-2);
+dt_tmp2 = logspace(-6, -2, 100);
 
-figure; 
+make_fig(6, 6);
 loglog(dt_tmp2, (1+1/epsilon)*ones(size(dt_tmp2)), '-b')
 hold on
 loglog(dt_tmp2, dt_tmp2/epsilon^2, '-r')
+loglog(dt_tmp2, 1+1/epsilon+dt_tmp2/epsilon^2, '-k')
+axis square
+xlabel('\tau - t')
+ylabel('$\| C \|$', 'interpreter','latex')
+h = legend('$\|C\|$', '$\| $e$_C\|$', '$\| \hat{C} \| = \| C\| + \|$e$_C \|$', 'location','southwest');
+ set(h,'Interpreter','latex');
+ set(h,'fontsize', 6);
+saveas(gcf, 'C_dt_analytical_linear.eps', 'epsc');
+
 
 figure; 
 loglog(dt_tmp2, 1+1/epsilon+dt_tmp2/epsilon^2)
[…]

### 2014-09-09 reducible_SDE/paper/SIADS_paper/make_figures_nonlinear.m
edited paper more

@@ -100,8 +100,8 @@ dy = logspace(-4, 1.5, 100);
 make_fig(3, 3.5);
 loglog(dy, 1.5*dy.^2,'-b')
 hold on
-loglog(dy, 3*dy.^4,'-r')
-loglog(dy, 1.5*dy.^2+3*dy.^4,'-k')
+loglog(dy, 1*dy.^4,'-r')
+loglog(dy, 1.5*dy.^2+1*dy.^4,'-k')
 for i=1:length(kernel_samplepoints)
     loglog(dy, kernel_samplepoints(i)*ones(size(dy)), 'linestyle', ':', 'color', 0.5*ones(1,3))
 end
@@ -114,7 +114,7 @@ set(h,'Interpreter','latex');
 set(h,'fontsize', 4);
 saveas(gcf, 'dist_dy_analytical_nonlinear.eps', 'epsc');
 
-
+return
 %%
 W = squareform(pdist(data1));
 minW = min(W(W > 0));

### 2014-09-12 reducible_SDE/paper/SIADS_paper/make_figures_linear.m
updated all the math

@@ -69,8 +69,9 @@ dt_burst_samplepoints = [1e-6 1e-5 1e-3];
 make_fig(3, 3);
 loglog(dt_tmp2, (1+1/epsilon)*ones(size(dt_tmp2)), '-b')
 hold on
-loglog(dt_tmp2, 2*dt_tmp2/epsilon^2, '-r')
-loglog(dt_tmp2, 1+1/epsilon+2*dt_tmp2/epsilon^2, '-k')
+% loglog(dt_tmp2, 2*dt_tmp2/epsilon^2, '-r')
+loglog(dt_tmp2, dt_tmp2/(epsilon^2), '-r')
+loglog(dt_tmp2, 1+1/epsilon+dt_tmp2/(epsilon^2), '-k')
 for i=1:length(dt_burst_samplepoints)
     loglog(dt_burst_samplepoints(i)*ones(size(dt_tmp2)), dt_tmp2/epsilon^2, 'linestyle', ':', 'color', 0.5*ones(1,3))
 end
@@ -82,6 +83,8 @@ h = legend('$\|C\|$', '$\| $e$_C\|$', '$\| C\| + \|$e$_C \|$', 'location','south
 set(h,'Interpreter','latex');
 set(h,'fontsize', 6);
 saveas(gcf, 'C_dt_analytical_linear.eps', 'epsc');
+
+return
 
 %%
 

### 2014-09-12 reducible_SDE/paper/SIADS_paper/make_figures_nonlinear.m

@@ -114,7 +114,6 @@ set(h,'Interpreter','latex');
 set(h,'fontsize', 4);
 saveas(gcf, 'dist_dy_analytical_nonlinear.eps', 'epsc');
 
-return
 %%
 W = squareform(pdist(data1));
 minW = min(W(W > 0));
@@ -153,8 +152,10 @@ dt_tmp2 = logspace(-9, 0, 100);
 make_fig(3, 3);
 loglog(dt_tmp2, (1+4/epsilon+1/epsilon)*ones(size(dt_tmp2)), '-b')
 hold on
-loglog(dt_tmp2, 8*sqrt(dt_tmp2)/epsilon^(3/2), '-r')
-loglog(dt_tmp2, (1+4/epsilon+1/epsilon)+8*sqrt(dt_tmp2)/epsilon^(3/2), '-k')
+% loglog(dt_tmp2, 8*sqrt(dt_tmp2)/epsilon^(3/2), '-r')
+loglog(dt_tmp2, 15*dt_tmp2/epsilon^2, '-r')
+% loglog(dt_tmp2, (1+4/epsilon+1/epsilon)+8*sqrt(dt_tmp2)/epsilon^(3/2), '-k')
+loglog(dt_tmp2, (1+4/epsilon+1/epsilon)+15*dt_tmp2/epsilon^2, '-k')
 for i=1:length(dt_burst_samplepoints)
     loglog(dt_burst_samplepoints(i)*ones(size(dt_tmp2)), 8*sqrt(dt_tmp2)/epsilon^(3/2), 'linestyle', ':', 'color', 0.5*ones(1,3))
 end
@@ -166,6 +167,8 @@ h = legend('$\|C\|$', '$\| $e$_C\|$', '$\| C\| + \|$e$_C \|$', 'location','south
 set(h,'Interpreter','latex');
 set(h,'fontsize', 6);
 saveas(gcf, 'C_dt_analytical_nonlinear.eps', 'epsc');
+
+return
 
 %%
 

### 2014-09-12 reducible_SDE/paper/SIADS_paper/make_figures_linear.m
updated all paper stuff

@@ -79,7 +79,7 @@ axis([10^-7 10^-2 10^0 10^4])
 axis square
 xlabel('\delta t')
 ylabel('Analytical covariance contributions')
-h = legend('$\|C\|$', '$\| $e$_C\|$', '$\| C\| + \|$e$_C \|$', 'location','southwest');
+h = legend('$\|C\|$', '$\| $e$_C\|$', '$\| C\| + \|$e$_C \|$', 'location','southeast');
 set(h,'Interpreter','latex');
 set(h,'fontsize', 6);
 saveas(gcf, 'C_dt_analytical_linear.eps', 'epsc');

### 2014-09-12 reducible_SDE/paper/SIADS_paper/make_figures_nonlinear.m

@@ -111,7 +111,7 @@ xlabel('$\|\vec{y}_2 - \vec{y}_1\|_2$', 'interpreter','latex')
 ylabel('Analytical distance contributions')
 h = legend('$\|\vec{y}_2 - \vec{y}_1\|_M^2$', '$\| $e$_M\|$',  '$ \|\vec{y}_2 - \vec{y}_1\|_M^2 + \|$e$_M \|$', 'location','southeast');
 set(h,'Interpreter','latex');
-set(h,'fontsize', 4);
+set(h,'fontsize', 6);
 saveas(gcf, 'dist_dy_analytical_nonlinear.eps', 'epsc');
 
 %%

### 2014-09-15 reducible_SDE/paper/SIADS_paper/make_figures_linear.m
added edits after talking with ROnen

@@ -79,7 +79,7 @@ axis([10^-7 10^-2 10^0 10^4])
 axis square
 xlabel('\delta t')
 ylabel('Analytical covariance contributions')
-h = legend('$\|C\|$', '$\| $e$_C\|$', '$\| C\| + \|$e$_C \|$', 'location','southeast');
+h = legend('$\|C\|$', '$\| $E$_C\|$', '$\| C\| + \|$e$_C \|$', 'location','southeast');
 set(h,'Interpreter','latex');
 set(h,'fontsize', 6);
 saveas(gcf, 'C_dt_analytical_linear.eps', 'epsc');

### 2014-12-06 chemotaxis/swiss_roll_example.tex
added swiss roll and testng for haromnics for chemotaxis paper

\documentclass[12pt]{article}

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage[margin=1in]{geometry}

\begin{document}

\section{Local Linear Regression}

Given the eigenvectors $\phi_1, \phi_2, \dots, \phi_{n-1} \in \mathbb{R}^n$, we would like to automatically deduce which ones are harmonics/functions of another. 
%
To accomplish this, we will fit a function $f(\phi_1, \dots, \phi_k)$ to $\phi_{k+1}$. 
%
If the resulting fit is good/accurate, then $\phi_{k+1}$ is most likely a harmonic of $\phi_1, \dots, \phi_k$.

We will use a local linear function 
\begin{equation}
f_{k+1}(\phi_1, \dots, \phi_k) = a_0 + a_1 \phi_1 + \dots + a_k \phi_k
\end{equation}
where the coefficients $a_0, \dots, a_k$ are functions of $\phi_1, \dots, \phi_k$.

To find these coefficients, we want to minimize the local weighted sum of squares
\begin{equation} \label{eq:opt_problem}
\sum_{i=1}^n w_i(x) (\phi_{k+1}(x) - f_{k+1} (\phi_1(x), \dots, \phi_{k}(x))
\end{equation}
where $w$ is a weighting function.
%
We will use
%
\begin{itemize}
\item $w_i(x) = \exp \left( - \frac{\|\Phi_{k}(x) - \Phi_{k} (x_i) \|^2}{\epsilon^2} \right)$, where $\Phi_k(x) = \begin{bmatrix}
\phi_1(x) \\
\vdots \\
\phi_k(x)
\end{bmatrix} $
\item $\epsilon = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k}(x)$
\end{itemize}

Let $\hat{f}_{k+1}(\phi_1(x_i), \dots, \phi_k(x_i))$ denote the minimum achieved in \eqref{eq:opt_problem}, when the minimization is performed over $x_j$ with $j \ne i$. 
%
We then define the leave-one-out- cross-validation error as
\begin{equation}
r_{k+1} = \sum_{i=1}^n \left( \phi_{k+1} (x_i) - \hat{f}_{k+1}(\phi_1(x_i), \dots, \phi_k(x_i)) \right)^2
\end{equation}
%
A small value of $r_{k+1}$ means that $\phi_{k+1}$ can be accurately approximated from $\phi_1, \dots, \phi_k$, and is therefore a harmonic of a previous mode.
%
A large value of $r_{k+1}$ indicates that $\phi_{k+1}$ parameterizes a new direction in the data.

\section{Eigenvalues}

The eigenvalues of the Laplacian with Neumann boundary conditions on a $d$-dimensional rectangular domain are given by
\begin{equation}
\mu_{k,l} = \frac{k^2 \pi^2}{L_l^2}
\end{equation}
for $k=1, 2, \dots$ and $l=1, 2, \dots, d$, and $L_l$ is the length of the domain along the $l^{th}$ dimension. 
%
These eigenvalues are related to the eigenvalues we compute in diffusion maps by 
\begin{equation}
\lambda_{k,l} = e^{-\mu_{k,l} t} = e^{-\frac{k^2 \pi^2}{L_l^2} t}
\end{equation}
%
Therefore, 
\begin{equation}
\sqrt{\frac{\log(\lambda_{1,l_1})}{\log(\lambda_{1,l_2})}} = 
\frac{L_{l_2}}{L_{l_1}}
\end{equation}
and so looking at the ratio of the log of the eigenvalues (modulo harmonics) should inform us as to the lengths of the different directions in the manifold.

\section{Example: Swiss Roll}

\centering
\includegraphics[height=3in]{swissroll1}
\includegraphics[height=3in]{swissroll2}\\
\includegraphics[height=2in]{swissroll1_cv}
\includegraphics[height=2in]{swissroll2_cv}\\
\includegraphics[height=3in]{swissroll1_color1}
\includegraphics[height=3in]{swissroll2_color1}\\
\includegraphics[height=3in]{swissroll1_color2}
\includegraphics[height=3in]{swissroll2_color2}

For first case: $\lambda_1 = 0.9982$, $\lambda_2 = 0.9959$

For second case: $\lambda_1 = 0.9977$, $\lambda_5 = 0.9652$.

\section{Example: Chemotaxis Data}

\centering
\includegraphics[height=2in]{lle_errors1}
\includegraphics[height=2in]{lle_errors2}

\includegraphics[height=2in]{lle_evals1}
\includegraphics[height=2in]{lle_evals2}

\includegraphics[height=2in]{lle_embed1_1}
\includegraphics[height=2in]{lle_embed2_2}

\hfill 
\includegraphics[height=2in]{lle_embed1_2}

\end{document}

### 2014-12-06 chemotaxis/compute_residuals_DMAPS.m

@@ -1,0 +1,10 @@
+function res = compute_residuals_DMAPS(V, eps_med_scale)
+
+n = size(V, 2);
+
+res = zeros(n,1);
+res(2) = 1;
+
+for i=3:n
+    [~, res(i)] = local_linear_regression(V(:,i), V(:, 2:i-1), eps_med_scale);
+end

### 2014-12-06 chemotaxis/estimate_dimension.m

@@ -1,0 +1,13 @@
+function d = estimate_dimension(V, D, res_tol, lambda_tol)
+
+numD = find(abs(diag(D)) > lambda_tol, 1, 'last');
+
+d = 1;
+for i=3:numD
+    [~, res] = local_linear_regression(V(:,i), V(:, 2:i-1));
+    if res > res_tol^2
+        d = d + 1;
+    end
+end
+
+

### 2014-12-06 chemotaxis/local_linear_regression.m

@@ -1,0 +1,19 @@
+function [fx, res] = local_linear_regression(y, X, eps_med_scale)
+
+n = size(X, 1);
+
+K = squareform(pdist(X));
+eps = median(K(:))/eps_med_scale;
+W = exp(-K.^2 / eps^2);
+
+L = zeros(n);
+for i=1:n
+    Xx = [ones(size(X,1),1) X-repmat(X(i,:), n, 1)];
+    Wx = diag(W(i,:));
+%     A = inv(Xx'*Wx*Xx)*Xx'*Wx;
+    A = (Xx'*Wx*Xx)\(Xx'*Wx);
+    L(i,:) = A(1,:);
+end
+
+fx = L*y;
+res = sqrt(mean(((y-fx)./(1-diag(L))).^2)) / std(y);

### 2014-12-06 chemotaxis/simulate_jump_process_EMD_test_lle.m

@@ -1,0 +1,290 @@
+clear all
+close all
+
+rng(123);
+markersize = 500;
+
+
+%% simulations
+
+% rate of switching velocity
+lambda = 100;
+% speed of particles
+s = 10;
+sim_num = 1;
+
+% % rate of switching velocity
+% lambda = 2500;
+% % speed of particles
+% s = 50;
+% sim_num = 2;
+
+% number of particles
+N = 1000;
+% time step
+dt = 1;
+% maximum simulation time
+tmax = 10;
+nsteps = floor(tmax / dt) + 1;
+
+%initial probility of a particle moving to the right (vel=+1)
+nsim = 10;
+pinit = linspace(0.1, 0.9, nsim);
+
+min_start = -10;
+max_start = 10;
+
+% histogram parameters
+%nbins = 128;
+nbins = 32;
+x_hist = linspace(min_start-s*tmax, max_start+s*tmax, nbins);
+hist_all = zeros(nbins, nsteps*nsim);
+hist_left = zeros(nbins, nsteps*nsim);
+hist_right = zeros(nbins, nsteps*nsim);
+all_time = zeros(1, nsteps*nsim);
+all_p = zeros(1, nsteps*nsim);
+
+%%
+subplot_dim1 = floor(sqrt(nsim));
+subplot_dim2 = ceil(nsim / subplot_dim1);
+
+figure;
+for j=1:nsim
+    % simulate velocity jump process
+    [pos, vel, time] = simulate_jumps(N, tmax, dt, lambda, s, pinit(j));
+    %[pos, vel, time] = simulate_jumps_unifdist(N, tmax, dt, lambda, s, pinit(j), min_start, max_start);
+    
+    % histogram of all particle positions as a function of time
+    for i=1:nsteps
+        
+        [hist_all(:,(j-1)*nsteps+i), ~] = hist(pos(:,i),x_hist);
+        
+        idx = (vel(:,i) == -1);
+        [hist_left(:,(j-1)*nsteps+i), ~] = hist(pos(idx,i),x_hist);
+        
+        idx = (vel(:,i) == 1);
+        [hist_right(:,(j-1)*nsteps+i), ~] = hist(pos(idx,i),x_hist);
+        
+    end
+    
[…]

### 2014-12-06 chemotaxis/simulate_samples.m

@@ -1,0 +1,70 @@
+function W2 = simulate_samples(lambda, s)
+
+
+% number of particles
+N = 1000;
+% time step
+dt = 1;
+% maximum simulation time
+tmax = 10;
+nsteps = floor(tmax / dt) + 1;
+
+%initial probility of a particle moving to the right (vel=+1)
+nsim = 10;
+pinit = linspace(0.1, 0.9, nsim);
+
+% histogram parameters
+nbins = 32;
+x_hist = linspace(-s*tmax, s*tmax, nbins);
+hist_all = zeros(nbins, nsteps*nsim);
+hist_left = zeros(nbins, nsteps*nsim);
+hist_right = zeros(nbins, nsteps*nsim);
+all_time = zeros(1, nsteps*nsim);
+all_p = zeros(1, nsteps*nsim);
+
+%%
+
+for j=1:nsim
+    % simulate velocity jump process
+    [pos, vel, time] = simulate_jumps(N, tmax, dt, lambda, s, pinit(j));
+    %[pos, vel, time] = simulate_jumps_unifdist(N, tmax, dt, lambda, s, pinit(j), min_start, max_start);
+    
+    % histogram of all particle positions as a function of time
+    for i=1:nsteps
+        
+        [hist_all(:,(j-1)*nsteps+i), ~] = hist(pos(:,i),x_hist);
+        
+        idx = (vel(:,i) == -1);
+        [hist_left(:,(j-1)*nsteps+i), ~] = hist(pos(idx,i),x_hist);
+        
+        idx = (vel(:,i) == 1);
+        [hist_right(:,(j-1)*nsteps+i), ~] = hist(pos(idx,i),x_hist);
+        
+    end
+    
+    % store times and probabilities
+    all_time((j-1)*nsteps+1:j*nsteps) = time;
+    all_p((j-1)*nsteps+1:j*nsteps) = pinit(j);
+    
+end
+
+% use data after initial "relaxation" time
+idx = (all_time > 1);
+
+%% compute EMD
+
+W2 = zeros(nsteps*nsim);
+
+for i1=1:nsteps*nsim
+    for i2=1:i1-1
+        
+        %[emd_dist, ~]= emd_hat_gd_metric_mex(hist_all(:,i1), hist_all(:,i2), D_for_emd);
+        emd_dist = sum(abs(cumsum(hist_all(:,i1)) - cumsum(hist_all(:,i2))));
+        
[…]

### 2014-12-06 reducible_SDE/paper/SIADS_paper/make_schematic.m
updated figures in paper

@@ -1,0 +1,49 @@
+clear all
+close all
+
+rng(123);
+
+
+%%
+
+ntraj = 10;
+npoints = 1000;
+x = repmat(linspace(0, 1, npoints), ntraj, 1);
+
+y = zeros(size(x));
+for i=1:ntraj
+    y(i, :) = randn * exp(-10*x(i,:));
+end
+
+% figure;
+% plot(x, y, '.')
+% set(gca, 'xtick', [])
+% set(gca, 'ytick', [])
+
+make_fig(3, 3);
+scatter(x(:), y(:), 50, -x(:),'.')
+set(gca, 'xtick', [])
+set(gca, 'ytick', [])
+set(gca, 'ylim', [-1 1])
+% axis off
+saveas(gcf, 'schematic_DS1.eps', 'epsc');
+
+% figure;
+% plot(-(y'+1).*cos((pi/2)*x'), (y'+1).*sin((pi/2)*x'))
+
+%%
+
+npoints = 2000;
+
+x = linspace(0, 1, npoints);
+y = randn(size(x));
+
+make_fig(3, 3);
+scatter(x, smooth(y, 11), 50, -x,'.')
+set(gca, 'xtick', [])
+set(gca, 'ytick', [])
+set(gca, 'ylim', [-1 1])
+% axis off
+saveas(gcf, 'schematic_DS2.eps', 'epsc');
+
+

### 2014-12-06 reducible_SDE/paper/SIADS_paper/make_figures_nonlinear.m
updated multiscale paper

@@ -1,4 +1,4 @@
-clear all
+% clear all
 close all
 
 %%
@@ -116,7 +116,6 @@ set(h,'Interpreter','latex');
 set(h,'fontsize', 6);
 set(gca,'layer','top')
 saveas(gcf, 'dist_dy_analytical_nonlinear.eps', 'epsc');
-return
 
 %%
 W = squareform(pdist(data1));
@@ -177,8 +176,6 @@ set(h,'fontsize', 6);
 loglog(dt_tmp2, (1+4/epsilon+1/epsilon)*ones(size(dt_tmp2)), '-b')
 set(gca,'layer','top')
 saveas(gcf, 'C_dt_analytical_nonlinear.eps', 'epsc');
-
-return
 
 %%
 
@@ -252,20 +249,21 @@ for j=1:length(dt_burst_samplepoints)
 %         xlabel(h,'\phi_1');
 %         saveas(gcf, sprintf('data_nonlinear_NIV_dt%d_kernel%d.eps', j, j2), 'epsc');
         
-        make_fig(3, 3);
+        make_fig(2.5, 2);
         loglog(dt_tmp, mean(norm_c, 2),'.')
         hold on
-        loglog(dt_burst_samplepoints(j)*ones(size(dt_tmp2)), dt_tmp2/epsilon^2, 'linestyle', ':', 'color', 'r')
+        loglog(dt_burst_samplepoints(j)*ones(size(dt_tmp2)), dt_tmp2/epsilon^2, 'linestyle', '--', 'color', 'r')
         axis([1e-9 1e-2 1e1 1e5])
         axis square
         xlabel('\delta t')
         ylabel('$\| \hat{C} \|$', 'interpreter','latex')
         saveas(gcf, sprintf('C_dt_nonlinear_dt%d_kernel%d.eps', j, j2), 'epsc');
         
-        make_fig(3, 3);
+        make_fig(2.25, 2);
         loglog(dy2, Dis_avg, '.')
         hold on
-        loglog(dy, kernel_samplepoints(j2)*ones(size(dy)), 'linestyle', ':', 'color', 'r')
+        loglog(dy2, 0.5*dy2.^2, '-b')
+        loglog(dy, kernel_samplepoints(j2)*ones(size(dy)), 'linestyle', '--', 'color', 'r')
         axis([1e-2 1e1 1e-6 1e2])
         axis square
         xlabel('$\| \vec{y}_2 - \vec{y}_1 \|_2$', 'interpreter','latex')

### 2014-12-06 reducible_SDE/paper/SIADS_paper/make_schematic.m

@@ -33,16 +33,16 @@ saveas(gcf, 'schematic_DS1.eps', 'epsc');
 
 %%
 
-npoints = 2000;
+npoints = 4000;
 
 x = linspace(0, 1, npoints);
-y = randn(size(x));
+y = randn(size(x)) .* (1+8*exp(-4*x));
 
 make_fig(3, 3);
 scatter(x, smooth(y, 11), 50, -x,'.')
 set(gca, 'xtick', [])
 set(gca, 'ytick', [])
-set(gca, 'ylim', [-1 1])
+set(gca, 'ylim', [-8 8])
 % axis off
 saveas(gcf, 'schematic_DS2.eps', 'epsc');
 

### 2014-12-10 reducible_SDE/paper/SIADS_paper/make_schematic.m
updated paper

@@ -11,8 +11,8 @@ npoints = 1000;
 x = repmat(linspace(0, 1, npoints), ntraj, 1);
 
 y = zeros(size(x));
-y(1, :) = exp(-20*x(1,:));
-y(2, :) = -exp(-20*x(2,:));
+y(1, :) = exp(-10*x(1,:));
+y(2, :) = -exp(-10*x(2,:));
 
 % figure;
 % plot(x, y, '.')
@@ -35,7 +35,7 @@ saveas(gcf, 'schematic_DS1.eps', 'epsc');
 npoints = 4000;
 
 x = linspace(0, 1, npoints);
-y = randn(size(x)) .* max(10*exp(-7*x), 2);
+y = randn(size(x)) .* max(10*exp(-8*x), 2);
 
 make_fig(3, 3);
 scatter(x, smooth(y, 11), 50, -x,'.')

### 2014-12-11 chemotaxis/paper/draft1.tex
moved old + new text to chemotaxis/detecting harmonics draft

--- 2014-07-29 chemotaxis/prl_paper.tex
@@
-\documentclass[prl, reprint]{revtex4-1}
+\documentclass[preprint]{elsarticle}
 
+\usepackage{amssymb}
 \usepackage{amsmath}
 \usepackage{amsfonts}
-
+%\usepackage{cite}
+%\usepackage{graphics}
+%\usepackage{epsf}
+%\usepackage{psfig}
+%\usepackage{float}
 \usepackage{graphicx}
+\usepackage{epstopdf}
+%\usepackage[small]{caption2}
+%\usepackage{bm}%
+%\usepackage{multirow}
+%\usepackage{subfigure}
+%\usepackage{wrapfig}
+%\usepackage{color}
+%\usepackage{dsfont}
+%\usepackage{amsthm}
+%\usepackage{algorithm}
+%\usepackage{subfigure}
 \usepackage{subcaption}
 
-\usepackage{epstopdf}
+\graphicspath{{../}}
+
+%\newtheorem{theorem}{Theorem}
+%\newtheorem{corollary}[theorem]{Corollary}
+%\newtheorem{lemma}[theorem]{Lemma}
+%\newtheorem{observation}[theorem]{Observation}
+%\newtheorem{proposition}[theorem]{Proposition}
+%\newtheorem{definition}[theorem]{Definition}
+%\newtheorem{claim}[theorem]{Claim}
+%\newtheorem{fact}[theorem]{Fact}
+%\newtheorem{assumption}[theorem]{Assumption}
+%%\newtheorem{example}[theorem]{Example}
+% \newtheorem{remark}[theorem]{Remark}
+
+\journal{Computers \& Chemical Engineering}
 
@@
 
-\title{Characterizing System Dynamics using Diffusion Maps: A Chemotaxis Case Study}
+\begin{frontmatter}
 
-\author{Carmeline J. Dsilva}
-\email{cdsilva@princeton.edu}
-\affiliation{Department of Chemical and Biological Engineering, Princeton University, Princeton, NJ, 08544}
+\title{Parsimonious Representation of Nonlinear Dynamical Systems using Diffusion Maps: A Chemotaxis Case Study}
 
-\author{Ronen Talmon}
-\email{ronen.talmon@yale.edu}
-\affiliation{Department of Mathematics, Yale University, New Haven, CT, 06520}
-
-\author{Ronald R. Coifman}
-\email{coifman@math.yale.edu}
-\affiliation{Department of Mathematics, Yale University, New Haven, CT, 06520}
-
-\author{Ioannis G. Kevrekidis}
-\email{yannis@princeton.edu}
-\affiliation{Department of Chemical and Biological Engineering, Princeton University, Princeton, NJ, 08544}
-\affiliation{Program in Applied and Computational Mathematics, Princeton University, Princeton, NJ, 08544}
-
-\date{\today}
+\author[princeton]{Carmeline~J.~Dsilva\corref{cor1}}
+\ead{cdsilva@princeton.edu}
+%
+\author[technion]{Ronen~Talmon}
+\ead{ ronen@ee.technion.ac.il}
+%
+\author[yale]{Ronald R. Coifman}
+\ead{coifman@math.yale.edu}
+%
+\author[princeton, princetonpacm]{Ioannis~G.~Kevrekidis \corref{cor1}}
+\ead{yannis@princeton.edu}
+%
+\address[princeton]{Department of Chemical and Biological Engineering, Princeton University, Princeton, NJ, 08540, USA}
+\address[technion]{Technion - Israel Institute of Technology, Haifa, 3200003, Israel}
+\address[yale]{Department of Mathematics, Yale University, New Haven, CT, 06520, USA}
+\address[princetonpacm]{Program in Applied and Computational Mathematics, Princeton University, Princeton, NJ, 08540, USA}
+%
+\cortext[cor1]{Corresponding author}
 
@@
 
-Many dynamical systems are governed by stochastic motions at the microscale which give rise to smooth, coherent dynamics at the macroscale.
-%
-The relationship between the microscopic and macroscopic dynamics is often not immediately obvious and either requires some modeling assumptions or extensive observation of the system.
-%
-We demonstrate automatically extracting a parameterization of microscale data using manifold learning techniques, and show that this parameterization is consistent with the governing macroscopic dynamical model.
-%
-These data-driven algorithms need no {\em a priori} knowledge about the appropriate macroscopic model, but instead, only require the appropriate observers and pairwise metric between observations.
-%
-We also show how these algorithms can detect changes in the macroscopic dynamical behavior as a function of the microscopic system parameters. 
-%
-We illustrate these ideas using a model which arises in studies of cellular chemotaxis.
-
+TODO
 \end{abstract}
 
-\keywords{diffusion maps, earth mover's distance, histograms}
+%\sloppy
 
-\maketitle
+\begin{keyword}
+TODO
+\end{keyword}
 
-%\section{Introduction}
+\end{frontmatter}
+
+\section{Introduction}
+
+\begin{itemize}
+
+\item
+Data driven analysis methods, and in particular, manifold learning, can do many things (dimensionality reduction, reduction in dynamical systems, etc).
+
+\item
+Manifold learning constructs parameterization of the data through the spectral analysis of a Laplace operator.
+
+\item
+Initially it was used almost exclusively for toy examples/synthetic data (``these people suck").
+
+\item
+Only recently, has it been applied to real (experimental/simulation) data, and this was possible due to advances in observers (data representation) and metrics (``we are awesome").
+
+\item
+However, a major limitation (of spectral methods based on a Laplace operator) is identifying the true principal components.
+
+\item
+In this paper, we propose an algorithm to automatically identify those principal components using locally linear regression.
+
+\item
+All of these considerations (recently developed plus what we propose here) provide a data analysis pipeline.
+We show that this pipeline can successfully analyze data from a complex stochastic dynamical system originated from the study of cellular dynamics (chemotaxis).
+
+\end{itemize}
+
+\section{Issues in Manifold Learning/Diffusion maps}
+
+\subsection{Diffusion maps}
+
+\subsection{Observers}
+Add physical examples, and Amit's bi-spectrum, scattering....
+
+Cite PNAS, stating that the histograms are good since they linearize any noise.
+
+\subsection{Metric}
+Euclidean (not good, especially for histograms), Mahalanobis (good, cite PNAS, since it noise robust). Here we use EMD since it is multiscale.
+Stability and regularity issues.
+
+\section{Identify the principal components}
+
+\subsection{Algorithm: Local Linear Regression}
+
+Given the eigenvectors $\phi_1, \phi_2, \dots, \phi_{n-1} \in \mathbb{R}^n$, we would like to automatically deduce which ones are harmonics/functions of another. 
+%
+To accomplish this, we will fit a function $f(\phi_1, \dots, \phi_k)$ to $\phi_{k+1}$. 
+%
+If the resulting fit is good/accurate, then $\phi_{k+1}$ is most likely a harmonic of $\phi_1, \dots, \phi_k$.
+
+We will use a local linear function 
+\begin{equation}
+f_{k+1}(\phi_1, \dots, \phi_k) = a_0 + a_1 \phi_1 + \dots + a_k \phi_k
+\end{equation}
+where the coefficients $a_0, \dots, a_k$ are functions of $\phi_1, \dots, \phi_k$.
+
+To find these coefficients, we want to minimize the local weighted sum of squares
+\begin{equation} \label{eq:opt_problem}
+\sum_{i=1}^n w_i(x) (\phi_{k+1}(x) - f_{k+1} (\phi_1(x), \dots, \phi_{k}(x))
+\end{equation}
+where $w$ is a weighting function.
+%
+We will use
+%
+\begin{itemize}
+\item $w_i(x) = \exp \left( - \frac{\|\Phi_{k}(x) - \Phi_{k} (x_i) \|^2}{\epsilon^2} \right)$, where $\Phi_k(x) = \begin{bmatrix}
+\phi_1(x) \\
+\vdots \\
+\phi_k(x)
+\end{bmatrix} $
+\item $\epsilon = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k}(x)$
+\end{itemize}
+
+Let $\hat{f}_{k+1}(\phi_1(x_i), \dots, \phi_k(x_i))$ denote the minimum achieved in \eqref{eq:opt_problem}, when the minimization is performed over $x_j$ with $j \ne i$. 
+%
+We then define the leave-one-out- cross-validation error as
+\begin{equation}
+r_{k+1} = \sum_{i=1}^n \left( \phi_{k+1} (x_i) - \hat{f}_{k+1}(\phi_1(x_i), \dots, \phi_k(x_i)) \right)^2
+\end{equation}
+%
+A small value of $r_{k+1}$ means that $\phi_{k+1}$ can be accurately approximated from $\phi_1, \dots, \phi_k$, and is therefore a harmonic of a previous mode.
+%
+A large value of $r_{k+1}$ indicates that $\phi_{k+1}$ parameterizes a new direction in the data.
+
+\subsection{Eigenvalues to Quantify Dimensions}
+
+The eigenvalues of the Laplacian with Neumann boundary conditions on a $d$-dimensional rectangular domain are given by
+\begin{equation}
+\mu_{k,l} = \frac{k^2 \pi^2}{L_l^2}
+\end{equation}
+for $k=1, 2, \dots$ and $l=1, 2, \dots, d$, and $L_l$ is the length of the domain along the $l^{th}$ dimension. 
+%
+These eigenvalues are related to the eigenvalues we compute in diffusion maps by 
+\begin{equation}
+\lambda_{k,l} = e^{-\mu_{k,l} t} = e^{-\frac{k^2 \pi^2}{L_l^2} t}
+\end{equation}
+%
+Therefore, 
+\begin{equation}
+\sqrt{\frac{\log(\lambda_{1,l_1})}{\log(\lambda_{1,l_2})}} = 
+\frac{L_{l_2}}{L_{l_1}}
+\end{equation}
+and so looking at the ratio of the log of the eigenvalues (modulo harmonics) should inform us as to the lengths of the different directions in the manifold.
+
+\subsection{Illustrative Examples}
+
+\subsubsection{Swiss Roll}
+
+\includegraphics[height=3in]{swissroll1_evals}
+\includegraphics[height=3in]{swissroll2}\\
+\includegraphics[height=2in]{swissroll1_cv}
+\includegraphics[height=2in]{swissroll2_cv}\\
+\includegraphics[height=3in]{swissroll1_color1}
+\includegraphics[height=3in]{swissroll2_color1}\\
+\includegraphics[height=3in]{swissroll1_color2}
+\includegraphics[height=3in]{swissroll2_color2}
+
+For first case: $\lambda_1 = 0.9982$, $\lambda_2 = 0.9959$
+
+For second case: $\lambda_1 = 0.9977$, $\lambda_5 = 0.9652$.
+
+\subsubsection{Torus}
+
+\section{Chemotaxis: a case study}
 
@@
 
-\begin{figure*}[t!]
-\def\figwidth{6.4cm}
-\begin{subfigure}{\figwidth}
+\begin{figure}[t!]
+\begin{subfigure}{6cm}
 \includegraphics[width=\textwidth]{EMD_withhist_p_1}
@@
 \end{subfigure}
-\begin{subfigure}{\figwidth}
+\begin{subfigure}{6cm}
 \includegraphics[width=\textwidth]{EMD_withhist_t_1}
@@
 \end{subfigure}
-\begin{subfigure}{\figwidth}
+\begin{subfigure}{6cm}
 \includegraphics[width=\textwidth]{EMD_withhist_p_400}
@@
 \end{subfigure}
-\begin{subfigure}{\figwidth}
+\begin{subfigure}{6cm}
 \includegraphics[width=\textwidth]{EMD_withhist_t_400}
@@
 \label{fig:dmaps_embed_emd}
-\end{figure*}
+\end{figure}
 
@@
 
-\begin{acknowledgments}
+\subsection{Detect dimensionality}
 
-The authors would like to thank...
+\includegraphics[height=2in]{lle_errors1}
+\includegraphics[height=2in]{lle_errors2}
 
-\end{acknowledgments}
+\includegraphics[height=2in]{lle_evals1}
+\includegraphics[height=2in]{lle_evals2}
 
-\bibliography{../../references/references}
+\includegraphics[height=2in]{lle_embed1_1}
+\includegraphics[height=2in]{lle_embed2_2}
+
+\hfill 
+\includegraphics[height=2in]{lle_embed1_2}
+
+\section*{Acknowledgment}
+The authors would like to thank TODO.
+
+\bibliographystyle{elsarticle-num}
+\bibliography{papers}
 
 \end{document}
-%\end{document}

### 2014-12-12 chemotaxis/paper/draft1.tex
updated chemotaxis paper
updated chemotaxis paper and figures

@@
 \subsection{Observers}
-Add physical examples, and Amit's bi-spectrum, scattering....
 
-Cite PNAS, stating that the histograms are good since they linearize any noise.
+\begin{itemize}
+\item Physical examples (e.g., $\alpha$-carbons for molecules)
+\item Amit's bispectrum
+\item Scattering transform
+\item Cite PNAS, stating that the histograms are good since they linearize any noise. 
+\end{itemize}
 
 \subsection{Metric}
-Euclidean (not good, especially for histograms), Mahalanobis (good, cite PNAS, since it noise robust). Here we use EMD since it is multiscale.
-Stability and regularity issues.
 
-\section{Identify the principal components}
+\begin{itemize}
+\item Euclidean (not good, especially for histograms)
+\item Mahalanobis (good, cite PNAS, since it noise robust)
+\item Here we use EMD since it is multiscale.
+\item Emphasize stability and regularity issues.
+\end{itemize}
+
+\section{Identifying the Principal Components}
+
+\begin{itemize}
+\item Discuss the issues: harmonics
+\item Give simple example (Laplacian on strip)
+\end{itemize}
 
@@
 
-\includegraphics[height=3in]{swissroll1_evals}
-\includegraphics[height=3in]{swissroll2}\\
-\includegraphics[height=2in]{swissroll1_cv}
-\includegraphics[height=2in]{swissroll2_cv}\\
-\includegraphics[height=3in]{swissroll1_color1}
-\includegraphics[height=3in]{swissroll2_color1}\\
-\includegraphics[height=3in]{swissroll1_color2}
-\includegraphics[height=3in]{swissroll2_color2}
+\begin{figure}
+%
+\begin{subfigure}{0.45\textwidth}
+\centering
+\includegraphics[height=1.5in]{swissroll1}\\
+\includegraphics[height=1.5in]{swissroll1_evals}\\
+\includegraphics[height=1.5in]{swissroll1_color1}\\
+\includegraphics[height=1.5in]{swissroll1_color2}
+\caption{}
+\end{subfigure}
+\hfill
+\begin{subfigure}{0.45\textwidth}
+\centering
+\includegraphics[height=1.5in]{swissroll2} \\
+\includegraphics[height=1.5in]{swissroll2_evals} \\
+\includegraphics[height=1.5in]{swissroll2_color1}\\
+\includegraphics[height=1.5in]{swissroll2_color2}
+\caption{}
+\end{subfigure}
+%
+\caption{}
+\end{figure}
 
-For first case: $\lambda_1 = 0.9982$, $\lambda_2 = 0.9959$
+%For first case: $\lambda_1 = 0.9982$, $\lambda_2 = 0.9959$
 
-For second case: $\lambda_1 = 0.9977$, $\lambda_5 = 0.9652$.
+%For second case: $\lambda_1 = 0.9977$, $\lambda_5 = 0.9652$.
 
 \subsubsection{Torus}
+
+\begin{figure}[t]
+%
+\includegraphics[width=2.5in]{torus1}
+\hfill
+\includegraphics[height=1.5in]{torus1_evals}\\
+%
+\includegraphics[width=2.5in]{torus2}
+\hfill
+\includegraphics[height=1.5in]{torus2_evals}\\
+%
+\includegraphics[width=2.5in]{torus3}
+\hfill
+\includegraphics[height=1.5in]{torus3_evals}
+%
+\caption{}
+\end{figure}
 
@@
 
-\includegraphics[height=2in]{lle_errors1}
-\includegraphics[height=2in]{lle_errors2}
-
-\includegraphics[height=2in]{lle_evals1}
-\includegraphics[height=2in]{lle_evals2}
-
-\includegraphics[height=2in]{lle_embed1_1}
-\includegraphics[height=2in]{lle_embed2_2}
-
-\hfill 
-\includegraphics[height=2in]{lle_embed1_2}
+\begin{figure}
+%
+\def\figheight{1.4in}
+%
+\begin{subfigure}[t]{0.31\textwidth}
+\includegraphics[height=\figheight]{chemotaxis1_evals}
+\includegraphics[height=\figheight]{chemotaxis1_embed_good}
+%\includegraphics[height=1.5in]{chemotaxis1_embed_bad}
+\end{subfigure}
+%
+\begin{subfigure}[t]{0.31\textwidth}
+\includegraphics[height=\figheight]{chemotaxis2_evals}
+\includegraphics[height=\figheight]{chemotaxis2_embed_good}
+\includegraphics[height=\figheight]{chemotaxis2_embed_bad}
+\end{subfigure}
+%
+\begin{subfigure}[t]{0.35\textwidth}
+\includegraphics[height=\figheight]{chemotaxis3_evals}
+\includegraphics[height=\figheight]{chemotaxis3_embed_good}
+\includegraphics[height=\figheight]{chemotaxis3_embed_bad}
+\end{subfigure}
+%
+\caption{}
+%
+\end{figure}
 

### 2014-12-15 chemotaxis/paper/draft1.tex
updated chemotaxis paper

@@
 
+Let $z(t) \in \mathbb{R}^n$ denote an $n$-dimensional observation of the system at time $t$. 
+%
+Assume we are given $m$ observations $z(1), \dots, z(m)$, which lie on a $d$-dimensional, nonlinear manifold $\mathcal{M}_d$, with $d \ll n$. 
+%
+Our goal is to uncover a $d$-dimensional parameterization of the $m$ observations that respects the underlying manifold geometry.
+%
+We first construct the matrix $W \in \mathbb{R}^{m \times m}$, with
+\begin{equation} \label{eq:W}
+W_{ij} = \exp \left( -\frac{\|z(i) - z(j) \|^2}{\epsilon^2} \right), \ i,j=1,\ldots,m
+\end{equation}
+where $\| \cdot \|$ denotes the appropriate norm for the observations, and $\epsilon$ is a characteristic distance between the observations. 
+%
+We typically take $\epsilon$ to be the median of the pairwise distances between the data points.
+%
+We then construct the diagonal matrix $D \in \mathbb{R}^{m \times m}$, with $D_{ii} = \sum_j W_{ij}$, and the matrix $A  = D^{-1} W.$
+%
+The eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ provide the embedding coordinates for the data, such that
+$\phi_{j}(i)$ is the $j^{th}$ embedding coordinate for $z(i)$.
+%
+The ``importance'' of coordinate $\phi_j$ is quantified by the corresponding eigenvalue $\mu_j$, e.g., the relative importance of $\phi_1$ versus $\phi_2$ can be measured by comparing $\mu_1$ and $\mu_2$.
+%
+We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$, and if the data are low-dimensional, we only need to retain a few of the leading eigenvectors to adequately describe the data.
+%
+Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a constant vector which is not used as an embedding coordinate.
+
+In the limit of infinite data ($m \rightarrow \infty$ and $\epsilon \rightarrow 0$ with the appropriate rate), the matrix $I-A$ approaches the Laplacian operator on the manifold $\mathcal{M}_d$ \cite{...}.
+%
+Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ approach the eigenfunctions of the Laplacian (with Neumann boundary conditions) on $\mathcal{M}_d$.
+%
+The eigenvalues are related to the eigenvalues we compute in diffusion maps by 
+\begin{equation}
+\mu_k = e^{-\tilde{\mu}_k t}.
+\end{equation}
+
 \subsection{Observers}
@@
 \item Cite PNAS, stating that the histograms are good since they linearize any noise. 
+For the chemotaxis example, we use histograms of the particle positions as observers \cite{talmon2013empirical}. 
+%
+Histograms are invariant to the indexing of the particles, while retaining the spatial locations of the particles.
 \end{itemize}
@@
 \item Here we use EMD since it is multiscale.
+%
+Instead of the standard Euclidean distance, we use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between pairs of histograms. 
+%
+Conceptually, the EMD measures how much ``work'' it takes to transform one probability density into another.
+%
+It therefore not only considers where the densities are inconsistent, but also how far apart the inconsistencies are.
+%
+Although the brute-force computation of the EMD is computationally expensive, there has been a plethora of work in developing efficient algorithms for its computation \cite{Pele-eccv2008, Pele-iccv2009}.
+%
+For the special case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions \cite{rubner2000perceptual}, which can be estimated from histograms as
+\begin{equation}
+\| z(i) - z(j) \|_{EMD} \approx \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|
+\end{equation}
+where the histograms $z(i), z(j)$ are defined on equally-spaced bins in $\mathbb{R}$. 
+%
 \item Emphasize stability and regularity issues.
@@
 \end{itemize}
+
+In diffusion maps, we order the eigenvectors by the magnitude of the corresponding eigenvalues, in the hopes that the top eigenvectors provide us with a parameterization of the underlying manifold.
+%
+However, often some eigenvectors are higher harmonics of previous eigenvectors and do not describe new directions in the data set. 
+%
+We will illustrate this issue with a simple two-dimensional strip where we know the analytical form of the eigenfunctions and can explicitly highlight the issue at hand.
+
+Consider the case where $\mathcal{M}_d$ is a $2$-dimensional rectangular domain with edge lengths $L_x$ and $L_y$. 
+%
+The eigenvalues of the Laplacian with Neumann boundary conditions are given by
+\begin{equation}
+\tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_x} \right)^2 + \left( \frac{k_2 \pi}{L_y} \right)^2
+\end{equation}
+for $k_1, k_2 = 0, 1, 2, \dots$.
+%
+The corresponding eigenfunctions are 
+\begin{equation}
+\tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi x}{L_x} \right) \cos \left( \frac{k_2 \pi y}{L_y} \right)
+\end{equation}
+where $x$ and $y$ denote the two coordinates of the rectangular domain. 
+%
+We note that $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi x}{L_x} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi y}{L_y} \right)$ are one-to-one with the $x$ and $y$ coordinates, respectively, and so the eigenfunctions of the Laplacian yield a parameterization of the underlying manifold.
+%
+However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to correspond to the two largest (non-trivial) eigenvalues (the largest eigenvalue will always be $\mu_{0,0} = 1$).
+%
+In fact, if $L_x > 2 L_y$, then $\mu_{2, 0} > \mu_{0, 1}$, and so the second (non-trivial) eigenvector (when ordered by eigenvalue) will be a harmonic of the first and still parameterize the $x$ direction.
+%
+We therefore require a methodology to automatically detect which eigenvectors are harmonics of each other.
+
+%From the eigenvalues, 
+%\begin{equation}
+%\sqrt{\frac{\log(\mu_{1,0})}{\log(\mu_{0,1})}} = 
+%\frac{L_y}{L_x}
+%\end{equation}
+%and so looking at the ratio of the log of the eigenvalues (modulo harmonics) should inform us as to the lengths of the different directions in the manifold.
 
@@
 %
-\begin{itemize}
-\item $w_i(x) = \exp \left( - \frac{\|\Phi_{k}(x) - \Phi_{k} (x_i) \|^2}{\epsilon^2} \right)$, where $\Phi_k(x) = \begin{bmatrix}
+\begin{equation}
+w_i(x) = \exp \left( - \frac{\|\Phi_{k}(x) - \Phi_{k} (x_i) \|^2}{\epsilon^2} \right), \text{where } \Phi_k(x) = \begin{bmatrix}
 \phi_1(x) \\
@@
 \phi_k(x)
-\end{bmatrix} $
-\item $\epsilon = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k}(x)$
-\end{itemize}
+\end{bmatrix} .
+\end{equation}
+%
+We take $\epsilon = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k}(x)$.
 
@@
 We then define the leave-one-out- cross-validation error as
-\begin{equation}
+\begin{equation} \label{eq:cv_error}
 r_{k+1} = \sum_{i=1}^n \left( \phi_{k+1} (x_i) - \hat{f}_{k+1}(\phi_1(x_i), \dots, \phi_k(x_i)) \right)^2
@@
 
-\subsection{Eigenvalues to Quantify Dimensions}
-
-The eigenvalues of the Laplacian with Neumann boundary conditions on a $d$-dimensional rectangular domain are given by
-\begin{equation}
-\mu_{k,l} = \frac{k^2 \pi^2}{L_l^2}
-\end{equation}
-for $k=1, 2, \dots$ and $l=1, 2, \dots, d$, and $L_l$ is the length of the domain along the $l^{th}$ dimension. 
-%
-These eigenvalues are related to the eigenvalues we compute in diffusion maps by 
-\begin{equation}
-\lambda_{k,l} = e^{-\mu_{k,l} t} = e^{-\frac{k^2 \pi^2}{L_l^2} t}
-\end{equation}
-%
-Therefore, 
-\begin{equation}
-\sqrt{\frac{\log(\lambda_{1,l_1})}{\log(\lambda_{1,l_2})}} = 
-\frac{L_{l_2}}{L_{l_1}}
-\end{equation}
-and so looking at the ratio of the log of the eigenvalues (modulo harmonics) should inform us as to the lengths of the different directions in the manifold.
-
 \subsection{Illustrative Examples}
@@
 \subsubsection{Swiss Roll}
+
+Our first illustrative example will consist of two different swiss roll data sets.
+%
+Each data set consists of $m=1500$ points uniformly distributed on the swiss roll. 
+%
+Both swiss rolls are the same in the angular/spiral direction.
+%
+However, height of the first swiss roll is $h = 15 \pi$, while the heigh of the second is $h = 6 \pi$. 
+%
+These two data sets are shown in Figure~\ref{fig:swiss_rolls}(a)~and~(b), respectively.
+%
+We expect to recover the eigenvector which parameterizes the height farther down in the spectrum for the second data set relative to the first, as this second data set is shorter.
+
+Figure~\ref{fig:swiss_rolls}~(c)~and~(d) shows the diffusion maps eigenvalue spectra for the two data sets.
+%
+The bars are colored by the leave-one-out cross-validation error as defined in \eqref{eq:cv_error}; a small value of $r_k$ indicate that the corresponding eigenvector is a harmonic of previous eigenvectors, while a large value of $r_k$ indicates that the corresponding eigenvector describes a new direction in the data.
+%
+From these plots, one can conclude that the first two eigenvectors $\phi_1$ and $\phi_2$ parameterize the first data set, while $\phi_1$ and $\phi_5$ parameterize the second. 
+%
+Figure~\ref{fig:swiss_rolls}~(e--h) shows the two data sets, colored by the two ``important'' eigenvectors. 
+%
+As expected, these eigenvectors parameterize the arc length along the spiral, and the height of the swiss roll.
 
@@
 \centering
-\includegraphics[height=1.5in]{swissroll1}\\
-\includegraphics[height=1.5in]{swissroll1_evals}\\
-\includegraphics[height=1.5in]{swissroll1_color1}\\
+\includegraphics[height=1.5in]{swissroll1}
+\caption{}
+\end{subfigure}
+\hfill
+\begin{subfigure}{0.45\textwidth}
+\centering
+\includegraphics[height=1.5in]{swissroll2}
+\caption{}
+\end{subfigure}
+\hfill
+\begin{subfigure}{0.45\textwidth}
+\centering
+\includegraphics[height=1.5in]{swissroll1_evals}
+\caption{}
+\end{subfigure}
+\hfill
+\begin{subfigure}{0.45\textwidth}
+\centering
+\includegraphics[height=1.5in]{swissroll2_evals}
+\caption{}
+\end{subfigure}
+\hfill
+\begin{subfigure}{0.45\textwidth}
+\centering
+\includegraphics[height=1.5in]{swissroll1_color1} 
+\caption{}
+\end{subfigure}
+\hfill
+\begin{subfigure}{0.45\textwidth}
+\centering
+\includegraphics[height=1.5in]{swissroll2_color1}
+\caption{}
+\end{subfigure}
+\hfill
+\begin{subfigure}{0.45\textwidth}
+\centering
 \includegraphics[height=1.5in]{swissroll1_color2}
@@
 \centering
-\includegraphics[height=1.5in]{swissroll2} \\
-\includegraphics[height=1.5in]{swissroll2_evals} \\
-\includegraphics[height=1.5in]{swissroll2_color1}\\
 \includegraphics[height=1.5in]{swissroll2_color2}
@@
 %
-\caption{}
+\caption{Swiss roll example. (a) Data set 1, with $h=6 \pi$. (b) Data set 2, with $h = 15 \pi$. (c) Diffusion maps eigenvalue spectrum for data set 1. } 
+\label{fig:swiss_rolls}	
 \end{figure}
@@
 \bibliographystyle{elsarticle-num}
-\bibliography{papers}
+\bibliography{../../../references/references}
 

### 2014-12-15 chemotaxis/swissroll.m

@@ -100,6 +100,51 @@ print('swissroll1_evals.eps','-depsc')
 
 %%
 
+figure;
+
+D1 = pdist(V(:,2:3));
+D2 = pdist(V(:,2:4));
+
+subplot(2,2,1)
+plot(V(:,2),V(:,3),'.')
+xlabel('\phi_2')
+ylabel('\phi_3')
+
+eps = logspace(-8, -3, 100);
+sumA1 = zeros(size(eps));
+sumA2 = zeros(size(eps));
+
+for i=1:length(eps)
+    sumA1(i) = sum(exp(-D1.^2/(2*eps(i))));
+    sumA2(i) = sum(exp(-D2.^2/(2*eps(i))));
+end
+
+subplot(2,2,2)
+loglog(eps, sumA1, '.')
+hold on
+loglog(eps, 1e10*eps)
+p = polyfit(log(eps), log(sumA1), 1);
+legend(sprintf('data; est dim = %2.2f', p(1)*2), 'dim=2 line','location','best')
+xlabel('\epsilon')
+ylabel('$\sum W_{ij}$', 'interpreter','latex')
+
+subplot(2,2,3)
+plot3(V(:,2),V(:,3),V(:,4),'.')
+xlabel('\phi_2')
+ylabel('\phi_3')
+zlabel('\phi_4')
+
+subplot(2,2,4)
+loglog(eps, sumA2, '.r')
+hold on
+loglog(eps, 1e10*eps, 'r')
+p = polyfit(log(eps), log(sumA2), 1);
+legend(sprintf('data; est dim = %2.2f', p(1)*2), 'dim=2 line','location','best')
+xlabel('\epsilon')
+ylabel('$\sum W_{ij}$', 'interpreter','latex')
+
+%%
+
 load swissroll2.mat
 
 make_fig(2,2);
@@ -153,27 +198,47 @@ print('swissroll2_evals.eps','-depsc')
 
 %%
 
-% D1 = pdist(V(:,2:3));
-% D2 = pdist(V(:,[2 6]));
-% 
-% 
-% eps = logspace(-8, -4, 100);
-% sumA1 = zeros(size(eps));
-% sumA2 = zeros(size(eps));
-% 
-% for i=1:length(eps)
-%     sumA1(i) = sum(exp(-D1.^2/(2*eps(i))));
-%     sumA2(i) = sum(exp(-D2.^2/(2*eps(i))));
-% end
-% 
[…]

### 2015-01-02 chemotaxis/paper/draft1.tex
edited paper

@@
 \graphicspath{{../}}
+
+\DeclareMathOperator*{\argmin}{\arg\!\min}
 
@@
 
+Often, systems which are stochastic and noisy on the microscale exhibit coherent dynamics on the macroscale which are governed by only a few parameters.
+%
+For example, fluid flow, which can be described via the position and velocity of every fluid particle, is modeled at the macroscale using the Navier-Stokes equations.
+%
+However, in general, this mapping from microscale to macroscale is not always obvious.
+%
+We would like to extract a parameterization of data collected at the microscale which is consistent with the macroscopic behavior, without any {\em a priori} knowledge of the appropriate microscopic or macroscopic model.
+
+In this letter, we will demonstrate how this goal can be achieved through a data-driven method based on geometric analysis and manifold learning. 
+%
+Specifically, we aim to emphasize two particular aspects of manifold learning from a dynamical systems' point of view: (a) choosing the appropriate observers, especially in presence of stochastic dynamics, and (b) selecting an appropriate metric for comparison \cite{mallat2012group}.
+%
+We will show that both are essential for data-driven algorithms to successfully uncover the underlying system dynamics.
+
+We will illustrate these points using a model problem which arises in studies of cellular chemotaxis \cite{othmer2000diffusion}.
+%
+In this example, the macroscopic dynamics vary depending on the value of a single system parameter.
+%
+We will demonstrate how we can use diffusion maps \cite{coifman2005geometric}, a manifold learning technique, to {\em automatically} detect changes in the system's macroscopic dynamics from microscale data.
+%
+This case study exhibits several key points which allow us to demonstrate the strength of our approach.
+%
+First, this system has random microscopic behavior with smooth macroscopic dynamics governed by a single parameter which determines the regime/mode of the system. 
+%
+Second, it consists of an ensemble of individual stochastic particles which must be described using the appropriate observers.
+%
+Third, it has an analytic macroscopic description, which serves as a ground truth and will allow us to verify our results which are obtained in an unsupervised manner.
+%
+We will show that manifold learning algorithms based on statistical observers and with the appropriate affinity metric between pairs of observations uncovers a parameterization of the microscopic data which is consistent with the macroscopic model.
+%
+Furthermore, these data-driven methods allow us to detect changes in dynamical behavior resulting from changes in the system parameters.
+
 \section{Issues in Manifold Learning/Diffusion maps}
@@
 %
-The eigenvalues are related to the eigenvalues we compute in diffusion maps by 
+The eigenvalues of the Laplacian operator $\tilde{\mu}_0, \tilde{\mu}_1, \dots, \tilde{\mu}_{m-1}$ are related to the eigenvalues we compute in diffusion maps by 
 \begin{equation}
@@
 \end{equation}
-where the histograms $z(i), z(j)$ are defined on equally-spaced bins in $\mathbb{R}$. 
+where the histograms $z(i), z(j)$ are defined on equally-spaced bins in $\mathbb{R}$, and $z_k$ denotes the $k^{th}$ bin. 
 %
@@
 
-Consider the case where $\mathcal{M}_d$ is a $2$-dimensional rectangular domain with edge lengths $L_x$ and $L_y$. 
+Consider the case where $\mathcal{M}_d$ is a $2$-dimensional rectangular domain with edge lengths $L_x  > L_y$. 
 %
@@
 %
-We note that $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi x}{L_x} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi y}{L_y} \right)$ are one-to-one with the $x$ and $y$ coordinates, respectively, and so the eigenfunctions of the Laplacian yield a parameterization of the underlying manifold.
+We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi x}{L_x} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi y}{L_y} \right)$ are one-to-one with the $x$ and $y$ coordinates, respectively, and therefore yield a parameterization of the underlying manifold.
 %
-However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to correspond to the two largest (non-trivial) eigenvalues (the largest eigenvalue will always be $\mu_{0,0} = 1$).
+However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to correspond to the two largest (non-trivial) eigenvalues (the largest eigenvalue will always be $\mu_{0,0} = 1$ and correspond to a constant eigenfunction).
 %
-In fact, if $L_x > 2 L_y$, then $\mu_{2, 0} > \mu_{0, 1}$, and so the second (non-trivial) eigenvector (when ordered by eigenvalue) will be a harmonic of the first and still parameterize the $x$ direction.
+In fact, if $L_x > 2 L_y$, then $\mu_{2, 0} > \mu_{0, 1}$, and so the second (non-trivial) eigenvector (when the eigenvectors are sorted by their corresponding eigenvalues) will be a harmonic of the first and still parameterize the $x$ direction.
 %
@@
 %
-To accomplish this, we will fit a function $f(\phi_1, \dots, \phi_k)$ to $\phi_{k+1}$. 
+To accomplish this, we will fit a function $f(\phi_1, \dots, \phi_{k-1})$ to $\phi_{k}$. 
 %
-If the resulting fit is good/accurate, then $\phi_{k+1}$ is most likely a harmonic of $\phi_1, \dots, \phi_k$.
+If the resulting fit is good/accurate, then $\phi_{k}$ is most likely a harmonic of $\phi_1, \dots, \phi_{k-1}$.
 
@@
 \begin{equation}
-f_{k+1}(\phi_1, \dots, \phi_k) = a_0 + a_1 \phi_1 + \dots + a_k \phi_k
+\phi_k \approx \alpha( \Phi_{k-1}) + \beta(\Phi_{k-1})^T \Phi_{k-1}
 \end{equation}
-where the coefficients $a_0, \dots, a_k$ are functions of $\phi_1, \dots, \phi_k$.
+where 
+%
+$\Phi_{k-1} = \begin{bmatrix} \phi_1 \\ \vdots \\ \phi_{k-1} \end{bmatrix}$
+and $\beta \in \mathbb{R}^{k-1}$. 
+%
+The coefficients $\alpha$ and $\beta$ are {\em functions} of $\phi_{k-1}$ because we use a {\em local} linear fit; therefore, the coefficients change as a function of the domain.
 
-To find these coefficients, we want to minimize the local weighted sum of squares
+At each point $\Phi_{k-1}(i)$, we fit a local linear function using the remaining $m-1$ data points. 
+%
+We solve the following optimization problem 
 \begin{equation} \label{eq:opt_problem}
-\sum_{i=1}^n w_i(x) (\phi_{k+1}(x) - f_{k+1} (\phi_1(x), \dots, \phi_{k}(x))
+\hat{\alpha} (i) , \hat{\beta}(i)  = \argmin_{\alpha, \beta} \sum_{j \ne i} w_i(\Phi_{k-1}(i), \Phi_{k-1}(j)) \left( \phi_{k}(j) - (\alpha + \beta^T \Phi_{k-1}(j)) \right)^2.
 \end{equation}
+%
 where $w$ is a weighting function.
 %
-We will use
+We will use Gaussian weights, 
 %
 \begin{equation}
-w_i(x) = \exp \left( - \frac{\|\Phi_{k}(x) - \Phi_{k} (x_i) \|^2}{\epsilon^2} \right), \text{where } \Phi_k(x) = \begin{bmatrix}
-\phi_1(x) \\
-\vdots \\
-\phi_k(x)
-\end{bmatrix} .
+w_i(\Phi_{k-1}(i), \Phi_{k-1}(j))  = \exp \left( - \frac{\|\Phi_{k-1}(i) - \Phi_{k-1} (j) \|^2}{\epsilon^2} \right).
 \end{equation}
 %
-We take $\epsilon = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k}(x)$.
+We take $\epsilon = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k-1}(i)$.
 
-Let $\hat{f}_{k+1}(\phi_1(x_i), \dots, \phi_k(x_i))$ denote the minimum achieved in \eqref{eq:opt_problem}, when the minimization is performed over $x_j$ with $j \ne i$. 
-%
-We then define the leave-one-out- cross-validation error as
+We then define the normalized leave-one-out cross-validation error as
 \begin{equation} \label{eq:cv_error}
-r_{k+1} = \sum_{i=1}^n \left( \phi_{k+1} (x_i) - \hat{f}_{k+1}(\phi_1(x_i), \dots, \phi_k(x_i)) \right)^2
+r_{k} = \sqrt{ \frac{\sum_{i=1}^n \left( \phi_{k} (i) - (\hat{\alpha}(i) + \hat{\beta}(i)^T \Phi_{k-1}(i))  \right)^2} {\sum_{i=1}^n  \left( \phi_{k} (i) \right)^2 }}
 \end{equation}
 %
-A small value of $r_{k+1}$ means that $\phi_{k+1}$ can be accurately approximated from $\phi_1, \dots, \phi_k$, and is therefore a harmonic of a previous mode.
+%We will use a local linear function 
+%\begin{equation}
+%f_{k}(\phi_1, \dots, \phi_{k-1}) = a_0 + a_1 \phi_1 + \dots + a_{k-1} \phi_{k-1}
+%\end{equation}
+%where the coefficients $a_0, \dots, a_{k-1}$ are functions of $\phi_1, \dots, \phi_{k-1}$. 
 %
-A large value of $r_{k+1}$ indicates that $\phi_{k+1}$ parameterizes a new direction in the data.
+%To find these coefficients, we want to minimize the local weighted sum of squares
+%\begin{equation} \label{eq:opt_problem}
+%\sum_{i=1}^n w_i(x) \left( \phi_{k}(x) - f_{k} (\phi_1(x), \dots, \phi_{k-1}(x)) \right)^2
+%\end{equation}
+%where $w$ is a weighting function.
+%%
+%We will use Gaussian weights, 
+%%
+%\begin{equation}
+%w_i(x) = \exp \left( - \frac{\|\Phi_{k-1}(x) - \Phi_{k-1} (x_i) \|^2}{\epsilon^2} \right), \text{where } \Phi_{k-1}(x) = \begin{bmatrix}
+%\phi_1(x) \\
+%\vdots \\
+%\phi_{k-1}(x)
+%\end{bmatrix} .
+%\end{equation}
+%%
+%We take $\epsilon = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k-1}(x)$.
+%
+%Let $\hat{f}_{k}(\phi_1(x_i), \dots, \phi_{k-1}(x_i))$ denote the minimum achieved in \eqref{eq:opt_problem}, when the minimization is performed over $x_j$ with $j \ne i$. 
+%
+%We then define the normalized leave-one-out cross-validation error as
+%\begin{equation} \label{eq:cv_error}
+%r_{k} = \sum_{i=1}^n \left( \phi_{k} (x_i) - \hat{f}_{k}(\phi_1(x_i), \dots, \phi_{k-1}(x_i)) \right)^2
+%\end{equation}
+%\begin{equation} \label{eq:cv_error}
+%r_{k} = \sqrt{ \frac{\sum_{i=1}^n \left( \phi_{k} (x_i) - \hat{f}_{k}(\Phi_{k-1}(x_i)) \right)^2} {\sum_{i=1}^n  \left( \phi_{k} (x_i) \right) }}
+%\end{equation}
+%
+When $r_{k}$ is small, $\phi_{k}$ can be accurately approximated from $\phi_1, \dots, \phi_{k-1}$, and is therefore a harmonic of a previous mode.
+%
+A large value of $r_{k}$ indicates that $\phi_{k}$ parameterizes a new direction in the data.
+%
+Because $r_{k} \approx 1$ if $\phi_{k}$ parameterizes a new direction, we define $r_1 = 1$.
 
@@
 %
-\caption{Swiss roll example. (a) Data set 1, with $h=6 \pi$. (b) Data set 2, with $h = 15 \pi$. (c) Diffusion maps eigenvalue spectrum for data set 1. } 
+\caption{Swiss roll example.
+(a) Data set 1; $h=6 \pi$.
+(b) Data set 2; $h = 15 \pi$.
+(c) Diffusion maps eigenvalue spectrum for data set 1.
+(d) Diffusion maps eigenvalue spectrum for data set 2.
+(e) Data set 1, colored by the first diffusion maps eigenvector.
+(f) Data set 2, colored by the first diffusion maps eigenvector.
+(g) Data set 1, colored by the second diffusion maps eigenvector which is not a harmonic of the first.
+(h) Data set 2, colored by the second diffusion maps eigenvector which is not a harmonic of the first.
+}
 \label{fig:swiss_rolls}	
@@
 
-%For first case: $\lambda_1 = 0.9982$, $\lambda_2 = 0.9959$
+\subsubsection{Torus}
 
-%For second case: $\lambda_1 = 0.9977$, $\lambda_5 = 0.9652$.
+For the second example, we consider a torus defined by
+%
+\begin{equation}
+\begin{aligned}
+x =& (r_1 + r_2 \cos \theta_2 ) \cos \theta_1 \\
+y =& (r_1 + r_2 \cos \theta_2 ) \sin \theta_1 \\
+z =& r_2 \sin \theta_2
+\end{aligned}
+\end{equation}
+%
+where $r_1 > r_2$ are the outer and inner radii, respectively, and $\theta_1, \theta_2 \sim Unif(0, 2 \pi)$. 
+%
+Changing the ratio of $r_1$ to $r_2$ changes the ratio of the eigenvalues corresponding to non-harmonic eigenvectors.
 
-\subsubsection{Torus}
+We expect to obtain two eigenfunctions ($\sin \theta_1$ and $\cos \theta_1$) which parmeterize the outer circle, and two eigenfunctions ($\sin \theta_2$ and $\cos \theta_2$) which parameterize the inner circle.
+%
+However, the number of harmonics which fall between these relevant/descriptive eigenfunctions depends on the ratio of the two radii.
+
+Figure~\ref{fig:torus} shows the diffusion maps eigenspectra for three different torii. 
+%
+Clearly, as the torus becomes thinner, the second pair of relevant eigenfunctions (corresponding to $\sin \theta_2$ and $\cos \theta_2$) moves farther down in the spectrum.
 
@@
 %
+\begin{subfigure}{\textwidth}
 \includegraphics[width=2.5in]{torus1}
-\hfill
-\includegraphics[height=1.5in]{torus1_evals}\\
+\includegraphics[height=1.5in]{torus1_evals}
+\caption{}
+\end{subfigure}
 %
+\begin{subfigure}{\textwidth}
 \includegraphics[width=2.5in]{torus2}
-\hfill
 \includegraphics[height=1.5in]{torus2_evals}\\
+\caption{}
+\end{subfigure}
 %
+\begin{subfigure}{\textwidth}
 \includegraphics[width=2.5in]{torus3}
-\hfill
 \includegraphics[height=1.5in]{torus3_evals}
+\caption{}
+\end{subfigure}
 %
-\caption{}
+\caption{Torus data sets (left) and corresponding diffusion maps eigenvalues (right).
+(a) $r_1 = 3$, $r_2 = 1$.
+(b) $r_1 = 5$, $r_2 = 1$.
+(c) $r_1 = 10$, $r_2 = 1$.
+In all three data sets, the first two eigenvalues/eigenvectors correspond to $\sin \theta_1$ and $\cos \theta_1$.
+The second pair of non-harmonic eigenvectors (corresponding to $\sin \theta_2$ and $\cos \theta_2$) are captured by components 7 and 8, 11 and 12, and 15 and 16, respectively.}
+%
+\label{fig:torus}
+%
 \end{figure}
@@
 \section{Chemotaxis: a case study}
-
-Often, systems which are stochastic and noisy on the microscale exhibit coherent dynamics on the macroscale which are governed by only a few parameters.
-%
-For example, fluid flow, which can be described via the position and velocity of every fluid particle, is modeled at the macroscale using the Navier-Stokes equations.
-%
-However, in general, this mapping from microscale to macroscale is not always obvious.
-%
-We would like to extract a parameterization of data collected at the microscale which is consistent with the macroscopic behavior, without any {\em a priori} knowledge of the appropriate microscopic or macroscopic model.
-
-In this letter, we will demonstrate how this goal can be achieved through a data-driven method based on geometric analysis and manifold learning. 
-%
-Specifically, we aim to emphasize two particular aspects of manifold learning from a dynamical systems' point of view: (a) choosing the appropriate observers, especially in presence of stochastic dynamics, and (b) selecting an appropriate metric for comparison \cite{mallat2012group}.
-%
-We will show that both are essential for data-driven algorithms to successfully uncover the underlying system dynamics.
-
-We will illustrate these points using a model problem which arises in studies of cellular chemotaxis \cite{othmer2000diffusion}.
-%
-In this example, the macroscopic dynamics vary depending on the value of a single system parameter.
-%
-We will demonstrate how we can use diffusion maps \cite{coifman2005geometric}, a manifold learning technique, to {\em automatically} detect changes in the system's macroscopic dynamics from microscale data.
-%
-This case study exhibits several key points which allow us to demonstrate the strength of our approach.
-%
-First, this system has random microscopic behavior with smooth macroscopic dynamics governed by a single parameter which determines the regime/mode of the system. 
-%
-Second, it consists of an ensemble of individual stochastic particles which must be described using the appropriate observers.
-%
-Third, it has an analytic macroscopic description, which serves as a ground truth and will allow us to verify our results which are obtained in an unsupervised manner.
-%
-We will show that manifold learning algorithms based on statistical observers and with the appropriate affinity metric between pairs of observations uncovers a parameterization of the microscopic data which is consistent with the macroscopic model.
-%
-Furthermore, these data-driven methods allow us to detect changes in dynamical behavior resulting from changes in the system parameters.
 
@@
 Two is a {\em distance metric} between the observations that captures a notion of locality: observations which we perceive to be similar should have a small distance. 
-
-Let $z(t) \in \mathbb{R}^n$ denote an $n$-dimensional observation of the system at time $t$. 
 %
-For the chemotaxis example, we use histograms of the particle positions as observers \cite{talmon2013empirical}. 
-%
-Histograms are invariant to the indexing of the particles, while retaining the spatial locations of the particles.
-%
-Instead of the standard Euclidean distance, we use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between pairs of histograms. 
-%
-Conceptually, the EMD measures how much ``work'' it takes to transform one probability density into another.
-%
-It therefore not only considers where the densities are inconsistent, but also how far apart the inconsistencies are.
-%
-Although the brute-force computation of the EMD is computationally expensive, there has been a plethora of work in developing efficient algorithms for its computation \cite{Pele-eccv2008, Pele-iccv2009}.
-%
-For the special case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions \cite{rubner2000perceptual}, which can be estimated from histograms as
-\begin{equation}
-\| z(i) - z(j) \|_{EMD} \approx \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|
-\end{equation}
-where the histograms $z(i), z(j)$ are defined on equally-spaced bins in $\mathbb{R}$.
-
-Assume we are given $m$ observations $z(1), \dots, z(m)$, which lie on a $d$-dimensional, nonlinear manifold, with $d \ll n$. 
-%
-Our goal is to uncover a $d$-dimensional parameterization of the $m$ observations that respects the underlying manifold geometry.
-%
-We first construct the matrix $W \in \mathbb{R}^{m \times m}$, with
-\begin{equation} \label{eq:W}
-W_{ij} = \exp \left( -\frac{\|z(i) - z(j) \|^2}{\epsilon^2} \right), \ i,j=1,\ldots,m
-\end{equation}
-where $\| \cdot \|$ denotes the appropriate norm for the observations, and $\epsilon$ is a characteristic distance between the observations. 
-%
-For the chemotaxis example, we use the EMD as the norm, and we take $\epsilon$ to be the median of the pairwise distances between the data points.
-%
-We then construct the diagonal matrix $D \in \mathbb{R}^{m \times m}$, with $D_{ii} = \sum_j W_{ij}$, and the matrix $A  = D^{-1} W.$
-%
-The eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ provide the embedding coordinates for the data, such that
-$\phi_{j}(i)$ is the $j^{th}$ embedding coordinate for $z(i)$.
-%
-The ``importance'' of coordinate $\phi_j$ is quantified by the corresponding eigenvalue $\mu_j$, e.g., the relative importance of $\phi_1$ versus $\phi_2$ can be measured by comparing $\mu_1$ and $\mu_2$.
-%
-We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$, and if the data are low-dimensional, we only need to retain a few of the leading eigenvectors to adequately describe the data.
-%
-Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a constant vector which is not used as an embedding coordinate.
+For this particular example, we will use histograms of the particle positions as the observations, and the EMD as the distance metric.
 
@@
 
+\subsection{Detect dimensionality}
+
+Often in dynamical systems, one would like to detect a change in the dimensionally of the system. 
+%
+We can uncover the true dimensionally of the data set by looking at the eigenvalues corresponding to eigenvectors which parameterize new directions in the data. 
+%
+For the specific chemotaxis example, we know that there are two directions in the data, corresponding to simulation parameters $p$ and $t$. 
+%
+However, at long times, we know the dynamics become diffusive and $p$ becomes unimportant.
+
+To detect this transition in the data, we look at the eigenvalues corresponding to the two independent eigenevctors. 
+%
+Let $\mu_{i_1} \ge \mu_{i_2}$ denote these two eigenvalues. 
+%
+We plot  
+\begin{equation}
+ \frac{\log \mu_{i_1}}{\log \mu_{i_2}} ;
+\end{equation}
+when this ratio becomes small, it indicates that the data is effectively one-dimensional, as the second dimension is very small compared with the first. 
+%
+This is plotted in Figure~\ref{fig:chemotaxis_compare_timescales_evals} as a function of $\lambda$.
+
+For this specific example, we can compare the empirically detected change in dimensionality with the analytically-predicted shift. 
+%
+From \cite{...}, the relevant time scales for the velocity jump process are
+\begin{equation}
+\begin{aligned}
+\tau_{drift} =& \frac{L}{s} \\
+\tau_{diff} =& \frac{L^2 \lambda}{s^2}
+\end{aligned}
+\end{equation}
+%
+Here, we take $L = t_{max} s$ as a characteristic length scale for the simulations. 
+%
+When $\frac{\tau_{diff}}{\tau_{drift}} =  t_{max}  \lambda \gg 1$, then the dynamics are effectively one-dimensional. 
+%
+This ratio is also plotted in Figure~\ref{fig:chemotaxis_compare_timescales_evals} as a function of $\lambda$, and the transition is consisted with the transition predicted by the diffusion maps eigenvalues.
+
+\begin{figure}
+%
+\def\figheight{1.4in}
+%
+\begin{subfigure}[t]{0.31\textwidth}
+\centering
+\includegraphics[height=\figheight]{chemotaxis1_evals}
+\includegraphics[height=\figheight]{chemotaxis1_embed_good}
+\vspace{1.2in}
+%\includegraphics[height=1.5in]{chemotaxis1_embed_bad}
+\caption{}
+\end{subfigure}
+%
+\begin{subfigure}[t]{0.31\textwidth}
+\centering
+\includegraphics[height=\figheight]{chemotaxis2_evals}
+\includegraphics[height=\figheight]{chemotaxis2_embed_good}
+\includegraphics[height=\figheight]{chemotaxis2_embed_bad}
+\caption{}
+\end{subfigure}
+%
+\begin{subfigure}[t]{0.35\textwidth}
+\centering
+\includegraphics[height=\figheight]{chemotaxis3_evals}
+\includegraphics[height=\figheight]{chemotaxis3_embed_good}
+\includegraphics[height=\figheight]{chemotaxis3_embed_bad}
+\caption{}
+\end{subfigure}
+%
+\caption{Diffusion maps embedding for three different chemotaxis simulations.
+(a) $\lambda = 100$, $s = 10$.
+(b) $\lambda = 2500$, $s = 50$.
+(c) $\lambda = 6400$, $s = 80$.
+For each simulation, the eigenvalue spectrum is shown.
+From the spectra, we can see that the first two components are informative for (a), components 1 and 3 are informative for (b), and components 1 and  are informative for (c).
+The corresponding embeddings are shown below.
+For comparison, the embeddings using the first two components are also shown in (b) and (c), where clearly, the second component is a harmonic of the first and does not describe a new direction in the data.
+}
+%
+\end{figure}
+
+\begin{figure}[t]
+%
+\centering
+\includegraphics[width=\textwidth]{chemotaxis_compare_timescales_evals}
+%
+\caption{}
+%
+\label{fig:chemotaxis_compare_timescales_evals}
+%
+\end{figure}
+
+\section{Conclusions}
+
 This paper aims to bridge the fields of data mining and dynamical systems. 
@@
 
-\subsection{Detect dimensionality}
-
-\begin{figure}
-%
-\def\figheight{1.4in}
-%
-\begin{subfigure}[t]{0.31\textwidth}
-\includegraphics[height=\figheight]{chemotaxis1_evals}
-\includegraphics[height=\figheight]{chemotaxis1_embed_good}
-%\includegraphics[height=1.5in]{chemotaxis1_embed_bad}
-\end{subfigure}
-%
-\begin{subfigure}[t]{0.31\textwidth}
-\includegraphics[height=\figheight]{chemotaxis2_evals}
-\includegraphics[height=\figheight]{chemotaxis2_embed_good}
-\includegraphics[height=\figheight]{chemotaxis2_embed_bad}
-\end{subfigure}
-%
-\begin{subfigure}[t]{0.35\textwidth}
-\includegraphics[height=\figheight]{chemotaxis3_evals}
-\includegraphics[height=\figheight]{chemotaxis3_embed_good}
-\includegraphics[height=\figheight]{chemotaxis3_embed_bad}
-\end{subfigure}
-%
-\caption{}
-%
-\end{figure}
-
 \section*{Acknowledgment}

### 2015-01-02 chemotaxis/paper/draft1.tex
edited paper

@@
 
-\section{Issues in Manifold Learning/Diffusion maps}
+\section{Background on Manifold Learning and Some Relevant Issues}
 
-\subsection{Diffusion maps}
+\subsection{Manifold learning algorithm: diffusion maps}
 
@@
 
-\subsection{Observers}
+\subsection{Constructing appropriate observers}
 
@@
 
-\subsection{Metric}
+\subsection{Defining a meaningful metric}
 
@@
 
-\section{Identifying the Principal Components}
+\section{Identifying the Informative Eigenvectors}
 
@@
 If the resulting fit is good/accurate, then $\phi_{k}$ is most likely a harmonic of $\phi_1, \dots, \phi_{k-1}$. 
-
+%
 We will use a local linear function 
@@
 %
-The coefficients $\alpha$ and $\beta$ are {\em functions} of $\phi_{k-1}$ because we use a {\em local} linear fit; therefore, the coefficients change as a function of the domain.
+The coefficients $\alpha$ and $\beta$ are functions of $\phi_{k-1}$ because we use a {\em local} linear fit; therefore, the coefficients change as a function of the domain.
 
-At each point $\Phi_{k-1}(i)$, we fit a local linear function using the remaining $m-1$ data points. 
+At each point $\Phi_{k-1}(i)$, we approximate $\phi_k(i)$ by fitting a local linear function using the remaining $m-1$ data points. 
 %
@@
 We take $\epsilon = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k-1}(i)$.
+%
+The solution of \eqref{eq:opt_problem} is given by the solution of an eigenvector problem \cite{...}.
 
@@
 
+We will demonstrate the local linear regression algorithm for identifying informative eigenvectors on two synthetic data sets. 
+%
 Our first illustrative example will consist of two different swiss roll data sets.
@@
 \section{Chemotaxis: a case study}
+
+The issues discussed above must all be considered and addressed in order to successfully analyze data using manifold learning algorithms. 
+%
+To illustrate this point, we will use a model problem arising in the study of cellular dynamics. 
+%
+In this example, constructing appropriate observers with an informative metric will allow us to perform informative analysis. 
+%
+Furthermore, automatically detecting the ``important'' components will allow us to deduce when the dimension of the underlying dynamical system changes. 
+%
+In this way, we can deduce how many variables are required to write an accurate reduced model.
 
@@
 %
-\caption{}
+\caption{Detecting the change in dimensionality.
+The ratio of the log of the two ``meaningful'' diffusion maps eigenvalues (those which correspond to modes which are not harmonics) is shown in blue, and the ratio of the diffusive and drift time scales are shown in red.
+When the diffusive time scale is significantly larger than the drift time scale, the data is expected to be effectively one-dimensional.
+This transition is consistent with the diffusion maps eigenvalues: the data is effectively one-dimensional if the first eigenvalue is significantly larger than the second.
+}
 %

### 2015-01-05 chemotaxis/paper/draft1.tex
updated chemotaxis paper

@@
 
-Often, systems which are stochastic and noisy on the microscale exhibit coherent dynamics on the macroscale which are governed by only a few parameters.
-%
-For example, fluid flow, which can be described via the position and velocity of every fluid particle, is modeled at the macroscale using the Navier-Stokes equations.
-%
-However, in general, this mapping from microscale to macroscale is not always obvious.
-%
-We would like to extract a parameterization of data collected at the microscale which is consistent with the macroscopic behavior, without any {\em a priori} knowledge of the appropriate microscopic or macroscopic model.
-
-In this letter, we will demonstrate how this goal can be achieved through a data-driven method based on geometric analysis and manifold learning. 
-%
-Specifically, we aim to emphasize two particular aspects of manifold learning from a dynamical systems' point of view: (a) choosing the appropriate observers, especially in presence of stochastic dynamics, and (b) selecting an appropriate metric for comparison \cite{mallat2012group}.
-%
-We will show that both are essential for data-driven algorithms to successfully uncover the underlying system dynamics.
-
-We will illustrate these points using a model problem which arises in studies of cellular chemotaxis \cite{othmer2000diffusion}.
-%
-In this example, the macroscopic dynamics vary depending on the value of a single system parameter.
-%
-We will demonstrate how we can use diffusion maps \cite{coifman2005geometric}, a manifold learning technique, to {\em automatically} detect changes in the system's macroscopic dynamics from microscale data.
-%
-This case study exhibits several key points which allow us to demonstrate the strength of our approach.
-%
-First, this system has random microscopic behavior with smooth macroscopic dynamics governed by a single parameter which determines the regime/mode of the system. 
-%
-Second, it consists of an ensemble of individual stochastic particles which must be described using the appropriate observers.
-%
-Third, it has an analytic macroscopic description, which serves as a ground truth and will allow us to verify our results which are obtained in an unsupervised manner.
-%
-We will show that manifold learning algorithms based on statistical observers and with the appropriate affinity metric between pairs of observations uncovers a parameterization of the microscopic data which is consistent with the macroscopic model.
-%
-Furthermore, these data-driven methods allow us to detect changes in dynamical behavior resulting from changes in the system parameters.
-
 \section{Background on Manifold Learning and Some Relevant Issues}
@@
 %
-Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a constant vector which is not used as an embedding coordinate.
+Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a trivial constant vector which is not used as an embedding coordinate.
 
@@
 %
-Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ approach the eigenfunctions of the Laplacian (with Neumann boundary conditions) on $\mathcal{M}_d$.
+Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ approximate the eigenfunctions of the Laplacian (with Neumann boundary conditions) on $\mathcal{M}_d$.
 %
 The eigenvalues of the Laplacian operator $\tilde{\mu}_0, \tilde{\mu}_1, \dots, \tilde{\mu}_{m-1}$ are related to the eigenvalues we compute in diffusion maps by 
-\begin{equation}
+\begin{equation} \label{eq:evals_relationship}
 \mu_k = e^{-\tilde{\mu}_k t}.
@@
 
-\begin{itemize}
-\item Discuss the issues: harmonics
-\item Give simple example (Laplacian on strip)
-\end{itemize}
-
 In diffusion maps, we order the eigenvectors by the magnitude of the corresponding eigenvalues, in the hopes that the top eigenvectors provide us with a parameterization of the underlying manifold.
@@
 \end{equation}
-for $k_1, k_2 = 0, 1, 2, \dots$.
-%
-The corresponding eigenfunctions are 
+for $k_1, k_2 = 0, 1, 2, \dots$,
+and the corresponding eigenfunctions are 
 \begin{equation}
@@
 %
-However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to correspond to the two largest (non-trivial) eigenvalues (the largest eigenvalue will always be $\mu_{0,0} = 1$ and correspond to a constant eigenfunction).
+Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide a measure of $L_x$ versus $L_y$: as the ratio between $L_x$ and $L_y$ increases, the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases.
 %
-In fact, if $L_x > 2 L_y$, then $\mu_{2, 0} > \mu_{0, 1}$, and so the second (non-trivial) eigenvector (when the eigenvectors are sorted by their corresponding eigenvalues) will be a harmonic of the first and still parameterize the $x$ direction.
+However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to correspond to the two smallest (non-trivial) eigenvalues (the largest eigenvalue will always be $\tilde{\mu}_{0,0} = 1$ and correspond to a constant eigenfunction; the eigenvalues $\tilde{\mu}_k$ are related to the diffusion maps eigenvalues $\mu_k$ via \eqref{eq:evals_relationship}).
+%
+In fact, if $L_x > 2 L_y$, then $\tilde{\mu}_{2, 0} < \tilde{\mu}_{0, 1}$, and so the second (non-trivial) eigenvector (when the eigenvectors are ordered by their corresponding eigenvalues) will be a harmonic of the first and still parameterize the $x$ direction.
 %
 We therefore require a methodology to automatically detect which eigenvectors are harmonics of each other. 
+%
+Looking at the diffusion maps eigenvalues which correspond to modes which are {\em not} higher harmonics will then provide us with a measure of the relative lengths of the data set along the principal directions.
 
@@
 %
-The coefficients $\alpha$ and $\beta$ are functions of $\phi_{k-1}$ because we use a {\em local} linear fit; therefore, the coefficients change as a function of the domain.
+The coefficients $\alpha$ and $\beta$ are functions of $\phi_{k-1}$ because we use a {\em local} linear fit, and so the coefficients change as a function of the domain.
 
@@
 \begin{equation} \label{eq:opt_problem}
-\hat{\alpha} (i) , \hat{\beta}(i)  = \argmin_{\alpha, \beta} \sum_{j \ne i} w_i(\Phi_{k-1}(i), \Phi_{k-1}(j)) \left( \phi_{k}(j) - (\alpha + \beta^T \Phi_{k-1}(j)) \right)^2.
+\hat{\alpha}_k (i) , \hat{\beta}_k(i)  = \argmin_{\alpha, \beta} \sum_{j \ne i} w_i(\Phi_{k-1}(i), \Phi_{k-1}(j)) \left( \phi_{k}(j) - (\alpha + \beta^T \Phi_{k-1}(j)) \right)^2.
 \end{equation}
@@
 \begin{equation}
-w_i(\Phi_{k-1}(i), \Phi_{k-1}(j))  = \exp \left( - \frac{\|\Phi_{k-1}(i) - \Phi_{k-1} (j) \|^2}{\epsilon^2} \right).
+w_i(\Phi_{k-1}(i), \Phi_{k-1}(j))  = \exp \left( - \frac{\|\Phi_{k-1}(i) - \Phi_{k-1} (j) \|^2}{\epsilon^2} \right),
 \end{equation}
 %
-We take $\epsilon = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k-1}(i)$.
+and take $\epsilon = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k-1}(i)$.
 %
@@
 
-We then define the normalized leave-one-out cross-validation error as
+We then define the normalized leave-one-out cross-validation error for this local linear fit as
 \begin{equation} \label{eq:cv_error}
-r_{k} = \sqrt{ \frac{\sum_{i=1}^n \left( \phi_{k} (i) - (\hat{\alpha}(i) + \hat{\beta}(i)^T \Phi_{k-1}(i))  \right)^2} {\sum_{i=1}^n  \left( \phi_{k} (i) \right)^2 }}
+r_{k} = \sqrt{ \frac{\sum_{i=1}^n \left( \phi_{k} (i) - (\hat{\alpha}_k(i) + \hat{\beta}_k(i)^T \Phi_{k-1}(i))  \right)^2} {\sum_{i=1}^n  \left( \phi_{k} (i) \right)^2 }}
 \end{equation}
@@
 %
-When $r_{k}$ is small, $\phi_{k}$ can be accurately approximated from $\phi_1, \dots, \phi_{k-1}$, and is therefore a harmonic of a previous mode.
+Note that a small value of $r_k$ implies that $\phi_{k}$ can be accurately approximated from $\phi_1, \dots, \phi_{k-1}$, and is therefore a harmonic of a previous mode.
 %
@@
 In this way, we can deduce how many variables are required to write an accurate reduced model.
+
+Often, systems which are stochastic and noisy on the microscale exhibit coherent dynamics on the macroscale which are governed by only a few parameters.
+%
+For example, fluid flow, which can be described via the position and velocity of every fluid particle, is modeled at the macroscale using the Navier-Stokes equations.
+%
+However, in general, this mapping from microscale to macroscale is not always obvious.
+%
+We would like to extract a parameterization of data collected at the microscale which is consistent with the macroscopic behavior, without any {\em a priori} knowledge of the appropriate microscopic or macroscopic model.
+
+In this letter, we will demonstrate how this goal can be achieved through a data-driven method based on geometric analysis and manifold learning. 
+%
+Specifically, we aim to emphasize two particular aspects of manifold learning from a dynamical systems' point of view: (a) choosing the appropriate observers, especially in presence of stochastic dynamics, and (b) selecting an appropriate metric for comparison \cite{mallat2012group}.
+%
+We will show that both are essential for data-driven algorithms to successfully uncover the underlying system dynamics.
+
+We will illustrate these points using a model problem which arises in studies of cellular chemotaxis \cite{othmer2000diffusion}.
+%
+In this example, the macroscopic dynamics vary depending on the value of a single system parameter.
+%
+We will demonstrate how we can use diffusion maps \cite{coifman2005geometric}, a manifold learning technique, to {\em automatically} detect changes in the system's macroscopic dynamics from microscale data.
+%
+This case study exhibits several key points which allow us to demonstrate the strength of our approach.
+%
+First, this system has random microscopic behavior with smooth macroscopic dynamics governed by a single parameter which determines the regime/mode of the system. 
+%
+Second, it consists of an ensemble of individual stochastic particles which must be described using the appropriate observers.
+%
+Third, it has an analytic macroscopic description, which serves as a ground truth and will allow us to verify our results which are obtained in an unsupervised manner.
+%
+We will show that manifold learning algorithms based on statistical observers and with the appropriate affinity metric between pairs of observations uncovers a parameterization of the microscopic data which is consistent with the macroscopic model.
+%
+Furthermore, these data-driven methods allow us to detect changes in dynamical behavior resulting from changes in the system parameters.
 
@@
 \begin{figure}[t] 
-\includegraphics[width=8cm]{detect_change_eigenvalues}
+\centering
+\includegraphics[width=\textwidth]{detect_change_eigenvalues}
 \caption{The first two eigenvalues from diffusion maps as a function of $\lambda$.
@@
 \end{equation}
-when this ratio becomes small, it indicates that the data is effectively one-dimensional, as the second dimension is very small compared with the first. 
+when this ratio becomes small, the data are effectively one-dimensional, as the second dimension is very small compared with the first. 
 %

### 2015-01-05 chemotaxis/simulate_jump_process_EMD_test_lle.m

@@ -267,13 +267,15 @@ tau_drift = L./s_all;
 tau_diff = L.^2.*lambda_all./(s_all.^2);
 
 make_fig(4,3);
-[ax,h1,h2] = plotyy(lambda_all, log(eval1)./log(eval2), lambda_all, log(tau_diff./tau_drift));
+[ax,h1,h2] = plotyy(lambda_all, log(eval1)./log(eval2), lambda_all, tau_drift./tau_diff, 'plot','semilogy');
 set(h1, 'marker', '.', 'color','b')
 set(h2, 'marker', '.', 'color','r')
+% set(h2, 'yscale', 'log')
 set(ax(1), 'ycolor','b')
 set(ax(2), 'ycolor','r')
 ylabel(ax(1), 'log(\mu_{i_1})/log(\mu_{i_2})');
-ylabel(ax(2), 'log(\tau_{diff}/\tau_{drift})');
+ylab = ylabel(ax(2), '\tau_{diff}/\tau_{drift}');
+set(ylab, 'Units', 'Normalized', 'Position', [1.02, 0.5, 0]);
 xlabel('\lambda')
 % axis tight
 % axis square

### 2015-01-09 reducible_SDE/paper/SIADS_paper/make_schematic.m
updated multiscale paper

@@ -12,7 +12,7 @@ x = repmat(linspace(0, 1, npoints), ntraj, 1);
 
 y = zeros(size(x));
 y(1, :) = exp(-10*x(1,:));
-y(2, :) = -exp(-10*x(2,:));
+y(2, :) = -exp(-20*x(2,:));
 
 % figure;
 % plot(x, y, '.')
@@ -20,11 +20,17 @@ y(2, :) = -exp(-10*x(2,:));
 % set(gca, 'ytick', [])
 
 make_fig(3, 3);
-scatter(x(:), y(:), 50, -x(:),'.')
+scatter(x(:), y(:), 50, x(:)/max(x(:)),'.')
 set(gca, 'xtick', [])
 set(gca, 'ytick', [])
 set(gca, 'ylim', [-1 1])
+xlabel('x_1')
+ylabel('x_2')
 % axis off
+colormap(flipud(jet))
+h = colorbar;
+xlabel(h,'t');
+set(h,'YTick',[0 1])
 saveas(gcf, 'schematic_DS1.eps', 'epsc');
 
 % figure;
@@ -38,11 +44,17 @@ x = linspace(0, 1, npoints);
 y = randn(size(x)) .* max(10*exp(-8*x), 2);
 
 make_fig(3, 3);
-scatter(x, smooth(y, 11), 50, -x,'.')
+scatter(x, smooth(y, 11), 50, x/max(x),'.')
 set(gca, 'xtick', [])
 set(gca, 'ytick', [])
 set(gca, 'ylim', [-3 3])
+xlabel('x_1')
+ylabel('x_2')
 % axis off
+colormap(flipud(jet))
+h = colorbar;
+xlabel(h,'t');
+set(h,'YTick',[0 1])
 saveas(gcf, 'schematic_DS2.eps', 'epsc');
 
 

### 2015-02-22 chemotaxis/paper/draft1.tex
updated chemotaxis paper

@@
 
+\item For the example of a square/strip, we know everything analytically, and identifying the ``true'' principal components is a problem/not obvious, and depends both on geometry and density
+
+\item While we can ignore/address nonuniform densities through normalization of the Laplace operator, we still have the problem of identifying the ``true'' components due to geometry.
+
 \item
@@
 
-\subsection{Manifold learning algorithm: diffusion maps}
+\subsection{Manifold learning and the Laplace operator}
 
-Let $z(t) \in \mathbb{R}^n$ denote an $n$-dimensional observation of the system at time $t$. 
+We consider a continuous $d$-dimensional manifold $\mathcal{M}_d$ embedded in $\mathbb{R}^n$, and we are interested in uncovering a parameterization of this manifold. 
 %
-Assume we are given $m$ observations $z(1), \dots, z(m)$, which lie on a $d$-dimensional, nonlinear manifold $\mathcal{M}_d$, with $d \ll n$. 
+Many algorithms have been proposed for such a task.
 %
-Our goal is to uncover a $d$-dimensional parameterization of the $m$ observations that respects the underlying manifold geometry.
+Historically, principal component analysis (PCA) \cite{shlens2005tutorial} has been used to parmeterize linear manifolds (planes and hyperplanes), where the principal components parameterize the underlying hyperplane.
+%
+However, in the case when the manifold is curved, such linear techniques can fail to capture the underlying structure.
+
+In recent years, it has been noted that eigenfunctions of the Laplace-Beltrami operator on the manifold with Neumann boundary conditions \cite{....} can provide coordinates which parameterize the manifold.
+%
+To provide intuition as to why this is the case, consider a two-dimensional strip with edge lengths $L_x$ and $L_y$. 
+%
+The eigenvalues of the Laplacian with Neumann boundary conditions are given by
+\begin{equation}
+\tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_x} \right)^2 + \left( \frac{k_2 \pi}{L_y} \right)^2
+\end{equation}
+for $k_1, k_2 = 0, 1, 2, \dots$,
+and the corresponding eigenfunctions are 
+\begin{equation}
+\tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi x}{L_x} \right) \cos \left( \frac{k_2 \pi y}{L_y} \right)
+\end{equation}
+where $x$ and $y$ denote the two coordinates of the rectangular domain. 
+%
+We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi x}{L_x} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi y}{L_y} \right)$ are one-to-one with the $x$ and $y$ coordinates, respectively, and therefore yield a parameterization of the underlying manifold.
+%
+Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide a measure of $L_x$ versus $L_y$: as the ratio between $L_x$ and $L_y$ increases, the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases.
+
+Although these eigenfunctions can only be written analytically for very special cases, it has been observed empirically that the eigenfunctions often yield appropriate/good coordinates to parameterize the underlying manifold.
+
+%From the eigenvalues, 
+%\begin{equation}
+%\sqrt{\frac{\log(\mu_{1,0})}{\log(\mu_{0,1})}} = 
+%\frac{L_y}{L_x}
+%\end{equation}
+%and so looking at the ratio of the log of the eigenvalues (modulo harmonics) should inform us as to the lengths of the different directions in the manifold.
+
+\subsection{Discrete approximation of the Laplacian: diffusion maps}
+
+In most applications, we are not given a description of the continuous manifold. 
+%
+Instead, we are often given data {\em sampled} from the underlying manifold $\mathcal{M}_d$, and we would like to uncover a parameterization of the data.
+%
+We will do this by constructing a matrix which approximates the Laplace-Beltrami operator with Neumann boundary conditions. 
+%
+The eigenvectors of this matrix then approximate the eigenfunctions of this operator.
+
+Assume we are given $m$ observations $z(1), \dots, z(m) \in \mathbb{R}^n$, which lie on a $d$-dimensional, nonlinear manifold $\mathcal{M}_d$, with $d \ll n$. 
 %
@@
 %
-The eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ provide the embedding coordinates for the data, such that
+In the limit of infinite data uniformly sampled from $\mathcal{M}_d$ ($m \rightarrow \infty$ and $\epsilon \rightarrow 0$ with the appropriate rate), the matrix $I-A$ approaches the Laplace-Beltrami operator on the manifold $\mathcal{M}_d$ \cite{...}.
+%
+Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ approximate the eigenfunctions of the Laplacian (with Neumann boundary conditions) on $\mathcal{M}_d$, such that
 $\phi_{j}(i)$ is the $j^{th}$ embedding coordinate for $z(i)$.
 %
-The ``importance'' of coordinate $\phi_j$ is quantified by the corresponding eigenvalue $\mu_j$, e.g., the relative importance of $\phi_1$ versus $\phi_2$ can be measured by comparing $\mu_1$ and $\mu_2$.
+The eigenvalues of the Laplacian operator $\tilde{\mu}_0, \tilde{\mu}_1, \dots, \tilde{\mu}_{m-1}$ are related to the eigenvalues of $A$, $\mu_0, \mu_1, \dots, \mu_{m-1}$, by 
+\begin{equation} \label{eq:evals_relationship}
+\mu_k = e^{-\tilde{\mu}_k t}.
+\end{equation}
 %
@@
 
-In the limit of infinite data ($m \rightarrow \infty$ and $\epsilon \rightarrow 0$ with the appropriate rate), the matrix $I-A$ approaches the Laplacian operator on the manifold $\mathcal{M}_d$ \cite{...}.
+However, the data need not be sampeld uniformly from the underlying manifold.
 %
-Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ approximate the eigenfunctions of the Laplacian (with Neumann boundary conditions) on $\mathcal{M}_d$.
-%
-The eigenvalues of the Laplacian operator $\tilde{\mu}_0, \tilde{\mu}_1, \dots, \tilde{\mu}_{m-1}$ are related to the eigenvalues we compute in diffusion maps by 
-\begin{equation} \label{eq:evals_relationship}
-\mu_k = e^{-\tilde{\mu}_k t}.
-\end{equation}
+If the data $z(1), \dots, z(m)$ are sampled from $\mathcal{M}_d$ with some density $U$, then the matrix $I-A$ approximates the Fokker-Planck operator with potential $2U$, e.g., $(I-A) \phi \rightarrow \nabla^2 \phi - 2\nabla\phi \cdot \nabla U$.
 
-\subsection{Constructing appropriate observers}
+TODO: add two sets of figures: one with data uniformly sampled, one with data with Gaussian distribiution
 
-\begin{itemize}
-\item Physical examples (e.g., $\alpha$-carbons for molecules)
-\item Amit's bispectrum
-\item Scattering transform
-\item Cite PNAS, stating that the histograms are good since they linearize any noise. 
-For the chemotaxis example, we use histograms of the particle positions as observers \cite{talmon2013empirical}. 
-%
-Histograms are invariant to the indexing of the particles, while retaining the spatial locations of the particles.
-\end{itemize}
+Common practice is to use these tools for nonlinear data where these things are coupled. Show swiss roll and torus.
 
-\subsection{Defining a meaningful metric}
-
-\begin{itemize}
-\item Euclidean (not good, especially for histograms)
-\item Mahalanobis (good, cite PNAS, since it noise robust)
-\item Here we use EMD since it is multiscale.
-%
-Instead of the standard Euclidean distance, we use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between pairs of histograms. 
-%
-Conceptually, the EMD measures how much ``work'' it takes to transform one probability density into another.
-%
-It therefore not only considers where the densities are inconsistent, but also how far apart the inconsistencies are.
-%
-Although the brute-force computation of the EMD is computationally expensive, there has been a plethora of work in developing efficient algorithms for its computation \cite{Pele-eccv2008, Pele-iccv2009}.
-%
-For the special case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions \cite{rubner2000perceptual}, which can be estimated from histograms as
-\begin{equation}
-\| z(i) - z(j) \|_{EMD} \approx \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|
-\end{equation}
-where the histograms $z(i), z(j)$ are defined on equally-spaced bins in $\mathbb{R}$, and $z_k$ denotes the $k^{th}$ bin. 
-%
-\item Emphasize stability and regularity issues.
-\end{itemize}
-
-\section{Identifying the Informative Eigenvectors}
+\subsection{Identifying the Informative Eigenvectors using Local Linear Regression}
 
@@
 %
-We will illustrate this issue with a simple two-dimensional strip where we know the analytical form of the eigenfunctions and can explicitly highlight the issue at hand.
-
-Consider the case where $\mathcal{M}_d$ is a $2$-dimensional rectangular domain with edge lengths $L_x  > L_y$. 
+Again, we consider the case where $\mathcal{M}_d$ is a $2$-dimensional rectangular domain with edge lengths $L_x  > L_y$. 
 %
@@
 Looking at the diffusion maps eigenvalues which correspond to modes which are {\em not} higher harmonics will then provide us with a measure of the relative lengths of the data set along the principal directions.
-
-%From the eigenvalues, 
-%\begin{equation}
-%\sqrt{\frac{\log(\mu_{1,0})}{\log(\mu_{0,1})}} = 
-%\frac{L_y}{L_x}
-%\end{equation}
-%and so looking at the ratio of the log of the eigenvalues (modulo harmonics) should inform us as to the lengths of the different directions in the manifold.
-
-\subsection{Algorithm: Local Linear Regression}
 
@@
 
-%\subsection{Stochastic particles}
+\subsection{Problem description}
 
@@
 \end{figure}
+
+\subsection{Observers and Metrics}
+
+\subsubsection{Constructing appropriate observers}
+
+\begin{itemize}
+\item Physical examples (e.g., $\alpha$-carbons for molecules)
+\item Amit's bispectrum
+\item Scattering transform
+\item Cite PNAS, stating that the histograms are good since they linearize any noise. 
+For the chemotaxis example, we use histograms of the particle positions as observers \cite{talmon2013empirical}. 
+%
+Histograms are invariant to the indexing of the particles, while retaining the spatial locations of the particles.
+\end{itemize}
+
+\subsubsection{Defining a meaningful metric}
+
+\begin{itemize}
+\item Euclidean (not good, especially for histograms)
+\item Mahalanobis (good, cite PNAS, since it noise robust)
+\item Here we use EMD since it is multiscale.
+%
+Instead of the standard Euclidean distance, we use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between pairs of histograms. 
+%
+Conceptually, the EMD measures how much ``work'' it takes to transform one probability density into another.
+%
+It therefore not only considers where the densities are inconsistent, but also how far apart the inconsistencies are.
+%
+Although the brute-force computation of the EMD is computationally expensive, there has been a plethora of work in developing efficient algorithms for its computation \cite{Pele-eccv2008, Pele-iccv2009}.
+%
+For the special case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions \cite{rubner2000perceptual}, which can be estimated from histograms as
+\begin{equation}
+\| z(i) - z(j) \|_{EMD} \approx \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|
+\end{equation}
+where the histograms $z(i), z(j)$ are defined on equally-spaced bins in $\mathbb{R}$, and $z_k$ denotes the $k^{th}$ bin. 
+%
+\item Emphasize stability and regularity issues.
+\end{itemize}
+
+Note: histograms are not essential for computing EMD, however, they make it computationally feasible. Furthermore, histograms are good because they are robust to noise (cite PNAS).
+
+\subsection{Results}
 

### 2015-02-22 chemotaxis/strip.m

@@ -1,0 +1,46 @@
+clear all
+close all
+
+%%
+
+Lx = 3;
+Ly = 1;
+
+
+delta = 0.1;
+
+[X, Y] = meshgrid(0:delta:Lx, 0:delta:Ly);
+
+figure;
+h = pcolor(X, Y, cos(pi*X/Lx));
+set(h, 'edgecolor','none');
+shading interp
+axis equal
+
+figure;
+h = pcolor(X, Y, cos(pi*Y/Ly));
+set(h, 'edgecolor','none');
+shading interp
+axis equal
+
+%%
+
+rng(321);
+
+n = 1000;
+data = rand(n, 2);
+data(:,1) = data(:,1) * Lx;
+data(:,2) = data(:,2) * Ly;
+
+W = squareform(pdist(data)).^2;
+eps = 0.5^2;
+
+[V, D] = dmaps(W, eps, 10);
+
+figure;
+scatter(data(:,1),data(:,2),50, V(:,2),'.')
+axis equal
+
+figure;
+scatter(data(:,1),data(:,2),50, V(:,4),'.')
+axis equal

### 2015-02-22 chemotaxis/swissroll2.m

@@ -1,0 +1,128 @@
+clear all
+close all
+
+%%
+
+N = 1500; % number of points considered
+t = rand(1,N);
+t = sort(4*pi*sqrt(t))'; 
+
+h = 15*pi;
+
+z = h*rand(N,1); % random heights
+x = (t+.1).*cos(t);
+y = (t+.1).*sin(t);
+data = [x,y,z]; % data
+
+W = squareform(pdist(data)).^2;
+eps = 5;
+
+neigs = 10;
+
+[V, D] = dmaps(W, eps, neigs);
+
+eps_med_scale = 3;
+res = compute_residuals_DMAPS(V, eps_med_scale);
+
+%%
+
+
+figure;
+plot3(x,y,z,'.', 'markersize', 10)
+view(-20, 70)
+xlabel('x')
+ylabel('y')
+zlabel('z')
+axis equal
+grid on
+
+figure;
+scatter3(x,y,z,50,sign(corr(V(:,2),t))*V(:,2),'.')
+view(-20, 75)
+xlabel('x')
+ylabel('y')
+zlabel('z')
+axis equal
+grid on
+cbar = colorbar('peer',gca); 
+set(get(cbar,'xlabel'),'String','\phi_1');
+
+figure;
+scatter3(x,y,z,50,sign(corr(V(:,3),z))*V(:,3),'.')
+view(-20, 75)
+xlabel('x')
+ylabel('y')
+zlabel('z')
+axis equal
+grid on
+cbar = colorbar('peer',gca); 
+set(get(cbar,'xlabel'),'String','\phi_2');
+
+figure;
+make_colored_bars(diag(D(2:end,2:end)), res(2:end))
+set(gca, 'ylim', [0.9 1])
+xlabel('k')
+ylabel('\mu_k')
+axis square
+
+%%
+
+h = 6*pi;
+
+z = h*rand(N,1); % random heights
+
+data = [x,y,z]; % data
+
+W = squareform(pdist(data)).^2;
+eps = 2;
+
+neigs = 10;
+
+[V, D] = dmaps(W, eps, neigs);
+
+eps_med_scale = 3;
+res = compute_residuals_DMAPS(V, eps_med_scale);
+
+%%
+
+
+figure;
+plot3(x,y,z,'.', 'markersize', 10)
+view(-20, 70)
+xlabel('x')
+ylabel('y')
+zlabel('z')
+axis equal
+grid on
+
+figure;
[…]

### 2015-02-22 chemotaxis/test_sum_evals.m

@@ -1,0 +1,193 @@
+clear all
+close all
+
+%%
+
+scale_vec = repmat(2:10, 1 ,3);
+
+lambda_ratio = zeros(size(scale_vec));
+
+n = 500;
+
+figure;
+for i=1:length(scale_vec)
+    
+    scale = scale_vec(i);
+    
+    
+    data = rand(n, 2);
+    data(:,1) = data(:,1) * scale;
+    
+    %     figure;
+    %     scatter(data(:,1),data(:,2))
+    %     axis equal
+    
+    W = squareform(pdist(data)).^2;
+    eps = median(W(:))/9;
+    
+    [V, D] = dmaps(W, eps, 15);
+    
+    eps_med_scale = 3;
+    res = compute_residuals_DMAPS(V, eps_med_scale);
+    
+    %     figure;
+    %     make_colored_bars(diag(D(2:end, 2:end)), res(2:end))
+    %     xlabel('k')
+    %     ylabel('\mu_k')
+    
+    %     idx = find(res > 0.2);
+    %     D(idx, idx);
+    [~, idx] = sort(res, 'descend');
+    idx = idx(1:2);
+    if idx(1) > idx(2)
+        tmp = idx(2);
+        idx(2) = idx(1);
+        idx(1) = tmp;
+    end
+    
+    subplot(length(scale_vec), 2, 2*i-1);
+    plot(data(:,1),V(:,idx(1)),'.')
+    
+    subplot(length(scale_vec), 2, 2*i);
+    plot(data(:,2),V(:,idx(2)),'.')
+    
+    lambda_ratio(i) = 1/sqrt(log(D(idx(1), idx(1)))/log(D(idx(2), idx(2))));
+    
+end
+
+figure;
+plot(scale_vec, lambda_ratio,'o');
+hold on
+
+%%
+scale_vec = repmat(2:10, 1 ,3);
+
+lambda_ratio = zeros(size(scale_vec));
+
+n = 500;
+
+figure;
+for i=1:length(scale_vec)
+    
+    scale = scale_vec(i);
+    
+    
+    data = rand(n, 2);
+    data(:,1) = data(:,1) * scale;
+    
+    %     figure;
+    %     scatter(data(:,1),data(:,2))
+    %     axis equal
+    
+    W = squareform(pdist(data)).^2;
+    eps = median(W(:))/9;
+    
[…]

### 2015-02-22 chemotaxis/paper/draft1.tex
updated chemotaxis paper

@@
 
+\begin{figure}
+\includegraphics[width=0.5\textwidth]{strip_cnts1}
+\includegraphics[width=0.5\textwidth]{strip_cnts2}
+\end{figure}
+
 %From the eigenvalues, 
@@
 
+\begin{figure}
+\includegraphics[width=0.5\textwidth]{strip_discrete1}
+\includegraphics[width=0.5\textwidth]{strip_discrete2}
+\includegraphics[width=0.5\textwidth]{strip_nonuniform1}
+\includegraphics[width=0.5\textwidth]{strip_nonuniform2}
+
+\end{figure}
+
 TODO: add two sets of figures: one with data uniformly sampled, one with data with Gaussian distribiution
@@
 %
-Again, we consider the case where $\mathcal{M}_d$ is a $2$-dimensional rectangular domain with edge lengths $L_x  > L_y$. 
+Again, for illustrative purposes, we consider the case where $\mathcal{M}_d$ is a $2$-dimensional rectangular domain with edge lengths $L_x  > L_y$. 
 %
-The eigenvalues of the Laplacian with Neumann boundary conditions are given by
-\begin{equation}
-\tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_x} \right)^2 + \left( \frac{k_2 \pi}{L_y} \right)^2
-\end{equation}
-for $k_1, k_2 = 0, 1, 2, \dots$,
-and the corresponding eigenfunctions are 
-\begin{equation}
-\tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi x}{L_x} \right) \cos \left( \frac{k_2 \pi y}{L_y} \right)
-\end{equation}
-where $x$ and $y$ denote the two coordinates of the rectangular domain. 
-%
-We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi x}{L_x} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi y}{L_y} \right)$ are one-to-one with the $x$ and $y$ coordinates, respectively, and therefore yield a parameterization of the underlying manifold.
-%
-Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide a measure of $L_x$ versus $L_y$: as the ratio between $L_x$ and $L_y$ increases, the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases.
+We recall that the eigenfunctions $\phi_{1,0} = \cos \left( \frac {\pi x}{L_x} \right)$ and  $\phi_{0,1} = \cos \left( \frac {\pi y}{L_y} \right)$ provide embedding coordinates for the manifold $|mathcal{M}_d$.
 %
@@
 
+We will demonstrate the local linear regression algorithm for identifying informative eigenvectors on two synthetic data sets. 
+%
+Both of these data sets are nonlinear manifolds which provide examples of the flexibility of diffusion maps. 
+%
+However, these data sets will also illustrate the issues of harmonics.
+
 \subsubsection{Swiss Roll}
 
-We will demonstrate the local linear regression algorithm for identifying informative eigenvectors on two synthetic data sets. 
-%
 Our first illustrative example will consist of two different swiss roll data sets.
@@
 %
-In this example, constructing appropriate observers with an informative metric will allow us to perform informative analysis. 
+%In this example, constructing appropriate observers with an informative metric will allow us to perform informative analysis. 
 %
-Furthermore, automatically detecting the ``important'' components will allow us to deduce when the dimension of the underlying dynamical system changes. 
+Automatically detecting the ``important'' components will allow us to deduce when the dimension of the underlying dynamical system changes. 
 %
@@
 We would like to extract a parameterization of data collected at the microscale which is consistent with the macroscopic behavior, without any {\em a priori} knowledge of the appropriate microscopic or macroscopic model.
+%
+Furthermore, determining the dimensionality of such a parameterization can inform us to the requisite dimensionality of a reduced model which captures the relevant macroscopic behavior.
 
-In this letter, we will demonstrate how this goal can be achieved through a data-driven method based on geometric analysis and manifold learning. 
+%In this letter, we will demonstrate how this goal can be achieved through a data-driven method based on geometric analysis and manifold learning. 
 %
-Specifically, we aim to emphasize two particular aspects of manifold learning from a dynamical systems' point of view: (a) choosing the appropriate observers, especially in presence of stochastic dynamics, and (b) selecting an appropriate metric for comparison \cite{mallat2012group}.
+%Specifically, we aim to emphasize two particular aspects of manifold learning from a dynamical systems' point of view: (a) choosing the appropriate observers, especially in presence of stochastic dynamics, and (b) selecting an appropriate metric for comparison \cite{mallat2012group}.
 %
-We will show that both are essential for data-driven algorithms to successfully uncover the underlying system dynamics.
+%We will show that both are essential for data-driven algorithms to successfully uncover the underlying system dynamics.
 
-We will illustrate these points using a model problem which arises in studies of cellular chemotaxis \cite{othmer2000diffusion}.
+We will illustrate these points using a model problem which arises in studies of cellular chemotaxis \cite{othmer2000diffusion}, where 
+biological cells work together and exhibit coherent macroscopic dynamics regulated by extracellular sensed signals, in order to accomplish tasks such as finding food or navigating away from toxins.
 %
+Several microscopic models have been proposed to describe chemotaxis dynamics \cite{othmer1988models, codling2008random}.
+%
+We will analyze one such model described by a one-dimensional velocity jump process \cite{othmer2000diffusion}.
+%
+In this model, each cell is initialized with a position and a velocity on a line, and the dynamics of each cell are governed by a stochastic process.
+%
+At random times, a cell will ``turn around'' and switch the direction of its velocity (this turning is controlled by extracellular signals). 
+%
+%We will analyze the dynamics of collections of such cells/particles.
+
 In this example, the macroscopic dynamics vary depending on the value of a single system parameter.
@@
 %\subsection{Chemotaxis ``story''}
-
-Biological cells are one common example of a system where microscopic entities work together and exhibit coherent macroscopic dynamics, in order to accomplish tasks such as finding food or navigating away from toxins.
-%
-The cellular dynamics are often governed by extracellular sensed signals; this movement is known as chemotaxis. 
-%
-Several microscopic models have been proposed to describe chemotaxis dynamics \cite{othmer1988models, codling2008random}.
-%
-We will analyze one such model described by a one-dimensional velocity jump process \cite{othmer2000diffusion}.
-%
-In this model, each cell is initialized with a position and a velocity on a line, and the dynamics of each cell are governed by a stochastic process.
-%
-At random times, a cell will ``turn around'' and switch the direction of its velocity (this turning is controlled by extracellular signals). 
-%
-%We will analyze the dynamics of collections of such cells/particles.
 
@@
 When $\lambda \rightarrow \infty$, the dynamics are described by one heat equation, and the initial conditions are insignificant -- the velocity distribution quickly equilibrates and we see purely diffusive behavior.
-
+%
 For general dynamical systems, the governing macroscopic variables and asymptotic behavior may not be immediately obvious, and instead need to be inferred from the microscopic data.
 %
-We expect the simulation data to lie on a low-dimensional manifold which is parameterized by the macroscopic variables.
+We will demonstrate how we can use diffusion maps to parameterize the simulation data, as well as detect the effective dimensionality of the data.
 %
-We use diffusion maps \cite{coifman2005geometric}, which is a manifold learning technique, to uncover such low-dimensional structure and to analyze the results of the microscopic chemotaxis simulations.
+%We expect the simulation data to lie on a low-dimensional manifold which is parameterized by the macroscopic variables.
 %
-Compared to principal components analysis (PCA) \cite{shlens2005tutorial}, a standard linear technique, diffusion maps provides a parameterization of data which lie on a low-dimensional, (possibly) {\em nonlinear} structure in a high-dimensional space.
+%We use diffusion maps \cite{coifman2005geometric}, which is a manifold learning technique, to uncover such low-dimensional structure and to analyze the results of the microscopic chemotaxis simulations.
 %
-Diffusion maps is one of many recently introduced manifold learning techniques that could be used to analyzed this data \cite{roweis2000nonlinear, tenenbaum2000global, Belkin2003}.
+%Compared to principal components analysis (PCA) \cite{shlens2005tutorial}, a standard linear technique, diffusion maps provides a parameterization of data which lie on a low-dimensional, (possibly) {\em nonlinear} structure in a high-dimensional space.
+%
+%Diffusion maps is one of many recently introduced manifold learning techniques that could be used to analyzed this data \cite{roweis2000nonlinear, tenenbaum2000global, Belkin2003}.
 
@@
 
-\subsubsection{Constructing appropriate observers}
+In general, manifold learning techniques have two essential components.
+%
+One is the appropriate {\em observers} of the system. 
+%
+These observers should be informative as to the state of the system, as well as insensitive to noise in the system. 
+%
+Two is a {\em distance metric} between the observations that captures a notion of locality: observations which we perceive to be similar should have a small distance. 
+%
+%For this particular example, we will use histograms of the particle positions as the observations, and the EMD as the distance metric.
 
-\begin{itemize}
-\item Physical examples (e.g., $\alpha$-carbons for molecules)
-\item Amit's bispectrum
-\item Scattering transform
-\item Cite PNAS, stating that the histograms are good since they linearize any noise. 
+Appropriate observers can be constructed in several different ways. 
+%
+For simple data, such as the illustrative examples in the previous section, the observers are simply the data points themselves. 
+%
+However, for more complex data, the raw data may contain other sources of variability which are not meaningful to the system dynamics. 
+%
+For some examples, physical intuition and {\em a priori} knowledge can aid in constructing appropriate observers. 
+%
+For example, when looking at a protein, the ``observers'' may be the locations of the $\alpha$-carbon atoms. 
+%
+Observers can also factor out other sources of variation. 
+%
+For example, the bispectrum \cite{...} and the scattering transform \cite{...} are both observers which are invariant to rotations. 
+%
+Observers should also be robust to noise. 
+%
+The scattering transform \cite{...} is robust to small deformations, and histograms serve to linearize noise \cite{...}.
+
+After constructing appropriate observers, we must consider what metric to use in the diffusion maps calculation. 
+%
+Again, the simplest option is to use the Euclidean distance, but this can often prove uninformative. 
+%
+For molecular simulation data, it is often appropriate to use the root-mean-squared displacement {\em after} aligning configurations with respect to rotations, as global rotations of the entire molecule are not meaningful.
+%
+For dynamical systems which are obscured by large variance noise, the Mahalanobis distance often yields a more informative parameterization of the data \cite{...}.
+%
+Another option is the Earth Mover's Distance, a multi scale distance metric which is stable and robust to noise.
+
 For the chemotaxis example, we use histograms of the particle positions as observers \cite{talmon2013empirical}. 
@@
 Histograms are invariant to the indexing of the particles, while retaining the spatial locations of the particles.
-\end{itemize}
-
-\subsubsection{Defining a meaningful metric}
-
-\begin{itemize}
-\item Euclidean (not good, especially for histograms)
-\item Mahalanobis (good, cite PNAS, since it noise robust)
-\item Here we use EMD since it is multiscale.
 %
@@
 %
-\item Emphasize stability and regularity issues.
-\end{itemize}
-
 Note: histograms are not essential for computing EMD, however, they make it computationally feasible. Furthermore, histograms are good because they are robust to noise (cite PNAS).
 
+%\subsubsection{Constructing appropriate observers}
+%
+%\begin{itemize}
+%\item Physical examples (e.g., $\alpha$-carbons for molecules)
+%\item Amit's bispectrum
+%\item Scattering transform
+%\item Cite PNAS, stating that the histograms are good since they linearize any noise. 
+%For the chemotaxis example, we use histograms of the particle positions as observers \cite{talmon2013empirical}. 
+%%
+%Histograms are invariant to the indexing of the particles, while retaining the spatial locations of the particles.
+%\end{itemize}
+%
+%\subsubsection{Defining a meaningful metric}
+%
+%
+%\begin{itemize}
+%\item Euclidean (not good, especially for histograms)
+%\item Mahalanobis (good, cite PNAS, since it noise robust)
+%\item Here we use EMD since it is multiscale.
+%%
+%Instead of the standard Euclidean distance, we use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between pairs of histograms. 
+%%
+%Conceptually, the EMD measures how much ``work'' it takes to transform one probability density into another.
+%%
+%It therefore not only considers where the densities are inconsistent, but also how far apart the inconsistencies are.
+%%
+%Although the brute-force computation of the EMD is computationally expensive, there has been a plethora of work in developing efficient algorithms for its computation \cite{Pele-eccv2008, Pele-iccv2009}.
+%%
+%For the special case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions \cite{rubner2000perceptual}, which can be estimated from histograms as
+%\begin{equation}
+%\| z(i) - z(j) \|_{EMD} \approx \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|
+%\end{equation}
+%where the histograms $z(i), z(j)$ are defined on equally-spaced bins in $\mathbb{R}$, and $z_k$ denotes the $k^{th}$ bin. 
+%%
+%\item Emphasize stability and regularity issues.
+%\end{itemize}
+
 \subsection{Results}
-
-In general, manifold learning techniques have two essential components.
-%
-One is the appropriate {\em observers} of the system. 
-%
-These observers should be informative as to the state of the system, as well as insensitive to noise in the system. 
-%
-Two is a {\em distance metric} between the observations that captures a notion of locality: observations which we perceive to be similar should have a small distance. 
-%
-For this particular example, we will use histograms of the particle positions as the observations, and the EMD as the distance metric.
 

### 2015-02-23 chemotaxis/paper/draft1.tex
made small edits to chemotaxis paper

@@
 
-We consider a continuous $d$-dimensional manifold $\mathcal{M}_d$ embedded in $\mathbb{R}^n$, and we are interested in uncovering a parameterization of this manifold. 
+We first consider the problem of parameterizing a continuous $d$-dimensional manifold $\mathcal{M}_d$ embedded in $\mathbb{R}^n$.
 %
-Many algorithms have been proposed for such a task.
+For linear hyperplanes,  the principal axes parameterize the manifold.
 %
-Historically, principal component analysis (PCA) \cite{shlens2005tutorial} has been used to parmeterize linear manifolds (planes and hyperplanes), where the principal components parameterize the underlying hyperplane.
-%
-However, in the case when the manifold is curved, such linear techniques can fail to capture the underlying structure.
+However, in the case when the manifold is curved, it is more difficult to define such a set of coordinates.
 
-In recent years, it has been noted that eigenfunctions of the Laplace-Beltrami operator on the manifold with Neumann boundary conditions \cite{....} can provide coordinates which parameterize the manifold.
+In recent years, it has been noted that eigenfunctions of the Laplace-Beltrami operator with Neumann boundary conditions on the manifold \cite{....} can parameterize the manifold. 
 %
@@
 %
-The eigenvalues of the Laplacian with Neumann boundary conditions are given by
-\begin{equation}
+The eigenvalues of the Laplace-Beltrami operator with Neumann boundary conditions are given by
+\begin{equation} \label{eq:evals}
 \tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_x} \right)^2 + \left( \frac{k_2 \pi}{L_y} \right)^2
@@
 and the corresponding eigenfunctions are 
-\begin{equation}
+\begin{equation} \label{eq:efuncs}
 \tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi x}{L_x} \right) \cos \left( \frac{k_2 \pi y}{L_y} \right)
@@
 \begin{figure}
-\includegraphics[width=0.5\textwidth]{strip_cnts1}
-\includegraphics[width=0.5\textwidth]{strip_cnts2}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{strip_cnts1}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{strip_cnts2}
+\end{subfigure}
+\caption{Two-dimensional strip colored by the eigenfunction (a) $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi x}{L_x} \right)$, and (b) $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi y}{L_y} \right)$ .}
+\label{fig:strip_efuncs}
 \end{figure}
@@
 %
-Instead, we are often given data {\em sampled} from the underlying manifold $\mathcal{M}_d$, and we would like to uncover a parameterization of the data.
+Instead, we are given data {\em sampled} from the underlying manifold $\mathcal{M}_d$, and we would like to uncover a parameterization of the data.
 %
@@
 
-Assume we are given $m$ observations $z(1), \dots, z(m) \in \mathbb{R}^n$, which lie on a $d$-dimensional, nonlinear manifold $\mathcal{M}_d$, with $d \ll n$. 
+Assume we are given $m$ observations $z(1), \dots, z(m) \in \mathcal{M}_d$. 
 %
@@
 %
-Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ approximate the eigenfunctions of the Laplacian (with Neumann boundary conditions) on $\mathcal{M}_d$, such that
-$\phi_{j}(i)$ is the $j^{th}$ embedding coordinate for $z(i)$.
-%
-The eigenvalues of the Laplacian operator $\tilde{\mu}_0, \tilde{\mu}_1, \dots, \tilde{\mu}_{m-1}$ are related to the eigenvalues of $A$, $\mu_0, \mu_1, \dots, \mu_{m-1}$, by 
+Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ approximate the eigenfunctions of the Laplace-Beltrami operator (with Neumann boundary conditions) on $\mathcal{M}_d$,
+and the eigenvalues $\mu_0, \mu_1, \dots, \mu_{m-1}$ of $A$ are related to the eigenvalues of the operator by 
 \begin{equation} \label{eq:evals_relationship}
@@
 %
-We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$, and if the data are low-dimensional, we only need to retain a few of the leading eigenvectors to adequately describe the data.
+where $t$ is ....
+%
+As discussed previously, the eigenfunctions provide a parameterization of the manifold, such that $\phi_{j}(i)$ yields the $j^{th}$ embedding coordinate for $z(i)$.
+%
+We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$, and if the data do lie on a low-dimensional manifold, we only need to retain a few of the leading eigenvectors to adequately describe the data.
 %
@@
 
-However, the data need not be sampeld uniformly from the underlying manifold.
+However, the data need not be sampled uniformly from the underlying manifold.
 %
-If the data $z(1), \dots, z(m)$ are sampled from $\mathcal{M}_d$ with some density $U$, then the matrix $I-A$ approximates the Fokker-Planck operator with potential $2U$, e.g., $(I-A) \phi \rightarrow \nabla^2 \phi - 2\nabla\phi \cdot \nabla U$.
+If the data $z(1), \dots, z(m)$ are sampled from $\mathcal{M}_d$ with some density (TODO: check density vs.
+potential) $U$, then the matrix $I-A$ approximates the Fokker-Planck operator with potential $2U$, e.g., $(I-A) \phi \rightarrow \nabla^2 \phi - 2\nabla\phi \cdot \nabla U$.
+%
+If the sampling density of the data on the manifold is not meaningful, it can also be factored out via the appropriate normalization of the matrix $W$. TODO: add details?
 
 \begin{figure}
-\includegraphics[width=0.5\textwidth]{strip_discrete1}
-\includegraphics[width=0.5\textwidth]{strip_discrete2}
-\includegraphics[width=0.5\textwidth]{strip_nonuniform1}
-\includegraphics[width=0.5\textwidth]{strip_nonuniform2}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{strip_discrete1}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{strip_discrete2}
+\end{subfigure}
 
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{strip_nonuniform1}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{strip_nonuniform2}
+\end{subfigure}
+\caption{(a, b) Two-dimensional strip with uniform sampling colored by the (a) first, and (b) fourth (non-trivial) eigenvector from diffusion maps.
+(c, d)Two-dimensional strip with data sampled from a Gaussian distribution in the x-direction and uniform sampling in the y-direction, colored by the (c) first, and (d) third (non-trivial) eigenvector from diffusion maps.
+Note that in both cases we uncover parameterizations which are one-to-one with the x- and y-directions.
+}
+\label{fig:strip_evecs}
 \end{figure}
@@
 
-In diffusion maps, we order the eigenvectors by the magnitude of the corresponding eigenvalues, in the hopes that the top eigenvectors provide us with a parameterization of the underlying manifold.
+In diffusion maps, we order the eigenvectors by the magnitude of the corresponding eigenvalues, in the hopes that the leading eigenvectors provide us with a parameterization of the underlying manifold.
 %
@@
 %
-We recall that the eigenfunctions $\phi_{1,0} = \cos \left( \frac {\pi x}{L_x} \right)$ and  $\phi_{0,1} = \cos \left( \frac {\pi y}{L_y} \right)$ provide embedding coordinates for the manifold $|mathcal{M}_d$.
+We recall that the eigenfunctions $\phi_{1,0} = \cos \left( \frac {\pi x}{L_x} \right)$ and  $\phi_{0,1} = \cos \left( \frac {\pi y}{L_y} \right)$ provide embedding coordinates for the manifold $\mathcal{M}_d$.
 %
@@
 %
-Looking at the diffusion maps eigenvalues which correspond to modes which are {\em not} higher harmonics will then provide us with a measure of the relative lengths of the data set along the principal directions.
+Looking at the diffusion maps eigenvectors which correspond to modes which are {\em not} higher harmonics will yield the most parsimonious representation of the data.
+%
+Furthermore, the corresponding eigenvalues provide us with a measure of the relative lengths of the data set along the principal directions.
 
@@
 %
-The solution of \eqref{eq:opt_problem} is given by the solution of an eigenvector problem \cite{...}.
+The solution of \eqref{eq:opt_problem} is given by the solution of an eigenvector problem \cite{...} (TODO: add details in appendix/at end?).
 
@@
 
-The issues discussed above must all be considered and addressed in order to successfully analyze data using manifold learning algorithms. 
+We will now demonstrate our methodology on simulation data from a model problem arising in the study of cellular dynamics. 
 %
-To illustrate this point, we will use a model problem arising in the study of cellular dynamics. 
-%
-%In this example, constructing appropriate observers with an informative metric will allow us to perform informative analysis. 
-%
-Automatically detecting the ``important'' components will allow us to deduce when the dimension of the underlying dynamical system changes. 
-%
-In this way, we can deduce how many variables are required to write an accurate reduced model.
-
-Often, systems which are stochastic and noisy on the microscale exhibit coherent dynamics on the macroscale which are governed by only a few parameters.
-%
-For example, fluid flow, which can be described via the position and velocity of every fluid particle, is modeled at the macroscale using the Navier-Stokes equations.
+Although the system consists of many stochastic and noisy particles on the microscale, it exhibits coherent dynamics on the macroscale which are governed by only a few parameters.
 %
@@
 %
-We would like to extract a parameterization of data collected at the microscale which is consistent with the macroscopic behavior, without any {\em a priori} knowledge of the appropriate microscopic or macroscopic model.
+We will show how we can use our methodology to extract a parameterization of data collected at the microscale which is consistent with the macroscopic behavior, without any {\em a priori} knowledge of the appropriate microscopic or macroscopic model.
 %
 Furthermore, determining the dimensionality of such a parameterization can inform us to the requisite dimensionality of a reduced model which captures the relevant macroscopic behavior.
+
+%For example, fluid flow, which can be described via the position and velocity of every fluid particle, is modeled at the macroscale using the Navier-Stokes equations.
+%
 
@@
 
-We will illustrate these points using a model problem which arises in studies of cellular chemotaxis \cite{othmer2000diffusion}, where 
+Our model problem arises in studies of cellular chemotaxis \cite{othmer2000diffusion}, where 
 biological cells work together and exhibit coherent macroscopic dynamics regulated by extracellular sensed signals, in order to accomplish tasks such as finding food or navigating away from toxins.
@@
 %
-For example, when looking at a protein, the ``observers'' may be the locations of the $\alpha$-carbon atoms. 
+For example, when looking at a protein, the observers may be the locations of the $\alpha$-carbon atoms, as these backbone atoms are known to describe most of the relevant dynamics within the protein.
 %
-Observers can also factor out other sources of variation. 
-%
-For example, the bispectrum \cite{...} and the scattering transform \cite{...} are both observers which are invariant to rotations. 
+Observers can also remove unimportant sources of variation; 
+for example, the bispectrum \cite{...} and the scattering transform \cite{...} are both invariant to rotations. 
 %
@@
 %
-For molecular simulation data, it is often appropriate to use the root-mean-squared displacement {\em after} aligning configurations with respect to rotations, as global rotations of the entire molecule are not meaningful.
+For molecular simulation data, it is often appropriate to use the Euclidean distance {\em after} aligning configurations with respect to rotations, as global rotations of the entire molecule are not meaningful.
 %
@@
 %
-Another option is the Earth Mover's Distance, a multi scale distance metric which is stable and robust to noise.
+Another option is the Earth Mover's Distance, a multiscale distance metric which is stable and robust to noise.
 
@@
 
+\subsubsection{Using the appropriate metric}
+
 Figure~\ref{fig:dmaps_embed_emd} shows the results of analyzing two sets of chemotaxis simulations using diffusion maps. 
@@
 
+\subsubsection{Detecting changes in dynamical behavior}
+
 We have seen that the eigenvalues $\mu_i$, which quantify the importance of each embedding coordinate, reveal the relative importance of the variables which govern the macroscopic dynamics.
@@
 %
-We showed that at the two asymptotic regimes ($\lambda \rightarrow 0$ and $\lambda \rightarrow \infty$), there is a gap between the two eigenvalues, as one of the two macroscopic variables is significantly more important to the dynamical behavior.
+We expect a gap between the two eigenvalues at the two asymptotic regimes ($\lambda \rightarrow 0$ and $\lambda \rightarrow \infty$), as one of the two macroscopic variables is significantly more important to the dynamical behavior.
 %
@@
 
-\subsection{Detect dimensionality}
+\subsection{Detecting changes in dimensionality}
 
@@
 %
-We plot  
+From the relationship in \ref{...}, we plot  
 \begin{equation}

### 2015-02-23 chemotaxis/paper/draft1.tex
updated chemotaxis paper

@@
 
-In recent years, it has been noted that eigenfunctions of the Laplace-Beltrami operator with Neumann boundary conditions on the manifold \cite{....} can parameterize the manifold. 
-%
-To provide intuition as to why this is the case, consider a two-dimensional strip with edge lengths $L_x$ and $L_y$. 
+In recent years, it has been noted that eigenfunctions of the Laplace-Beltrami operator with Neumann boundary conditions can parameterize the manifold \cite{Belkin2003, coifman2005geometric, singer2008non}.
+%
+To provide intuition as to why this is the case, consider a two-dimensional strip with edge lengths $L_1$ and $L_2$. 
 %
@@
 \begin{equation} \label{eq:evals}
-\tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_x} \right)^2 + \left( \frac{k_2 \pi}{L_y} \right)^2
+\tilde{\mu}_{k_1, k_2} = \left( \frac{k_1 \pi}{L_1} \right)^2 + \left( \frac{k_2 \pi}{L_2} \right)^2
 \end{equation}
@@
 \begin{equation} \label{eq:efuncs}
-\tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi x}{L_x} \right) \cos \left( \frac{k_2 \pi y}{L_y} \right)
-\end{equation}
-where $x$ and $y$ denote the two coordinates of the rectangular domain. 
-%
-We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi x}{L_x} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi y}{L_y} \right)$ are one-to-one with the $x$ and $y$ coordinates, respectively, and therefore yield a parameterization of the underlying manifold.
-%
-Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide a measure of $L_x$ versus $L_y$: as the ratio between $L_x$ and $L_y$ increases, the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases.
+\tilde{\phi}_{k_1, k_2} = \cos \left( \frac{k_1 \pi z_1}{L_1} \right) \cos \left( \frac{k_2 \pi z_2}{L_2} \right)
+\end{equation}
+where $z_1$ and $z_2$ denote the two coordinates of the rectangular domain. 
+%
+We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi z_1}{L_1} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi z_2}{L_2} \right)$ are one-to-one with the $z_1$ and $z_2$ coordinates, respectively, and therefore yield a parameterization of the underlying manifold (see Figure~\ref{fig:strip_efuncs}).
+%
+Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide a measure of $L_1$ versus $L_2$: as the ratio between $L_1$ and $L_2$ increases, the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases.
 
@@
 
-\begin{figure}
+\begin{figure}[t]
 \begin{subfigure}{0.5\textwidth}
@@
 \end{subfigure}
-\caption{Two-dimensional strip colored by the eigenfunction (a) $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi x}{L_x} \right)$, and (b) $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi y}{L_y} \right)$ .}
+\caption{Two-dimensional strip colored by the eigenfunction (a) $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi z_1}{L_1} \right)$, and (b) $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi z_2}{L_2} \right)$ .}
 \label{fig:strip_efuncs}
@@
 %
-Instead, we are given data {\em sampled} from the underlying manifold $\mathcal{M}_d$, and we would like to uncover a parameterization of the data.
+Instead, we are given data {\em sampled} from the underlying manifold, and we would like to uncover a parameterization of the data.
 %
@@
 %
-In the limit of infinite data uniformly sampled from $\mathcal{M}_d$ ($m \rightarrow \infty$ and $\epsilon \rightarrow 0$ with the appropriate rate), the matrix $I-A$ approaches the Laplace-Beltrami operator on the manifold $\mathcal{M}_d$ \cite{...}.
+In the limit of infinite data uniformly sampled from $\mathcal{M}_d$ ($m \rightarrow \infty$ and $\epsilon \rightarrow 0$ with the appropriate rate), the matrix $A-I$ approaches the Laplace-Beltrami operator on the manifold $\mathcal{M}_d$ \cite{coifman2006geometric}.
 %
 Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ approximate the eigenfunctions of the Laplace-Beltrami operator (with Neumann boundary conditions) on $\mathcal{M}_d$,
-and the eigenvalues $\mu_0, \mu_1, \dots, \mu_{m-1}$ of $A$ are related to the eigenvalues of the operator by 
+and the eigenvalues $\mu_0, \mu_1, \dots, \mu_{m-1}$ of $A$ are related to the eigenvalues of the continuous operator by 
 \begin{equation} \label{eq:evals_relationship}
-\mu_k = e^{-\tilde{\mu}_k t}.
-\end{equation}
-%
-where $t$ is ....
+\mu_k = \exp \left( -\frac{\epsilon^2}{4} \tilde{\mu}_{k_1, k_2}  \right).
+\end{equation}
 %
@@
 Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a trivial constant vector which is not used as an embedding coordinate.
+TODO: check sign , I-A or A-I??
 
@@
 %
-If the data $z(1), \dots, z(m)$ are sampled from $\mathcal{M}_d$ with some density (TODO: check density vs.
-potential) $U$, then the matrix $I-A$ approximates the Fokker-Planck operator with potential $2U$, e.g., $(I-A) \phi \rightarrow \nabla^2 \phi - 2\nabla\phi \cdot \nabla U$.
-%
-If the sampling density of the data on the manifold is not meaningful, it can also be factored out via the appropriate normalization of the matrix $W$. TODO: add details?
-
-\begin{figure}
+If the data $z(1), \dots, z(m)$ are sampled from $\mathcal{M}_d$ with some density $p$, then the matrix $A-I$ approximates the Fokker-Planck operator, $(A-I) \phi \rightarrow \nabla^2 \phi - \nabla U \cdot \nabla \phi$, where $U = -2 \log p$.
+%
+Alternatively, one can factor out the density effects in the diffusion maps matrix.
+%
+Given data $z(1), \dots, z(m) \in \mathcal{M}_d$, we construct the matrix
+%
+\begin{equation}
+\tilde{W} = D^{-1} W D^{-1}.
+\end{equation}
+%
+Pre- and post-multiplication by $D^{-1}$ removes any density effects from $W$. 
+%
+We then construct the diagonal matrix $\tilde{D} \in \mathbb{R}^{m \times m}$, with $\tilde{D}_{ii} = \sum_j \tilde{W}_{ij}$, and the matrix $\tilde{A}  = \tilde{D}^{-1} \tilde{W}.$
+%
+The matrix $\tilde{A}$ is comparable to the matrix $A$, but with any sampling density effects removed, such that $A-I$ approaches the Laplace-Beltrami operator even if the data is nonuniformly sampled \cite{coifman2005geometric}.
+
+Figure~\ref{fig:strip_evecs} shows data sampled from a strip, colored by diffusion maps eigenvectors. 
+%
+In both the case of uniform and nonuniform sampling, the eigenvectors are one-to-one with $z_1$ and $z_2$, and thus parameterize the manifold. 
+%
+Although we have considered a very simple example for illustrative purposes, common practice is to use these tools for more complex, nonlinear data sets.
+
+\begin{figure}[t]
 \begin{subfigure}{0.5\textwidth}
@@
 \begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{strip_discrete2}
+\includegraphics[width=\textwidth]{strip_discrete4}
 \end{subfigure}
@@
 \caption{(a, b) Two-dimensional strip with uniform sampling colored by the (a) first, and (b) fourth (non-trivial) eigenvector from diffusion maps.
-(c, d)Two-dimensional strip with data sampled from a Gaussian distribution in the x-direction and uniform sampling in the y-direction, colored by the (c) first, and (d) third (non-trivial) eigenvector from diffusion maps.
-Note that in both cases we uncover parameterizations which are one-to-one with the x- and y-directions.
+(c, d)Two-dimensional strip with data sampled from a Gaussian distribution in $z_1$ and sampled uniformly in $z_2$, colored by the (c) first, and (d) third (non-trivial) eigenvector from diffusion maps.
+Note that in both cases we uncover parameterizations which are one-to-one with $z_1$ and $z_2$.
 }
@@
 
-TODO: add two sets of figures: one with data uniformly sampled, one with data with Gaussian distribiution
-
-Common practice is to use these tools for nonlinear data where these things are coupled. Show swiss roll and torus.
-
 \subsection{Identifying the Informative Eigenvectors using Local Linear Regression}
@@
 %
-Again, for illustrative purposes, we consider the case where $\mathcal{M}_d$ is a $2$-dimensional rectangular domain with edge lengths $L_x  > L_y$. 
-%
-We recall that the eigenfunctions $\phi_{1,0} = \cos \left( \frac {\pi x}{L_x} \right)$ and  $\phi_{0,1} = \cos \left( \frac {\pi y}{L_y} \right)$ provide embedding coordinates for the manifold $\mathcal{M}_d$.
+Again, for illustrative purposes, we consider the case where $\mathcal{M}_d$ is a $2$-dimensional rectangular domain with edge lengths $L_1  > L_2$. 
+%
+We recall that the eigenfunctions $\tilde{\phi}_{1,0} = \cos \left( \frac {\pi z_1}{L_1} \right)$ and  $\tilde{\phi}_{0,1} = \cos \left( \frac {\pi z_2}{L_2} \right)$ provide embedding coordinates for the manifold $\mathcal{M}_d$.
 %
@@
 %
-In fact, if $L_x > 2 L_y$, then $\tilde{\mu}_{2, 0} < \tilde{\mu}_{0, 1}$, and so the second (non-trivial) eigenvector (when the eigenvectors are ordered by their corresponding eigenvalues) will be a harmonic of the first and still parameterize the $x$ direction.
+In fact, if $L_1 > 2 L_2$, then $\tilde{\mu}_{2, 0} < \tilde{\mu}_{0, 1}$, and so the second (non-trivial) eigenvector (when the eigenvectors are ordered by their corresponding eigenvalues) will be a harmonic of the first and still parameterize $z_1$ (see Figure~\ref{fig:strip_harmonics}).
 %
@@
 
-Given the eigenvectors $\phi_1, \phi_2, \dots, \phi_{n-1} \in \mathbb{R}^n$, we would like to automatically deduce which ones are harmonics/functions of another. 
+\begin{figure}[t]
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{strip_discrete1}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{strip_discrete2}
+\end{subfigure}
+
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{strip_discrete3}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{strip_discrete4}
+\end{subfigure}
+\caption{ Two-dimensional strip with uniform sampling colored by the first four (non-trivial) eigenvectors from diffusion maps.
+Note that the first and fourth eigenvectors are one-to-one with $z_1$ and $z_2$, respectively.
+However, the second and third eigenvectors are higher harmonics of the first eigenvector and do not capture any additional structure within the data set.
+}
+\label{fig:strip_harmonics}
+\end{figure}
+
+Given the eigenvectors $\phi_1, \phi_2, \dots, \phi_{n-1} \in \mathbb{R}^m$, we would like to automatically deduce which ones are harmonics/functions of another. 
 %
@@
 %
-The coefficients $\alpha$ and $\beta$ are functions of $\phi_{k-1}$ because we use a {\em local} linear fit, and so the coefficients change as a function of the domain.
+The coefficients $\alpha$ and $\beta$ are functions of $\Phi_{k-1}$ because we use a {\em local} linear fit, and so the coefficients change as a function of the domain.
 
@@
 \begin{equation} \label{eq:opt_problem}
-\hat{\alpha}_k (i) , \hat{\beta}_k(i)  = \argmin_{\alpha, \beta} \sum_{j \ne i} w_i(\Phi_{k-1}(i), \Phi_{k-1}(j)) \left( \phi_{k}(j) - (\alpha + \beta^T \Phi_{k-1}(j)) \right)^2.
-\end{equation}
-%
-where $w$ is a weighting function.
-%
-We will use Gaussian weights, 
-%
-\begin{equation}
-w_i(\Phi_{k-1}(i), \Phi_{k-1}(j))  = \exp \left( - \frac{\|\Phi_{k-1}(i) - \Phi_{k-1} (j) \|^2}{\epsilon^2} \right),
+\hat{\alpha}_k (i) , \hat{\beta}_k(i)  = \argmin_{\alpha, \beta} \sum_{j \ne i} K(\Phi_{k-1}(i), \Phi_{k-1}(j)) \left( \phi_{k}(j) - (\alpha + \beta^T \Phi_{k-1}(j)) \right)^2.
+\end{equation}
+%
+where $K$ is a kernel weighting function.
+%
+We will use a Gaussian kernel, 
+%
+\begin{equation}
+K(\Phi_{k-1}(i), \Phi_{k-1}(j))  = \exp \left( - \frac{\|\Phi_{k-1}(i) - \Phi_{k-1} (j) \|^2}{\epsilon^2} \right),
 \end{equation}
@@
 %
-The solution of \eqref{eq:opt_problem} is given by the solution of an eigenvector problem \cite{...} (TODO: add details in appendix/at end?).
-
 We then define the normalized leave-one-out cross-validation error for this local linear fit as
@@
 %
-%We will use a local linear function 
-%\begin{equation}
-%f_{k}(\phi_1, \dots, \phi_{k-1}) = a_0 + a_1 \phi_1 + \dots + a_{k-1} \phi_{k-1}
-%\end{equation}
-%where the coefficients $a_0, \dots, a_{k-1}$ are functions of $\phi_1, \dots, \phi_{k-1}$. 
-%
-%To find these coefficients, we want to minimize the local weighted sum of squares
-%\begin{equation} \label{eq:opt_problem}
-%\sum_{i=1}^n w_i(x) \left( \phi_{k}(x) - f_{k} (\phi_1(x), \dots, \phi_{k-1}(x)) \right)^2
-%\end{equation}
-%where $w$ is a weighting function.
-%%
-%We will use Gaussian weights, 
-%%
-%\begin{equation}
-%w_i(x) = \exp \left( - \frac{\|\Phi_{k-1}(x) - \Phi_{k-1} (x_i) \|^2}{\epsilon^2} \right), \text{where } \Phi_{k-1}(x) = \begin{bmatrix}
-%\phi_1(x) \\
-%\vdots \\
-%\phi_{k-1}(x)
-%\end{bmatrix} .
-%\end{equation}
-%%
-%We take $\epsilon = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k-1}(x)$.
-%
-%Let $\hat{f}_{k}(\phi_1(x_i), \dots, \phi_{k-1}(x_i))$ denote the minimum achieved in \eqref{eq:opt_problem}, when the minimization is performed over $x_j$ with $j \ne i$. 
-%
-%We then define the normalized leave-one-out cross-validation error as
-%\begin{equation} \label{eq:cv_error}
-%r_{k} = \sum_{i=1}^n \left( \phi_{k} (x_i) - \hat{f}_{k}(\phi_1(x_i), \dots, \phi_{k-1}(x_i)) \right)^2
-%\end{equation}
-%\begin{equation} \label{eq:cv_error}
-%r_{k} = \sqrt{ \frac{\sum_{i=1}^n \left( \phi_{k} (x_i) - \hat{f}_{k}(\Phi_{k-1}(x_i)) \right)^2} {\sum_{i=1}^n  \left( \phi_{k} (x_i) \right) }}
-%\end{equation}
-%
 Note that a small value of $r_k$ implies that $\phi_{k}$ can be accurately approximated from $\phi_1, \dots, \phi_{k-1}$, and is therefore a harmonic of a previous mode.
@@
 
+\eqref{eq:cv_error} can easily be computed \cite{wasserman2006all}.
+%
+We begin by computing the matrix
+\begin{equation}
+X_{k-1}(i) = \begin{bmatrix}
+1 & \phi_1(1) - \phi_1(i) & \dots & \phi_{k-1}(1)- \phi_{k-1}(i) \\
+1 & \phi_1(2) - \phi_1(i) & \dots & \phi_{k-1}(2)- \phi_{k-1}(i) \\
+\vdots & \vdots & \ddots & \vdots \\
+1 & \phi_1(m) - \phi_1(i) & \dots & \phi_{k-1}(m)- \phi_{k-1}(i) 
+\end{bmatrix}
+\end{equation}
+%
+and the matrix $W_{k-1}(i) = diag \left( K(\Phi_{k-1}(i), \Phi_{k-1}(1)), \dots, K(\Phi_{k-1}(i), \Phi_{k-1}(m)) \right)$.
+%
+We then calculate the smoothing matrix $L_{k-1} \in \mathbb{R}^{m \times m}$, where 
+\begin{equation}
+L_{k-1} =
+\begin{bmatrix}
+e_1^T \left( X_{k-1}^T(1) W_{k-1}(1) X_{k-1}(1) \right) ^{-1} X_{k-1}^T(1) W_{k-1}(1) \\
+e_1^T \left( X_{k-1}^T(2) W_{k-1}(2) X_{k-1}(2) \right) ^{-1} X_{k-1}^T(2) W_{k-1}(2) \\
+\vdots \\
+e_1^T \left( X_{k-1}^T(m) W_{k-1}(m) X_{k-1}(m) \right) ^{-1} X_{k-1}^T(m) W_{k-1}(m) \\
+\end{bmatrix},
+\end{equation}
+%
+where $e_1^T = (1, 0, \dots, 0)$. 
+%
+We can then write
+%
+\begin{equation} 
+r_{k} = \sqrt{ \frac{\sum_{i=1}^n \left( \frac{ \phi_{k} (i) - \sum_{j=1}^m L_{k-1}(i,j) \phi_{k}(j) }{1-L_{k-1}(i,i)} \right)^2} {\sum_{i=1}^n  \left( \phi_{k} (i) \right)^2 }}
+\end{equation}
+
 \subsection{Illustrative Examples}
@@
 %
-Each data set consists of $m=1500$ points uniformly distributed on the swiss roll. 
-%
-Both swiss rolls are the same in the angular/spiral direction.
-%
-However, height of the first swiss roll is $h = 15 \pi$, while the heigh of the second is $h = 6 \pi$. 
-%
-These two data sets are shown in Figure~\ref{fig:swiss_rolls}(a)~and~(b), respectively.
+The data are
+\begin{equation}
+\begin{aligned}
+z_1 =& (4 \pi \sqrt{t_1} + 1) \cos(4 \pi \sqrt{t_1}) \\
+z_2 =& (4 \pi \sqrt{t_1} + 1) \sin(4 \pi \sqrt{t_1}) \\
+z_3 =& h t_2
+\end{aligned}
+\end{equation}
+%
+where $t_1, t_2 \sim Unif(0,1)$.
+%
+The height of the first swiss roll is $h = 15 \pi$, while the height of the second is $h = 6 \pi$. 
+%
+Each data set consists of $m=1500$ points, shown shown in Figure~\ref{fig:swiss_rolls}(a)~and~(b).
 %
@@
 
-\begin{figure}
+\begin{figure}[t]
 %
@@
 \centering
-\includegraphics[height=1.5in]{swissroll1}
+\includegraphics[height=1.25in]{swissroll1}
 \caption{}
@@
 \centering
-\includegraphics[height=1.5in]{swissroll2}
+\includegraphics[height=1.25in]{swissroll2}
 \caption{}
@@
 \centering
-\includegraphics[height=1.5in]{swissroll1_evals}
+\includegraphics[height=1.25in]{swissroll1_evals}
 \caption{}
@@
 \centering
-\includegraphics[height=1.5in]{swissroll2_evals}
+\includegraphics[height=1.25in]{swissroll2_evals}
 \caption{}
@@
 \centering
-\includegraphics[height=1.5in]{swissroll1_color1} 
+\includegraphics[height=1.25in]{swissroll1_color1} 
 \caption{}
@@
 \centering
-\includegraphics[height=1.5in]{swissroll2_color1}
+\includegraphics[height=1.25in]{swissroll2_color1}
 \caption{}
@@
 \centering
-\includegraphics[height=1.5in]{swissroll1_color2}
+\includegraphics[height=1.25in]{swissroll1_color2}
 \caption{}
@@
 \centering
-\includegraphics[height=1.5in]{swissroll2_color2}
+\includegraphics[height=1.25in]{swissroll2_color2}
 \caption{}
@@
 \begin{aligned}
-x =& (r_1 + r_2 \cos \theta_2 ) \cos \theta_1 \\
-y =& (r_1 + r_2 \cos \theta_2 ) \sin \theta_1 \\
-z =& r_2 \sin \theta_2
+z_1 =& (r_1 + r_2 \cos \theta_2 ) \cos \theta_1 \\
+z_2 =& (r_1 + r_2 \cos \theta_2 ) \sin \theta_1 \\
+z_3 =& r_2 \sin \theta_2
 \end{aligned}
@@
 %
-Although the system consists of many stochastic and noisy particles on the microscale, it exhibits coherent dynamics on the macroscale which are governed by only a few parameters.
-%
-However, in general, this mapping from microscale to macroscale is not always obvious.
+Although on the microscale the system consists of many stochastic and noisy particles, it exhibits coherent dynamics on the macroscale which are governed by only a few parameters.
+%
+In general, this mapping from microscale to macroscale is not always obvious.
 %
@@
 %
+We therefore expect the observed dimensionality to change from two to one as a function of $\lambda$. 
+%
 For general dynamical systems, the governing macroscopic variables and asymptotic behavior may not be immediately obvious, and instead need to be inferred from the microscopic data.
@@
 Observers can also remove unimportant sources of variation; 
-for example, the bispectrum \cite{...} and the scattering transform \cite{...} are both invariant to rotations. 
+for example, the bispectrum \cite{zhao2014rotationally} and the scattering transform \cite{mallat2012group} are both invariant to rotations. 
 %
@@
 %
-The scattering transform \cite{...} is robust to small deformations, and histograms serve to linearize noise \cite{...}.
+The scattering transform \cite{mallat2012group} is robust to small deformations, and histograms serve to linearize noise \cite{talmon2013empirical}.
 
@@
 %
-For dynamical systems which are obscured by large variance noise, the Mahalanobis distance often yields a more informative parameterization of the data \cite{...}.
+For dynamical systems which are obscured by large variance noise, the Mahalanobis distance often yields a more informative parameterization of the data \cite{talmon2013empirical}.
 %
@@
 %
-From the relationship in \ref{...}, we plot  
-\begin{equation}
- \frac{\log \mu_{i_1}}{\log \mu_{i_2}} ;
+According to the relationship in \eqref{eq:evals} and \eqref{eq:evals_relationship}, we plot  
+\begin{equation}
+ \frac{\log \mu_{i_2}}{\log \mu_{i_1}} ;
 \end{equation}
@@
 %
-From \cite{...}, the relevant time scales for the velocity jump process are
+From \cite{othmer2000diffusion}, the relevant time scales for the velocity jump process are
+%\begin{equation}
+%\begin{aligned}
+%\tau_{drift} =& \frac{L}{s} \\
+%\tau_{diff} =& \frac{L^2 \lambda}{s^2}
+%\end{aligned}
+%\end{equation}
+%%
+%Here, we take $L = t_{max} s$ as a characteristic length scale for the simulations. 
+%
 \begin{equation}
 \begin{aligned}
-\tau_{drift} =& \frac{L}{s} \\
-\tau_{diff} =& \frac{L^2 \lambda}{s^2}
+\tau_{drift} =& t_{max} \\
+\tau_{diff} =& t_{max}^2 \lambda
 \end{aligned}
 \end{equation}
-%
-Here, we take $L = t_{max} s$ as a characteristic length scale for the simulations. 
 %

### 2015-02-23 chemotaxis/test_torus.m

@@ -36,9 +36,9 @@ load('torus1.mat')
 make_fig(3,2);
 plot3(x,y,z,'.', 'markersize', 8)
 % view(-20, 70)
-xlabel('x')
-ylabel('y')
-zlabel('z')
+xlabel('z_1')
+ylabel('z_2')
+zlabel('z_3')
 axis equal
 grid on
 print('torus1.eps','-depsc')
@@ -99,9 +99,9 @@ load('torus2.mat')
 make_fig(3,2);
 plot3(x,y,z,'.', 'markersize', 8)
 % view(-20, 75)
-xlabel('x')
-ylabel('y')
-zlabel('z')
+xlabel('z_1')
+ylabel('z_2')
+zlabel('z_3')
 axis equal
 grid on
 print('torus2.eps','-depsc')
@@ -163,9 +163,9 @@ load('torus3.mat')
 make_fig(3,2);
 plot3(x,y,z,'.', 'markersize', 8)
 % view(-20, 70)
-xlabel('x')
-ylabel('y')
-zlabel('z')
+xlabel('z_1')
+ylabel('z_2')
+zlabel('z_3')
 axis equal
 grid on
 print('torus3.eps','-depsc')

### 2015-02-24 chemotaxis/paper/draft1.tex
update dpaper

@@
 \begin{aligned}
-z_1 =& (4 \pi \sqrt{t_1} + 1) \cos(4 \pi \sqrt{t_1}) \\
-z_2 =& (4 \pi \sqrt{t_1} + 1) \sin(4 \pi \sqrt{t_1}) \\
-z_3 =& h t_2
+z_1 =& \theta \cos \theta \\
+z_2 =& \theta \sin \theta \\
+z_3 =& h t
 \end{aligned}
@@
 %
-where $t_1, t_2 \sim Unif(0,1)$.
-%
-The height of the first swiss roll is $h = 15 \pi$, while the height of the second is $h = 6 \pi$. 
+where $\theta$ and $t$ are random variables, sampled such that the data is uniformly distributed along the swiss roll. 
+%
+The height of the first swiss roll is $h = 40$, while the height of the second is $h = 20$. 
 %
@@
 \caption{Swiss roll example.
-(a) Data set 1; $h=6 \pi$.
-(b) Data set 2; $h = 15 \pi$.
+(a) Data set 1; $h= 40$.
+(b) Data set 2; $h = 20$.
 (c) Diffusion maps eigenvalue spectrum for data set 1.
@@
 
-\subsection{Detecting changes in dimensionality}
+\subsubsection{Detecting changes in dimensionality}
 
@@
 \begin{equation}
- \frac{\log \mu_{i_2}}{\log \mu_{i_1}} ;
+ \frac{\log \mu_{i_1}}{\log \mu_{i_2}} ;
 \end{equation}
@@
 %
-This is plotted in Figure~\ref{fig:chemotaxis_compare_timescales_evals} as a function of $\lambda$.
-
+Figure~\ref{fig:chemotaxis_simulations_harmonics} shows the eigenvalue spectra for three different sets of simulations. 
+%
+It now becomes clear that looking at the diffusion maps coordinates {\em modulo} higher harmonics is essential for extracting an informative parameterization of the data, as well as characterizing the dimensionality of the data.
+%
+From Figures~\ref{fig:chemotaxis_simulations_harmonics}(b-c), we can see that the eigenvalue spectrum does not exhibit any large spectral gaps.
+%
+However, there is a significant gap between the eigenvalues which correspond to new directions in the data, and the corresponding eigenvectors provide us with a two-dimensional parameterization of the data which is consistent with the expected macroscopic dynamics.
+
+Figure~\ref{fig:chemotaxis_compare_timescales_evals} shows the diffusion maps eigenvalue ratio as a function of $\lambda$. 
+%
 For this specific example, we can compare the empirically detected change in dimensionality with the analytically-predicted shift. 
@@
 %
+\label{fig:chemotaxis_simulations_harmonics}
 \end{figure}

### 2015-02-24 chemotaxis/local_linear_regression.m

@@ -9,9 +9,10 @@ W = exp(-K.^2 / eps^2);
 L = zeros(n);
 for i=1:n
     Xx = [ones(size(X,1),1) X-repmat(X(i,:), n, 1)];
-    Wx = diag(W(i,:));
-%     A = inv(Xx'*Wx*Xx)*Xx'*Wx;
-    A = (Xx'*Wx*Xx)\(Xx'*Wx);
+%     Wx = diag(W(i,:));
+%     A = (Xx'*Wx*Xx)\(Xx'*Wx);
+    Xx2 = Xx'.*repmat(W(i,:), size(Xx, 2), 1);
+    A = (Xx2*Xx)\Xx2;
     L(i,:) = A(1,:);
 end
 

### 2015-02-25 chemotaxis/paper/draft1.tex
small edits to paper
updated figures for chemotaxis

@@
 
-\section{Background on Manifold Learning and Some Relevant Issues}
+\section{Manifold Learning and Some Relevant Issues}
 
@@
 
-\subsection{Identifying the Informative Eigenvectors using Local Linear Regression}
+\begin{figure}[t]
+\centering
+\begin{subfigure}{0.3\textwidth}
+\includegraphics[height=3cm]{strip_data_L2}
+\includegraphics[height=3.5cm]{strip_spectrum_L2}
+\caption{}
+\end{subfigure}
+%
+\begin{subfigure}{0.3\textwidth}
+\includegraphics[height=3cm]{strip_data_L4}
+\includegraphics[height=3.5cm]{strip_spectrum_L4}
+\caption{}
+\end{subfigure}
+%
+\begin{subfigure}{0.35\textwidth}
+\includegraphics[height=3cm]{strip_data_L8}
+\includegraphics[height=3.5cm]{strip_spectrum_L8}
+\caption{}
+\end{subfigure}
+%
+\caption{Data sets (top) and eigenvalue spectra (bottom) for strips with (a) $L_1 = 2$, $L_2 = 1$, (b) $L_1 = 4$, $L_2 = 1$, (c) $L_1 = 8$, $L_2 = 1$.
+The empirical eigenvalues are plotted in blue, and the analytical eigenvalues are plotted in red.
+From the eigenvalues which are identified as important/non-harmonics, the estimated length ratio is (a) 2.0, (b) 4.4, (c) 8.8.
+}
+\label{fig:strip_compare_analytic}
+\end{figure}
+
+\section{Identifying the Informative Eigenvectors }
 
@@
 Furthermore, the corresponding eigenvalues provide us with a measure of the relative lengths of the data set along the principal directions.
+
+\subsection{Algorithm: Local Linear Regression}
 
@@
 
+\subsection{Calculating the relative lengths of the manifold}
+
+Once we have determined which eigenvectors parameterize independent directions along the manifold, we can use the corresponding eigenvalues to calculate the relative lengths along these directions. 
+%
+Let $\mu_{i_1}, \mu_{i_2}, \dots$ denote the eigenvalues which correspond to eigenvectors which are {\em not} higher harmonics. 
+%
+From \eqref{eq:evals} and \eqref{eq:evals_relationship}, we propose to use
+\begin{equation} \label{eq:est_lengths}
+L_j \propto \frac{1}{\sqrt{-\log \mu_{i_j}}}
+\end{equation}
+where $L_1, L_2, \dots$ are the relative lengths along the principal directions of the manifold. 
+%
+We can then use this measure to determine how many components are required to retain most of the information within the data set.
+
 \subsection{Illustrative Examples}
 
-We will demonstrate the local linear regression algorithm for identifying informative eigenvectors on two synthetic data sets. 
-%
-Both of these data sets are nonlinear manifolds which provide examples of the flexibility of diffusion maps. 
+We will demonstrate the local linear regression algorithm for identifying informative eigenvectors on three synthetic data sets. 
+%
+The first data set is the strip where the solution is known analytically, and the second and third data sets are nonlinear manifolds which provide examples of the flexibility of diffusion maps. 
 %
@@
 
+\subsubsection{Strip}
+
+Our first illustrative example consists of three different strip data sets. 
+%
+Each of these data sets contains $m=1500$ data points uniformly distributed along the strip. 
+%
+Figure~\ref{fig:strip_compare_analytic} shows the data sets and corresponding eigenspectra.
+%
+The diffusion maps eigenvalues are consistent with the known analytic eigenvalues of the Laplacian, and using \eqref{eq:est_lengths}, we can accurately estimate the relative lengths of the two principal directions in each data set.
+
 \subsubsection{Swiss Roll}
 
-Our first illustrative example will consist of two different swiss roll data sets.
+Our second illustrative example will consist of two different swiss roll data sets.
 %
@@
 %
-where $\theta$ and $t$ are random variables, sampled such that the data is uniformly distributed along the swiss roll. 
+where $\theta$ is uniformly sampled along the arclength of the spiral, and $t \sim Unif(0,1)$. 
 %
@@
 
-For the second example, we consider a torus defined by
+For the third example, we consider a torus defined by
 %
@@
 %
-\begin{subfigure}{\textwidth}
-\includegraphics[width=2.5in]{torus1}
-\includegraphics[height=1.5in]{torus1_evals}
-\caption{}
-\end{subfigure}
-%
-\begin{subfigure}{\textwidth}
-\includegraphics[width=2.5in]{torus2}
-\includegraphics[height=1.5in]{torus2_evals}\\
-\caption{}
-\end{subfigure}
-%
-\begin{subfigure}{\textwidth}
-\includegraphics[width=2.5in]{torus3}
-\includegraphics[height=1.5in]{torus3_evals}
+\begin{subfigure}{0.3\textwidth}
+\includegraphics[height=0.75in]{torus1}
+\includegraphics[height=1in]{torus1_evals}
+\caption{}
+\end{subfigure}
+%
+\begin{subfigure}{0.32\textwidth}
+\includegraphics[height=0.75in]{torus2}
+\includegraphics[height=1in]{torus2_evals}
+\caption{}
+\end{subfigure}
+%
+\begin{subfigure}{0.35\textwidth}
+\includegraphics[height=0.75in]{torus3}
+\includegraphics[height=1in]{torus3_evals}
 \caption{}
@@
 %
-According to the relationship in \eqref{eq:evals} and \eqref{eq:evals_relationship}, we plot  
-\begin{equation}
- \frac{\log \mu_{i_1}}{\log \mu_{i_2}} ;
+According to \eqref{eq:est_lengths}, we plot  
+\begin{equation}
+ \sqrt{\frac{\log \mu_{i_1}}{\log \mu_{i_2}}} ;
 \end{equation}

### 2015-02-25 chemotaxis/strip.m

@@ -18,7 +18,7 @@ shading interp
 axis equal
 xlabel('z_1')
 ylabel('z_2')
-cbar = colorbar('peer',gca); 
+cbar = colorbar('peer',gca);
 set(get(cbar,'xlabel'),'String','$\tilde{\phi}_{1,0}$', 'interpreter','latex');
 print('strip_cnts1.eps','-depsc')
 
@@ -29,7 +29,7 @@ shading interp
 axis equal
 xlabel('z_1')
 ylabel('z_2')
-cbar = colorbar('peer',gca); 
+cbar = colorbar('peer',gca);
 set(get(cbar,'xlabel'),'String','$\tilde{\phi}_{0,1}$', 'interpreter','latex');
 print('strip_cnts2.eps','-depsc')
 
@@ -57,14 +57,14 @@ if corr(V(:,5), data(:,2)) > 0
 end
 
 for i=2:5
-make_fig(3,2);
-scatter(data(:,1),data(:,2),50, V(:,i)/max(V(:,i)),'.')
-axis equal
-xlabel('z_1')
-ylabel('z_2')
-cbar = colorbar('peer',gca); 
-set(get(cbar,'xlabel'),'String',sprintf('\\phi_{%d}', i-1));
-print(sprintf('strip_discrete%d.eps', i-1),'-depsc')
+    make_fig(3,2);
+    scatter(data(:,1),data(:,2),50, V(:,i)/max(V(:,i)),'.')
+    axis equal
+    xlabel('z_1')
+    ylabel('z_2')
+    cbar = colorbar('peer',gca);
+    set(get(cbar,'xlabel'),'String',sprintf('\\phi_{%d}', i-1));
+    print(sprintf('strip_discrete%d.eps', i-1),'-depsc')
 end
 
 %%
@@ -93,7 +93,7 @@ scatter(data(:,1),data(:,2),50, V(:,2)/max(V(:,2)),'.')
 axis equal
 xlabel('z_1')
 ylabel('z_2')
-cbar = colorbar('peer',gca); 
+cbar = colorbar('peer',gca);
 set(get(cbar,'xlabel'),'String','\phi_{1}');
 print('strip_nonuniform1.eps','-depsc')
 
@@ -102,7 +102,64 @@ scatter(data(:,1),data(:,2),50, V(:,4)/max(V(:,4)),'.')
 axis equal
 xlabel('z_1')
 ylabel('z_2')
-cbar = colorbar('peer',gca); 
[…]

### 2015-02-25 chemotaxis/swissroll_uniform.m

@@ -1,0 +1,142 @@
+clear all
+close all
+
+%%
+
+N = 1500; % number of points considered
+
+% archemedian spiral
+
+a = 1;
+
+theta_vec = linspace(0, 4*pi, 100);
+s = 0.5*a*(theta_vec.*sqrt(1+theta_vec.^2)+log(theta_vec+sqrt(1+theta_vec.^2)));
+
+theta = interp1(s, theta_vec, rand(N, 1)*max(s));
+
+h = 40;
+
+neigs = 10;
+
+%%
+
+rng(321);
+
+z = h*rand(N,1); % random heights
+x = a * cos(theta) .* theta;
+y = a * sin(theta) .* theta;
+
+
+%%
+
+data = [x y z]; % data
+
+W = squareform(pdist(data)).^2;
+eps = 5;
+
+[V, D] = dmaps(W, eps, neigs);
+
+eps_med_scale = 3;
+res = compute_residuals_DMAPS(V, eps_med_scale);
+
+%%
+
+make_fig(2,2);
+plot3(x,y,z,'.', 'markersize', 10)
+view(-20, 80)
+xlabel('z_1')
+ylabel('z_2')
+zlabel('z_3')
+axis equal
+grid on
+print('swissroll1.eps','-depsc')
+
+make_fig(3,2);
+scatter3(x,y,z,50,sign(corr(V(:,2),theta))*V(:,2),'.')
+view(-20, 80)
+xlabel('z_1')
+ylabel('z_2')
+zlabel('z_3')
+axis equal
+grid on
+cbar = colorbar('peer',gca); 
+set(get(cbar,'xlabel'),'String','\phi_1');
+print('swissroll1_color1.eps','-depsc')
+
+make_fig(3,2);
+scatter3(x,y,z,50,sign(corr(V(:,3),z))*V(:,3),'.')
+view(-20, 80)
+xlabel('z_1')
+ylabel('z_2')
+zlabel('z_3')
+axis equal
+grid on
+cbar = colorbar('peer',gca); 
+set(get(cbar,'xlabel'),'String','\phi_2');
+print('swissroll1_color2.eps','-depsc')
+
+make_fig(3,2);
+make_colored_bars(diag(D(2:end,2:end)), res(2:end))
+set(gca, 'ylim', [0.9 1])
+xlabel('k')
+ylabel('\mu_k')
+axis square
+print('swissroll1_evals.eps','-depsc')
+
+%%
+
[…]

### 2015-02-26 chemotaxis/paper/draft1.tex
updated text for chemotaxis

@@
 
-\subsection{Manifold learning and the Laplace operator}
+\subsection{Eigenfunctions of the Laplace-Beltrami operator}
+
+\begin{figure}[t]
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{strip_cnts1}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{strip_cnts2}
+\end{subfigure}
+\caption{Two-dimensional strip colored by the eigenfunction (a) $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi z_1}{L_1} \right)$, and (b) $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi z_2}{L_2} \right)$ .}
+\label{fig:strip_efuncs}
+\end{figure}
 
@@
 %
-However, in the case when the manifold is curved, it is more difficult to define such a set of coordinates.
-
-In recent years, it has been noted that eigenfunctions of the Laplace-Beltrami operator with Neumann boundary conditions can parameterize the manifold \cite{Belkin2003, coifman2005geometric, singer2008non}.
+However, in the case when the manifold is curved, it is more difficult to define a set of coordinates.
+
+In recent years, it has been noted that eigenfunctions of the Laplace-Beltrami operator with Neumann boundary conditions can be used to parameterize the manifold \cite{Belkin2003, coifman2005geometric, singer2008non}.
 %
@@
 \end{equation}
-where $z_1$ and $z_2$ denote the two coordinates of the rectangular domain. 
+where $z_1$ and $z_2$ denote the two coordinates of the rectangular domain \cite{singer2008non}. 
 %
@@
 %
-Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide a measure of $L_1$ versus $L_2$: as the ratio between $L_1$ and $L_2$ increases, the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases.
-
-Although these eigenfunctions can only be written analytically for very special cases, it has been observed empirically that the eigenfunctions often yield appropriate/good coordinates to parameterize the underlying manifold.
+Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide a measure of $L_1$ versus $L_2$: as the ratio between $L_1$ and $L_2$ increases, the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases (this will be discussed further in Section~\ref{sec:relative_lengths}).
+%
+Although these eigenfunctions can only be written analytically for very special cases, it has been observed empirically that the eigenfunctions often provide appropriate coordinates to parameterize more complex, nonlinear manifolds.
+
+\subsection{Discrete approximation of the Laplace-Beltrami operator: diffusion maps}
 
@@
 \begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{strip_cnts1}
+\includegraphics[width=\textwidth]{strip_discrete1}
 \end{subfigure}
 \begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{strip_cnts2}
-\end{subfigure}
-\caption{Two-dimensional strip colored by the eigenfunction (a) $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi z_1}{L_1} \right)$, and (b) $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi z_2}{L_2} \right)$ .}
-\label{fig:strip_efuncs}
+\includegraphics[width=\textwidth]{strip_discrete4}
+\end{subfigure}
+
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{strip_nonuniform1}
+\end{subfigure}
+\begin{subfigure}{0.5\textwidth}
+\includegraphics[width=\textwidth]{strip_nonuniform2}
+\end{subfigure}
+\caption{(a, b) Two-dimensional strip with uniform sampling colored by the (a) first, and (b) fourth (non-trivial) eigenvector from diffusion maps.
+(c, d) Two-dimensional strip with data sampled from a Gaussian distribution in $z_1$ and sampled uniformly in $z_2$, colored by the (c) first, and (d) third (non-trivial) eigenvector from diffusion maps.
+Note that in both cases we uncover parameterizations which are one-to-one with $z_1$ and $z_2$.
+}
+\label{fig:strip_evecs}
 \end{figure}
-
-%From the eigenvalues, 
-%\begin{equation}
-%\sqrt{\frac{\log(\mu_{1,0})}{\log(\mu_{0,1})}} = 
-%\frac{L_y}{L_x}
-%\end{equation}
-%and so looking at the ratio of the log of the eigenvalues (modulo harmonics) should inform us as to the lengths of the different directions in the manifold.
-
-\subsection{Discrete approximation of the Laplacian: diffusion maps}
 
@@
 \begin{equation} \label{eq:W}
-W_{ij} = \exp \left( -\frac{\|z(i) - z(j) \|^2}{\epsilon^2} \right), \ i,j=1,\ldots,m
+W_{ij} = \exp \left( -\frac{\|z(i) - z(j) \|^2}{\epsilon^2} \right), \ i,j=1,\ldots,m,
 \end{equation}
@@
 %
-We typically take $\epsilon$ to be the median of the pairwise distances between the data points.
+$\epsilon$ can be chosen using several heuristics \cite{coifman2008graph, rohrdanz2011determination}; we often take $\epsilon$ to be the median of the pairwise distances between the data points.
 %
@@
 %
-In the limit of infinite data uniformly sampled from $\mathcal{M}_d$ ($m \rightarrow \infty$ and $\epsilon \rightarrow 0$ with the appropriate rate), the matrix $A-I$ approaches the Laplace-Beltrami operator on the manifold $\mathcal{M}_d$ \cite{coifman2006geometric}.
+In the limit of infinite data uniformly sampled from $\mathcal{M}_d$ ($m \rightarrow \infty$ and $\epsilon \rightarrow 0$ with the appropriate rate), the matrix $I-A$ approaches the Laplace-Beltrami operator on the manifold $\mathcal{M}_d$ \cite{coifman2006geometric}.
 %
@@
 Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a trivial constant vector which is not used as an embedding coordinate.
-TODO: check sign , I-A or A-I??
 
@@
 %
-If the data $z(1), \dots, z(m)$ are sampled from $\mathcal{M}_d$ with some density $p$, then the matrix $A-I$ approximates the Fokker-Planck operator, $(A-I) \phi \rightarrow \nabla^2 \phi - \nabla U \cdot \nabla \phi$, where $U = -2 \log p$.
+If the data $z(1), \dots, z(m)$ are sampled from $\mathcal{M}_d$ with some density $q$, then the matrix $I-A$ approximates the Fokker-Planck operator, 
+\begin{equation}
+(I-A) \phi \rightarrow \nabla^2 \phi - \nabla U \cdot \nabla \phi, 
+\end{equation}
+where $U = -2 \log q$. 
 %
@@
 %
-The matrix $\tilde{A}$ is comparable to the matrix $A$, but with any sampling density effects removed, such that $A-I$ approaches the Laplace-Beltrami operator even if the data is nonuniformly sampled \cite{coifman2005geometric}.
+The matrix $\tilde{A}$ is comparable to the matrix $A$, but with sampling density effects removed, such that $I-\tilde{A}$ approaches the Laplace-Beltrami operator even if the data is nonuniformly sampled from the underlying manifold \cite{coifman2005geometric}.
 
@@
 %
-In both the case of uniform and nonuniform sampling, the eigenvectors are one-to-one with $z_1$ and $z_2$, and thus parameterize the manifold. 
-%
-Although we have considered a very simple example for illustrative purposes, common practice is to use these tools for more complex, nonlinear data sets.
-
-\begin{figure}[t]
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{strip_discrete1}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{strip_discrete4}
-\end{subfigure}
-
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{strip_nonuniform1}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
-\includegraphics[width=\textwidth]{strip_nonuniform2}
-\end{subfigure}
-\caption{(a, b) Two-dimensional strip with uniform sampling colored by the (a) first, and (b) fourth (non-trivial) eigenvector from diffusion maps.
-(c, d)Two-dimensional strip with data sampled from a Gaussian distribution in $z_1$ and sampled uniformly in $z_2$, colored by the (c) first, and (d) third (non-trivial) eigenvector from diffusion maps.
-Note that in both cases we uncover parameterizations which are one-to-one with $z_1$ and $z_2$.
-}
-\label{fig:strip_evecs}
-\end{figure}
-
-\begin{figure}[t]
-\centering
-\begin{subfigure}{0.3\textwidth}
-\includegraphics[height=3cm]{strip_data_L2}
-\includegraphics[height=3.5cm]{strip_spectrum_L2}
-\caption{}
-\end{subfigure}
-%
-\begin{subfigure}{0.3\textwidth}
-\includegraphics[height=3cm]{strip_data_L4}
-\includegraphics[height=3.5cm]{strip_spectrum_L4}
-\caption{}
-\end{subfigure}
-%
-\begin{subfigure}{0.35\textwidth}
-\includegraphics[height=3cm]{strip_data_L8}
-\includegraphics[height=3.5cm]{strip_spectrum_L8}
-\caption{}
-\end{subfigure}
-%
-\caption{Data sets (top) and eigenvalue spectra (bottom) for strips with (a) $L_1 = 2$, $L_2 = 1$, (b) $L_1 = 4$, $L_2 = 1$, (c) $L_1 = 8$, $L_2 = 1$.
-The empirical eigenvalues are plotted in blue, and the analytical eigenvalues are plotted in red.
-From the eigenvalues which are identified as important/non-harmonics, the estimated length ratio is (a) 2.0, (b) 4.4, (c) 8.8.
-}
-\label{fig:strip_compare_analytic}
-\end{figure}
+In cases of both uniform and nonuniform sampling, the eigenvectors are one-to-one with $z_1$ and $z_2$, and thus parameterize the manifold. 
+%
+Although we have considered a very simple example for illustrative purposes, common practice is to use these tools for high-dimensional, nonlinear data sets.
 
@@
 %
-However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to correspond to the two smallest (non-trivial) eigenvalues (the largest eigenvalue will always be $\tilde{\mu}_{0,0} = 1$ and correspond to a constant eigenfunction; the eigenvalues $\tilde{\mu}_k$ are related to the diffusion maps eigenvalues $\mu_k$ via \eqref{eq:evals_relationship}).
+However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to correspond to the two smallest (non-trivial) eigenvalues (the largest eigenvalue will always be $\tilde{\mu}_{0,0} = 1$ and correspond to a constant eigenfunction; these eigenvalues are related to the diffusion maps eigenvalues $\mu_k$ via \eqref{eq:evals_relationship}).
 %
@@
 %
-Looking at the diffusion maps eigenvectors which correspond to modes which are {\em not} higher harmonics will yield the most parsimonious representation of the data.
+Utilizing only the diffusion maps eigenvectors which correspond to modes which are {\em not} higher harmonics yields the most parsimonious representation of the data.
 %
@@
 
-\subsection{Algorithm: Local Linear Regression}
+\subsection{Algorithm: local linear regression}
 
@@
 
-Given the eigenvectors $\phi_1, \phi_2, \dots, \phi_{n-1} \in \mathbb{R}^m$, we would like to automatically deduce which ones are harmonics/functions of another. 
-%
-To accomplish this, we will fit a function $f(\phi_1, \dots, \phi_{k-1})$ to $\phi_{k}$. 
-%
-If the resulting fit is good/accurate, then $\phi_{k}$ is most likely a harmonic of $\phi_1, \dots, \phi_{k-1}$. 
+Given the eigenvectors $\phi_1, \phi_2, \dots, \phi_{n-1} \in \mathbb{R}^m$, we would like to automatically deduce which ones are higher harmonics of previous eigenvectors. 
+%
+We will attempt to fit a function $f(\phi_1, \dots, \phi_{k-1})$ to $\phi_{k}$. 
+%
+If the resulting fit is accurate, then $\phi_{k}$ is most likely a harmonic of $\phi_1, \dots, \phi_{k-1}$. 
 %
@@
 \end{equation}
-where 
+%
+as our functional approximation, where 
 %
@@
 %
-A large value of $r_{k}$ indicates that $\phi_{k}$ parameterizes a new direction in the data.
-%
-Because $r_{k} \approx 1$ if $\phi_{k}$ parameterizes a new direction, we define $r_1 = 1$.
+Conversely, a large value of $r_{k}$ indicates that $\phi_{k}$ parameterizes a new direction in the data.
+%
+Because $r_{k} \approx 1$ if $\phi_{k}$ parameterizes a new direction, we choose to define $r_1 = 1$.
 
@@
 \begin{equation} 
-r_{k} = \sqrt{ \frac{\sum_{i=1}^n \left( \frac{ \phi_{k} (i) - \sum_{j=1}^m L_{k-1}(i,j) \phi_{k}(j) }{1-L_{k-1}(i,i)} \right)^2} {\sum_{i=1}^n  \left( \phi_{k} (i) \right)^2 }}
-\end{equation}
-
-\subsection{Calculating the relative lengths of the manifold}
+r_{k} = \sqrt{ \frac{\sum_{i=1}^n \left( \frac{ \phi_{k} (i) - \sum_{j=1}^m L_{k-1}(i,j) \phi_{k}(j) }{1-L_{k-1}(i,i)} \right)^2} {\sum_{i=1}^n  \left( \phi_{k} (i) \right)^2 }} .
+\end{equation}
+
+\subsection{Calculating the relative lengths of the manifold} \label{sec:relative_lengths}
 
@@
 %
-Let $\mu_{i_1}, \mu_{i_2}, \dots$ denote the eigenvalues which correspond to eigenvectors which are {\em not} higher harmonics. 
-%
-From \eqref{eq:evals} and \eqref{eq:evals_relationship}, we propose to use
+Let $\mu_{i_1}, \mu_{i_2}, \dots$ denote the eigenvalues which correspond to eigenvectors which are {\em not} higher harmonics (i.e., those eigenvectors where $r_{i_j}$ is large). 
+%
+From \eqref{eq:evals} and \eqref{eq:evals_relationship}, we propose to approximate the relative lengths along the manifold by
 \begin{equation} \label{eq:est_lengths}
@@
 %
-We can then use this measure to determine how many components are required to retain most of the information within the data set.
-
-\subsection{Illustrative Examples}
+We can use this measure to determine how many components are required to retain most of the information within the data set: if $L_j$ is small, then this component can be disregarded.
+
+\subsection{Illustrative examples} \label{sec:illustrative_examples}
 
@@
 %
-The first data set is the strip where the solution is known analytically, and the second and third data sets are nonlinear manifolds which provide examples of the flexibility of diffusion maps. 
-%
-However, these data sets will also illustrate the issues of harmonics.
+The first data set is a two-dimensional strip where the eigenvalues and eigenvectors are known analytically.
+%
+The second and third data sets are nonlinear manifolds which demonstrate the flexibility of our approach.
 
@@
 
-Our first illustrative example consists of three different strip data sets. 
-%
-Each of these data sets contains $m=1500$ data points uniformly distributed along the strip. 
+\begin{figure}[t]
+\centering
+\begin{subfigure}{0.3\textwidth}
+\includegraphics[height=3cm]{strip_data_L2}
+\includegraphics[height=3.5cm]{strip_spectrum_L2}
+\caption{}
+\end{subfigure}
+%
+\begin{subfigure}{0.3\textwidth}
+\includegraphics[height=3cm]{strip_data_L4}
+\includegraphics[height=3.5cm]{strip_spectrum_L4}
+\caption{}
+\end{subfigure}
+%
+\begin{subfigure}{0.35\textwidth}
+\includegraphics[height=3cm]{strip_data_L8}
+\includegraphics[height=3.5cm]{strip_spectrum_L8}
+\caption{}
+\end{subfigure}
+%
+\caption{Data sets (top) and eigenvalue spectra (bottom) for strips with (a) $L_1 = 2$, $L_2 = 1$, (b) $L_1 = 4$, $L_2 = 1$, (c) $L_1 = 8$, $L_2 = 1$.
+The empirical eigenvalues are plotted in blue, and the analytical eigenvalues are plotted in red.
+From the eigenvalues which are identified as important/non-harmonics, the estimated length ratio is (a) 2.0, (b) 4.4, (c) 8.8.
+}
+\label{fig:strip_compare_analytic}
+\end{figure}
+
+Our first illustrative example consists of three different two-dimensional strip data sets. 
+%
+Each data set contains $m=1500$ data points uniformly sampled from the strip. 
 %
@@
 %
-The diffusion maps eigenvalues are consistent with the known analytic eigenvalues of the Laplacian, and using \eqref{eq:est_lengths}, we can accurately estimate the relative lengths of the two principal directions in each data set.
-
-\subsubsection{Swiss Roll}
-
-Our second illustrative example will consist of two different swiss roll data sets.
-%
-The data are
-\begin{equation}
-\begin{aligned}
-z_1 =& \theta \cos \theta \\
-z_2 =& \theta \sin \theta \\
-z_3 =& h t
-\end{aligned}
-\end{equation}
-%
-where $\theta$ is uniformly sampled along the arclength of the spiral, and $t \sim Unif(0,1)$. 
-%
-The height of the first swiss roll is $h = 40$, while the height of the second is $h = 20$. 
-%
-Each data set consists of $m=1500$ points, shown shown in Figure~\ref{fig:swiss_rolls}(a)~and~(b).
-%
-We expect to recover the eigenvector which parameterizes the height farther down in the spectrum for the second data set relative to the first, as this second data set is shorter.
-
-Figure~\ref{fig:swiss_rolls}~(c)~and~(d) shows the diffusion maps eigenvalue spectra for the two data sets.
-%
-The bars are colored by the leave-one-out cross-validation error as defined in \eqref{eq:cv_error}; a small value of $r_k$ indicate that the corresponding eigenvector is a harmonic of previous eigenvectors, while a large value of $r_k$ indicates that the corresponding eigenvector describes a new direction in the data.
-%
-From these plots, one can conclude that the first two eigenvectors $\phi_1$ and $\phi_2$ parameterize the first data set, while $\phi_1$ and $\phi_5$ parameterize the second. 
-%
-Figure~\ref{fig:swiss_rolls}~(e--h) shows the two data sets, colored by the two ``important'' eigenvectors. 
-%
-As expected, these eigenvectors parameterize the arc length along the spiral, and the height of the swiss roll.
-
-\begin{figure}[t]
+The eigenvalues are colored by the leave-one-out cross-validation error as defined in \eqref{eq:cv_error}; a small value of $r_k$ indicate that the corresponding eigenvector is a harmonic of previous eigenvectors, while a large value of $r_k$ indicates that the corresponding eigenvector describes a new direction in the data.
+%
+As expected, the gap between the two meaningful eigenvalues increases as the strip becomes narrower. 
+%
+The diffusion maps eigenvalues are consistent with the known analytic eigenvalues of the Laplacian (see \eqref{eq:evals} and \eqref{eq:evals_relationship}), and using \eqref{eq:est_lengths}, we can accurately estimate the relative lengths of the two principal directions in each data set.
+
+\subsubsection{Swiss roll}
+
+\begin{figure}[th]
 %
@@
 
+Our second illustrative example consists of two different swiss roll data sets.
+%
+The data are
+\begin{equation}
+\begin{aligned}
+z_1 =& \theta \cos \theta \\
+z_2 =& \theta \sin \theta \\
+z_3 =& h t
+\end{aligned}
+\end{equation}
+%
+such that $z_1, z_2$ are uniformly sampled along the arclength of the spiral, and $t \sim Unif(0,1)$. 
+%
+The height of the first swiss roll is $h = 40$, while the height of the second is $h = 20$. 
+%
+Each data set consists of $m=1500$ points, shown shown in Figure~\ref{fig:swiss_rolls}(a)~and~(b).
+%
+We expect to recover the eigenvector which parameterizes the height farther down in the spectrum for the second data set relative to the first, as this second data set is shorter.
+
+Figure~\ref{fig:swiss_rolls}~(c)~and~(d) shows the diffusion maps eigenvalue spectra for the two data sets.
+%
+Again, the bars are colored by the leave-one-out cross-validation error, where a small value of $r_k$ indicates that the corresponding eigenvector is a harmonic of previous eigenvectors. 
+%
+From these plots, one can conclude that the first two eigenvectors $\phi_1$ and $\phi_2$ parameterize the first data set, while $\phi_1$ and $\phi_5$ parameterize the second. 
+%
+Figure~\ref{fig:swiss_rolls}~(e--h) shows the two data sets, colored by the two ``important'' eigenvectors. 
+%
+As expected, these eigenvectors parameterize the arc length along the spiral, and the height of the swiss roll.
+
 \subsubsection{Torus}
-
-For the third example, we consider a torus defined by
-%
-\begin{equation}
-\begin{aligned}
-z_1 =& (r_1 + r_2 \cos \theta_2 ) \cos \theta_1 \\
-z_2 =& (r_1 + r_2 \cos \theta_2 ) \sin \theta_1 \\
-z_3 =& r_2 \sin \theta_2
-\end{aligned}
-\end{equation}
-%
-where $r_1 > r_2$ are the outer and inner radii, respectively, and $\theta_1, \theta_2 \sim Unif(0, 2 \pi)$. 
-%
-Changing the ratio of $r_1$ to $r_2$ changes the ratio of the eigenvalues corresponding to non-harmonic eigenvectors.
-
-We expect to obtain two eigenfunctions ($\sin \theta_1$ and $\cos \theta_1$) which parmeterize the outer circle, and two eigenfunctions ($\sin \theta_2$ and $\cos \theta_2$) which parameterize the inner circle.
-%
-However, the number of harmonics which fall between these relevant/descriptive eigenfunctions depends on the ratio of the two radii.
-
-Figure~\ref{fig:torus} shows the diffusion maps eigenspectra for three different torii. 
-%
-Clearly, as the torus becomes thinner, the second pair of relevant eigenfunctions (corresponding to $\sin \theta_2$ and $\cos \theta_2$) moves farther down in the spectrum.
 
@@
 
-\section{Chemotaxis: a case study}
-
-We will now demonstrate our methodology on simulation data from a model problem arising in the study of cellular dynamics. 
-%
-Although on the microscale the system consists of many stochastic and noisy particles, it exhibits coherent dynamics on the macroscale which are governed by only a few parameters.
+For the third example, we consider a torus defined by
+%
+\begin{equation}
+\begin{aligned}
+z_1 =& (r_1 + r_2 \cos \theta_2 ) \cos \theta_1 \\
+z_2 =& (r_1 + r_2 \cos \theta_2 ) \sin \theta_1 \\
+z_3 =& r_2 \sin \theta_2
+\end{aligned}
+\end{equation}
+%
+where $r_1 > r_2$ are the outer and inner radii, respectively, and $\theta_1, \theta_2 \sim Unif(0, 2 \pi)$. 
+%
+Changing the ratio of $r_1$ to $r_2$ changes the ratio of the eigenvalues corresponding to non-harmonic eigenvectors.
+
+We expect to obtain two eigenfunctions ($\sin \theta_1$ and $\cos \theta_1$) which parmeterize the outer circle, and two eigenfunctions ($\sin \theta_2$ and $\cos \theta_2$) which parameterize the inner circle.
+%
+However, the number of harmonics which fall between the two pairs of descriptive eigenfunctions depends on the ratio of the two radii.
+
+Figure~\ref{fig:torus} shows the diffusion maps eigenspectra for three different torii, again colored by the leave-one-out cross-validation error. 
+%
+Clearly, as the torus becomes thinner, the second pair of relevant eigenfunctions (corresponding to $\sin \theta_2$ and $\cos \theta_2$) moves farther down in the spectrum.
+
+\section{Chemotaxis: A Case Study}
+
+Our main motivation for factoring out the harmonics in diffusion maps is to facilitate the analysis of complex data sets where the underlying structure is not readily apparent. 
+%
+In data from complex dynamical systems, noisy microscopic behavior often gives rise to cohernt macroscopic dynamics. 
+%
+%Although on the microscale the system consists of many stochastic and noisy particles, it exhibits coherent dynamics on the macroscale which are governed by only a few parameters.
 %
@@
 %
-We will show how we can use our methodology to extract a parameterization of data collected at the microscale which is consistent with the macroscopic behavior, without any {\em a priori} knowledge of the appropriate microscopic or macroscopic model.
-%
-Furthermore, determining the dimensionality of such a parameterization can inform us to the requisite dimensionality of a reduced model which captures the relevant macroscopic behavior.
+We will show how we can use diffusion maps to extract a parameterization of data collected at the microscale which is consistent with the macroscopic behavior, without any {\em a priori} knowledge of the appropriate microscopic or macroscopic model.
+%
+Furthermore, determining the {\em true} dimensionality of such a parameterization by factoring out higher harmonics in the diffusion maps analysis uncovers the requisite dimensionality of a reduced model which captures the relevant macroscopic behavior.
+%
+Knowing appropriate macroscopic variables and the true dimensionality can help inform modeling efforts and aid in the writing of accurate macroscale models.
 
@@
 
-Our model problem arises in studies of cellular chemotaxis \cite{othmer2000diffusion}, where 
-biological cells work together and exhibit coherent macroscopic dynamics regulated by extracellular sensed signals, in order to accomplish tasks such as finding food or navigating away from toxins.
+Our model problem comes from studies of cellular chemotaxis \cite{othmer2000diffusion}, where biological cells work together and exhibit coherent macroscopic dynamics regulated by extracellular sensed signals in order to accomplish tasks such as finding food or navigating away from toxins.
 %
@@
 
-In this example, the macroscopic dynamics vary depending on the value of a single system parameter.
-%
-We will demonstrate how we can use diffusion maps \cite{coifman2005geometric}, a manifold learning technique, to {\em automatically} detect changes in the system's macroscopic dynamics from microscale data.
-%
-This case study exhibits several key points which allow us to demonstrate the strength of our approach.
-%
-First, this system has random microscopic behavior with smooth macroscopic dynamics governed by a single parameter which determines the regime/mode of the system. 
-%
-Second, it consists of an ensemble of individual stochastic particles which must be described using the appropriate observers.
-%
-Third, it has an analytic macroscopic description, which serves as a ground truth and will allow us to verify our results which are obtained in an unsupervised manner.
+%In this example, the macroscopic dynamics of the group of cells vary depending on the value of a single system parameter.
+%
+%We will demonstrate how we can use diffusion maps \cite{coifman2005geometric}, a manifold learning technique, to {\em automatically} detect changes in the system's macroscopic dynamics from microscale data.
+%
+%This case study exhibits several key points which allow us to demonstrate the strength of our diffusion maps approach.
+%
+%First, this system has random microscopic behavior with smooth macroscopic dynamics governed by a single parameter which determines the regime/mode of the system. 
+%
+%Second, it consists of an ensemble of individual stochastic particles which must be described using the appropriate observers.
+%
+%Third, it has an analytic macroscopic description, which serves as a ground truth and will allow us to verify our results which are obtained in an unsupervised manner.
+This specific example has an analytic macroscopic description in which the macroscopic dynamics of the group of cells vary depending on the value of a single system parameter.
+%
+This model serves as a ground truth and will allow us to verify our results which are obtained in an unsupervised manner.
 %
@@
 %
-We would like to note that we have chosen a very specific one-parameter family of initial conditions, which lead to simple dynamics and straightforward illustration of our main points.
-%
-More complex initial conditions could easily be chosen.
+We would like to note that we have chosen a very specific one-parameter family of initial conditions which lead to simple dynamics and straightforward illustration of our main points;
+more complex initial conditions could easily be chosen.
 
@@
 
-\begin{figure}[t!]
-\begin{subfigure}{6cm}
-\includegraphics[width=\textwidth]{EMD_withhist_p_1}
-\caption{}
-\label{subfig:small_lambda_p}
-\end{subfigure}
-\begin{subfigure}{6cm}
-\includegraphics[width=\textwidth]{EMD_withhist_t_1}
-\caption{}
-\label{subfig:small_lambda_t}
-\end{subfigure}
-\begin{subfigure}{6cm}
-\includegraphics[width=\textwidth]{EMD_withhist_p_400}
-\caption{}
-\label{subfig:large_lambda_p}
-\end{subfigure}
-\begin{subfigure}{6cm}
-\includegraphics[width=\textwidth]{EMD_withhist_t_400}
-\caption{}
-\label{subfig:large_lambda_t}
-\end{subfigure}
-\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with (a,b) $\lambda=1$, $s=1$, and (c,d) $\lambda=400$, $s=20$.
-For each set, we run many simulations, each with $N=1000$ particles.
-%
-We vary the initial conditions such that $0.1 \le p  \le 0.9$, and allow each simulation to evolve for $10$ time units.
-%
-We discard the initial portion of each simulation trajectory ($t < 1$), which corresponds to the initial relaxation of the system.
-%
-The observation $z(i) \in \mathbb{R}^n$ is the histogram of $x_1(i), x_2(i), \dots, x_N(i)$ into $n = 32$ bins.
-%
-The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions.
-The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time.
-Representative histograms are shown for selected data points.}
-\label{fig:dmaps_embed_emd}
-\end{figure}
-
-\subsection{Observers and Metrics}
+\subsection{Observers and metrics}
 
@@
 %
-For simple data, such as the illustrative examples in the previous section, the observers are simply the data points themselves. 
+For simple data, such as the illustrative examples in Section~\ref{sec:illustrative_examples}, the observers are simply the data points themselves. 
 %
@@
 %
-For example, when looking at a protein, the observers may be the locations of the $\alpha$-carbon atoms, as these backbone atoms are known to describe most of the relevant dynamics within the protein.
+For example, when looking at a protein, the observers may be the locations of the $\alpha$-carbons, as these backbone atoms are known to capture most of the relevant dynamics. 
 %
@@
 %
-Observers should also be robust to noise. 
-%
-The scattering transform \cite{mallat2012group} is robust to small deformations, and histograms serve to linearize noise \cite{talmon2013empirical}.
-
-After constructing appropriate observers, we must consider what metric to use in the diffusion maps calculation. 
-%
-Again, the simplest option is to use the Euclidean distance, but this can often prove uninformative. 
-%
-For molecular simulation data, it is often appropriate to use the Euclidean distance {\em after} aligning configurations with respect to rotations, as global rotations of the entire molecule are not meaningful.
+Observers should also be robust to noise, which motivates the scattering transform \cite{mallat2012group} and histograms \cite{talmon2013empirical}.
+
+After constructing appropriate observers, one must choose the appropriate metric for the diffusion maps calculation. 
+%
+The simplest option (used for the illustrative examples in Section~\ref{sec:illustrative_examples}) is to use the Euclidean distance, but this can often prove uninformative. 
+%
+For molecular simulation data, it is often appropriate to use the Euclidean distance {\em after} aligning configurations with respect to rotations, as global rotations of the entire molecule are not meaningful \cite{ferguson2010systematic}.
 %
@@
 %
-Another option is the Earth Mover's Distance, a multiscale distance metric which is stable and robust to noise.
+Another option is the Earth Mover's Distance, a multiscale distance metric which is stable and robust to noise \cite{rubner2000earth}.
 
@@
 %
-Note: histograms are not essential for computing EMD, however, they make it computationally feasible. Furthermore, histograms are good because they are robust to noise (cite PNAS).
+We note that although histograms are not essential for using EMD, histograms make the computation feasible. 
+%
+The histograms are also robust to noise \cite{talmon2013empirical}.
 
@@
 
+\begin{figure}[t!]
+\begin{subfigure}{6cm}
+\includegraphics[width=\textwidth]{EMD_withhist_p_1}
+\caption{}
+\label{subfig:small_lambda_p}
+\end{subfigure}
+\begin{subfigure}{6cm}
+\includegraphics[width=\textwidth]{EMD_withhist_t_1}
+\caption{}
+\label{subfig:small_lambda_t}
+\end{subfigure}
+\begin{subfigure}{6cm}
+\includegraphics[width=\textwidth]{EMD_withhist_p_400}
+\caption{}
+\label{subfig:large_lambda_p}
+\end{subfigure}
+\begin{subfigure}{6cm}
+\includegraphics[width=\textwidth]{EMD_withhist_t_400}
+\caption{}
+\label{subfig:large_lambda_t}
+\end{subfigure}
+\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with (a,b) $\lambda=1$, $s=1$, and (c,d) $\lambda=400$, $s=20$.
+For each set, we run many simulations, each with $N=1000$ particles.
+%
+We vary the initial conditions such that $0.1 \le p  \le 0.9$, and allow each simulation to evolve for $10$ time units.
+%
+We discard the initial portion of each simulation trajectory ($t < 1$), which corresponds to the initial relaxation of the system.
+%
+The observation $z(i) \in \mathbb{R}^n$ is the histogram of $x_1(i), x_2(i), \dots, x_N(i)$ into $n = 32$ bins.
+%
+The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions.
+The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time.
+Representative histograms are shown for selected data points.}
+\label{fig:dmaps_embed_emd}
+\end{figure}
+
 Figure~\ref{fig:dmaps_embed_emd} shows the results of analyzing two sets of chemotaxis simulations using diffusion maps. 
@@
 In contrast, the correlations between the embedding coordinates and the macroscopic variables when using EMD in the diffusion maps calculation are all greater than 90\%.
+
+\subsubsection{Detecting changes in dynamical behavior}
 
@@
 \end{figure}
-
-\subsubsection{Detecting changes in dynamical behavior}
 
@@
 
+\begin{figure}
+%
+\def\figheight{1.4in}
+%
+\begin{subfigure}[t]{0.31\textwidth}
+\centering
+\includegraphics[height=\figheight]{chemotaxis1_evals}
+\includegraphics[height=\figheight]{chemotaxis1_embed_good}
+\vspace{1.2in}
+%\includegraphics[height=1.5in]{chemotaxis1_embed_bad}
+\caption{}
+\end{subfigure}
+%
+\begin{subfigure}[t]{0.31\textwidth}
+\centering
+\includegraphics[height=\figheight]{chemotaxis2_evals}
+\includegraphics[height=\figheight]{chemotaxis2_embed_good}
+\includegraphics[height=\figheight]{chemotaxis2_embed_bad}
+\caption{}
+\end{subfigure}
+%
+\begin{subfigure}[t]{0.35\textwidth}
+\centering
+\includegraphics[height=\figheight]{chemotaxis3_evals}
+\includegraphics[height=\figheight]{chemotaxis3_embed_good}
+\includegraphics[height=\figheight]{chemotaxis3_embed_bad}
+\caption{}
+\end{subfigure}
+%
+\caption{Diffusion maps embedding for three different chemotaxis simulations.
+(a) $\lambda = 100$, $s = 10$.
+(b) $\lambda = 2500$, $s = 50$.
+(c) $\lambda = 6400$, $s = 80$.
+For each simulation, the eigenvalue spectrum is shown.
+From the spectra, we can see that the first two components are informative for (a), components 1 and 3 are informative for (b), and components 1 and  are informative for (c).
+The corresponding embeddings are shown below.
+For comparison, the embeddings using the first two components are also shown in (b) and (c), where clearly, the second component is a harmonic of the first and does not describe a new direction in the data.
+}
+%
+\label{fig:chemotaxis_simulations_harmonics}
+\end{figure}
+
+\begin{figure}[t]
+%
+\centering
+\includegraphics[width=\textwidth]{chemotaxis_compare_timescales_evals}
+%
+\caption{Detecting the change in dimensionality.
+The ratio of the log of the two ``meaningful'' diffusion maps eigenvalues (those which correspond to modes which are not harmonics) is shown in blue, and the ratio of the diffusive and drift time scales are shown in red.
+When the diffusive time scale is significantly larger than the drift time scale, the data is expected to be effectively one-dimensional.
+This transition is consistent with the diffusion maps eigenvalues: the data is effectively one-dimensional if the first eigenvalue is significantly larger than the second.
+}
+%
+\label{fig:chemotaxis_compare_timescales_evals}
+%
+\end{figure}
+
 Often in dynamical systems, one would like to detect a change in the dimensionally of the system. 
@@
 %
-However, at long times, we know the dynamics become diffusive and $p$ becomes unimportant.
+However, for large values of $\lambda$, we know the dynamics become diffusive and $p$ becomes unimportant; we therefore expect the data to look approximately one-dimensional when $\lambda$ is large.
 
@@
 %
-This ratio is also plotted in Figure~\ref{fig:chemotaxis_compare_timescales_evals} as a function of $\lambda$, and the transition is consisted with the transition predicted by the diffusion maps eigenvalues.
-
-\begin{figure}
-%
-\def\figheight{1.4in}
-%
-\begin{subfigure}[t]{0.31\textwidth}
-\centering
-\includegraphics[height=\figheight]{chemotaxis1_evals}
-\includegraphics[height=\figheight]{chemotaxis1_embed_good}
-\vspace{1.2in}
-%\includegraphics[height=1.5in]{chemotaxis1_embed_bad}
-\caption{}
-\end{subfigure}
-%
-\begin{subfigure}[t]{0.31\textwidth}
-\centering
-\includegraphics[height=\figheight]{chemotaxis2_evals}
-\includegraphics[height=\figheight]{chemotaxis2_embed_good}
-\includegraphics[height=\figheight]{chemotaxis2_embed_bad}
-\caption{}
-\end{subfigure}
-%
-\begin{subfigure}[t]{0.35\textwidth}
-\centering
-\includegraphics[height=\figheight]{chemotaxis3_evals}
-\includegraphics[height=\figheight]{chemotaxis3_embed_good}
-\includegraphics[height=\figheight]{chemotaxis3_embed_bad}
-\caption{}
-\end{subfigure}
-%
-\caption{Diffusion maps embedding for three different chemotaxis simulations.
-(a) $\lambda = 100$, $s = 10$.
-(b) $\lambda = 2500$, $s = 50$.
-(c) $\lambda = 6400$, $s = 80$.
-For each simulation, the eigenvalue spectrum is shown.
-From the spectra, we can see that the first two components are informative for (a), components 1 and 3 are informative for (b), and components 1 and  are informative for (c).
-The corresponding embeddings are shown below.
-For comparison, the embeddings using the first two components are also shown in (b) and (c), where clearly, the second component is a harmonic of the first and does not describe a new direction in the data.
-}
-%
-\label{fig:chemotaxis_simulations_harmonics}
-\end{figure}
-
-\begin{figure}[t]
-%
-\centering
-\includegraphics[width=\textwidth]{chemotaxis_compare_timescales_evals}
-%
-\caption{Detecting the change in dimensionality.
-The ratio of the log of the two ``meaningful'' diffusion maps eigenvalues (those which correspond to modes which are not harmonics) is shown in blue, and the ratio of the diffusive and drift time scales are shown in red.
-When the diffusive time scale is significantly larger than the drift time scale, the data is expected to be effectively one-dimensional.
-This transition is consistent with the diffusion maps eigenvalues: the data is effectively one-dimensional if the first eigenvalue is significantly larger than the second.
-}
-%
-\label{fig:chemotaxis_compare_timescales_evals}
-%
-\end{figure}
+This ratio is also plotted in Figure~\ref{fig:chemotaxis_compare_timescales_evals} as a function of $\lambda$, and the transition is consistent with the transition predicted by the diffusion maps eigenvalues.
 

### 2015-03-01 chemotaxis/paper/draft1.tex
updated chemotaxis
incorporated some of Ronen's comments

@@
 
-\item While we can ignore/address nonuniform densities through normalization of the Laplace operator, we still have the problem of identifying the ``true'' components due to geometry.
+\item While we can ignore/address nonuniform densities through normalization of the Laplace operator, we still have the problem of repeated eigendirections.
+
+\item Add text about the meaning of the eigenvectors (obtained in the form of eigenvectors of a Laplace operator), as demonstrated in the chemotaxis example.
+The main idea is the ability to obtain, in a data driven manner, the "true" parsimonious description of the system, which is certain cases (such as particular regimes/modes of the chemotaxis system) may be significantly different from the typical representation and choice of variables used to describe the system (e.g.
+in the chemotaxis, it will not necessarily be the two fluxes).
 
@@
 
-\section{Manifold Learning and Some Relevant Issues}
+\section{Manifold Learning Based on Laplace Operators}
 
@@
 \begin{figure}[t]
-\begin{subfigure}{0.5\textwidth}
+\centering
+\begin{subfigure}{0.45\textwidth}
 \includegraphics[width=\textwidth]{strip_cnts1}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
+\caption{}
+\end{subfigure}
+%
+\begin{subfigure}{0.45\textwidth}
 \includegraphics[width=\textwidth]{strip_cnts2}
+\caption{}
 \end{subfigure}
@@
 
-In recent years, it has been noted that eigenfunctions of the Laplace-Beltrami operator with Neumann boundary conditions can be used to parameterize the manifold \cite{Belkin2003, coifman2005geometric, singer2008non}.
+In recent years, it has been noted that eigenfunctions of the Laplace-Beltrami operator can be used to parameterize the manifold \cite{Belkin2003, coifman2005geometric, singer2008non}. 
 %
@@
 \begin{figure}[t]
-\begin{subfigure}{0.5\textwidth}
+\centering
+\begin{subfigure}{0.45\textwidth}
 \includegraphics[width=\textwidth]{strip_discrete1}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
+\caption{}
+\end{subfigure}
+%
+\begin{subfigure}{0.45\textwidth}
 \includegraphics[width=\textwidth]{strip_discrete4}
-\end{subfigure}
-
-\begin{subfigure}{0.5\textwidth}
+\caption{}
+\end{subfigure}
+
+\begin{subfigure}{0.45\textwidth}
 \includegraphics[width=\textwidth]{strip_nonuniform1}
-\end{subfigure}
-\begin{subfigure}{0.5\textwidth}
+\caption{}
+\end{subfigure}
+%
+\begin{subfigure}{0.45\textwidth}
 \includegraphics[width=\textwidth]{strip_nonuniform2}
+\caption{}
 \end{subfigure}
@@
 %
-We will do this by constructing a matrix which approximates the Laplace-Beltrami operator with Neumann boundary conditions. 
-%
-The eigenvectors of this matrix then approximate the eigenfunctions of this operator.
+We will do this by constructing a matrix which approximates the Laplace-Beltrami operator with Neumann boundary conditions (TODO: is it explicit in the construction?). 
+%
+The eigenvectors of this matrix then approximate the eigenfunctions of this operator \cite{...}.
 
@@
 %
-However, often some eigenvectors are higher harmonics of previous eigenvectors and do not describe new directions in the data set.
+However, often some eigenvectors are higher harmonics of previous eigenvectors and do not describe new directions in the data set \cite{gerber2007robust}.
 %
@@
 %
-In fact, if $L_1 > 2 L_2$, then $\tilde{\mu}_{2, 0} < \tilde{\mu}_{0, 1}$, and so the second (non-trivial) eigenvector (when the eigenvectors are ordered by their corresponding eigenvalues) will be a harmonic of the first and still parameterize $z_1$ (see Figure~\ref{fig:strip_harmonics}).
+In fact, if $L_1 > 2 L_2$, then $\tilde{\mu}_{2, 0} < \tilde{\mu}_{0, 1}$, and so the second (non-trivial) eigenvector (when the eigenvectors are ordered by their corresponding eigenvalues) will be a repeated eigendirection of the first and still parameterize $z_1$ (see Figure~\ref{fig:strip_harmonics}).
 %
@@
 %
+We will show how we can automatically detect these modes to obtain a meaningful representation of the data.
+%
 Furthermore, the corresponding eigenvalues provide us with a measure of the relative lengths of the data set along the principal directions.
@@
 \begin{figure}[t]
-\begin{subfigure}{0.5\textwidth}
+\centering
+\begin{subfigure}{0.45\textwidth}
 \includegraphics[width=\textwidth]{strip_discrete1}
 \end{subfigure}
-\begin{subfigure}{0.5\textwidth}
+%
+\begin{subfigure}{0.45\textwidth}
 \includegraphics[width=\textwidth]{strip_discrete2}
@@
 
-\begin{subfigure}{0.5\textwidth}
+\begin{subfigure}{0.45\textwidth}
 \includegraphics[width=\textwidth]{strip_discrete3}
 \end{subfigure}
-\begin{subfigure}{0.5\textwidth}
+%
+\begin{subfigure}{0.45\textwidth}
 \includegraphics[width=\textwidth]{strip_discrete4}
@@
 
-Given the eigenvectors $\phi_1, \phi_2, \dots, \phi_{n-1} \in \mathbb{R}^m$, we would like to automatically deduce which ones are higher harmonics of previous eigenvectors. 
+Given the eigenvectors $\phi_1, \phi_2, \dots, \phi_{n-1} \in \mathbb{R}^m$, we would like to automatically deduce which ones capture new directions in the data, and which ones are merely repeated eigendirections.
+%
+This problem was addressed previously in \cite{...} by performing successive iterations of diffusion maps, interspersed with advection along the first principal direction at each iteration. 
+%
+However, both the advection procedure and the successive eigendecompositions are very expensive and intractable for larger data set. 
+%
+Here, we propose an alternative approach to address the problem of repeated eigendirections. 
 %
@@
 \begin{equation}
-K(\Phi_{k-1}(i), \Phi_{k-1}(j))  = \exp \left( - \frac{\|\Phi_{k-1}(i) - \Phi_{k-1} (j) \|^2}{\epsilon^2} \right),
-\end{equation}
-%
-and take $\epsilon = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k-1}(i)$.
+K(\Phi_{k-1}(i), \Phi_{k-1}(j))  = \exp \left( - \frac{\|\Phi_{k-1}(i) - \Phi_{k-1} (j) \|^2}{\epsilon_{reg}^2} \right),
+\end{equation}
+%
+where $\epsilon_{reg}$ is the kernel scale. 
+%
+We typically take $\epsilon_{reg} = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k-1}(i)$, as we empirically found this choice to yield good results. 
 %
@@
 %
-and the matrix $W_{k-1}(i) = diag \left( K(\Phi_{k-1}(i), \Phi_{k-1}(1)), \dots, K(\Phi_{k-1}(i), \Phi_{k-1}(m)) \right)$.
+and the matrix 
+\begin{equation}
+W_{k-1}(i) = diag \left( K(\Phi_{k-1}(i), \Phi_{k-1}(1)), \dots, K(\Phi_{k-1}(i), \Phi_{k-1}(m)) \right).
+\end{equation}
 %
@@
 %
-We can use this measure to determine how many components are required to retain most of the information within the data set: if $L_j$ is small, then this component can be disregarded.
+We can use this measure to determine how many components are required to retain most of the information within the data set: if $L_j$ is small, then this component can be disregarded (TODO: define small).
 
@@
 \begin{subfigure}{0.3\textwidth}
-\includegraphics[height=3cm]{strip_data_L2}
-\includegraphics[height=3.5cm]{strip_spectrum_L2}
-\caption{}
-\end{subfigure}
+\includegraphics[height=2.5cm]{strip_data_L2}
+\includegraphics[height=3cm]{strip_spectrum_L2}
+\caption{}
+\end{subfigure}
+%
+\hfill
 %
 \begin{subfigure}{0.3\textwidth}
-\includegraphics[height=3cm]{strip_data_L4}
-\includegraphics[height=3.5cm]{strip_spectrum_L4}
-\caption{}
-\end{subfigure}
+\includegraphics[height=2.5cm]{strip_data_L4}
+\includegraphics[height=3cm]{strip_spectrum_L4}
+\caption{}
+\end{subfigure}
+%
+\hfill
 %
 \begin{subfigure}{0.35\textwidth}
-\includegraphics[height=3cm]{strip_data_L8}
-\includegraphics[height=3.5cm]{strip_spectrum_L8}
+\includegraphics[height=2.5cm]{strip_data_L8}
+\includegraphics[height=3cm]{strip_spectrum_L8}
 \caption{}
@@
 
-\begin{figure}[th]
+\begin{figure}[!th]
 %
@@
 \begin{subfigure}{0.3\textwidth}
+\centering
 \includegraphics[height=0.75in]{torus1}
@@
 %
+\hfill
+%
 \begin{subfigure}{0.32\textwidth}
+\centering
 \includegraphics[height=0.75in]{torus2}
@@
 %
+\hfill
+%
 \begin{subfigure}{0.35\textwidth}
+\centering
 \includegraphics[height=0.75in]{torus3}
@@
 \end{subfigure}
+%
+\hfill
 %
@@
 \begin{figure}[t!]
-\begin{subfigure}{6cm}
+\centering
+\begin{subfigure}{5cm}
 \includegraphics[width=\textwidth]{EMD_withhist_p_1}
@@
 \end{subfigure}
-\begin{subfigure}{6cm}
+\begin{subfigure}{5cm}
 \includegraphics[width=\textwidth]{EMD_withhist_t_1}
@@
 \end{subfigure}
-\begin{subfigure}{6cm}
+\begin{subfigure}{5cm}
 \includegraphics[width=\textwidth]{EMD_withhist_p_400}
@@
 \end{subfigure}
-\begin{subfigure}{6cm}
+\begin{subfigure}{5cm}
 \includegraphics[width=\textwidth]{EMD_withhist_t_400}
@@
 \centering
-\includegraphics[width=\textwidth]{detect_change_eigenvalues}
+\includegraphics[width=0.5\textwidth]{detect_change_eigenvalues}
 \caption{The first two eigenvalues from diffusion maps as a function of $\lambda$.
@@
 \centering
-\includegraphics[width=\textwidth]{chemotaxis_compare_timescales_evals}
+\includegraphics[width=0.5\textwidth]{chemotaxis_compare_timescales_evals}
 %

### 2015-03-07 chemotaxis/paper/draft1.tex
updated stuff after ronen visited

@@
 
-\journal{Computers \& Chemical Engineering}
+%\journal{Computers \& Chemical Engineering}
+\journal{Applied and Computational Harmonic Analysis Letter}
 
@@
 
-\begin{itemize}
-
-\item
-Data driven analysis methods, and in particular, manifold learning, can do many things (dimensionality reduction, reduction in dynamical systems, etc).
-
-\item
-Manifold learning constructs parameterization of the data through the spectral analysis of a Laplace operator.
-
-\item
-Initially it was used almost exclusively for toy examples/synthetic data (``these people suck").
-
-\item
-Only recently, has it been applied to real (experimental/simulation) data, and this was possible due to advances in observers (data representation) and metrics (``we are awesome").
-
-\item
-However, a major limitation (of spectral methods based on a Laplace operator) is identifying the true principal components.
-
-\item For the example of a square/strip, we know everything analytically, and identifying the ``true'' principal components is a problem/not obvious, and depends both on geometry and density
-
-\item While we can ignore/address nonuniform densities through normalization of the Laplace operator, we still have the problem of repeated eigendirections.
-
-\item Add text about the meaning of the eigenvectors (obtained in the form of eigenvectors of a Laplace operator), as demonstrated in the chemotaxis example.
+In recent years, data mining algorithms have proven useful for many disciplines and applications \cite{....}. 
+%
+In particular, for dynamical systems, data-driven methodologies have been shown to be essential when a simple macroscopic model cannot be written analytically due to the complexity of the system.
+%
+For simulations, the governing model is often very high-dimensional and the macroscale dynamics are not obvious; for experimental systems, an explicit model is often unknown. 
+%
+For such cases, data obtained from observations and/or simulations of the dynamical system combined with these methodologies give rise to a low-dimensional description which can not only provide insight into the underlying dynamics, but also serve as a first step in constructing macroscale models which are consistent with the observed microscale behavior.
+
+%For many engineering applications, it is often advantageous or even essential to effectively parameterize a system's dynamics. 
+%TODO: add reasons why, potential applications, etc.
+%%
+%Often, such a parameterization cannot be obtained analytically, due to the complexity of the system.
+%%
+%For simulations, the governing model is often very high-dimensional and the macroscale dynamics are not obvious; for experimental systems, an explicit model is often unknown. 
+%%
+%In such cases, we utilize data obtained from observations and/or simulations of the dynamical system, and turn to data-driven analysis methods to provide an accurate low-dimensional description of this data.
+%%
+%This low-dimensional description can not only provide insight into the underlying dynamics, but also serve as a first step in constructing macroscale models which are consistent with the observed microscale behavior.
+%
+
+Due to the complexity and range of microscale behaviors possible in dynamical systems, we use manifold learning, a nonlinear data-driven analysis technique, to analyze data.
+%
+Most manifold learning algorithms \cite{...} construct parameterizations of the data through the spectral analysis of a Laplace operator.
+%
+The data is then embedded in a new, low-dimensional coordinate system given by the eigenvectors of this operator. 
+%
+Our hope is that these coordinates, obtained in a data-driven manner, will describe the variables which govern the macroscale dynamical behavior of the system and enable us to introduce a reduced macroscale ab-initio model.
+
+These algorithms were first applied to synthetic data sets to illustrate their geometric properties and flexibility. 
+%
+Only recently have they been applied to experimental and simulation data, which was possible due to advances in data representation (observers) and metrics. 
+%
+From a geometric perspective, not all eigenvectors are guaranteed to parameterize unique directions within the data; some eigenvectors are 	``repeated eigendirections'' which describe the same mode. 
+%
+Identifying those eigenvectors is critical for obtaining an accurate parameterization of the system which captures the true dimensionality of the macroscale dynamics.  
+%
+This is a challenging task which is often done manually, and few methods have been proposed to automate the identification of the unique eigendirections.
+
+In this paper, we propose an algorithm to automatically identify the unique eigendirections using local linear regression.
+%
+We first demonstrate our algorithm on synthetic examples where we have a closed-form solution for the eigenvectors and eigenvalues of the Laplace operator.  
+%
+This will allow us to validate our proposed approach.
+%
+We then consider a complex data set from a stochastic dynamical system which models cellular chemotaxis. 
+%
+The recent advances in observers and metrics, coupled with the proposed approach for factoring out repeated eigendirections, provide a data analysis pipeline which can successfully analyze data from this simulation.
+%
+This pipeline will allow us to detect changes in the regime/mode of the system, as well as changes in the underlying dimensionality of the macroscale dynamics.
+
+%\begin{itemize}
+%
+%\item
+%Data driven analysis methods, and in particular, manifold learning, can do many things (dimensionality reduction, reduction in dynamical systems, etc). 
+%
+%\item
+%Manifold learning constructs parameterization of the data through the spectral analysis of a Laplace operator.
+%
+%\item
+%Initially it was used almost exclusively for toy examples/synthetic data (``these people suck").
+%
+%\item
+%Only recently, has it been applied to real (experimental/simulation) data, and this was possible due to advances in observers (data representation) and metrics (``we are awesome"). 
+%
+%\item
+%However, a major limitation (of spectral methods based on a Laplace operator) is identifying the true principal components.
+%
+%\item For the example of a square/strip, we know everything analytically, and identifying the ``true'' principal components is a problem/not obvious, and depends both on geometry and density
+%
+%\item While we can ignore/address nonuniform densities through normalization of the Laplace operator, we still have the problem of repeated eigendirections. 
+%
+%\item Add text about the meaning of the eigenvectors (obtained in the form of eigenvectors of a Laplace operator), as demonstrated in the chemotaxis example.
 The main idea is the ability to obtain, in a data driven manner, the "true" parsimonious description of the system, which is certain cases (such as particular regimes/modes of the chemotaxis system) may be significantly different from the typical representation and choice of variables used to describe the system (e.g.
 in the chemotaxis, it will not necessarily be the two fluxes).
-
-\item
-In this paper, we propose an algorithm to automatically identify those principal components using locally linear regression.
-
-\item
-All of these considerations (recently developed plus what we propose here) provide a data analysis pipeline.
+%
+%\item
+%In this paper, we propose an algorithm to automatically identify those principal components using locally linear regression.
+%
+%\item
+%All of these considerations (recently developed plus what we propose here) provide a data analysis pipeline.
 We show that this pipeline can successfully analyze data from a complex stochastic dynamical system originated from the study of cellular dynamics (chemotaxis).
-
-\end{itemize}
+%
+%\end{itemize}
 
 \section{Manifold Learning Based on Laplace Operators}
+
+Data from dynamical systems is most often high-dimensional, since it is collected at the microscale.
+%
+However, the macroscale dynamics often evolve to a low-dimensional description. 
+%
+In such cases, the high-dimensional microscale data will be restricted to a low-dimensional manifold embedded in the high-dimensional space. 
+%
+Extracting a parameterization of this manifold corresponds to parameterizing the low-dimensional, macroscale dynamics \cite{...}.
+
+Let $z(1), \dots, z(m) \in \mathbb{R}^n$ denote $m$ observations sampled from the dynamical system. 
+%
+We assume the $n$-dimensional observations $z(i)$ lie on a $d$-dimensional manifold $\mathcal{M}_d$, where $d < n$. 
+%
+We will show how to extract a $d$-dimensional parameterization of the observations which is consistent with the geometry of the manifold.
 
@@
 
-We first consider the problem of parameterizing a continuous $d$-dimensional manifold $\mathcal{M}_d$ embedded in $\mathbb{R}^n$.
+We start by considering a continuous setting where rigorous analysis is available for specific examples. 
+%
+In such an example, we will show that the existence of repeated eigendirections is inherent to the manifold learning setup based on Laplace operators, and that the identification of the unique eigendirections is nontrivial.
+%
+The existence of these challenges in the continuous setting implies that, even in the limit of infinite data, such repeated eigendirections still pose a problem for analysis. 
+%
+Thus, we consider the problem of parameterizing a continuous $d$-dimensional manifold $\mathcal{M}_d$ embedded in $\mathbb{R}^n$.
 %
@@
 %
-However, in the case when the manifold is curved, it is more difficult to define a set of coordinates.
-
-In recent years, it has been noted that eigenfunctions of the Laplace-Beltrami operator can be used to parameterize the manifold \cite{Belkin2003, coifman2005geometric, singer2008non}. 
-%
-To provide intuition as to why this is the case, consider a two-dimensional strip with edge lengths $L_1$ and $L_2$. 
+However, in the case when the manifold is nonlinear, the set of coordinates is not readily apparent.
+
+In recent years, the eigenfunctions of the Laplace-Beltrami operator have been used to parameterize the manifold \cite{Belkin2003, coifman2005geometric, singer2008non}. 
+%
+To illustrate why these eigenfunctions provide appropriate coordinates, consider a two-dimensional strip with edge lengths $L_1$ and $L_2$. 
 %
@@
 \end{equation}
-where $z_1$ and $z_2$ denote the two coordinates of the rectangular domain \cite{singer2008non}. 
+where $z_1$ and $z_2$ denote the two coordinates of the strip \cite{singer2008non}. 
 %
@@
 %
-Although these eigenfunctions can only be written analytically for very special cases, it has been observed empirically that the eigenfunctions often provide appropriate coordinates to parameterize more complex, nonlinear manifolds.
+The analytic form of the eigenfunctions in \eqref{eq:efuncs} illustrates the two issues we address in this paper. 
+%
+First, $z_1$ and $z_2$ are not necessarily decoupled in subsequent eigenfunctions. 
+TODO: address this in later section
+%
+Second, eigenfunctions with $k_1+k_2 \ge 2$ do not describe any additional directions along the strip; we will refer to these as ``repeated eigendirections,'' and we will refer to the eigenfunctions with $k_1+k_2 =1$ as ``unique eigendirections.''
+%
+We note that, although the eigenfunctions can only be written analytically for very special cases, it has been observed empirically that the eigenfunctions often provide appropriate coordinates to parameterize more complex, nonlinear manifolds.
 
@@
 %
-We will do this by constructing a matrix which approximates the Laplace-Beltrami operator with Neumann boundary conditions (TODO: is it explicit in the construction?). 
-%
-The eigenvectors of this matrix then approximate the eigenfunctions of this operator \cite{...}.
-
-Assume we are given $m$ observations $z(1), \dots, z(m) \in \mathcal{M}_d$. 
-%
-We first construct the matrix $W \in \mathbb{R}^{m \times m}$, with
+We will do this by constructing a matrix which approximates the Laplace-Beltrami operator. 
+%
+It was shown in \cite{...} that, in the limit of infinite data, the discrete Laplacian constructed from data converges pointwise to the continuous Laplace-Beltrami operator on the manifold. 
+%
+As a result, the eigenvectors of the discrete Laplacian approximate the eigenfunctions of this continuous operator \cite{...}.
+
+Given observations $z(1), \dots, z(m) \in \mathcal{M}_d$, we construct the matrix $W \in \mathbb{R}^{m \times m}$, with
 \begin{equation} \label{eq:W}
@@
 %
-$\epsilon$ can be chosen using several heuristics \cite{coifman2008graph, rohrdanz2011determination}; we often take $\epsilon$ to be the median of the pairwise distances between the data points.
+The kernel's scale $\epsilon$ can be chosen using several heuristics \cite{coifman2008graph, rohrdanz2011determination}; we often take $\epsilon$ to be the median of the pairwise distances between the data points.
 %
@@
 %
-Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ approximate the eigenfunctions of the Laplace-Beltrami operator (with Neumann boundary conditions) on $\mathcal{M}_d$,
+Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ approximate the eigenfunctions of the Laplace-Beltrami operator on $\mathcal{M}_d$,
 and the eigenvalues $\mu_0, \mu_1, \dots, \mu_{m-1}$ of $A$ are related to the eigenvalues of the continuous operator by 
@@
 %
-As discussed previously, the eigenfunctions provide a parameterization of the manifold, such that $\phi_{j}(i)$ yields the $j^{th}$ embedding coordinate for $z(i)$.
-%
-We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$, and if the data do lie on a low-dimensional manifold, we only need to retain a few of the leading eigenvectors to adequately describe the data.
-%
-Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a trivial constant vector which is not used as an embedding coordinate.
 
@@
 
+As discussed previously, the eigenfunctions provide a parameterization of the manifold, such that $\phi_{j}(i)$ yields the $j^{th}$ embedding coordinate for $z(i)$; this embedding is called diffusion maps \cite{...}.
+%
+We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$.%, and if the data do lie on a low-dimensional manifold, we only need to retain a few of the leading eigenvectors to adequately describe the data.
+%
+Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a trivial constant vector.
+
 Figure~\ref{fig:strip_evecs} shows data sampled from a strip, colored by diffusion maps eigenvectors. 
@@
 
+From the previous section, we know that the eigenfunctions with $(k_1, k_2) =(1, 0)$ and $(k_1, k_2) =(0, 1)$ provide embedding coordinates for the manifold; these two eigenfunctions are both uncoupled and not repeated.
+%
+From \eqref{eq:evals}, we see that if we sort the eigenvectors by the magnitude of the corresponding eigenvalues, these eigenvectors are guaranteed to appear before any coupled or repeated eigendirections.
+%
+However, these two eigenvectors are {\em not} guaranteed to appear as the first two (non-trivial) eigenvectors, as harmonics of the first eigendirection (i.e., $\cos \left( n \pi z_1 / L_1 \right)$ with $n > 1$) could appear before the second.
+%
+In practice, identifying which eigenvectors correspond to unique eigendirections is nontrivial, as the correspondence between the original data and the coordinates which parameterize the manifold are unknown.
+%
+We note that, in contrast to our simple illustrative example, for most data sets of interest, the coordinates for the underlying manifold cannot easily be obtained from the coordinates of the original data.
+
 \section{Identifying the Informative Eigenvectors }
 
-In diffusion maps, we order the eigenvectors by the magnitude of the corresponding eigenvalues, in the hopes that the leading eigenvectors provide us with a parameterization of the underlying manifold.
-%
-However, often some eigenvectors are higher harmonics of previous eigenvectors and do not describe new directions in the data set \cite{gerber2007robust}.
-%
-Again, for illustrative purposes, we consider the case where $\mathcal{M}_d$ is a $2$-dimensional rectangular domain with edge lengths $L_1  > L_2$. 
-%
-We recall that the eigenfunctions $\tilde{\phi}_{1,0} = \cos \left( \frac {\pi z_1}{L_1} \right)$ and  $\tilde{\phi}_{0,1} = \cos \left( \frac {\pi z_2}{L_2} \right)$ provide embedding coordinates for the manifold $\mathcal{M}_d$.
-%
-However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to correspond to the two smallest (non-trivial) eigenvalues (the largest eigenvalue will always be $\tilde{\mu}_{0,0} = 1$ and correspond to a constant eigenfunction; these eigenvalues are related to the diffusion maps eigenvalues $\mu_k$ via \eqref{eq:evals_relationship}).
+Common practice is to order the eigenvectors by the magnitude of the corresponding eigenvalues, assuming that the leading eigenvectors provide a parameterization of the underlying manifold.
+%
+However, as discussed in the previous section, some eigenvectors are higher harmonics of previous eigenvectors and do not describe new directions in the data set \cite{gerber2007robust}.
+%
+For the case where $\mathcal{M}_d$ is a $2$-dimensional strip with edge lengths $L_1  > L_2$,
+recall that the eigenfunctions $\tilde{\phi}_{1,0} = \cos \left(  {\pi z_1}/{L_1} \right)$ and  $\tilde{\phi}_{0,1} = \cos \left(  {\pi z_2}/{L_2} \right)$ provide embedding coordinates for the manifold $\mathcal{M}_d$.
+%
+However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to correspond to the two smallest (non-trivial) eigenvalues (the largest eigenvalue will always be $\tilde{\mu}_{0,0} = 1$ and correspond to a constant eigenfunction; these eigenvalues are related to the eigenvalues $\mu_k$ of the discrete operator via \eqref{eq:evals_relationship}).
 %
@@
 %
-Utilizing only the diffusion maps eigenvectors which correspond to modes which are {\em not} higher harmonics yields the most parsimonious representation of the data.
-%
-We will show how we can automatically detect these modes to obtain a meaningful representation of the data.
-%
-Furthermore, the corresponding eigenvalues provide us with a measure of the relative lengths of the data set along the principal directions.
+Utilizing only the eigenvectors $\phi_i$ which correspond to unique eigendirections yields the most parsimonious representation of the data.
+%
+We will show how we can automatically detect these eigenvectors to obtain a meaningful representation of the data.
+%
+Furthermore, the corresponding eigenvalues provide us with a measure of the relative lengths of the data set along these unique eigendirections.
 
@@
 
-Given the eigenvectors $\phi_1, \phi_2, \dots, \phi_{n-1} \in \mathbb{R}^m$, we would like to automatically deduce which ones capture new directions in the data, and which ones are merely repeated eigendirections.
-%
-This problem was addressed previously in \cite{...} by performing successive iterations of diffusion maps, interspersed with advection along the first principal direction at each iteration. 
+Given the eigenvectors $\phi_1, \phi_2, \dots, \phi_{m-1} \in \mathbb{R}^m$, we would like to automatically deduce which ones capture new directions in the data, and which ones are merely repeated eigendirections.
+%
+This problem was addressed previously in \cite{...} by performing successive iterations of diffusion maps, interspersed with advection along the first eigendirection at each iteration. 
 %
 However, both the advection procedure and the successive eigendecompositions are very expensive and intractable for larger data set. 
+TODO: find other advantages of our proposed method. 
 %
@@
 %
-If the resulting fit is accurate, then $\phi_{k}$ is most likely a harmonic of $\phi_1, \dots, \phi_{k-1}$. 
+If the resulting fit is accurate, then we assume $\phi_{k}$ is a repeated eigendirection of $\phi_1, \dots, \phi_{k-1}$. 
 %
@@
 \begin{equation}
-\phi_k \approx \alpha( \Phi_{k-1}) + \beta(\Phi_{k-1})^T \Phi_{k-1}
+\phi_k \approx \alpha + \beta^T \Phi_{k-1}
 \end{equation}
@@
 %
-$\Phi_{k-1} = \begin{bmatrix} \phi_1 \\ \vdots \\ \phi_{k-1} \end{bmatrix}$
-and $\beta \in \mathbb{R}^{k-1}$. 
+$\Phi_{k-1} = \begin{bmatrix} \phi_1 & \dots & \phi_{k-1} \end{bmatrix}^T$,
+$\alpha \in \mathbb{R}$, and $\beta \in \mathbb{R}^{k-1}$. 
 %
@@
 
-\eqref{eq:cv_error} can easily be computed \cite{wasserman2006all}.
+The error in \eqref{eq:cv_error} can easily be computed \cite{wasserman2006all}.
 %
@@
 
-Once we have determined which eigenvectors parameterize independent directions along the manifold, we can use the corresponding eigenvalues to calculate the relative lengths along these directions. 
-%
-Let $\mu_{i_1}, \mu_{i_2}, \dots$ denote the eigenvalues which correspond to eigenvectors which are {\em not} higher harmonics (i.e., those eigenvectors where $r_{i_j}$ is large). 
-%
-From \eqref{eq:evals} and \eqref{eq:evals_relationship}, we propose to approximate the relative lengths along the manifold by
+Once the eigenvectors which parameterize unique eigendirections along the manifold have been determined, the corresponding eigenvalues can be used to calculate the relative lengths along these directions.
+%
+Let $\mu_{i_1}, \mu_{i_2}, \dots$ denote the eigenvalues corresponding to eigenvectors which parameterize unique eigendirections (i.e., those eigenvectors where $r_{i_j}$ is large). 
+%
+From \eqref{eq:evals} and \eqref{eq:evals_relationship}, we propose to approximate the relative lengths $L_j$  along the manifold by
 \begin{equation} \label{eq:est_lengths}
-L_j \propto \frac{1}{\sqrt{-\log \mu_{i_j}}}
-\end{equation}
-where $L_1, L_2, \dots$ are the relative lengths along the principal directions of the manifold. 
-%
-We can use this measure to determine how many components are required to retain most of the information within the data set: if $L_j$ is small, then this component can be disregarded (TODO: define small).
+L_j \propto \frac{1}{\sqrt{-\log \mu_{i_j}}}.
+\end{equation}
+%
+This measure is used to determine how many components are required to retain most of the information within the data set.
+%
+For example, if $L_j$ is smaller than a predefined threshold, then this component can be disregarded.
 
@@
 
-We will demonstrate the local linear regression algorithm for identifying informative eigenvectors on three synthetic data sets. 
-%
-The first data set is a two-dimensional strip where the eigenvalues and eigenvectors are known analytically.
+We will demonstrate our proposed approach on three synthetic data sets. 
+%
+The first data set is the two-dimensional strip discussed previously, where the eigenvalues and eigenvectors are known analytically.
 %
@@
 From the eigenvalues which are identified as important/non-harmonics, the estimated length ratio is (a) 2.0, (b) 4.4, (c) 8.8.
-}
+TODO: normalize colors}
 \label{fig:strip_compare_analytic}
@@
 %
-The eigenvalues are colored by the leave-one-out cross-validation error as defined in \eqref{eq:cv_error}; a small value of $r_k$ indicate that the corresponding eigenvector is a harmonic of previous eigenvectors, while a large value of $r_k$ indicates that the corresponding eigenvector describes a new direction in the data.
-%
-As expected, the gap between the two meaningful eigenvalues increases as the strip becomes narrower. 
-%
-The diffusion maps eigenvalues are consistent with the known analytic eigenvalues of the Laplacian (see \eqref{eq:evals} and \eqref{eq:evals_relationship}), and using \eqref{eq:est_lengths}, we can accurately estimate the relative lengths of the two principal directions in each data set.
+The eigenvalues are colored by the leave-one-out cross-validation error as defined in \eqref{eq:cv_error}; a small value of $r_k$ indicate that the corresponding eigenvector is a repeated eigendirection, while a large value of $r_k$ indicates that the corresponding eigenvector describes a new direction in the data.
+%
+The analytic eigenvalues of the Laplacian (see \eqref{eq:evals} and \eqref{eq:evals_relationship}) are also indicated in red.
+
+We observe that the eigenvalues are consistent with the known analytic eigenvalues of the Laplacian (see \eqref{eq:evals} and \eqref{eq:evals_relationship}).
+%
+Furthermore, the two unique eigendirections can be identified, since their corresponding regression error $r_k$ is large. 
+%
+As expected, we observe that the gap between the two meaningful eigenvalues increases as the strip becomes longer. 
+%
+Using \eqref{eq:est_lengths}, we can accurately estimate the relative lengths of the two unique eigendirections in each data set.
 
@@
 (b) Data set 2; $h = 20$.
-(c) Diffusion maps eigenvalue spectrum for data set 1.
-(d) Diffusion maps eigenvalue spectrum for data set 2.
-(e) Data set 1, colored by the first diffusion maps eigenvector.
-(f) Data set 2, colored by the first diffusion maps eigenvector.
-(g) Data set 1, colored by the second diffusion maps eigenvector which is not a harmonic of the first.
-(h) Data set 2, colored by the second diffusion maps eigenvector which is not a harmonic of the first.
+(c) Eigenvalue spectrum from the analysis of data set 1.
+(d) Eigenvalue spectrum from the analysis data set 2.
+(e) Data set 1, colored by the first eigenvector.
+(f) Data set 2, colored by the first eigenvector.
+(g) Data set 1, colored by the second eigenvector.
+(h) Data set 2, colored by the fifth eigenvector.
 }
@@
 %
-The data are
+The data are sampled according to
 \begin{equation}
@@
 %
+Note that, in this example, $z_1, z_2, z_3$ are the original coordinates; however, the data lie on a two-dimensional manifold parameterized by the $\theta$ and $t$. 
+%
 The height of the first swiss roll is $h = 40$, while the height of the second is $h = 20$. 
@@
 %
-We expect to recover the eigenvector which parameterizes the height farther down in the spectrum for the second data set relative to the first, as this second data set is shorter.
-
-Figure~\ref{fig:swiss_rolls}~(c)~and~(d) shows the diffusion maps eigenvalue spectra for the two data sets.
-%
-Again, the bars are colored by the leave-one-out cross-validation error, where a small value of $r_k$ indicates that the corresponding eigenvector is a harmonic of previous eigenvectors. 
+%We expect to recover the eigenvector which parameterizes the height farther down in the spectrum for the second data set relative to the first, as this second data set is shorter.
+
+Figure~\ref{fig:swiss_rolls}~(c)~and~(d) shows the eigenvalue spectra from the analysis of the two data sets.
+%
+Similar to Figure~\ref{fig:strip_compare_analytic}, the bars are colored by the leave-one-out cross-validation error, where a small value of $r_k$ indicates that the corresponding eigenvector is a repeated eigendirection.
 %
@@
 %
-Figure~\ref{fig:swiss_rolls}~(e--h) shows the two data sets, colored by the two ``important'' eigenvectors. 
-%
-As expected, these eigenvectors parameterize the arc length along the spiral, and the height of the swiss roll.
+Figure~\ref{fig:swiss_rolls}~(e--h) shows the two data sets, colored by the two eigenvectors identified by our method as parameterizing the unique eigendirections. 
+%
+As expected, these eigenvectors are one-to-one with the arc length along the spiral, and the height of the swiss roll.
 
@@
 %
-\caption{Torus data sets (left) and corresponding diffusion maps eigenvalues (right).
+\caption{Torii data sets (left) and corresponding eigenvalues (right).
 (a) $r_1 = 3$, $r_2 = 1$.
@@
 In all three data sets, the first two eigenvalues/eigenvectors correspond to $\sin \theta_1$ and $\cos \theta_1$.
-The second pair of non-harmonic eigenvectors (corresponding to $\sin \theta_2$ and $\cos \theta_2$) are captured by components 7 and 8, 11 and 12, and 15 and 16, respectively.}
+The second pair of unique eigendirections (corresponding to $\sin \theta_2$ and $\cos \theta_2$) are captured by components 7 and 8, 11 and 12, and 15 and 16, respectively.}
 %
@@
 %
-Changing the ratio of $r_1$ to $r_2$ changes the ratio of the eigenvalues corresponding to non-harmonic eigenvectors.
-
-We expect to obtain two eigenfunctions ($\sin \theta_1$ and $\cos \theta_1$) which parmeterize the outer circle, and two eigenfunctions ($\sin \theta_2$ and $\cos \theta_2$) which parameterize the inner circle.
-%
-However, the number of harmonics which fall between the two pairs of descriptive eigenfunctions depends on the ratio of the two radii.
-
-Figure~\ref{fig:torus} shows the diffusion maps eigenspectra for three different torii, again colored by the leave-one-out cross-validation error. 
-%
-Clearly, as the torus becomes thinner, the second pair of relevant eigenfunctions (corresponding to $\sin \theta_2$ and $\cos \theta_2$) moves farther down in the spectrum.
+In this example, $z_1, z_2, z_3$ are the coordinates of the data; however, we expect to obtain from our analysis two eigenfunctions ($\sin \theta_1$ and $\cos \theta_1$) which parameterize the outer circle, and two eigenfunctions ($\sin \theta_2$ and $\cos \theta_2$) which parameterize the inner circle.
+%
+Intuitively, increasing $r_1$ makes the eigendirections which parameterize the outer circle ($\cos \theta_1$ and $\sin \theta_1$) more dominant compared to the eigendirections with parameterize the inner circle.
+%
+Increasing $r_1$ can be viewed as analogous to increasing the ratio of $L_1$ to $L_2$ in the strip example.
+
+Figure~\ref{fig:torus} shows the eigenspectra from the analysis of three different torii, colored by the leave-one-out cross-validation error. 
+%
+We observe that as the torus becomes thinner ($r_1$ increases), the second pair of eigenvalues corresponding to unique eigendirections (corresponding to $\sin \theta_2$ and $\cos \theta_2$) moves farther down in the spectrum.
 
@@
 
-Our main motivation for factoring out the harmonics in diffusion maps is to facilitate the analysis of complex data sets where the underlying structure is not readily apparent. 
+Our main motivation for identifying the unique eigendirections is to facilitate the analysis of complex data sets where the underlying structure is not readily apparent. 
 %
@@
 %
-We will show how we can use diffusion maps to extract a parameterization of data collected at the microscale which is consistent with the macroscopic behavior, without any {\em a priori} knowledge of the appropriate microscopic or macroscopic model.
-%
-Furthermore, determining the {\em true} dimensionality of such a parameterization by factoring out higher harmonics in the diffusion maps analysis uncovers the requisite dimensionality of a reduced model which captures the relevant macroscopic behavior.
+We will show how we can use our proposed methodology to extract a parameterization of data collected at the microscale which is consistent with the macroscopic behavior, without any {\em a priori} knowledge of the appropriate microscopic or macroscopic model.
+%
+Furthermore, determining the {\em true} dimensionality of such a parameterization by identifying the unique eigendirections in the analysis uncovers the requisite dimensionality of a reduced model which captures the relevant macroscopic behavior.
 %
@@
 
-Our model problem comes from studies of cellular chemotaxis \cite{othmer2000diffusion}, where biological cells work together and exhibit coherent macroscopic dynamics regulated by extracellular sensed signals in order to accomplish tasks such as finding food or navigating away from toxins.
+Our model problem describes the process of cellular chemotaxis \cite{othmer2000diffusion}, where biological cells work together and exhibit coherent macroscopic dynamics regulated by extracellular sensed signals in order to accomplish tasks such as finding food or navigating away from toxins.
 %
@@
 %
-In this model, each cell is initialized with a position and a velocity on a line, and the dynamics of each cell are governed by a stochastic process.
-%
-At random times, a cell will ``turn around'' and switch the direction of its velocity (this turning is controlled by extracellular signals). 
+%In this model, each cell is initialized with a position and a velocity on a line, and the dynamics of each cell are governed by a stochastic process.
+%
+%At random times, a cell will ``turn around'' and switch the direction of its velocity (this turning is controlled by extracellular signals). 
 %
 %We will analyze the dynamics of collections of such cells/particles. 
-
+%
 %In this example, the macroscopic dynamics of the group of cells vary depending on the value of a single system parameter.
@@
 %
-We will show that manifold learning algorithms based on statistical observers and with the appropriate affinity metric between pairs of observations uncovers a parameterization of the microscopic data which is consistent with the macroscopic model.
-%
-Furthermore, these data-driven methods allow us to detect changes in dynamical behavior resulting from changes in the system parameters.
+Thus far, we only considered synthetic data sets for which the Euclidean distance between data points served as an informative metric. 
+%
+Here, we will show that our approach, when utilizing the appropriate statistical observers and affinity metric between pairs of observations, uncovers a parameterization of the microscopic data which is consistent with the macroscopic model.
+%
+Furthermore, we will show that changes in dynamical behavior resulting from changes in the system parameters can be automatically detected.
 
@@
 
-The microscopic model consists of a collection of $N$ particles whose states are defined by their positions and velocities. 
-%
-Let $x_i(t)$ and $v_i(t)$ denote the position and velocity, respectively, of particle $i$ at time $t$.
-%
-The velocity of each particle is either $\pm s$, where $s$ is a (fixed) speed. 
-%
-We initialize the particles such that
+The microscopic model consists of a collection of $N$ cells whose states are defined by their positions and velocities on a line, and the dynamics of each cell are governed by a stochastic process.
+%
+Let $x_i(t)$ and $v_i(t)$ denote the position and velocity, respectively, of cell $i$ at time $t$.
+%
+The velocity of each cell is $\pm s$, where $s$ is a (fixed) speed. 
+%
+We initialize the cells such that
 \begin{equation}\label{eqn:system}
@@
 \end{equation}
-where $0 < p < 1$ is the probability of a particle initially moving to the right.
-%
-The velocity of each particle randomly switches between $\pm s$ following an (independent) Poisson process with rate $\lambda$.
-%
-We would like to note that we have chosen a very specific one-parameter family of initial conditions which lead to simple dynamics and straightforward illustration of our main points;
-more complex initial conditions could easily be chosen.
-
-This particular example has a known analytic macroscopic equation that governs the overall system behavior.
-%
-For a large collection of particles ($N \rightarrow \infty$), the system can be described by the probability density of the particles.
-%
-Let $\rho(x, t)$ denote the probability density of the particles, and let $\rho^-(x, t)$ and $\rho^+(x, t)$ denote the densities of the particles moving towards the positive and negative axis directions, respectively.
-%
-It can be shown that, as $N \rightarrow \infty$, the densities obey the following set of partial differential equations (PDEs) \cite{othmer2000diffusion}:
-\begin{equation} \label{eqn:coupled_pdes}
-\begin{aligned}
-\frac{\partial \rho^+}{\partial t} + s \frac{\partial \rho^+}{\partial x} & = -\lambda \rho^+ +\lambda \rho^- \\
-\frac{\partial \rho^-}{\partial t} - s \frac{\partial \rho^-}{\partial x} & = \lambda \rho^+ -\lambda \rho^- 
-\end{aligned}
-\end{equation}
-%
-Alternatively, \eqref{eqn:coupled_pdes} can be rewritten as one, second--order PDE:
-\begin{equation} \label{eq:second_order_pde}
-\frac{\partial^2 \rho}{\partial t^2} + 2 \lambda \frac{\partial \rho}{\partial t} = s^2 \frac{\partial ^2 \rho}{\partial x^2}
-\end{equation}
-%
-%We assume that $s^2/\lambda = D$ is constant, so that the dynamics of the probability density are governed by a {\em single} parameter $\lambda$.
-%
-From \eqref{eq:second_order_pde}, we see that, for fixed values of $\lambda$ and $s$, the macroscopic state of the system (the probability density of the particles) is a function of two independent variables: $p$, which controls the initial distribution of the particles, and $t$, the time.
-%
-This implies that for fixed $\lambda$ and $s$, the {\em microscopic} data in a high-dimensional ambient space (e.g., the positions of all $N$ particles) should lie on a two-dimensional manifold parameterized by $p$ and $t$,
-and uncovering the low-dimensional structure of the microscopic data reveals this manifold.
-
-We consider two asymptotic regimes of simulation.
-%
-When $\lambda \rightarrow 0$, the right-hand side of \eqref{eqn:coupled_pdes} tends to 0, and \eqref{eqn:coupled_pdes} becomes two uncoupled wave equations,
-\begin{equation}
-\begin{aligned}
-\frac{\partial \rho^+}{\partial t} + s \frac{\partial \rho^+}{\partial x} & = 0 \\
-\frac{\partial \rho^-}{\partial t} - s \frac{\partial \rho^-}{\partial x} & = 0.
-\end{aligned}
-\end{equation}
-%alternatively, \eqref{eq:second_order_pde} becomes the second order wave equation,
-%\begin{equation}
-%\frac{\partial^2 \rho}{\partial t^2} = s^2 \frac{\partial ^2 \rho}{\partial x^2}.
-%\end{equation}
-
-%Dividing \eqref{eq:second_order_pde} by $\lambda > 0$ yields
-%\[
-%\frac{1}{\lambda} \frac{\partial^2 \rho}{\partial t^2} + 2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2}
-%\]
-When $\lambda \rightarrow \infty$, \eqref{eq:second_order_pde} approaches the heat equation,
-\begin{equation}
-2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2}.
-\end{equation}
-%
-%Two variables determine the macroscopic state of the system: the initial distribution, which is controlled by a single parameter $p$, and the time, $t$.
-%
-The above analysis shows that the initial distribution of velocities of the particles (determined by $p$ in the microscopic simulations) plays a very different role depending on the value of $\lambda$.
-%
-When $\lambda \rightarrow 0$, the dynamics are described by two wave equations, and the initial distribution persists throughout the trajectory.
-%
-When $\lambda \rightarrow \infty$, the dynamics are described by one heat equation, and the initial conditions are insignificant -- the velocity distribution quickly equilibrates and we see purely diffusive behavior.
-%
-We therefore expect the observed dimensionality to change from two to one as a function of $\lambda$. 
-%
-For general dynamical systems, the governing macroscopic variables and asymptotic behavior may not be immediately obvious, and instead need to be inferred from the microscopic data.
-%
-We will demonstrate how we can use diffusion maps to parameterize the simulation data, as well as detect the effective dimensionality of the data.
-%
-%We expect the simulation data to lie on a low-dimensional manifold which is parameterized by the macroscopic variables.
-%
-%We use diffusion maps \cite{coifman2005geometric}, which is a manifold learning technique, to uncover such low-dimensional structure and to analyze the results of the microscopic chemotaxis simulations.
-%
-%Compared to principal components analysis (PCA) \cite{shlens2005tutorial}, a standard linear technique, diffusion maps provides a parameterization of data which lie on a low-dimensional, (possibly) {\em nonlinear} structure in a high-dimensional space.
-%
-%Diffusion maps is one of many recently introduced manifold learning techniques that could be used to analyzed this data \cite{roweis2000nonlinear, tenenbaum2000global, Belkin2003}.
+where $0 < p < 1$ is the probability of a cell initially moving to the right.
+%
+At random times, a cell will ``turn around'' and switch the direction of its velocity (this turning is controlled by extracellular signals). 
+%
+The velocity of each cell randomly switches between $\pm s$ following an (independent) Poisson process with rate $\lambda$.
+%
+We note that we have chosen a very specific one-parameter family of initial conditions which lead to simple dynamics and allow us to illustrate of our main points;
+however, our analysis is not restricted to such specific cases and more complex initial conditions could be used.
 
@@
 
-Appropriate observers can be constructed in several different ways. 
-%
-For simple data, such as the illustrative examples in Section~\ref{sec:illustrative_examples}, the observers are simply the data points themselves. 
-%
-However, for more complex data, the raw data may contain other sources of variability which are not meaningful to the system dynamics. 
-%
-For some examples, physical intuition and {\em a priori} knowledge can aid in constructing appropriate observers. 
-%
-For example, when looking at a protein, the observers may be the locations of the $\alpha$-carbons, as these backbone atoms are known to capture most of the relevant dynamics. 
-%
-Observers can also remove unimportant sources of variation; 
-for example, the bispectrum \cite{zhao2014rotationally} and the scattering transform \cite{mallat2012group} are both invariant to rotations. 
-%
-Observers should also be robust to noise, which motivates the scattering transform \cite{mallat2012group} and histograms \cite{talmon2013empirical}.
-
-After constructing appropriate observers, one must choose the appropriate metric for the diffusion maps calculation. 
-%
-The simplest option (used for the illustrative examples in Section~\ref{sec:illustrative_examples}) is to use the Euclidean distance, but this can often prove uninformative. 
-%
-For molecular simulation data, it is often appropriate to use the Euclidean distance {\em after} aligning configurations with respect to rotations, as global rotations of the entire molecule are not meaningful \cite{ferguson2010systematic}.
-%
-For dynamical systems which are obscured by large variance noise, the Mahalanobis distance often yields a more informative parameterization of the data \cite{talmon2013empirical}.
-%
-Another option is the Earth Mover's Distance, a multiscale distance metric which is stable and robust to noise \cite{rubner2000earth}.
+%Appropriate observers can be constructed in several different ways. 
+%%
+%For simple data, such as the illustrative examples in Section~\ref{sec:illustrative_examples}, the observers are simply the data points themselves. 
+%%
+%However, for more complex data, the raw data may contain other sources of variability which are not meaningful to the system dynamics. 
+%%
+%For some examples, physical intuition and {\em a priori} knowledge can aid in constructing appropriate observers. 
+%%
+%For example, when looking at a protein, the observers may be the locations of the $\alpha$-carbons, as these backbone atoms are known to capture most of the relevant dynamics. 
+%%
+%Observers can also remove unimportant sources of variation; 
+%for example, the bispectrum \cite{zhao2014rotationally} and the scattering transform \cite{mallat2012group} are both invariant to rotations. 
+%%
+%Observers should also be robust to noise, which motivates the scattering transform \cite{mallat2012group} and histograms \cite{talmon2013empirical}.
+%
+%After constructing appropriate observers, one must choose the appropriate metric for the diffusion maps calculation. 
+%%
+%The simplest option (used for the illustrative examples in Section~\ref{sec:illustrative_examples}) is to use the Euclidean distance, but this can often prove uninformative. 
+%%
+%For molecular simulation data, it is often appropriate to use the Euclidean distance {\em after} aligning configurations with respect to rotations, as global rotations of the entire molecule are not meaningful \cite{ferguson2010systematic}.
+%%
+%For dynamical systems which are obscured by large variance noise, the Mahalanobis distance often yields a more informative parameterization of the data \cite{talmon2013empirical}.
+%%
+%Another option is the Earth Mover's Distance, a multiscale distance metric which is stable and robust to noise \cite{rubner2000earth}.
 
@@
 %
+The histograms are also robust to noise \cite{talmon2013empirical}. 
+%
 Instead of the standard Euclidean distance, we use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between pairs of histograms. 
@@
 %
-Although the brute-force computation of the EMD is computationally expensive, there has been a plethora of work in developing efficient algorithms for its computation \cite{Pele-eccv2008, Pele-iccv2009}.
-%
-For the special case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions \cite{rubner2000perceptual}, which can be estimated from histograms as
+Although the brute-force computation of the EMD is computationally expensive, there has been a plethora of work in developing efficient algorithms for its approximation \cite{Pele-eccv2008, Pele-iccv2009}.
+%
+For the special case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions of the data \cite{rubner2000perceptual}, which can be approximated from histograms as
 \begin{equation}
@@
 We note that although histograms are not essential for using EMD, histograms make the computation feasible.
-%
-The histograms are also robust to noise \cite{talmon2013empirical}.
 
@@
 
-\subsubsection{Using the appropriate metric}
+\subsubsection{Identifying the unique eigendirections}
 
@@
 
-Figure~\ref{fig:dmaps_embed_emd} shows the results of analyzing two sets of chemotaxis simulations using diffusion maps. 
+\begin{figure}
+%
+\def\figheight{1.4in}
+%
+\begin{subfigure}[t]{0.31\textwidth}
+\centering
+\includegraphics[height=\figheight]{chemotaxis1_evals}
+\includegraphics[height=\figheight]{chemotaxis1_embed_good}
+\vspace{1.2in}
+%\includegraphics[height=1.5in]{chemotaxis1_embed_bad}
+\caption{}
+\end{subfigure}
+%
+\begin{subfigure}[t]{0.31\textwidth}
+\centering
+\includegraphics[height=\figheight]{chemotaxis2_evals}
+\includegraphics[height=\figheight]{chemotaxis2_embed_good}
+\includegraphics[height=\figheight]{chemotaxis2_embed_bad}
+\caption{}
+\end{subfigure}
+%
+\begin{subfigure}[t]{0.35\textwidth}
+\centering
+\includegraphics[height=\figheight]{chemotaxis3_evals}
+\includegraphics[height=\figheight]{chemotaxis3_embed_good}
+\includegraphics[height=\figheight]{chemotaxis3_embed_bad}
+\caption{}
+\end{subfigure}
+%
+\caption{Diffusion maps embedding for three different chemotaxis simulations.
+(a) $\lambda = 100$, $s = 10$.
+(b) $\lambda = 2500$, $s = 50$.
+(c) $\lambda = 6400$, $s = 80$.
+For each simulation, the eigenvalue spectrum is shown.
+From the spectra, we can see that the first two components are informative for (a), components 1 and 3 are informative for (b), and components 1 and 4 are informative for (c).
+The corresponding embeddings are shown below.
+For comparison, the embeddings using the first two components are also shown in (b) and (c).
+TODO: change (a) to larger $\lambda$ value so that the correspondences are the same across figures.}
+%
+\label{fig:chemotaxis_simulations_harmonics}
+\end{figure}
+
+Figure~\ref{fig:dmaps_embed_emd} shows the results of analyzing two sets of chemotaxis simulations. 
 %
@@
 %
-For both sets of simulations, the two macroscopic variables, $p$ and $t$, are well-correlated with the first two diffusion maps coordinates, $\phi_1$ and $\phi_2$. 
+For both sets of simulations, the macroscopic variables $p$, which controls the initial distribution of the cells, and $t$, the time, are well-correlated with the eigenvectors $\phi_1$ and $\phi_2$. 
 %
@@
 %
-$\phi_1$, the ``more important'' coordinate, is correlated with $p$ for the small $\lambda$ case (Figure~\ref{subfig:small_lambda_p}), and correlated with $t$ for the large $\lambda$ case (Figure~\ref{subfig:large_lambda_t}), indicating that the relative importance of $p$ and $t$ changes in the two simulations;
-this is consistent with the analytic macroscopic description. 
-%
-Furthermore, in the small $\lambda$ regime (wave equation), shown in Figures \ref{subfig:small_lambda_p} and \ref{subfig:small_lambda_t}, the points corresponding to small times are more tightly clustered than the points corresponding to large times.
-%
-This is in agreement with the macroscopic model: at small times, the particles are more condensed around $x=0$, and it is more difficult to distinguish the particles moving to the left from the particles moving to the right.
-%
-On the other hand, at large times, once the particles evolve from the origin, this separation is clear.  
-%
-For the large $\lambda$ case (heat equation), shown in Figures \ref{subfig:large_lambda_p} and \ref{subfig:large_lambda_t}, we observe that at small times, the initial distribution $p$ is well organized in the embedding in Figure \ref{subfig:large_lambda_p}, as the skew of the velocity distribution can be seen in the initial displacements.
-%
-On the other hand, for large times, we observe that the initial distribution $p$ is less organized in Figure \ref{subfig:large_lambda_p}, as the velocities have equilibrated and the initial distribution is less detectable in the particle density.
-%
-Overall, in both cases, we obtain, in an unsupervised data-driven manner, an accurate picture of the macroscopic variables that govern the system dynamics.
-
-%correlations
-%small lambda
-%Euclidean: 3.3848e-04, -0.1861
-%EMD: 0.9000, 0.9807
- %   
-%large lambda
-%Euclidean: -0.5466, 0.3329
-%EMD: 0.9329, 0.9832
-
-To emphasize the importance of using the correct distance metric, we applied diffusion maps to the two sets of simulation data, using the standard Euclidean distance between the histograms to compute the distances in \eqref{eq:W}.
-%
-There is no appreciable correlation between the embedding coordinates and the macroscopic variables $p$ and $t$ (the correlations between the embedding coordinates and the governing macroscopic variables are all  less than 60\%, with the correlations for the small $\lambda$ case being less than 20\%).
-%
-This is due to using an inappropriate metric which does not accurately describe the distances between histograms.
-%
-In contrast, the correlations between the embedding coordinates and the macroscopic variables when using EMD in the diffusion maps calculation are all greater than 90\%.
+The dominant coordinate $\phi_1$ is correlated with $p$ for the small $\lambda$ case (Figure~\ref{subfig:small_lambda_p}), and correlated with $t$ for the large $\lambda$ case (Figure~\ref{subfig:large_lambda_t}), indicating that the relative importance of $p$ and $t$ changes in the two simulations.
+%
+
+As illustrated in the synthetic data sets we discussed previously, the two leading eigenvectors are not guaranteed to correspond to unique eigendirections. 
+%
+Figure~\ref{fig:chemotaxis_simulations_harmonics} shows the results of analyzing three different simulations which span a larger range of $\lambda$ values compared to Figure~\ref{fig:dmaps_embed_emd}.
+%
+We see from the eigenspectra and the cross-validation errors $r_k$ that the second unique eigendirection moves down in the spectrum as $\lambda$ increases. 
+%
+Plotting the data using the two eigenvectors identified as corresponding to unique eigendirections yields two-dimensional embeddings which capture the macroscopic variables $p$ and $t$ (we only show the correspondence of the two identified unique eigendirections with $p$ in the middle row of Figure~\ref{fig:chemotaxis_simulations_harmonics} and report that the correspondence between these eigendirections and $t$ is also consistent).
+%
+In contrast, plotting the data using the two leading eigenvectors $\phi_1$ and $\phi_2$ produces uninformative embeddings for large values of $\lambda$, as $\phi_2$ is a repeated eigendirection which does not parameterize any new directions in the data.
+%
+From the top row of Figure~\ref{fig:chemotaxis_simulations_harmonics}, we can see that the eigenvalue spectrum does not exhibit any large spectral gaps.
+%
+However, there is a significant gap between the eigenvalues which correspond to new directions in the data, and the corresponding eigenvectors provide us with an informative two-dimensional parameterization of the data.
+%which is consistent with the expected macroscopic dynamics.
+%
+It now becomes clear that looking at the eigenvectors {\em modulo} repeated eigendirections is essential for extracting an informative parameterization of the data, as well as characterizing the dimensionality of the data.
 
@@
 \includegraphics[width=0.5\textwidth]{detect_change_eigenvalues}
-\caption{The first two eigenvalues from diffusion maps as a function of $\lambda$.
-Each point is averaged over $10$ simulation experiments (as in Figure~\ref{fig:dmaps_embed_emd}, one experiment consists of data collected at varying $p$ and $t$, for a fixed value of $\lambda$), with error bars denoting one standard deviation.
+\caption{The first two eigenvalues from our analysis as a function of $\lambda$.
+Each point is averaged over $10$ experiments (as in Figure~\ref{fig:dmaps_embed_emd}, one experiment consists of data collected at varying $p$ and $t$, for a fixed value of $\lambda$), with error bars denoting one standard deviation.
 The convergence of the eigenvalues at $\lambda \approx 125$ indicates switching of the dynamical ``mode''.}
@@
 
-We have seen that the eigenvalues $\mu_i$, which quantify the importance of each embedding coordinate, reveal the relative importance of the variables which govern the macroscopic dynamics.
-%
-We can also use these eigenvalues to detect changes in dynamical behavior.
-%
-We expect a gap between the two eigenvalues at the two asymptotic regimes ($\lambda \rightarrow 0$ and $\lambda \rightarrow \infty$), as one of the two macroscopic variables is significantly more important to the dynamical behavior.
-%
-For intermediate $\lambda$ values, we expect the gap between $|\mu_1|$ and $|\mu_2|$ to narrow, with the $\lambda$ value where $|\mu_1| \approx |\mu_2|$ corresponding to the transition between ``wave-like'' and ``heat-like'' behavior.
-
-To demonstrate this point, we performed a new set of experiments, where we simulated the velocity jump process system over a range of $\lambda$ values. 
-% 
-Figure~\ref{fig:detect_change} shows the first two diffusion maps eigenvalues, $\mu_1$ and $\mu_2$, computed from these simulations, as a function of $\lambda$.
-%
-For small $\lambda$ (the wave-equation regime, corresponding to Figures~\ref{subfig:small_lambda_p}~and~\ref{subfig:small_lambda_t}), the two eigenvalues are well-separated, as the first mode (which parameterizes $p$) is significantly more important than the second mode (which parameterizes $t$).
+%We have seen that the eigenvalues $\mu_i$, which quantify the importance of each embedding coordinate, reveal the relative importance of the variables which govern the macroscopic dynamics.
+%
+We can use the eigenvalues from our analysis to detect changes in dynamical behavior.
+%
+From Figure~\ref{fig:dmaps_embed_emd}, we can see that the dominant eigendirection is correlated with $p$ when $\lambda$ is small, and correlated with $t$ when $\lambda$ is large. 
+%
+%We expect an intermediate $\lambda$ value where $p$ and $t$ are equally important. 
+%%
+%From the analytic macroscopic description, we expect a gap between the two eigenvalues at the two asymptotic regimes ($\lambda \rightarrow 0$, where the dynamics are described by two uncoupled wave equations \eqref{...}, and $\lambda \rightarrow \infty$, where the dynamics are described by a single heat equation \eqref{...}), as one of the two macroscopic variables is significantly more important to the dynamical behavior.
+%%
+%For intermediate $\lambda$ values, we expect the gap between $|\mu_1|$ and $|\mu_2|$ to narrow, with the $\lambda$ value where $|\mu_1| \approx |\mu_2|$ corresponding to the transition between ``wave-like'' and ``heat-like'' behavior.
+%
+We performed a new set of simulations over a range of $\lambda$ values;  
+Figure~\ref{fig:detect_change} shows the first two eigenvalues, $\mu_1$ and $\mu_2$, from our analysis as a function of $\lambda$.
+%
+For small $\lambda$ (corresponding to Figures~\ref{subfig:small_lambda_p}~and~\ref{subfig:small_lambda_t}), the two eigenvalues are well-separated, as the first mode (which parameterizes $p$) is significantly more important than the second mode (which parameterizes $t$).
 %
@@
 %
-The eigenvalues then separate again at large $\lambda$ (the heat equation regime, corresponding to Figures~\ref{subfig:large_lambda_p}~and~\ref{subfig:large_lambda_t}), where the $t$ effects dominate the dynamics.
+The eigenvalues then separate again at large $\lambda$ (corresponding to Figures~\ref{subfig:large_lambda_p}~and~\ref{subfig:large_lambda_t}), where the $t$ effects dominate the dynamics.
 %
@@
 
+\subsubsection{Analytic macroscopic description}
+
+This particular example has a known analytic macroscopic equation that governs the overall system behavior.
+%
+For a large collection of cells ($N \rightarrow \infty$), the system can be described by the probability density of the cells.
+%
+Let $\rho(x, t)$ denote the probability density of the cells, and let $\rho^-(x, t)$ and $\rho^+(x, t)$ denote the densities of the cells moving in the negative and positive axis directions, respectively.
+%
+It can be shown that, as $N \rightarrow \infty$, the densities satisfy the following set of partial differential equations (PDEs) \cite{othmer2000diffusion}:
+\begin{equation} \label{eqn:coupled_pdes}
+\begin{aligned}
+\frac{\partial \rho^+}{\partial t} + s \frac{\partial \rho^+}{\partial x} & = -\lambda \rho^+ +\lambda \rho^- \\
+\frac{\partial \rho^-}{\partial t} - s \frac{\partial \rho^-}{\partial x} & = \lambda \rho^+ -\lambda \rho^- 
+\end{aligned}
+\end{equation}
+%
+Alternatively, \eqref{eqn:coupled_pdes} can be rewritten as one, second--order PDE:
+\begin{equation} \label{eq:second_order_pde}
+\frac{\partial^2 \rho}{\partial t^2} + 2 \lambda \frac{\partial \rho}{\partial t} = s^2 \frac{\partial ^2 \rho}{\partial x^2}
+\end{equation}
+%
+%We assume that $s^2/\lambda = D$ is constant, so that the dynamics of the probability density are governed by a {\em single} parameter $\lambda$.
+%
+From \eqref{eq:second_order_pde}, we see that, for fixed values of $\lambda$ and $s$, the macroscopic state of the system (the probability density of the cells) is indeed a function of two variables $p$ and $t$.
+%
+This implies that for fixed $\lambda$ and $s$, the {\em microscopic} data in a high-dimensional ambient space (e.g., the positions of all $N$ cells) should lie on a two-dimensional manifold parameterized by $p$ and $t$,
+and uncovering the low-dimensional structure of the microscopic data reveals this manifold.
+%
+This is consistent with the results presented in Figures~\ref{fig:dmaps_embed_emd}~and~\ref{fig:chemotaxis_simulations_harmonics}, where the two-dimensional embeddings obtained from microscopic simulation data are one-to-one with $p$ and $t$.
+
+We consider two asymptotic regimes of simulation.
+%
+When $\lambda \rightarrow 0$, the right-hand side of \eqref{eqn:coupled_pdes} tends to 0, and \eqref{eqn:coupled_pdes} becomes two uncoupled wave equations,
+\begin{equation}
+\begin{aligned}
+\frac{\partial \rho^+}{\partial t} + s \frac{\partial \rho^+}{\partial x} & = 0 \\
+\frac{\partial \rho^-}{\partial t} - s \frac{\partial \rho^-}{\partial x} & = 0.
+\end{aligned}
+\end{equation}
+%alternatively, \eqref{eq:second_order_pde} becomes the second order wave equation,
+%\begin{equation}
+%\frac{\partial^2 \rho}{\partial t^2} = s^2 \frac{\partial ^2 \rho}{\partial x^2}.
+%\end{equation}
+
+%Dividing \eqref{eq:second_order_pde} by $\lambda > 0$ yields
+%\[
+%\frac{1}{\lambda} \frac{\partial^2 \rho}{\partial t^2} + 2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2}
+%\]
+When $\lambda \rightarrow \infty$, \eqref{eq:second_order_pde} approaches the heat equation,
+\begin{equation}
+2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2}.
+\end{equation}
+%
+%Two variables determine the macroscopic state of the system: the initial distribution, which is controlled by a single parameter $p$, and the time, $t$.
+%
+The above analysis shows that the initial distribution of velocities of the particles (determined by $p$ in the microscopic simulations) plays a very different role depending on the value of $\lambda$.
+%
+When $\lambda \rightarrow 0$, the dynamics are described by two wave equations, and the initial distribution persists throughout the trajectory.
+%
+When $\lambda \rightarrow \infty$, the dynamics are described by one heat equation, and the initial conditions are insignificant -- the velocity distribution quickly equilibrates and we see purely diffusive behavior.
+%
+
+The relative importance of $p$ and $t$ from the analytic description are consistent with the results in Figure~\ref{fig:dmaps_embed_emd}, where $p$ appears as the ``more important'' coordinate when $\lambda$ is small and as the ``less important'' coordinate when $\lambda$ is large.
+%
+Furthermore, in the small $\lambda$ regime (wave equation), shown in Figures \ref{subfig:small_lambda_p} and \ref{subfig:small_lambda_t}, the points corresponding to small times are more tightly clustered than the points corresponding to large times.
+%
+This is in agreement with the macroscopic model: at small times, the particles are more condensed around $x=0$, and it is more difficult to distinguish the particles moving to the left from the particles moving to the right.
+%
+On the other hand, at large times, once the particles evolve from the origin, this separation is clear.  
+%
+For the large $\lambda$ case (heat equation), shown in Figures \ref{subfig:large_lambda_p} and \ref{subfig:large_lambda_t}, we observe that at small times, the initial distribution $p$ is well organized in the embedding in Figure \ref{subfig:large_lambda_p}, as the skew of the velocity distribution can be seen in the initial displacements.
+%
+On the other hand, for large times, we observe that the initial distribution $p$ is less organized in Figure \ref{subfig:large_lambda_p}, as the velocities have equilibrated and the initial distribution is less detectable in the particle density.
+%
+Overall, in both cases, we obtain, in an unsupervised data-driven manner, an accurate picture of the macroscopic variables that govern the system dynamics.
+%
+TODO: address fluxes
+
+%
+%We will demonstrate how we can use diffusion maps to parameterize the simulation data, as well as detect the effective dimensionality of the data.
+%
+%We expect the simulation data to lie on a low-dimensional manifold which is parameterized by the macroscopic variables.
+%
+%We use diffusion maps \cite{coifman2005geometric}, which is a manifold learning technique, to uncover such low-dimensional structure and to analyze the results of the microscopic chemotaxis simulations.
+%
+%Compared to principal components analysis (PCA) \cite{shlens2005tutorial}, a standard linear technique, diffusion maps provides a parameterization of data which lie on a low-dimensional, (possibly) {\em nonlinear} structure in a high-dimensional space.
+%
+%Diffusion maps is one of many recently introduced manifold learning techniques that could be used to analyzed this data \cite{roweis2000nonlinear, tenenbaum2000global, Belkin2003}.
+
+%correlations
+%small lambda
+%Euclidean: 3.3848e-04, -0.1861
+%EMD: 0.9000, 0.9807
+ %   
+%large lambda
+%Euclidean: -0.5466, 0.3329
+%EMD: 0.9329, 0.9832
+
+\subsubsection{Using an appropriate metric}
+
+To emphasize the importance of using an appropriate distance metric, we analyzed the two sets of simulation data using the standard Euclidean distance between the histograms to compute the distances in \eqref{eq:W}.
+%
+We empirically found that there is no appreciable correlation between the embedding coordinates and the macroscopic variables $p$ and $t$ (the correlations between the embedding coordinates and the governing macroscopic variables are all  less than 60\%, with the correlations for the small $\lambda$ case being less than 20\%).
+%
+In contrast, the correlations between the embedding coordinates and the macroscopic variables when using EMD in the diffusion maps calculation are all greater than 90\%.
+%
+This emphasizes the importance of using a metric which accurately describes the distances between observations.
+
 \subsubsection{Detecting changes in dimensionality}
 
-\begin{figure}
-%
-\def\figheight{1.4in}
-%
-\begin{subfigure}[t]{0.31\textwidth}
-\centering
-\includegraphics[height=\figheight]{chemotaxis1_evals}
-\includegraphics[height=\figheight]{chemotaxis1_embed_good}
-\vspace{1.2in}
-%\includegraphics[height=1.5in]{chemotaxis1_embed_bad}
-\caption{}
-\end{subfigure}
-%
-\begin{subfigure}[t]{0.31\textwidth}
-\centering
-\includegraphics[height=\figheight]{chemotaxis2_evals}
-\includegraphics[height=\figheight]{chemotaxis2_embed_good}
-\includegraphics[height=\figheight]{chemotaxis2_embed_bad}
-\caption{}
-\end{subfigure}
-%
-\begin{subfigure}[t]{0.35\textwidth}
-\centering
-\includegraphics[height=\figheight]{chemotaxis3_evals}
-\includegraphics[height=\figheight]{chemotaxis3_embed_good}
-\includegraphics[height=\figheight]{chemotaxis3_embed_bad}
-\caption{}
-\end{subfigure}
-%
-\caption{Diffusion maps embedding for three different chemotaxis simulations.
-(a) $\lambda = 100$, $s = 10$.
-(b) $\lambda = 2500$, $s = 50$.
-(c) $\lambda = 6400$, $s = 80$.
-For each simulation, the eigenvalue spectrum is shown.
-From the spectra, we can see that the first two components are informative for (a), components 1 and 3 are informative for (b), and components 1 and  are informative for (c).
-The corresponding embeddings are shown below.
-For comparison, the embeddings using the first two components are also shown in (b) and (c), where clearly, the second component is a harmonic of the first and does not describe a new direction in the data.
+\begin{figure}[t]
+%
+\centering
+\includegraphics[width=0.5\textwidth]{chemotaxis_compare_timescales_evals}
+%
+\caption{Detecting the change in dimensionality.
+The ratio of the log of the two eigenvalues which correspond to unique eigendirections is shown in blue, and the ratio of the diffusive and drift time scales are shown in red.
+When the diffusive time scale is significantly larger than the drift time scale, the data is expected to be effectively one-dimensional.
+This transition is consistent with the empirical eigenvalues: the data is effectively one-dimensional if the first eigenvalue is significantly larger than the second.
 }
 %
-\label{fig:chemotaxis_simulations_harmonics}
+\label{fig:chemotaxis_compare_timescales_evals}
+%
 \end{figure}
 
-\begin{figure}[t]
-%
-\centering
-\includegraphics[width=0.5\textwidth]{chemotaxis_compare_timescales_evals}
-%
-\caption{Detecting the change in dimensionality.
-The ratio of the log of the two ``meaningful'' diffusion maps eigenvalues (those which correspond to modes which are not harmonics) is shown in blue, and the ratio of the diffusive and drift time scales are shown in red.
-When the diffusive time scale is significantly larger than the drift time scale, the data is expected to be effectively one-dimensional.
-This transition is consistent with the diffusion maps eigenvalues: the data is effectively one-dimensional if the first eigenvalue is significantly larger than the second.
-}
-%
-\label{fig:chemotaxis_compare_timescales_evals}
-%
-\end{figure}
-
-Often in dynamical systems, one would like to detect a change in the dimensionally of the system. 
-%
-We can uncover the true dimensionally of the data set by looking at the eigenvalues corresponding to eigenvectors which parameterize new directions in the data. 
-%
-For the specific chemotaxis example, we know that there are two directions in the data, corresponding to simulation parameters $p$ and $t$. 
-%
-However, for large values of $\lambda$, we know the dynamics become diffusive and $p$ becomes unimportant; we therefore expect the data to look approximately one-dimensional when $\lambda$ is large.
-
-To detect this transition in the data, we look at the eigenvalues corresponding to the two independent eigenevctors. 
+Detecting changes in dimensionality is essential for the analysis and modeling of dynamical systems. 
+%
+For this example, we show that the true dimensionally of a data set can be estimated by looking at the eigenvalues corresponding to unique eigendirections. 
+%
+%For the specific chemotaxis example, we know that there are two directions in the data, corresponding to simulation parameters $p$ and $t$. 
+%%
+%However, for large values of $\lambda$, we know the dynamics become diffusive and $p$ becomes unimportant; we therefore expect the data to look approximately one-dimensional when $\lambda$ is large. 
+%
+%To detect this transition in the data, we look at the eigenvalues corresponding to the unique eigendirections. 
 %
@@
 According to \eqref{eq:est_lengths}, we plot  
-\begin{equation}
+\begin{equation}\label{eq:chemotaxis_eval_ratio}
  \sqrt{\frac{\log \mu_{i_1}}{\log \mu_{i_2}}} ;
@@
 %
-Figure~\ref{fig:chemotaxis_simulations_harmonics} shows the eigenvalue spectra for three different sets of simulations. 
-%
-It now becomes clear that looking at the diffusion maps coordinates {\em modulo} higher harmonics is essential for extracting an informative parameterization of the data, as well as characterizing the dimensionality of the data.
-%
-From Figures~\ref{fig:chemotaxis_simulations_harmonics}(b-c), we can see that the eigenvalue spectrum does not exhibit any large spectral gaps.
-%
-However, there is a significant gap between the eigenvalues which correspond to new directions in the data, and the corresponding eigenvectors provide us with a two-dimensional parameterization of the data which is consistent with the expected macroscopic dynamics.
-
-Figure~\ref{fig:chemotaxis_compare_timescales_evals} shows the diffusion maps eigenvalue ratio as a function of $\lambda$. 
-%
-For this specific example, we can compare the empirically detected change in dimensionality with the analytically-predicted shift. 
+
+Figure~\ref{fig:chemotaxis_compare_timescales_evals} shows the eigenvalue ratio in \eqref{eq:chemotaxis_eval_ratio} as a function of $\lambda$. 
+%
+We observe changes in the estimated dimensionality as a function of $\lambda$.
+%
+%We therefore expect the observed dimensionality to change from two to one as a function of $\lambda$. 
+%
+%For general dynamical systems, the governing macroscopic variables and asymptotic behavior may not be immediately obvious, and instead need to be inferred from the microscopic data.
+%
+For this specific example, these changes are consistent with the analytically-predicted shift. 
 %

### 2015-03-07 chemotaxis/nonuniform_sampling_example.m

@@ -1,60 +1,0 @@
-clear all
-close all
-
-%%
-[X, Y] = meshgrid(1:10, 1:100);
-X = (X - mean(X(:)));
-Y = (Y - mean(Y(:)))/100;
-
-data = [X(:) Y(:)];
-
-figure;
-
-subplot(2,3,1)
-plot(data(:,1),data(:,2),'.')
-axis equal
-title('data')
-
-%%
-
-[V, D] = PCA(data, 2);
-
-subplot(2,3,2)
-scatter(data(:,1),data(:,2),50, data*V(:,1),'.')
-axis equal
-cbar = colorbar('peer',gca); 
-set(get(cbar,'xlabel'),'String','PC 1');
-title('PCA')
-
-subplot(2,3,3)
-scatter(data(:,1),data(:,2),50, data*V(:,2),'.')
-axis equal
-cbar = colorbar('peer',gca); 
-set(get(cbar,'xlabel'),'String','PC 2');
-title('PCA')
-
-%%
-W = squareform(pdist(data)).^2;
-eps = median(W(:));
-[V, D] = dmaps(W, eps, 10);
-
-subplot(2,3,4)
-scatter(data(:,1),data(:,2),50, V(:,2),'.')
-axis equal
-cbar = colorbar('peer',gca); 
-set(get(cbar,'xlabel'),'String','\phi_1');
-title('DMAPS')
-
-subplot(2,3,5)
-scatter(data(:,1),data(:,2),50, V(:,3),'.')
-axis equal
-cbar = colorbar('peer',gca); 
-set(get(cbar,'xlabel'),'String','\phi_2');
-title('DMAPS')
-
-subplot(2,3,6)
-scatter(data(:,1),data(:,2),50, V(:,6),'.')
-axis equal
-cbar = colorbar('peer',gca); 
-set(get(cbar,'xlabel'),'String','\phi_5');
-title('DMAPS')

### 2015-03-07 chemotaxis/reorder_DMAPS_noharmonics.m

@@ -1,14 +1,0 @@
-function [V2, D2] = reorder_DMAPS_noharmonics(V, D)
-
-D2 = D;
-n = size(D, 2);
-
-for i=3:n
-    [~, res] = local_linear_regression(V(:,i), V(:, 2:i-1));
-    D2(i,i) = D(i,i) * res;
-end
-
-[~, I] = sort(abs(diag(D2)), 'descend');
-
-V2 = V(:,I);
-D2 = D2(I,I);

### 2015-03-07 chemotaxis/simulate_and_calculate_embedding.m

@@ -1,84 +1,0 @@
-function [V2, D2] = simulate_and_calculate_embedding(lambda, s)
-
-
-% number of particles
-N = 1000;
-% time step
-dt = 1;
-% maximum simulation time
-tmax = 10;
-nsteps = floor(tmax / dt) + 1;
-
-%initial probility of a particle moving to the right (vel=+1)
-nsim = 10;
-pinit = linspace(0.1, 0.9, nsim);
-
-% histogram parameters
-nbins = 32;
-x_hist = linspace(-s*tmax, s*tmax, nbins);
-hist_all = zeros(nbins, nsteps*nsim);
-hist_left = zeros(nbins, nsteps*nsim);
-hist_right = zeros(nbins, nsteps*nsim);
-all_time = zeros(1, nsteps*nsim);
-all_p = zeros(1, nsteps*nsim);
-
-%%
-
-for j=1:nsim
-    % simulate velocity jump process
-    [pos, vel, time] = simulate_jumps(N, tmax, dt, lambda, s, pinit(j));
-    %[pos, vel, time] = simulate_jumps_unifdist(N, tmax, dt, lambda, s, pinit(j), min_start, max_start);
-    
-    % histogram of all particle positions as a function of time
-    for i=1:nsteps
-        
-        [hist_all(:,(j-1)*nsteps+i), ~] = hist(pos(:,i),x_hist);
-        
-        idx = (vel(:,i) == -1);
-        [hist_left(:,(j-1)*nsteps+i), ~] = hist(pos(idx,i),x_hist);
-        
-        idx = (vel(:,i) == 1);
-        [hist_right(:,(j-1)*nsteps+i), ~] = hist(pos(idx,i),x_hist);
-        
-    end
-    
-    % store times and probabilities
-    all_time((j-1)*nsteps+1:j*nsteps) = time;
-    all_p((j-1)*nsteps+1:j*nsteps) = pinit(j);
-    
-end
-
-% use data after initial "relaxation" time
-idx = (all_time > 1);
-
-%% compute EMD
-
-W2 = zeros(nsteps*nsim);
-
-D_for_emd = pdist2(x_hist', x_hist');
-
-for i1=1:nsteps*nsim
-    for i2=1:i1-1
-        
[…]

### 2015-03-07 chemotaxis/simulate_jump_process.m

@@ -1,59 +1,0 @@
-clear all
-close all
-
-%% simulations
-
-% rate of switching velocity
-lambda = 10;
-% speed of particles
-s = 10;
-
-% number of particles
-N = 1000;
-% time step
-dt = 10;
-% maximum simulation time
-tmax = 100;
-%initial probility of a particle moving to the right (vel=+1)
-pinit = 0.1; 
-
-% simulate velocity jump process
-[pos, vel, time] = simulate_jumps(N, tmax, dt, lambda, s, pinit);
-
-%% plot results
-
-cmap = jet;
-cmap = cmap(floor(linspace(1, size(cmap,1), length(time))), :);
-
-nbins = 25;
-x_hist = linspace(-s*tmax, s*tmax, nbins);
-hist_all = zeros(nbins, length(time));
-hist_left = zeros(nbins, length(time));
-hist_right = zeros(nbins, length(time));
-
-
-% histogram of all particle positions as a function of time
-for i=1:length(time)
-    
-    [hist_all(:,i), ~] = hist(pos(:,i),x_hist);
-    
-    idx = (vel(:,i) == -1);
-    [hist_left(:,i), ~] = hist(pos(idx,i),x_hist);
-    
-    idx = (vel(:,i) == 1);
-    [hist_right(:,i), ~] = hist(pos(idx,i),x_hist);
-    
-end
-
-figure;
-set(gcf,'DefaultAxesColorOrder',cmap)
-plot(x_hist, hist_all)
-
-
-figure;
-set(gcf,'DefaultAxesColorOrder',cmap)
-plot(x_hist, hist_left)
-
-figure;
-set(gcf,'DefaultAxesColorOrder',cmap)
-plot(x_hist, hist_right)

### 2015-03-07 chemotaxis/simulate_jump_process_EMD.m

@@ -170,7 +170,6 @@ corr(V2(:,2),all_p(idx)')
 corr(V2(:,3),all_time(idx)')
 
 
-return
 %%
 
 num_hist = 6;

### 2015-03-07 chemotaxis/simulate_jump_process_EMD2.m

@@ -170,7 +170,6 @@ corr(V2(:,2),all_time(idx)')
 
 corr(V2(:,3),all_p(idx)')
 
-return
 %%
 
 num_hist = 6;

### 2015-03-07 chemotaxis/simulate_jump_process_EMD_calc_vol.m

@@ -1,62 +1,0 @@
-clear all
-close all
-
-rng(321);
-
-nsamples = 20;
-
-D = 1;
-lambda = linspace(1, 4000, nsamples);
-s = sqrt(lambda / D);
-
-neps = 100;
-eps_range = logspace(2, 10, neps);
-
-vol_data = zeros(nsamples, neps);
-
-for i=1:nsamples
-    W2 = simulate_samples(lambda(i), s(i));
-    for j=1:neps
-        vol_data(i,j) = sum(exp(-W2(:)/(2*eps_range(j))));
-    end
-end
-
-%%
-figure; 
-loglog(eps_range, vol_data)
-hold on
-loglog(eps_range, 0.01*eps_range)
-loglog(eps_range, 0.01*eps_range.^(1/2))
-
-%%
-
-slopes = zeros(nsamples);
-
-% figure;
-for i=1:nsamples
-    vol1 = vol_data(i,:);
-    idx = find(vol1 < (max(vol1)^0.6)*(min(vol1)^0.4) & vol1 > (min(vol1)^0.6)*(max(vol1)^0.4));
-%     loglog(eps_range, vol1, '.')
-%     hold on
-%     loglog(eps_range(idx), vol1(idx), '.r')
-%     hold off
-%     pause
-    P = polyfit(log(eps_range(idx)), log(vol1(idx)), 1);
-    slopes(i) = P(1);
-%     slopes(i,:) = (log(vol1(2:end))-log(vol1(1:end-1)))./(log(eps_range(2:end))-log(eps_range(1:end-1)));
-    
-end
-
-figure;
-plot(2*slopes)
-
-% figure;
-% plot(max(slopes,[],2))
-% 
-% figure;
-% for i=1:nsamples
-%     plot(slopes(i,:))
-%     pause
-% end
-
-

### 2015-03-07 chemotaxis/simulate_jump_process_EMD_detect_change2.m

@@ -1,89 +1,0 @@
-clear all
-close all
-
-rng(321);
-
-markersize = 500;
-
-%nsamples = 25;
-nsamples = 5;
-nreplicates = 5;
-
-D = 1;
-lambda = linspace(1, 2500, nsamples);
-s = sqrt(lambda / D);
-neigs = 4;
-
-
-D_all = zeros(nreplicates, nsamples, neigs);
-D2_all = zeros(nreplicates, nsamples, neigs);
-res = zeros(nreplicates, nsamples, neigs);
-for i=1:nsamples
-    for j=1:nreplicates
-        [V_tmp, D_tmp] = simulate_and_calculate_embedding(lambda(i), s(i));
-        D_all(j,i,:) = diag(D_tmp(1:neigs,1:neigs));
-        [V2, D2] = reorder_DMAPS_noharmonics(V_tmp(:,1:neigs), D_tmp(1:neigs,1:neigs));
-        D2_all(j,i,:) = diag(D2);
-        res(j,i,:) = compute_residuals_DMAPS(V_tmp(:,1:neigs), D_tmp(1:neigs,1:neigs));
-
-    end
-end
-
-%%
-figure;
-errorbar(lambda, mean(D_all(:,:,2)), std(D_all(:,:,2)))
-hold on
-errorbar(lambda, mean(D_all(:,:,3)), std(D_all(:,:,3)))
-errorbar(lambda, mean(D_all(:,:,4)), std(D_all(:,:,4)))
-xlabel('\lambda')
-ylabel('eigenvalues')
-
-figure;
-errorbar(lambda, mean(D2_all(:,:,2)), std(D2_all(:,:,2)))
-hold on
-errorbar(lambda, mean(D2_all(:,:,3)), std(D2_all(:,:,3)))
-errorbar(lambda, mean(D2_all(:,:,4)), std(D2_all(:,:,4)))
-xlabel('\lambda')
-ylabel('adjusted eigenvalues')
-
-figure;
-plot(lambda, mean(D_all(:,:,2)) - mean(D_all(:,:,3)))
-
-figure;
-plot(lambda, mean(D2_all(:,:,2)) - mean(D2_all(:,:,3)))
-
-% save('simulate_jump_process_EMD_detect_change_data')
-%
-% %%
-% figure;
-% set(gcf, 'papersize', [8.6 4.3], 'units','centimeters')
-% set(gcf, 'paperposition', [0 0 8.6 4.3], 'units','centimeters')
[…]

### 2015-03-07 chemotaxis/simulate_jump_process_varyp.m

@@ -1,303 +1,0 @@
-clear all
-close all
-
-%% simulations
-
-% rate of switching velocity
-nsiml = 10;
-lambda = logspace(1,3,nsiml);
-%lambda = linspace(10,1000,nsiml);
-% speed of particles
-s = 10;
-
-% number of particles
-N = 1000;
-% time step
-dt = 1;
-% maximum simulation time
-tmax = 10;
-nsteps = floor(tmax / dt) + 1;
-
-%initial probility of a particle moving to the right (vel=+1)
-nsimp = 10;
-pinit = linspace(0.1, 0.9, nsimp); 
-
-% histogram parameters
-nbins = 128;
-x_hist = linspace(-s*tmax, s*tmax, nbins);
-hist_all = zeros(nbins, nsteps*nsimp*nsiml);
-hist_left = zeros(nbins, nsteps*nsimp*nsiml);
-hist_right = zeros(nbins, nsteps*nsimp*nsiml);
-all_time = zeros(1, nsteps*nsimp*nsiml);
-all_p = zeros(1, nsteps*nsimp*nsiml);
-all_lambda = zeros(1, nsteps*nsimp*nsiml);
-
-subplot_dim1 = floor(sqrt(nsimp));
-subplot_dim2 = ceil(nsimp / subplot_dim1);
-
-figure;
-for k=1:nsiml
-    for j=1:nsimp
-        % simulate velocity jump process
-        [pos, vel, time] = simulate_jumps(N, tmax, dt, lambda(k), s, pinit(j));
-
-        % histogram of all particle positions as a function of time
-        for i=1:nsteps
-
-            [hist_all(:,(k-1)*nsteps*nsimp+(j-1)*nsteps+i), ~] = hist(pos(:,i),x_hist);
-
-            idx = (vel(:,i) == -1);
-            [hist_left(:,(k-1)*nsteps*nsimp+(j-1)*nsteps+i), ~] = hist(pos(idx,i),x_hist);
-
-            idx = (vel(:,i) == 1);
-            [hist_right(:,(k-1)*nsteps*nsimp+(j-1)*nsteps+i), ~] = hist(pos(idx,i),x_hist);
-
-        end
-
-        % store times and probabilities
-        all_time((k-1)*nsteps*nsimp+(j-1)*nsteps+1:(k-1)*nsteps*nsimp+j*nsteps) = time;
[…]

### 2015-03-07 chemotaxis/simulate_jumps_unifdist.m

@@ -1,59 +1,0 @@
-% this function siulates the velocity jump process from  "The diffusion
-% limit of transport equations deroived from velocity-jump processes," 
-% Hillen and Othmer, SIAM J. Applied Math, 2000
-% N is the number of particles
-% tmax is the time to simulate
-% dt is the time step of simulation
-% lambda is the rate of switching of left/right velocities
-% s is the speed (constant)
-% pos stores the positions of all the particles at each timestep (N x nsteps matrix)
-% vel stores the velocities (+/-1) of all the particles at each timestep (N x nsteps matrix)
-% time stores the times at each timestep
-% pinit is the initial probability of a particle having velocity +1
-function [pos, vel, time] = simulate_jumps_unifdist(N, tmax, dt, lambda, s, pinit, dist_min, dist_max)
-
-%% arrays to store times, positions, and velocities
-time = 0:dt:tmax;
-pos = zeros(N, length(time));
-vel = zeros(N, length(time));
-
-%% keep track of current time, position, and velocity during iterations
-
-% start time at 0
-ti = 0;
-% start position at uniform distribution
-posi = unifrnd(dist_min, dist_max, N,1);
-% intial velocities are  +/-1 with probability pinit of +1
-veli =  2 * binornd(1,pinit, [N 1]) - 1;
-
-% store initial configurations
-pos(:,1) = posi;
-vel(:,1) = veli;
-
-%% simulate/update positions of particles, change velocity of random particle every time an "event" occurs
-
-% time of next event (time between events is exponentially distributed)
-tevent = ti + exprnd(1/lambda);
-
-% update position of particles at each timestep
-for i=2:length(time)
-    % if event occurs before next timestep, update pos/vel and draw
-    % another event
-    while time(i) > tevent
-        % update positions
-        posi = posi + veli * s * (tevent - ti);
-        % switch velocity of random particle
-        switch_idx = randi(N, 1);
-        veli(switch_idx) = -veli(switch_idx);
-        % update current time of simulation
-        ti = tevent;
-        % draw new event
-        tevent = ti + exprnd(1/lambda);
-    end
-    % update positions/velocities to current timestep
-    posi = posi + veli * s * (time(i) - ti);
-    pos(:,i) = posi;
-    vel(:,i) = veli;
-    % update current time
-    ti = time(i);
-end 

### 2015-03-07 chemotaxis/simulate_samples.m

@@ -1,70 +1,0 @@
-function W2 = simulate_samples(lambda, s)
-
-
-% number of particles
-N = 1000;
-% time step
-dt = 1;
-% maximum simulation time
-tmax = 10;
-nsteps = floor(tmax / dt) + 1;
-
-%initial probility of a particle moving to the right (vel=+1)
-nsim = 10;
-pinit = linspace(0.1, 0.9, nsim);
-
-% histogram parameters
-nbins = 32;
-x_hist = linspace(-s*tmax, s*tmax, nbins);
-hist_all = zeros(nbins, nsteps*nsim);
-hist_left = zeros(nbins, nsteps*nsim);
-hist_right = zeros(nbins, nsteps*nsim);
-all_time = zeros(1, nsteps*nsim);
-all_p = zeros(1, nsteps*nsim);
-
-%%
-
-for j=1:nsim
-    % simulate velocity jump process
-    [pos, vel, time] = simulate_jumps(N, tmax, dt, lambda, s, pinit(j));
-    %[pos, vel, time] = simulate_jumps_unifdist(N, tmax, dt, lambda, s, pinit(j), min_start, max_start);
-    
-    % histogram of all particle positions as a function of time
-    for i=1:nsteps
-        
-        [hist_all(:,(j-1)*nsteps+i), ~] = hist(pos(:,i),x_hist);
-        
-        idx = (vel(:,i) == -1);
-        [hist_left(:,(j-1)*nsteps+i), ~] = hist(pos(idx,i),x_hist);
-        
-        idx = (vel(:,i) == 1);
-        [hist_right(:,(j-1)*nsteps+i), ~] = hist(pos(idx,i),x_hist);
-        
-    end
-    
-    % store times and probabilities
-    all_time((j-1)*nsteps+1:j*nsteps) = time;
-    all_p((j-1)*nsteps+1:j*nsteps) = pinit(j);
-    
-end
-
-% use data after initial "relaxation" time
-idx = (all_time > 1);
-
-%% compute EMD
-
-W2 = zeros(nsteps*nsim);
-
-for i1=1:nsteps*nsim
-    for i2=1:i1-1
-        
-        %[emd_dist, ~]= emd_hat_gd_metric_mex(hist_all(:,i1), hist_all(:,i2), D_for_emd);
-        emd_dist = sum(abs(cumsum(hist_all(:,i1)) - cumsum(hist_all(:,i2))));
-        
[…]

### 2015-03-07 chemotaxis/swissroll.m

@@ -1,334 +1,0 @@
-clear all
-close all
-
-%%
-% 
-% N = 1500; % number of points considered
-% t = rand(1,N);
-% t = sort(4*pi*sqrt(t))'; 
-% 
-% h = 15*pi;
-% 
-% z = h*rand(N,1); % random heights
-% x = (t+.1).*cos(t);
-% y = (t+.1).*sin(t);
-% data = [x,y,z]; % data
-% 
-% %%
-% figure;
-% plot3(x,y,z,'.')
-% 
-% axis equal
-% 
-% 
-% %%
-% 
-% W = squareform(pdist(data)).^2;
-% eps = 5;
-% 
-% neigs = 10;
-% 
-% [V, D] = dmaps(W, eps, neigs);
-% 
-% 
-% %%
-% figure;
-% scatter3(x,y,z,50,V(:,2),'.')
-% 
-% figure;
-% scatter3(x,y,z,50,V(:,3),'.')
-% 
-% %%
-% 
-% eps_med_scale = 3;
-% res = compute_residuals_DMAPS(V, eps_med_scale);
-% 
-% figure;
-% plot(res, 'o-')
-
-%%
-
-load swissroll1.mat
-
-make_fig(2,2);
-plot3(x,y,z,'.', 'markersize', 10)
-view(-20, 80)
-xlabel('z_1')
-ylabel('z_2')
-zlabel('z_3')
-axis equal
-grid on
-print('swissroll1.eps','-depsc')
-
-% 
-% figure;
-% bar(res(2:end))
-% ylabel('cross-validation error')
-% print('swissroll1_cv.eps','-depsc')
-% 
-make_fig(3,2);
-scatter3(x,y,z,50,sign(corr(V(:,2),t))*V(:,2),'.')
-view(-20, 80)
-xlabel('z_1')
-ylabel('z_2')
-zlabel('z_3')
-axis equal
-grid on
-cbar = colorbar('peer',gca); 
-set(get(cbar,'xlabel'),'String','\phi_1');
-print('swissroll1_color1.eps','-depsc')
-
-make_fig(3,2);
-scatter3(x,y,z,50,sign(corr(V(:,3),z))*V(:,3),'.')
-view(-20, 80)
-xlabel('z_1')
-ylabel('z_2')
-zlabel('z_3')
-axis equal
-grid on
-cbar = colorbar('peer',gca); 
-set(get(cbar,'xlabel'),'String','\phi_2');
[…]

### 2015-03-07 chemotaxis/test.m

@@ -1,37 +1,17 @@
 clear all
 close all
 
-rng(321);
-
+rng(123);
 markersize = 500;
+
 
 %% simulations
 
-% % rate of switching velocity
-% lambda = 0.01;
-% % speed of particles
-% s = 0.1;
 
-% % rate of switching velocity
-% lambda = 1;
-% % speed of particles
-% s = 1;
-
-% % rate of switching velocity
-% lambda = 400;
-% % speed of particles
-% s = 20;
-
-% % rate of switching velocity
-% lambda = 2500;
-% % speed of particles
-% s = 50;
-
-% % rate of switching velocity
-% lambda = 4900;
-% % speed of particles
-% s = 70;
-
+% rate of switching velocity
+lambda = 1000;
+% speed of particles
+s = sqrt(lambda);
 
 % number of particles
 N = 1000;
@@ -49,24 +29,20 @@ min_start = -10;
 max_start = 10;
 
 % histogram parameters
+%nbins = 128;
 nbins = 32;
+x_hist = linspace(min_start-s*tmax, max_start+s*tmax, nbins);
 hist_all = zeros(nbins, nsteps*nsim);
 hist_left = zeros(nbins, nsteps*nsim);
 hist_right = zeros(nbins, nsteps*nsim);
 all_time = zeros(1, nsteps*nsim);
 all_p = zeros(1, nsteps*nsim);
-mean_pos = zeros(1, nsteps*nsim);
-mean_vel = zeros(1, nsteps*nsim);
-
-figure;
-
-for lambda = [1 400 2500 4900]
-    s = sqrt(lambda);
-    
-    x_hist = linspace(min_start-s*tmax, max_start+s*tmax, nbins);
 
 %%
+subplot_dim1 = floor(sqrt(nsim));
+subplot_dim2 = ceil(nsim / subplot_dim1);
 
+figure;
 for j=1:nsim
     % simulate velocity jump process
     [pos, vel, time] = simulate_jumps(N, tmax, dt, lambda, s, pinit(j));
@@ -83,20 +59,28 @@ for j=1:nsim
         idx = (vel(:,i) == 1);
         [hist_right(:,(j-1)*nsteps+i), ~] = hist(pos(idx,i),x_hist);
         
-        mean_pos((j-1)*nsteps+i) = mean(pos(:,i));
[…]

### 2015-03-07 chemotaxis/test_eval_spectrum_variations.m

@@ -1,277 +1,0 @@
-function test_eval_spectrum_variations
-
-clear all
-close all
-
-%% strips of different ratios with uniform points
-
-ymax = 1;
-
-for xmax = [2 4 8]
-    
-    spacing = 0.025*xmax;
-    [X, Y] = meshgrid(0:spacing:xmax, 0:spacing:ymax);
-    data = [X(:) Y(:)];
-    
-    
-    W = squareform(pdist(data)).^2;
-    eps = median(W(:));
-    
-    [V, D] = dmaps(W, eps, 15);
-    
-    eps_med_scale = 3;
-    res = compute_residuals_DMAPS(V, eps_med_scale);
-    
-    make_plot(data, V, D, res)
-
-end
-
-%% strips of different ratios with uniformly random points
-
-n = 500;
-ymax = 1;
-
-rng(12321);
-
-for xmax = [2 4 8]
-    
-    data = rand(n, 2);
-    data(:,1) = data(:,1) * xmax;
-    
-
-    
-    W = squareform(pdist(data)).^2;
-    eps = median(W(:));
-    
-    [V, D] = dmaps(W, eps, 15);
-    
-    eps_med_scale = 3;
-    res = compute_residuals_DMAPS(V, eps_med_scale);
-
-        make_plot(data, V, D, res)
-
-end
-
-%% strips of different ratios with points dense in x
-
-ymax = 1;
-spacingx = 0.05;
-spacingy = 0.5;
-
-for xmax = [2 4 8]
-    
-    [X, Y] = meshgrid(0:spacingx:xmax, 0:spacingy:ymax);
-    data = [X(:) Y(:)];
-
-    
-    W = squareform(pdist(data)).^2;
-    eps = median(W(:));
-    
-    [V, D] = dmaps(W, eps, 15);
-    
-    eps_med_scale = 3;
-    res = compute_residuals_DMAPS(V, eps_med_scale);
-    
-        make_plot(data, V, D, res)
-
-    
-end
-
-%% strips of different ratios with points dense in x; alpha = 1
-
-alpha = 1;
-
-ymax = 1;
-spacingx = 0.05;
-spacingy = 0.5;
-
-for xmax = [2 4 8]
-    
-    
-    [X, Y] = meshgrid(0:spacingx:xmax, 0:spacingy:ymax);
-    data = [X(:) Y(:)];
-    
-    
-    W = squareform(pdist(data)).^2;
-    eps = median(W(:));
-    
[…]

### 2015-03-07 chemotaxis/test_sum_evals.m

@@ -1,193 +1,0 @@
-clear all
-close all
-
-%%
-
-scale_vec = repmat(2:10, 1 ,3);
-
-lambda_ratio = zeros(size(scale_vec));
-
-n = 500;
-
-figure;
-for i=1:length(scale_vec)
-    
-    scale = scale_vec(i);
-    
-    
-    data = rand(n, 2);
-    data(:,1) = data(:,1) * scale;
-    
-    %     figure;
-    %     scatter(data(:,1),data(:,2))
-    %     axis equal
-    
-    W = squareform(pdist(data)).^2;
-    eps = median(W(:))/9;
-    
-    [V, D] = dmaps(W, eps, 15);
-    
-    eps_med_scale = 3;
-    res = compute_residuals_DMAPS(V, eps_med_scale);
-    
-    %     figure;
-    %     make_colored_bars(diag(D(2:end, 2:end)), res(2:end))
-    %     xlabel('k')
-    %     ylabel('\mu_k')
-    
-    %     idx = find(res > 0.2);
-    %     D(idx, idx);
-    [~, idx] = sort(res, 'descend');
-    idx = idx(1:2);
-    if idx(1) > idx(2)
-        tmp = idx(2);
-        idx(2) = idx(1);
-        idx(1) = tmp;
-    end
-    
-    subplot(length(scale_vec), 2, 2*i-1);
-    plot(data(:,1),V(:,idx(1)),'.')
-    
-    subplot(length(scale_vec), 2, 2*i);
-    plot(data(:,2),V(:,idx(2)),'.')
-    
-    lambda_ratio(i) = 1/sqrt(log(D(idx(1), idx(1)))/log(D(idx(2), idx(2))));
-    
-end
-
-figure;
-plot(scale_vec, lambda_ratio,'o');
-hold on
-
-%%
-scale_vec = repmat(2:10, 1 ,3);
-
-lambda_ratio = zeros(size(scale_vec));
-
-n = 500;
-
-figure;
-for i=1:length(scale_vec)
-    
-    scale = scale_vec(i);
-    
-    
-    data = rand(n, 2);
-    data(:,1) = data(:,1) * scale;
-    
-    %     figure;
-    %     scatter(data(:,1),data(:,2))
-    %     axis equal
-    
-    W = squareform(pdist(data)).^2;
-    eps = median(W(:))/9;
-    
[…]

### 2015-03-07 chemotaxis/test_torus.m

@@ -13,24 +13,22 @@ neigs = 20;
 
 %%
 
-% r1 = 3;
-% r2 = 1;
-% 
-% x = (r1 + r2 * cos(theta2)) .* cos(theta1);
-% y = (r1 + r2 * cos(theta2)) .* sin(theta1);
-% z = r2 * sin(theta2);
-% 
-% data = [x y z];
-% W = squareform(pdist(data));
-% eps = median(W(:))/3;
-% 
-% [V, D] = dmaps(W.^2, eps.^2, neigs);
-% 
-% eps_med_scale = 3;
-% res = compute_residuals_DMAPS(V, eps_med_scale);
-% 
-% save('torus1.mat')
-load('torus1.mat')
+r1 = 3;
+r2 = 1;
+
+x = (r1 + r2 * cos(theta2)) .* cos(theta1);
+y = (r1 + r2 * cos(theta2)) .* sin(theta1);
+z = r2 * sin(theta2);
+
+data = [x y z];
+W = squareform(pdist(data));
+eps = median(W(:))/3;
+
+[V, D] = dmaps(W.^2, eps.^2, neigs);
+
+eps_med_scale = 3;
+res = compute_residuals_DMAPS(V, eps_med_scale);
+
 
 %
 make_fig(3,2);
@@ -43,30 +41,6 @@ axis equal
 grid on
 print('torus1.eps','-depsc')
 
-% make_fig(3,3);
-% scatter3(x,y,z,50,sign(corr(V(:,2),t))*V(:,2),'.')
-% view(-20, 75)
-% xlabel('x')
-% ylabel('y')
-% zlabel('z')
-% axis equal
-% grid on
-% cbar = colorbar('peer',gca); 
-% set(get(cbar,'xlabel'),'String','\phi_1');
-% print('swissroll1_color1.eps','-depsc')
-% 
-% make_fig(3,3);
-% scatter3(x,y,z,50,sign(corr(V(:,3),z))*V(:,3),'.')
-% view(-20, 75)
-% xlabel('x')
-% ylabel('y')
-% zlabel('z')
-% axis equal
-% grid on
-% cbar = colorbar('peer',gca); 
-% set(get(cbar,'xlabel'),'String','\phi_2');
-% print('swissroll1_color2.eps','-depsc')
-
 make_fig(3,2);
 make_colored_bars(diag(D(2:end,2:end)), res(2:end))
 set(gca, 'xtick', (6:5:size(D,1))-1)
[…]

### 2015-03-11 chemotaxis/paper/draft1.tex
transition simulations, etc.

@@
 However, both the advection procedure and the successive eigendecompositions are very expensive and intractable for larger data set. 
-TODO: find other advantages of our proposed method. 
+TODO: elaborate on their method and find other advantages of our proposed method (for example, theirs is ad hoc). 
 %
@@
 %
-If the resulting fit is accurate, then we assume $\phi_{k}$ is a repeated eigendirection of $\phi_1, \dots, \phi_{k-1}$. 
+Our approach is motivated by simple trigonometric arguments, such as $\cos2x$ can be written as a function of $\cos x$. 
+%
+Thus, if the resulting fit is accurate, then we assume $\phi_{k}$ is a repeated eigendirection of $\phi_1, \dots, \phi_{k-1}$. 
 %
@@
 %
-The coefficients $\alpha$ and $\beta$ are functions of $\Phi_{k-1}$ because we use a {\em local} linear fit, and so the coefficients change as a function of the domain.
+The coefficients $\alpha$ and $\beta$ are functions of $\Phi_{k-1}$ because we use a {\em local} linear fit in the $k-1$-dimensional $\Phi_{k-1}$ space, and so the coefficients change as a function of the domain.
 
@@
 %
-Note that a small value of $r_k$ implies that $\phi_{k}$ can be accurately approximated from $\phi_1, \dots, \phi_{k-1}$, and is therefore a harmonic of a previous mode.
+Note that a small value of $r_k$ implies that $\phi_{k}$ can be accurately approximated from $\phi_1, \dots, \phi_{k-1}$.
+We will assume it means that $\phi_k$ is a harmonic of previous modes, i.e., a repeated eigendirction.
 %
@@
 %
-Because $r_{k} \approx 1$ if $\phi_{k}$ parameterizes a new direction, we choose to define $r_1 = 1$.
+We choose to set $r_1 = 1$.
 
@@
 
-Once the eigenvectors which parameterize unique eigendirections along the manifold have been determined, the corresponding eigenvalues can be used to calculate the relative lengths along these directions.
+For the strip example, once the eigenvectors which parameterize unique eigendirections along the manifold have been determined, the corresponding eigenvalues can be used to calculate the relative lengths along these directions.
+%
+For more general manifolds, we believe the eigenvalues can still provide a measure of the relative significance of the unique eigendirections.
 %
@@
 %
-From \eqref{eq:evals} and \eqref{eq:evals_relationship}, we propose to approximate the relative lengths $L_j$  along the manifold by
+For the strip example, from \eqref{eq:evals} and \eqref{eq:evals_relationship}, we propose to approximate the relative lengths $L_j$  along the manifold by
 \begin{equation} \label{eq:est_lengths}
@@
 The empirical eigenvalues are plotted in blue, and the analytical eigenvalues are plotted in red.
-From the eigenvalues which are identified as important/non-harmonics, the estimated length ratio is (a) 2.0, (b) 4.4, (c) 8.8.
-TODO: normalize colors}
+From the eigenvalues which are identified as important/non-harmonics, the estimated length ratio is (a) 2.2, (b) 4.1, (c) 8.7.}
 \label{fig:strip_compare_analytic}
@@
 
-Our model problem describes the process of cellular chemotaxis \cite{othmer2000diffusion}, where biological cells work together and exhibit coherent macroscopic dynamics regulated by extracellular sensed signals in order to accomplish tasks such as finding food or navigating away from toxins.
+Our model problem describes the process of cellular chemotaxis \cite{othmer2000diffusion}, where biological cells exhibit coherent macroscopic dynamics regulated by extracellular sensed signals in order to accomplish tasks such as finding food or navigating away from toxins.
 %
@@
 %Third, it has an analytic macroscopic description, which serves as a ground truth and will allow us to verify our results which are obtained in an unsupervised manner.
-This specific example has an analytic macroscopic description in which the macroscopic dynamics of the group of cells vary depending on the value of a single system parameter.
+This specific example has an analytic macroscopic description in which the macroscopic dynamics of the group of cells depend on the value of a single system parameter.
 %
@@
 
-The microscopic model consists of a collection of $N$ cells whose states are defined by their positions and velocities on a line, and the dynamics of each cell are governed by a stochastic process.
+The microscopic model consists of a collection of $N$ cells (for our simulations, we take $N=1000$) whose states are defined by their positions and velocities on a line, and the dynamics of each cell are governed by a stochastic process.
 %
@@
 %
+For our specific simulations, we set $s^2/\lambda=1$, which is consistent with the analysis presented in \cite{...}. 
+%
 We note that we have chosen a very specific one-parameter family of initial conditions which lead to simple dynamics and allow us to illustrate of our main points;
@@
 We note that although histograms are not essential for using EMD, histograms make the computation feasible. 
+%
+For our specific simulations, we will take $n=32$.
 
@@
 \begin{figure}[t!]
-\centering
-\begin{subfigure}{5cm}
+\def \figwidth {3.5cm}
+
+\centering
+\begin{subfigure}{\figwidth}
 \includegraphics[width=\textwidth]{EMD_withhist_p_1}
@@
 \end{subfigure}
-\begin{subfigure}{5cm}
+\begin{subfigure}{\figwidth}
 \includegraphics[width=\textwidth]{EMD_withhist_t_1}
@@
 \end{subfigure}
-\begin{subfigure}{5cm}
+\begin{subfigure}{\figwidth}
+\includegraphics[width=\textwidth]{EMD_withhist_rho_1}
+\caption{}
+\label{subfig:small_lambda_rho}
+\end{subfigure}
+
+\begin{subfigure}{\figwidth}
 \includegraphics[width=\textwidth]{EMD_withhist_p_400}
@@
 \end{subfigure}
-\begin{subfigure}{5cm}
+\begin{subfigure}{\figwidth}
 \includegraphics[width=\textwidth]{EMD_withhist_t_400}
@@
 \end{subfigure}
-\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with (a,b) $\lambda=1$, $s=1$, and (c,d) $\lambda=400$, $s=20$.
-For each set, we run many simulations, each with $N=1000$ particles.
-%
-We vary the initial conditions such that $0.1 \le p  \le 0.9$, and allow each simulation to evolve for $10$ time units.
-%
-We discard the initial portion of each simulation trajectory ($t < 1$), which corresponds to the initial relaxation of the system.
-%
-The observation $z(i) \in \mathbb{R}^n$ is the histogram of $x_1(i), x_2(i), \dots, x_N(i)$ into $n = 32$ bins.
-%
+\begin{subfigure}{\figwidth}
+\includegraphics[width=\textwidth]{EMD_withhist_rho_400}
+\caption{}
+\label{subfig:large_lambda_rho}
+\end{subfigure}
+\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with (a--c) $\lambda=1$, $s=1$, and (d--f) $\lambda=400$, $s=20$.
+Each data set consists of $10$ simulations, with initial conditions uniformly chosen such that $0.1 \le p  \le 0.9$.
+We allow each simulation to evolve for $10$ time units, with $dt=1$, and use data with $t > 0$ for analysis.
 The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions.
-The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time.
-Representative histograms are shown for selected data points.}
+The data are colored by (a, d) $p$, the initial probability of a particle moving to the right, (b, e) $t$, time.
+Representative histograms are shown for selected data points, and (c, f) $E \rho^+ - E \rho^-$, the difference between the average position of the left- and right-moving particles.}
 \label{fig:dmaps_embed_emd}
@@
 %
-\caption{Diffusion maps embedding for three different chemotaxis simulations.
+\caption{Diffusion maps embedding for three different chemotaxis simulation data sets.
 (a) $\lambda = 100$, $s = 10$.
-(b) $\lambda = 2500$, $s = 50$.
+(b) $\lambda = 1600$, $s = 40$.
 (c) $\lambda = 6400$, $s = 80$.
-For each simulation, the eigenvalue spectrum is shown.
+Each data set consists of $10$ simulations, with initial conditions uniformly chosen such that $0.1 \le p  \le 0.9$.
+We allow each simulation to evolve for $10$ time units, with $dt=1$, and use data with $t > 0$ for analysis.
+The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions.
+For each data set, the eigenvalue spectrum, colored by the cross-validation error, is shown.
 From the spectra, we can see that the first two components are informative for (a), components 1 and 3 are informative for (b), and components 1 and 4 are informative for (c).
 The corresponding embeddings are shown below.
-For comparison, the embeddings using the first two components are also shown in (b) and (c).
-TODO: change (a) to larger $\lambda$ value so that the correspondences are the same across figures.}
+For comparison, the embeddings using the first two components are also shown in (b) and (c).}
 %
@@
 %
-One set of simulations (Figures~\ref{subfig:small_lambda_p}~and~\ref{subfig:small_lambda_t}) is for a small value of $\lambda$, and the other set (Figures~\ref{subfig:large_lambda_p}~and~\ref{subfig:large_lambda_t}) is for a large value of $\lambda$.
+One set of simulations (Figures~\ref{subfig:small_lambda_p}--\ref{subfig:small_lambda_rho}) is for a small value of $\lambda$, and the other set (Figures~\ref{subfig:large_lambda_p}--\ref{subfig:large_lambda_rho}) is for a large value of $\lambda$.
 %
@@
 %
-In both cases, $\phi_1$ is significantly more important than $\phi_2$, as indicated by the separation between the first two eigenvalues (for the small $\lambda$ case, $\mu_1 = 0.5138$, and $\mu_2 = 0.3943$, and for the large $\lambda$ case, $\mu_1 = 0.5543$, and $\mu_2 = 0.3807$).
+%In both cases, $\phi_1$ is significantly more important than $\phi_2$, as indicated by the separation between the first two eigenvalues (for the small $\lambda$ case, $\mu_1 = 0.4828$, and $\mu_2 = 0.4287$, and for the large $\lambda$ case, $\mu_1 = 0.6288$, and $\mu_2 = 0.3236$).
 %
@@
 %
+TODO: should we mention fluxes here??
 
@@
 %
-However, there is a significant gap between the eigenvalues which correspond to new directions in the data, and the corresponding eigenvectors provide us with an informative two-dimensional parameterization of the data.
+However, there is a significant gap between the eigenvalues which correspond to new directions in the data (colored in blue), and the corresponding eigenvectors provide us with an informative two-dimensional parameterization of the data.
 %which is consistent with the expected macroscopic dynamics.
@@
 
-\subsubsection{Detecting changes in dynamical behavior}
-
-\begin{figure}[t] 
-\centering
-\includegraphics[width=0.5\textwidth]{detect_change_eigenvalues}
-\caption{The first two eigenvalues from our analysis as a function of $\lambda$.
+%\subsubsection{Detecting changes in dynamical behavior}
+%
+%
+%%\begin{figure}[t] 
+%%\centering
+%%\includegraphics[width=0.5\textwidth]{detect_change_eigenvalues}
+%%\caption{The first two eigenvalues from our analysis as a function of $\lambda$.
 Each point is averaged over $10$ experiments (as in Figure~\ref{fig:dmaps_embed_emd}, one experiment consists of data collected at varying $p$ and $t$, for a fixed value of $\lambda$), with error bars denoting one standard deviation.
-The convergence of the eigenvalues at $\lambda \approx 125$ indicates switching of the dynamical ``mode''.}
-\label{fig:detect_change}
-\end{figure}
-
-%We have seen that the eigenvalues $\mu_i$, which quantify the importance of each embedding coordinate, reveal the relative importance of the variables which govern the macroscopic dynamics.
-%
-We can use the eigenvalues from our analysis to detect changes in dynamical behavior.
-%
-From Figure~\ref{fig:dmaps_embed_emd}, we can see that the dominant eigendirection is correlated with $p$ when $\lambda$ is small, and correlated with $t$ when $\lambda$ is large. 
-%
-%We expect an intermediate $\lambda$ value where $p$ and $t$ are equally important. 
-%%
-%From the analytic macroscopic description, we expect a gap between the two eigenvalues at the two asymptotic regimes ($\lambda \rightarrow 0$, where the dynamics are described by two uncoupled wave equations \eqref{...}, and $\lambda \rightarrow \infty$, where the dynamics are described by a single heat equation \eqref{...}), as one of the two macroscopic variables is significantly more important to the dynamical behavior.
-%%
-%For intermediate $\lambda$ values, we expect the gap between $|\mu_1|$ and $|\mu_2|$ to narrow, with the $\lambda$ value where $|\mu_1| \approx |\mu_2|$ corresponding to the transition between ``wave-like'' and ``heat-like'' behavior.
-%
-We performed a new set of simulations over a range of $\lambda$ values;  
-Figure~\ref{fig:detect_change} shows the first two eigenvalues, $\mu_1$ and $\mu_2$, from our analysis as a function of $\lambda$.
-%
-For small $\lambda$ (corresponding to Figures~\ref{subfig:small_lambda_p}~and~\ref{subfig:small_lambda_t}), the two eigenvalues are well-separated, as the first mode (which parameterizes $p$) is significantly more important than the second mode (which parameterizes $t$).
-%
-At intermediate $\lambda$ values, the eigenvalues come together as the system dynamics change and both $p$ and $t$ are of similar importance.
-%
-The eigenvalues then separate again at large $\lambda$ (corresponding to Figures~\ref{subfig:large_lambda_p}~and~\ref{subfig:large_lambda_t}), where the $t$ effects dominate the dynamics.
-%
-We can therefore detect changes in dynamical behavior using data-driven techniques, without any knowledge of the governing macroscopic model.
-
-%\begin{figure}[t]
-%\def\figwidth{4.2cm}
-%\begin{subfigure}{\figwidth}
-%\includegraphics[width=\textwidth]{rawhist_p_1}
-%\caption{}
-%\end{subfigure}
-%\begin{subfigure}{\figwidth}
-%\includegraphics[width=\textwidth]{rawhist_t_1}
-%\caption{}
-%\end{subfigure}
-%\begin{subfigure}{\figwidth}
-%\includegraphics[width=\textwidth]{rawhist_p_400}
-%\caption{}
-%\end{subfigure}
-%\begin{subfigure}{\figwidth}
-%\includegraphics[width=\textwidth]{rawhist_t_400}
-%\caption{}
-%\end{subfigure}
-%\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with (a,b) $\lambda=1$, $s=1$, and (c,d) $\lambda=400$, $s=20$.
+%%The convergence of the eigenvalues at $\lambda \approx 125$ indicates switching of the dynamical ``mode''.}
+%%\label{fig:detect_change}
+%%\end{figure}
+%
+%%We have seen that the eigenvalues $\mu_i$, which quantify the importance of each embedding coordinate, reveal the relative importance of the variables which govern the macroscopic dynamics.
+%%
+%We can use the eigenvalues from our analysis to detect changes in dynamical behavior.
+%%
+%From Figure~\ref{fig:dmaps_embed_emd}, we can see that the dominant eigendirection is correlated with $p$ when $\lambda$ is small, and correlated with $t$ when $\lambda$ is large. 
+%%
+%%We expect an intermediate $\lambda$ value where $p$ and $t$ are equally important. 
+%%%
+%%From the analytic macroscopic description, we expect a gap between the two eigenvalues at the two asymptotic regimes ($\lambda \rightarrow 0$, where the dynamics are described by two uncoupled wave equations \eqref{...}, and $\lambda \rightarrow \infty$, where the dynamics are described by a single heat equation \eqref{...}), as one of the two macroscopic variables is significantly more important to the dynamical behavior.
+%%%
+%%For intermediate $\lambda$ values, we expect the gap between $|\mu_1|$ and $|\mu_2|$ to narrow, with the $\lambda$ value where $|\mu_1| \approx |\mu_2|$ corresponding to the transition between ``wave-like'' and ``heat-like'' behavior.
+%%
+%We performed a new set of simulations over a range of $\lambda$ values;  
+%Figure~\ref{fig:detect_change} shows the first two eigenvalues, $\mu_1$ and $\mu_2$, from our analysis as a function of $\lambda$.
+%%
+%For small $\lambda$ (corresponding to Figures~\ref{subfig:small_lambda_p}~and~\ref{subfig:small_lambda_t}), the two eigenvalues are well-separated, as the first mode (which parameterizes $p$) is significantly more important than the second mode (which parameterizes $t$).
+%%
+%At intermediate $\lambda$ values, the eigenvalues come together as the system dynamics change and both $p$ and $t$ are of similar importance.
+%%
+%The eigenvalues then separate again at large $\lambda$ (corresponding to Figures~\ref{subfig:large_lambda_p}~and~\ref{subfig:large_lambda_t}), where the $t$ effects dominate the dynamics.
+%%
+%We can therefore detect changes in dynamical behavior using data-driven techniques, without any knowledge of the governing macroscopic model. 
+%%
+%In the next section, we will show that the detected changes in dynamical behavior is consistent with the analytic macroscale model.
+%
+%%\begin{figure}[t]
+%%\def\figwidth{4.2cm}
+%%\begin{subfigure}{\figwidth}
+%%\includegraphics[width=\textwidth]{rawhist_p_1}
+%%\caption{}
+%%\end{subfigure}
+%%\begin{subfigure}{\figwidth}
+%%\includegraphics[width=\textwidth]{rawhist_t_1}
+%%\caption{}
+%%\end{subfigure}
+%%\begin{subfigure}{\figwidth}
+%%\includegraphics[width=\textwidth]{rawhist_p_400}
+%%\caption{}
+%%\end{subfigure}
+%%\begin{subfigure}{\figwidth}
+%%\includegraphics[width=\textwidth]{rawhist_t_400}
+%%\caption{}
+%%\end{subfigure}
+%%\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with (a,b) $\lambda=1$, $s=1$, and (c,d) $\lambda=400$, $s=20$.
 The distances used in the diffusion maps kernel are the Euclidean distances between the histograms of particle positions.
 The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time.}
-%\label{fig:dmaps_embed_noemd}
-%\end{figure}
+%%\label{fig:dmaps_embed_noemd}
+%%\end{figure}
 
@@
 \begin{equation}
-2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2}.
-\end{equation}
+2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2},
+\end{equation}
+%
+where $D=s^2/\lambda$.
 %
@@
 Overall, in both cases, we obtain, in an unsupervised data-driven manner, an accurate picture of the macroscopic variables that govern the system dynamics.
-%
-TODO: address fluxes
+
+Furthermore, this analytic macroscopic descrption shows us that when $\lambda$ is small, the system dynamics can be described by the densities $\rho^+$ and $\rho^-$, and when $\lambda$ is large, the dynamics are described by $\rho = \rho^+ + \rho^-$.
+%
+Figures~\ref{subfig:small_lambda_rho}~and~\ref{subfig:large_lambda_rho} show the data, colored by the difference in the average position of the left- and right-moving particles. 
+%
+Clearly, for small $\lambda$, where both $\rho^+$ and $\rho^-$ are required to describe the dynamics, the data is organized by the differences in the two densities. 
+%
+However, for large $\lambda$, the two densities rapidly equilibrate, and no organization based on the differences between the two densities is visible.
 
@@
 %
-In contrast, the correlations between the embedding coordinates and the macroscopic variables when using EMD in the diffusion maps calculation are all greater than 90\%.
+In contrast, the correlations between the embedding coordinates and the macroscopic variables when using EMD in the diffusion maps calculation are all greater than 80\%.
 %
@@
 \centering
-\includegraphics[width=0.5\textwidth]{chemotaxis_compare_timescales_evals}
+\includegraphics[width=0.5\textwidth]{tmax_lambda_transition}
 %
 \caption{Detecting the change in dimensionality.
-The ratio of the log of the two eigenvalues which correspond to unique eigendirections is shown in blue, and the ratio of the diffusive and drift time scales are shown in red.
+The ratio of the log of the two eigenvalues which correspond to unique eigendirections is plotted as a function of the timescale of the simulation ($t_{max}$) and $\lambda$.
+Each data point is the average eigenvalue ratio over three data sets.
+Each data set consists of $10$ simulations, with initial conditions uniformly chosen such that $0.1 \le p  \le 0.9$.
+We allow each simulation to evolve for $t_{max}$ time units, with $dt=t_{max}/10$, and use data with $t > 0$ for analysis.
+The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions.
+The curve $1/\lambda = t_{max}/N$ is shown in white.
 When the diffusive time scale is significantly larger than the drift time scale, the data is expected to be effectively one-dimensional.
-This transition is consistent with the empirical eigenvalues: the data is effectively one-dimensional if the first eigenvalue is significantly larger than the second.
+This transition is consistent with the empirical eigenvalues: the data are effectively one-dimensional when $t_{max}$ and/or $\lambda$ are large, since the first eigenvalue is significantly larger than the second.
 }
@@
 
-Figure~\ref{fig:chemotaxis_compare_timescales_evals} shows the eigenvalue ratio in \eqref{eq:chemotaxis_eval_ratio} as a function of $\lambda$. 
-%
-We observe changes in the estimated dimensionality as a function of $\lambda$.
+Figure~\ref{fig:chemotaxis_compare_timescales_evals} shows the eigenvalue ratio in \eqref{eq:chemotaxis_eval_ratio} as a function of $\lambda$ and the time scale of observation $t_{max}$, and we observe changes in the estimated dimensionality.
 %
@@
 %
-For this specific example, these changes are consistent with the analytically-predicted shift. 
-%
-From \cite{othmer2000diffusion}, the relevant time scales for the velocity jump process are
+For this specific example, these changes are consistent with the analytically-predicted shift when 
+\begin{equation}
+\frac{t_{max}}{N} = \frac{1}{\lambda}.
+\end{equation} 
+%
+For small $\lambda$ and/or short $t_{max}$, the data are effectively two-dimensional, as both $p$ and $t$ are important to the observed dynamics. 
+%
+However, when $\lambda$ and/or $t_{max}$ is large, the system rapidly equilbrates and the data are effectively one-dimensional, parameterized by $t$. 
+%
+%From \cite{othmer2000diffusion}, the relevant time scales for the velocity jump process are
+%%\begin{equation}
+%%\begin{aligned}
+%%\tau_{drift} =& \frac{L}{s} \\
+%%\tau_{diff} =& \frac{L^2 \lambda}{s^2}
+%%\end{aligned}
+%%\end{equation}
+%%%
+%%Here, we take $L = t_{max} s$ as a characteristic length scale for the simulations. 
+%%
 %\begin{equation}
 %\begin{aligned}
-%\tau_{drift} =& \frac{L}{s} \\
-%\tau_{diff} =& \frac{L^2 \lambda}{s^2}
+%\tau_{drift} =& t_{max} \\
+%\tau_{diff} =& t_{max}^2 \lambda
 %\end{aligned}
@@
 %%
-%Here, we take $L = t_{max} s$ as a characteristic length scale for the simulations. 
-%
-\begin{equation}
-\begin{aligned}
-\tau_{drift} =& t_{max} \\
-\tau_{diff} =& t_{max}^2 \lambda
-\end{aligned}
-\end{equation}
-%
-When $\frac{\tau_{diff}}{\tau_{drift}} =  t_{max}  \lambda \gg 1$, then the dynamics are effectively one-dimensional. 
-%
+%When $\frac{\tau_{diff}}{\tau_{drift}} =  t_{max}  \lambda \gg 1$, then the dynamics are effectively one-dimensional. 
+%%
 This ratio is also plotted in Figure~\ref{fig:chemotaxis_compare_timescales_evals} as a function of $\lambda$, and the transition is consistent with the transition predicted by the diffusion maps eigenvalues.

### 2015-03-11 chemotaxis/simulate_jump_process_EMD_detect_transition.m

@@ -6,177 +6,176 @@ rng(123);
 
 %% simulations
 
-nlambda = 10;
-
-lambda_all = linspace(1, 1000, nlambda);
-s_all = sqrt(lambda_all);
+% nlambda = 10;
+% lambda_all = linspace(1, 400, nlambda);
+lambda_all = 1:10:200;
+nlambda = length(lambda_all);
 
 % number of particles
 N = 1000;
-% time step
-dt = 1;
-% maximum simulation time
-tmax = 100;
-nsteps = floor(tmax / dt) + 1;
 
-t_range_all = 10:10:tmax;
-ntime = length(t_range_all);
+% ntmax = 10;
+% tmax_all = linspace(10, 400, ntmax);
+tmax_all = 10:10:200;
+ntmax = length(tmax_all);
 
 %initial probility of a particle moving to the right (vel=+1)
 nsim = 10;
 pinit = linspace(0.1, 0.9, nsim);
 
 % histogram parameters
-nbins = 256;
-% nbins = 32;
-hist_all = zeros(nbins, nsteps*nsim);
-hist_left = zeros(nbins, nsteps*nsim);
-hist_right = zeros(nbins, nsteps*nsim);
-all_time = zeros(1, nsteps*nsim);
-all_p = zeros(1, nsteps*nsim);
+nbins = 32;
 
-niter = 1;
+niter = 3;
 
-eval1 = zeros(nlambda, ntime, niter);
-eval2 = zeros(nlambda, ntime, niter);
-t_dom = zeros(nlambda, ntime, niter);
+eval1 = zeros(nlambda, ntmax, niter);
+eval2 = zeros(nlambda, ntmax, niter);
+t_dom = zeros(nlambda, ntmax, niter);
 
-f1 = figure;
-f2 = figure;
-fig_idx = 1;
+% f1 = figure;
+% f2 = figure;
+% fig_idx = 1;
 
-%%
 
-for k=1:niter
-    for sim_num = 1:nlambda
-        
-        sim_num
-        
-        lambda = lambda_all(sim_num);
-        s = s_all(sim_num);
-        
-        x_hist = linspace(-s*tmax, s*tmax, nbins);
-        
-        for j=1:nsim
-            % simulate velocity jump process
-            [pos, vel, time] = simulate_jumps(N, tmax, dt, lambda, s, pinit(j));
+for j1=1:nlambda
+    j1
+    for j2=1:ntmax
+        for k=1:niter
             
-            % histogram of all particle positions as a function of time
[…]

### 2015-03-11 chemotaxis/simulate_jump_process_EMD_detect_transition2.m

@@ -1,176 +1,0 @@
-clear all
-close all
-
-rng(123);
-
-
-%% simulations
-
-% nlambda = 10;
-% lambda_all = linspace(1, 400, nlambda);
-lambda_all = 1:10:200;
-nlambda = length(lambda_all);
-
-% number of particles
-N = 1000;
-
-% ntmax = 10;
-% tmax_all = linspace(10, 400, ntmax);
-tmax_all = 10:10:200;
-ntmax = length(tmax_all);
-
-%initial probility of a particle moving to the right (vel=+1)
-nsim = 10;
-pinit = linspace(0.1, 0.9, nsim);
-
-% histogram parameters
-nbins = 32;
-
-niter = 3;
-
-eval1 = zeros(nlambda, ntmax, niter);
-eval2 = zeros(nlambda, ntmax, niter);
-t_dom = zeros(nlambda, ntmax, niter);
-
-% f1 = figure;
-% f2 = figure;
-% fig_idx = 1;
-
-
-for j1=1:nlambda
-    for j2=1:ntmax
-        for k=1:niter
-            
-            % rate of switching velocity
-            lambda = lambda_all(j1);
-            % speed of particles
-            s = sqrt(lambda);
-            
-            % maximum simulation time
-            tmax = tmax_all(j2);
-            %             tmin = 0.1*tmax;
-            tmin = tmax/10;
-            
-            % time step
-            dt = (tmax - tmin) / 9;
-            nsteps = floor(tmax / dt+1e-6) + 1;
-            
-            
-            % histogram parameters
-            x_hist = linspace(-s*tmax, s*tmax, nbins);
-            hist_all = zeros(nbins, nsteps*nsim);
-            hist_left = zeros(nbins, nsteps*nsim);
-            hist_right = zeros(nbins, nsteps*nsim);
-            all_time = zeros(1, nsteps*nsim);
-            all_p = zeros(1, nsteps*nsim);
-            
-            for j=1:nsim
-                % simulate velocity jump process
-                [pos, vel, time] = simulate_jumps(N, tmax, dt, lambda, s, pinit(j));
-                
-                % histogram of all particle positions as a function of time
-                for i=1:nsteps
-                    
-                    [hist_all(:,(j-1)*nsteps+i), ~] = hist(pos(:,i),x_hist);
-                    
-                    idx = (vel(:,i) == -1);
-                    [hist_left(:,(j-1)*nsteps+i), ~] = hist(pos(idx,i),x_hist);
-                    
[…]

### 2015-03-11 chemotaxis/simulate_jump_process_EMD_test_llr.m

@@ -1,13 +1,13 @@
 clear all
 close all
 
-rng(123);
+rng(321);
 markersize = 500;
 
 
 %% simulations
 
-lambda_all = [100 2500 6400];
+lambda_all = [100 1600 6400];
 s_all = sqrt(lambda_all);
 comp2 = [3 4 5];
 
@@ -71,7 +71,7 @@ for sim_num = 1:length(lambda_all)
     
     
     % use data after initial "relaxation" time
-    idx = (all_time > 1);
+    idx = (all_time > 0);
     
     %% compute EMD between histograms
     
@@ -99,12 +99,6 @@ for sim_num = 1:length(lambda_all)
     [V2, D2] = dmaps(W2, eps2, 10);
     
     
-    if corr(V2(:,2), all_time(idx)') < 0
-        V2(:,2) = -V2(:,2);
-    end
-    if corr(V2(:,3), all_p(idx)') < 0
-        V2(:,3) = -V2(:,3);
-    end
     
     %%
     eps_med_scale = 3;
@@ -119,6 +113,15 @@ for sim_num = 1:length(lambda_all)
     ylabel('\mu_k')
     axis square
     print(sprintf('chemotaxis%d_evals.eps', sim_num),'-depsc')
+    
+    
+    if corr(V2(:,2), all_time(idx)') < 0
+        V2(:,2) = -V2(:,2);
+    end
+    if corr(V2(:,comp2(sim_num)), all_p(idx)') < 0
+        V2(:,comp2(sim_num)) = -V2(:,comp2(sim_num));
+    end
+    
     
     make_fig(3,2);
     scatter(V2(:,2),V2(:,3),50, all_p(idx), '.')
@@ -143,6 +146,7 @@ for sim_num = 1:length(lambda_all)
     axis tight
     axis square
     print(sprintf('chemotaxis%d_embed_good.eps', sim_num),'-depsc')
+    
     
     %
     % figure;
@@ -169,6 +173,7 @@ for sim_num = 1:length(lambda_all)
     
 end
 
+return
 
 %%
 
@@ -215,7 +220,7 @@ for sim_num = 1:prod(size(lambda_all))
     
     
     % use data after initial "relaxation" time
-    idx = (all_time > 1);
+    idx = (all_time > 0);
     
     % compute EMD between histograms
     
@@ -243,8 +248,8 @@ for sim_num = 1:prod(size(lambda_all))
     eps_med_scale = 3;
[…]

### 2015-03-11 chemotaxis/strip.m

@@ -110,7 +110,7 @@ print('strip_nonuniform2.eps','-depsc')
 
 n = 2000;
 
-rng(321);
+rng(123);
 
 for Lx = [2 4 8]
     
@@ -119,7 +119,7 @@ for Lx = [2 4 8]
     data(:,2) = data(:,2) * Ly;
     
     W = squareform(pdist(data)).^2;
-    eps = 0.25^2;
+    eps = 0.15^2;
     
     [V, D] = dmaps(W, 2*eps, 10);
     

### 2015-03-11 chemotaxis/test.m

@@ -9,30 +9,27 @@ markersize = 500;
 
 
 % rate of switching velocity
-lambda = 1;
+lambda = 400;
 % speed of particles
 s = sqrt(lambda);
 
 % number of particles
 N = 1000;
 % maximum simulation time
-tmax = 10000;
-tmin = 0.1*tmax;
+tmax = 10;
+tmin = 0;
 % time step
-dt = (tmax - tmin) / 9;
+dt = 1;
 nsteps = floor(tmax / dt) + 1;
 
 %initial probility of a particle moving to the right (vel=+1)
 nsim = 10;
 pinit = linspace(0.1, 0.9, nsim);
 
-min_start = -10;
-max_start = 10;
-
 % histogram parameters
 % nbins = 512;
 nbins = 32;
-x_hist = linspace(min_start-s*tmax, max_start+s*tmax, nbins);
+x_hist = linspace(-s*tmax, s*tmax, nbins);
 hist_all = zeros(nbins, nsteps*nsim);
 hist_left = zeros(nbins, nsteps*nsim);
 hist_right = zeros(nbins, nsteps*nsim);
@@ -81,7 +78,15 @@ idx = (all_time > tmin);
 
 %%
 
-tmp_diff = sum((hist_left-hist_right).^2);
+rho_diff = sum((hist_left-hist_right).^2);
+rho_mean_left = x_hist * hist_left;
+rho_mean_right = x_hist * hist_right;
+rho_mean = x_hist * hist_all;
+rho_mean_der = x_hist * (hist_all(:, 2:end) - hist_all(:, 1:end-1));
+rho_mean_der = [0 rho_mean_der];
+
+rho_mean_left_norm = x_hist * (hist_left./repmat(sum(hist_left), nbins, 1));
+rho_mean_right_norm = x_hist * (hist_right./repmat(sum(hist_right), nbins, 1));
 
 %% compute EMD between histograms
 
@@ -141,11 +146,55 @@ corr(V2(:,2),all_time(idx)')
 
 corr(V2(:,3),all_p(idx)')
 
+return
+
 %%
 
+% figure;
+% scatter(V2(:,2),V2(:,3),markersize,rho_diff(idx), '.')
+% xlabel('$\phi_1$', 'interpreter','latex')
+% ylabel('$\phi_2$', 'interpreter','latex')
+% h = colorbar;
+% set(get(h,'xlabel'),'String', 'FLUXES');
+% 
+% figure;
[…]

### 2015-03-11 chemotaxis/test_torus.m

@@ -4,7 +4,7 @@ close all
 %%
 rng(123);
 
-n = 1000;
+n = 2000;
 
 theta1 = 2*pi*rand(n, 1);
 theta2 = 2*pi*rand(n, 1);

### 2015-03-11 chemotaxis/paper/draft1.tex
text edits to chemotaxis paper

@@
 \begin{abstract}
-
-TODO
+Nonlinear manifold learning algorithms have proven useful for the anlaysis of complex data sets. 
+%
+However, these algorithms still encounter several challenges when faced with real data. 
 \end{abstract}
@@
 %
-In particular, for dynamical systems, data-driven methodologies have been shown to be essential when a simple macroscopic model cannot be written analytically due to the complexity of the system.
+In particular, for dynamical systems, data-driven methodologies have been essential when a simple macroscopic model cannot be written analytically due to the complexity of the system.
 %
@@
 %
-For such cases, data obtained from observations and/or simulations of the dynamical system combined with these methodologies give rise to a low-dimensional description which can not only provide insight into the underlying dynamics, but also serve as a first step in constructing macroscale models which are consistent with the observed microscale behavior.
+For such cases, data obtained from observations and/or simulations of the dynamical system combined with such methodologies give rise to a low-dimensional description which can not only provide insight into the underlying dynamics, but also serve as a first step in constructing macroscale models which are consistent with the observed microscale behavior.
 
@@
 
-Due to the complexity and range of microscale behaviors possible in dynamical systems, we use manifold learning, a nonlinear data-driven analysis technique, to analyze data.
+Due to the complexity and range of microscale behaviors possible in dynamical systems, we use manifold learning, a nonlinear data-driven technique, to analyze data.
 %
@@
 %
-Our hope is that these coordinates, obtained in a data-driven manner, will describe the variables which govern the macroscale dynamical behavior of the system and enable us to introduce a reduced macroscale ab-initio model.
-
-These algorithms were first applied to synthetic data sets to illustrate their geometric properties and flexibility. 
-%
-Only recently have they been applied to experimental and simulation data, which was possible due to advances in data representation (observers) and metrics. 
-%
-From a geometric perspective, not all eigenvectors are guaranteed to parameterize unique directions within the data; some eigenvectors are 	``repeated eigendirections'' which describe the same mode. 
-%
-Identifying those eigenvectors is critical for obtaining an accurate parameterization of the system which captures the true dimensionality of the macroscale dynamics.  
-%
-This is a challenging task which is often done manually, and few methods have been proposed to automate the identification of the unique eigendirections.
+Our hope is that these coordinates, obtained in a data-driven manner, will describe the variables which govern the macroscale dynamical behavior of the system and enable us to introduce a reduced macroscale {\em ab initio} model.
+
+These manifold learning algorithms were first applied to synthetic data sets to illustrate their geometric properties and flexibility. 
+%
+Only recently have they been applied to experimental and simulation data, which was possible due to advances in data representation (observers) and metrics \cite{...}. 
+%
+A critical shortcoming of these methods is that from a geometric perspective, not all eigenvectors are guaranteed to parameterize unique directions within the data; some eigenvectors are ``repeated eigendirections'' which describe the same mode.
+%
+Identifying those eigenvectors is critical for obtaining a parameterization of the system which captures the true dimensionality of the macroscale dynamics.  
+%
+This is a challenging task which is often done manually, and few methods have been proposed to automate the identification of the unique eigendirections \cite{...}.
 
@@
 %
-We first demonstrate our algorithm on synthetic examples where we have a closed-form solution for the eigenvectors and eigenvalues of the Laplace operator.  
-%
-This will allow us to validate our proposed approach.
+We first demonstrate our algorithm on synthetic examples where we have a closed-form solution for the eigenvectors and eigenvalues of the Laplace operator, which will allow us to validate our proposed approach.
 %
@@
 %
-The recent advances in observers and metrics, coupled with the proposed approach for factoring out repeated eigendirections, provide a data analysis pipeline which can successfully analyze data from this simulation.
+The recent advances in observers and metrics, coupled with the proposed approach for identifying the unique eigendirections, provide a data analysis pipeline which can successfully analyze data from this simulation.
 %
@@
 
-Data from dynamical systems is most often high-dimensional, since it is collected at the microscale.
-%
-However, the macroscale dynamics often evolve to a low-dimensional description. 
+Although data from dynamical systems is most often high-dimensional, since it is collected at the microscale, the macroscale dynamics often evolve to a low-dimensional description. 
 %
@@
 \end{subfigure}
-\caption{Two-dimensional strip colored by the eigenfunction (a) $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi z_1}{L_1} \right)$, and (b) $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi z_2}{L_2} \right)$ .}
+\caption{Two-dimensional strip colored by the eigenfunction (a) $\tilde{\phi}_{1, 0} = \cos \left( {\pi z_1}/{L_1} \right)$, and (b) $\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$ .}
 \label{fig:strip_efuncs}
@@
 %
-We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( \frac{\pi z_1}{L_1} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( \frac{\pi z_2}{L_2} \right)$ are one-to-one with the $z_1$ and $z_2$ coordinates, respectively, and therefore yield a parameterization of the underlying manifold (see Figure~\ref{fig:strip_efuncs}).
+We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( {\pi z_1}/{L_1} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$ are one-to-one with the $z_1$ and $z_2$ coordinates, respectively, and therefore yield a parameterization of the underlying manifold (see Figure~\ref{fig:strip_efuncs}).
 %
@@
 First, $z_1$ and $z_2$ are not necessarily decoupled in subsequent eigenfunctions. 
-TODO: address this in later section
+% TODO: address this in later section
 %
@@
 %
-Alternatively, one can factor out the density effects in the diffusion maps matrix.
+Alternatively, one can factor out the density effects in the weight matrix.
 %
@@
 \begin{equation}
-\tilde{W} = D^{-1} W D^{-1}.
-\end{equation}
-%
-Pre- and post-multiplication by $D^{-1}$ removes any density effects from $W$. 
+\tilde{W} = D^{-1} W D^{-1},
+\end{equation}
+%
+where pre- and post-multiplication by $D^{-1}$ removes any density effects from $W$. 
 %
@@
 %
-The matrix $\tilde{A}$ is comparable to the matrix $A$, but with sampling density effects removed, such that $I-\tilde{A}$ approaches the Laplace-Beltrami operator even if the data is nonuniformly sampled from the underlying manifold \cite{coifman2005geometric}.
+The matrix $\tilde{A}$ is comparable to the matrix $A$, but with sampling density effects removed, such that $I-\tilde{A}$ approaches the Laplace-Beltrami operator on $\mathcal{M}_d$ even if the data is nonuniformly sampled from the underlying manifold \cite{coifman2005geometric}.
 
@@
 Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a trivial constant vector.
-
-Figure~\ref{fig:strip_evecs} shows data sampled from a strip, colored by diffusion maps eigenvectors. 
-%
-In cases of both uniform and nonuniform sampling, the eigenvectors are one-to-one with $z_1$ and $z_2$, and thus parameterize the manifold. 
+%
+Figure~\ref{fig:strip_evecs} shows data sampled from a strip, colored by eigenvectors of $A$. 
+%
+In cases of both uniform and nonuniform sampling, the selected eigenvectors are one-to-one with $z_1$ and $z_2$, and thus parameterize the manifold. 
 %
@@
 %
-From \eqref{eq:evals}, we see that if we sort the eigenvectors by the magnitude of the corresponding eigenvalues, these eigenvectors are guaranteed to appear before any coupled or repeated eigendirections.
+From \eqref{eq:evals}, we see that sorting the eigenvectors by the magnitude of the corresponding eigenvalues implies that these eigenvectors are guaranteed to appear before any coupled or repeated eigendirections.
 %
@@
 %
-We will attempt to fit a function $f(\phi_1, \dots, \phi_{k-1})$ to $\phi_{k}$. 
-%
-Our approach is motivated by simple trigonometric arguments, such as $\cos2x$ can be written as a function of $\cos x$. 
-%
-Thus, if the resulting fit is accurate, then we assume $\phi_{k}$ is a repeated eigendirection of $\phi_1, \dots, \phi_{k-1}$. 
+Our approach is motivated by simple trigonometric arguments, such as the fact that $\cos2x$ (a repeated eigendirection) can be written as a function of $\cos x$ (a unique eigendirection). 
+%
+We will therefore attempt to fit a function $f(\phi_1, \dots, \phi_{k-1})$ to $\phi_{k}$. 
+%
+If the resulting fit is accurate, we will assume $\phi_{k}$ is a repeated eigendirection of $\phi_1, \dots, \phi_{k-1}$. 
 %
@@
 \begin{equation} \label{eq:cv_error}
-r_{k} = \sqrt{ \frac{\sum_{i=1}^n \left( \phi_{k} (i) - (\hat{\alpha}_k(i) + \hat{\beta}_k(i)^T \Phi_{k-1}(i))  \right)^2} {\sum_{i=1}^n  \left( \phi_{k} (i) \right)^2 }}
+r_{k} = \sqrt{ \frac{\sum_{i=1}^n \left( \phi_{k} (i) - (\hat{\alpha}_k(i) + \hat{\beta}_k(i)^T \Phi_{k-1}(i))  \right)^2} {\sum_{i=1}^n  \left( \phi_{k} (i) \right)^2 }}.
 \end{equation}
@@
 Note that a small value of $r_k$ implies that $\phi_{k}$ can be accurately approximated from $\phi_1, \dots, \phi_{k-1}$. 
-We will assume it means that $\phi_k$ is a harmonic of previous modes, i.e., a repeated eigendirction.
-%
-Conversely, a large value of $r_{k}$ indicates that $\phi_{k}$ parameterizes a new direction in the data.
+%
+We will assume this implies that $\phi_k$ is a harmonic of previous modes, i.e., a repeated eigendirction, and conversely, a large value of $r_{k}$ indicates that $\phi_{k}$ parameterizes a unique eigendirection in the data.
 %
@@
 %
-We begin by computing the matrix
+We begin by constructing the matrix
 \begin{equation}
@@
 
-For the strip example, once the eigenvectors which parameterize unique eigendirections along the manifold have been determined, the corresponding eigenvalues can be used to calculate the relative lengths along these directions.
-%
-For more general manifolds, we believe the eigenvalues can still provide a measure of the relative significance of the unique eigendirections.
+For a two-dimensional strip, once the eigenvectors which parameterize unique eigendirections have been determined, the corresponding eigenvalues can be used to calculate the relative lengths along these directions.
+%
+For more general manifolds, we presume that the eigenvalues can still provide a measure of the relative significance of the unique eigendirections.
 %
@@
 %
-This measure is used to determine how many components are required to retain most of the information within the data set.
-%
-For example, if $L_j$ is smaller than a predefined threshold, then this component can be disregarded.
+These lengths are used to determine how many components are required to retain most of the information within the data set; 
+for example, if $L_j$ is smaller than a predefined threshold, then this component can be disregarded.
 
@@
 %
-Each data set contains $m=1500$ data points uniformly sampled from the strip. 
+Each data set contains $m=2000$ data points uniformly sampled from the strip. 
 %
@@
 %
-The eigenvalues are colored by the leave-one-out cross-validation error as defined in \eqref{eq:cv_error}; a small value of $r_k$ indicate that the corresponding eigenvector is a repeated eigendirection, while a large value of $r_k$ indicates that the corresponding eigenvector describes a new direction in the data.
+The eigenvalues are colored by the leave-one-out cross-validation error as defined in \eqref{eq:cv_error}; a small value of $r_k$ indicates that the corresponding eigenvector is a repeated eigendirection, while a large value of $r_k$ indicates that the corresponding eigenvector describes a new direction in the data.
 %
@@
 
-We observe that the eigenvalues are consistent with the known analytic eigenvalues of the Laplacian (see \eqref{eq:evals} and \eqref{eq:evals_relationship}).
-%
-Furthermore, the two unique eigendirections can be identified, since their corresponding regression error $r_k$ is large. 
+We observe that the eigenvalues are consistent with the known analytic eigenvalues of the Laplacian.
+%
+Furthermore, the two unique eigendirections can easily be identified, since their corresponding regression error $r_k$ is large. 
 %
@@
 \caption{}
+\label{subfig:swissroll1}
+\end{subfigure}
+%\hfill
+\begin{subfigure}{0.45\textwidth}
+\centering
+\includegraphics[height=1.25in]{swissroll2}
+\caption{}
+\label{subfig:swissroll2}
+\end{subfigure}
+%\hfill
+\begin{subfigure}{0.45\textwidth}
+\centering
+\includegraphics[height=1.25in]{swissroll1_evals}
+\caption{}
+\label{subfig:swissroll1_evals}
+\end{subfigure}
+%\hfill
+\begin{subfigure}{0.45\textwidth}
+\centering
+\includegraphics[height=1.25in]{swissroll2_evals}
+\caption{}
+\label{subfig:swissroll2_evals}
+\end{subfigure}
+%\hfill
+\begin{subfigure}{0.45\textwidth}
+\centering
+\includegraphics[height=1.25in]{swissroll1_color1} 
+\caption{}
+\label{subfig:swissroll1_color1}
 \end{subfigure}
@@
 \centering
-\includegraphics[height=1.25in]{swissroll2}
-\caption{}
+\includegraphics[height=1.25in]{swissroll2_color1}
+\caption{}
+\label{subfig:swissroll2_color1}
 \end{subfigure}
@@
 \centering
-\includegraphics[height=1.25in]{swissroll1_evals}
-\caption{}
+\includegraphics[height=1.25in]{swissroll1_color2}
+\caption{}
+\label{subfig:swissroll1_color2}
 \end{subfigure}
@@
 \centering
-\includegraphics[height=1.25in]{swissroll2_evals}
-\caption{}
-\end{subfigure}
-\hfill
-\begin{subfigure}{0.45\textwidth}
-\centering
-\includegraphics[height=1.25in]{swissroll1_color1} 
-\caption{}
-\end{subfigure}
-\hfill
-\begin{subfigure}{0.45\textwidth}
-\centering
-\includegraphics[height=1.25in]{swissroll2_color1}
-\caption{}
-\end{subfigure}
-\hfill
-\begin{subfigure}{0.45\textwidth}
-\centering
-\includegraphics[height=1.25in]{swissroll1_color2}
-\caption{}
-\end{subfigure}
-\hfill
-\begin{subfigure}{0.45\textwidth}
-\centering
 \includegraphics[height=1.25in]{swissroll2_color2}
 \caption{}
+\label{subfig:swissroll2_color2}
 \end{subfigure}
@@
 %
-such that $z_1, z_2$ are uniformly sampled along the arclength of the spiral, and $t \sim Unif(0,1)$. 
+where $\theta$ is sampled such that $z_1, z_2$ are uniformly sampled along the arclength of the spiral, and $t \sim Unif(0,1)$. 
 %
@@
 %
-Each data set consists of $m=1500$ points, shown shown in Figure~\ref{fig:swiss_rolls}(a)~and~(b).
+Each data set consists of $m=1500$ points, shown shown in Figures~\ref{subfig:swissroll1}--\ref{subfig:swissroll2}.
 %
@@
 
-Figure~\ref{fig:swiss_rolls}~(c)~and~(d) shows the eigenvalue spectra from the analysis of the two data sets.
+Figures~\ref{subfig:swissroll1_evals}--\ref{subfig:swissroll2_evals} shows the eigenvalue spectra from the analysis of the two data sets.
 %
@@
 %
-Figure~\ref{fig:swiss_rolls}~(e--h) shows the two data sets, colored by the two eigenvectors identified by our method as parameterizing the unique eigendirections. 
+Figures~\ref{subfig:swissroll1_color1}--\ref{subfig:swissroll2_color2} shows the two data sets, colored by the two eigenvectors identified by our method as parameterizing the unique eigendirections. 
 %
@@
 
-For the chemotaxis example, we use histograms of the particle positions as observers \cite{talmon2013empirical}. 
+For the chemotaxis example, we use histograms of the particle positions as observers. 
 %
@@
 \begin{equation}
-\| z(i) - z(j) \|_{EMD} \approx \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|
-\end{equation}
-where the histograms $z(i), z(j)$ are defined on equally-spaced bins in $\mathbb{R}$, and $z_k$ denotes the $k^{th}$ bin. 
-%
-We note that although histograms are not essential for using EMD, histograms make the computation feasible. 
-%
-For our specific simulations, we will take $n=32$.
+\| z(i) - z(j) \|_{EMD} \approx \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|,
+\end{equation}
+where the histograms $z(i), z(j)$ are defined on $n$ equally-spaced bins in $\mathbb{R}$ (for our specific simulations, we will take $n=32$), and $z_k$ denotes the $k^{th}$ bin. 
+%
+We note that although histograms are not essential for using EMD, they make the computation feasible. 
+%
 
@@
 %
-Plotting the data using the two eigenvectors identified as corresponding to unique eigendirections yields two-dimensional embeddings which capture the macroscopic variables $p$ and $t$ (we only show the correspondence of the two identified unique eigendirections with $p$ in the middle row of Figure~\ref{fig:chemotaxis_simulations_harmonics} and report that the correspondence between these eigendirections and $t$ is also consistent).
+Plotting the data using the two eigenvectors identified as corresponding to unique eigendirections yields two-dimensional embeddings which capture the macroscopic variables $p$ and $t$ (we only show the correspondence of the two identified unique eigendirections with $p$ in the middle row of Figure~\ref{fig:chemotaxis_simulations_harmonics}; the correspondence between these eigendirections and $t$ is similarly consistent).
 %
@@
 %
-From the top row of Figure~\ref{fig:chemotaxis_simulations_harmonics}, we can see that the eigenvalue spectrum does not exhibit any large spectral gaps.
-%
-However, there is a significant gap between the eigenvalues which correspond to new directions in the data (colored in blue), and the corresponding eigenvectors provide us with an informative two-dimensional parameterization of the data.
+From the top row of Figure~\ref{fig:chemotaxis_simulations_harmonics}, we see that the eigenvalue spectra do not exhibit any large spectral gaps.
+%
+However, the gap between the eigenvalues which correspond to new directions in the data (colored in blue) increases with $\lambda$, indicating that the underlying manifold becomes more narrow/one-dimensional, and the corresponding eigenvectors provide us with an informative two-dimensional parameterization of the data.
 %which is consistent with the expected macroscopic dynamics.
@@
 %\]
-When $\lambda \rightarrow \infty$, \eqref{eq:second_order_pde} approaches the heat equation,
+When $\lambda \rightarrow \infty$ with $s^2/\lambda$ fixed, \eqref{eq:second_order_pde} approaches the heat equation,
 \begin{equation}
@@
 
-Figure~\ref{fig:chemotaxis_compare_timescales_evals} shows the eigenvalue ratio in \eqref{eq:chemotaxis_eval_ratio} as a function of $\lambda$ and the time scale of observation $t_{max}$, and we observe changes in the estimated dimensionality.
+Figure~\ref{fig:chemotaxis_compare_timescales_evals} shows the eigenvalue ratio in \eqref{eq:chemotaxis_eval_ratio} as a function of $\lambda$ and the time scale of observation $t_{max}$.
+%
+From this eigenvalue ratio, we detect changes in the estimated dimensionality as we vary the relevant parameters.
+%
+or small $\lambda$ and/or short $t_{max}$, the data are effectively two-dimensional, as both $p$ and $t$ are important to the observed dynamics. 
+%
+However, when $\lambda$ and/or $t_{max}$ is large, the system rapidly equilbrates and the data are effectively one-dimensional, parameterized by $t$. 
 %
@@
 %
-For this specific example, these changes are consistent with the analytically-predicted shift when 
+For this specific example, these changes are consistent with the analytically-predicted shift at 
 \begin{equation}
@@
 \end{equation} 
-%
-For small $\lambda$ and/or short $t_{max}$, the data are effectively two-dimensional, as both $p$ and $t$ are important to the observed dynamics. 
-%
-However, when $\lambda$ and/or $t_{max}$ is large, the system rapidly equilbrates and the data are effectively one-dimensional, parameterized by $t$. 
 %

### 2015-03-12 chemotaxis/paper/draft1.tex
updated paper

@@
-\documentclass[preprint]{elsarticle}
+\documentclass[3p]{elsarticle}
 
@@
 
-\title{Parsimonious Representation of Nonlinear Dynamical Systems using Diffusion Maps: A Chemotaxis Case Study}
-
-\author[princeton]{Carmeline~J.~Dsilva\corref{cor1}}
+\title{Parsimonious Representation of Nonlinear Dynamical Systems Through Manifold Learning: A Chemotaxis Case Study}
+
+\author[princeton]{Carmeline~J.~Dsilva}
 \ead{cdsilva@princeton.edu}
@@
 %
-\author[princeton, princetonpacm]{Ioannis~G.~Kevrekidis \corref{cor1}}
+\author[princeton,princetonpacm]{Ioannis~G.~Kevrekidis \corref{cor1}}
 \ead{yannis@princeton.edu}
-%
+
 \address[princeton]{Department of Chemical and Biological Engineering, Princeton University, Princeton, NJ, 08540, USA}
@@
 \begin{abstract}
-Nonlinear manifold learning algorithms have proven useful for the anlaysis of complex data sets. 
+Nonlinear manifold learning algorithms have been effective for the analysis of complex data sets. 
 %
 However, these algorithms still encounter several challenges when faced with real data.
+%
+A key shortcoming is the existence of repeated eigendirections, where several embedding coordinates parameterize the same direction in the data set, therefore obscuring the true dimensionality of the underlying manifold.
+%
+In this paper, we propose an algorithm, based on local linear regression, to automatically detect which coordinates are repeated eigendirections. 
+%
+We first demonstrate the utility and flexibility of our approach on synthetic data sets. 
+%
+We then apply our algorithm to data collected from a stochastic simulation of cellular chemotaxis. 
+%
+Here, we show that our approach for factoring out repeated eigendirections can allow us to detect changes in dynamical behavior and the underlying system dimensionality directly from data. 
+%
 \end{abstract}
@@
 \begin{keyword}
-TODO
+diffusion maps, repeated eigendirections, chemotaxis
 \end{keyword}
@@
 
-In recent years, data mining algorithms have proven useful for many disciplines and applications \cite{....}. 
-%
-In particular, for dynamical systems, data-driven methodologies have been essential when a simple macroscopic model cannot be written analytically due to the complexity of the system.
+In recent years, data mining algorithms have proven useful for many disciplines and applications \cite{gepshtein2013image, fernandez2014diffusion, singer2011viewing, yuan2014automated, zhao2003face, trapnell2014dynamics, kemelmacher2011exploring, sifre2013rotation}.
+%
+In particular, for dynamical systems, data-driven methodologies have been essential when a simple macroscopic model cannot be written analytically due to the complexity of the system \cite{talmon2014intrinsic,berry2013time,singer2009detecting,ferguson2010systematic}.
 %
@@
 %
-For such cases, data obtained from observations and/or simulations of the dynamical system combined with such methodologies give rise to a low-dimensional description which can not only provide insight into the underlying dynamics, but also serve as a first step in constructing macroscale models which are consistent with the observed microscale behavior.
-
-%For many engineering applications, it is often advantageous or even essential to effectively parameterize a system's dynamics. 
-%TODO: add reasons why, potential applications, etc.
-%%
-%Often, such a parameterization cannot be obtained analytically, due to the complexity of the system.
-%%
-%For simulations, the governing model is often very high-dimensional and the macroscale dynamics are not obvious; for experimental systems, an explicit model is often unknown. 
-%%
-%In such cases, we utilize data obtained from observations and/or simulations of the dynamical system, and turn to data-driven analysis methods to provide an accurate low-dimensional description of this data.
-%%
-%This low-dimensional description can not only provide insight into the underlying dynamics, but also serve as a first step in constructing macroscale models which are consistent with the observed microscale behavior.
-%
-
-Due to the complexity and range of microscale behaviors possible in dynamical systems, we use manifold learning, a nonlinear data-driven technique, to analyze data.
-%
-Most manifold learning algorithms \cite{...} construct parameterizations of the data through the spectral analysis of a Laplace operator.
+For such cases, data obtained from observations and/or simulations of the dynamical system combined with such methodologies gives rise to a low-dimensional description which can not only provide insight into the underlying dynamics, but also serve as a first step in constructing macroscale models which are consistent with the observed microscale behavior.
+
+Due to the complexity and range of microscale behaviors possible in dynamical systems, we use manifold learning, a class of nonlinear data-driven techniques, to analyze data.
+%
+Most manifold learning algorithms construct parameterizations of the data through the spectral analysis of a Laplace operator \cite{Belkin2003,coifman2005geometric,coifman2006geometric,singer2008non}.
 %
@@
 
-These manifold learning algorithms were first applied to synthetic data sets to illustrate their geometric properties and flexibility. 
-%
-Only recently have they been applied to experimental and simulation data, which was possible due to advances in data representation (observers) and metrics \cite{...}. 
+These manifold learning algorithms were first applied to synthetic data sets to illustrate their geometric properties and flexibility \cite{coifman2005geometric, nadler2006diffusion}. 
+%
+Only recently have they been applied to experimental and simulation data, which was possible due to advances in data representation (observers) and metrics \cite{rubner2000earth,mallat2012group,talmon2013empirical,zhao2014rotationally,rohrdanz2011determination.}.
 %
@@
 %
-Identifying those eigenvectors is critical for obtaining a parameterization of the system which captures the true dimensionality of the macroscale dynamics.  
-%
-This is a challenging task which is often done manually, and few methods have been proposed to automate the identification of the unique eigendirections \cite{...}.
-
-In this paper, we propose an algorithm to automatically identify the unique eigendirections using local linear regression.
-%
-We first demonstrate our algorithm on synthetic examples where we have a closed-form solution for the eigenvectors and eigenvalues of the Laplace operator, which will allow us to validate our proposed approach.
-%
-We then consider a complex data set from a stochastic dynamical system which models cellular chemotaxis. 
-%
-The recent advances in observers and metrics, coupled with the proposed approach for identifying the unique eigendirections, provide a data analysis pipeline which can successfully analyze data from this simulation.
-%
-This pipeline will allow us to detect changes in the regime/mode of the system, as well as changes in the underlying dimensionality of the macroscale dynamics.
-
-%\begin{itemize}
-%
-%\item
-%Data driven analysis methods, and in particular, manifold learning, can do many things (dimensionality reduction, reduction in dynamical systems, etc). 
-%
-%\item
-%Manifold learning constructs parameterization of the data through the spectral analysis of a Laplace operator.
-%
-%\item
-%Initially it was used almost exclusively for toy examples/synthetic data (``these people suck").
-%
-%\item
-%Only recently, has it been applied to real (experimental/simulation) data, and this was possible due to advances in observers (data representation) and metrics (``we are awesome"). 
-%
-%\item
-%However, a major limitation (of spectral methods based on a Laplace operator) is identifying the true principal components.
-%
-%\item For the example of a square/strip, we know everything analytically, and identifying the ``true'' principal components is a problem/not obvious, and depends both on geometry and density
-%
-%\item While we can ignore/address nonuniform densities through normalization of the Laplace operator, we still have the problem of repeated eigendirections. 
-%
-%\item Add text about the meaning of the eigenvectors (obtained in the form of eigenvectors of a Laplace operator), as demonstrated in the chemotaxis example.
-The main idea is the ability to obtain, in a data driven manner, the "true" parsimonious description of the system, which is certain cases (such as particular regimes/modes of the chemotaxis system) may be significantly different from the typical representation and choice of variables used to describe the system (e.g.
-in the chemotaxis, it will not necessarily be the two fluxes).
-%
-%\item
-%In this paper, we propose an algorithm to automatically identify those principal components using locally linear regression.
-%
-%\item
-%All of these considerations (recently developed plus what we propose here) provide a data analysis pipeline.
-We show that this pipeline can successfully analyze data from a complex stochastic dynamical system originated from the study of cellular dynamics (chemotaxis).
-%
-%\end{itemize}
+Identifying those eigenvectors is critical for obtaining a parsimonious parameterization of the system which captures the true dimensionality of the macroscale dynamics.  
+%
+This is a challenging task which is often done manually, and few methods have been proposed to automate the identification of the unique eigendirections \cite{gerber2007robust}.
+
+In this paper, we propose an algorithm to automatically identify the unique eigendirections using local linear regression \cite{wasserman2006all}.
+%
+We first demonstrate our algorithm on synthetic examples where a closed-form solution for the eigenvectors and eigenvalues of the Laplace operator is known, which will allow us to validate our proposed approach.
+%
+We then consider a complex data set from a stochastic dynamical system simulation which models cellular chemotaxis \cite{othmer1988models}. 
+%
+The recent advances in observers and metrics, coupled with the proposed approach for identifying the unique eigendirections, provide a data analysis pipeline which successfully analyzes data from this simulation.
+%
+This pipeline allows us to detect changes in the regime/mode of the system, as well as changes in the underlying dimensionality of the macroscale dynamics.
 
@@
 
-Although data from dynamical systems is most often high-dimensional, since it is collected at the microscale, the macroscale dynamics often evolve to a low-dimensional description. 
-%
-In such cases, the high-dimensional microscale data will be restricted to a low-dimensional manifold embedded in the high-dimensional space. 
-%
-Extracting a parameterization of this manifold corresponds to parameterizing the low-dimensional, macroscale dynamics \cite{...}.
-
+Although data from dynamical systems is often high-dimensional, since it is collected at the microscale, the macroscale dynamics often evolve to a low-dimensional description. 
+%
+In such cases, the high-dimensional microscale data is restricted to a low-dimensional manifold embedded in the high-dimensional space, and extracting a parameterization of this manifold corresponds to parameterizing the low-dimensional, macroscale dynamics.
+%
 Let $z(1), \dots, z(m) \in \mathbb{R}^n$ denote $m$ observations sampled from the dynamical system. 
@@
 \centering
-\begin{subfigure}{0.45\textwidth}
+\begin{subfigure}{0.3\textwidth}
 \includegraphics[width=\textwidth]{strip_cnts1}
-\caption{}
-\end{subfigure}
-%
-\begin{subfigure}{0.45\textwidth}
 \includegraphics[width=\textwidth]{strip_cnts2}
 \caption{}
-\end{subfigure}
-\caption{Two-dimensional strip colored by the eigenfunction (a) $\tilde{\phi}_{1, 0} = \cos \left( {\pi z_1}/{L_1} \right)$, and (b) $\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$ .}
-\label{fig:strip_efuncs}
+\label{subfig:strip_efuncs}
+\end{subfigure}
+%
+\begin{subfigure}{0.3\textwidth}
+\includegraphics[width=\textwidth]{strip_discrete1}
+\includegraphics[width=\textwidth]{strip_discrete4}
+\caption{}
+\label{subfig:strip_evecs_uniform}
+\end{subfigure}
+%
+\begin{subfigure}{0.3\textwidth}
+\includegraphics[width=\textwidth]{strip_nonuniform1}
+\includegraphics[width=\textwidth]{strip_nonuniform2}
+\caption{}
+\label{subfig:strip_evecs_nonuniform}
+\end{subfigure}
+%
+\caption{(a) Two-dimensional continuous strip colored by the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( {\pi z_1}/{L_1} \right)$, and $\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$.
+(b) Two-dimensional strip with uniform sampling colored by the first and fourth (non-trivial) eigenvectors of the discrete Laplacian.
+(c) Two-dimensional strip with data sampled from a Gaussian distribution in $z_1$ and sampled uniformly in $z_2$, colored by the first and third (non-trivial) eigenvectors of the discrete Laplacian.
+Note that in all cases we uncover parameterizations which are one-to-one with $z_1$ and $z_2$.}
 \end{figure}
@@
 %
-In such an example, we will show that the existence of repeated eigendirections is inherent to the manifold learning setup based on Laplace operators, and that the identification of the unique eigendirections is nontrivial.
+Using such an example, we will show that the existence of repeated eigendirections is inherent to the manifold learning setup based on Laplace operators, and that the identification of the unique eigendirections is nontrivial.
 %
 The existence of these challenges in the continuous setting implies that, even in the limit of infinite data, such repeated eigendirections still pose a problem for analysis.
-%
-Thus, we consider the problem of parameterizing a continuous $d$-dimensional manifold $\mathcal{M}_d$ embedded in $\mathbb{R}^n$.
+
+We consider the problem of parameterizing a continuous $d$-dimensional manifold $\mathcal{M}_d$ embedded in $\mathbb{R}^n$.
 %
@@
 However, in the case when the manifold is nonlinear, the set of coordinates is not readily apparent. 
-
+%
 In recent years, the eigenfunctions of the Laplace-Beltrami operator have been used to parameterize the manifold \cite{Belkin2003, coifman2005geometric, singer2008non}. 
@@
 %
-We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( {\pi z_1}/{L_1} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$ are one-to-one with the $z_1$ and $z_2$ coordinates, respectively, and therefore yield a parameterization of the underlying manifold (see Figure~\ref{fig:strip_efuncs}).
+We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( {\pi z_1}/{L_1} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$ are one-to-one with the $z_1$ and $z_2$ coordinates, respectively, and therefore yield a parameterization of the underlying manifold (see Figure~\ref{subfig:strip_efuncs}).
 %
@@
 
-\begin{figure}[t]
-\centering
-\begin{subfigure}{0.45\textwidth}
-\includegraphics[width=\textwidth]{strip_discrete1}
-\caption{}
-\end{subfigure}
-%
-\begin{subfigure}{0.45\textwidth}
-\includegraphics[width=\textwidth]{strip_discrete4}
-\caption{}
-\end{subfigure}
-
-\begin{subfigure}{0.45\textwidth}
-\includegraphics[width=\textwidth]{strip_nonuniform1}
-\caption{}
-\end{subfigure}
-%
-\begin{subfigure}{0.45\textwidth}
-\includegraphics[width=\textwidth]{strip_nonuniform2}
-\caption{}
-\end{subfigure}
-\caption{(a, b) Two-dimensional strip with uniform sampling colored by the (a) first, and (b) fourth (non-trivial) eigenvector from diffusion maps.
-(c, d) Two-dimensional strip with data sampled from a Gaussian distribution in $z_1$ and sampled uniformly in $z_2$, colored by the (c) first, and (d) third (non-trivial) eigenvector from diffusion maps.
-Note that in both cases we uncover parameterizations which are one-to-one with $z_1$ and $z_2$.
-}
-\label{fig:strip_evecs}
-\end{figure}
-
 In most applications, we are not given a description of the continuous manifold. 
@@
 %
-It was shown in \cite{...} that, in the limit of infinite data, the discrete Laplacian constructed from data converges pointwise to the continuous Laplace-Beltrami operator on the manifold. 
-%
-As a result, the eigenvectors of the discrete Laplacian approximate the eigenfunctions of this continuous operator \cite{...}.
-
-Given observations $z(1), \dots, z(m) \in \mathcal{M}_d$, we construct the matrix $W \in \mathbb{R}^{m \times m}$, with
+It was shown in \cite{coifman2006geometric} that, in the limit of infinite data, the discrete Laplacian matrix constructed from data converges pointwise to the continuous Laplace-Beltrami operator on the manifold.
+%
+As a result, the eigenvectors of the discrete Laplacian approximate the eigenfunctions of this continuous operator.
+
+Given observations $z(1), \dots, z(m) \in \mathcal{M}_d$, we first construct the weight matrix $W \in \mathbb{R}^{m \times m}$, with
 \begin{equation} \label{eq:W}
@@
 %
-In the limit of infinite data uniformly sampled from $\mathcal{M}_d$ ($m \rightarrow \infty$ and $\epsilon \rightarrow 0$ with the appropriate rate), the matrix $I-A$ approaches the Laplace-Beltrami operator on the manifold $\mathcal{M}_d$ \cite{coifman2006geometric}.
+In the limit of infinite data uniformly sampled from $\mathcal{M}_d$ ($m \rightarrow \infty$ and $\epsilon \rightarrow 0$ with the appropriate rate), the discrete Laplacian matrix $I-A$ approaches the Laplace-Beltrami operator on the manifold $\mathcal{M}_d$ \cite{coifman2006geometric}.
 %
@@
 \end{equation}
-%
 
@@
 If the data $z(1), \dots, z(m)$ are sampled from $\mathcal{M}_d$ with some density $q$, then the matrix $I-A$ approximates the Fokker-Planck operator, 
-\begin{equation}
+$%\begin{equation}
 (I-A) \phi \rightarrow \nabla^2 \phi - \nabla U \cdot \nabla \phi, 
-\end{equation}
+$ %\end{equation}
 where $U = -2 \log q$. 
@@
 
-As discussed previously, the eigenfunctions provide a parameterization of the manifold, such that $\phi_{j}(i)$ yields the $j^{th}$ embedding coordinate for $z(i)$; this embedding is called diffusion maps \cite{...}.
-%
-We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$.%, and if the data do lie on a low-dimensional manifold, we only need to retain a few of the leading eigenvectors to adequately describe the data.
+As discussed previously, the eigenfunctions provide a parameterization of the manifold, such that $\phi_{j}(i)$ yields the $j^{th}$ embedding coordinate for $z(i)$; this embedding is called diffusion maps \cite{coifman2005geometric, coifman2006geometric}.
+%
+We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$.
 %
@@
 %
-Figure~\ref{fig:strip_evecs} shows data sampled from a strip, colored by eigenvectors of $A$. 
+Figures~\ref{subfig:strip_evecs_uniform}--\ref{subfig:strip_evecs_nonuniform} shows data sampled from a strip, colored by eigenvectors of $A$. 
 %
@@
 %
-From \eqref{eq:evals}, we see that sorting the eigenvectors by the magnitude of the corresponding eigenvalues implies that these eigenvectors are guaranteed to appear before any coupled or repeated eigendirections.
-%
-However, these two eigenvectors are {\em not} guaranteed to appear as the first two (non-trivial) eigenvectors, as harmonics of the first eigendirection (i.e., $\cos \left( n \pi z_1 / L_1 \right)$ with $n > 1$) could appear before the second.
-%
-In practice, identifying which eigenvectors correspond to unique eigendirections is nontrivial, as the correspondence between the original data and the coordinates which parameterize the manifold are unknown.
-%
-We note that, in contrast to our simple illustrative example, for most data sets of interest, the coordinates for the underlying manifold cannot easily be obtained from the coordinates of the original data.
+From \eqref{eq:evals}, we see that sorting the eigenvectors by the magnitude of the corresponding eigenvalues implies that these two eigenvectors are guaranteed to appear before any coupled or repeated eigendirections.
+%
+However, these eigenvectors are {\em not} guaranteed to appear as the first two (non-trivial) eigenvectors, as harmonics of the first eigendirection (i.e., $\cos \left( n \pi z_1 / L_1 \right)$ with $n > 1$) could appear before the second.
+%
+We note that, in contrast to our simple illustrative example, for most data sets of interest, the coordinates for the underlying manifold are unknown and cannot easily be obtained from the coordinates of the original data, and identifying which eigenvectors correspond to unique eigendirections is nontrivial.
 
@@
 %
-For the case where $\mathcal{M}_d$ is a $2$-dimensional strip with edge lengths $L_1  > L_2$,
-recall that the eigenfunctions $\tilde{\phi}_{1,0} = \cos \left(  {\pi z_1}/{L_1} \right)$ and  $\tilde{\phi}_{0,1} = \cos \left(  {\pi z_2}/{L_2} \right)$ provide embedding coordinates for the manifold $\mathcal{M}_d$.
-%
-However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to correspond to the two smallest (non-trivial) eigenvalues (the largest eigenvalue will always be $\tilde{\mu}_{0,0} = 1$ and correspond to a constant eigenfunction; these eigenvalues are related to the eigenvalues $\mu_k$ of the discrete operator via \eqref{eq:evals_relationship}).
+For the case where $\mathcal{M}_d$ is a $2$-dimensional strip with edge lengths $L_1  > L_2$, recall that the eigenfunctions $\tilde{\phi}_{1,0} = \cos \left(  {\pi z_1}/{L_1} \right)$ and  $\tilde{\phi}_{0,1} = \cos \left(  {\pi z_2}/{L_2} \right)$ provide embedding coordinates for the manifold $\mathcal{M}_d$.
+%
+However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to correspond to the two smallest (non-trivial) eigenvalues (the smallest eigenvalue will always be $\tilde{\mu}_{0,0} = 0$ and correspond to a constant eigenfunction; these eigenvalues are related to the eigenvalues $\mu_k$ of the discrete operator via \eqref{eq:evals_relationship}).
 %
@@
 \centering
-\begin{subfigure}{0.45\textwidth}
+\begin{subfigure}{0.24\textwidth}
 \includegraphics[width=\textwidth]{strip_discrete1}
@@
 %
-\begin{subfigure}{0.45\textwidth}
+\begin{subfigure}{0.24\textwidth}
 \includegraphics[width=\textwidth]{strip_discrete2}
 \end{subfigure}
-
-\begin{subfigure}{0.45\textwidth}
+%
+\begin{subfigure}{0.24\textwidth}
 \includegraphics[width=\textwidth]{strip_discrete3}
@@
 %
-\begin{subfigure}{0.45\textwidth}
+\begin{subfigure}{0.24\textwidth}
 \includegraphics[width=\textwidth]{strip_discrete4}
@@
 %
-This problem was addressed previously in \cite{...} by performing successive iterations of diffusion maps, interspersed with advection along the first eigendirection at each iteration. 
-%
-However, both the advection procedure and the successive eigendecompositions are very expensive and intractable for larger data set. 
-TODO: elaborate on their method and find other advantages of our proposed method (for example, theirs is ad hoc). 
+This problem was addressed previously in \cite{gerber2007robust} by performing successive iterations of diffusion maps, interspersed with advection along the first eigendirection at each iteration. 
+%
+However, this algorithm is {\em ad hoc}, and both the advection procedure and the successive eigendecompositions are very expensive and intractable for larger data set. 
+%TODO: elaborate on their method and find other advantages of our proposed method (for example, theirs is ad hoc). 
 %
@@
 %
-We will therefore attempt to fit a function $f(\phi_1, \dots, \phi_{k-1})$ to $\phi_{k}$. 
-%
-If the resulting fit is accurate, we will assume $\phi_{k}$ is a repeated eigendirection of $\phi_1, \dots, \phi_{k-1}$. 
-%
-We will use a local linear function 
+We therefore attempt to fit a function $f(\phi_1, \dots, \phi_{k-1})$ to $\phi_{k}$; if the resulting fit is accurate, we assume $\phi_{k}$ is a repeated eigendirection of $\phi_1, \dots, \phi_{k-1}$.
+%
+We use a local linear function 
 \begin{equation}
@@
 %
-We will use a Gaussian kernel, 
+We use a Gaussian kernel, 
 %
@@
 %
-We will assume this implies that $\phi_k$ is a harmonic of previous modes, i.e., a repeated eigendirction, and conversely, a large value of $r_{k}$ indicates that $\phi_{k}$ parameterizes a unique eigendirection in the data.
+We assume small value of $r_k$ implies that $\phi_k$ is a harmonic of previous modes, i.e., a repeated eigendirction, and conversely, a large value of $r_{k}$ indicates that $\phi_{k}$ parameterizes a unique eigendirection in the data.
 %
@@
 %
-We then calculate the smoothing matrix $L_{k-1} \in \mathbb{R}^{m \times m}$, where 
-\begin{equation}
-L_{k-1} =
+We then calculate the smoothing matrix $S_{k-1} \in \mathbb{R}^{m \times m}$, where 
+\begin{equation}
+S_{k-1} =
 \begin{bmatrix}
@@
 \begin{equation} 
-r_{k} = \sqrt{ \frac{\sum_{i=1}^n \left( \frac{ \phi_{k} (i) - \sum_{j=1}^m L_{k-1}(i,j) \phi_{k}(j) }{1-L_{k-1}(i,i)} \right)^2} {\sum_{i=1}^n  \left( \phi_{k} (i) \right)^2 }} .
+r_{k} = \sqrt{ \frac{\sum_{i=1}^n \left( \frac{ \phi_{k} (i) - \sum_{j=1}^m S_{k-1}(i,j) \phi_{k}(j) }{1-S_{k-1}(i,i)} \right)^2} {\sum_{i=1}^n  \left( \phi_{k} (i) \right)^2 }} .
 \end{equation}
@@
 %
-For more general manifolds, we presume that the eigenvalues can still provide a measure of the relative significance of the unique eigendirections.
+For more general manifolds, we postulate that the eigenvalues can still provide a measure of the relative significance of the unique eigendirections.
 %
@@
 %
-These lengths are used to determine how many components are required to retain most of the information within the data set; 
+These lengths can then be used to determine how many components are required to retain most of the information within the data set; 
 for example, if $L_j$ is smaller than a predefined threshold, then this component can be disregarded.
@@
 
-We will demonstrate our proposed approach on three synthetic data sets. 
+We demonstrate our proposed approach on three synthetic data sets. 
 %
@@
 
-\begin{figure}[t]
-\centering
-\begin{subfigure}{0.3\textwidth}
+\begin{figure}[!t]
+\centering
+\begin{subfigure}{0.25\textwidth}
 \includegraphics[height=2.5cm]{strip_data_L2}
@@
 %
-\hfill
-%
-\begin{subfigure}{0.3\textwidth}
+%\hfill
+%
+\begin{subfigure}{0.25\textwidth}
 \includegraphics[height=2.5cm]{strip_data_L4}
@@
 %
-\hfill
-%
-\begin{subfigure}{0.35\textwidth}
+%\hfill
+%
+\begin{subfigure}{0.3\textwidth}
 \includegraphics[height=2.5cm]{strip_data_L8}
@@
 %
-\caption{Data sets (top) and eigenvalue spectra (bottom) for strips with (a) $L_1 = 2$, $L_2 = 1$, (b) $L_1 = 4$, $L_2 = 1$, (c) $L_1 = 8$, $L_2 = 1$.
+\caption{Data sets (top) and eigenvalue spectra from diffusion maps analysis (bottom) for strips with (a) $L_1 = 2$, $L_2 = 1$, (b) $L_1 = 4$, $L_2 = 1$, (c) $L_1 = 8$, $L_2 = 1$.
 The empirical eigenvalues are plotted in blue, and the analytical eigenvalues are plotted in red.
-From the eigenvalues which are identified as important/non-harmonics, the estimated length ratio is (a) 2.2, (b) 4.1, (c) 8.7.}
+From the eigenvalues which are identified as parameterizing unique eigendirections (indicated by the darker blue bars), the estimated length ratio is (a) 2.2, (b) 4.1, (c) 8.7.}
 \label{fig:strip_compare_analytic}
@@
 %
-Figure~\ref{fig:strip_compare_analytic} shows the data sets and corresponding eigenspectra.
+Figure~\ref{fig:strip_compare_analytic} shows the data sets and the eigenspectra of the corresponding discrete Laplacians.
 %
@@
 %
-The analytic eigenvalues of the Laplacian (see \eqref{eq:evals} and \eqref{eq:evals_relationship}) are also indicated in red.
-
-We observe that the eigenvalues are consistent with the known analytic eigenvalues of the Laplacian.
+
+We observe that the eigenvalues are consistent with the known analytic eigenvalues of the Laplacian (see \eqref{eq:evals} and \eqref{eq:evals_relationship}), shown in red.
 %
@@
 
-\begin{figure}[!th]
-%
-\begin{subfigure}{0.45\textwidth}
-\centering
-\includegraphics[height=1.25in]{swissroll1}
+\begin{figure}[!t]
+%
+\begin{subfigure}{0.2\textwidth}
+\centering
+\includegraphics[width=\textwidth]{swissroll1}
 \caption{}
@@
 \end{subfigure}
-%\hfill
-\begin{subfigure}{0.45\textwidth}
-\centering
-\includegraphics[height=1.25in]{swissroll2}
+%
+\begin{subfigure}{0.25\textwidth}
+\centering
+\includegraphics[width=\textwidth]{swissroll1_evals}
+\caption{}
+\label{subfig:swissroll1_evals}
+\end{subfigure}
+%
+\begin{subfigure}{0.25\textwidth}
+\centering
+\includegraphics[width=\textwidth]{swissroll1_color1} 
+\caption{}
+\label{subfig:swissroll1_color1}
+\end{subfigure}
+%
+\begin{subfigure}{0.25\textwidth}
+\centering
+\includegraphics[width=\textwidth]{swissroll1_color2}
+\caption{}
+\label{subfig:swissroll1_color2}
+\end{subfigure}
+
+\begin{subfigure}{0.2\textwidth}
+\centering
+\includegraphics[width=\textwidth]{swissroll2}
 \caption{}
 \label{subfig:swissroll2}
-\end{subfigure}
-%\hfill
-\begin{subfigure}{0.45\textwidth}
-\centering
-\includegraphics[height=1.25in]{swissroll1_evals}
-\caption{}
-\label{subfig:swissroll1_evals}
-\end{subfigure}
-%\hfill
-\begin{subfigure}{0.45\textwidth}
-\centering
-\includegraphics[height=1.25in]{swissroll2_evals}
+\end{subfigure}%
+%
+\begin{subfigure}{0.25\textwidth}
+\centering
+\includegraphics[width=\textwidth]{swissroll2_evals}
 \caption{}
@@
 \end{subfigure}
-%\hfill
-\begin{subfigure}{0.45\textwidth}
-\centering
-\includegraphics[height=1.25in]{swissroll1_color1} 
-\caption{}
-\label{subfig:swissroll1_color1}
-\end{subfigure}
-\hfill
-\begin{subfigure}{0.45\textwidth}
-\centering
-\includegraphics[height=1.25in]{swissroll2_color1}
+%
+\begin{subfigure}{0.25\textwidth}
+\centering
+\includegraphics[width=\textwidth]{swissroll2_color1}
 \caption{}
@@
 \end{subfigure}
-\hfill
-\begin{subfigure}{0.45\textwidth}
-\centering
-\includegraphics[height=1.25in]{swissroll1_color2}
-\caption{}
-\label{subfig:swissroll1_color2}
-\end{subfigure}
-\hfill
-\begin{subfigure}{0.45\textwidth}
-\centering
-\includegraphics[height=1.25in]{swissroll2_color2}
+%
+\begin{subfigure}{0.25\textwidth}
+\centering
+\includegraphics[width=\textwidth]{swissroll2_color2}
 \caption{}
@@
 (a) Data set 1; $h= 40$.
-(b) Data set 2; $h = 20$.
-(c) Eigenvalue spectrum from the analysis of data set 1.
-(d) Eigenvalue spectrum from the analysis data set 2.
-(e) Data set 1, colored by the first eigenvector.
-(f) Data set 2, colored by the first eigenvector.
-(g) Data set 1, colored by the second eigenvector.
-(h) Data set 2, colored by the fifth eigenvector.
+(b) Eigenvalue spectrum from the diffusion maps analysis of data set 1.
+(c) Data set 1, colored by the first diffusion maps eigenvector.
+(d) Data set 1, colored by the second diffusion maps eigenvector.
+(e) Data set 2; $h = 20$.
+(f) Eigenvalue spectrum from the diffusion maps analysis of data set 2.
+(g) Data set 2, colored by the first diffusion maps eigenvector.
+(h) Data set 2, colored by the fifth diffusion maps eigenvector.
 }
@@
 %
-Note that, in this example, $z_1, z_2, z_3$ are the original coordinates; however, the data lie on a two-dimensional manifold parameterized by the $\theta$ and $t$. 
+Note that, in this example, $z_1, z_2, z_3$ are the original data coordinates; however, the data lie on a two-dimensional manifold parameterized by the $\theta$ and $t$. 
 %
@@
 
-Figures~\ref{subfig:swissroll1_evals}--\ref{subfig:swissroll2_evals} shows the eigenvalue spectra from the analysis of the two data sets.
+Figures~\ref{subfig:swissroll1_evals}~and~\ref{subfig:swissroll2_evals} shows the eigenvalue spectra from the analysis of the two data sets.
 %
@@
 %
-Figures~\ref{subfig:swissroll1_color1}--\ref{subfig:swissroll2_color2} shows the two data sets, colored by the two eigenvectors identified by our method as parameterizing the unique eigendirections. 
+Figures~\ref{subfig:swissroll1_color1},\ref{subfig:swissroll1_color2},\ref{subfig:swissroll2_color1},~and~\ref{subfig:swissroll2_color2} shows the two data sets, colored by the two eigenvectors identified as parameterizing the unique eigendirections.
 %
@@
 \begin{figure}[t]
-%
-\begin{subfigure}{0.3\textwidth}
+\centering
+\begin{subfigure}{1.5in}
 \centering
@@
 %
-\hfill
-%
-\begin{subfigure}{0.32\textwidth}
+%\hfill
+%
+\begin{subfigure}{1.5in}
 \centering
@@
 %
-\hfill
-%
-\begin{subfigure}{0.35\textwidth}
+%\hfill
+%
+\begin{subfigure}{1.5in}
 \centering
@@
 %
-\caption{Torii data sets (left) and corresponding eigenvalues (right).
+\caption{Torii data sets (top, see \eqref{eq:torus}) and corresponding diffusion maps eigenvalues (bottom).
 (a) $r_1 = 3$, $r_2 = 1$.
@@
 \end{aligned}
+\label{eq:torus}
 \end{equation}
@@
 %
-Intuitively, increasing $r_1$ makes the eigendirections which parameterize the outer circle ($\cos \theta_1$ and $\sin \theta_1$) more dominant compared to the eigendirections with parameterize the inner circle.
-%
-Increasing $r_1$ can be viewed as analogous to increasing the ratio of $L_1$ to $L_2$ in the strip example.
-
+Intuitively, increasing $r_1$ can be viewed as analogous to increasing the ratio of $L_1$ to $L_2$ in the strip example and makes the eigendirections which parameterize the outer circle ($\cos \theta_1$ and $\sin \theta_1$) more dominant compared to the eigendirections with parameterize the inner circle.
+%
 Figure~\ref{fig:torus} shows the eigenspectra from the analysis of three different torii, colored by the leave-one-out cross-validation error. 
@@
 %
-%Although on the microscale the system consists of many stochastic and noisy particles, it exhibits coherent dynamics on the macroscale which are governed by only a few parameters.
-%
 In general, this mapping from microscale to macroscale is not always obvious.
@@
 
-%For example, fluid flow, which can be described via the position and velocity of every fluid particle, is modeled at the macroscale using the Navier-Stokes equations.
-%
-
-%In this letter, we will demonstrate how this goal can be achieved through a data-driven method based on geometric analysis and manifold learning. 
-%
-%Specifically, we aim to emphasize two particular aspects of manifold learning from a dynamical systems' point of view: (a) choosing the appropriate observers, especially in presence of stochastic dynamics, and (b) selecting an appropriate metric for comparison \cite{mallat2012group}.
-%
-%We will show that both are essential for data-driven algorithms to successfully uncover the underlying system dynamics.
-
 Our model problem describes the process of cellular chemotaxis \cite{othmer2000diffusion}, where biological cells exhibit coherent macroscopic dynamics regulated by extracellular sensed signals in order to accomplish tasks such as finding food or navigating away from toxins.
@@
 %
-%In this model, each cell is initialized with a position and a velocity on a line, and the dynamics of each cell are governed by a stochastic process.
-%
-%At random times, a cell will ``turn around'' and switch the direction of its velocity (this turning is controlled by extracellular signals). 
-%
-%We will analyze the dynamics of collections of such cells/particles. 
-%
-%In this example, the macroscopic dynamics of the group of cells vary depending on the value of a single system parameter.
-%
-%We will demonstrate how we can use diffusion maps \cite{coifman2005geometric}, a manifold learning technique, to {\em automatically} detect changes in the system's macroscopic dynamics from microscale data.
-%
-%This case study exhibits several key points which allow us to demonstrate the strength of our diffusion maps approach.
-%
-%First, this system has random microscopic behavior with smooth macroscopic dynamics governed by a single parameter which determines the regime/mode of the system. 
-%
-%Second, it consists of an ensemble of individual stochastic particles which must be described using the appropriate observers.
-%
-%Third, it has an analytic macroscopic description, which serves as a ground truth and will allow us to verify our results which are obtained in an unsupervised manner.
 This specific example has an analytic macroscopic description in which the macroscopic dynamics of the group of cells depend on the value of a single system parameter.
@@
 Furthermore, we will show that changes in dynamical behavior resulting from changes in the system parameters can be automatically detected.
-
-%The remainder of this letter is organized as follows. 
-%
-%We begin by describing the chemotaxis problem, the illustrative example used to demonstrate the methodologies.
-%%
-%We present the stochastic microscopic model, and use diffusion maps, a nonlinear dimensionality reduction technique, to analyze microscopic simulation results.
-%%
-%We show that the governing macroscopic variables can be recovered, provided that the appropriate observers and distance metric are used.
-%%
-%We then describe how, for this example, the microscopic model gives rise to a compact macroscopic description governed by a single parameter, and show that the diffusion maps results are consistent with this macroscopic description.
-%%
-%Finally, we demonstrate that diffusion maps can also identify the relative importance of system variables in different parameter regimes, and detect changes in macroscopic dynamical behavior.
-
-%\section{Problem formulation}
-
-%\subsection{Chemotaxis ``story''}
 
@@
 %
-For our specific simulations, we set $s^2/\lambda=1$, which is consistent with the analysis presented in \cite{...}. 
+For our specific simulations, we set $s^2/\lambda=1$, which is consistent with the analysis presented in \cite{othmer1988models}. 
 %
@@
 
-\subsection{Observers and metrics}
-
-In general, manifold learning techniques have two essential components.
-%
-One is the appropriate {\em observers} of the system. 
-%
-These observers should be informative as to the state of the system, as well as insensitive to noise in the system. 
-%
-Two is a {\em distance metric} between the observations that captures a notion of locality: observations which we perceive to be similar should have a small distance. 
-%
-%For this particular example, we will use histograms of the particle positions as the observations, and the EMD as the distance metric.
-
-%Appropriate observers can be constructed in several different ways. 
-%%
-%For simple data, such as the illustrative examples in Section~\ref{sec:illustrative_examples}, the observers are simply the data points themselves. 
-%%
-%However, for more complex data, the raw data may contain other sources of variability which are not meaningful to the system dynamics. 
-%%
-%For some examples, physical intuition and {\em a priori} knowledge can aid in constructing appropriate observers. 
-%%
-%For example, when looking at a protein, the observers may be the locations of the $\alpha$-carbons, as these backbone atoms are known to capture most of the relevant dynamics. 
-%%
-%Observers can also remove unimportant sources of variation; 
-%for example, the bispectrum \cite{zhao2014rotationally} and the scattering transform \cite{mallat2012group} are both invariant to rotations. 
-%%
-%Observers should also be robust to noise, which motivates the scattering transform \cite{mallat2012group} and histograms \cite{talmon2013empirical}.
-%
-%After constructing appropriate observers, one must choose the appropriate metric for the diffusion maps calculation. 
-%%
-%The simplest option (used for the illustrative examples in Section~\ref{sec:illustrative_examples}) is to use the Euclidean distance, but this can often prove uninformative. 
-%%
-%For molecular simulation data, it is often appropriate to use the Euclidean distance {\em after} aligning configurations with respect to rotations, as global rotations of the entire molecule are not meaningful \cite{ferguson2010systematic}.
-%%
-%For dynamical systems which are obscured by large variance noise, the Mahalanobis distance often yields a more informative parameterization of the data \cite{talmon2013empirical}.
-%%
-%Another option is the Earth Mover's Distance, a multiscale distance metric which is stable and robust to noise \cite{rubner2000earth}.
-
-For the chemotaxis example, we use histograms of the particle positions as observers. 
-%
-Histograms are invariant to the indexing of the particles, while retaining the spatial locations of the particles.
-%
-The histograms are also robust to noise \cite{talmon2013empirical}. 
-%
-Instead of the standard Euclidean distance, we use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between pairs of histograms. 
-%
-Conceptually, the EMD measures how much ``work'' it takes to transform one probability density into another.
-%
-It therefore not only considers where the densities are inconsistent, but also how far apart the inconsistencies are.
-%
-Although the brute-force computation of the EMD is computationally expensive, there has been a plethora of work in developing efficient algorithms for its approximation \cite{Pele-eccv2008, Pele-iccv2009}.
-%
-For the special case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions of the data \cite{rubner2000perceptual}, which can be approximated from histograms as
-\begin{equation}
-\| z(i) - z(j) \|_{EMD} \approx \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|,
-\end{equation}
-where the histograms $z(i), z(j)$ are defined on $n$ equally-spaced bins in $\mathbb{R}$ (for our specific simulations, we will take $n=32$), and $z_k$ denotes the $k^{th}$ bin. 
-%
-We note that although histograms are not essential for using EMD, they make the computation feasible. 
-%
-
-%\subsubsection{Constructing appropriate observers}
-%
-%\begin{itemize}
-%\item Physical examples (e.g., $\alpha$-carbons for molecules)
-%\item Amit's bispectrum
-%\item Scattering transform
-%\item Cite PNAS, stating that the histograms are good since they linearize any noise. 
-%For the chemotaxis example, we use histograms of the particle positions as observers \cite{talmon2013empirical}. 
-%%
-%Histograms are invariant to the indexing of the particles, while retaining the spatial locations of the particles.
-%\end{itemize}
-%
-%\subsubsection{Defining a meaningful metric}
-%
-%
-%\begin{itemize}
-%\item Euclidean (not good, especially for histograms)
-%\item Mahalanobis (good, cite PNAS, since it noise robust)
-%\item Here we use EMD since it is multiscale.
-%%
-%Instead of the standard Euclidean distance, we use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between pairs of histograms. 
-%%
-%Conceptually, the EMD measures how much ``work'' it takes to transform one probability density into another.
-%%
-%It therefore not only considers where the densities are inconsistent, but also how far apart the inconsistencies are.
-%%
-%Although the brute-force computation of the EMD is computationally expensive, there has been a plethora of work in developing efficient algorithms for its computation \cite{Pele-eccv2008, Pele-iccv2009}.
-%%
-%For the special case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions \cite{rubner2000perceptual}, which can be estimated from histograms as
-%\begin{equation}
-%\| z(i) - z(j) \|_{EMD} \approx \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|
-%\end{equation}
-%where the histograms $z(i), z(j)$ are defined on equally-spaced bins in $\mathbb{R}$, and $z_k$ denotes the $k^{th}$ bin. 
-%%
-%\item Emphasize stability and regularity issues.
-%\end{itemize}
-
-\subsection{Results}
-
-\subsubsection{Identifying the unique eigendirections}
-
 \begin{figure}[t!]
-\def \figwidth {3.5cm}
-
+\def \figwidth {0.25\textwidth}
 \centering
@@
 \end{subfigure}
-\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with (a--c) $\lambda=1$, $s=1$, and (d--f) $\lambda=400$, $s=20$.
+\caption{Embeddings using the first two eigenvectors of the (unnormalized) Laplacian $A$ computed from simulation data of the velocity jump process with (a--c) $\lambda=1$, $s=1$, and (d--f) $\lambda=400$, $s=20$.
 Each data set consists of $10$ simulations, with initial conditions uniformly chosen such that $0.1 \le p  \le 0.9$.
@@
 
-\begin{figure}
-%
-\def\figheight{1.4in}
-%
-\begin{subfigure}[t]{0.31\textwidth}
+\subsection{Observers and metrics}
+
+In general, manifold learning techniques have two essential components.
+%
+One is the appropriate {\em observers} of the system. 
+%
+These observers should be informative as to the state of the system, as well as insensitive to noise in the system. 
+%
+Two is a {\em distance metric} between the observations that captures a notion of locality: observations which we perceive to be similar should have a small distance.
+
+For the chemotaxis example, we use histograms of the particle positions as observers. 
+%
+Histograms are invariant to the indexing of the particles, while retaining the spatial locations of the particles.
+%
+The histograms are also robust to noise \cite{talmon2013empirical}. 
+%
+Instead of the standard Euclidean distance, we use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between pairs of histograms. 
+%
+Conceptually, the EMD measures how much ``work'' it takes to transform one probability density into another.
+%
+It therefore not only considers where the densities are inconsistent, but also how far apart the inconsistencies are.
+%
+Although the brute-force computation of the EMD is computationally expensive, there has been a plethora of work in developing efficient algorithms for its approximation \cite{Pele-eccv2008, Pele-iccv2009}.
+%
+For the special case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions of the data \cite{rubner2000perceptual}, which can be approximated from histograms as
+\begin{equation}
+\| z(i) - z(j) \|_{EMD} \approx \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|,
+\end{equation}
+where the histograms $z(i), z(j)$ are defined on $n$ equally-spaced bins in $\mathbb{R}$ (for our specific simulations, we will take $n=32$), and $z_k$ denotes the $k^{th}$ bin. 
+%
+We note that although histograms are not essential for using EMD, they make the computation feasible.
+
+\subsection{Results}
+
+\subsubsection{Identifying the unique eigendirections}
+
+\begin{figure}[t]
+\centering
+\def\figheight{1.3in}
+%
+\begin{subfigure}[t]{1.5in}
 \centering
@@
 %
-\begin{subfigure}[t]{0.31\textwidth}
+\begin{subfigure}[t]{1.5in}
 \centering
@@
 %
-\begin{subfigure}[t]{0.35\textwidth}
+\begin{subfigure}[t]{2in}
 \centering
@@
 %
-TODO: should we mention fluxes here??
+%TODO: should we mention fluxes here??
 
@@
 It now becomes clear that looking at the eigenvectors {\em modulo} repeated eigendirections is essential for extracting an informative parameterization of the data, as well as characterizing the dimensionality of the data.
-
-%\subsubsection{Detecting changes in dynamical behavior}
-%
-%
-%%\begin{figure}[t] 
-%%\centering
-%%\includegraphics[width=0.5\textwidth]{detect_change_eigenvalues}
-%%\caption{The first two eigenvalues from our analysis as a function of $\lambda$.
-Each point is averaged over $10$ experiments (as in Figure~\ref{fig:dmaps_embed_emd}, one experiment consists of data collected at varying $p$ and $t$, for a fixed value of $\lambda$), with error bars denoting one standard deviation.
-%%The convergence of the eigenvalues at $\lambda \approx 125$ indicates switching of the dynamical ``mode''.}
-%%\label{fig:detect_change}
-%%\end{figure}
-%
-%%We have seen that the eigenvalues $\mu_i$, which quantify the importance of each embedding coordinate, reveal the relative importance of the variables which govern the macroscopic dynamics.
-%%
-%We can use the eigenvalues from our analysis to detect changes in dynamical behavior.
-%%
-%From Figure~\ref{fig:dmaps_embed_emd}, we can see that the dominant eigendirection is correlated with $p$ when $\lambda$ is small, and correlated with $t$ when $\lambda$ is large. 
-%%
-%%We expect an intermediate $\lambda$ value where $p$ and $t$ are equally important. 
-%%%
-%%From the analytic macroscopic description, we expect a gap between the two eigenvalues at the two asymptotic regimes ($\lambda \rightarrow 0$, where the dynamics are described by two uncoupled wave equations \eqref{...}, and $\lambda \rightarrow \infty$, where the dynamics are described by a single heat equation \eqref{...}), as one of the two macroscopic variables is significantly more important to the dynamical behavior.
-%%%
-%%For intermediate $\lambda$ values, we expect the gap between $|\mu_1|$ and $|\mu_2|$ to narrow, with the $\lambda$ value where $|\mu_1| \approx |\mu_2|$ corresponding to the transition between ``wave-like'' and ``heat-like'' behavior.
-%%
-%We performed a new set of simulations over a range of $\lambda$ values;  
-%Figure~\ref{fig:detect_change} shows the first two eigenvalues, $\mu_1$ and $\mu_2$, from our analysis as a function of $\lambda$.
-%%
-%For small $\lambda$ (corresponding to Figures~\ref{subfig:small_lambda_p}~and~\ref{subfig:small_lambda_t}), the two eigenvalues are well-separated, as the first mode (which parameterizes $p$) is significantly more important than the second mode (which parameterizes $t$).
-%%
-%At intermediate $\lambda$ values, the eigenvalues come together as the system dynamics change and both $p$ and $t$ are of similar importance.
-%%
-%The eigenvalues then separate again at large $\lambda$ (corresponding to Figures~\ref{subfig:large_lambda_p}~and~\ref{subfig:large_lambda_t}), where the $t$ effects dominate the dynamics.
-%%
-%We can therefore detect changes in dynamical behavior using data-driven techniques, without any knowledge of the governing macroscopic model. 
-%%
-%In the next section, we will show that the detected changes in dynamical behavior is consistent with the analytic macroscale model.
-%
-%%\begin{figure}[t]
-%%\def\figwidth{4.2cm}
-%%\begin{subfigure}{\figwidth}
-%%\includegraphics[width=\textwidth]{rawhist_p_1}
-%%\caption{}
-%%\end{subfigure}
-%%\begin{subfigure}{\figwidth}
-%%\includegraphics[width=\textwidth]{rawhist_t_1}
-%%\caption{}
-%%\end{subfigure}
-%%\begin{subfigure}{\figwidth}
-%%\includegraphics[width=\textwidth]{rawhist_p_400}
-%%\caption{}
-%%\end{subfigure}
-%%\begin{subfigure}{\figwidth}
-%%\includegraphics[width=\textwidth]{rawhist_t_400}
-%%\caption{}
-%%\end{subfigure}
-%%\caption{Diffusion maps embeddings computed from simulation data of the velocity jump process with (a,b) $\lambda=1$, $s=1$, and (c,d) $\lambda=400$, $s=20$.
-The distances used in the diffusion maps kernel are the Euclidean distances between the histograms of particle positions.
-The data are colored by (a, c) $p$, the initial probability of a particle moving to the right, and (b, d) $t$, time.}
-%%\label{fig:dmaps_embed_noemd}
-%%\end{figure}
 
@@
 %
-%We assume that $s^2/\lambda = D$ is constant, so that the dynamics of the probability density are governed by a {\em single} parameter $\lambda$.
-%
 From \eqref{eq:second_order_pde}, we see that, for fixed values of $\lambda$ and $s$, the macroscopic state of the system (the probability density of the cells) is indeed a function of two variables $p$ and $t$.
@@
 \end{equation}
-%alternatively, \eqref{eq:second_order_pde} becomes the second order wave equation,
-%\begin{equation}
-%\frac{\partial^2 \rho}{\partial t^2} = s^2 \frac{\partial ^2 \rho}{\partial x^2}.
-%\end{equation}
-
-%Dividing \eqref{eq:second_order_pde} by $\lambda > 0$ yields
-%\[
-%\frac{1}{\lambda} \frac{\partial^2 \rho}{\partial t^2} + 2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2}
-%\]
+%
 When $\lambda \rightarrow \infty$ with $s^2/\lambda$ fixed, \eqref{eq:second_order_pde} approaches the heat equation,
@@
 %
-%Two variables determine the macroscopic state of the system: the initial distribution, which is controlled by a single parameter $p$, and the time, $t$.
-%
 The above analysis shows that the initial distribution of velocities of the particles (determined by $p$ in the microscopic simulations) plays a very different role depending on the value of $\lambda$.
@@
 When $\lambda \rightarrow \infty$, the dynamics are described by one heat equation, and the initial conditions are insignificant -- the velocity distribution quickly equilibrates and we see purely diffusive behavior.
-%
 
@@
 However, for large $\lambda$, the two densities rapidly equilibrate, and no organization based on the differences between the two densities is visible.
-
-%
-%We will demonstrate how we can use diffusion maps to parameterize the simulation data, as well as detect the effective dimensionality of the data.
-%
-%We expect the simulation data to lie on a low-dimensional manifold which is parameterized by the macroscopic variables.
-%
-%We use diffusion maps \cite{coifman2005geometric}, which is a manifold learning technique, to uncover such low-dimensional structure and to analyze the results of the microscopic chemotaxis simulations.
-%
-%Compared to principal components analysis (PCA) \cite{shlens2005tutorial}, a standard linear technique, diffusion maps provides a parameterization of data which lie on a low-dimensional, (possibly) {\em nonlinear} structure in a high-dimensional space.
-%
-%Diffusion maps is one of many recently introduced manifold learning techniques that could be used to analyzed this data \cite{roweis2000nonlinear, tenenbaum2000global, Belkin2003}.
-
-%correlations
-%small lambda
-%Euclidean: 3.3848e-04, -0.1861
-%EMD: 0.9000, 0.9807
- %   
-%large lambda
-%Euclidean: -0.5466, 0.3329
-%EMD: 0.9329, 0.9832
 
@@
 %
-%For the specific chemotaxis example, we know that there are two directions in the data, corresponding to simulation parameters $p$ and $t$. 
-%%
-%However, for large values of $\lambda$, we know the dynamics become diffusive and $p$ becomes unimportant; we therefore expect the data to look approximately one-dimensional when $\lambda$ is large. 
-%
-%To detect this transition in the data, we look at the eigenvalues corresponding to the unique eigendirections. 
-%
 Let $\mu_{i_1} \ge \mu_{i_2}$ denote these two eigenvalues. 
@@
 when this ratio becomes small, the data are effectively one-dimensional, as the second dimension is very small compared with the first.
-%
 
@@
 %
-%We therefore expect the observed dimensionality to change from two to one as a function of $\lambda$. 
-%
-%For general dynamical systems, the governing macroscopic variables and asymptotic behavior may not be immediately obvious, and instead need to be inferred from the microscopic data.
-%
 For this specific example, these changes are consistent with the analytically-predicted shift at 
@@
 %
-%From \cite{othmer2000diffusion}, the relevant time scales for the velocity jump process are
-%%\begin{equation}
-%%\begin{aligned}
-%%\tau_{drift} =& \frac{L}{s} \\
-%%\tau_{diff} =& \frac{L^2 \lambda}{s^2}
-%%\end{aligned}
-%%\end{equation}
-%%%
-%%Here, we take $L = t_{max} s$ as a characteristic length scale for the simulations. 
-%%
-%\begin{equation}
-%\begin{aligned}
-%\tau_{drift} =& t_{max} \\
-%\tau_{diff} =& t_{max}^2 \lambda
-%\end{aligned}
-%\end{equation}
-%%
-%When $\frac{\tau_{diff}}{\tau_{drift}} =  t_{max}  \lambda \gg 1$, then the dynamics are effectively one-dimensional. 
-%%
 This ratio is also plotted in Figure~\ref{fig:chemotaxis_compare_timescales_evals} as a function of $\lambda$, and the transition is consistent with the transition predicted by the diffusion maps eigenvalues.

### 2015-03-13 chemotaxis/paper/draft1.tex
edited chemotaxis paper

@@
 %
+We propose using only the eigenvectors which correspond to unique eigendirections to embed the data, and show that this reduced diffusion maps embedding induces a metric which is equivalent to the standard diffusion distance.
+%
 We first demonstrate the utility and flexibility of our approach on synthetic data sets. 
@@
 
-As discussed previously, the eigenfunctions provide a parameterization of the manifold, such that $\phi_{j}(i)$ yields the $j^{th}$ embedding coordinate for $z(i)$; this embedding is called diffusion maps \cite{coifman2005geometric, coifman2006geometric}.
+As discussed previously, the eigenfunctions provide a parameterization of the manifold, such that $\phi_{j}(i)$ yields the $j^{th}$ embedding coordinate for $z(i)$; this is the standard diffusion maps embedding \cite{coifman2005geometric, coifman2006geometric}.
 %
@@
 
+We propose using {\em only} those eigenvectors for which $r_k$ is large to embed the data, as this will yield a more parsimonious representation of the data. 
+%
+However, it also preserves some features of the standard diffusion map embedding. 
+%
+In standard diffusion maps, we define the diffusion distance as
+\begin{equation}
+D^2(z(i), z(j)) = \sum_{k=0}^{m-1} \mu_k^t \left( \phi_k(i) - \phi_k(j)  \right)^2.
+\end{equation}
+%
+Let $I = \{i_1, i_2, \dots, i_d \}$ denote the indices of the unique eigenvectors along the manifold (i.e.,$r_{i_1}, \dots, r_{i_d}$ are large). 
+%
+Under the assumption that $\phi_j = f \left( \phi_{i_1}, \dots, \phi_{i_d} \right)$ for $j \not\in I$, and if $f$ is Lipshitz with Lipshitz constant $K$, then one can show that 
+\begin{equation}
+\sqrt{\frac{1}{1+K \sum_{k \not\in I} \mu_k^t}} D^2(z(i), z(j)) \le \sum_{k \in I} \mu_k^t \left( \phi_k(i) - \phi_k(j)  \right)^2 \le D^2(z(i), z(j)).
+\end{equation}
+%
+Therefore, provided the eigenvalues $\mu_k$ decay sufficiently fast, the distance induced by the $d$ eigenvectors identified as parameterizing unique eigendirections $\phi_{i_1}, \dots, \phi_{i_d}$ is equivalent to the standard diffusion distance.
+%
+We will refer to this distance 
+\begin{equation}
+\tilde{D}^2(z(i), z(j)) = \sum_{k \in I} \mu_k^t \left( \phi_k(i) - \phi_k(j)  \right)^2
+\end{equation}
+%
+as the {\em reduced} diffusion distance, and the embedding obtained from the corresponding eigenvectors as the reduced diffusion maps embedding. 
+%
+
 \subsection{Calculating the relative lengths of the manifold} \label{sec:relative_lengths}
@@
 %
-These lengths can then be used to determine how many components are required to retain most of the information within the data set; 
+These lengths can then be used to truncate the reduced diffusion maps embedding and determine how many components are required to retain most of the information within the data set; 
 for example, if $L_j$ is smaller than a predefined threshold, then this component can be disregarded.
@@
 however, our analysis is not restricted to such specific cases and more complex initial conditions could be used. 
+%
+Each data set consists of $10$ simulations, with initial conditions uniformly chosen such that $0.1 \le p  \le 0.9$. 
+%
+We allow each simulation to evolve for $t_{max}$ time units with a fixed time step $dt$, and use data with $t > 0$ for analysis.
 
@@
 \end{subfigure}
-\caption{Embeddings using the first two eigenvectors of the (unnormalized) Laplacian $A$ computed from simulation data of the velocity jump process with (a--c) $\lambda=1$, $s=1$, and (d--f) $\lambda=400$, $s=20$.
-Each data set consists of $10$ simulations, with initial conditions uniformly chosen such that $0.1 \le p  \le 0.9$.
-We allow each simulation to evolve for $10$ time units, with $dt=1$, and use data with $t > 0$ for analysis.
+\caption{Embeddings using the first two diffusion maps eigenvectors computed from simulation data of the velocity jump process with (a--c) $\lambda=1$, $s=1$, and (d--f) $\lambda=400$, $s=20$.
+We set $t_{max} = 10$ and $dt=1$.
 The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions.
@@
 %
-Histograms are invariant to the indexing of the particles, while retaining the spatial locations of the particles.
-%
-The histograms are also robust to noise \cite{talmon2013empirical}. 
+Histograms are invariant to the indexing of the particles, while retaining information about the spatial locations of the particles.
+%
+Histograms are also robust to noise \cite{talmon2013empirical}. 
 %
@@
 %
-For the special case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions of the data \cite{rubner2000perceptual}, which can be approximated from histograms as
+For the specific case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions of the data \cite{rubner2000perceptual}, which can be approximated from histograms as
 \begin{equation}
@@
 \end{equation}
-where the histograms $z(i), z(j)$ are defined on $n$ equally-spaced bins in $\mathbb{R}$ (for our specific simulations, we will take $n=32$), and $z_k$ denotes the $k^{th}$ bin. 
-%
-We note that although histograms are not essential for using EMD, they make the computation feasible.
+where the histograms $z(i), z(j)$ are defined on $n$ equally-spaced bins in $\mathbb{R}$ (for our simulations, we will take $n=32$), and $z_k$ denotes the $k^{th}$ bin. 
+%
+We note that although histograms are not theoretically essential for using EMD, they make the computation feasible.
 
@@
 %
-\caption{Diffusion maps embedding for three different chemotaxis simulation data sets.
+\caption{Diffusion maps embeddings for three different chemotaxis simulation data sets.
 (a) $\lambda = 100$, $s = 10$.
@@
 (c) $\lambda = 6400$, $s = 80$.
-Each data set consists of $10$ simulations, with initial conditions uniformly chosen such that $0.1 \le p  \le 0.9$.
-We allow each simulation to evolve for $10$ time units, with $dt=1$, and use data with $t > 0$ for analysis.
+We set $t_{max} = 10$ and $dt=1$.
 The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions.
@@
 From the spectra, we can see that the first two components are informative for (a), components 1 and 3 are informative for (b), and components 1 and 4 are informative for (c).
-The corresponding embeddings are shown below.
-For comparison, the embeddings using the first two components are also shown in (b) and (c).}
+The corresponding reduced diffusion maps embeddings are shown in the middle row.
+For comparison, the standard diffusion maps embeddings using the first two components are also shown in the bottom row in (b) and (c).}
 %
@@
 %
-Plotting the data using the two eigenvectors identified as corresponding to unique eigendirections yields two-dimensional embeddings which capture the macroscopic variables $p$ and $t$ (we only show the correspondence of the two identified unique eigendirections with $p$ in the middle row of Figure~\ref{fig:chemotaxis_simulations_harmonics}; the correspondence between these eigendirections and $t$ is similarly consistent).
-%
-In contrast, plotting the data using the two leading eigenvectors $\phi_1$ and $\phi_2$ produces uninformative embeddings for large values of $\lambda$, as $\phi_2$ is a repeated eigendirection which does not parameterize any new directions in the data.
-%
-From the top row of Figure~\ref{fig:chemotaxis_simulations_harmonics}, we see that the eigenvalue spectra do not exhibit any large spectral gaps.
-%
-However, the gap between the eigenvalues which correspond to new directions in the data (colored in blue) increases with $\lambda$, indicating that the underlying manifold becomes more narrow/one-dimensional, and the corresponding eigenvectors provide us with an informative two-dimensional parameterization of the data.
+The two-dimensional reduced diffusion maps embeddings capture the macroscopic variables $p$ and $t$ (we only show the correspondence of the two identified unique eigendirections with $p$ in the middle row of Figure~\ref{fig:chemotaxis_simulations_harmonics}; the correspondence between these eigendirections and $t$ is similarly consistent).
+%
+In contrast, the standard diffusion maps embeddings using the two leading eigenvectors $\phi_1$ and $\phi_2$ produce uninformative embeddings for large values of $\lambda$, as $\phi_2$ is a repeated eigendirection which does not parameterize any new directions in the data.
+%
+Furthermore, although the eigenvalue spectra do not exhibit any large spectral gaps, the gap between the eigenvalues which correspond to new directions in the data (colored in blue) increases with $\lambda$, indicating that the underlying manifold becomes more narrow/one-dimensional, and the corresponding eigenvectors provide us with an informative two-dimensional parameterization of the data.
 %which is consistent with the expected macroscopic dynamics.
@@
 %
-This implies that for fixed $\lambda$ and $s$, the {\em microscopic} data in a high-dimensional ambient space (e.g., the positions of all $N$ cells) should lie on a two-dimensional manifold parameterized by $p$ and $t$,
+This implies that for fixed $\lambda$ and $s$, the {\em microscopic} data in a high-dimensional ambient space (e.g., the positions of all $N$ cells) should lie on a two-dimensional manifold parameterized by $p$ (which controls the initial ratio of $\rho^+$ and $\rho^-$) and $t$ (which describes the evolution),
 and uncovering the low-dimensional structure of the microscopic data reveals this manifold.
@@
 
-The relative importance of $p$ and $t$ from the analytic description are consistent with the results in Figure~\ref{fig:dmaps_embed_emd}, where $p$ appears as the ``more important'' coordinate when $\lambda$ is small and as the ``less important'' coordinate when $\lambda$ is large.
+The relative importance of $p$ and $t$ from the analytic description are consistent with the results in Figure~\ref{fig:dmaps_embed_emd}, where $p$ is correlated with the dominant coordinate when $\lambda$ is small and correlated with the subdominant coordinate when $\lambda$ is large.
 %
@@
 
-Furthermore, this analytic macroscopic descrption shows us that when $\lambda$ is small, the system dynamics can be described by the densities $\rho^+$ and $\rho^-$, and when $\lambda$ is large, the dynamics are described by $\rho = \rho^+ + \rho^-$.
+Furthermore, this analytic macroscopic descrption shows us that when $\lambda$ is small, the system dynamics can be described by the two densities $\rho^+$ and $\rho^-$, and when $\lambda$ is large, the dynamics are described by a single density $\rho = \rho^+ + \rho^-$.
 %
@@
 %
-However, for large $\lambda$, the two densities rapidly equilibrate, and no organization based on the differences between the two densities is visible.
-
-\subsubsection{Using an appropriate metric}
-
-To emphasize the importance of using an appropriate distance metric, we analyzed the two sets of simulation data using the standard Euclidean distance between the histograms to compute the distances in \eqref{eq:W}.
+However, for large $\lambda$, the two densities rapidly equilibrate, and negligible organization based on the differences between the two densities is visible.
+
+%\subsubsection{Using an appropriate metric}
+
+This analytic macroscopic description also allows us to emphasize the importance of using an appropriate distance metric.
+%
+We analyzed the two sets of simulation data from Figure~\ref{fig:dmaps_embed_emd} using the standard Euclidean distance between the histograms to compute the distances in \eqref{eq:W}.
 %
@@
 %
-In contrast, the correlations between the embedding coordinates and the macroscopic variables when using EMD in the diffusion maps calculation are all greater than 80\%.
-%
-This emphasizes the importance of using a metric which accurately describes the distances between observations.
+In contrast, the correlations between the embedding coordinates and the macroscopic variables when using EMD in the diffusion maps calculation are all greater than 80\%, and so using a metric which accurately describes the distances between observations is essential for obtaining an informative parameterization of data.
 
@@
 \caption{Detecting the change in dimensionality.
-The ratio of the log of the two eigenvalues which correspond to unique eigendirections is plotted as a function of the timescale of the simulation ($t_{max}$) and $\lambda$.
-Each data point is the average eigenvalue ratio over three data sets.
-Each data set consists of $10$ simulations, with initial conditions uniformly chosen such that $0.1 \le p  \le 0.9$.
-We allow each simulation to evolve for $t_{max}$ time units, with $dt=t_{max}/10$, and use data with $t > 0$ for analysis.
+The ratio of the log of the two eigenvalues which correspond to unique eigendirections is plotted as a function of $t_{max}$ and $\lambda$.
+Each data point is the average eigenvalue ratio over three data sets, and in each simulation, $dt=t_{max}/10$.
 The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions.
 The curve $1/\lambda = t_{max}/N$ is shown in white.
-When the diffusive time scale is significantly larger than the drift time scale, the data is expected to be effectively one-dimensional.
-This transition is consistent with the empirical eigenvalues: the data are effectively one-dimensional when $t_{max}$ and/or $\lambda$ are large, since the first eigenvalue is significantly larger than the second.
 }
@@
 %
-For this example, we show that the true dimensionally of a data set can be estimated by looking at the eigenvalues corresponding to unique eigendirections. 
-%
-Let $\mu_{i_1} \ge \mu_{i_2}$ denote these two eigenvalues. 
+For this example, the true dimensionally of a data set can be estimated by looking at the two eigenvalues $\mu_{i_1} \ge \mu_{i_2}$ corresponding to unique eigendirections. 
 %
@@
 \end{equation}
-when this ratio becomes small, the data are effectively one-dimensional, as the second dimension is very small compared with the first.
+when this ratio becomes small, the data are effectively one-dimensional, as the second direction is very small compared with the first.
 
@@
 %
-or small $\lambda$ and/or short $t_{max}$, the data are effectively two-dimensional, as both $p$ and $t$ are important to the observed dynamics. 
+For small $\lambda$ and/or short $t_{max}$, the data are effectively two-dimensional, as both $p$ and $t$ are important to the observed dynamics. 
 %
@@
 %
-This ratio is also plotted in Figure~\ref{fig:chemotaxis_compare_timescales_evals} as a function of $\lambda$, and the transition is consistent with the transition predicted by the diffusion maps eigenvalues.
+This curve is also plotted in Figure~\ref{fig:chemotaxis_compare_timescales_evals} as a function of $\lambda$ and is consistent with the transition predicted by the diffusion maps eigenvalues.
 
@@
 
-This paper aims to bridge the fields of data mining and dynamical systems. 
-%
-For data mining methods to be informative with regards to the underlying system dynamics, processing the appropriate statistical moments or observables instead of individual particles is essential. 
-%
-In addition, one must also define the appropriate distance metric between the observations.
-%
-These two components induce the ``correct" Riemannian geometry and allow for informative analyses, such as meaningful and compact parameterization, phase shift detection, and dynamical characterization, through manifold learning.
-%
-In this paper, these concepts were demonstrated using a dynamical model of cellular chemotaxis with a known macroscopic behavior.
-%
-We showed that, using histograms as observers and EMD as a distance metric, diffusion maps can uncover a parameterization of the microscale data which is consistent with the analytic macroscopic model.
-%
-We are confident that such techniques can help inform modeling efforts for systems where the macroscopic dynamics and behavior are not currently known.
+This paper addresses the issue of repeated eigendirections, one of the key shortcomings in applying diffusion maps analysis to real complex data. 
+%
+We show that, using local linear regression, we can automatically detect which eigenvectors correspond to unique eigendirections in the data. 
+%
+From this detection, we can then obtain a reduced diffusion maps embedding which is both parsimonious and endures a metric which is equivalent to the standard diffusion distance. 
+%
+We then showed that these algorithms can allow us to analyze data from a complex dynamical system, enabling the extraction of good reduced coordinates and also the detection of changes in the dimensionality of the macroscopic dynamics.
+%
+In the future, we are confident that our proposed methodology will aid in the analysis of more complex data sets for which the dimensionality of the underlying manifold is unknown.
 

### 2015-05-14 chemotaxis/paper/draft1.tex
chemotaxis : yannis edits

@@
 \author[princeton]{Carmeline~J.~Dsilva}
-\ead{cdsilva@princeton.edu}
+%\ead{cdsilva@princeton.edu}
 %
 \author[technion]{Ronen~Talmon}
-\ead{ ronen@ee.technion.ac.il}
+%\ead{ ronen@ee.technion.ac.il}
 %
 \author[yale]{Ronald R. Coifman}
-\ead{coifman@math.yale.edu}
+%\ead{coifman@math.yale.edu}
 %
@@
 \begin{abstract}
-Nonlinear manifold learning algorithms have been effective for the analysis of complex data sets. 
-%
-However, these algorithms still encounter several challenges when faced with real data.
-%
-A key shortcoming is the existence of repeated eigendirections, where several embedding coordinates parameterize the same direction in the data set, therefore obscuring the true dimensionality of the underlying manifold.
-%
-In this paper, we propose an algorithm, based on local linear regression, to automatically detect which coordinates are repeated eigendirections. 
-%
-We propose using only the eigenvectors which correspond to unique eigendirections to embed the data, and show that this reduced diffusion maps embedding induces a metric which is equivalent to the standard diffusion distance.
+%
+Nonlinear manifold learning algorithms, such as diffusion maps, have been fruitfully
+applied in recent years to the analysis of large and complex data sets.
+%
+However, such algorithms still encounter challenges when faced with real data.
+%
+One such 
+challenge is the existence of ``repeated eigendirections," which obscures the detection of the 
+true dimensionality of the underlying manifold and arises when several embedding coordinates parametrize
+the same direction in the intrinsic geometry of the data set. 
+%
+We propose an algorithm, based on local linear regression, 
+to automatically detect coordinates corresponding to repeated eigendirections.
+%
+We construct a more parsimonious embedding using only the eigenvectors corresponding to unique eigendirections, and we show that this reduced diffusion maps embedding induces 
+a metric which is equivalent to the standard diffusion distance.
 %
@@
 %
-We then apply our algorithm to data collected from a stochastic simulation of cellular chemotaxis. 
-%
-Here, we show that our approach for factoring out repeated eigendirections can allow us to detect changes in dynamical behavior and the underlying system dimensionality directly from data. 
+We then apply our algorithm to data collected from a stochastic simulation of cellular chemotaxis, where our approach for factoring out repeated eigendirections allows us to detect changes in dynamical behavior and the underlying
+intrinsic system dimensionality directly from data.
 %
@@
 %
-In particular, for dynamical systems, data-driven methodologies have been essential when a simple macroscopic model cannot be written analytically due to the complexity of the system \cite{talmon2014intrinsic,berry2013time,singer2009detecting,ferguson2010systematic}.
-%
-For simulations, the governing model is often very high-dimensional and the macroscale dynamics are not obvious; for experimental systems, an explicit model is often unknown. 
-%
-For such cases, data obtained from observations and/or simulations of the dynamical system combined with such methodologies gives rise to a low-dimensional description which can not only provide insight into the underlying dynamics, but also serve as a first step in constructing macroscale models which are consistent with the observed microscale behavior.
-
-Due to the complexity and range of microscale behaviors possible in dynamical systems, we use manifold learning, a class of nonlinear data-driven techniques, to analyze data.
-%
-Most manifold learning algorithms construct parameterizations of the data through the spectral analysis of a Laplace operator \cite{Belkin2003,coifman2005geometric,coifman2006geometric,singer2008non}.
-%
-The data is then embedded in a new, low-dimensional coordinate system given by the eigenvectors of this operator. 
-%
-Our hope is that these coordinates, obtained in a data-driven manner, will describe the variables which govern the macroscale dynamical behavior of the system and enable us to introduce a reduced macroscale {\em ab initio} model.
-
-These manifold learning algorithms were first applied to synthetic data sets to illustrate their geometric properties and flexibility \cite{coifman2005geometric, nadler2006diffusion}. 
-%
-Only recently have they been applied to experimental and simulation data, which was possible due to advances in data representation (observers) and metrics \cite{rubner2000earth,mallat2012group,talmon2013empirical,zhao2014rotationally,rohrdanz2011determination.}.
-%
-A critical shortcoming of these methods is that from a geometric perspective, not all eigenvectors are guaranteed to parameterize unique directions within the data; some eigenvectors are ``repeated eigendirections'' which describe the same mode.
-%
-Identifying those eigenvectors is critical for obtaining a parsimonious parameterization of the system which captures the true dimensionality of the macroscale dynamics.  
-%
-This is a challenging task which is often done manually, and few methods have been proposed to automate the identification of the unique eigendirections \cite{gerber2007robust}.
+For multiscale dynamical systems in particular, data-driven methodologies are essential for
+assisting in model reduction when simple macroscopic models cannot be analytically obtained
+due to the system complexity \cite{talmon2014intrinsic,berry2013time,singer2009detecting,ferguson2010systematic}.
+%
+For simulations or experimental observations of such systems, the detailed, microscale evolving system state is
+very high-dimensional, and the reduction to useful macroscale dynamics is not obvious.
+%
+In such cases, data obtained from observations and/or simulations of the dynamical system combined with
+data mining methodologies can lead to a low-dimensional description which not only provides insight into the underlying dynamics, but also serves as a first step in constructing macroscale models consistent with the observed microscale behavior.
+
+%Due to the complexity and range of microscale behaviors possible in dynamical systems,
+%we use manifold learning, a class of nonlinear data-driven techniques, to analyze data.
+%
+Several manifold learning algorithms obtain parametrizations of the data through
+the spectral analysis of a Laplace operator \cite{Belkin2003,coifman2005geometric,coifman2006geometric,singer2008non}.
+%
+The data is then embedded in a new low-dimensional coordinate system
+given by the eigenvectors of this operator.
+%
+The premise is that these coordinates, obtained in a data-driven manner, are the right macroscopic ``observables",
+the variables which parametrize the macroscale dynamical behavior of the system, thus enabling the construction of a reduced macroscale model.
+
+Such manifold learning algorithms were initially applied to synthetic data sets, to illustrate
+their geometric properties and flexibility \cite{coifman2005geometric, nadler2006diffusion}.
+%
+More recently they have been applied to experimental and simulation data,
+enabled by advances in data representation (observers) and
+metrics \cite{rubner2000earth,mallat2012group,talmon2013empirical,zhao2014rotationally, rohrdanz2011determination}.
+%
+A nontrivial shortcoming of these methods is that, from a geometric perspective,
+not all eigenvectors are guaranteed to parametrize unique directions within the data;
+some eigenvectors are ``repeated eigendirections'' which describe the same coordinate in the intrinsic geometry
+of the data.
+%
+Identifying these eigenvectors is critical for obtaining a
+parametrization of the system which captures the true dimensionality of the macroscale dynamics.
+%
+This task is often done manually, and few methods have been proposed to automate the identification of the unique eigendirections \cite{gerber2007robust}.
 
@@
 %
-We first demonstrate our algorithm on synthetic examples where a closed-form solution for the eigenvectors and eigenvalues of the Laplace operator is known, which will allow us to validate our proposed approach.
-%
-We then consider a complex data set from a stochastic dynamical system simulation which models cellular chemotaxis \cite{othmer1988models}. 
-%
-The recent advances in observers and metrics, coupled with the proposed approach for identifying the unique eigendirections, provide a data analysis pipeline which successfully analyzes data from this simulation.
-%
-This pipeline allows us to detect changes in the regime/mode of the system, as well as changes in the underlying dimensionality of the macroscale dynamics.
+We first demonstrate and validate our algorithm on synthetic examples, where a closed-form solution for the
+eigenvectors and eigenvalues of the Laplace operator is known.
+%
+We then consider a simulation data set from a stochastic dynamical system modeling
+cellular chemotaxis \cite{othmer1988models}.
+%
+Recent advances in observers and metrics, coupled with the proposed approach for identifying the unique eigendirections,
+provide a data analysis pipeline which successfully analyzes this simulation in a purely data-driven manner.
+%
+We will show that this pipeline allows us to detect changes in the dimensionality of the underlying macroscale dynamics, which are
+related to changes in the regime/mode of the system.
 
 \section{Manifold Learning Based on Laplace Operators}
-
-Although data from dynamical systems is often high-dimensional, since it is collected at the microscale, the macroscale dynamics often evolve to a low-dimensional description. 
-%
-In such cases, the high-dimensional microscale data is restricted to a low-dimensional manifold embedded in the high-dimensional space, and extracting a parameterization of this manifold corresponds to parameterizing the low-dimensional, macroscale dynamics.
-%
-Let $z(1), \dots, z(m) \in \mathbb{R}^n$ denote $m$ observations sampled from the dynamical system. 
+%Although data from dynamical systems is often high-dimensional, since it is collected at the microscale, the macroscale dynamics often evolve to a low-dimensional description.
+%
+%In such cases, the high-dimensional microscale data is restricted to a low-dimensional manifold embedded in the high-dimensional space, and extracting a parametrization of this manifold corresponds to parametrizing the low-dimensional, macroscale dynamics.
+%
+
+Let $z(1), \dots, z(m) \in \mathbb{R}^n$ denote $m$ state observations sampled from an evolving autonomous dynamical system.
 %
@@
 %
-We will show how to extract a $d$-dimensional parameterization of the observations which is consistent with the geometry of the manifold.
+We therefore consider the problem of parametrizing the continuous $d$-dimensional manifold $\mathcal{M}_d$ embedded in $\mathbb{R}^n$ from data.
+%
+For linear hyperplanes, the principal axes parametrize the manifold.
+%
+However, in the case when the manifold is nonlinear, a set of appropriate coordinates is not readily apparent.
+%
+By using a particular manifold learning technique based on the construction of a Laplace operator, namely, diffusion maps, we will show how to extract a $d$-dimensional parametrization of the observations which is consistent with the geometry of the manifold \cite{Belkin2003, coifman2005geometric, singer2008non}.
 
@@
 (c) Two-dimensional strip with data sampled from a Gaussian distribution in $z_1$ and sampled uniformly in $z_2$, colored by the first and third (non-trivial) eigenvectors of the discrete Laplacian.
-Note that in all cases we uncover parameterizations which are one-to-one with $z_1$ and $z_2$.}
+Note that in all cases we uncover parametrizations which are one-to-one with $z_1$ and $z_2$.}
 \end{figure}
 
-We start by considering a continuous setting where rigorous analysis is available for specific examples. 
-%
-Using such an example, we will show that the existence of repeated eigendirections is inherent to the manifold learning setup based on Laplace operators, and that the identification of the unique eigendirections is nontrivial.
-%
-The existence of these challenges in the continuous setting implies that, even in the limit of infinite data, such repeated eigendirections still pose a problem for analysis.
-
-We consider the problem of parameterizing a continuous $d$-dimensional manifold $\mathcal{M}_d$ embedded in $\mathbb{R}^n$.
-%
-For linear hyperplanes, the principal axes parameterize the manifold.
-%
-However, in the case when the manifold is nonlinear, the set of coordinates is not readily apparent. 
-%
-In recent years, the eigenfunctions of the Laplace-Beltrami operator have been used to parameterize the manifold \cite{Belkin2003, coifman2005geometric, singer2008non}. 
-%
+In recent years it has been noted that the eigenfunctions of the \emph{continuous}
+Laplace-Beltrami operator provide ``good" coordinates for a manifold \cite{jones2008}.
+%
+We start by considering such a continuous setting, where rigorous analysis is possible for specific examples.
+%
+Using such an example, we will demonstrate that the existence of repeated eigendirections is
+inherent to the manifold learning setup based on Laplace operators,
+and that the identification of the unique eigendirections is nontrivial.
+%
+The fact that these challenges arise in the continuous setting implies that, even in the limit of infinite data, such repeated eigendirections still pose a problem for data analysis.
+
 To illustrate why these eigenfunctions provide appropriate coordinates, consider a two-dimensional strip with edge lengths $L_1$ and $L_2$.
@@
 %
-We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( {\pi z_1}/{L_1} \right)$ and $\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$ are one-to-one with the $z_1$ and $z_2$ coordinates, respectively, and therefore yield a parameterization of the underlying manifold (see Figure~\ref{subfig:strip_efuncs}).
-%
-Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide a measure of $L_1$ versus $L_2$: as the ratio between $L_1$ and $L_2$ increases, the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases (this will be discussed further in Section~\ref{sec:relative_lengths}).
+We note that the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( {\pi z_1}/{L_1} \right)$ and
+$\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$ are one-to-one with the $z_1$ and $z_2$ coordinates,
+respectively, and therefore provide a parametrization of the underlying manifold (see Figure~\ref{subfig:strip_efuncs}).
+%
+Furthermore, the corresponding eigenvalues $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ provide an estimate of the
+relative magnitude of $L_1$ versus $L_2$: as the ratio between $L_1$ and $L_2$ increases,
+the gap between $\tilde{\mu}_{1,0}$ and $\tilde{\mu}_{0,1}$ also increases
+(this will be discussed further in Section~\ref{sec:relative_lengths}).
 %
@@
 %
-First, $z_1$ and $z_2$ are not necessarily decoupled in subsequent eigenfunctions. 
+First, $z_1$ and $z_2$ are not necessarily decoupled in subsequent eigenfunctions,
+and a proper parameterization of the manifold is not necessarily given by the $d$ eigenfunctions associated with the
+smallest $d$ eigenvalues.
 % TODO: address this in later section
 %
-Second, eigenfunctions with $k_1+k_2 \ge 2$ do not describe any additional directions along the strip; we will refer to these as ``repeated eigendirections,'' and we will refer to the eigenfunctions with $k_1+k_2 =1$ as ``unique eigendirections.''
-%
-We note that, although the eigenfunctions can only be written analytically for very special cases, it has been observed empirically that the eigenfunctions often provide appropriate coordinates to parameterize more complex, nonlinear manifolds.
+Second, eigenfunctions with $k_1+k_2 \ge 2$ do not describe any
+additional directions intrinsic to the strip geometry; we will refer to these as ``repeated eigendirections,''
+and we will refer to the eigenfunctions with $k_1+k_2 =1$ as ``unique eigendirections.''
+%
+Clearly, this problem of repeated eigendirections also arises for more complex manifolds of more than two variables.
+%
+We note that, although the eigenfunctions can only be written analytically for very special cases,
+it has been observed empirically that they often provide appropriate coordinates
+to usefully parametrize more complex, nonlinear manifolds.
+%
 
@@
 
-In most applications, we are not given a description of the continuous manifold. 
-%
-Instead, we are given data {\em sampled} from the underlying manifold, and we would like to uncover a parameterization of the data.
-%
-We will do this by constructing a matrix which approximates the Laplace-Beltrami operator. 
-%
-It was shown in \cite{coifman2006geometric} that, in the limit of infinite data, the discrete Laplacian matrix constructed from data converges pointwise to the continuous Laplace-Beltrami operator on the manifold.
+In most applications, we are not given a description of the continuous manifold on which the data lie.
+%
+Instead, we are given data {\em sampled} from the underlying manifold,
+and the parameterization of the manifold needs to be uncovered from the data.
+%
+This can be accomplished by constructing a matrix which approximates the Laplace-Beltrami operator.
+%
+It was shown in \cite{coifman2006geometric} that, in the limit of infinite data,
+this discrete Laplacian matrix constructed from data converges pointwise to the continuous Laplace-Beltrami operator on the manifold.
 %
@@
 
-Given observations $z(1), \dots, z(m) \in \mathcal{M}_d$, we first construct the weight matrix $W \in \mathbb{R}^{m \times m}$, with
+Given observations $z(1), \dots, z(m) \in \mathcal{M}_d$, we first construct the weight matrix $\mathbf{W} \in \mathbb{R}^{m \times m}$, with
 \begin{equation} \label{eq:W}
-W_{ij} = \exp \left( -\frac{\|z(i) - z(j) \|^2}{\epsilon^2} \right), \ i,j=1,\ldots,m,
+\mathbf{W}_{ij} = \exp \left( -\frac{\|z(i) - z(j) \|^2}{\epsilon^2} \right), \ i,j=1,\ldots,m,
 \end{equation}
@@
 %
-The kernel's scale $\epsilon$ can be chosen using several heuristics \cite{coifman2008graph, rohrdanz2011determination}; we often take $\epsilon$ to be the median of the pairwise distances between the data points.
-%
-We then construct the diagonal matrix $D \in \mathbb{R}^{m \times m}$, with $D_{ii} = \sum_j W_{ij}$, and the matrix $A  = D^{-1} W.$
-%
-In the limit of infinite data uniformly sampled from $\mathcal{M}_d$ ($m \rightarrow \infty$ and $\epsilon \rightarrow 0$ with the appropriate rate), the discrete Laplacian matrix $I-A$ approaches the Laplace-Beltrami operator on the manifold $\mathcal{M}_d$ \cite{coifman2006geometric}.
-%
-Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $A$ approximate the eigenfunctions of the Laplace-Beltrami operator on $\mathcal{M}_d$,
-and the eigenvalues $\mu_0, \mu_1, \dots, \mu_{m-1}$ of $A$ are related to the eigenvalues of the continuous operator by 
+The kernel's scale $\epsilon$ can be chosen using several heuristics \cite{rohrdanz2011determination, coifman2008graph};
+we often take $\epsilon$ to be the median of the pairwise distances between the data points.
+%
+We then construct the diagonal matrix $\mathbf{D} \in \mathbb{R}^{m \times m}$, with $\mathbf{D}_{ii} = \sum_j \mathbf{W}_{ij}$,
+and form the matrix $\widetilde{\mathbf{W}} = \mathbf{D}^{-\alpha} \mathbf{W} \mathbf{D}^{-\alpha}$, where $0 < \alpha < 1$.
+%
+Next, we construct the diagonal matrix $\widetilde{\mathbf{D}} \in \mathbb{R}^{m \times m}$, with $\widetilde{\mathbf{D}}_{ii} = \sum_j \widetilde{\mathbf{W}}_{ij}$, and the matrix $\mathbf{A}  = \widetilde{\mathbf{D}}^{-1} \widetilde{\mathbf{W}}.$
+
+If the data $z(1), \dots, z(m)$ are sampled from $\mathcal{M}_d$ with some density $q$, then, for $\epsilon \rightarrow 0$ and $m \rightarrow \infty$ (with the appropriate rates), the discrete matrix converges to the following continuous limit operators with Neumann boundary conditions (as discussed in the previous section) \cite{coifman2006geometric}
+\begin{align}
+\frac{1}{\epsilon^2}(\mathbf{I}-\mathbf{A}) \phi &\rightarrow \nabla^2 \phi - 2\nabla U \cdot \nabla \phi, &&\alpha = 0 \\
+\frac{1}{\epsilon^2}(\mathbf{I}-\mathbf{A}) \phi &\rightarrow \nabla^2 \phi - \nabla U \cdot \nabla \phi, &&\alpha = 1/2 \\
+\frac{1}{\epsilon^2}(\mathbf{I}-\mathbf{A}) \phi &\rightarrow \nabla^2 \phi, &&\alpha = 1
+\end{align}
+where $U = - \log q$.
+%
+The different limit operators, depending on the choice of $\alpha$, imply that nonuniform
+sampling on the manifold may have different effects.
+%
+In particular, by setting $\alpha=1$, one can factor out the density effects in the
+weight matrix, and the discrete Laplacian matrix approaches the Laplace-Beltrami operator on the manifold $\mathcal{M}_d$.
+%
+Therefore, the eigenvectors $\phi_0, \phi_1, \dots, \phi_{m-1}$ of $\mathbf{A}$ approximate the eigenfunctions of the Laplace-Beltrami operator on $\mathcal{M}_d$,
+and the eigenvalues $\mu_0, \mu_1, \dots, \mu_{m-1}$ of $\mathbf{A}$ are related to the eigenvalues of the continuous operator by
 \begin{equation} \label{eq:evals_relationship}
@@
 
-However, the data need not be sampled uniformly from the underlying manifold.
-%
-If the data $z(1), \dots, z(m)$ are sampled from $\mathcal{M}_d$ with some density $q$, then the matrix $I-A$ approximates the Fokker-Planck operator, 
-$%\begin{equation}
-(I-A) \phi \rightarrow \nabla^2 \phi - \nabla U \cdot \nabla \phi, 
-$ %\end{equation}
-where $U = -2 \log q$. 
-%
-Alternatively, one can factor out the density effects in the weight matrix.
-%
-Given data $z(1), \dots, z(m) \in \mathcal{M}_d$, we construct the matrix
+As discussed previously, the eigenfunctions provide a parametrization of the manifold,
+such that $\mu_j^\tau \phi_{j}(i)$ yields the $j^{th}$ embedding coordinate for $z(i)$, where $\tau \ge 0$ is
+a parameter (in our examples, we will take $\tau=0$).
+%
+This is the standard diffusion maps embedding \cite{coifman2005geometric, coifman2006geometric}, where
+we order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$.
+%
+Because the matrix $\mathbf{A}$ is row-stochastic ($\sum_j \mathbf{A}_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a trivial constant vector.
+%
+The distance induced by this mapping is called the standard diffusion distance,
 %
 \begin{equation}
-\tilde{W} = D^{-1} W D^{-1},
-\end{equation}
-%
-where pre- and post-multiplication by $D^{-1}$ removes any density effects from $W$. 
-%
-We then construct the diagonal matrix $\tilde{D} \in \mathbb{R}^{m \times m}$, with $\tilde{D}_{ii} = \sum_j \tilde{W}_{ij}$, and the matrix $\tilde{A}  = \tilde{D}^{-1} \tilde{W}.$
-%
-The matrix $\tilde{A}$ is comparable to the matrix $A$, but with sampling density effects removed, such that $I-\tilde{A}$ approaches the Laplace-Beltrami operator on $\mathcal{M}_d$ even if the data is nonuniformly sampled from the underlying manifold \cite{coifman2005geometric}.
-
-As discussed previously, the eigenfunctions provide a parameterization of the manifold, such that $\phi_{j}(i)$ yields the $j^{th}$ embedding coordinate for $z(i)$; this is the standard diffusion maps embedding \cite{coifman2005geometric, coifman2006geometric}.
-%
-We order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$.
-%
-Because the matrix $A$ is row-stochastic ($\sum_j A_{ij} = 1$),  $\mu_0 = 1$ and $\phi_0$ is a trivial constant vector.
-%
-Figures~\ref{subfig:strip_evecs_uniform}--\ref{subfig:strip_evecs_nonuniform} shows data sampled from a strip, colored by eigenvectors of $A$. 
-%
-In cases of both uniform and nonuniform sampling, the selected eigenvectors are one-to-one with $z_1$ and $z_2$, and thus parameterize the manifold. 
+D^2_\tau(z(i), z(j)) = \sum_{k=1}^{m-1} \mu_k^{2 \tau} \left( \phi_k(i) - \phi_k(j)  \right)^2.
+\end{equation}
+%
+Figures~\ref{subfig:strip_evecs_uniform}--\ref{subfig:strip_evecs_nonuniform} shows
+data sampled from a strip, colored by eigenvectors of $\mathbf{A}$.
+%
+In cases of both uniform and nonuniform sampling, the selected eigenvectors are one-to-one with $z_1$ and $z_2$,
+and thus parametrize the manifold.
+%
+%TODO: maybe we can put here the new figures for the different values of $\alpha$ instead of these figures.
 %
@@
 
-From the previous section, we know that the eigenfunctions with $(k_1, k_2) =(1, 0)$ and $(k_1, k_2) =(0, 1)$ provide embedding coordinates for the manifold; these two eigenfunctions are both uncoupled and not repeated.
-%
-From \eqref{eq:evals}, we see that sorting the eigenvectors by the magnitude of the corresponding eigenvalues implies that these two eigenvectors are guaranteed to appear before any coupled or repeated eigendirections.
-%
-However, these eigenvectors are {\em not} guaranteed to appear as the first two (non-trivial) eigenvectors, as harmonics of the first eigendirection (i.e., $\cos \left( n \pi z_1 / L_1 \right)$ with $n > 1$) could appear before the second.
-%
-We note that, in contrast to our simple illustrative example, for most data sets of interest, the coordinates for the underlying manifold are unknown and cannot easily be obtained from the coordinates of the original data, and identifying which eigenvectors correspond to unique eigendirections is nontrivial.
+From the previous section, we know that the eigenfunctions with $(k_1, k_2) =(1, 0)$ and $(k_1, k_2) =(0, 1)$
+provide embedding coordinates for the manifold; these two eigenfunctions are both uncoupled and not repeated.
+%
+From \eqref{eq:evals}, we see that sorting the eigenvectors by the magnitude of
+the corresponding eigenvalues implies that these two eigenvectors are guaranteed to
+appear before any coupled or repeated eigendirections.
+%
+However, these eigenvectors are {\em not} guaranteed to appear as the first two (non-trivial) eigenvectors,
+as harmonics of the first eigendirection (i.e., $\cos \left( n \pi z_1 / L_1 \right)$ with $n > 1$) could appear before the second.
+%
+We note that, in contrast to our simple illustrative example, for most data sets of interest, the coordinates for the underlying manifold are unknown and cannot easily be obtained from the coordinates of the original data, so that identifying which eigenvectors correspond to unique eigendirections is nontrivial.
 
@@
 
-Common practice is to order the eigenvectors by the magnitude of the corresponding eigenvalues, assuming that the leading eigenvectors provide a parameterization of the underlying manifold.
-%
-However, as discussed in the previous section, some eigenvectors are higher harmonics of previous eigenvectors and do not describe new directions in the data set \cite{gerber2007robust}.
-%
-For the case where $\mathcal{M}_d$ is a $2$-dimensional strip with edge lengths $L_1  > L_2$, recall that the eigenfunctions $\tilde{\phi}_{1,0} = \cos \left(  {\pi z_1}/{L_1} \right)$ and  $\tilde{\phi}_{0,1} = \cos \left(  {\pi z_2}/{L_2} \right)$ provide embedding coordinates for the manifold $\mathcal{M}_d$.
-%
-However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to correspond to the two smallest (non-trivial) eigenvalues (the smallest eigenvalue will always be $\tilde{\mu}_{0,0} = 0$ and correspond to a constant eigenfunction; these eigenvalues are related to the eigenvalues $\mu_k$ of the discrete operator via \eqref{eq:evals_relationship}).
-%
-In fact, if $L_1 > 2 L_2$, then $\tilde{\mu}_{2, 0} < \tilde{\mu}_{0, 1}$, and so the second (non-trivial) eigenvector (when the eigenvectors are ordered by their corresponding eigenvalues) will be a repeated eigendirection of the first and still parameterize $z_1$ (see Figure~\ref{fig:strip_harmonics}).
-%
-We therefore require a methodology to automatically detect which eigenvectors are harmonics of each other. 
-%
-Utilizing only the eigenvectors $\phi_i$ which correspond to unique eigendirections yields the most parsimonious representation of the data.
-%
-We will show how we can automatically detect these eigenvectors to obtain a meaningful representation of the data.
-%
-Furthermore, the corresponding eigenvalues provide us with a measure of the relative lengths of the data set along these unique eigendirections.
+Common practice is to order the eigenvectors by the magnitude of the corresponding eigenvalues,
+assuming that the leading eigenvectors provide a parametrization of the underlying manifold.
+%
+However, as discussed in the previous section, some eigenvectors are higher harmonics of
+previous eigenvectors and do not describe new directions in the data set \cite{gerber2007robust}.
+%
+For the case where $\mathcal{M}_d$ is a $2$-dimensional strip with edge lengths $L_1  > L_2$, recall
+that the eigenfunctions $\tilde{\phi}_{1,0} = \cos \left(  {\pi z_1}/{L_1} \right)$ and
+$\tilde{\phi}_{0,1} = \cos \left(  {\pi z_2}/{L_2} \right)$ provide embedding coordinates for the manifold $\mathcal{M}_d$.
+%
+However, these two eigenvectors $\tilde{\phi}_{1, 0}$ and $\tilde{\phi}_{0, 1}$ are not guaranteed to
+correspond to the two smallest (non-trivial) eigenvalues (the smallest eigenvalue will always be $\tilde{\mu}_{0,0} = 0$ and
+correspond to a constant eigenfunction; these eigenvalues are related to the eigenvalues $\mu_k$ of the discrete
+operator via \eqref{eq:evals_relationship}).
+%
+In fact, if $L_1 > 2 L_2$, then $\tilde{\mu}_{2, 0} < \tilde{\mu}_{0, 1}$, and so the
+second (non-trivial) eigenvector (when the eigenvectors are ordered by their corresponding eigenvalues)
+will be a ``repeated eigendirection" of the first and still parametrize $z_1$ (see Figure~\ref{fig:strip_harmonics}).
+%
+Automatic detection of which eigenvectors are harmonics of each other would be clearly useful.
+%
+Utilizing only the eigenvectors $\phi_i$ which correspond to unique eigendirections yields
+the most parsimonious representation of the data.
+%
+We will show how we can automatically detect these eigenvectors to obtain a meaningful representation of the data,
+and that using only these eigenvectors yields a reduced embedding which is equivalent to the standard diffusion maps embedding.
+%
+Furthermore, the corresponding eigenvalues provide us with an estimate of the relative lengths of the data set along these unique eigendirections.
 
@@
 
-Given the eigenvectors $\phi_1, \phi_2, \dots, \phi_{m-1} \in \mathbb{R}^m$, we would like to automatically deduce which ones capture new directions in the data, and which ones are merely repeated eigendirections.
-%
-This problem was addressed previously in \cite{gerber2007robust} by performing successive iterations of diffusion maps, interspersed with advection along the first eigendirection at each iteration. 
-%
-However, this algorithm is {\em ad hoc}, and both the advection procedure and the successive eigendecompositions are very expensive and intractable for larger data set. 
+Given the eigenvectors $\phi_1, \phi_2, \dots, \phi_{m-1} \in \mathbb{R}^m$, we would like to automatically deduce
+which ones capture new directions in the data, and which ones are merely repeated eigendirections.
+%
+This problem was addressed previously in \cite{gerber2007robust} by performing
+successive iterations of diffusion maps, interspersed with advection along the first eigendirection at each iteration.
+%
+However, both the advection
+procedure and the successive eigendecompositions are expensive and soon become intractable for larger data sets, 
+and the approach is somewhat {\em ad hoc}.
 %TODO: elaborate on their method and find other advantages of our proposed method (for example, theirs is ad hoc).
 %
-Here, we propose an alternative approach to address the problem of repeated eigendirections. 
-%
-Our approach is motivated by simple trigonometric arguments, such as the fact that $\cos2x$ (a repeated eigendirection) can be written as a function of $\cos x$ (a unique eigendirection). 
-%
-We therefore attempt to fit a function $f(\phi_1, \dots, \phi_{k-1})$ to $\phi_{k}$; if the resulting fit is accurate, we assume $\phi_{k}$ is a repeated eigendirection of $\phi_1, \dots, \phi_{k-1}$.
-%
-We use a local linear function 
+Here, we propose an alternative approach to address the problem of repeated eigendirections
+motivated by simple trigonometric arguments (for example, the repeated eigendirection $\cos2x$ 
+can be written as a function of the unique eigendirection $\cos x$).
+%
+We therefore attempt to fit a function $f(\phi_1, \dots, \phi_{k-1})$ to $\phi_{k}$; if the
+resulting fit is accurate, we assume $\phi_{k}$ is a repeated eigendirection of $\phi_1, \dots, \phi_{k-1}$.
+%
+We use a {\em local} linear function
 \begin{equation}
@@
 %
-We typically take $\epsilon_{reg} = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k-1}(i)$, as we empirically found this choice to yield good results. 
+We typically take $\epsilon_{reg} = M / 3$, where $M$ is the median of the pairwise distances between $\Phi_{k-1}(i)$,
+as we empirically found this choice to yield good results.
 %
@@
 %
-We assume small value of $r_k$ implies that $\phi_k$ is a harmonic of previous modes, i.e., a repeated eigendirction, and conversely, a large value of $r_{k}$ indicates that $\phi_{k}$ parameterizes a unique eigendirection in the data.
+We reason that a small value of $r_k$ implies that $\phi_k$ is a harmonic of previous modes, i.e.,
+a repeated eigendirection, and conversely, a large value of $r_{k}$ indicates that $\phi_{k}$
+parametrizes a new unique eigendirection in the data.
 %
 We choose to set $r_1 = 1$.
-
-The error in \eqref{eq:cv_error} can easily be computed \cite{wasserman2006all}.
-%
-We begin by constructing the matrix
+%
+The error in \eqref{eq:cv_error} can easily be computed (see Section~5.4 of \cite{wasserman2006all}; code is also available at ...).
+
+%The error in \eqref{eq:cv_error} can easily be computed \cite{wasserman2006all}.
+%%
+%We begin by constructing the matrix
+%\begin{equation}
+%X_{k-1}(i) = \begin{bmatrix}
+%1 & \phi_1(1) - \phi_1(i) & \dots & \phi_{k-1}(1)- \phi_{k-1}(i) \\
+%1 & \phi_1(2) - \phi_1(i) & \dots & \phi_{k-1}(2)- \phi_{k-1}(i) \\
+%\vdots & \vdots & \ddots & \vdots \\
+%1 & \phi_1(m) - \phi_1(i) & \dots & \phi_{k-1}(m)- \phi_{k-1}(i)
+%\end{bmatrix}
+%\end{equation}
+%%
+%and the matrix
+%\begin{equation}
+%W_{k-1}(i) = diag \left( K(\Phi_{k-1}(i), \Phi_{k-1}(1)), \dots, K(\Phi_{k-1}(i), \Phi_{k-1}(m)) \right).
+%\end{equation}
+%%
+%We then calculate the smoothing matrix $S_{k-1} \in \mathbb{R}^{m \times m}$, where
+%\begin{equation}
+%S_{k-1} =
+%\begin{bmatrix}
+%e_1^T \left( X_{k-1}^T(1) W_{k-1}(1) X_{k-1}(1) \right) ^{-1} X_{k-1}^T(1) W_{k-1}(1) \\
+%e_1^T \left( X_{k-1}^T(2) W_{k-1}(2) X_{k-1}(2) \right) ^{-1} X_{k-1}^T(2) W_{k-1}(2) \\
+%\vdots \\
+%e_1^T \left( X_{k-1}^T(m) W_{k-1}(m) X_{k-1}(m) \right) ^{-1} X_{k-1}^T(m) W_{k-1}(m) \\
+%\end{bmatrix},
+%\end{equation}
+%%
+%where $e_1^T = (1, 0, \dots, 0)$.
+%%
+%We can then write
+%%
+%\begin{equation}
+%r_{k} = \sqrt{ \frac{\sum_{i=1}^n \left( \frac{ \phi_{k} (i) - \sum_{j=1}^m S_{k-1}(i,j) \phi_{k}(j) }{1-S_{k-1}(i,i)} \right)^2} {\sum_{i=1}^n  \left( \phi_{k} (i) \right)^2 }} .
+%\end{equation}
+
+We propose using {\em only} those eigenvectors for which $r_k$ is large to embed the data,
+as this will yield a more parsimonious representation.
+%
+However, it is important to note that such an embedding also preserves some features of the standard diffusion map embedding.
+%
+Let $I = \{i_1, i_2, \dots, i_d \}$ denote the indices of the identified unique eigendirections (i.e., $r_{i_1}, \dots, r_{i_d}$ are large).
+%
+Under the assumption that $\phi_j = f \left( \phi_{i_1}, \dots, \phi_{i_d} \right)$ for $j \not\in I$, where $f$ is Lipschitz continuous with Lipschitz constant $K$, one can show that
 \begin{equation}
-X_{k-1}(i) = \begin{bmatrix}
-1 & \phi_1(1) - \phi_1(i) & \dots & \phi_{k-1}(1)- \phi_{k-1}(i) \\
-1 & \phi_1(2) - \phi_1(i) & \dots & \phi_{k-1}(2)- \phi_{k-1}(i) \\
-\vdots & \vdots & \ddots & \vdots \\
-1 & \phi_1(m) - \phi_1(i) & \dots & \phi_{k-1}(m)- \phi_{k-1}(i) 
-\end{bmatrix}
-\end{equation}
-%
-and the matrix 
-\begin{equation}
-W_{k-1}(i) = diag \left( K(\Phi_{k-1}(i), \Phi_{k-1}(1)), \dots, K(\Phi_{k-1}(i), \Phi_{k-1}(m)) \right).
-\end{equation}
-%
-We then calculate the smoothing matrix $S_{k-1} \in \mathbb{R}^{m \times m}$, where 
-\begin{equation}
-S_{k-1} =
-\begin{bmatrix}
-e_1^T \left( X_{k-1}^T(1) W_{k-1}(1) X_{k-1}(1) \right) ^{-1} X_{k-1}^T(1) W_{k-1}(1) \\
-e_1^T \left( X_{k-1}^T(2) W_{k-1}(2) X_{k-1}(2) \right) ^{-1} X_{k-1}^T(2) W_{k-1}(2) \\
-\vdots \\
-e_1^T \left( X_{k-1}^T(m) W_{k-1}(m) X_{k-1}(m) \right) ^{-1} X_{k-1}^T(m) W_{k-1}(m) \\
-\end{bmatrix},
-\end{equation}
-%
-where $e_1^T = (1, 0, \dots, 0)$. 
-%
-We can then write
-%
-\begin{equation} 
-r_{k} = \sqrt{ \frac{\sum_{i=1}^n \left( \frac{ \phi_{k} (i) - \sum_{j=1}^m S_{k-1}(i,j) \phi_{k}(j) }{1-S_{k-1}(i,i)} \right)^2} {\sum_{i=1}^n  \left( \phi_{k} (i) \right)^2 }} .
-\end{equation}
-
-We propose using {\em only} those eigenvectors for which $r_k$ is large to embed the data, as this will yield a more parsimonious representation of the data. 
-%
-However, it also preserves some features of the standard diffusion map embedding. 
-%
-In standard diffusion maps, we define the diffusion distance as
-\begin{equation}
-D^2(z(i), z(j)) = \sum_{k=0}^{m-1} \mu_k^t \left( \phi_k(i) - \phi_k(j)  \right)^2.
-\end{equation}
-%
-Let $I = \{i_1, i_2, \dots, i_d \}$ denote the indices of the unique eigenvectors along the manifold (i.e.,$r_{i_1}, \dots, r_{i_d}$ are large). 
-%
-Under the assumption that $\phi_j = f \left( \phi_{i_1}, \dots, \phi_{i_d} \right)$ for $j \not\in I$, and if $f$ is Lipshitz with Lipshitz constant $K$, then one can show that 
-\begin{equation}
-\sqrt{\frac{1}{1+K \sum_{k \not\in I} \mu_k^t}} D^2(z(i), z(j)) \le \sum_{k \in I} \mu_k^t \left( \phi_k(i) - \phi_k(j)  \right)^2 \le D^2(z(i), z(j)).
-\end{equation}
-%
-Therefore, provided the eigenvalues $\mu_k$ decay sufficiently fast, the distance induced by the $d$ eigenvectors identified as parameterizing unique eigendirections $\phi_{i_1}, \dots, \phi_{i_d}$ is equivalent to the standard diffusion distance.
+\frac{1}{1+K^2 \sum_{k \not\in I} \mu_k^{2\tau}} D^2_\tau(z(i), z(j)) \le \sum_{k \in I} \mu_k^{2 \tau} \left( \phi_k(i) - \phi_k(j)  \right)^2 \le D^2_\tau(z(i), z(j)).
+\end{equation}
+%
+Therefore (for finite data, or provided the eigenvalues $\mu_k$ decay sufficiently fast in the limit of infinite data),
+the distance induced by the $d$ eigenvectors $\phi_{i_1}, \dots, \phi_{i_d}$ identified as
+parametrizing unique eigendirections is equivalent to the standard diffusion distance.
 %
@@
 \begin{equation}
-\tilde{D}^2(z(i), z(j)) = \sum_{k \in I} \mu_k^t \left( \phi_k(i) - \phi_k(j)  \right)^2
-\end{equation}
-%
-as the {\em reduced} diffusion distance, and the embedding obtained from the corresponding eigenvectors as the reduced diffusion maps embedding. 
-%
-
-\subsection{Calculating the relative lengths of the manifold} \label{sec:relative_lengths}
-
-For a two-dimensional strip, once the eigenvectors which parameterize unique eigendirections have been determined, the corresponding eigenvalues can be used to calculate the relative lengths along these directions.
-%
-For more general manifolds, we postulate that the eigenvalues can still provide a measure of the relative significance of the unique eigendirections.
-%
-Let $\mu_{i_1}, \mu_{i_2}, \dots$ denote the eigenvalues corresponding to eigenvectors which parameterize unique eigendirections (i.e., those eigenvectors where $r_{i_j}$ is large). 
+\tilde{D}^2_\tau(z(i), z(j)) = \sum_{k \in I} \mu_k^{2 \tau} \left( \phi_k(i) - \phi_k(j)  \right)^2
+\end{equation}
+%
+as the {\em reduced} diffusion distance,
+and the embedding obtained from the corresponding eigenvectors as the reduced diffusion maps embedding.
+%
+
+\subsection{Estimating the relative lengths} \label{sec:relative_lengths}
+
+For a two-dimensional strip, once the eigenvectors which parametrize unique eigendirections have been determined,
+the corresponding eigenvalues can be used to calculate the relative lengths along these directions.
+%
+For more general manifolds, we postulate that the eigenvalues can still provide a measure of the relative
+significance of the unique eigendirections.
+%
+Let $\mu_{i_1}, \mu_{i_2}, \dots$ denote the eigenvalues corresponding to
+eigenvectors which parametrize unique eigendirections (i.e., those eigenvectors where $r_{i_j}$ is large).
 %
@@
 %
-These lengths can then be used to truncate the reduced diffusion maps embedding and determine how many components are required to retain most of the information within the data set; 
-for example, if $L_j$ is smaller than a predefined threshold, then this component can be disregarded.
+These lengths can then be used to evaluate how many components are required to retain
+most of the information within the data set;
+for example, if $L_j$ is smaller than a predefined threshold, then this component can be disregarded in the reduced diffusion maps embedding.
 
@@
 %
-The second and third data sets are nonlinear manifolds which demonstrate the flexibility of our approach.
+The second and third data sets involve nonlinear manifolds which demonstrate the flexibility of our approach.
 
@@
 The empirical eigenvalues are plotted in blue, and the analytical eigenvalues are plotted in red.
-From the eigenvalues which are identified as parameterizing unique eigendirections (indicated by the darker blue bars), the estimated length ratio is (a) 2.2, (b) 4.1, (c) 8.7.}
+The bars are colored by the cross-validation error $r_k$, as defined in \eqref{eq:cv_error}.
+From the eigenvalues which are identified as parametrizing unique eigendirections (indicated by the darker blue bars), the estimated length ratio from \eqref{eq:est_lengths} is (a) 2.2, (b) 4.1, (c) 8.7.
+}
 \label{fig:strip_compare_analytic}
@@
 
-Our first illustrative example consists of three different two-dimensional strip data sets. 
+We consider three different two-dimensional strip data sets.
 %
@@
 %
-Figure~\ref{fig:strip_compare_analytic} shows the data sets and the eigenspectra of the corresponding discrete Laplacians.
-%
-The eigenvalues are colored by the leave-one-out cross-validation error as defined in \eqref{eq:cv_error}; a small value of $r_k$ indicates that the corresponding eigenvector is a repeated eigendirection, while a large value of $r_k$ indicates that the corresponding eigenvector describes a new direction in the data.
-%
-
-We observe that the eigenvalues are consistent with the known analytic eigenvalues of the Laplacian (see \eqref{eq:evals} and \eqref{eq:evals_relationship}), shown in red.
+Figure~\ref{fig:strip_compare_analytic} shows the data sets and the diffusion maps eigenspectra.
+%
+The eigenvalues are colored by the leave-one-out cross-validation error as defined in \eqref{eq:cv_error};
+a small value of $r_k$ indicates that the corresponding eigenvector is a repeated eigendirection, while a
+large value of $r_k$ indicates that the corresponding eigenvector describes a new direction in the data.
+%
+
+The eigenvalues are consistent with the known analytic eigenvalues of
+the Laplacian (see \eqref{eq:evals} and \eqref{eq:evals_relationship}), shown in red.
 %
@@
 %
-As expected, we observe that the gap between the two meaningful eigenvalues increases as the strip becomes longer. 
+As expected, the gap between the two meaningful eigenvalues increases as the strip becomes longer.
 %
@@
 
-Our second illustrative example consists of two different swiss roll data sets.
+Our second illustrative example consists of two different Swiss roll data sets.
 %
@@
 %
-where $\theta$ is sampled such that $z_1, z_2$ are uniformly sampled along the arclength of the spiral, and $t \sim Unif(0,1)$. 
-%
-Note that, in this example, $z_1, z_2, z_3$ are the original data coordinates; however, the data lie on a two-dimensional manifold parameterized by the $\theta$ and $t$. 
-%
-The height of the first swiss roll is $h = 40$, while the height of the second is $h = 20$. 
-%
-Each data set consists of $m=1500$ points, shown shown in Figures~\ref{subfig:swissroll1}--\ref{subfig:swissroll2}.
-%
-%We expect to recover the eigenvector which parameterizes the height farther down in the spectrum for the second data set relative to the first, as this second data set is shorter.
-
-Figures~\ref{subfig:swissroll1_evals}~and~\ref{subfig:swissroll2_evals} shows the eigenvalue spectra from the analysis of the two data sets.
-%
-Similar to Figure~\ref{fig:strip_compare_analytic}, the bars are colored by the leave-one-out cross-validation error, where a small value of $r_k$ indicates that the corresponding eigenvector is a repeated eigendirection.
-%
-From these plots, one can conclude that the first two eigenvectors $\phi_1$ and $\phi_2$ parameterize the first data set, while $\phi_1$ and $\phi_5$ parameterize the second. 
-%
-Figures~\ref{subfig:swissroll1_color1},\ref{subfig:swissroll1_color2},\ref{subfig:swissroll2_color1},~and~\ref{subfig:swissroll2_color2} shows the two data sets, colored by the two eigenvectors identified as parameterizing the unique eigendirections.
-%
-As expected, these eigenvectors are one-to-one with the arc length along the spiral, and the height of the swiss roll.
+where $\theta$ is sampled such that $z_1, z_2$ are uniformly sampled along the arclength of the spiral, and $t$ is uniformly sampled from $[0,1]$.
+%
+Note that, in this example, $z_1, z_2, z_3$ are the original data coordinates; however, the data lie on a
+two-dimensional manifold parametrized by the $\theta$ and $t$.
+%
+The height of the first Swiss roll is $h = 40$, while the height of the second is $h = 20$.
+%
+Each data set consists of $m=1500$ points, shown in Figures~\ref{subfig:swissroll1}~and~\ref{subfig:swissroll2}.
+%
+%We expect to recover the eigenvector which parametrizes the height farther down in the spectrum for the second data set relative to the first, as this second data set is shorter.
+
+Figures~\ref{subfig:swissroll1_evals}~and~\ref{subfig:swissroll2_evals} show the eigenvalue spectra from the analysis of the two data sets.
+%
+Similar to Figure~\ref{fig:strip_compare_analytic}, the bars are colored by the leave-one-out cross-validation error, 
+where a small value of $r_k$ indicates that the corresponding eigenvector is a repeated eigendirection.
+%
+From these plots, one can conclude that the first two eigenvectors $\phi_1$ and $\phi_2$ parametrize the first data set, 
+while $\phi_1$ and $\phi_5$ parametrize the second.
+%
+Figures~\ref{subfig:swissroll1_color1},\ref{subfig:swissroll1_color2},\ref{subfig:swissroll2_color1},~and~\ref{subfig:swissroll2_color2} 
+show the two data sets, colored by the two eigenvectors identified as parametrizing the unique eigendirections.
+%
+As expected, these eigenvectors are one-to-one with the arc length along the spiral, and the height of the Swiss roll.
 
@@
 %
-\caption{Torii data sets (top, see \eqref{eq:torus}) and corresponding diffusion maps eigenvalues (bottom).
-(a) $r_1 = 3$, $r_2 = 1$.
+\caption{Tori data sets (top, see \eqref{eq:torus}) and corresponding diffusion maps eigenvalues (bottom) for (a) $r_1 = 3$, $r_2 = 1$.
 (b) $r_1 = 5$, $r_2 = 1$.
 (c) $r_1 = 10$, $r_2 = 1$.
+The torii are parameterized by two angles: $\theta_1$, which parameterizes the large outer circle, and $\theta_2$, which parameterizes the smaller inner circle.
 In all three data sets, the first two eigenvalues/eigenvectors correspond to $\sin \theta_1$ and $\cos \theta_1$.
@@
 %
-where $r_1 > r_2$ are the outer and inner radii, respectively, and $\theta_1, \theta_2 \sim Unif(0, 2 \pi)$. 
-%
-In this example, $z_1, z_2, z_3$ are the coordinates of the data; however, we expect to obtain from our analysis two eigenfunctions ($\sin \theta_1$ and $\cos \theta_1$) which parameterize the outer circle, and two eigenfunctions ($\sin \theta_2$ and $\cos \theta_2$) which parameterize the inner circle.
-%
-Intuitively, increasing $r_1$ can be viewed as analogous to increasing the ratio of $L_1$ to $L_2$ in the strip example and makes the eigendirections which parameterize the outer circle ($\cos \theta_1$ and $\sin \theta_1$) more dominant compared to the eigendirections with parameterize the inner circle.
-%
-Figure~\ref{fig:torus} shows the eigenspectra from the analysis of three different torii, colored by the leave-one-out cross-validation error. 
+where $r_1 > r_2$ are the outer and inner radii, respectively, and $\theta_1$ and $\theta_2$ are uniformly and independently sampled from $[0, 2\pi)$.
+%
+In this example, $z_1, z_2, z_3$ are the coordinates of the data; however, we expect to obtain from our analysis 
+two eigenfunctions ($\sin \theta_1$ and $\cos \theta_1$) which parametrize the outer circle, and 
+two eigenfunctions ($\sin \theta_2$ and $\cos \theta_2$) which parametrize the inner circle.
+%
+Intuitively, increasing $r_1$ can be viewed as analogous to increasing the ratio of $L_1$ to $L_2$ in the 
+strip example and makes the eigendirections which parametrize the outer circle ($\cos \theta_1$ and $\sin \theta_1$) more 
+dominant compared to the eigendirections with parametrize the inner circle.
+%
+Figure~\ref{fig:torus} shows the eigenspectra from the analysis of three different tori, colored again by the leave-one-out cross-validation error.
 %
@@
 
-Our main motivation for identifying the unique eigendirections is to facilitate the analysis of complex data sets where the underlying structure is not readily apparent. 
-%
-In data from complex dynamical systems, noisy microscopic behavior often gives rise to cohernt macroscopic dynamics. 
+Our main motivation for identifying the unique eigendirections is to facilitate the analysis of 
+complex data sets where the underlying dimensionality and structure is not readily apparent.
+%
+In data from complex dynamical systems, noisy microscopic behavior often gives rise to coherent macroscopic dynamics.
 %
@@
 %
-We will show how we can use our proposed methodology to extract a parameterization of data collected at the microscale which is consistent with the macroscopic behavior, without any {\em a priori} knowledge of the appropriate microscopic or macroscopic model.
-%
-Furthermore, determining the {\em true} dimensionality of such a parameterization by identifying the unique eigendirections in the analysis uncovers the requisite dimensionality of a reduced model which captures the relevant macroscopic behavior.
-%
-Knowing appropriate macroscopic variables and the true dimensionality can help inform modeling efforts and aid in the writing of accurate macroscale models.
-
-Our model problem describes the process of cellular chemotaxis \cite{othmer2000diffusion}, where biological cells exhibit coherent macroscopic dynamics regulated by extracellular sensed signals in order to accomplish tasks such as finding food or navigating away from toxins.
-%
-Several microscopic models have been proposed to describe chemotaxis dynamics \cite{othmer1988models, codling2008random}.
+We will show how we can use our proposed methodology to extract a parametrization of data collected 
+at the microscale which is consistent with a certain macroscopic behavior, without any {\em a priori} knowledge of the appropriate 
+microscopic or macroscopic model.
+%
+Furthermore, determining the {\em true} dimensionality of such a parametrization by identifying 
+the unique eigendirections reveals the requisite dimensionality of a reduced model which captures the relevant macroscopic behavior.
+%
+Knowing appropriate macroscopic variables and the true dimensionality can help inform modeling efforts and aid in 
+positing useful macroscale models.
+
+Our model problem describes the process of cellular chemotaxis \cite{othmer2000diffusion}, 
+where biological cells exhibit coherent macroscopic dynamics regulated by extracellular sensed signals 
+in order to accomplish tasks such as finding food or navigating away from toxins.
+%
+Several microscopic models have been proposed to describe chemotactic dynamics \cite{othmer1988models, codling2008random}.
 %
@@
 %
-This specific example has an analytic macroscopic description in which the macroscopic dynamics of the group of cells depend on the value of a single system parameter.
-%
-This model serves as a ground truth and will allow us to verify our results which are obtained in an unsupervised manner.
+This specific example has an analytic macroscopic description in which the macroscopic 
+dynamics of a large ensemble of cells depend on the value of a single macroscopic system parameter.
+%
+This description serves as a ``ground truth" and will allow us to validate our results which are obtained in an unsupervised manner.
 %
@@
 %
-Here, we will show that our approach, when utilizing the appropriate statistical observers and affinity metric between pairs of observations, uncovers a parameterization of the microscopic data which is consistent with the macroscopic model.
-%
-Furthermore, we will show that changes in dynamical behavior resulting from changes in the system parameters can be automatically detected.
+Here, we will show that our approach, {\em when utilizing the appropriate statistical observers and affinity metric between pairs of observations}, 
+uncovers a parametrization of the microscopic data which is consistent with the macroscopic model.
+%
+Furthermore, we will show that changes in the ``mode" of dynamical behavior resulting from changes in the system parameters can be automatically detected.
 
@@
 
-The microscopic model consists of a collection of $N$ cells (for our simulations, we take $N=1000$) whose states are defined by their positions and velocities on a line, and the dynamics of each cell are governed by a stochastic process.
+The microscopic model consists of a collection of $N$ cells (for our simulations, we take $N=1000$) whose states are 
+defined by their positions and velocities on a line, and the dynamics of each cell are governed by a stochastic process.
 %
@@
 %
-At random times, a cell will ``turn around'' and switch the direction of its velocity (this turning is controlled by extracellular signals). 
-%
-The velocity of each cell randomly switches between $\pm s$ following an (independent) Poisson process with rate $\lambda$.
+The velocity of each cell randomly switches between $\pm s$ following an (independent) Poisson process with rate $\lambda$ (this switching is physically controlled by extracellular signals).
 %
@@
 %
-We note that we have chosen a very specific one-parameter family of initial conditions which lead to simple dynamics and allow us to illustrate of our main points;
-however, our analysis is not restricted to such specific cases and more complex initial conditions could be used. 
+We note that we have chosen a very specific one-parameter family of initial conditions which lead to simple dynamics and allow us to illustrate our main points;
+however, our analysis is not restricted to such specific cases and more complex sets of initial conditions could be used.
 %
@@
 \end{subfigure}
-\caption{Embeddings using the first two diffusion maps eigenvectors computed from simulation data of the velocity jump process with (a--c) $\lambda=1$, $s=1$, and (d--f) $\lambda=400$, $s=20$.
-We set $t_{max} = 10$ and $dt=1$.
-The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions.
-The data are colored by (a, d) $p$, the initial probability of a particle moving to the right, (b, e) $t$, time.
-Representative histograms are shown for selected data points, and (c, f) $E \rho^+ - E \rho^-$, the difference between the average position of the left- and right-moving particles.}
+\caption{Embeddings using the first two diffusion maps eigenvectors computed from simulation data of the velocity jump process with $t_{max} = 10$,k $dt=1$, and (a--c) $\lambda=1$, $s=1$, (d--f) $\lambda=400$, $s=20$.
+The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of cell positions.
+The data are colored by (a, d) $p$, the initial probability of a cell moving to the right, (b, e) $t$, time, and (c, f) $E \rho^+ - E \rho^-$, the difference between the average position of the left- and right-moving cells.
+Representative histograms of the cell positions are shown for selected data points.
+}
 \label{fig:dmaps_embed_emd}
@@
 %
-One is the appropriate {\em observers} of the system. 
+One is the appropriate {\em observers} of the system state.
 %
@@
 %
-Two is a {\em distance metric} between the observations that captures a notion of locality: observations which we perceive to be similar should have a small distance.
-
-For the chemotaxis example, we use histograms of the particle positions as observers. 
-%
-Histograms are invariant to the indexing of the particles, while retaining information about the spatial locations of the particles.
+The second is a {\em distance metric} between the observations that captures a notion of locality: 
+observations which we perceive to be similar should yield a small value for this distance.
+
+For the chemotaxis example, we use histograms of the cell positions as observers.
+%
+Histograms are invariant to the indexing of the cells, while retaining information about the relative spatial locations of the cells.
 %
@@
 %
-Instead of the standard Euclidean distance, we use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between pairs of histograms. 
+Instead of the standard Euclidean distance, we use the Earth Mover's Distance (EMD) \cite{rubner2000earth} as the metric between pairs of histograms.
 %
@@
 %
-It therefore not only considers where the densities are inconsistent, but also how far apart the inconsistencies are.
-%
-Although the brute-force computation of the EMD is computationally expensive, there has been a plethora of work in developing efficient algorithms for its approximation \cite{Pele-eccv2008, Pele-iccv2009}.
-%
-For the specific case of one-dimensional data, the EMD is equivalent to the $L_1$-norm between the cumulative distribution functions of the data \cite{rubner2000perceptual}, which can be approximated from histograms as
+It therefore not only considers where the densities are inconsistent, but also how far apart the inconsistencies are from each other.
+%
+Although the brute-force computation of the EMD is computationally expensive, 
+there has been a plethora of work in developing efficient algorithms for its approximation \cite{Pele-eccv2008, Pele-iccv2009, leeb2014lipschitz}.
+%
+For the specific case of spatially one-dimensional data, the EMD is equivalent to the $L_1$-norm between the 
+cumulative distribution functions of the data \cite{rubner2000perceptual}, which can be approximated from histograms as
 \begin{equation}
-\| z(i) - z(j) \|_{EMD} \approx \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|,
-\end{equation}
-where the histograms $z(i), z(j)$ are defined on $n$ equally-spaced bins in $\mathbb{R}$ (for our simulations, we will take $n=32$), and $z_k$ denotes the $k^{th}$ bin. 
-%
-We note that although histograms are not theoretically essential for using EMD, they make the computation feasible.
+\| z(i) - z(j) \|_{EMD} \approx \sum_{l=1}^{n} \left| \sum_{k=1}^l z_k(i) - \sum_{k=1}^l z_k(j) \right|;
+\end{equation}
+the histograms $z(i), z(j)$ are defined on $n$ equally-spaced bins in $\mathbb{R}$ (for our simulations, we will take $n=32$), and $z_k$ denotes the $k^{th}$ bin.
+%
+We note that although histograms are not essential to the theory of EMD, they make the computation practical.
 
@@
 %
-\caption{Diffusion maps embeddings for three different chemotaxis simulation data sets.
-(a) $\lambda = 100$, $s = 10$.
-(b) $\lambda = 1600$, $s = 40$.
-(c) $\lambda = 6400$, $s = 80$.
-We set $t_{max} = 10$ and $dt=1$.
-The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions.
-For each data set, the eigenvalue spectrum, colored by the cross-validation error, is shown.
-From the spectra, we can see that the first two components are informative for (a), components 1 and 3 are informative for (b), and components 1 and 4 are informative for (c).
-The corresponding reduced diffusion maps embeddings are shown in the middle row.
-For comparison, the standard diffusion maps embeddings using the first two components are also shown in the bottom row in (b) and (c).}
+\caption{Analysis of three different chemotaxis simulation data sets. (a) $\lambda = 100$, $s = 10$. 
+(b) $\lambda = 1600$, $s = 40$. (c) $\lambda = 6400$, $s = 80$. We set $t_{max} = 10$ and $dt=1$, and the 
+distances used in the diffusion maps kernel are the earth mover's distances between the histograms of cell positions. 
+For each data set, the eigenvalue spectrum, colored by the cross-validation error, is shown in the top row. 
+From the spectra, we can see that the first two components are informative for (a), that components 1 and 3 are informative for (b), 
+and that components 1 and 4 are informative for (c). 
+The corresponding reduced diffusion maps embeddings are shown in the middle row. For comparison, 
+the standard diffusion maps embeddings using the first two components are also shown in the bottom row in (b) and (c).}
 %
@@
 %
-One set of simulations (Figures~\ref{subfig:small_lambda_p}--\ref{subfig:small_lambda_rho}) is for a small value of $\lambda$, and the other set (Figures~\ref{subfig:large_lambda_p}--\ref{subfig:large_lambda_rho}) is for a large value of $\lambda$.
-%
-For both sets of simulations, the macroscopic variables $p$, which controls the initial distribution of the cells, and $t$, the time, are well-correlated with the eigenvectors $\phi_1$ and $\phi_2$. 
+One set of simulations (Figures~\ref{subfig:small_lambda_p}--\ref{subfig:small_lambda_rho}) is for a relatively small value of $\lambda$, 
+while the other set (Figures~\ref{subfig:large_lambda_p}--\ref{subfig:large_lambda_rho}) is for a larger value of $\lambda$.
+%
+For both sets of simulations, the macroscopic variables $p$, which controls the initial distribution of cell velocities, and $t$, the time, 
+are well-correlated with the eigenvectors $\phi_1$ and $\phi_2$.
+%
+In particular, $p$ characterizes the fraction of initially right-moving cells (alternatively, the initial flux of the right-moving cells). 
 %
@@
 %
-The dominant coordinate $\phi_1$ is correlated with $p$ for the small $\lambda$ case (Figure~\ref{subfig:small_lambda_p}), and correlated with $t$ for the large $\lambda$ case (Figure~\ref{subfig:large_lambda_t}), indicating that the relative importance of $p$ and $t$ changes in the two simulations.
+The dominant coordinate $\phi_1$ is correlated with $p$ for the small $\lambda$ case (Figure~\ref{subfig:small_lambda_p}), 
+and correlated with $t$ for the large $\lambda$ case (Figure~\ref{subfig:large_lambda_t}), indicating that the relative 
+importance of $p$ and $t$ changes in the two simulations.
 %
@@
 
-As illustrated in the synthetic data sets we discussed previously, the two leading eigenvectors are not guaranteed to correspond to unique eigendirections. 
-%
-Figure~\ref{fig:chemotaxis_simulations_harmonics} shows the results of analyzing three different simulations which span a larger range of $\lambda$ values compared to Figure~\ref{fig:dmaps_embed_emd}.
-%
-We see from the eigenspectra and the cross-validation errors $r_k$ that the second unique eigendirection moves down in the spectrum as $\lambda$ increases. 
-%
-The two-dimensional reduced diffusion maps embeddings capture the macroscopic variables $p$ and $t$ (we only show the correspondence of the two identified unique eigendirections with $p$ in the middle row of Figure~\ref{fig:chemotaxis_simulations_harmonics}; the correspondence between these eigendirections and $t$ is similarly consistent).
-%
-In contrast, the standard diffusion maps embeddings using the two leading eigenvectors $\phi_1$ and $\phi_2$ produce uninformative embeddings for large values of $\lambda$, as $\phi_2$ is a repeated eigendirection which does not parameterize any new directions in the data.
-%
-Furthermore, although the eigenvalue spectra do not exhibit any large spectral gaps, the gap between the eigenvalues which correspond to new directions in the data (colored in blue) increases with $\lambda$, indicating that the underlying manifold becomes more narrow/one-dimensional, and the corresponding eigenvectors provide us with an informative two-dimensional parameterization of the data.
-%which is consistent with the expected macroscopic dynamics.
-%
-It now becomes clear that looking at the eigenvectors {\em modulo} repeated eigendirections is essential for extracting an informative parameterization of the data, as well as characterizing the dimensionality of the data.
+As illustrated in the synthetic data sets we discussed previously, the two leading eigenvectors are not 
+guaranteed to correspond to unique eigendirections.
+%
+Figure~\ref{fig:chemotaxis_simulations_harmonics} shows the results of analyzing three different simulations 
+which span a larger range of $\lambda$ values compared to Figure~\ref{fig:dmaps_embed_emd}.
+%
+We see from the eigenspectra and the cross-validation errors $r_k$ that the second unique eigendirection moves 
+down in the spectrum as $\lambda$ increases.
+%
+The two-dimensional reduced diffusion maps embeddings obtained from eigenvectors which parametrize unique eigendirections 
+capture the macroscopic variables $p$ and $t$ (we only show the correspondence of the two identified unique eigendirections 
+with $p$ in the middle row of Figure~\ref{fig:chemotaxis_simulations_harmonics}; the correspondence 
+between these eigendirections and $t$ is similarly consistent).
+%
+In contrast, using the two leading eigenvectors $\phi_1$ and $\phi_2$ produces 
+uninformative embeddings for large values of $\lambda$, as $\phi_2$ is a repeated eigendirection which 
+does not parametrize any new directions in the data.
+%
+Furthermore, although the eigenvalue spectra do not exhibit any large spectral gaps, 
+the gap between the eigenvalues which correspond to new directions in the data (indicated by the dark blue) increases with $\lambda$, 
+indicating that the underlying manifold becomes narrower, and the corresponding eigenvectors 
+provide us with an informative two-dimensional parametrization of the data. %which is consistent with the expected macroscopic dynamics.
+%
+It now becomes clear that looking at the eigenvectors {\em modulo} repeated eigendirections is essential for extracting an informative parametrization of the data, as well as characterizing the effective dimensionality of the data.
 
@@
 
-This particular example has a known analytic macroscopic equation that governs the overall system behavior.
-%
-For a large collection of cells ($N \rightarrow \infty$), the system can be described by the probability density of the cells.
-%
-Let $\rho(x, t)$ denote the probability density of the cells, and let $\rho^-(x, t)$ and $\rho^+(x, t)$ denote the densities of the cells moving in the negative and positive axis directions, respectively.
+This particular example can be usefully modelled through a known analytic macroscopic equation that governs the overall system behavior.
+%
+For a large collection of cells ($N \rightarrow \infty$), the system can be described using the evolution of the cell density.
+%
+Let $\rho(x, t)$ denote this evolving cell density field, and let $\rho^-(x, t)$ and $\rho^+(x, t)$ denote the 
+densities of the left-moving and of the right-moving cells, respectively.
 %
@@
 %
-Alternatively, \eqref{eqn:coupled_pdes} can be rewritten as one, second--order PDE:
+Alternatively, \eqref{eqn:coupled_pdes} can be rewritten as a single, second--order PDE (the telegrapher equation):
 \begin{equation} \label{eq:second_order_pde}
@@
 %
-From \eqref{eq:second_order_pde}, we see that, for fixed values of $\lambda$ and $s$, the macroscopic state of the system (the probability density of the cells) is indeed a function of two variables $p$ and $t$.
-%
-This implies that for fixed $\lambda$ and $s$, the {\em microscopic} data in a high-dimensional ambient space (e.g., the positions of all $N$ cells) should lie on a two-dimensional manifold parameterized by $p$ (which controls the initial ratio of $\rho^+$ and $\rho^-$) and $t$ (which describes the evolution),
-and uncovering the low-dimensional structure of the microscopic data reveals this manifold.
+From \eqref{eq:second_order_pde}, we see that, for fixed values of $\lambda$ and $s$, the macroscopic 
+state of the system (the probability density of the cells) is indeed a function of the initial condition (parametrized by $p$) and the time $t$.
+%
+This implies that for fixed $\lambda$ and $s$, the {\em microscopic} data in a high-dimensional ambient space (e.g., the positions of all $N$ cells) should lie on a two-dimensional manifold parametrized by $p$ (which controls the initial ratio of $\rho^+$ and $\rho^-$) and $t$ (which describes the evolution);
+uncovering the low-dimensional structure of the microscopic data helps reveal this manifold, leading to a subsequent physically meaningful
+parametrization of it.
 %
@@
 
-We consider two asymptotic regimes of simulation.
+Guided by the known theory of the model macroscopic PDE system, we consider two asymptotic regimes for the simulation.
 %
@@
 %
-When $\lambda \rightarrow \infty$ with $s^2/\lambda$ fixed, \eqref{eq:second_order_pde} approaches the heat equation,
-\begin{equation}
+When $\lambda \rightarrow \infty$ with $s^2/\lambda$ fixed, \eqref{eq:second_order_pde} approaches the diffusion equation,
+\begin{equation} \label{eq:diff_eqn}
 2 \frac{\partial \rho}{\partial t} = D \frac{\partial ^2 \rho}{\partial x^2},
@@
 %
-The above analysis shows that the initial distribution of velocities of the particles (determined by $p$ in the microscopic simulations) plays a very different role depending on the value of $\lambda$.
-%
-When $\lambda \rightarrow 0$, the dynamics are described by two wave equations, and the initial distribution persists throughout the trajectory.
-%
-When $\lambda \rightarrow \infty$, the dynamics are described by one heat equation, and the initial conditions are insignificant -- the velocity distribution quickly equilibrates and we see purely diffusive behavior.
-
-The relative importance of $p$ and $t$ from the analytic description are consistent with the results in Figure~\ref{fig:dmaps_embed_emd}, where $p$ is correlated with the dominant coordinate when $\lambda$ is small and correlated with the subdominant coordinate when $\lambda$ is large.
-%
-Furthermore, in the small $\lambda$ regime (wave equation), shown in Figures \ref{subfig:small_lambda_p} and \ref{subfig:small_lambda_t}, the points corresponding to small times are more tightly clustered than the points corresponding to large times.
-%
-This is in agreement with the macroscopic model: at small times, the particles are more condensed around $x=0$, and it is more difficult to distinguish the particles moving to the left from the particles moving to the right.
-%
-On the other hand, at large times, once the particles evolve from the origin, this separation is clear.  
-%
-For the large $\lambda$ case (heat equation), shown in Figures \ref{subfig:large_lambda_p} and \ref{subfig:large_lambda_t}, we observe that at small times, the initial distribution $p$ is well organized in the embedding in Figure \ref{subfig:large_lambda_p}, as the skew of the velocity distribution can be seen in the initial displacements.
-%
-On the other hand, for large times, we observe that the initial distribution $p$ is less organized in Figure \ref{subfig:large_lambda_p}, as the velocities have equilibrated and the initial distribution is less detectable in the particle density.
-%
-Overall, in both cases, we obtain, in an unsupervised data-driven manner, an accurate picture of the macroscopic variables that govern the system dynamics.
-
-Furthermore, this analytic macroscopic descrption shows us that when $\lambda$ is small, the system dynamics can be described by the two densities $\rho^+$ and $\rho^-$, and when $\lambda$ is large, the dynamics are described by a single density $\rho = \rho^+ + \rho^-$.
-%
-Figures~\ref{subfig:small_lambda_rho}~and~\ref{subfig:large_lambda_rho} show the data, colored by the difference in the average position of the left- and right-moving particles. 
-%
-Clearly, for small $\lambda$, where both $\rho^+$ and $\rho^-$ are required to describe the dynamics, the data is organized by the differences in the two densities. 
-%
-However, for large $\lambda$, the two densities rapidly equilibrate, and negligible organization based on the differences between the two densities is visible.
+The above analysis shows that the initial distribution of velocities of the cells 
+(determined by $p$ in the microscopic simulations) plays a very different role depending on the value of $\lambda$.
+%
+When $\lambda \rightarrow 0$, the dynamics are described by the telegrapher equation, and the effect of the initial velocity distribution 
+persists throughout the trajectory.
+%
+When $\lambda \rightarrow \infty$, the dynamics are described by a single heat equation, and the initial conditions 
+for the velocity are insignificant -- the velocity distribution quickly equilibrates and we see purely diffusive behavior.
+
+The relative importance of $p$ and $t$ from the analytic description are consistent with the results 
+in Figure~\ref{fig:dmaps_embed_emd}, where $p$ is correlated with the dominant diffusion map coordinate 
+when $\lambda$ is small, and becomes correlated with the subdominant coordinate when $\lambda$ is large.
+%
+Furthermore, in the small $\lambda$ regime (wave equation), shown in 
+Figures \ref{subfig:small_lambda_p}--\ref{subfig:small_lambda_rho}, the points corresponding to 
+small times are more tightly clustered than the points corresponding to large times.
+%
+This is in agreement with the macroscopic model: at small times, the cells are more 
+condensed around $x=0$, and it is more difficult to distinguish the cells moving to the left from the cells moving to the right.
+%%%YGK
+%%% that is because we did not use the velocities in our diffusion maps.....
+%%%
+%
+On the other hand, at large times, once the cells move away from the origin, this separation between left-moving and
+right-moving cells is clear.
+%
+For the large $\lambda$ case (heat equation), shown in Figures \ref{subfig:large_lambda_p}--\ref{subfig:large_lambda_rho}, 
+we observe that at small times (before the cell velocities have sufficient opportunity to switch and equilibrate), the initial velocity distribution $p$ can be well perceived in the
+embedding in Figure \ref{subfig:large_lambda_p}, as the initial imbalance in left-right velocities affects
+the initial displacements.
+%
+On the other hand, for longer times, we observe that the initial distribution $p$ cannot be well perceived
+in the embedding in Figure \ref{subfig:large_lambda_p}, 
+as the velocities have equilibrated and their initial distribution is practically forgotten.
+%
+Overall, in both cases, we obtain, in an unsupervised data-driven manner, a useful low-dimensional embedding of the data.
+%
+Correlating this embedding with potential candidate macroscopic observables (such as time, densities, fluxes, and initial condition statistics) helps elucidate
+an accurate picture of the macroscopic variables that govern the system dynamics.
+
+The analytic macroscopic model suggests that when $\lambda$ is small, the system dynamics 
+can be described by the two densities $\rho^+$ and $\rho^-$, 
+and when $\lambda$ is large, the dynamics can be described by a single density $\rho = \rho^+ + \rho^-$.
+%
+Figures~\ref{subfig:small_lambda_rho}~and~\ref{subfig:large_lambda_rho} show the data, 
+colored by the difference in the average position of the left- and right-moving cells.
+%
+Clearly, for small $\lambda$, where both $\rho^+$ and $\rho^-$ are required to describe the dynamics, 
+the difference between the two densities is a useful observable in parametrizing the data manifold
+(and thus a good macroscopic descriptor, a good ``candidate macroscopic variable").
+%
+However, for large $\lambda$, the two densities rapidly equilibrate, and the difference between the two 
+densities is {\em not} a useful observable in parametrizing the data manifold.
 
@@
 
-This analytic macroscopic description also allows us to emphasize the importance of using an appropriate distance metric.
-%
-We analyzed the two sets of simulation data from Figure~\ref{fig:dmaps_embed_emd} using the standard Euclidean distance between the histograms to compute the distances in \eqref{eq:W}.
-%
-We empirically found that there is no appreciable correlation between the embedding coordinates and the macroscopic variables $p$ and $t$ (the correlations between the embedding coordinates and the governing macroscopic variables are all less than 60\%, with the correlations for the small $\lambda$ case being less than 20\%).
-%
-In contrast, the correlations between the embedding coordinates and the macroscopic variables when using EMD in the diffusion maps calculation are all greater than 80\%, and so using a metric which accurately describes the distances between observations is essential for obtaining an informative parameterization of data.
+This analytic macroscopic description allows us to emphasize the importance of using an appropriate distance metric.
+%
+We also analyzed the two sets of simulation data from Figure~\ref{fig:dmaps_embed_emd} using the standard 
+Euclidean distance between the histograms to compute the distances in \eqref{eq:W}.
+%
+We empirically found that there is no appreciable correlation between the embedding 
+coordinates and what we know to be meaningful macroscopic variables, $p$ and $t$ (the correlations between the embedding 
+coordinates and the governing macroscopic variables are all less than 60\%, with the correlations for 
+the small $\lambda$ case being less than 20\%) with this similarity measure.
+%
+%In contrast, the correlations between the embedding coordinates and the macroscopic variables when using EMD in the diffusion maps 
+calculation are all greater than 80\%
+%
+Clearly, using a metric which meaningfully describes the distances between observations 
+is essential for obtaining an informative parametrization of data.
+%
+We note that the metrics we used in this work only employed observations of cell positions (and not of cell instantaneous
+velocities); if the instantaneous velocities are readily measurable, they could be incorporated in a useful
+and possibly even more informative metric for data analysis.
 
@@
 \caption{Detecting the change in dimensionality.
-The ratio of the log of the two eigenvalues which correspond to unique eigendirections is plotted as a function of $t_{max}$ and $\lambda$.
-Each data point is the average eigenvalue ratio over three data sets, and in each simulation, $dt=t_{max}/10$.
-The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of particle positions.
+The appropriate ratio the two eigenvalues (see \eqref{eq:chemotaxis_eval_ratio}) which correspond to unique eigendirections is plotted as a function of $t_{max}$ and $\lambda$.
+Each data point is the average eigenvalue ratio over three replicate data sets, and in each simulation, $dt=t_{max}/10$.
+The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of cell positions.
 The curve $1/\lambda = t_{max}/N$ is shown in white.
@@
 
-Detecting changes in dimensionality is essential for the analysis and modeling of dynamical systems. 
-%
-For this example, the true dimensionally of a data set can be estimated by looking at the two eigenvalues $\mu_{i_1} \ge \mu_{i_2}$ corresponding to unique eigendirections. 
+Detecting changes in effective dimensionality is essential for the analysis and modeling of (simulated or experimentally observed) dynamical systems.
+%
+The approximate dimensionally of a data set can be estimated by examining the 
+two eigenvalues $\mu_{i_1} \ge \mu_{i_2}$ corresponding to unique eigendirections.
 %
@@
 \end{equation}
-when this ratio becomes small, the data are effectively one-dimensional, as the second direction is very small compared with the first.
-
-Figure~\ref{fig:chemotaxis_compare_timescales_evals} shows the eigenvalue ratio in \eqref{eq:chemotaxis_eval_ratio} as a function of $\lambda$ and the time scale of observation $t_{max}$.
-%
-From this eigenvalue ratio, we detect changes in the estimated dimensionality as we vary the relevant parameters.
+when this ratio becomes small, the data are effectively one-dimensional, as the extent of data along the second direction is very 
+small compared with the first.
+
+Figure~\ref{fig:chemotaxis_compare_timescales_evals} shows the eigenvalue ratio in \eqref{eq:chemotaxis_eval_ratio} as 
+a function of $\lambda$ and the time scale of observation $t_{max}$.
+%
+From this eigenvalue ratio, we can easily detect changes in the estimated dimensionality as we vary the relevant parameters.
 %
@@
 %
-However, when $\lambda$ and/or $t_{max}$ is large, the system rapidly equilbrates and the data are effectively one-dimensional, parameterized by $t$. 
-%
-For this specific example, these changes are consistent with the analytically-predicted shift at 
+However, when $\lambda$ and/or $t_{max}$ is large, the system rapidly equilibrates and the data are effectively one-dimensional, parametrized by $t$.
+%
+For this specific example, these changes are consistent with the analytically-predicted  shift in the dynamical behavior from the telegrapher equation to the diffusion equation.
+%
+From \eqref{eq:second_order_pde}, we can see that $t_{max}$, the time scale of observation, should be larger than the time scale $1/\lambda$ for the diffusion equation limit in \eqref{eq:diff_eqn} to hold; we therefore expect a transition in the dynamical behavior at
 \begin{equation}
-\frac{t_{max}}{N} = \frac{1}{\lambda}.
-\end{equation} 
-%
-This curve is also plotted in Figure~\ref{fig:chemotaxis_compare_timescales_evals} as a function of $\lambda$ and is consistent with the transition predicted by the diffusion maps eigenvalues.
+\frac{t_{max}}{N} \approx \frac{1}{\lambda}.
+\end{equation}
+%
+This curve is also plotted in Figure~\ref{fig:chemotaxis_compare_timescales_evals} and is consistent with the transition predicted by the diffusion maps eigenvalues.
 
@@
 
-This paper addresses the issue of repeated eigendirections, one of the key shortcomings in applying diffusion maps analysis to real complex data. 
-%
-We show that, using local linear regression, we can automatically detect which eigenvectors correspond to unique eigendirections in the data. 
-%
-From this detection, we can then obtain a reduced diffusion maps embedding which is both parsimonious and endures a metric which is equivalent to the standard diffusion distance. 
-%
-We then showed that these algorithms can allow us to analyze data from a complex dynamical system, enabling the extraction of good reduced coordinates and also the detection of changes in the dimensionality of the macroscopic dynamics.
-%
-In the future, we are confident that our proposed methodology will aid in the analysis of more complex data sets for which the dimensionality of the underlying manifold is unknown.
-
-\section*{Acknowledgment}
-The authors would like to thank TODO.
+This paper addresses the issue of repeated eigendirections, a critical issue in applying diffusion maps to the analysis of complex dynamic data.
+%
+We show that, using local linear regression, we can automate the detection of eigenvectors corresponding to unique directions in the
+geometry of the data.
+%
+From this detection, we can then obtain a {\em reduced diffusion maps embedding} which is both parsimonious and 
+induces a metric which is equivalent to the standard diffusion distance.
+%
+We showed that this algorithm enables us to better analyze data from a complex dynamical system, 
+enabling the extraction of good reduced coordinates and the detection of changes in the dimensionality of the macroscopic dynamics.
+%
+We are confident that the proposed methodology will be helpful in the analysis of complex data sets for which 
+the dimensionality of the underlying manifold is unknown.
+
+\section*{Acknowledgments}
+ C.J.D.
+was supported by the Department of Energy Computational Science Graduate Fellowship (CSGF), grant number DE-FG02-97ER25308, and the National Science Foundation Graduate Research Fellowship, grant number DGE 1148900.
+%
+I.G.K. was supported by the National Science Foundation (CS\&E program).
 
@@
 \bibliography{../../../references/references}
+%\bibliography{references}
 

### 2015-05-20 chemotaxis/paper/draft1.tex
updated chemotaxis with Yannis' edits

@@
 \usepackage{subcaption}
+\usepackage{hyperref}
 
@@
 \address[princeton]{Department of Chemical and Biological Engineering, Princeton University, Princeton, NJ, 08540, USA}
-\address[technion]{Technion - Israel Institute of Technology, Haifa, 3200003, Israel}
+\address[technion]{Department of Electrical Engineering, Technion - Israel Institute of Technology, Haifa, 3200003, Israel}
 \address[yale]{Department of Mathematics, Yale University, New Haven, CT, 06520, USA}
@@
 %
-We then apply our algorithm to data collected from a stochastic simulation of cellular chemotaxis, where our approach for factoring out repeated eigendirections allows us to detect changes in dynamical behavior and the underlying
+We then apply our algorithm to data collected from a stochastic model of cellular chemotaxis, where our approach for factoring out repeated eigendirections allows us to detect changes in dynamical behavior and the underlying
 intrinsic system dimensionality directly from data.
@@
 %
-The premise is that these coordinates, obtained in a data-driven manner, are the right macroscopic ``observables",
+The premise is that these coordinates, obtained in a data-driven manner, are the right macroscopic ``observables", i.e., 
 the variables which parametrize the macroscale dynamical behavior of the system, thus enabling the construction of a reduced macroscale model.
@@
 %
-More recently they have been applied to experimental and simulation data,
+More recently they have been applied to experimental as well as simulation data,
 enabled by advances in data representation (observers) and
-metrics \cite{rubner2000earth,mallat2012group,talmon2013empirical,zhao2014rotationally, rohrdanz2011determination}.
+metrics \cite{rubner2000earth,mallat2012group,talmon2013empirical,zhao2014rotationally, rohrdanz2011determination, talmon2015manifold}.
 %
@@
 
-In this paper, we propose an algorithm to automatically identify the unique eigendirections using local linear regression \cite{wasserman2006all}.
+In this paper, we propose an algorithm to automatically identify such unique eigendirections using local linear regression \cite{wasserman2006all}.
 %
@@
 %
-By using a particular manifold learning technique based on the construction of a Laplace operator, namely, diffusion maps, we will show how to extract a $d$-dimensional parametrization of the observations which is consistent with the geometry of the manifold \cite{Belkin2003, coifman2005geometric, singer2008non}.
+By using a particular manifold learning technique based on the construction of a Laplace operator, namely, {\em diffusion maps}, we will show how to extract a $d$-dimensional parametrization of the observations which is consistent with the geometry of the manifold \cite{Belkin2003, coifman2005geometric, singer2008non}.
 
@@
 \caption{(a) Two-dimensional continuous strip colored by the eigenfunctions $\tilde{\phi}_{1, 0} = \cos \left( {\pi z_1}/{L_1} \right)$, and $\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$.
-(b) Two-dimensional strip with uniform sampling colored by the first and fourth (non-trivial) eigenvectors of the discrete Laplacian.
-(c) Two-dimensional strip with data sampled from a Gaussian distribution in $z_1$ and sampled uniformly in $z_2$, colored by the first and third (non-trivial) eigenvectors of the discrete Laplacian.
-Note that in all cases we uncover parametrizations which are one-to-one with $z_1$ and $z_2$.}
+(b) Data, uniformly sampled from a two-dimensional strip, colored by the first and fourth (non-trivial) eigenvectors of the discrete Laplacian.
+(c) Data, sampled from a Gaussian distribution in $z_1$ and sampled uniformly in $z_2$, colored by the first and third (non-trivial) eigenvectors of the discrete Laplacian.
+Note that in all cases we uncover parametrizations which are one-to-one with $z_1$ and $z_2$ on the domain.}
 \end{figure}
@@
 inherent to the manifold learning setup based on Laplace operators,
-and that the identification of the unique eigendirections is nontrivial.
+and that the identification of the unique eigendirections requires some additional effort.
 %
@@
 
-To illustrate why these eigenfunctions provide appropriate coordinates, consider a two-dimensional strip with edge lengths $L_1$ and $L_2$.
+To review and illustrate why these eigenfunctions provide appropriate coordinates, consider a two-dimensional strip with edge lengths $L_1$ and $L_2$.
 %
@@
 $\tilde{\phi}_{0, 1} = \cos \left( {\pi z_2}/{L_2} \right)$ are one-to-one with the $z_1$ and $z_2$ coordinates,
-respectively, and therefore provide a parametrization of the underlying manifold (see Figure~\ref{subfig:strip_efuncs}).
+respectively, on the domain, and therefore provide a parametrization of the underlying manifold (see Figure~\ref{subfig:strip_efuncs}).
 %
@@
 %
-Clearly, this problem of repeated eigendirections also arises for more complex manifolds of more than two variables.
+Clearly, this problem of repeated eigendirections also arises for more curved of more than two variables.
 %
@@
 it has been observed empirically that they often provide appropriate coordinates
-to usefully parametrize more complex, nonlinear manifolds.
+to usefully parametrize more complicated, nonlinear manifolds.
 %
@@
 %
+The underlying assumption is that, even in cases where the samples embedded in the high dimensional space lie on a highly nonlinear manifold, within $\epsilon$-local neighborhoods, the Euclidean distance respects the tangent plane to the manifold and thus locally conveys meaningful relationships,.
+%
 We then construct the diagonal matrix $\mathbf{D} \in \mathbb{R}^{m \times m}$, with $\mathbf{D}_{ii} = \sum_j \mathbf{W}_{ij}$,
-and form the matrix $\widetilde{\mathbf{W}} = \mathbf{D}^{-\alpha} \mathbf{W} \mathbf{D}^{-\alpha}$, where $0 < \alpha < 1$.
+and form the matrix $\widetilde{\mathbf{W}} = \mathbf{D}^{-\alpha} \mathbf{W} \mathbf{D}^{-\alpha}$, where $0 \le \alpha \le 1$.
 %
@@
 \begin{align} 
+\label{eq:limiting_op1}
 \frac{1}{\epsilon^2}(\mathbf{I}-\mathbf{A}) \phi &\rightarrow \nabla^2 \phi - 2\nabla U \cdot \nabla \phi, &&\alpha = 0 \\
+\label{eq:limiting_op2}
 \frac{1}{\epsilon^2}(\mathbf{I}-\mathbf{A}) \phi &\rightarrow \nabla^2 \phi - \nabla U \cdot \nabla \phi, &&\alpha = 1/2 \\
+\label{eq:limiting_op3}
 \frac{1}{\epsilon^2}(\mathbf{I}-\mathbf{A}) \phi &\rightarrow \nabla^2 \phi, &&\alpha = 1
@@
 The different limit operators, depending on the choice of $\alpha$, imply that nonuniform
-sampling on the manifold may have different effects.
+sampling on the manifold may have different effects via the potential $U$ in \eqref{eq:limiting_op1}--\eqref{eq:limiting_op3}.
 %
@@
 %
-However, these eigenvectors are {\em not} guaranteed to appear as the first two (non-trivial) eigenvectors,
+However, these eigenvectors are {\em not} guaranteed to appear as {\em the first two} (non-trivial) eigenvectors,
 as harmonics of the first eigendirection (i.e., $\cos \left( n \pi z_1 / L_1 \right)$ with $n > 1$) could appear before the second.
@@
 %
-Automatic detection of which eigenvectors are harmonics of each other would be clearly useful.
+Automatic detection of which eigenvectors are harmonics of each other is clearly useful.
 %
@@
 %
-Furthermore, the corresponding eigenvalues provide us with an estimate of the relative lengths of the data set along these unique eigendirections.
+Furthermore, when the data is uniformly sampled from the underlying manifold, the corresponding eigenvalues provide us with an estimate of the relative lengths of the data set along these unique eigendirections.
 
@@
 \end{subfigure}
-\caption{ Two-dimensional strip with uniform sampling colored by the first four (non-trivial) eigenvectors from diffusion maps.
+\caption{ Data, uniformly sampled from a two-dimensional strip, colored by the first four (non-trivial) eigenvectors from diffusion maps.
 Note that the first and fourth eigenvectors are one-to-one with $z_1$ and $z_2$, respectively.
@@
 However, both the advection
-procedure and the successive eigendecompositions are expensive and soon become intractable for larger data sets, 
-and the approach is somewhat {\em ad hoc}.
-%TODO: elaborate on their method and find other advantages of our proposed method (for example, theirs is ad hoc).
+procedure and the successive eigendecompositions are expensive and soon become intractable for larger data sets.
 %
@@
 \begin{equation}
-\phi_k \approx \alpha + \beta^T \Phi_{k-1}
+\phi_k(i) \approx \alpha_k(i) + \beta_k^T(i) \Phi_{k-1}(i)
 \end{equation}
@@
 %
-$\Phi_{k-1} = \begin{bmatrix} \phi_1 & \dots & \phi_{k-1} \end{bmatrix}^T$,
-$\alpha \in \mathbb{R}$, and $\beta \in \mathbb{R}^{k-1}$.
-%
-The coefficients $\alpha$ and $\beta$ are functions of $\Phi_{k-1}$ because we use a {\em local} linear fit in the $k-1$-dimensional $\Phi_{k-1}$ space, and so the coefficients change as a function of the domain.
+$\Phi_{k-1}(i) = \begin{bmatrix} \phi_1(i) & \dots & \phi_{k-1}(i) \end{bmatrix}^T$,
+$\alpha_k(i) \in \mathbb{R}$, and $\beta_k(i) \in \mathbb{R}^{k-1}$.
+%
+The coefficients $\alpha_k(i)$ and $\beta_k(i)$ are not constant because we use a {\em local} linear fit in the $k-1$-dimensional $\Phi_{k-1}$ space, and so the coefficients change as a function of the domain.
 
@@
 %
-where $\epsilon_{reg}$ is the kernel scale.
+where $\epsilon_{reg}$ is the kernel scale for the regression algorithm.
 %
@@
 %
-The error in \eqref{eq:cv_error} can easily be computed (see Section~5.4 of \cite{wasserman2006all}; code is also available at ...).
+The error in \eqref{eq:cv_error} can easily be computed (see Section~5.4 of \cite{wasserman2006all}; code is also available at \url{http://ronen.eew.technion.ac.il/sample-page/journal/}).
 
@@
 %
-However, it is important to note that such an embedding also preserves some features of the standard diffusion map embedding.
+However, it is important to note that such an embedding also preserves certain features of the standard diffusion map embedding.
 %
@@
 as the {\em reduced} diffusion distance,
-and the embedding obtained from the corresponding eigenvectors as the reduced diffusion maps embedding.
-%
-
-\subsection{Estimating the relative lengths} \label{sec:relative_lengths}
+and the embedding obtained from the corresponding eigenvectors as the {\em reduced} diffusion maps embedding.
+%
+
+\subsection{Estimating the ``relative lengths''} \label{sec:relative_lengths}
 
@@
 %
-For more general manifolds, we postulate that the eigenvalues can still provide a measure of the relative
+For more general manifolds (at least for the case of uniform sampling), we postulate that the eigenvalues can still provide a measure of the relative
 significance of the unique eigendirections.
@@
 These lengths can then be used to evaluate how many components are required to retain
-most of the information within the data set;
-for example, if $L_j$ is smaller than a predefined threshold, then this component can be disregarded in the reduced diffusion maps embedding.
+most of the information within the data set, similar to the eigenvalues in principal component analysis.
+%for example, if $L_j$ is smaller than a predefined threshold, then this component can be disregarded in the reduced diffusion maps embedding.
 
@@
 The empirical eigenvalues are plotted in blue, and the analytical eigenvalues are plotted in red.
-The bars are colored by the cross-validation error $r_k$, as defined in \eqref{eq:cv_error}.
+The bars are colored by $r_k$, as defined in \eqref{eq:cv_error}.
 From the eigenvalues which are identified as parametrizing unique eigendirections (indicated by the darker blue bars), the estimated length ratio from \eqref{eq:est_lengths} is (a) 2.2, (b) 4.1, (c) 8.7.
@@
 %
-The eigenvalues are colored by the leave-one-out cross-validation error as defined in \eqref{eq:cv_error};
+The eigenvalues are colored by the leave-one-out cross-validation error $r_k$ as defined in \eqref{eq:cv_error};
 a small value of $r_k$ indicates that the corresponding eigenvector is a repeated eigendirection, while a
@@
 \begin{equation}
-\begin{aligned}
-z_1 =& \theta \cos \theta \\
-z_2 =& \theta \sin \theta \\
-z_3 =& h t
-\end{aligned}
+(z_1, z_2, z_3) = (\theta \cos \theta , \theta \sin \theta, h t)
 \end{equation}
@@
 %
-As expected, these eigenvectors are one-to-one with the arc length along the spiral, and the height of the Swiss roll.
+As expected, these eigenvectors are one-to-one with the arclength along the spiral, and the height of the Swiss roll.
 
@@
 (c) $r_1 = 10$, $r_2 = 1$.
-The torii are parameterized by two angles: $\theta_1$, which parameterizes the large outer circle, and $\theta_2$, which parameterizes the smaller inner circle.
+The tori are parameterized by two angles: $\theta_1$, which parameterizes the large outer circle, and $\theta_2$, which parameterizes the smaller inner circle.
 In all three data sets, the first two eigenvalues/eigenvectors correspond to $\sin \theta_1$ and $\cos \theta_1$.
@@
 %
-\begin{equation}
-\begin{aligned}
-z_1 =& (r_1 + r_2 \cos \theta_2 ) \cos \theta_1 \\
-z_2 =& (r_1 + r_2 \cos \theta_2 ) \sin \theta_1 \\
-z_3 =& r_2 \sin \theta_2
-\end{aligned}
-\label{eq:torus}
+\begin{equation} \label{eq:torus}
+\left( z_1, z_2, z_3 \right) = \left(  
+(r_1 + r_2 \cos \theta_2 ) \cos \theta_1, 
+(r_1 + r_2 \cos \theta_2 ) \sin \theta_1, 
+r_2 \sin \theta_2
+\right) 
 \end{equation}
@@
 %
-Figure~\ref{fig:torus} shows the eigenspectra from the analysis of three different tori, colored again by the leave-one-out cross-validation error.
+Figure~\ref{fig:torus} shows the eigenspectra from the analysis of three different tori, colored again by the leave-one-out cross-validation error $r_k$.
 %
@@
 %
-This specific example has an analytic macroscopic description in which the macroscopic 
-dynamics of a large ensemble of cells depend on the value of a single macroscopic system parameter.
+This specific example has an analytic macroscopic description in which the 
+dynamics of a large ensemble of cells depend on the value of a single system parameter.
 %
 This description serves as a ``ground truth" and will allow us to validate our results which are obtained in an unsupervised manner.
-%
+
 Thus far, we only considered synthetic data sets for which the Euclidean distance between data points served as an informative metric.
@@
 %
-Each data set consists of $10$ simulations, with initial conditions uniformly chosen such that $0.1 \le p  \le 0.9$.
-%
-We allow each simulation to evolve for $t_{max}$ time units with a fixed time step $dt$, and use data with $t > 0$ for analysis.
+Each data set consists of $10$ stochastic simulations, with initial conditions uniformly chosen such that $0.1 \le p  \le 0.9$.
+%
+We allow each simulation to evolve for $t_{max}$ time units, and record the positions of the $N$ cells at $\Delta t$ intervals; we use data with $t > 0$ for analysis.
+%
+We will show that the parameter $t_{obs} \equiv t_{max}/N$, which determines the time scale of observation, is of critical importance when we address the limiting asymptotic behavior.
 
@@
 \end{subfigure}
-\caption{Embeddings using the first two diffusion maps eigenvectors computed from simulation data of the velocity jump process with $t_{max} = 10$,k $dt=1$, and (a--c) $\lambda=1$, $s=1$, (d--f) $\lambda=400$, $s=20$.
+\caption{Embeddings using the first two diffusion maps eigenvectors computed from simulation data of the velocity jump process with $t_{max} = 10$, $\Delta t=1$, and (a--c) $\lambda=1$, $s=1$, (d--f) $\lambda=400$, $s=20$.
 The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of cell positions.
@@
 %
+Here, we choose to only use the positions of the cells in the distance computation; utilizing both the positions and velocities produces qualitatively similar results (not shown).
+%
 We note that although histograms are not essential to the theory of EMD, they make the computation practical.
@@
 \caption{Analysis of three different chemotaxis simulation data sets. (a) $\lambda = 100$, $s = 10$. 
-(b) $\lambda = 1600$, $s = 40$. (c) $\lambda = 6400$, $s = 80$. We set $t_{max} = 10$ and $dt=1$, and the 
+(b) $\lambda = 1600$, $s = 40$. (c) $\lambda = 6400$, $s = 80$. We set $t_{max} = 10$ and $\Delta t=1$, and the 
 distances used in the diffusion maps kernel are the earth mover's distances between the histograms of cell positions. 
-For each data set, the eigenvalue spectrum, colored by the cross-validation error, is shown in the top row. 
+For each data set, the eigenvalue spectrum, colored by $r_k$, is shown in the top row. 
 From the spectra, we can see that the first two components are informative for (a), that components 1 and 3 are informative for (b), 
@@
 %
-In particular, $p$ characterizes the fraction of initially right-moving cells (alternatively, the initial flux of the right-moving cells). 
+In particular, $p$ characterizes the fraction of initially right-moving cells (alternatively, the initial {\em flux} of the right-moving cells). 
 %
@@
 
-This particular example can be usefully modelled through a known analytic macroscopic equation that governs the overall system behavior.
+This particular model system can be usefully described through a known analytic macroscopic equation that governs the overall system behavior.
 %
@@
 %
-It can be shown that, as $N \rightarrow \infty$, the densities satisfy the following set of partial differential equations (PDEs) \cite{othmer2000diffusion}:
+It can be shown that, as $N \rightarrow \infty$, the densities satisfy the following set of coupled partial differential equations (PDEs) \cite{othmer2000diffusion}:
 \begin{equation} \label{eqn:coupled_pdes}
@@
 \frac{\partial \rho^+}{\partial t} + s \frac{\partial \rho^+}{\partial x} & = -\lambda \rho^+ +\lambda \rho^- \\
-\frac{\partial \rho^-}{\partial t} - s \frac{\partial \rho^-}{\partial x} & = \lambda \rho^+ -\lambda \rho^-
+\frac{\partial \rho^-}{\partial t} - s \frac{\partial \rho^-}{\partial x} & = \lambda \rho^+ -\lambda \rho^- .
 \end{aligned}
@@
 \begin{equation} \label{eq:second_order_pde}
-\frac{\partial^2 \rho}{\partial t^2} + 2 \lambda \frac{\partial \rho}{\partial t} = s^2 \frac{\partial ^2 \rho}{\partial x^2}
+\frac{\partial^2 \rho}{\partial t^2} + 2 \lambda \frac{\partial \rho}{\partial t} = s^2 \frac{\partial ^2 \rho}{\partial x^2} .
 \end{equation}
@@
 When $\lambda \rightarrow \infty$, the dynamics are described by a single heat equation, and the initial conditions 
-for the velocity are insignificant -- the velocity distribution quickly equilibrates and we see purely diffusive behavior.
+for the velocity are not especially important, since the velocity distribution quickly equilibrates and we see purely diffusive behavior.
 
@@
 %
-Clearly, for small $\lambda$, where both $\rho^+$ and $\rho^-$ are required to describe the dynamics, 
+Clearly, for small $\lambda$, both $\rho^+$ and $\rho^-$ are required to describe the dynamics, and
 the difference between the two densities is a useful observable in parametrizing the data manifold
@@
 %
-%In contrast, the correlations between the embedding coordinates and the macroscopic variables when using EMD in the diffusion maps 
-calculation are all greater than 80\%
+In contrast, the correlations between the embedding coordinates and the macroscopic variables when using EMD in the diffusion maps 
+calculation are all greater than 80\%.
 %
@@
 %
-We note that the metrics we used in this work only employed observations of cell positions (and not of cell instantaneous
+We note again that the metrics we used in this work only employed observations of cell positions (and not of cell instantaneous
 velocities); if the instantaneous velocities are readily measurable, they could be incorporated in a useful
@@
 The appropriate ratio the two eigenvalues (see \eqref{eq:chemotaxis_eval_ratio}) which correspond to unique eigendirections is plotted as a function of $t_{max}$ and $\lambda$.
-Each data point is the average eigenvalue ratio over three replicate data sets, and in each simulation, $dt=t_{max}/10$.
+Each data point is the average eigenvalue ratio over three replicate data sets, and in each simulation, $\Delta t=t_{max}/10$.
 The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of cell positions.
@@
 \end{equation}
-when this ratio becomes small, the data are effectively one-dimensional, as the extent of data along the second direction is very 
+when this ratio becomes small, the data are {\em effectively} one-dimensional, as the extent of data along the second direction is very 
 small compared with the first.
@@
 %
-For small $\lambda$ and/or short $t_{max}$, the data are effectively two-dimensional, as both $p$ and $t$ are important to the observed dynamics.
-%
-However, when $\lambda$ and/or $t_{max}$ is large, the system rapidly equilibrates and the data are effectively one-dimensional, parametrized by $t$.
+For small $\lambda$ and/or short $t_{obs} = t_{max} / N$, the data are effectively two-dimensional, as both $p$ and $t$ are important to the observed dynamics.
+%
+However, when $\lambda$ and/or $t_{obs}$ is large, the system rapidly equilibrates and the data are effectively one-dimensional, parametrized by time.
 %
@@
 %
-From \eqref{eq:second_order_pde}, we can see that $t_{max}$, the time scale of observation, should be larger than the time scale $1/\lambda$ for the diffusion equation limit in \eqref{eq:diff_eqn} to hold; we therefore expect a transition in the dynamical behavior at
+From \eqref{eq:second_order_pde}, we can see that $t_{obs}$, the time scale of observation, should be larger than the time scale $1/\lambda$ for the diffusion equation limit in \eqref{eq:diff_eqn} to hold; we therefore expect a transition in the dynamical behavior at
 \begin{equation}
-\frac{t_{max}}{N} \approx \frac{1}{\lambda}.
+t_{obs} \approx \frac{1}{\lambda}.
 \end{equation}
@@
 
-This paper addresses the issue of repeated eigendirections, a critical issue in applying diffusion maps to the analysis of complex dynamic data.
+This paper addresses repeated eigendirections, a critical issue in applying diffusion maps to the analysis of complex dynamic data.
 %
@@
 %
-We showed that this algorithm enables us to better analyze data from a complex dynamical system, 
+We showed that this algorithm enables us to more fruitfully analyze data from a complex dynamical system, 
 enabling the extraction of good reduced coordinates and the detection of changes in the dimensionality of the macroscopic dynamics.
 %
-We are confident that the proposed methodology will be helpful in the analysis of complex data sets for which 
+We are confident that the proposed methodology will be helpful in the analysis of high-dimensional data sets for which 
 the dimensionality of the underlying manifold is unknown.
@@
 %
+R.T. was supported by the European Union’s Seventh Framework Programme (FP7) under Marie Curie Grant 630657 and by the Horev Fellowship.
+%
 I.G.K. was supported by the National Science Foundation (CS\&E program).

### 2015-05-22 chemotaxis/paper/manuscript.tex
updated chemotaxis paper for submission

--- 2015-05-20 chemotaxis/paper/draft1.tex
@@
 First, $z_1$ and $z_2$ are not necessarily decoupled in subsequent eigenfunctions,
-and a proper parameterization of the manifold is not necessarily given by the $d$ eigenfunctions associated with the
+and a proper parametrization of the manifold is not necessarily given by the $d$ eigenfunctions associated with the
 smallest $d$ eigenvalues.
@@
 Instead, we are given data {\em sampled} from the underlying manifold,
-and the parameterization of the manifold needs to be uncovered from the data.
+and the parametrization of the manifold needs to be uncovered from the data.
 %
@@
 As discussed previously, the eigenfunctions provide a parametrization of the manifold,
-such that $\mu_j^\tau \phi_{j}(i)$ yields the $j^{th}$ embedding coordinate for $z(i)$, where $\tau \ge 0$ is
+such that $\mu_j^\tau \phi_{j}(i)$ yields the $j^{th}$ embedding coordinate for $z(i)$, where $\tau \in \mathbb{N}$ is
 a parameter (in our examples, we will take $\tau=0$).
 %
-This is the standard diffusion maps embedding \cite{coifman2005geometric, coifman2006geometric}, where
+The parameter $\tau$ corresponds to the number of iterates of the Laplace operator, or in other words, the number of diffusion steps (for more details, see \cite{coifman2006geometric}). 
+%
+%In practice, it has a similar, but not equivalent, effect as the kernel scale \epsilon. 
+% 
+%
+This mapping $z(i) \mapsto \left(\mu_1^\tau \phi_{1}(i), \dots, \mu_{m-1}^\tau \phi_{m-1}(i)\right)$ is the standard diffusion maps embedding \cite{coifman2005geometric, coifman2006geometric}, where
 we order the eigenvectors such that $|\mu_0| \ge |\mu_1| \ge \dots \ge |\mu_{m-1}|$.
@@
 (c) $r_1 = 10$, $r_2 = 1$.
-The tori are parameterized by two angles: $\theta_1$, which parameterizes the large outer circle, and $\theta_2$, which parameterizes the smaller inner circle.
+The tori are parametrized by two angles: $\theta_1$, which parametrizes the large outer circle, and $\theta_2$, which parametrizes the smaller inner circle.
 In all three data sets, the first two eigenvalues/eigenvectors correspond to $\sin \theta_1$ and $\cos \theta_1$.
@@
 %
-We allow each simulation to evolve for $t_{max}$ time units, and record the positions of the $N$ cells at $\Delta t$ intervals; we use data with $t > 0$ for analysis.
+We allow each simulation to evolve for $t_{max}$ time units, and record the states of the $N$ cells every $\Delta t$ time units; we use data with $t > 0$ for analysis.
 %
@@
 %
-Instead of the standard Euclidean distance, we use the Earth Mover's Distance (EMD) \cite{rubner2000earth} as the metric between pairs of histograms.
+Instead of the standard Euclidean distance, we use the earth mover's distance (EMD) \cite{rubner2000earth} as the metric between pairs of histograms.
 %
@@
 %
-When $\lambda \rightarrow \infty$, the dynamics are described by a single heat equation, and the initial conditions 
+When $\lambda \rightarrow \infty$, the dynamics are described by a single diffusion equation, and the initial conditions 
 for the velocity are not especially important, since the velocity distribution quickly equilibrates and we see purely diffusive behavior.
@@
 %
-For the large $\lambda$ case (heat equation), shown in Figures \ref{subfig:large_lambda_p}--\ref{subfig:large_lambda_rho}, 
+For the large $\lambda$ case (diffusion equation), shown in Figures \ref{subfig:large_lambda_p}--\ref{subfig:large_lambda_rho}, 
 we observe that at small times (before the cell velocities have sufficient opportunity to switch and equilibrate), the initial velocity distribution $p$ can be well perceived in the
@@
 \caption{Detecting the change in dimensionality.
-The appropriate ratio the two eigenvalues (see \eqref{eq:chemotaxis_eval_ratio}) which correspond to unique eigendirections is plotted as a function of $t_{max}$ and $\lambda$.
+The appropriate ratio the two eigenvalues (see \eqref{eq:chemotaxis_eval_ratio}) which correspond to unique eigendirections is plotted as a function of $t_{obs}$ and $\lambda$.
 Each data point is the average eigenvalue ratio over three replicate data sets, and in each simulation, $\Delta t=t_{max}/10$.
 The distances used in the diffusion maps kernel are the earth mover's distances between the histograms of cell positions.
-The curve $1/\lambda = t_{max}/N$ is shown in white.
+The curve $t_{obs} = 1/\lambda$ is shown in white.
 }
@@
 Figure~\ref{fig:chemotaxis_compare_timescales_evals} shows the eigenvalue ratio in \eqref{eq:chemotaxis_eval_ratio} as 
-a function of $\lambda$ and the time scale of observation $t_{max}$.
+a function of $\lambda$ and the time scale of observation $t_{obs} = t_{max} / N$.
 %
@@
 %
-For small $\lambda$ and/or short $t_{obs} = t_{max} / N$, the data are effectively two-dimensional, as both $p$ and $t$ are important to the observed dynamics.
+For small $\lambda$ and/or short $t_{obs}$, the data are effectively two-dimensional, as both $p$ and $t$ are important to the observed dynamics.
 %

### 2015-05-22 chemotaxis/strip.m

@@ -6,7 +6,6 @@ close all
 Lx = 4;
 Ly = 1;
 
-
 delta = 0.1;
 
 [X, Y] = meshgrid(0:delta:Lx, 0:delta:Ly);
@@ -22,7 +21,6 @@ cbar = colorbar('peer',gca);
 set(get(cbar,'xlabel'),'String','$\tilde{\phi}_{1,0}$', 'interpreter','latex');
 print('strip_cnts1.eps','-depsc')
 
-return
 make_fig(3,2);
 h = pcolor(X, Y, cos(pi*Y/Ly));
 set(h, 'edgecolor','none');
@@ -131,7 +129,7 @@ for Lx = [2 4 8]
     idx = sort(idx(1:2));
     
     sqrt(log(D(idx(2), idx(2)))/log(D(idx(1), idx(1))))
-
+    
     m = 0:10;
     m1 = reshape(repmat(m, length(m), 1), [], 1);
     m2 = reshape(repmat(m', 1, length(m)), [], 1);
@@ -162,5 +160,79 @@ for Lx = [2 4 8]
     print(sprintf('strip_spectrum_L%d.eps', Lx),'-depsc')
     
     
+end
+
+%%
+
+
+figure;
+
+n = 2000;
+rng(321);
+
+Lx = 4;
+Ly = 1;
+
+data = rand(n, 2);
+data(:,1) = data(:,1) * Lx;
+data(:,2) = data(:,2) * Ly;
+
+plot_idx = 1;
+
+subplot(2,4,plot_idx)
+plot_idx = plot_idx + 1;
+scatter(data(:,1),data(:,2),25, 'b','.')
+axis square
+axis equal
+xlabel('z_1')
+ylabel('z_2')
+
+W = squareform(pdist(data)).^2;
+eps = 0.15^2;
+eps_med_scale = 3;
+
+for alpha=[0 0.5 1]
+    [V, D] = dmaps_weight(W, alpha, 2*eps, 10);
+    res = compute_residuals_DMAPS(V, eps_med_scale);
+    [~, idx] = sort(res, 'descend');
+    idx = sort(idx(1:2));
+    
+    subplot(2,4,plot_idx)
+    plot_idx = plot_idx + 1;
+    make_colored_bars(diag(D(2:end,2:end)), res(2:end))
[…]

### 2015-05-27 reducible_SDE/paper/SIADS_paper/SIADS_manuscript.tex
edits before initial SIADS submission

%%
%% This is file `docultexmm.tex',
%% Documentation for siam multimedia macros for use with LaTeX 2e
%%
%% December 19, 2013
%%
%% Version 1.0.1
%%
%% You are not allowed to change this file.
%%
%% You are allowed to distribute this file under the condition that
%% it is distributed together with all of the files in the siam macro
%% distribution. These are:
%%
%%  siamltexmm.cls (this file)
%%  siam11.clo   (required size option for 11pt papers)
%%  subeqn.clo   (allows equation numbers with lettered subelements)
%%  siam.bst     (bibliographic style file for BibTeX)
%%  docultexmm.tex (documentation file)
%%
%% If you receive only some of these files from someone, please contact:
%% multimedia@siam.org
%%
%% You are not allowed to distribute this file alone. You are not
%% allowed to take money for the distribution or use of either this
%% file or a changed version, except for a nominal charge for copying
%% etc.
%%
%% \CharacterTable
%%  {Upper-case    \A\B\C\D\E\F\G\H\I\J\K\L\M\N\O\P\Q\R\S\T\U\V\W\X\Y\Z
%%   Lower-case    \a\b\c\d\e\f\g\h\i\j\k\l\m\n\o\p\q\r\s\t\u\v\w\x\y\z
%%   Digits        \0\1\2\3\4\5\6\7\8\9
%%   Exclamation   \!     Double quote  \"     Hash (number) \#
%%   Dollar        \$     Percent       \%     Ampersand     \&
%%   Acute accent  \'     Left paren    \(     Right paren   \)
%%   Asterisk      \*     Plus          \+     Comma         \,
%%   Minus         \-     Point         \.     Solidus       \/
%%   Colon         \:     Semicolon     \;     Less than     \<
%%   Equals        \=     Greater than  \>     Question mark \?
%%   Commercial at \@     Left bracket  \[     Backslash     \\
%%   Right bracket \]     Circumflex    \^     Underscore    \_
%%   Grave accent  \`     Left brace    \{     Vertical bar  \|
%%   Right brace   \}     Tilde         \~}

\documentclass[final,leqno,onefignum,onetabnum]{siamltexmm}

\usepackage{amsmath}
\usepackage{amsfonts}

\usepackage{subcaption}

\usepackage{epsfig}

\title{Data-Driven Reduction for Multiscale Stochastic Dynamical Systems \thanks{
%
C.J.D. was supported by the Department of Energy Computational Science Graduate Fellowship (CSGF), grant number DE-FG02-97ER25308, and the National Science Foundation Graduate Research Fellowship, Grant No. DGE 1148900.
%
R.T. was supported by the European Union's Seventh Framework Programme (FP7) under Marie Curie Grant 630657 and by the Horev Fellowship.
%
R.T and R.R.C. were supported by the National Science Foundation, Award No. 1309858.
%
%C.W.G. was supported by...
%
I.G.K. was supported by the NSF and the AFOSR.}}

\author{Carmeline~J.~Dsilva\footnotemark[2] \and Ronen~Talmon\footnotemark[3] \and C.~William~Gear\footnotemark[2] \and Ronald~R.~Coifman\footnotemark[4] \and Ioannis~G.~Kevrekidis\footnotemark[2]\ \footnotemark[5]}

\begin{document}
\maketitle
\newcommand{\slugmaster}{%
\slugger{siads}{xxxx}{xx}{x}{x--x}}%slugger should be set to juq, siads, sifin, or siims

\renewcommand{\thefootnote}{\fnsymbol{footnote}}

\footnotetext[2]{Department of Chemical and Biological Engineering, Princeton University, Princeton, New Jersey, 08544, USA}
\footnotetext[3]{Department of Electrical Engineering, Technion – Israel Institute of Technology, Haifa, Israel, 3200003}
\footnotetext[4]{Department of Mathematics, Yale University, New Haven, Connecticut, 06520, USA}
\footnotetext[5]{Program in Applied and Computational Mathematics, Princeton University, Princeton, New Jersey, 08544, USA}

\renewcommand{\thefootnote}{\arabic{footnote}}

\begin{abstract}

Multiple time scale stochastic dynamical systems are ubiquitous in science and engineering,
and the reduction of such systems and their models to only their slow components is often
essential for scientific computation and further analysis.
%
Rather than being available in the form of an explicit analytical model, often
such systems can only be observed as a data set which exhibits dynamics on several time scales.
%
We will focus on applying and adapting data mining and manifold learning techniques to detect the slow components in such multiscale data.
%
Traditional data mining methods are based on metrics (and thus, geometries) which are
not informed of the multiscale nature of the underlying system dynamics;
such methods cannot successfully recover the slow variables.
%
Here, we present an approach which utilizes  both the local geometry and the {\em local dynamics} within the data set through a
metric which is both insensitive to the fast variables and more general than simple statistical averaging.
%
Our analysis of the approach provides conditions for successfully recovering the underlying slow variables, as well as an empirical protocol guiding the selection of the method parameters.

\end{abstract}

\begin{keywords}
multiscale dynamical systems, Mahalanobis distance, diffusion maps
\end{keywords}

\begin{AMS}
37M10, 62-07
\end{AMS}

\pagestyle{myheadings}
\thispagestyle{plain}
\markboth{C.~J. DSILVA {\it ET AL}}{DATA-DRIVEN REDUCTION OF SDES}

\section{Introduction}

Dynamical systems of engineering interest often contain several disparate time scales.
%
When the evolving variables are strongly coupled, resolving the
dynamics at all relevant scales can be computationally challenging and pose problems for analysis.
%
Often, the goal is to write a reduced system of equations which accurately captures the dynamics on the slow time scales.
%
These reduced models can greatly accelerate simulation efforts, and are more appropriate for integration into larger modeling frameworks.

Following the methods of Mori \cite{mori1965transport}, Zwanzig \cite{zwanzig1961memory}, and others \cite{brey1981nonlinear, chorin2000optimal, hijon2010mori}, one can reduce the number of variables needed to describe a system of differential equations.
%
However, in general, this reduction introduces memory terms.
%
It transforms a system of differential equations into a system of (lower-dimensional) {\em integro-}differential equations,
so that the reduction of the number of variables is counterpoised by the increased complexity of the  reduced model.
%
Here, we will study the special case of evolution equations which contain an inherent time scale separation;
in this case, it is possible, in principle, to obtain a reduced system of differential equations in only the slow variables {\em without memory terms}.
%
Such an analysis crucially hinges on knowing in which variables (or, more generally, functions of variables) one can write such a reduced system of slow evolution equations.

Moving averages and subsampling have often been used in simple cases as appropriate functions of variables in which to formulate slow lower-dimensional models \cite{pavliotis2007parameter}.
%
However, if the underlying dynamics are sufficiently nonlinear, such statistics may
fail to capture the relevant structures and time scales within the data (see Figure~\ref{fig:schematic_fastslow} for a schematic illustration).
%
For well-studied systems, one often has some {\em a priori} knowledge of the appropriate observables (such as phase field variables) with which to formulate the reduced dynamics \cite{chen2002phase, wheeler1992phase}.
%
However, such observables may not be immediately obvious upon inspection for new complex systems, and so we require an automated approach to construct such slow variables.

Given an explicit system of ordinary differential equations,
one can make numerical approximations, such as the quasi-steady state approximation \cite{segel1989quasi} or
the partial equilibrium approximation \cite{gallagher1986combined}, to reduce the system dimensionality without introducing memory terms.
%
There has been some recent analytical work on extending and
generalizing such ideas to more complex systems of equations \cite{ait2008closed, calderon2007fitting, contou2011model, dong2007simplification, givon2004extracting, pavliotis2007parameter,  sotiropoulos2009model}.
%
However, in many instances, closed form, analytical models are not given explicitly,
but can only be inferred from simulation and/or experimental data.
%
We therefore turn to data-driven techniques to analyze such systems and uncover the relevant dynamical modes.
%
In particular, we will use a manifold-learning based approach, as such methods can accommodate nonlinear structures in high-dimensional data.

The core of most manifold learning methods is having a notion of similarity between data points,
usually through a distance metric \cite{Belkin2003, Coifman2006, coifman2005geometric, roweis2000nonlinear, tenenbaum2000global}.
%
The distances are then integrated into a global parametrization of the data, typically through the solution of an eigenproblem.
%
In this paper, we will analyze multiple time scale stochastic dynamical systems using data-driven methods.
%
Standard ``off-the-shelf'' manifold learning techniques which utilize the Euclidean distance are not appropriate
for analyzing data from such multiscale systems, since this metric does not account for the disparate time scales.
%
Research efforts have addressed the construction of more informative distance metrics, which are less sensitive to noise
and can better recover the true underlying structure in the data by suppressing unimportant sources of variability \cite{berry2013time, gepshtein2013image, rubner2000earth, simonyan2013fisher, xing2002distance}.
%
The Mahalanobis distance is one such metric.
%
It was shown that the Mahalanobis distance can remove the effect of {\em observing}
the underlying system variables through a complex, nonlinear function \cite{dsilva2013nonlinear, singer2008non, talmon2013empirical}.
%
Here, we will show the analogy between removing the effects of such nonlinear observation functions (in the context of data analysis), and reducing a dynamical system to remove the effects of the fast variables.
%
Our approach will build a parametrization of the data which is consistent with the underlying slow variables.
%
Because our approach is data-driven, we require no explicit description of the model, and can extract the underlying slow variables
from either simulation or experimental data.
%
Furthermore, the approach implicitly identifies the slow variables within the data and
does not require any {\em a priori} knowledge of the fast or slow variability sources.
%
Even when the underlying dynamical system is complex with nonlinear coupling between the fast and slow variables, we will show that our approach has the potential to isolate the underlying slow modes.

We will present detailed analysis for our method, and provide conditions under which it will successfully recover the slow variables.
%
Furthermore, based on this analysis, we will present data-driven protocols to tune the parameters of the method appropriately.
%
Our presentation and discussion will address two-time-scale stochastic systems; however, we claim that
our framework and analysis readily extends to systems with multiple time scale separations.
%

\begin{figure}[t]

\centering

\begin{subfigure}{0.4\textwidth}
\centering
\epsfig{width=2in, file=schematic_DS1.eps}
\caption{}
\end{subfigure}
%
\begin{subfigure}{0.4\textwidth}
\centering
\epsfig{width=2in, file=schematic_DS2.eps}
\caption{}
\end{subfigure}
%
\caption{(a) Schematic of a two-dimensional, two-time scale ($\tau_1 = 50$ and $\tau_2=2$) ordinary differential equation system where the value of $x_2$ becomes slaved to the value of $x_1$.
%
In such an example, traditional data mining algorithms are sufficient to recover the slow variable. (b) Schematic of a two-scale two-dimensional stochastic dynamical system where {\em the statistics} of $x_2$ become slaved to $x_1$.
%
In such an example, traditional data mining algorithms will not recover the slow variable if the variance in the fast variable is too large. }
\label{fig:schematic_fastslow}
\end{figure}

\section{Multiscale Stochastic Systems} \label{subsec:multiscale_SDE}

Consider the following two-time-scale system of stochastic differential equations (SDEs),
\begin{equation} \label{eq:general_SDE}
\begin{aligned}
dx_i(t) &= a_i(\mathbf{x}(t)) dt + dW_i(t), & \: 1 \le i \le m \\
dx_i(t) &= \frac{a_i(\mathbf{x}(t))}{\epsilon} dt + \frac{1}{\sqrt{\epsilon}} dW_i(t) , & \: m+1 \le i \le n
\end{aligned}
\end{equation}
where $W_i(t)$ are independent standard Brownian motions, $\mathbf{x}(t)  = \begin{bmatrix} x_1(t) & \cdots & x_n(t) \end{bmatrix}^T \in \mathbb{R}^n$, and $\epsilon \ll 1$.
%
In the simple case of a linear drift function, i.e., when $a_i(\mathbf{x}(t)) = \mu _i x_i$ with $\mu_i < 0$, the probability density function of $x_i$ approaches a Gaussian with
(finite) variance $\mu_i$.
%
The time constant of the approach of the variance to equilibrium is $-1/\mu_i$ for $i=1,\ldots,m$ and $-\epsilon/\mu_i$ for $i=m+1,\ldots,n$ \cite{lelievre2013optimal}.
%
Thus, the last $n-m$ variables of \eqref{eq:general_SDE} rapidly approach a local equilibrium measure and exhibit fast dynamics, while the first $m$ variables exhibit slow dynamics.
%
A short burst of simulation will yield a cloud of points which is broadly distributed in the fast directions but narrowly distributed in the slow ones.
%
With the appropriate conditions on $a_i(\mathbf{x})$, the same can be said for more general drift
functions, where $\mu_i$ are the eigenvalues of the Jacobian of $\mathbf{a}(\mathbf{x}) = \begin{bmatrix} a_1(\mathbf{x}) & \cdots & a_n(\mathbf{x}) \end{bmatrix}^T$ \cite{villani2009hypocoercivity}.
%
Therefore, \eqref{eq:general_SDE} defines an $n$-dimensional stochastic system with $m$ slow
variables and $n-m$ fast variables, and $\epsilon$ defines the time scale separation.
%
The ratio of the powers of $\epsilon$ in the drift and diffusion terms in \eqref{eq:general_SDE} is essential,
as we require the square of the diffusivity to be of the same order as the drift as $\epsilon \rightarrow 0$ \cite{berglund2003geometric}.
%
If the diffusivity is larger, then, as $\epsilon \rightarrow 0$, the equilibrium measure will be
unbounded.
%
Conversely, if the diffusivity is smaller, the equilibrium measure will go to $0$ as $\epsilon \rightarrow 0$.

Assuming the sample average of $a_i(\mathbf{x})$ converges to a distribution which is only a function of the slow variables, then by the averaging principle \cite{freidlin2012random}, we can write a reduced SDE in {\em only} the slow variables $x_1, \dots, x_m$.
%
The aim of our work is to show how we can detect such slow variables {\em automatically} from data, in order to help inform modeling efforts and aid in the writing of such reduced stochastic models.
%
In general, we are not given the variables $\mathbf{x}(t)$ from the original SDE system, but instead, we are given some {\em observations}
 in the form $\mathbf{y}(t) = \mathbf{f} (\mathbf{x}(t))$.
%
We assume that $\mathbf{f}: \mathbb{R}^n \mapsto \mathbb{R}^d$, $n \le d$, is a deterministic (possibly nonlinear) function whose image is an $n$-dimensional manifold $\mathcal{M}$ in $\mathbb{R}^d$.
%
% TODO: check whether the inverse condition is not too strong (and we only need it to apply to the domain of the samples)
%
For our analysis, we require $\mathbf{g} = \mathbf{f} ^{-1}$ to be well-defined on $\mathcal{M}$, and both $\mathbf{f}$ and $\mathbf{g}$ to be continuously differentiable to fourth order.
%
Given data $\mathbf{y}(t_1),\ldots,\mathbf{y}(t_N)$ on $\mathcal{M}$ we would like to recover a parametrization of the data that is one-to-one with the
slow variables $x_1, \dots, x_m$.

\section{Local Invariant Metrics}

In order to recover the slow variables from data, we will utilize a local metric that collapses the fast directions.
%
Typically, such a metric averages out the fast variables.
%
However, simple averages are inadequate to describe data which is observed through a complicated nonlinear function.
%
Instead, we propose to use the Mahalanobis distance, which measures distances normalized by the respective variances in each local principal direction.
%
Using this metric, we still retain information about both the fast and slow directions and can
more clearly observe complex dynamic behavior within the data set.

If two points $\mathbf{x}(t_1)$ and $\mathbf{x}(t_2)$ are drawn from an $n$-dimensional
Gaussian distribution with covariance $\mathbf{C}_x$, the Mahalanobis distance between the points is defined as \cite{mahalanobis1936generalized}
\begin{equation}
	\| \mathbf{x}(t_1) - \mathbf{x}(t_2) \| _M = \sqrt{ (\mathbf{x}(t_1) - \mathbf{x}(t_2))^T \mathbf{C}_x^{-1} (\mathbf{x}(t_1) - \mathbf{x}(t_2) )  }.
\end{equation}
In particular, 
%%%YGK+CDS if $\mathbf{x}(t_1)$ and $\mathbf{x}(t_2)$ are samples 
for \eqref{eq:general_SDE}, whose covariance does not depend on $\mathbf{x}$,  $\mathbf{C}_x^{-1} = \mathrm{diag}(e_1, \ldots, e_n)$ is a constant matrix where
\begin{equation} \label{eq:e_def}
\begin{aligned}
e_i =& 1, \: & 1 \le i \le m \\
e_i =& \epsilon, \: & m+1 \le i \le n,
\end{aligned}
\end{equation}
and the Mahalanobis distance between samples is
\begin{equation} \label{eq:rescale_x_dist}
\| \mathbf{x}(t_2) - \mathbf{x}(t_1) \|^2_M = \sum_{i=1}^n e_i \left( x_i(t_2) - x_i(t_1) \right)^2.
\end{equation}
Note that in \eqref{eq:rescale_x_dist}, the fast variables are collapsed and become $\mathcal{O}(\sqrt{\epsilon})$ small,
and so this metric is implicitly insensitive to variations in the fast variables.
%
The metric \eqref{eq:rescale_x_dist} can be rewritten as
\begin{equation} \label{eq:norm_z}
\| \mathbf{x}(t_2) - \mathbf{x}(t_1) \|^2_M = \| \mathbf{z}(t_2) - \mathbf{z}(t_1) \|^2_2
\end{equation}
where
\begin{equation} \label{eq:general_rescale}
z_i(t) = \sqrt{e_i} x_i(t).
\end{equation}
$\mathbf{z}(t)$ is a stochastic process of the same dimension as $\mathbf{x}(t)$, rescaled so that each variable has unit diffusivity.
%
This rescaling transforms our problem from one of detecting the slow variables within dynamic data to one of traditional data mining.
%
The Mahalanobis distance incorporates information about the dynamics and relevant time scales, so that using traditional data mining techniques with this metric will allow us to detect the slow variables in our data \cite{singer2009detecting}.
%
It is important to note that, in practice, we {\em never construct} $\mathbf{z}(t)$ {\em explicitly}.
%
It was shown in \cite{singer2008non} that, assuming $\mathbf{f}$ is bilipschitz, the Mahalanobis distance can be extended to approximate (to fourth order) the Euclidean distance between the rescaled samples $\mathbf{z}(t)$ from accessible $\mathbf{y}(t) = \mathbf{f} (\mathbf{x}(t))$,
%
\begin{equation} \label{eq:mahalanobis}
\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|^2_M = \| \mathbf{z}(t_2) - \mathbf{z}(t_1) \|^2_2 + \mathcal{O}(\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|^4_2).
\end{equation}
%
This approximation is accurate when $\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|$ is small.
%
Because we will integrate these distances into a manifold learning algorithm which only considers local distances, we can recover a parametrization of the data which is consistent with the underlying system variables $\mathbf{x}(t)$, even when the data are obscured by a function $\mathbf{f}$.
%
In Section~\ref{sec:analysis}, we will show how we can approximate this distance directly from data $\mathbf{y}(t)$.
%

\section{Diffusion Maps for Global Parametrization}

From pairwise distances, we want to extract a {\em global} parametrization of the data that represents the slow variables.
%
We will use diffusion maps \cite{Coifman2006, coifman2005geometric}, a kernel-based manifold learning technique, to extract a global parametrization using the local distances that we described in the previous section.
%
Given data $\mathbf{y}(t_1), \dots, \mathbf{y}(t_N)$, we first construct the kernel matrix $\mathbf{W} \in \mathbb{R}^{N \times N}$, where
\begin{equation} \label{eq:dmaps_kernel}
W_{ij} = \exp \left( -\frac{\|\mathbf{y}(t_i) - \mathbf{y}(t_j) \|^2}{\sigma_{kernel}^2} \right).
\end{equation}
Here, $\| \cdot \|$ denotes the appropriate norm (in our case, the Mahalanobis distance), and $\sigma_{kernel}$ is the kernel scale
and denotes a characteristic distance within the data set.
%
Note that $\sigma_{kernel}$ induces a notion of locality: if $\|\mathbf{y}(t_i) - \mathbf{y}(t_j) \| \gg \sigma_{kernel}$, then $W_{ij}$ is negligible.
%
Therefore, we only need our metric to be informative within a ball of radius $\sigma _{kernel}$.
%
We then construct the diagonal matrix $\mathbf{D} \in \mathbb{R}^{N \times N}$, with
\begin{equation}
D_{ii} = \sum_{j=1}^N W_{ij}.
\end{equation}
%
We compute the eigenvalues $\lambda_0, \dots, \lambda_{N-1}$ and eigenvectors $\phi_0, \dots, \phi_{N-1}$ of the matrix $\mathbf{A} = \mathbf{D}^{-1}\mathbf{W}$, and order them such that $1 = \lambda_0 \ge |\lambda_1| \ge \dots \ge |\lambda_{N-1}|$.
%
$\phi_0 = \begin{bmatrix} 1 & 1 & \cdots & 1 \end{bmatrix}^T$
is the trivial eigenvector; the next few eigenvectors provide embedding coordinates for the data, so that $\phi_j(i)$, the $i^{th}$ entry of $\phi_j$, provides the $j^{th}$ embedding coordinate for $\mathbf{y}(t_i)$ (modulo higher harmonics which characterize
the same direction in the data \cite{ferguson2010systematic}).
%

\section{Estimation of the Mahalanobis Distance} \label{sec:analysis}

\begin{figure}[t]
\centering
\epsfig{width=\textwidth, file=schematic.eps}
\caption{Illustration of how to choose $\delta t$ and $\sigma_{kernel}$ appropriately. 	Curvature effects and other nonlinearities should be negligible within a time window $\delta t$ and within a ball of radius $\sigma_{kernel}$.}
\label{fig:schematic}
\end{figure}

As previously mentioned, we do not have access to the original variables $\mathbf{x}(t)$ from the underlying original SDE system.
%
Instead, we only have measurements $\mathbf{y}(t) = \mathbf{f} (\mathbf{x}(t))$, and we want to estimate the Mahalanobis distance between the $\mathbf{x}$ variables from observations $\mathbf{y}(t)$.
%
The traditional Mahalanobis distance is defined for a fixed distribution,
whereas here we are dealing with a distribution that possibly changes as a function of position due to
nonlinearities in the observation function $\mathbf{f}$ and in the drift $\mathbf{a}(\mathbf{x})$.
%
Consequently, we use the following modified definition for the Mahalanobis distance between two points,
\begin{equation} \label{eq:mahalanobis_distance}
 \| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|^2_M =
 \frac{1}{2} (\mathbf{y}(t_2) - \mathbf{y}(t_1))^T \left( \mathbf{C}^{\dagger}(\mathbf{y}(t_1)) + \mathbf{C}^{\dagger}(\mathbf{y}(t_2)) \right) (\mathbf{y}(t_2) - \mathbf{y}(t_1)),
 \end{equation}
where $\mathbf{C}(\mathbf{y}(t))$ is the covariance of the observed stochastic process {\em at the point} $\mathbf{y}(t)$,
%
and $\dagger$ denotes the Moore-Penrose pseudoinverse (since $d$ may exceed $n$).

To motivate this definition of the Mahalanobis distance, we first consider the simple linear case where $\mathbf{f} (\mathbf{x}) = \mathbf{A} \mathbf{x}$, with $\mathbf{A} \in \mathbb{R}^{d \times n}$.
%
The covariance of the observed stochastic process $\mathbf{f} (\mathbf{x})$ is given by $\mathbf{C}=\mathbf{AC}_x\mathbf{A}^T$.
%
Let $\mathbf{A}= \mathbf{U} \mathbf{\Lambda} \mathbf{V}^T$ be the singular value decomposition (SVD) of $\mathbf{A}$, where $\mathbf{U} \in \mathbb{R}^{d \times n}$, $\mathbf{\Lambda} \in \mathbb{R}^{n \times n}$, and $\mathbf{V} \in \mathbb{R}^{n \times n}$.
%
The pseudoinverse of the covariance matrix is $\mathbf{C}^{\dagger} = \mathbf{U} \mathbf{\Lambda}^{-1} \mathbf{V}^T \mathbf{C}_x^{-1} \mathbf{V \Lambda} ^{-1} \mathbf{U}^T$. Consequently, the Mahalanobis distance \eqref{eq:mahalanobis_distance} is reduced to
\begin{equation}
\begin{aligned}
\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|^2_M &= (\mathbf{y}(t_2) - \mathbf{y}(t_1))^T  \mathbf{C}^{\dagger} (\mathbf{y}(t_2) - \mathbf{y}(t_1)) \\
 &= (\mathbf{x}(t_2) - \mathbf{x}(t_1))^T \mathbf{A}^T \mathbf{C}_x^{-1} \mathbf{A} (\mathbf{x}(t_2) - \mathbf{x}(t_1)) \\
 &= (\mathbf{x}(t_2) - \mathbf{x}(t_1))^T \mathbf{V \Lambda U}^T \mathbf{ U \Lambda}^{-1} \mathbf{V}^T \mathbf{C}_x^{-1} \mathbf{V \Lambda} ^{-1} \mathbf{U}^T \mathbf{U \Lambda V}^T (\mathbf{x}(t_2) - \mathbf{x}(t_1)) \\
 &= (\mathbf{x}(t_2) - \mathbf{x}(t_1))^T \mathbf{C}_x^{-1}  (\mathbf{x}(t_2) - \mathbf{x}(t_1)) \\
 &= \| \mathbf{x}(t_2) - \mathbf{x}(t_1) \|^2_M =  \| \mathbf{z}(t_2) - \mathbf{z}(t_1) \|^2_2 .
\end{aligned}
\end{equation}
Hence evaluating the Mahalanobis distances of the observations $\mathbf{y}(t) = \mathbf{f}(\mathbf{x}(t))$ using \eqref{eq:mahalanobis_distance} allows us to estimate the Euclidean distances of the rescaled variables $\mathbf{z}$ (in which the fast coordinates are collapsed).

Following \cite{singer2008non}, we will show via Taylor expansion that the Mahalanobis distance between the observations \eqref{eq:mahalanobis_distance} approximates the Euclidean
distance {\em in the rescaled variables} for general nonlinear observation functions $\mathbf{f}$ (provided $\mathbf{f}$ is bilipschitz and both $\mathbf{f}$ and $\mathbf{f}^{-1}$ are differentiable to fourth order).
%Our approximation of the pairwise distances relies on the Taylor expansion of the measurement function $\mathbf{f}$.
%
\eqref{eq:mahalanobis_distance} cannot be evaluated directly since we do not have access to the covariance matrices, so we will instead estimate the covariances directly from data.
%
We can estimate the covariance $\mathbf{C}(\mathbf{y}(t_0))$ empirically from a set of values $\mathbf{y}(t_1), \dots, \mathbf{y}(t_q)$ drawn from the local distribution at $\mathbf{y}(t_0)$.
%
One way to obtain such a set of points is to run $q$ simulations for a short time, $\delta t$, each starting from $\mathbf{y}(t_0)$.
%
Alternatively, we can consider a single time series of length $q \delta t$ starting from $\mathbf{y}(t_0)$, and then estimate the covariance
from the increments $\Delta \mathbf{y}(t_i) = \mathbf{y}(t_i) -\mathbf{y}(t_{i-1})$.
%
Although we will present analysis and results for the first type of estimation, the second case is often more practical in practice.
%
%It can be shown that if $q \delta t$ is sufficiently small, the errors introduced are of sufficiently low order.

Errors in our estimation of the Mahalanobis distance arise from three sources.
%
One source of error is approximating the function $\mathbf{f}$ locally as a linear function by truncating the Taylor expansion of $\mathbf{f}$ at first order.
%
An additional source of error arises from disregarding the drift in the stochastic process, and assuming that samples are drawn from a Gaussian distribution.
%
The third source comes from finite sampling effects.
%
In this work, we will address and discuss the first two sources of error (the finite sampling effects are the subject of future research).
%
We can control the effects of the errors due to truncation of the Taylor expansion by adjusting $\sigma_{kernel}$; the higher-order terms in this expansion will be small for points which are close, such that adjusting $\sigma_{kernel}$ will allow us to only consider distances which are
sufficiently accurate in our overall computation scheme.
%
Furthermore, we can control the errors incurred by disregarding the drift by adjusting the time scale of our simulation bursts $\delta t$.
%
Figure~\ref{fig:schematic} illustrates some of the issues in choosing the sizes $\delta t$ (or $q \delta t$ if the alternate method is used) and the parameter $\sigma_{kernel}$.
%
We will present both analytical results for the error bounds, as well as an empirical methodology to set the parameters $\sigma_{kernel}$ and $\delta t$ for our method to accurately recover the slow variable(s).

\subsection{Error due to the observation function $\mathbf{f}$}

We want to relate the distance in the rescaled space, $\|\mathbf{z}(t_2) - \mathbf{z}(t_1)\|_2$, to the estimated Mahalanobis distance between the observations $\| \mathbf{y}(t_2) - \mathbf{y}(t_1)\|_M$.
%
We define the error incurred by using the Mahalanobis distance to approximate the true distance as
\begin{equation}
E_M(\mathbf{y}(t_1), \mathbf{y}(t_2)) = \|\mathbf{z}(t_2) - \mathbf{z}(t_1)\|_2^2 - \| \mathbf{y}(t_2) - \mathbf{y}(t_1)\|^2_M .
\end{equation}
%
By Taylor expansion of $\mathbf{g}(y) = \mathbf{f}^{-1}(y)$ around $\mathbf{y}(t_1)$ and $\mathbf{y}(t_2)$ and averaging the two expansions, we obtain
%
\begin{equation} \label{eq:mahanaobis_error}
\begin{aligned}
E_M\left( \mathbf{y}(t_1), \mathbf{y}(t_2) \right)
 =&
\begin{aligned}[t]
 \frac{1}{2} \sum_{i=1}^n \sum_{jkl=1}^{d} &
\left( g_{i, (j)} (\mathbf{y}(t_1)) g_{i, (k,l)} (\mathbf{y}(t_1)) -  g_{i, (j)} (\mathbf{y}(t_2)) g_{i, (k,l)} (\mathbf{y}(t_2)) \right) \times \\
& ({y}_j(t_2) - {y}_j(t_1)) ({y}_k(t_2) - {y}_k(t_1))({y}_l(t_2) - {y}_l(t_1))
\end{aligned} \\
+&
\begin{aligned}[t]
\frac{1}{8} \sum_{i=1}^n \sum_{jklm=1}^d  &
\left( g_{i, (j,k)} (\mathbf{y}(t_1)) g_{i, (l,m)} (\mathbf{y}(t_1)) +  g_{i, (j,k)} (\mathbf{y}(t_2)) g_{i, (l,m)} (\mathbf{y}(t_2)) \right) \times
 \\
&({y}_j(t_2) - {y}_j(t_1))  ({y}_k(t_2) - {y}_k(t_1))({y}_l(t_2) - {y}_l(t_1)) ({y}_m(t_2) - {y}_m(t_1))
\end{aligned} \\
+&
\begin{aligned} [t]
\frac{1}{6} \sum_{i=1}^n \sum_{jklm=1}^d &
\left( g_{i, (j)} (\mathbf{y}(t_1)) g_{i, (k,l,m)} (\mathbf{y}(t_1)) +  g_{i, (j)} (\mathbf{y}(t_2)) g_{i, (k,l,m)} (\mathbf{y}(t_2)) \right) \times \\
& ({y}_j(t_2) - {y}_j(t_1))  ({y}_k(t_2) - {y}_k(t_1))({y}_l(t_2) - {y}_l(t_1))({y}_m(t_2) - {y}_m(t_1))
\end{aligned} \\
+& \mathcal{O} \left(\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|^6_2 \right) ,
\end{aligned}
\end{equation}
%
where
%
\begin{equation}
\begin{aligned}
g_{i,(j)} &= \sqrt{e_i} \frac{\partial g_i}{\partial y_j}
\\
g_{i,(j,k)} &= \sqrt{e_i}  \frac{\partial^2 g_i}{\partial y_j \partial y_k}
\\
g_{i,(j,k,l)} &= \sqrt{e_i}  \frac{\partial^3 g_i}{\partial y_j \partial y_k \partial y_l} .
\end{aligned}
\end{equation}
%
In \cite{singer2008non}, it was shown that the error incurred by using the Mahalanobis distance to approximate the $L_2$-distance between points $\mathbf{z}(t)$ is $\mathcal{O} (\|\mathbf{y}_1 - \mathbf{y}_2 \|_2^4 )$ (see the Supplementary Materials for details).
%
We now see from \eqref{eq:mahanaobis_error} that the error is an explicit function of the second- and higher-order derivatives of $\mathbf{g} = \mathbf{f}^{-1}$ and the distance between samples $\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|_2$.
%
We would like to note that this error does not depend on the dynamics of the underlying stochastic process (as we assume the covariances at each point on the manifold are known), but is only a function of the measurement function $\mathbf{f}$.
%
The parameter $\sigma_{kernel}$ in the diffusion maps calculation determines how much $E_M$ contributes to the overall analysis.
%
From \eqref{eq:dmaps_kernel}, distances which are much greater than $\sigma_{kernel}$ are negligible in the diffusion maps computation because of the exponential kernel.
%
Therefore, we want to choose $\sigma_{kernel}^2$ on the order of $\|\mathbf{y}(t_2) - \mathbf{y}(t_1)\|^2_M$ in a regime where $| E_M(\mathbf{y}(t_1), \mathbf{y}(t_2))|  \ll \|\mathbf{y}(t_2) - \mathbf{y}(t_1)\|^2_M$.
%
This is illustrated in Figure~\ref{fig:schematic}, where we want to choose $\sigma_{kernel}$ small enough so that the curvature and other nonlinear effects (captured in the error term $E_M$) are negligible.
%
This will ensure that the errors in the Mahalanobis distance approximation do not greatly effect our overall analysis.

On first inspection, it would appear that our analysis indicates that $\sigma_{kernel}$ should be chosen arbitrarily small.
%
However, to obtain a meaningful parametrization of the data set, there must be a nonnegligible number of data points within a ball of radius $\sigma_{kernel}$ around each sample.
%
Therefore, the sampling density on the underlying manifold provides a lower bound for $\sigma_{kernel}$.

\subsection{Error due to the dynamics} \label{subsec:cov_est}

To compute the Mahalanobis distance in \eqref{eq:mahalanobis_distance}, we require $\mathbf{C}$, the covariance of the observed stochastic process $\mathbf{y}(t) = \mathbf{f}( \mathbf{x}(t))$.
%
We will use simulation bursts to locally explore the dynamics on the manifold of observations in order to estimate the covariance at a point $\mathbf{y}(t)$ from data \cite{talmon2014manifold, talmon2014intrinsic}.
%
We write the elements of the estimated covariance $\hat{\mathbf{C}}(\mathbf{y}(t), \delta t)$ as
\begin{equation}\label{eq:estimated_cov_expected_value}
\hat{C}_{ij}(\mathbf{y}(t), \delta t)
=
\frac{1}{\delta t} \left( \mathbb{E} \left[ y_i (t+\delta t) y_j (t+ \delta t) \mid \mathbf{y}(t) \right]
- \mathbb{E} \left[ y_i (t+\delta t) \mid \mathbf{y}(t) \right] \mathbb{E} \left[ y_j (t+\delta t) \mid \mathbf{y}(t) \right] \right) ,
\end{equation}
%
where $\delta t > 0$ is the length of the simulation burst.

Due to the drift in the stochastic process and the (perhaps nonlinear) measurement function $\mathbf{f}$, we incur some error by approximating the covariance at a point $\mathbf{y}(t)$ using simulations of length $\delta t > 0$.
%
Define the error in this approximation as
\begin{equation}
\mathbf{E}_C(\mathbf{y}(t), \delta t) = \hat{\mathbf{C}}(\mathbf{y}(t), \delta t) - \mathbf{C}(\mathbf{y}(t)).
\end{equation}
%
%We know that, for a stochastic process following \eqref{eq:general_SDE} and observed through a function $\mathbf{f}$, the covariance $C$ is given by
%\begin{equation} \label{eq:analytical_cov}
%C_{ij}(\mathbf{y}(t)) =
%\sum_{k=1}^n \frac{1}{e_k} \left. \frac{\partial f_{i}}{\partial x_k} \right|_{\mathbf{x}(t)} \left. \frac{\partial f_{j}}{\partial x_k} \right|_{\mathbf{x}(t)}
%\end{equation}
%
By It\^{o}-Taylor expansion of $\mathbf{f}$ and $\mathbf{x}(t)$ \cite{kloeden1992numerical},
%
\begin{equation} \label{eq:cov_error}
\begin{aligned}
E_{C, ij} (\mathbf{x}(t), \delta t) =&
 \frac{1}{\delta t} \sum_{k=1}^n f_{i,(k)}(\mathbf{x}(t)) \mathbb{E} \left[ \int_t^{t+\delta t} \left( \int_{s_2}^{t+\delta t} f_{j,(k,0)}(\mathbf{x}(s_1)) ds_1
+ \int_t^{s_2} f_{j,(0,k)}(\mathbf{x}(s_1)) ds_1 \right) ds_2 \right] \\
&+  \frac{1}{\delta t} \sum_{k=1}^n f_{j,(k)}(\mathbf{x}(t))  \mathbb{E} \left[ \int_t^{t+\delta t} \left( \int_{s_2}^{t + \delta t} f_{i,(k,0)}(\mathbf{x}(s_1)) ds_1
+  \int_t^{s_2} f_{i,(0,k)}(\mathbf{x}(s_1)) ds_1 \right) ds_2 \right] \\
&+  \frac{1}{\delta t} \sum_{k,l=1}^n \mathbb{E} \left[ \int_t^{t+\delta t}\left( \int_t^{s_2} f_{i,(k,l)}(\mathbf{x}(s_1)) dW_{s_1, k}  \right) \left(  \int_t^{s_2} f_{j,(k,l)}(\mathbf{x}(s_1)) dW_{s_1, k} \right) ds_2 \right] \\
&+ \mathcal{O} (\delta t^{3/2})
\end{aligned}
\end{equation}
%
where
\begin{equation}
\begin{aligned}
f_{i,(k)} &= \frac{1}{\sqrt{e_k}} \frac{\partial f_i}{\partial x_k}
\\
f_{i,(k,l)} &= \frac{1}{\sqrt{e_k e_l}} \frac{\partial^2 f_i}{\partial x_k \partial x_l}
\\
f_{i,(k,0)} &= \frac{1}{\sqrt{e_k}} \sum_{l=1}^n \left( \frac{\partial}{\partial x_k} \left( \frac{a_l(\mathbf{x})}{e_l} \frac{\partial f_i}{\partial x_l} \right) + \frac{1}{2 e_l} \frac{\partial^3 f_i}{\partial x_k \partial x_l^2} \right)
\\
f_{i,(0, k)} &= \frac{1}{\sqrt{e_k}} \sum_{l=1}^n \left( \frac{a_l(\mathbf{x})}{e_l} \frac{\partial^2 f_i}{\partial x_k \partial x_l} +\frac{1}{2 e_l}  \frac{\partial^3 f_i}{\partial x_k \partial^2 x_l} \right).
\end{aligned}
\end{equation}
%
From \eqref{eq:cov_error}, the error in the covariance is $\mathcal{O}(\delta t)$ (as the $ds$ integrals are each $\mathcal{O}(\delta t)$ and the $dW$ integrals are each $\mathcal{O}(\sqrt{\delta t})$) and a function of the derivatives of the observation function $\mathbf{f}$ and the drift $\mathbf{a}$.
%
We want to set $\delta t$ such that $\|\mathbf{E}_C \| \ll \| \mathbf{C} \|$
(this is illustrated in Figure~\ref{fig:schematic}), so that the estimated covariances are accurate.
%
Note that in practice, we compute $\hat{\mathbf{C}}$ by running many simulations of length $\delta t$ starting from $\mathbf{x}(t)$, and use the sample average to approximate the expected values in \eqref{eq:estimated_cov_expected_value}.
%
We therefore incur additional error due to finite sampling; this error is ignored for the purposes of this analysis, and quantifying this error is the subject of future research.

Our analysis reveals that the errors decrease with decreasing $\delta t$; at first inspection, one would want to set $\delta t$ arbitrarily small to obtain the highest accuracy possible.
%
However, often in practice, one cannot obtain an arbitrarily refined sampling rate, such that a smaller $\delta t$ results in fewer samples with which to approximate the local covariance.%
When also accounting for these finite sampling errors, and one should take $\delta t$ as long as possible while still maintaining negligable errors from the observation function $\mathbf{f}$ and the drift $\mathbf{a}$.

\section{Illustrative Examples}

For illustrative purposes, we consider the following two-dimensional SDE
\begin{equation} \label{eq:specific_SDE}
\begin{aligned}
dx_1(t) &=& adt &+& dW_1(t)\\
dx_2(t) &=& -\frac{x_2(t)}{\epsilon} dt &+& \frac{1}{\sqrt{\epsilon}} dW_2(t)
\end{aligned}
\end{equation}
%
where $a$ is an $\mathcal{O}(1)$ constant, as a specific example of \eqref{eq:general_SDE}.
%
$x_1$ is the slow variable, and $x_2$ is a fast noise whose equilibrium measure is bounded and $\mathcal{O}(1)$.
%
Figure~\ref{fig:initial_data} shows data simulated from this SDE colored by time.
%
We would like to recover a parametrization of this data which is one-to-one with the slow variable $x_1$.

\begin{figure}[t]
\centering
\epsfig{width=0.5\textwidth, file=data_init.eps}
%
\caption{Data, simulated from \eqref{eq:specific_SDE} with $a=3$ and $\epsilon = 10^{-3}$, for $3000$ time steps with $dt = 10^{-4}$. The data are colored by time.}
\label{fig:initial_data}
\end{figure}

\subsection{Linear function} \label{subsec:linear_example}

In the first example, our observation function $\mathbf{f}$ will be the identity function,
%
\begin{equation} \label{eq:linear_transform}
\begin{aligned}
\begin{bmatrix}
y_1(t) \\ y_2(t)
\end{bmatrix} &=&
\mathbf{f}(\mathbf{x}(t)) &=&
\begin{bmatrix} x_1(t) \\ x_2(t) \end{bmatrix} \\
\mathbf{g}(\mathbf{y}(t)) &=& \mathbf{f}^{-1} (\mathbf{y}(t)) &=& \begin{bmatrix} y_1(t) \\ y_2(t) \end{bmatrix}
\end{aligned}
\end{equation}
%
where the fast and slow variables remain uncoupled.
%
In this case, there is no error incurred due to the measurement function $\mathbf{f}$ ($E_M = 0$), as the second- and higher-order derivatives of $\mathbf{g}$ are identically 0.

\subsubsection{Importance of using the Mahalanobis distance}

We want to demonstrate the utility of using the Mahalanobis distance compared to the typical Euclidean distance.
%
We compute the diffusion map embedding for the data in Figure~\ref{fig:initial_data},
using both the standard Euclidean distance and the Mahalanobis distance for the computation of the kernel in \eqref{eq:dmaps_kernel}.
%
The data, colored by $\phi_1$ using the two different metrics, are shown in Figure~\ref{fig:NIV_versus_DMAPS}.
%
When using the standard Euclidean distance which does not account for the underlying dynamics, the first diffusion maps recovers the fast variable $x_2$, suggesting the fast modes is the dominant scale purely in terms of data analysis (Figure~\ref{fig:NIV_versus_DMAPS}(a)).
%
In contrast, the slow variable is recovered when using the Mahalanobis distance, as the coloring in Figure~\ref{fig:NIV_versus_DMAPS}(b) (where the data are colored by the first diffusion maps variable) is consistent with the coloring in Figure~\ref{fig:initial_data} (where the data are colored by time).

\begin{figure}[t]
\centering
\begin{subfigure}{0.4\textwidth}
\centering
\epsfig{height=2in, file=data_linear_DMAPS.eps}
\caption{}
\end{subfigure}
\begin{subfigure}{0.4\textwidth}
\centering
\epsfig{height=2in, file=data_linear_NIV.eps}
\caption{}
\end{subfigure}
%
\caption{Comparison of using the Euclidean distance and the Mahalanobis distance in multiscale data mining. (a) The data from Figure~\ref{fig:initial_data}, colored by the first diffusion map coordinate when using the Euclidean distance in the kernel in \eqref{eq:dmaps_kernel}. Note that we do {\em not} recover the slow variable. (b) The data from Figure~\ref{fig:initial_data}, colored by the first diffusion map coordinate when using the Mahalanobis distance in the kernel in \eqref{eq:dmaps_kernel}. The good correspondence between this coordinate and the slow variable is visually obvious.}
\label{fig:NIV_versus_DMAPS}
\end{figure}

\subsubsection{Errors in covariance estimation}

For the example in \eqref{eq:linear_transform}, the analytical covariance is
 \begin{equation} \label{eq:cov_linear_example}
\mathbf{C}(\mathbf{x}(t)) =
\begin{bmatrix}
1 & 0 \\
0 & \frac{1}{\epsilon}
\end{bmatrix}.
\end{equation}
%
From \eqref{eq:cov_error}, we find
%
\begin{equation}
\mathbf{E}_C(\mathbf{x}(t), \delta t) =
\begin{bmatrix}
0 & 0 \\
0 & -\frac{\delta t}{\epsilon^2}
\end{bmatrix}
+ \mathcal{O} (\delta t^{3/2}) .
\end{equation}
%
Therefore, $\| \mathbf{C} \| = \mathcal{O} \left( \frac{1}{\epsilon} \right)$ and $\|\mathbf{E}_C \| = \mathcal{O}\left(\frac{\delta t}{\epsilon^2} \right)$ (provided $\frac{1}{\epsilon^2} \gg \sqrt{\delta t}$; this will be discussed further in Section~\ref{subsec:fastvar}).
%
These terms are shown in Figure~\ref{fig:cov_error}(a) as a function of $\delta t$.
%
We want to choose $\delta t$ in a regime where $\| \mathbf{E}_C \| \ll \| \mathbf{C} \|$ (the yellow shaded region in Figure~\ref{fig:cov_error} indicates where $\| \mathbf{E}_C \| < \| \mathbf{C} \|$), so that the errors in the estimated covariance are small with respect to the covariance.

When we do not analytically know the functions $\mathbf{f}$ or $\mathbf{g}$, we can find such a regime empirically by
estimating the covariance for several values of $\delta t$.
%
This provides an estimate of $\hat{\mathbf{C}} = \mathbf{C} + \mathbf{E}_C$ as a function of $\delta t$.
%
From Figure~\ref{fig:cov_error}(a), we expect a ``knee" in the plot of $\| \hat{\mathbf{C}} \|$ versus $\delta t$ when $\| \mathbf{E}_C \|$ becomes larger than $\| \mathbf{C}\|$.
%
Figure~\ref{fig:cov_error}(b) shows the empirical $\| \hat{\mathbf{C}} \|$ as a function of $\delta t$ for the data in Figure~\ref{fig:initial_data}, and the knee in this curve is consistent with the intersection in Figure~\ref{fig:cov_error}(a).

\begin{figure}[t]
\centering
\begin{subfigure}{0.4\textwidth}
\centering
\epsfig{height=2in, file=C_dt_analytical_linear.eps}
\caption{}
\end{subfigure}
%
\begin{subfigure}{0.4\textwidth}
\centering
\epsfig{height=2in, file=C_dt_linear.eps}
\caption{}
\end{subfigure}
%
\caption{Errors in covariance estimation for linear example. (a) The analytical contributions to the covariance for the example in Section~\ref{subsec:linear_example} as a function of $\delta t$. (b) The average estimated covariance $\| \hat{ \mathbf{C}} \|$ as a function of $\delta t$. The average is computed over $10$ data points and using $50$ sample points to estimate each covariance. The shaded yellow region indicates the range of $\delta t$ over which the errors in the estimated covariance are less than the norm of the covariance. }
\label{fig:cov_error}
\end{figure}

\subsubsection{Recovery of the fast variable} \label{subsec:fastvar}

Note that, for the example in \eqref{eq:linear_transform}, $\mathbf{E}_C$ is a constant diagonal matrix.
%
Therefore, taking $\delta t$ too large will not lead to nonlinear effects or mixing of the fast and slow variables.
%
Rather, changing $\delta t$ will only affect the perceived ratio of the fast and slow timescales.

To see this behavior in our diffusion maps results, we must first discuss the interpretation of the diffusion maps eigenspectrum.
%
The diffusion maps eigenvectors provide embedding coordinates for the data, and the corresponding eigenvalues provide a measure of the importance of each coordinate.
%
However, some eigenvectors can be harmonics of previous eigenvectors; for example, for a data set parameterized by a variable $x$, both $\cos x$ and $\cos 2x$ will appear as diffusion maps eigenvectors (see \cite{ferguson2010systematic} for a more detailed discussion).
%
These harmonics do not capture any new direction within the data set, but do appear as additional eigenvector/eigenvalue pairs.
%
Therefore, for the two-dimensional data considered here, the fast variable will not necessarily appear as the second (non-trivial) eigenvector.
%
As the time scale separation increases, the relative importance of the slow and fast directions will also increase.
%
This implies that the eigenvalue corresponding to the eigenvector which parameterizes the fast direction will decrease, and the number of harmonics of the slow mode which appear before the fast mode will increase.

Figure~\ref{fig:recover_fast} shows results for three different values of $\delta t$ (the corresponding values are indicated by the dashed lines in Figure~\ref{fig:cov_error}).
%
When the time scale of the simulation burst used to estimate the local covariance (indicated by the red clouds in the top row of figures), is sufficiently shorter than that of the equilibration time of the fast variable, the estimated local covariance is accurate and the fast variable is collapsed significantly relative to the slow variable.
%
This means that the fast variable is recovered {\em very} far down in the diffusion maps eigenvectors.
%
The left two columns of Figure~\ref{fig:recover_fast} show that, for this example, when the simulation burst is shorter than the equilibration time, the fast variable is recovered as $\phi_{10}$.
%
However, if the time scale of the burst is {\em longer} than the saturation time of the fast variable, the estimated covariance changes: the variance in the slow direction continues to grow, while the variance in the fast direction is fixed.
%
This means that the apparent time scale separation is smaller, the collapse of the fast variable is less pronounced relative to the slow variable, and the fast variable is recovered in an earlier eigenvector (in our ordering of the spectrum).
%
The right column of Figure~\ref{fig:recover_fast} shows that, when the burst is now longer than the equilibration time, the fast variable appears earlier in the eigenvalue spectrum and is recovered as $\phi_6$.

\begin{figure}[t]
\epsfig{width=0.3\textwidth, file=data_linear_burst1.eps}
\epsfig{width=0.3\textwidth, file=data_linear_burst2.eps}
\epsfig{width=0.3\textwidth, file=data_linear_burst3.eps}

\epsfig{width=0.3\textwidth, file=data_linear_slow1.eps}
\epsfig{width=0.3\textwidth, file=data_linear_slow2.eps}
\epsfig{width=0.3\textwidth, file=data_linear_slow3.eps}

\epsfig{width=0.3\textwidth, file=data_linear_fast1.eps}
\epsfig{width=0.3\textwidth, file=data_linear_fast2.eps}
\epsfig{width=0.3\textwidth, file=data_linear_fast3.eps}

\epsfig{width=0.3\textwidth, file=data_linear_evals1.eps}
\epsfig{width=0.3\textwidth, file=data_linear_evals2.eps}
\epsfig{width=0.3\textwidth, file=data_linear_evals3.eps}

\caption{Relationship between changing $\delta t$ and recovery of the variables. From left to right, the columns correspond to $\delta t = 10^{-6}, 10^{-5}, 10^{-3}$.  (Row~1) Data (gray) and representative burst (red) used to estimate the local covariance. (Row~2) Correlation between first diffusion maps coordinate and the slow variable $x_1$. (Row~3) Correlation between the relevant diffusion maps coordinate and the fast variable $x_2$. Note that for $\delta t = 10^{-6}$ and $\delta t = 10^{-5}$, $x_2$ is correlated with $\phi_{10}$. When $\delta t = 10^{-3}$, $x_2$ is correlated with $\phi_6$. (Row 4) Diffusion maps eigenvalue spectra. The eigenvalues corresponding to the coordinates for the slow and fast modes are indicated by red circles. Note that when $\delta t$ is too large, the apparent time scale separation decreases and the coordinate corresponding to the fast variable appears earlier in the spectrum. }
\label{fig:recover_fast}
\end{figure}

\subsection{Nonlinear observation function} \label{subsec:nonlinear_example}

In the second example, our data will be warped into ``half-moon" shapes via the function
\begin{equation} \label{eq:nonlinear_function}
\begin{aligned}
\begin{bmatrix}
y_1(t) \\ y_2(t)
\end{bmatrix} &=&
\mathbf{f}(\mathbf{x}(t)) &=&
\begin{bmatrix}
x_1(t) + x_2^2(t) \\
x_2(t)
\end{bmatrix}\\
\mathbf{g}(\mathbf{y}(t)) &=& \mathbf{f}^{-1} (\mathbf{y}(t)) &=& \begin{bmatrix} y_1(t) - y_2^2(t) \\ y_2(t) \end{bmatrix} .
\end{aligned}
\end{equation}
%
Figure~\ref{fig:initial_data_nonlinear} shows the data from Figure~\ref{fig:initial_data} transformed by the function $\mathbf{f}$ in \eqref{eq:nonlinear_function} and colored by time.
%
It is important to note that this is a difficult class of problem in practice, as none of the observed
variables are purely fast or slow, and the observed system appears, at first inspection, to possess no separation
of time scales.
%
For this example, the analytical covariance and inverse covariance are
\begin{equation}
\begin{aligned}
\mathbf{C}(\mathbf{x}(t)) =&
\frac{1}{\epsilon}
 \begin{bmatrix}
\epsilon + 4x_2^2(t) & 2x_2(t) \\
2x_2(t) & 1
\end{bmatrix}\\
\mathbf{C}^{\dagger}(\mathbf{x}(t)) =&
\begin{bmatrix}
1 & -2 x_2(t) \\
-2 x_2(t) & \epsilon+ 4 x_2^2(t)
\end{bmatrix} .
\end{aligned}
\end{equation}

The fast and slow variables are now coupled through the function $\mathbf{f}$, and the Euclidean distance is not informative about the fast {\em or} the slow variables.
%
We need to use the Mahalanobis distance to obtain a parametrization that is consistent with the underlying fast-slow dynamics.

\begin{figure}[t]
\centering
\epsfig{width=0.5\textwidth, file=data_init_nonlinear.eps}
%
\caption{The data from Figure~\ref{fig:initial_data}, transformed by $\mathbf{f}$ in \eqref{eq:nonlinear_function}}
\label{fig:initial_data_nonlinear}
\end{figure}

\subsubsection{Errors in Mahalanobis distance}

We can bound the Mahalanobis distance by the eigenvalues of $\mathbf{C}^{\dagger}$,
\begin{equation}
\lambda_{C^{\dagger},1} \| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|_2^2
\le
\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|^2_M
\le
\lambda_{C^{\dagger},2} \| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|_2^2
\end{equation}
where $\lambda_{C^{\dagger},1} \le \lambda_{C^{\dagger},2}$ are the two eigenvalues of $\mathbf{C}^{\dagger}$.
%
Therefore, for the example in \eqref{eq:nonlinear_function}, we have
\begin{equation}
E_M(\mathbf{y}(t_1), \mathbf{y}(t_2)) = - (y_2(t_2) - y_2(t_1))^4.
\end{equation}

Figure~\ref{fig:cov_error_nonlinear}(a) shows $\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|^2_M$ and $| E_M |$ as a function of $\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|_2$.
%
The Mahalanobis distance is an accurate approximation to the true intrinsic distance $\| \mathbf{z}(t_2) - \mathbf{z}(t_1) \|_2$ when $|E_M| \ll \| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|^2_M$ (the shaded yellow region in the plot indicates where $|E_M| < \| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|^2_M$).
%
We want to choose $\sigma_{kernel}^2$ in a regime where $|E_M(\mathbf{y}(t_1), \mathbf{y}(t_2))| \ll \| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|^2_M$, so that the distances we utilize in the diffusion maps calculation are accurate.
%
We can find such a regime empirically by plotting $\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|^2_M$ as a function of $\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|_2$, and assessing when the relationship deviates from quadratic.
%
This is shown in Figure~\ref{fig:cov_error_nonlinear}(b), and the deviation from quadratic behavior is consistent with the intersection of the analytical expressions plotted in Figure~\ref{fig:cov_error_nonlinear}(a).

Figures~\ref{fig:colored_data_nonlinear_cases}(a) and (b) show the data from Figure~\ref{fig:initial_data_nonlinear}, colored by $\phi_1$ for two different values of $\sigma_{kernel}$.
%
The corresponding values of $\sigma_{kernel}^2$ are indicated by the dashed lines.
%
When $\sigma_{kernel}^2$ corresponds to a region where $ |E_M | \ll \|\mathbf{y}(t_2) - \mathbf{y}(t_1) \|_M^2$, $\phi_1$ is well correlated with the slow variable.
%
However, when $\sigma_{kernel}^2$ corresponds to a region where $|E_M | \gg \|\mathbf{y}(t_2) - \mathbf{y}(t_1) \|_M^2$, the slow variable is no longer recovered.

\begin{figure}[t]
\centering
\begin{subfigure}{0.4\textwidth}
\centering
\epsfig{height=2in, file=dist_dy_analytical_nonlinear.eps}
\caption{}
\end{subfigure}
%
\begin{subfigure}{0.4\textwidth}
\centering
\epsfig{height=2in, file=dist_dy_nonlinear.eps}
\caption{}
\end{subfigure}

\begin{subfigure}{0.4\textwidth}
\centering
\epsfig{height=2in, file=C_dt_analytical_nonlinear.eps}
\caption{}
\end{subfigure}
%
\begin{subfigure}{0.4\textwidth}
\centering
\epsfig{height=2in, file=C_dt_nonlinear.eps}
\caption{}
\end{subfigure}
%
\caption{Errors in the Mahalanobis distance and the covariance estimation for the nonlinear example in \eqref{eq:nonlinear_function}. (a) The analytical expressions for the contributions to the distance approximation as a function of $\| \mathbf{y}_2 - \mathbf{y}_1\|_2$. (b) The average estimated Mahalanobis distance $\| \mathbf{y}_2 - \mathbf{y}_1\|^2_M$ as a function of the distance $\| \mathbf{y}_2 - \mathbf{y}_1\|_2$. The line $\| \mathbf{y}_2 - \mathbf{y}_1\|^2_M = \| \mathbf{y}_2 - \mathbf{y}_1\|_2$ is shown for reference. The yellow region indicates the range in which $\sigma_{kernel}$ should be chosen. (c) The analytical expressions for the contributions to the covariance as a function of $\delta t$. (d) The average estimated covariance $\| \hat{\mathbf{C}} \|$ as a function of $\delta t$. The yellow region indicates the range of $\delta t$ over which the errors in the estimated covariance are small relative to the norm of the covariance. }
\label{fig:cov_error_nonlinear}
\end{figure}

\subsubsection{Errors in covariance estimation}

From \eqref{eq:cov_error}, we find that, for the example in \eqref{eq:nonlinear_function},
%
\begin{equation}
\small
\begin{aligned}
E_{C,11} (\mathbf{x}(t), \delta t)
=&
\frac{2 \delta t}{\epsilon^2}
- \frac{8 x_2(t)}{\epsilon^2 \delta t} \mathbb{E} \left[ \int_t^{t+\delta t} \left( \int_{s_2}^{t+\delta t} 2 x_2(s_1) ds_1
+  \int_t^{s_2} x_2(s_1) ds_1 \right) ds_2\right]  + \mathcal{O} (\delta t^{3/2}) \\
%%
E_{C, 12} (\mathbf{x}(t), \delta t)
= &
E_{C, 21} (\mathbf{x}(t), \delta t)\\
=&
- \frac{x_2(t) \delta t}{\epsilon^2}
- \frac{2}{\epsilon^2 \delta t} \mathbb{E} \left[ \int_t^{t+\delta t} \left( \int_{s_2}^{t + \delta t} 2 x_2(s_1) ds_1 + \int_t^{s_2} x_2(s_1) ds_1 \right) ds_2 \right] + \mathcal{O} (\delta t^{3/2})\\
%%
E_{C, 22} (\mathbf{x}(t), \delta t)
=&
-\frac{\delta t}{\epsilon^2} + \mathcal{O} (\delta t^{3/2})
\end{aligned}
\end{equation}
%
The error in the covariance is $\mathcal{O} \left( \frac{\delta t}{\epsilon^2} \right)$.
%
As expected, the error grows with increasing $\delta t$.
%
We can also see the explicit dependence of the covariance error on the time scale separation $\epsilon$; larger time scale separation results in a larger covariance error, as a more refined simulation burst is required to estimate the covariance of the fast directions.
%
$\|\mathbf{C} \|$ and $\| \mathbf{E}_C\|  $ are plotted as a function of $\delta t$ in Figure~\ref{fig:cov_error_nonlinear}(c); the shaded yellow portion denotes the region where $\| \mathbf{E}_C \| < \| \mathbf{C} \|$.
%
As in the previous example, we can empirically find where $\| \mathbf{E}_C \| \ll \| \mathbf{C} \|$  by plotting $\| \hat{\mathbf{C}} \|$ as a function of $\delta t $ and looking for a knee in the plot.
%
These results are shown in Figure~\ref{fig:cov_error_nonlinear}(d).

Figures~\ref{fig:colored_data_nonlinear_cases}(a) and (c) show the data from Figure~\ref{fig:initial_data_nonlinear}, colored by $\phi_1$ for two different values of $\delta t$.
%
The corresponding values of $\delta t$ are indicated by the dashed lines in Figure~\ref{fig:cov_error_nonlinear}(c)~and~(d).
%
When $\delta t$ corresponds to a region where $\|\mathbf{E}_C \| \ll \| \mathbf{C} \|$, the slow variable is recovered by the first diffusion maps coordinate.
%
However, when $\delta t$ corresponds to a region where $\|\mathbf{E}_C \| \gg \| \mathbf{C} \|$, the slow variable is no longer recovered.

\def \figheight {1.6in}

\begin{figure}[t]

\begin{subfigure}{\textwidth}
\epsfig{height=\figheight, file=data_nonlinear_NIV_dt1_kernel1.eps}
\hfill
\epsfig{height=\figheight, file=C_dt_nonlinear_dt1_kernel1.eps}
\hfill
\epsfig{height=\figheight, file=dist_dy_nonlinear_dt1_kernel1.eps}
\end{subfigure}

\begin{subfigure}{\textwidth}
\epsfig{height=\figheight, file=data_nonlinear_NIV_dt1_kernel2.eps}
\hfill
\epsfig{height=\figheight, file=C_dt_nonlinear_dt1_kernel2.eps}
\hfill
\epsfig{height=\figheight, file=dist_dy_nonlinear_dt1_kernel2.eps}
\end{subfigure}

\begin{subfigure}{\textwidth}
\epsfig{height=\figheight, file=data_nonlinear_NIV_dt2_kernel1.eps}
\hfill
\epsfig{height=\figheight, file=C_dt_nonlinear_dt2_kernel1.eps}
\hfill
\epsfig{height=\figheight, file=dist_dy_nonlinear_dt2_kernel1.eps}
\end{subfigure}

\caption{Data from Figure~\ref{fig:initial_data_nonlinear}, colored by the first diffusion maps variable $\phi_1$ using the Mahalanobis distance for three different parameter settings. The relevant values of $\delta t$ and $\sigma_{kernel}$ are indicated by the red dashed lines on the corresponding plots.  (Row~1) $\delta t = 10^{-7}$ and $\sigma_{kernel}^2 = 10^{-2}$. Note that the parametrization is one-to-one with the slow variable.  (Row~2) $\delta t = 10^{-7}$ and $\sigma_{kernel}^2 = 10^{1}$. We do not recover the slow variable because $\sigma_{kernel}$ is too large. (Row~3)  $\delta t = 10^{-3}$ and $\sigma_{kernel}^2 = 10^{-2}$. We do not recover the slow variable because $\delta t$ is too large.  }
\label{fig:colored_data_nonlinear_cases}
\end{figure}

\section{Conclusions}

We have presented a methodology to compute a parametrization of a data set which respects the slow variables in the underlying dynamical system.
%
The approach utilizes diffusion maps, a kernel-based manifold learning technique, with the Mahalanobis distance as the metric.
%
We showed that the Mahalanobis distance collapses the fast directions within a data set, allowing for successful recovery of the slow variables.
%
Furthermore, we showed how to estimate the covariances (required for the Mahalanobis distance) directly from data.
%
A key point in our approach is that the embedding coordinates we compute are not only insensitive to the fast variables, but are also invariant to nonlinear observation functions.
%
Therefore, the approach can be used for data fusion: data collected from the same system via different measurement functions can be combined and merged into a single coordinate system.

In the examples presented, the initial data came from a single trajectory of a dynamical system, and the local covariance at each point in the trajectory was estimated using brief simulation bursts.
%
However, the initial data need not be collected from a single trajectory, and other sampling schemes could be employed.
%
Brief time series are required to estimate the local covariances, but given a simulator, one could reinitialize brief simulation bursts which are sufficiently short and refined from each sample point.

In our examples, we controlled the time scale of sampling and could therefore set the time scale over which to estimate the covariance and the simulation time step arbitrarily small.
%
However, in some settings, such as previously collected historical data, it is not uncommon to have a fixed sampling rate and be unable to reinitialize simulations.
%
In such cases, it is possible that we cannot find an appropriate kernel scale given the fixed $\delta t$ such that we can accurately recover the slow variables.
%
For these cases, the data cannot be processed as given,
and it is necessary to construct intermediate observers,
such as histograms, Fourier coefficients, or scattering transform coefficients \cite{mallat2012group, talmon2014intrinsic, talmon2014manifold}.
%
Such intermediates are more complex statistical functions than simple averages and can capture additional structure within the data.
%
They also reduce the effects of noise and permit a larger time step.
%
However, constructing such intermediates often requires additional {\em a priori} knowledge about the system dynamics and noise structure.

Clearly, in our analysis, we have ignored the finite sampling effects in our estimation.
%
In reality, both the number of samples used to estimate the covariances, as well as the density of sampled points on the manifold, affect the recovered parametrization and provide additional constraints on $\delta t$ and $\sigma_{kernel}$.
%
Future work involves extending our analysis to the finite sample case, and providing guidelines for the amount of data required to apply our methodology.

The methods presented here provide a bridge between traditional data mining and multiple time scale dynamical systems.
%
With this interface established, one can now consider using such data-driven methodologies to extract reduced models (either explicitly, or implicitly via an equation-free framework \cite{erban2006gene, kevrekidis2004equation, kevrekidis2003equation,  kevrekidis2009equation}) which also respect the underlying slow dynamics and geometry of the data.
%
Such reduced models hold the promise of accelerated analysis and reduced simulation of dynamical systems whose effective dynamics are obscure upon simple inspection.

\bibliographystyle{siam}
\bibliography{../../../../references/references}

\end{document}

### 2015-05-28 reducible_SDE/paper/SIADS_paper/SIADS_revisions.tex
added some edits to SIADS revisions

--- 2015-05-27 reducible_SDE/paper/SIADS_paper/SIADS_manuscript.tex
@@
 %
-However, in general, this reduction introduces memory terms.
+However, in general, this reduction introduces memory terms \cite{roberts2008normal}.
 %
@@
 Here, we will study the special case of evolution equations which contain an inherent time scale separation;
-in this case, it is possible, in principle, to obtain a reduced system of differential equations in only the slow variables {\em without memory terms}.
+in this case, it is possible (for weak models or for models with linear noise) to obtain a reduced system of differential equations in only the slow variables {\em without memory terms} \cite{roberts2008normal}.
 %
@@
 
-Given an explicit system of ordinary differential equations,
+Given a system of ordinary differential equations with an explicit functional form,
 one can make numerical approximations, such as the quasi-steady state approximation \cite{segel1989quasi} or
-the partial equilibrium approximation \cite{gallagher1986combined}, to reduce the system dimensionality without introducing memory terms.
+the partial equilibrium approximation \cite{gallagher1986combined}, to reduce the system dimensionality.
 %
@@
 
-We will present detailed analysis for our method, and provide conditions under which it will successfully recover the slow variables.
-%
-Furthermore, based on this analysis, we will present data-driven protocols to tune the parameters of the method appropriately.
-%
-Our presentation and discussion will address two-time-scale stochastic systems; however, we claim that
-our framework and analysis readily extends to systems with multiple time scale separations.
+We present detailed analysis for our method, and provide conditions under which it will successfully recover the slow variables.
+%
+Furthermore, based on this analysis, we present data-driven protocols to tune the parameters of the method appropriately.
+%
+Our presentation and discussion addresses two-time-scale stochastic systems; however, we claim that our framework and analysis readily extends to systems with multiple time scale separations.
 %
@@
 
-Consider the following two-time-scale system of stochastic differential equations (SDEs),
+\begin{figure}[t]
+\begin{subfigure}{0.3\textwidth}
+\includegraphics[width=\textwidth]{data_init_nonlinear}
+\caption{}
+\end{subfigure}
+%
+\begin{subfigure}{0.3\textwidth}
+\includegraphics[width=\textwidth]{data_init}
+\caption{}
+\end{subfigure}
+%
+\begin{subfigure}{0.3\textwidth}
+\includegraphics[width=\textwidth]{data_rescaled}
+\caption{}
+\end{subfigure}
+\caption{}
+\label{fig:data_rescaling}
+\end{figure}
+
+In this work, we assume that we observe data $\mathbf{y}(t_1), \dots, \mathbf{y}(t_N)$, where $t_i = t_0 + i \Delta t$, collected from a single trajectory of a stochastic process $\mathbf{y}(t)$. 
+%
+We further assume that there exists a nonlinear coordinate transform $\mathbf{f}$ such that $\mathbf{y}(t) = \mathbf{f} \left( \mathbf{x} (t) \right)$, where $\mathbf{x}$ comes from the following autonomous two-time-scale system of stochastic differential equations (SDEs),
 \begin{equation} \label{eq:general_SDE}
@@
 %
-In the simple case of a linear drift function, i.e., when $a_i(\mathbf{x}(t)) = \mu _i x_i$ with $\mu_i < 0$, the probability density function of $x_i$ approaches a Gaussian with
-(finite) variance $\mu_i$.
-%
-The time constant of the approach of the variance to equilibrium is $-1/\mu_i$ for $i=1,\ldots,m$ and $-\epsilon/\mu_i$ for $i=m+1,\ldots,n$ \cite{lelievre2013optimal}.
-%
-Thus, the last $n-m$ variables of \eqref{eq:general_SDE} rapidly approach a local equilibrium measure and exhibit fast dynamics, while the first $m$ variables exhibit slow dynamics.
-%
-A short burst of simulation will yield a cloud of points which is broadly distributed in the fast directions but narrowly distributed in the slow ones.
-%
-With the appropriate conditions on $a_i(\mathbf{x})$, the same can be said for more general drift
-functions, where $\mu_i$ are the eigenvalues of the Jacobian of $\mathbf{a}(\mathbf{x}) = \begin{bmatrix} a_1(\mathbf{x}) & \cdots & a_n(\mathbf{x}) \end{bmatrix}^T$ \cite{villani2009hypocoercivity}.
-%
-Therefore, \eqref{eq:general_SDE} defines an $n$-dimensional stochastic system with $m$ slow
-variables and $n-m$ fast variables, and $\epsilon$ defines the time scale separation.
-%
-The ratio of the powers of $\epsilon$ in the drift and diffusion terms in \eqref{eq:general_SDE} is essential,
-as we require the square of the diffusivity to be of the same order as the drift as $\epsilon \rightarrow 0$ \cite{berglund2003geometric}.
-%
-If the diffusivity is larger, then, as $\epsilon \rightarrow 0$, the equilibrium measure will be
-unbounded.
-%
-Conversely, if the diffusivity is smaller, the equilibrium measure will go to $0$ as $\epsilon \rightarrow 0$.
-
-Assuming the sample average of $a_i(\mathbf{x})$ converges to a distribution which is only a function of the slow variables, then by the averaging principle \cite{freidlin2012random}, we can write a reduced SDE in {\em only} the slow variables $x_1, \dots, x_m$.
-%
-The aim of our work is to show how we can detect such slow variables {\em automatically} from data, in order to help inform modeling efforts and aid in the writing of such reduced stochastic models.
-%
-In general, we are not given the variables $\mathbf{x}(t)$ from the original SDE system, but instead, we are given some {\em observations}
- in the form $\mathbf{y}(t) = \mathbf{f} (\mathbf{x}(t))$.
-%
-We assume that $\mathbf{f}: \mathbb{R}^n \mapsto \mathbb{R}^d$, $n \le d$, is a deterministic (possibly nonlinear) function whose image is an $n$-dimensional manifold $\mathcal{M}$ in $\mathbb{R}^d$.
+Here, we restrict ourselves to the case where $\mathbf{x}(t)$ comes from an SDE with additive noise; the case of multiplicative noise is the subject of future research.
+%
+The parameter $\epsilon$ induces a separation of time scales, such that $\mathcal{O}(1)$ changes in $x_1, \dots, x_m$ occur over times scales $\mathcal{O}(1)$, and $\mathcal{O}(1)$ changes in $x_{m+1}, \dots, x_n$ occur over time scales $\mathcal{O}(\epsilon)$.
+
+Our goal is extract a parameterization of the observed data $\mathbf{y}(t)$ which is consistent with the underlying {\em slow} variables $x_1, \dots, x_m$. 
+%
+We opt to use a data-driven technique (diffusion maps) which does not directly utilize any information about the underlying dynamics.
+%
+The main component in the diffusion maps algorithm is defining the relevant pairwise distance metric.
+%
+Therefore, to analyze dynamic data using diffusion maps, we first need to transform our data to (1) factor out the effects of the (perhaps nonlinear) function $\mathbf{f}$, and (2) rescale the data so that the length scale fluctuations are commensurate with the time scale fluctuations (see Figure~\ref{fig:data_rescaling}).
+%
+In this work, we propose a distance metric which incorporates these two transformations in a single step.
+
+Such a rescaling transformation of the SDEs in \eqref{eq:general_SDE} is guaranteed to exist for Khasminskii systems; however, we suspect that our method will be applicable for a broader class of multiscale SDEs.
+%
+We are interested in the more general case where we observe $\mathbf{y}(t) = \mathbf{f}( \mathbf{x}(t))$. 
+%
+In order to factor out the effects of $\mathbf{f}$, we assume that $\mathbf{g} = \mathbf{f} ^{-1}$ is well-defined on the image of $\mathbf{f}$, and both $\mathbf{f}$ and $\mathbf{g}$ be continuously differentiable to fourth order.
+%
+For the examples presented here, we have $\mathbf{f}: \mathbb{R}^d \mapsto \mathbb{R}^d$; however, we note that in general, the function $\mathbf{f}: \mathbb{R}^n \mapsto \mathbb{R}^d$, $n \le d$, can be a deterministic (possibly nonlinear) function whose image is an $n$-dimensional manifold $\mathcal{M}$ in $\mathbb{R}^d$.
+%
+We further note that the data must be sampled such that $N \Delta t \gg \epsilon$; this condition ensures that fast variables have equilbrated, and the slow variables have evolved appreciably.
+
+%In the simple case of a linear drift function, i.e., when $a_i(\mathbf{x}(t)) = \mu _i x_i$ with $\mu_i < 0$, the probability density function of $x_i$ approaches a Gaussian with (finite) variance $\mu_i$.
+%
+%The time constant of the approach of the variance to equilibrium is $-1/\mu_i$ for $i=1,\ldots,m$ and $-\epsilon/\mu_i$ for $i=m+1,\ldots,n$ \cite{lelievre2013optimal}.
+%
+%Thus, the last $n-m$ variables of \eqref{eq:general_SDE} rapidly approach a local equilibrium measure and exhibit fast dynamics, while the first $m$ variables exhibit slow dynamics.
+%
+%Because of the separation of time scales in the diffusion terms, a short burst of simulation (with $\delta t = o \left( \epsilon \right)$) will yield a cloud of points which is broadly distributed ($\mathcal{O} \left( 1/\sqrt{\epsilon} \right)$) in the last $n-m$ fast directions but narrowly distributed ($\mathcal{O} \left( 1 \right)$) in the first $m$ slow ones.
+%
+%With the appropriate conditions on $a_i(\mathbf{x})$, the same can be said for more general drift functions, where $\mu_i$ are the eigenvalues of the Jacobian of $\mathbf{a}(\mathbf{x}) = \begin{bmatrix} a_1(\mathbf{x}) & \cdots & a_n(\mathbf{x}) \end{bmatrix}^T$ \cite{villani2009hypocoercivity}.
+%
+%Therefore, \eqref{eq:general_SDE} defines an $n$-dimensional stochastic system with $m$ slow variables and $n-m$ fast variables, and $\epsilon$ defines the time scale separation.
+%
+%The ratio of the powers of $\epsilon$ in the drift and diffusion terms in \eqref{eq:general_SDE} is essential, as we require the square of the diffusivity to be of the same order as the drift as $\epsilon \rightarrow 0$ \cite{berglund2003geometric}.
+%
+%If the diffusivity is larger, then, as $\epsilon \rightarrow 0$, the equilibrium measure will be unbounded.
+%
+%Conversely, if the diffusivity is smaller, the equilibrium measure will go to $0$ as $\epsilon \rightarrow 0$.
+
+%Assuming the ensemble average of $a_i(\mathbf{x})$, conditioned on the slow variables $x_1, \dots, x_m$, converges to a distribution which is {\em only} a function of the slow variables, then by the averaging principle \cite{freidlin2012random}, we can write a reduced SDE in {\em only} the slow variables $x_1, \dots, x_m$.
+%
+%We assume that our data is collected over a time scale much larger than $\epsilon$, such that the $n-m$ fast variables have saturated and we have observed appreciable evolution in the slow variables.
+%
+%The aim of our work is to show how we can detect such slow variables {\em automatically} from data, in order to help inform modeling efforts and aid in the writing of such reduced stochastic models.
+%
+%In general, we are not given the variables $\mathbf{x}(t)$ from the original SDE system, but instead, we are given some {\em observations} in the form $\mathbf{y}(t) = \mathbf{f} (\mathbf{x}(t))$.
+%
+%We assume that $\mathbf{f}: \mathbb{R}^n \mapsto \mathbb{R}^d$, $n \le d$, is a deterministic (possibly nonlinear) function whose image is an $n$-dimensional manifold $\mathcal{M}$ in $\mathbb{R}^d$.
 %
@@
 %
-For our analysis, we require $\mathbf{g} = \mathbf{f} ^{-1}$ to be well-defined on $\mathcal{M}$, and both $\mathbf{f}$ and $\mathbf{g}$ to be continuously differentiable to fourth order.
-%
-Given data $\mathbf{y}(t_1),\ldots,\mathbf{y}(t_N)$ on $\mathcal{M}$ we would like to recover a parametrization of the data that is one-to-one with the
-slow variables $x_1, \dots, x_m$.
+%For our analysis, we require $\mathbf{g} = \mathbf{f} ^{-1}$ to be well-defined on $\mathcal{M}$, and both $\mathbf{f}$ and $\mathbf{g}$ to be continuously differentiable to fourth order.
+%
+%Given data $\mathbf{y}(t_1),\ldots,\mathbf{y}(t_N)$ on $\mathcal{M}$ we would like to recover a parametrization of the data that is one-to-one with the slow variables $x_1, \dots, x_m$.
 
@@
 \end{equation}
-and the Mahalanobis distance between samples is
+and a short burst of simulation (with $\delta t = o \left( \epsilon \right)$) yields a cloud of points which is broadly distributed ($\mathcal{O} \left( 1/\sqrt{\epsilon} \right)$) in the last $n-m$ fast directions but narrowly distributed ($\mathcal{O} \left( 1 \right)$) in the first $m$ slow ones.
+%
+The Mahalanobis distance between samples is
 \begin{equation} \label{eq:rescale_x_dist}
@@
 \end{equation}
-Note that in \eqref{eq:rescale_x_dist}, the fast variables are collapsed and become $\mathcal{O}(\sqrt{\epsilon})$ small,
-and so this metric is implicitly insensitive to variations in the fast variables.
+Note that the metric in \eqref{eq:rescale_x_dist} is insensitive to variations in the fast variables.
 %
@@
 %
-This rescaling transforms our problem from one of detecting the slow variables within dynamic data to one of traditional data mining.
+%This rescaling transforms our problem from one of detecting the slow variables within dynamic data to one of traditional data mining.
 %
@@
 %
-It is important to note that, in practice, we {\em never construct} $\mathbf{z}(t)$ {\em explicitly}.
+%It is important to note that, in practice, we {\em never construct} $\mathbf{z}(t)$ {\em explicitly}.
 %
@@
 \begin{equation} \label{eq:mahalanobis}
+%\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|^2_M = \| \mathbf{z}(t_2) - \mathbf{z}(t_1) \|^2_2 + \mathcal{O}(\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|^4_2).
 \| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|^2_M = \| \mathbf{z}(t_2) - \mathbf{z}(t_1) \|^2_2 + \mathcal{O}(\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|^4_2).
@@
 %
-This approximation is accurate when $\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|$ is small.
-%
-Because we will integrate these distances into a manifold learning algorithm which only considers local distances, we can recover a parametrization of the data which is consistent with the underlying system variables $\mathbf{x}(t)$, even when the data are obscured by a function $\mathbf{f}$.
-%
-In Section~\ref{sec:analysis}, we will show how we can approximate this distance directly from data $\mathbf{y}(t)$.
+Clearly, this approximation is accurate when $\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|_2^4 \ll \| \mathbf{z}(t_2) - \mathbf{z}(t_1) \|^2_2$; in Section~\ref{subsec:obsfn}, we provide a more detailed analysis of the errors incurred in this approximation.
+%
+Because we integrate these distances into a manifold learning algorithm which only considers local distances, we recover a parametrization of the data which is consistent with the underlying system variables $\mathbf{x}(t)$, even when the data are obscured by a function $\mathbf{f}$.
+%
+Section~\ref{subsec:cov_est} shows how we approximate this distance directly from data $\mathbf{y}(t)$.
 %
@@
 
-Following \cite{singer2008non}, we will show via Taylor expansion that the Mahalanobis distance between the observations \eqref{eq:mahalanobis_distance} approximates the Euclidean
-distance {\em in the rescaled variables} for general nonlinear observation functions $\mathbf{f}$ (provided $\mathbf{f}$ is bilipschitz and both $\mathbf{f}$ and $\mathbf{f}^{-1}$ are differentiable to fourth order).
+Following \cite{singer2008non}, we show via Taylor expansion that the Mahalanobis distance between the observations \eqref{eq:mahalanobis_distance} approximates the Euclidean distance {\em in the rescaled variables} for general nonlinear observation functions $\mathbf{f}$ (provided $\mathbf{f}$ is bilipschitz and both $\mathbf{f}$ and $\mathbf{f}^{-1}$ are differentiable to fourth order).
 %Our approximation of the pairwise distances relies on the Taylor expansion of the measurement function $\mathbf{f}$.
@@
 %
-We can estimate the covariance $\mathbf{C}(\mathbf{y}(t_0))$ empirically from a set of values $\mathbf{y}(t_1), \dots, \mathbf{y}(t_q)$ drawn from the local distribution at $\mathbf{y}(t_0)$.
-%
-One way to obtain such a set of points is to run $q$ simulations for a short time, $\delta t$, each starting from $\mathbf{y}(t_0)$.
-%
-Alternatively, we can consider a single time series of length $q \delta t$ starting from $\mathbf{y}(t_0)$, and then estimate the covariance
+We estimate the covariance $\mathbf{C}(\mathbf{y}(t_0))$ empirically from a set of values $\mathbf{y}(t_1), \dots, \mathbf{y}(t_q)$ drawn from the local distribution at $\mathbf{y}(t_0)$, by running $q$ simulations for a short time $\delta t = o (\epsilon) $, each starting from $\mathbf{y}(t_0)$; we refer to this as a {\em simulation burst}.
+%
+Alternatively, we can consider a single time series of length $q \delta t = o (\epsilon) $ starting from $\mathbf{y}(t_0)$, and then estimate the covariance
 from the increments $\Delta \mathbf{y}(t_i) = \mathbf{y}(t_i) -\mathbf{y}(t_{i-1})$.
@@
 %
-One source of error is approximating the function $\mathbf{f}$ locally as a linear function by truncating the Taylor expansion of $\mathbf{f}$ at first order.
-%
-An additional source of error arises from disregarding the drift in the stochastic process, and assuming that samples are drawn from a Gaussian distribution.
-%
-The third source comes from finite sampling effects.
-%
-In this work, we will address and discuss the first two sources of error (the finite sampling effects are the subject of future research).
-%
-We can control the effects of the errors due to truncation of the Taylor expansion by adjusting $\sigma_{kernel}$; the higher-order terms in this expansion will be small for points which are close, such that adjusting $\sigma_{kernel}$ will allow us to only consider distances which are
-sufficiently accurate in our overall computation scheme.
-%
-Furthermore, we can control the errors incurred by disregarding the drift by adjusting the time scale of our simulation bursts $\delta t$.
+One source of error, discussed in Section~\ref{subsec:obsfn}, is approximating the function $\mathbf{f}$ locally as a linear function by truncating the Taylor expansion of $\mathbf{f}$ at first order.
+%
+An additional source of error, discussed in Section~\ref{subsec:cov_est}, arises from disregarding the drift in the stochastic process, and assuming that samples are drawn from a Gaussian distribution.
+%
+The third source comes from finite sampling effects, which is the subject of future research.
+%
+In this work, we address and discuss the first two sources of error (the finite sampling effects are the subject of future research).
+%
+We control the effects of the errors due to truncation of the Taylor expansion by adjusting $\sigma_{kernel}$; the higher-order terms in this expansion will be small for points which are close, such that adjusting $\sigma_{kernel}$ will allow us to only consider distances which are sufficiently accurate in our overall computation scheme.
+%
+Furthermore, we control the errors incurred by disregarding the drift by adjusting the time scale of our simulation bursts $\delta t$.
 %
@@
 %
-We will present both analytical results for the error bounds, as well as an empirical methodology to set the parameters $\sigma_{kernel}$ and $\delta t$ for our method to accurately recover the slow variable(s).
-
-\subsection{Error due to the observation function $\mathbf{f}$}
+We present both analytical results for the error bounds, as well as an empirical methodology to set the parameters $\sigma_{kernel}$ and $\delta t$ for our method to accurately recover the slow variable(s).
+
+\subsection{Error due to the observation function $\mathbf{f}$} \label{subsec:obsfn}
 
@@
 %
-In \cite{singer2008non}, it was shown that the error incurred by using the Mahalanobis distance to approximate the $L_2$-distance between points $\mathbf{z}(t)$ is $\mathcal{O} (\|\mathbf{y}_1 - \mathbf{y}_2 \|_2^4 )$ (see the Supplementary Materials for details).
+Singer and Coifman \cite{singer2008non} showed that the error incurred by using the Mahalanobis distance to approximate the $L_2$-distance between points $\mathbf{z}(t)$ is $\mathcal{O} (\|\mathbf{y}_1 - \mathbf{y}_2 \|_2^4 )$ (see the Supplementary Materials for details).
 %
@@
 %
-We would like to note that this error does not depend on the dynamics of the underlying stochastic process (as we assume the covariances at each point on the manifold are known), but is only a function of the measurement function $\mathbf{f}$.
+We would like to note that this error does not depend on the dynamics of the underlying stochastic process (as we assume the covariances at each point on the manifold are known), but is only a function of the measurement function $\mathbf{f}$ and the distance between samples $\| \mathbf{y}(t_2) - \mathbf{y}(t_1) \|_2$.
 %
@@
 %
-We will use simulation bursts to locally explore the dynamics on the manifold of observations in order to estimate the covariance at a point $\mathbf{y}(t)$ from data \cite{talmon2014manifold, talmon2014intrinsic}.
-%
-We write the elements of the estimated covariance $\hat{\mathbf{C}}(\mathbf{y}(t), \delta t)$ as
+We use simulation bursts to locally explore the dynamics on the manifold of observations in order to estimate the covariance at a point $\mathbf{y}(t)$ from data \cite{talmon2014manifold, talmon2014intrinsic}.
+%
+We write the elements of the estimated covariance matrix $\hat{\mathbf{C}}(\mathbf{y}(t), \delta t)$ as
 \begin{equation}\label{eq:estimated_cov_expected_value}
@@
 %
-where $\delta t > 0$ is the length of the simulation burst.
+where $\delta t = o (\epsilon)$ is the length of the simulation burst.
 
@@
 %
-By It\^{o}-Taylor expansion of $\mathbf{f}$ and $\mathbf{x}(t)$ \cite{kloeden1992numerical},
+By It\^{o}-Taylor expansion of $\mathbf{f}$ and $\mathbf{x}(t)$ \cite{kloeden1992numerical}, the entries in the error matrix are given by
 %
@@
 \begin{aligned}
-dx_1(t) &=& adt &+& dW_1(t)\\
-dx_2(t) &=& -\frac{x_2(t)}{\epsilon} dt &+& \frac{1}{\sqrt{\epsilon}} dW_2(t)
-\end{aligned}
-\end{equation}
-%
-where $a$ is an $\mathcal{O}(1)$ constant, as a specific example of \eqref{eq:general_SDE}.
+dx_1(t) &=& 3dt &+& dW_1(t)\\
+dx_2(t) &=& -\frac{x_2(t)}{\epsilon} dt &+& \frac{1}{\sqrt{\epsilon}} dW_2(t),
+\end{aligned}
+\end{equation}
+%
+with $\epsilon=10^{-3}$, as a specific example of \eqref{eq:general_SDE}.
 %
@@
 %
-\caption{Data, simulated from \eqref{eq:specific_SDE} with $a=3$ and $\epsilon = 10^{-3}$, for $3000$ time steps with $dt = 10^{-4}$. The data are colored by time.}
+\caption{Data, simulated from \eqref{eq:specific_SDE} ($\epsilon = 10^{-3}$) for $3000$ time steps with $dt = 10^{-4}$. The data are colored by time.}
 \label{fig:initial_data}
@@
 
-For the example in \eqref{eq:linear_transform}, the analytical covariance is
+For the example in \eqref{eq:linear_transform}, the analytical covariance matrix is
  \begin{equation} \label{eq:cov_linear_example}
@@
 %
-From \eqref{eq:cov_error}, we find
+From \eqref{eq:cov_error}, we find the error matrix
 %
@@
 %
-Therefore, $\| \mathbf{C} \| = \mathcal{O} \left( \frac{1}{\epsilon} \right)$ and $\|\mathbf{E}_C \| = \mathcal{O}\left(\frac{\delta t}{\epsilon^2} \right)$ (provided $\frac{1}{\epsilon^2} \gg \sqrt{\delta t}$; this will be discussed further in Section~\ref{subsec:fastvar}).
+Therefore, $\| \mathbf{C} \| = \mathcal{O} \left( \frac{1}{\epsilon} \right)$ and $\|\mathbf{E}_C \| = \mathcal{O}\left(\frac{\delta t}{\epsilon^2} \right)$.
+%(provided $\frac{1}{\epsilon^2} \gg \sqrt{\delta t}$; this will be discussed further in Section~\ref{subsec:fastvar}).
 %
@@
 %
-We want to choose $\delta t$ in a regime where $\| \mathbf{E}_C \| \ll \| \mathbf{C} \|$ (the yellow shaded region in Figure~\ref{fig:cov_error} indicates where $\| \mathbf{E}_C \| < \| \mathbf{C} \|$), so that the errors in the estimated covariance are small with respect to the covariance.
-
-When we do not analytically know the functions $\mathbf{f}$ or $\mathbf{g}$, we can find such a regime empirically by
+We want to choose $\delta t = o(\epsilon)$, so that $\| \mathbf{E}_C \| \ll \| \mathbf{C} \|$ (the yellow shaded region in Figure~\ref{fig:cov_error} indicates where $\| \mathbf{E}_C \| < \| \mathbf{C} \|$), so that the errors in the estimated covariance are small with respect to the covariance.
+
+When we do not analytically know the functions $\mathbf{f}$ or $\mathbf{g}$ or the value of $\epsilon$, we can find such a regime empirically by
 estimating the covariance for several values of $\delta t$.
@@
 \begin{figure}[t]
-\epsfig{width=0.3\textwidth, file=data_linear_burst1.eps}
-\epsfig{width=0.3\textwidth, file=data_linear_burst2.eps}
-\epsfig{width=0.3\textwidth, file=data_linear_burst3.eps}
-
-\epsfig{width=0.3\textwidth, file=data_linear_slow1.eps}
-\epsfig{width=0.3\textwidth, file=data_linear_slow2.eps}
-\epsfig{width=0.3\textwidth, file=data_linear_slow3.eps}
-
-\epsfig{width=0.3\textwidth, file=data_linear_fast1.eps}
-\epsfig{width=0.3\textwidth, file=data_linear_fast2.eps}
-\epsfig{width=0.3\textwidth, file=data_linear_fast3.eps}
-
-\epsfig{width=0.3\textwidth, file=data_linear_evals1.eps}
-\epsfig{width=0.3\textwidth, file=data_linear_evals2.eps}
-\epsfig{width=0.3\textwidth, file=data_linear_evals3.eps}
+\centering
+\epsfig{width=0.2\textwidth, file=data_linear_burst1.eps}
+\epsfig{width=0.2\textwidth, file=data_linear_burst2.eps}
+\epsfig{width=0.2\textwidth, file=data_linear_burst3.eps}
+
+\epsfig{width=0.2\textwidth, file=data_linear_slow1.eps}
+\epsfig{width=0.2\textwidth, file=data_linear_slow2.eps}
+\epsfig{width=0.2\textwidth, file=data_linear_slow3.eps}
+
+\epsfig{width=0.2\textwidth, file=data_linear_fast1.eps}
+\epsfig{width=0.2\textwidth, file=data_linear_fast2.eps}
+\epsfig{width=0.2\textwidth, file=data_linear_fast3.eps}
+
+\epsfig{width=0.2\textwidth, file=data_linear_evals1.eps}
+\epsfig{width=0.2\textwidth, file=data_linear_evals2.eps}
+\epsfig{width=0.2\textwidth, file=data_linear_evals3.eps}
 
@@
 
-In the second example, our data will be warped into ``half-moon" shapes via the function
+In the second example, our data from Section~\ref{subsec:linear_example} will be warped into ``half-moon" shapes via the function
 \begin{equation} \label{eq:nonlinear_function}
@@
 %
-It is important to note that this is a difficult class of problem in practice, as none of the observed
+This is a difficult class of problem in practice, as none of the observed
 variables are purely fast or slow, and the observed system appears, at first inspection, to possess no separation
